### 2021-05-27 paper/paper-misspec-supp.tex
init paper

%% LyX 1.3 created this file.  For more info, see http://www.lyx.org/.
%% Do not edit unless you really know what you are doing.
\documentclass[english, 12pt]{article}
\usepackage{times}
%\usepackage{algorithm2e}
\usepackage{url}
\usepackage{bbm}
\usepackage[T1]{fontenc}
\usepackage[latin1]{inputenc}
\usepackage{geometry}
\geometry{verbose,letterpaper,tmargin=2cm,bmargin=2cm,lmargin=1.5cm,rmargin=1.5cm}
\usepackage{rotating}
\usepackage{color}
\usepackage{graphicx}
\usepackage{amsmath, amsthm, amssymb}
\usepackage{setspace}
\usepackage{lineno}
\usepackage{hyperref}
\usepackage{bbm}
\usepackage{makecell}
\usepackage{placeins}
\usepackage{subcaption}

%\renewcommand{\arraystretch}{1.8}

%\linenumbers
%\doublespacing
\onehalfspacing
%\usepackage[authoryear]{natbib}
\usepackage{natbib} \bibpunct{(}{)}{;}{author-year}{}{,}

%Pour les rajouts
\usepackage{color}
\definecolor{trustcolor}{rgb}{0,0,1}

\usepackage{dsfont}
\usepackage[warn]{textcomp}
\usepackage{adjustbox}
\usepackage{multirow}
\usepackage{subcaption}
\usepackage{graphicx}
\graphicspath{{../figures/}}
\DeclareMathOperator*{\argmin}{\arg\!\min}

\let\tabbeg\tabular
\let\tabend\endtabular
\renewenvironment{tabular}{\begin{adjustbox}{max width=0.95\textwidth}\tabbeg}{\tabend\end{adjustbox}}

\makeatletter

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% LyX specific LaTeX commands.
%% Bold symbol macro for standard LaTeX users
%\newcommand{\boldsymbol}[1]{\mbox{\boldmath $#1$}}

%% Because html converters don't know tabularnewline
\providecommand{\tabularnewline}{\\}
\renewcommand*{\arraystretch}{1.2}

\usepackage{babel}
\makeatother

\begin{document}

\renewcommand{\thefigure}{S\arabic{figure}}
\setcounter{figure}{0}
\renewcommand{\thetable}{S\arabic{table}}
\setcounter{table}{0}
\renewcommand{\theequation}{S\arabic{equation}}
\setcounter{equation}{0}

\section*{Supplementary Materials}

\vspace*{5em}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\clearpage

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\FloatBarrier
%\clearpage

\bibliographystyle{natbib}
\bibliography{refs}

\end{document}

### 2021-05-27 paper/paper-misspec.tex
init paper

%% LyX 1.3 created this file.  For more info, see http://www.lyx.org/.
%% Do not edit unless you really know what you are doing.
\documentclass[english, 12pt]{article}
\usepackage{times}
%\usepackage{algorithm2e}
\usepackage{url}
\usepackage{bbm}
\usepackage[T1]{fontenc}
\usepackage[latin1]{inputenc}
\usepackage{geometry}
\geometry{verbose,letterpaper,tmargin=2cm,bmargin=2cm,lmargin=1.5cm,rmargin=1.5cm}
\usepackage{rotating}
\usepackage{color}
\usepackage{graphicx}
\usepackage{amsmath, amsthm, amssymb}
\usepackage{setspace}
\usepackage{lineno}
\usepackage{hyperref}
\usepackage{bbm}
\usepackage{makecell}

\renewcommand{\arraystretch}{1.3}

\usepackage{xr}
\externaldocument{paper-misspec-supp}

%\linenumbers
%\doublespacing
\onehalfspacing
%\usepackage[authoryear]{natbib}
\usepackage{natbib} \bibpunct{(}{)}{;}{author-year}{}{,}

%Pour les rajouts
\usepackage{color}
\definecolor{trustcolor}{rgb}{0,0,1}

\usepackage{dsfont}
\usepackage[warn]{textcomp}
\usepackage{adjustbox}
\usepackage{multirow}
\usepackage{graphicx}
\graphicspath{{../figures/}}
\DeclareMathOperator*{\argmin}{\arg\!\min}
\usepackage{algorithm}
\usepackage{algpseudocode}

\let\tabbeg\tabular
\let\tabend\endtabular
\renewenvironment{tabular}{\begin{adjustbox}{max width=0.9\textwidth}\tabbeg}{\tabend\end{adjustbox}}

\makeatletter

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% LyX specific LaTeX commands.
%% Bold symbol macro for standard LaTeX users
%\newcommand{\boldsymbol}[1]{\mbox{\boldmath $#1$}}

%% Because html converters don't know tabularnewline
\providecommand{\tabularnewline}{\\}

\usepackage{babel}
\makeatother

\begin{document}

\title{Overcoming multiple sources of misspecification\\in GWAS summary statistics}
\author{Florian Priv\'e,$^{\text{1,}*}$ Timothy S. H. Mak,$^{\text{2}}$ and Bjarni J. Vilhj\'almsson$^{\text{1,3}}$}

\date{~ }
\maketitle

\noindent$^{\text{\sf 1}}$National Centre for Register-Based Research, Aarhus University, Aarhus, 8210, Denmark. \\
\noindent$^{\text{\sf 2}}$Fano Labs, Hong Kong \\
\noindent$^{\text{\sf 3}}$Bioinformatics Research Centre, Aarhus University, Aarhus, 8000, Denmark. \\
\noindent$^\ast$To whom correspondence should be addressed.\\

\noindent Contact: \url{florian.prive.21@gmail.com}

\vspace*{8em}

\abstract{
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\clearpage

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section*{Introduction}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section*{Results}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section*{Discussion}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section*{Materials and Methods}

\subsection*{New implementation of lassosum in bigsnpr}

From \cite{mak2017polygenic}, the solution from lassosum can be obtained by iteratively updating 
\[
\beta_j^{(t)} =
\begin{cases}
\text{sign}\left(u_j^{(t)}\right) \left(\left|u_j^{(t)}\right| - \lambda\right) / \left(\widetilde{X}_j^T \widetilde{X}_j + s\right) & \text{if } \left|u_j^{(t)}\right| > \lambda ~, \\
0 & \text{otherwise.}
\end{cases}
\]
where 
\[
u_j^{(t)} = r_j - \widetilde{X}_j^T \left( \widetilde{X} \beta^{(t-1)} - \widetilde{X}_j \beta_j^{(t-1)} \right) ~.
\]
Following notations from \cite{prive2020ldpred2} and denoting $\widetilde{X} = \sqrt{\frac{1-s}{n-1}} C_n G S^{-1}$, when $G$ is the genotype matrix, $C_n$ is the centering matrix and $S$ is the diagonal matrix of standard deviations of $G$, then $\widetilde{X}_j^T \widetilde{X} = (1-s) R_{j,.} = (1-s) R_{.,j}^T$ where $R$ is the correlation matrix between variants.
Then $\widetilde{X}_j^T \widetilde{X}_j + s = 1$ and
\[
u_j^{(t)} = \widehat{\beta}_j + (1-s) \left( \beta_j^{(t-1)} - R_{.,j}^T \beta^{(t-1)} \right) ~,
\]
where $r_j = \widehat{\beta}_j =  \dfrac{\widehat{\gamma}_j}{\sqrt{n_j ~ \text{se}(\widehat{\gamma}_j)^2 + \widehat{\gamma}_j^2}}$ and $\widehat{\gamma}_j$ is the GWAS effect of variant $j$ and $n$ is the GWAS sample size \cite[]{mak2017polygenic,prive2021high}.
Then the most time-consuming part is computing $R_{.,j}^T \beta^{(t-1)}$.
To make this faster, instead of computing $R_{.,j}^T \beta^{(t-1)}$ at each iteration ($j$ and $t$), we can start with a vector with only 0s initially (for all $j$) since $\beta^{(0)} \equiv 0$, and then updating this vector when $\beta_j^{(t)} \neq \beta_j^{(t-1)}$ only. Note that only positions $k$ for which $R_{k,j} \neq 0$ must be updated in this vector $R_{.,j}^T \beta^{(t-1)}$.

In this new implementation of the lassosum model, the input parameters are the correlation matrix $R$, the GWAS summary statistics ($\widehat{\gamma}_j$, $\text{se}(\widehat{\gamma}_j)$ and $n_j$), and the two hyper-parameters $\lambda$ and $s$. 
Therefore, except for the hyper-parameters, this is the exact same input as for LDpred2 \cite[]{prive2020ldpred2}.
For $s$, we now try $\{0.1, 0.2, \dots, 1.0\}$ by default, instead of only $\{0.2, 0.5, 0.8, 1.0\}$.
For $\lambda$, the default in lassosum uses a sequence of 20 values equally spaced on a log scale between 0.1 and 0.001; we now use a sequence between $\lambda_0$ and $\lambda_0 / 100$ by default in lassosum2, where $\lambda_0 = \max_j \left|\widehat{\beta}_j\right|$ is the minimum $\lambda$ for which no variable enter the model because the L1-regularization is too strong.
Note that we do not provide an ``auto'' version using pseudo-validation (as in \cite{mak2017polygenic}) as we have not found it to be very robust (Figures \ref{fig:auc} and \ref{fig:pseudoval}).
Also note that, as in LDpred2, we run lassosum2 genome-wide using a sparse correlation matrix which assumes that variants further away than 3 cM are not correlated, and therefore we do not require splitting the genome into independent LD blocks anymore (as in lassosum).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\clearpage
%\vspace*{5em}

\section*{Code and results availability}

All code used for this paper is available at %\url{https://github.com/privefl/paper-lassosum2/tree/master/code}.
Latest versions of R package bigsnpr can be installed from GitHub.
A tutorial on running lassosum2 along with LDpred2 using R package bigsnpr is available at \url{https://privefl.github.io/bigsnpr-extdoc/polygenic-scores-pgs.html}.
We have extensively used R packages bigstatsr and bigsnpr \cite[]{prive2017efficient} for analyzing large genetic data, packages from the future framework \cite[]{bengtsson2020unifying} for easy scheduling and parallelization of analyses on the HPC cluster, and packages from the tidyverse suite \cite[]{wickham2019welcome} for shaping and visualizing results.

\section*{Acknowledgements}

Authors thank GenomeDK and Aarhus University for providing computational resources and support that contributed to these research results.
This research has been conducted using the UK Biobank Resource under Application Number 58024.

\section*{Funding}

F.P.\ and B.J.V.\ are supported by the Danish National Research Foundation (Niels Bohr Professorship to Prof. John McGrath).
B.J.V.\ is also supported by a Lundbeck Foundation Fellowship (R335-2019-2339).

\section*{Declaration of Interests}

The authors declare no competing interests.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\clearpage

\bibliographystyle{natbib}
\bibliography{refs}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\end{document}

### 2021-05-27 paper/paper-misspec.tex
change with delta

@@
 
+\subsection{Misspecification of GWAS sample sizes}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-From \cite{mak2017polygenic}, the solution from lassosum can be obtained by iteratively updating 
+Instead of using a regularized version of the correlation matrix $R$ parameterized by $s$, $R_s = (1 - s) R + s I$, we use $R_{\delta} = R + \delta I$, which makes it clearer that lassosum is also using L2-regularization (therefore elastic-net).
+Then, from \cite{mak2017polygenic}, the solution from lassosum can be obtained by iteratively updating 
 \[
@@
 \begin{cases}
-\text{sign}\left(u_j^{(t)}\right) \left(\left|u_j^{(t)}\right| - \lambda\right) / \left(\widetilde{X}_j^T \widetilde{X}_j + s\right) & \text{if } \left|u_j^{(t)}\right| > \lambda ~, \\
+\text{sign}\left(u_j^{(t)}\right) \left(\left|u_j^{(t)}\right| - \lambda\right) / \left(\widetilde{X}_j^T \widetilde{X}_j + \delta\right) & \text{if } \left|u_j^{(t)}\right| > \lambda ~, \\
 0 & \text{otherwise.}
@@
 \]
-Following notations from \cite{prive2020ldpred2} and denoting $\widetilde{X} = \sqrt{\frac{1-s}{n-1}} C_n G S^{-1}$, when $G$ is the genotype matrix, $C_n$ is the centering matrix and $S$ is the diagonal matrix of standard deviations of $G$, then $\widetilde{X}_j^T \widetilde{X} = (1-s) R_{j,.} = (1-s) R_{.,j}^T$ where $R$ is the correlation matrix between variants.
-Then $\widetilde{X}_j^T \widetilde{X}_j + s = 1$ and
+Following notations from \cite{prive2020ldpred2} and denoting $\widetilde{X} = \frac{1}{\sqrt{n-1}} C_n G S^{-1}$, where $G$ is the genotype matrix, $C_n$ is the centering matrix and $S$ is the diagonal matrix of standard deviations of the columns of $G$.
+Then $\widetilde{X}_j^T \widetilde{X} = R_{j,.} = R_{.,j}^T$, $\widetilde{X}_j^T \widetilde{X}_j = 1$ and
 \[
-u_j^{(t)} = \widehat{\beta}_j + (1-s) \left( \beta_j^{(t-1)} - R_{.,j}^T \beta^{(t-1)} \right) ~,
+u_j^{(t)} = \beta_j^{(t-1)} + \widehat{\beta}_j - R_{.,j}^T \beta^{(t-1)} ~,
 \]
@@
 
-In this new implementation of the lassosum model, the input parameters are the correlation matrix $R$, the GWAS summary statistics ($\widehat{\gamma}_j$, $\text{se}(\widehat{\gamma}_j)$ and $n_j$), and the two hyper-parameters $\lambda$ and $s$.
+In this new implementation of the lassosum model, the input parameters are the correlation matrix $R$, the GWAS summary statistics ($\widehat{\gamma}_j$, $\text{se}(\widehat{\gamma}_j)$ and $n_j$), and the two hyper-parameters $\lambda$ and $\delta$.
 Therefore, except for the hyper-parameters, this is the exact same input as for LDpred2 \cite[]{prive2020ldpred2}.
-For $s$, we now try $\{0.1, 0.2, \dots, 1.0\}$ by default, instead of only $\{0.2, 0.5, 0.8, 1.0\}$.
+We try $\delta \in \{0.001, 0.005, 0.02, 0.1, 0.6, 3\}$ by default in lassosum2, instead of $s \in \{0.2, 0.5, 0.8, 1.0\}$ in lassosum.
 For $\lambda$, the default in lassosum uses a sequence of 20 values equally spaced on a log scale between 0.1 and 0.001; we now use a sequence between $\lambda_0$ and $\lambda_0 / 100$ by default in lassosum2, where $\lambda_0 = \max_j \left|\widehat{\beta}_j\right|$ is the minimum $\lambda$ for which no variable enter the model because the L1-regularization is too strong.
-Note that we do not provide an ``auto'' version using pseudo-validation (as in \cite{mak2017polygenic}) as we have not found it to be very robust (Figures \ref{fig:auc} and \ref{fig:pseudoval}).
-Also note that, as in LDpred2, we run lassosum2 genome-wide using a sparse correlation matrix which assumes that variants further away than 3 cM are not correlated, and therefore we do not require splitting the genome into independent LD blocks anymore (as in lassosum).
+Note that we do not provide an ``auto'' version using pseudo-validation (as in \cite{mak2017polygenic}) as we have not found it to be very robust (Figure \ref{fig:pseudoval} [TODO: REGENERATE THIS FIG AND CODE]).
+Also note that, as in LDpred2, we run lassosum2 genome-wide using a sparse correlation matrix which assumes that variants further away than 3 cM are not correlated, and therefore we do not require splitting the genome into independent LD blocks anymore (as is required in lassosum).
 

### 2021-05-27 paper/paper-misspec.tex
describe data

@@
 
-\subsection{Misspecification of GWAS sample sizes}
+\subsection*{Misspecification of GWAS sample sizes}
 
@@
 \section*{Materials and Methods}
+
+\subsection*{Data for simulations}
+
+We use the UK Biobank imputed (BGEN) data \cite[]{bycroft2018uk}. 
+We restrict individuals to the ones used for computing the principal components (PCs) in the UK Biobank (Field 22020).
+These individuals are unrelated and have passed some quality control including removing samples with a missing rate on autosomes larger than 0.02, having a mismatch between inferred sex and self-reported sex, and outliers based on heterozygosity (more details can be found in section S3 of \cite{bycroft2018uk}).
+To get a set of genetically homogeneous individuals, we compute a robust Mahalanobis distance based on the first 16 PCs and further restrict individuals to those within a log-distance of 5 \cite[]{prive2020efficient}.
+This results in 362,307 individuals.
+We sample 300,000 individuals to form a training set (e.g.\ to run GWAS), 10,000 individuals to form a validation set (to tune hyper-parameters), and use the remaining 52,307 individuals to form a test set (to evaluate final predictive models).
+
+Among genetic variants with a minor allele frequency larger than 0.01 and an imputation INFO score larger than 0.3, we sample 20,000 of them according to the inverse of the INFO score density so that they have varying levels of imputation accuracy (Figure \ref{fig:hist-info}).
+We read the UK Biobank data into two different datasets using function \texttt{snp\_readBGEN} from R package bigsnpr \cite[]{prive2017efficient}, one by reading the BGEN data at random according to imputation probabilities, and another one reading it as dosages (i.e.\ expected values according to imputation probabilities).
+The first dataset is used as what could be the real genotype calls and the second dataset as what would be its imputed version; this technique was used in \cite{prive2019making}.
+
+\subsection*{Data for real analyses}
+
+We also use the UK Biobank data, and use the same individuals as described in the previous subsection.
+We sample 10,000 individuals to form a validation set and use the remaining 352,307 individuals as test set.
+We restrict to the genetic variants to the 1,054,315 variants used in the LD reference provided in \cite{prive2020ldpred2}.
 
@@
 We try $\delta \in \{0.001, 0.005, 0.02, 0.1, 0.6, 3\}$ by default in lassosum2, instead of $s \in \{0.2, 0.5, 0.8, 1.0\}$ in lassosum.
-For $\lambda$, the default in lassosum uses a sequence of 20 values equally spaced on a log scale between 0.1 and 0.001; we now use a sequence between $\lambda_0$ and $\lambda_0 / 100$ by default in lassosum2, where $\lambda_0 = \max_j \left|\widehat{\beta}_j\right|$ is the minimum $\lambda$ for which no variable enter the model because the L1-regularization is too strong.
+For $\lambda$, the default in lassosum uses a sequence of 20 values equally spaced on a log scale between 0.1 and 0.001; we now use a sequence between $\lambda_0$ and $\lambda_0 / 100$ by default in lassosum2, where $\lambda_0 = \max_j \left|\widehat{\beta}_j\right|$ is the minimum $\lambda$ for which no variable enters the model because the L1-regularization is too strong.
 Note that we do not provide an ``auto'' version using pseudo-validation (as in \cite{mak2017polygenic}) as we have not found it to be very robust (Figure \ref{fig:pseudoval} [TODO: REGENERATE THIS FIG AND CODE]).

### 2021-05-28 paper/paper-misspec-supp.tex
describe data
add simu misN

@@
 
+\begin{figure}[h]
+	\centerline{\includegraphics[width=0.8\textwidth]{simu-qc-plot}}
+	\caption{Quality control plot, as proposed in \cite{prive2020ldpred2}, for the simulations with sample size misspecification.
+The standard deviations are derived from the summary statistics assuming the same global GWAS sample size for all variants (300,000).}
+	\label{fig:qcplot}
+\end{figure}
+
+\begin{figure}[h]
+	\centerline{\includegraphics[width=0.95\textwidth]{simu-misN}}
+	\caption{Results for the simulations with sample size misspecification.
+The GWAS sample size is ``true" when providing the true per-variant sample size, ``max'' when providing instead the maximum sample size as a unique value to be used for all variants,  ``imputed'' (see Methods in main text), or ``any'' when the method does not use this information (the case for C+T).}
+	\label{fig:simu-misN}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{figure}[htb]
+	\centerline{\includegraphics[width=0.8\textwidth]{simu_hist_info}}
+	\caption{Histogram of the imputation INFO scores of the 20,000 variants from chromosome 22 of the UK Biobank data used in the simulations. \label{fig:hist-info}}
+\end{figure}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
+%\begin{figure}[h]
+%	\centerline{\includegraphics[width=0.85\textwidth]{pseudoval}}
+%	\caption{For one simulation, scores from validation vs.\ scores from pseudo-validation as described in \cite{mak2017polygenic} using either correlations (the default in lassosum) or p-values for computing local false discovery rates.}
+%	\label{fig:pseudoval}
+%\end{figure}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \FloatBarrier
-%\clearpage
+\clearpage
 

### 2021-05-28 paper/paper-misspec.tex
add simu misN

@@
 
+We design simulations where variants have different GWAS sample sizes, which is often the case when meta-analyzing GWAS from multiple cohorts without the same genome coverage.
+Using 20,000 variants from chromosome 22 (Methods), we simulate quantitative phenotypes assuming an heritability of 20\% and 2000 causal variants.
+For half of the variants, we use 100\% of 300,000 individuals for GWAS, but use only 80\% for one quarter of them and 60\% for the remaining quarter. 
+We then run C+T, LDpred2-inf, LDpred2(-grid), LDpred2-auto, lassosum, lassosum2, PRS-CS and SBayesR by using either the true per-variant GWAS sample size, or the total sample size. 
+We do not show results for SBayesR because it always diverged and for PRS-CS because the overlap with the LD reference they provide is too small.
+Averaged over 10 simulations, when providing true per-variant GWAS sample sizes, squared correlations (in \%) between the polygenic scores and the simulated phenotypes are of [TODO: COMPLETE] (Figure \ref{fig:simu-misN}).
+Results for LDpred2-auto are the same as with LDpred2-inf, as it always converges to an infinitesimal model (p = 1) in these simulations.
+Note that C+T does not use this sample size information.
+For methods that use the per-variant GWAS sample sizes, predictive performance slightly decreases [TODO: REPLACE NUMBERS] from 17.5 to 17.4 for lassosum, from 18.0 to 17.5 for lassosum2, but dramatically decreases for LDpred2 with new values of 11.5 for LDpred2-grid, 5.7 for LDpred2-auto and 6.8 for LDpred2-inf (Figure \ref{fig:simu-misN}).
+Therefore, this extreme simulation scenario shows that LDpred2 can be sensitive to GWAS sample size misspecification, whereas lassosum (and lassosum2) seems little affected by this.
+
+[TODO: COMPLETE WITH LOOK AT EACH OF THEM]
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2021-05-30 paper/paper-misspec-supp.tex
add investigate mis N

@@
 
+\FloatBarrier
+
 \begin{figure}[h]
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\FloatBarrier
+
+\begin{figure}[p]
+	\begin{subfigure}{\textwidth}
+		\centering
+		\includegraphics[width=.8\linewidth]{lassosum2-mis-maxN}
+	\end{subfigure}\vspace{2em}
+	\begin{subfigure}{\textwidth}
+		\centering
+		\includegraphics[width=.8\linewidth]{lassosum2-mis-imputeN}
+	\end{subfigure}
+	\caption{Multiple results (squared correlation r2 between polygenic score and phenotype) over a grid of parameters ($\lambda$ and $\delta$) for lassosum2 in one of the simulations with misspecified GWAS sample sizes (N).
+The top subfigure is using the maximum N, whereas the bottom subfigure is using the imputed N (cf.\ figure \ref{fig:simu-misN}).
+Missing points represent models that were detected as divergent.}
+	\label{fig:lassosum2-misN}
+\end{figure}
+
+\begin{figure}[p]
+	\begin{subfigure}{\textwidth}
+		\centering
+		\includegraphics[width=.8\linewidth]{ldpred2-mis-maxN}
+	\end{subfigure}\vspace{2em}
+	\begin{subfigure}{\textwidth}
+		\centering
+		\includegraphics[width=.8\linewidth]{ldpred2-mis-imputeN}
+	\end{subfigure}
+	\caption{Multiple results (squared correlation r2 between polygenic score and phenotype) over a grid of parameters (the SNP heritability h2 and the proportion of causal variants p) for LDpred2(-grid) in one of the simulations with misspecified GWAS sample sizes (N).
+The top subfigure is using the maximum N, whereas the bottom subfigure is using the imputed N (cf.\ figure \ref{fig:simu-misN}).
+Note that the small h2 values (0.002, 0.02, 0.06) were not considered in the results of figure \ref{fig:simu-misN}.
+Missing points represent models that were detected as divergent.}
+	\label{fig:ldpred2-misN}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\FloatBarrier
 

### 2021-05-30 paper/paper-misspec.tex
add investigate mis N

@@
 Averaged over 10 simulations, when providing true per-variant GWAS sample sizes, squared correlations (in \%) between the polygenic scores and the simulated phenotypes are of [TODO: COMPLETE] (Figure \ref{fig:simu-misN}).
-Results for LDpred2-auto are the same as with LDpred2-inf, as it always converges to an infinitesimal model (p = 1) in these simulations.
 Note that C+T does not use this sample size information.
@@
 
-[TODO: COMPLETE WITH LOOK AT EACH OF THEM]
+We then conduct further investigations to explain results of figure \ref{fig:simu-misN}.
+First, the reason why results for LDpred2-auto are the same as with LDpred2-inf is because it always converges to an infinitesimal model (p = 1) in these simulations.
+Second, for lassosum2, results for a grid of parameters (over $\lambda$ and $\delta$) and quite smooth compared to LDpred2 (Figures \ref{fig:lassosum2-misN} and \ref{fig:ldpred2-misN}).
+Results slightly improve globally for both lassosum2 and LDpred2 when using imputed instead of maximum GWAS sample sizes.
+Finally, in these simulations with misspecified sample sizes, it seems highly beneficial to use a small value for the SNP heritability hyper-parameter in LDpred2, e.g.\ a value of 0.02 or even 0.002 when the true value is 0.2 (Figure \ref{fig:ldpred2-misN}).
+We recall that using a small value for this hyper-parameter induces a larger regularization on the effect sizes, and this benefits LDpred2 even when using imputed sample sizes.
 
@@
 
-Among genetic variants with a minor allele frequency larger than 0.01 and an imputation INFO score larger than 0.3, we sample 20,000 of them according to the inverse of the INFO score density so that they have varying levels of imputation accuracy (Figure \ref{fig:hist-info}).
+Among genetic variants with a minor allele frequency larger than 0.01 and an imputation INFO score larger than 0.4, we sample 40,000 of them according to the inverse of the INFO score density so that they have varying levels of imputation accuracy (Figure \ref{fig:hist-info}).
 We read the UK Biobank data into two different datasets using function \texttt{snp\_readBGEN} from R package bigsnpr \cite[]{prive2017efficient}, one by reading the BGEN data at random according to imputation probabilities, and another one reading it as dosages (i.e.\ expected values according to imputation probabilities).

### 2021-06-02 paper/paper-misspec-supp.tex
save

@@
 
-\begin{figure}[h]
+\begin{figure}[p]
 	\centerline{\includegraphics[width=0.95\textwidth]{simu-misN}}
@@
 
-\begin{figure}[htb]
+\begin{figure}[p]
+	\centerline{\includegraphics[width=0.8\textwidth]{compare-sd}}
+	\caption{Standard deviations from ``true'' genotype calls versus from imputed dosages, colored by INFO scores. \label{fig:compare-sd}}
+\end{figure}
+
+\begin{figure}[p]
+	\centerline{\includegraphics[width=0.8\textwidth]{compare-beta}}
+	\caption{GWAS effect sizes from ``true'' genotype calls versus from imputed dosages, colored by INFO scores. \label{fig:compare-beta}}
+\end{figure}
+
+\begin{figure}[p]
+	\centerline{\includegraphics[width=0.8\textwidth]{compare-beta-se}}
+	\caption{Standard errors of GWAS effect sizes from ``true'' genotype calls versus from imputed dosages, colored by INFO scores. \label{fig:compare-beta-se}}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{figure}[p]
+	\centerline{\includegraphics[width=0.8\textwidth]{compare-info}}
+	\caption{INFO scores reported in the UK Biobank vs.\ INFO scores recomputed from the subset of 362,307 European individuals used in this paper. \label{fig:compare-info}}
+\end{figure}
+
+\begin{figure}[p]
+	\centerline{\includegraphics[width=0.95\textwidth]{simu-info}}
+	\caption{Results for the simulations using dosage data.
+The GWAS sample size is ``true" when providing the true per-variant sample size, ``max'' when providing instead the maximum sample size as a unique value to be used for all variants,  ``imputed'' (see Methods in main text), or ``any'' when the method does not use this information (the case for C+T).}
+	\label{fig:simu-info}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\FloatBarrier
+
+\begin{figure}[p]
 	\centerline{\includegraphics[width=0.8\textwidth]{simu_hist_info}}
-	\caption{Histogram of the imputation INFO scores of the 20,000 variants from chromosome 22 of the UK Biobank data used in the simulations. \label{fig:hist-info}}
+	\caption{Histogram of the imputation INFO scores of the 40,000 variants from chromosome 22 of the UK Biobank data used in the simulations. \label{fig:hist-info}}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{figure}[p]
+	\centerline{\includegraphics[width=0.8\textwidth]{compare-mult-imp}}
+	\caption{GWAS effect sizes and Z-scores from imputed dosages versus from multiple imputation (MI, 10 draws used), ``corrected'' and colored by INFO scores.
+This shows only one tenth of the simulated variants, because MI takes some time to run.
+\label{fig:compare-mi}}
 \end{figure}
@@
 
-%\begin{figure}[h]
-%	\centerline{\includegraphics[width=0.85\textwidth]{pseudoval}}
-%	\caption{For one simulation, scores from validation vs.\ scores from pseudo-validation as described in \cite{mak2017polygenic} using either correlations (the default in lassosum) or p-values for computing local false discovery rates.}
-%	\label{fig:pseudoval}
-%\end{figure}
+\FloatBarrier
+
+\begin{figure}[p]
+	\centerline{\includegraphics[width=0.85\textwidth]{pseudoval}}
+	\caption{For one simulation, scores from validation vs.\ scores from pseudo-validation as described in \cite{mak2017polygenic} using either correlations (the default in lassosum) or p-values when computing local false discovery rates.}
+	\label{fig:pseudoval}
+\end{figure}
 

### 2021-06-02 paper/paper-misspec.tex
save

@@
 We design simulations where variants have different GWAS sample sizes, which is often the case when meta-analyzing GWAS from multiple cohorts without the same genome coverage.
-Using 20,000 variants from chromosome 22 (Methods), we simulate quantitative phenotypes assuming an heritability of 20\% and 2000 causal variants.
+Using 40,000 variants from chromosome 22 (Methods), we simulate quantitative phenotypes assuming an heritability of 20\% and 2000 causal variants.
 For half of the variants, we use 100\% of 300,000 individuals for GWAS, but use only 80\% for one quarter of them and 60\% for the remaining quarter. 
@@
 We recall that using a small value for this hyper-parameter induces a larger regularization on the effect sizes, and this benefits LDpred2 even when using imputed sample sizes.
+
+\subsection*{GWAS using expected genotypes from imputation}
+
+When using the expected genotypes from imputation (dosages), their standard deviations are usually lower than the expected value under Hardy-Weinberg equilibrium (cf.\ Methods section ``Standard deviations and imputation measures'').
+In simulations (cf.\ Methods section ``Data for simulations''), we show that $\text{sd}(G_j)^{true} \approx \text{sd}(G_j)^{imp} / \sqrt{\text{INFO}_j}$, $\text{se}(\hat{\beta}_j)^{true} \approx \text{se}(\hat{\beta}_j)^{imp} \cdot \sqrt{\text{INFO}_j}$, and $\hat{\beta}_j^{true} \approx \hat{\beta}_j^{imp} \cdot \sqrt{\text{INFO}_j}$ (Figures \ref{fig:compare-sd}, \ref{fig:compare-beta}, and \ref{fig:compare-beta-se}).
+This is the first correction of summary statistics we try in the simulations below. 
+Note that we recompute INFO scores for the subset of 362,307 European individuals used in this paper as they sometimes differ from the ones reported by the UK Biobank (Figure \ref{fig:compare-info}).
+Instead of using dosages to perform the GWAS, it has been argued that using multiple imputation (MI) would be more appropriate \cite[]{Palmer2016}.
+In simulations, we show that $Z_j^{MI} \approx Z_j^{imp} \cdot \text{INFO}_j$ and $\hat{\beta}_j^{MI} \approx \hat{\beta}_j^{imp} \cdot \text{INFO}_j$ (Figure \ref{fig:compare-mi}).
+This is the second correction of summary statistics we try in the simulations below, along with $n_j = N \cdot \text{INFO}_j$.
+
+Using 40,000 variants from chromosome 22, we simulate quantitative phenotypes assuming an heritability of 20\% and 2000 causal variants using the ``true'' dataset (cf.\ Methods section ``Data for simulations'').
+We run GWAS on the dosage dataset and use these summary statistics to run LDpred2 and lassosum2 with either no correction of the summary statistics, or with one of the two described above.
+The LD reference used by LDpred2 and lassosum2 is computed from the validation set using the ``true'' dataset.
 
@@
 
+\subsection*{GWAS sample size imputation}
+
+In this paper, we extensively use the following formula
+\begin{equation}\label{eq:approx-sd-lin}
+\text{sd}(G_j) \approx \dfrac{\text{sd}(y)}{\sqrt{n_j ~ \text{se}(\hat{\beta}_j)^2 + \hat{\beta}_j^2}} ~,
+\end{equation}
+where $\hat{\beta}_j$ is the marginal (GWAS) effect of variant $j$, $n_j$ is the GWAS sample size associated with variant $j$, $y$ is the vector of phenotypes and ${G}_j$ is the vector of genotypes for variant $j$.
+This formula is used in LDpred2 \cite{prive2020ldpred2,prive2021high}. Note that, for a binary trait for which logistic regression is used, we have instead
+\begin{equation}\label{eq:approx-sd-log}
+\text{sd}(G_j) \approx \dfrac{2}{\sqrt{n_j^\text{eff} ~ \text{se}(\hat{\beta}_j)^2 + \hat{\beta}_j^2}} ~,
+\end{equation}
+where $n_j^\text{eff} = \frac{4}{1 / n_j^\text{cases} + 1 / n_j^\text{controls}}$.
+
+We can then impute $n_j$ from equation \eqref{eq:approx-sd-lin} using
+\begin{equation}\label{eq:imputeN}
+n_j \approx \dfrac{\text{var}(y) / \text{var}(G_j) - \hat{\beta}_j^2}{\text{se}(\hat{\beta}_j)^2} ~.
+\end{equation}
+In practice, we also bound this estimate to be between $0.5 \cdot N$ and $1.1 \cdot N$, where $N$ is the global sample size.
+
+\subsection*{Standard deviations and imputation measures}
+
+The MACH $\hat{r}_j^2$ measure for variant $j$ is given by $\hat{r}_j^2 = \frac{var(G_j)}{2 \hat{\theta} (1 - \hat{\theta})}$, where $\hat{\theta}$ is the estimated allele frequency \cite[]{marchini2010genotype}.
+This measure is also highly concordant with the IMPUTE INFO measure \cite[]{marchini2010genotype}.
+Therefore $\text{INFO}_j \approx \frac{var(G_j)}{2 \hat{\theta} (1 - \hat{\theta})}$.
+
 \subsection*{New implementation of lassosum in bigsnpr}
@@
 \]
-where $r_j = \widehat{\beta}_j =  \dfrac{\widehat{\gamma}_j}{\sqrt{n_j ~ \text{se}(\widehat{\gamma}_j)^2 + \widehat{\gamma}_j^2}}$ and $\widehat{\gamma}_j$ is the GWAS effect of variant $j$ and $n$ is the GWAS sample size \cite[]{mak2017polygenic,prive2021high}.
+where $r_j = \widehat{\beta}_j =  \dfrac{\widehat{\beta}_j}{\sqrt{n_j ~ \text{se}(\widehat{\beta}_j)^2 + \widehat{\beta}_j^2}}$ and $\widehat{\beta}_j$ is the GWAS effect of variant $j$ and $n$ is the GWAS sample size \cite[]{mak2017polygenic,prive2021high}.
 Then the most time-consuming part is computing $R_{.,j}^T \beta^{(t-1)}$.
@@
 
-In this new implementation of the lassosum model, the input parameters are the correlation matrix $R$, the GWAS summary statistics ($\widehat{\gamma}_j$, $\text{se}(\widehat{\gamma}_j)$ and $n_j$), and the two hyper-parameters $\lambda$ and $\delta$.
+In this new implementation of the lassosum model, the input parameters are the correlation matrix $R$, the GWAS summary statistics ($\widehat{\beta}_j$, $\text{se}(\widehat{\beta}_j)$ and $n_j$), and the two hyper-parameters $\lambda$ and $\delta$.
 Therefore, except for the hyper-parameters, this is the exact same input as for LDpred2 \cite[]{prive2020ldpred2}.
@@
 For $\lambda$, the default in lassosum uses a sequence of 20 values equally spaced on a log scale between 0.1 and 0.001; we now use a sequence between $\lambda_0$ and $\lambda_0 / 100$ by default in lassosum2, where $\lambda_0 = \max_j \left|\widehat{\beta}_j\right|$ is the minimum $\lambda$ for which no variable enters the model because the L1-regularization is too strong.
-Note that we do not provide an ``auto'' version using pseudo-validation (as in \cite{mak2017polygenic}) as we have not found it to be very robust (Figure \ref{fig:pseudoval} [TODO: REGENERATE THIS FIG AND CODE]).
+Note that we do not provide an ``auto'' version using pseudo-validation (as in \cite{mak2017polygenic}) as we have not found it to be very robust (Figure \ref{fig:pseudoval}).
 Also note that, as in LDpred2, we run lassosum2 genome-wide using a sparse correlation matrix which assumes that variants further away than 3 cM are not correlated, and therefore we do not require splitting the genome into independent LD blocks anymore (as is required in lassosum).

### 2021-06-04 paper/paper-misspec-supp.tex
add brca analysis

@@
 	\caption{Results for the simulations with sample size misspecification.
-The GWAS sample size is ``true" when providing the true per-variant sample size, ``max'' when providing instead the maximum sample size as a unique value to be used for all variants,  ``imputed'' (see Methods in main text), or ``any'' when the method does not use this information (the case for C+T).}
+The GWAS sample size is ``true'' when providing the true per-variant sample size, ``max'' when providing instead the maximum sample size as a unique value to be used for all variants,  ``imputed'' (see Methods in main text), or ``any'' when the method does not use this information (the case for C+T).}
 	\label{fig:simu-misN}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\FloatBarrier
+
 \begin{figure}[p]
-	\centerline{\includegraphics[width=0.95\textwidth]{simu-info}}
-	\caption{Results for the simulations using dosage data.
-The GWAS sample size is ``true" when providing the true per-variant sample size, ``max'' when providing instead the maximum sample size as a unique value to be used for all variants,  ``imputed'' (see Methods in main text), or ``any'' when the method does not use this information (the case for C+T).}
-	\label{fig:simu-info}
+	\centerline{\includegraphics[width=0.8\textwidth]{compare-mult-imp}}
+	\caption{GWAS effect sizes and Z-scores from imputed dosages versus from multiple imputation (MI, 10 draws used), ``corrected'' and colored by INFO scores.
+This shows only one tenth of the simulated variants, because MI takes some time to run.
+\label{fig:compare-mi}}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\FloatBarrier
+
+\begin{figure}[p]
+	\centerline{\includegraphics[width=0.8\textwidth]{brca_onco_hist_info}}
+	\caption{Histogram of the imputation INFO scores of the HapMap3 variants from the OncoArray GWAS summary statistics for breast cancer \cite[]{michailidou2017association}. \label{fig:hist-info-brca}}
+\end{figure}
+
+\begin{figure}[p]
+\centerline{\includegraphics[width=0.8\textwidth]{brca_onco_hist_info}}
+\caption{Histogram of the imputation INFO scores of the HapMap3 variants from the iCOGS GWAS summary statistics for breast cancer \cite[]{michailidou2013large}. \label{fig:hist-info-brca2}}
+\end{figure}
+
+\begin{figure}[p]
+	\centerline{\includegraphics[width=0.8\textwidth]{hist_bad_pos}}
+	\caption{Histogram of the positions of outlier variants from Figure 2B [TODO: VERIF NUMBER AT THE END] in the main text. \label{fig:hist-bad-pos}}
+\end{figure}
+
+\begin{figure}[h]
+	\centerline{\includegraphics[width=0.85\textwidth]{gwas_bad_pc}}
+	\caption{Standard deviations inferred from the simulated GWAS summary statistics (\textbf{A:} with no covariate; \textbf{B:} with PC19 from the UK Biobank as covariate) versus the ones inferred from the allele frequencies.
+Only HapMap3 variants from chromosome 6 are represented.}
+	\label{fig:gwas_bad_pc}
+\end{figure}
+
+\begin{figure}[h]
+	\centerline{\includegraphics[width=0.95\textwidth]{brca_icogs_qc}}
+	\caption{Standard deviations inferred from the iCOGS breast cancer GWAS summary statistics versus the ones inferred from the reported GWAS allele frequencies.
+Only 100,000 variants are represented, at random.}
+	\label{fig:brca_icogs_qc}
 \end{figure}
@@
 	\caption{Histogram of the imputation INFO scores of the 40,000 variants from chromosome 22 of the UK Biobank data used in the simulations. \label{fig:hist-info}}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{figure}[p]
-	\centerline{\includegraphics[width=0.8\textwidth]{compare-mult-imp}}
-	\caption{GWAS effect sizes and Z-scores from imputed dosages versus from multiple imputation (MI, 10 draws used), ``corrected'' and colored by INFO scores.
-This shows only one tenth of the simulated variants, because MI takes some time to run.
-\label{fig:compare-mi}}
 \end{figure}

### 2021-06-04 paper/paper-misspec.tex
add brca analysis

@@
 
-\title{Overcoming multiple sources of misspecification\\in GWAS summary statistics}
+\title{Identifying and correcting multiple sources of misspecification\\in GWAS summary statistics}
 \author{Florian Priv\'e,$^{\text{1,}*}$ Timothy S. H. Mak,$^{\text{2}}$ and Bjarni J. Vilhj\'almsson$^{\text{1,3}}$}
@@
 The LD reference used by LDpred2 and lassosum2 is computed from the validation set using the ``true'' dataset.
+For lassosum2 and LDpred2 which tune parameters using validation set and can possibly use strong regularization, the two corrections just slighly improve predictive performance in these simulations (Figure \ref{fig:simu-info}).
+Correction ``sqrt\_info'', which is equivalent to postprocessing PGS effects by $\sqrt{\text{INFO}}$ slighly improves predictive performance for all methods, but correction ``info'' (using $\hat{\beta}_j^{imp} \cdot \text{INFO}_j$  and $N \cdot \text{INFO}_j$) provides much better predictive performance for LDpred2-auto, with a mean squared correlation of 15.4\% compared to 12.9\% with no correction and 13.3\% with correction ``sqrt\_info'' (Figure \ref{fig:simu-info}).
+
+\begin{figure}[h]
+	\centerline{\includegraphics[width=0.95\textwidth]{simu-info}}
+	\caption{Results for the simulations using dosage data.
+Correction ``sqrt\_info'' corresponds to using $\hat{\beta}_j^{imp} \cdot \sqrt{\text{INFO}_j}$ and $\text{se}(\hat{\beta}_j)^{imp} \cdot \sqrt{\text{INFO}_j}$.
+	Correction ``info'' corresponds to using $\hat{\beta}_j^{imp} \cdot \text{INFO}_j$  and $N \cdot \text{INFO}_j$.}
+	\label{fig:simu-info}
+\end{figure}
+
+\subsection*{Application to breast cancer summary statistics}
+
+Breast cancer summary statistics are interesting because they come from two mega analyses [TODO: VERIF] \cite[]{michailidou2013large,michailidou2017association}, which means there is some larger precision in the parameters reported, such as the INFO scores and the sample sizes.
+INFO scores for the OncoArray summary statistics (after having restricted to HapMap3 variants) are globally very good (Figure \ref{fig:hist-info-brca}), but the ones from iCOGS are a bit worse (Figure \ref{fig:hist-info-brca2}), probably because they are older.
+For the OncoArray summary statistics, we first compare the standard deviations inferred from the reported allele frequencies (i.e.\ $\sqrt{2 f (1 - f)}$ where $f$ is the allele frequency, and denoted as sd\_af) versus the ones inferred from the GWAS summary statistics (Equation \eqref{eq:approx-sd-log}, and denoted as sd\_ss).
+When coloring by INFO scores, we see a clear trend with sd\_ss being lower and lower than sd\_af as INFO decreases.
+Using $\text{sd\_ss} / \sqrt{\text{INFO}}$ instead, we get a very good fit with sd\_af, except for some variants of chromosome 6 and 8.
+Most of these outlier variants are in regions 25-33 Mbp of chromosome 6 and 8-12 Mbp of chromosome 8 (Figure \ref{fig:hist-bad-pos}), which are two known long-range LD regions \cite[]{price2008long}.
+We hypothesize that this is due to using principal components (PCs) that capture LD structure as covariates in GWAS \cite[]{prive2020efficient}.
+To validate this hypothesis, we simulate a phenotype using HapMap3 variants of chromosome 6 for 10,000 individuals from the UK Biobank, then we run GWAS with or without PC19 as covariate.
+PC19 from the UK Biobank was previously reported to capture LD structure in region 70-91 Mbp of chromsome 6 \cite[]{prive2020efficient}.
+In these simulations, the same bias as in figure \ref{fig:hist-bad-pos}B is observed for the variants in this region (Figure \ref{fig:gwas_bad_pc}), therefore confirming our hypothesis.
+For the iCOGS summary statistics, we also see a clear trend with sd\_ss being lower and lower than sd\_af as INFO decreases, but they do not have issues with PCs  (Figure \ref{fig:brca_icogs_qc}).
+
+\begin{figure}[h]
+	\centerline{\includegraphics[width=0.95\textwidth]{brca_onco_qc}}
+	\caption{Standard deviations inferred from the OncoArray breast cancer GWAS summary statistics versus the ones inferred from the reported GWAS allele frequencies.
+Only 100,000 variants are represented, at random.}
+	\label{fig:brca_onco_qc}
+\end{figure}
 
@@
 
-Among genetic variants with a minor allele frequency larger than 0.01 and an imputation INFO score larger than 0.4, we sample 40,000 of them according to the inverse of the INFO score density so that they have varying levels of imputation accuracy (Figure \ref{fig:hist-info}).
+Among genetic variants with a minor allele frequency larger than 0.01 and an imputation INFO score larger than 0.4 (as reported in the UK Biobank), we sample 40,000 of them according to the inverse of the INFO score density so that they have varying levels of imputation accuracy (Figure \ref{fig:hist-info}).
 We read the UK Biobank data into two different datasets using function \texttt{snp\_readBGEN} from R package bigsnpr \cite[]{prive2017efficient}, one by reading the BGEN data at random according to imputation probabilities, and another one reading it as dosages (i.e.\ expected values according to imputation probabilities).

### 2021-06-06 paper/paper-misspec.tex
add res brca

@@
 
-Breast cancer summary statistics are interesting because they come from two mega analyses [TODO: VERIF] \cite[]{michailidou2013large,michailidou2017association}, which means there is some larger precision in the parameters reported, such as the INFO scores and the sample sizes.
-INFO scores for the OncoArray summary statistics (after having restricted to HapMap3 variants) are globally very good (Figure \ref{fig:hist-info-brca}), but the ones from iCOGS are a bit worse (Figure \ref{fig:hist-info-brca2}), probably because they are older.
+[TALK ABOUT THE SAMPLE SIZES BEFORE/AFTER QC?]
+
+Breast cancer summary statistics are interesting because they include results from two mega analyses \cite[]{michailidou2013large,michailidou2015genome,michailidou2017association}, which means there is some larger precision in the parameters reported, such as the INFO scores and the sample sizes.
+INFO scores for the OncoArray summary statistics (after having restricted to HapMap3 variants) are globally very good (Figure \ref{fig:hist-info-brca}), but the ones from iCOGS are a bit worse (Figure \ref{fig:hist-info-brca2}), probably because the chip used included around 200K variants only, compared to more than 500K variants for the OncoArray.
 For the OncoArray summary statistics, we first compare the standard deviations inferred from the reported allele frequencies (i.e.\ $\sqrt{2 f (1 - f)}$ where $f$ is the allele frequency, and denoted as sd\_af) versus the ones inferred from the GWAS summary statistics (Equation \eqref{eq:approx-sd-log}, and denoted as sd\_ss).
@@
 
+[ALSO ADD THE GWAS META-ANALYSIS NEFF \textasciitilde20K?]
+
 \begin{figure}[h]
@@
 	\label{fig:brca_onco_qc}
+\end{figure}
+
+Therefore, having the INFO score enables both to correct for it in the QC step to be able to pinpoint possible other problems, and possibly to correct for it and improve prediction as shown in the simulations before.
+We try this for the two breast cancer summary statistics.
+We first compare the standard QC proposed in \cite{prive2020ldpred2} (which ends up filtering on MAF here, which we call ``qc1'').
+We then also filter out the two long-range LD regions of chromosome 6 and 8 for the OncoArray summary statistics and remove around 500 variants when filtering on differences of MAFs between summary statistics and the validation dataset.
+As for the correction for the INFO scores, we use $\hat{\beta}_j^{imp} \cdot \text{INFO}_j$ and $N \cdot \text{INFO}_j$ as we have shown to perform well in simulations.
+Although results for both QC used are very similar, correcting for the INFO score really improves prediction, especially for LDpred2-auto [TODO: NUMBERS] (Figure \ref{fig:res_brca}).
+
+[SHOULD JUST REPORT ONE PREDICTION INSTEAD OF ALL? THEN MAKE A TABLE?]
+
+\begin{figure}[h]
+	\centerline{\includegraphics[width=0.95\textwidth]{res_brca}}
+	\caption{AUC (for \textit{all} models) for predicting breast cancer in the UK Biobank when using either the OncoArray or the iCOGS summary statistics.}
+	\label{fig:res_brca}
 \end{figure}

### 2021-08-02 paper/paper-misspec-supp.tex
save

@@
 
-\begin{figure}[p]
-	\centerline{\includegraphics[width=0.95\textwidth]{simu-misN}}
-	\caption{Results for the simulations with sample size misspecification.
-The GWAS sample size is ``true'' when providing the true per-variant sample size, ``max'' when providing instead the maximum sample size as a unique value to be used for all variants,  ``imputed'' (see Methods in main text), or ``any'' when the method does not use this information (the case for C+T).}
-	\label{fig:simu-misN}
-\end{figure}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 Note that the small h2 values (0.002, 0.02, 0.06) were not considered in the results of figure \ref{fig:simu-misN}.
-Missing points represent models that were detected as divergent.}
+Missing points represent models that were detected as divergent.
+[TODO: REPLACE BY CONTOUR PLOT?
+REMOVE SPARSITY COLORING?]}
 	\label{fig:ldpred2-misN}

### 2021-08-02 paper/paper-misspec.tex
save

@@
 
-\title{Identifying and correcting multiple sources of misspecification\\in GWAS summary statistics}
-\author{Florian Priv\'e,$^{\text{1,}*}$ Timothy S. H. Mak,$^{\text{2}}$ and Bjarni J. Vilhj\'almsson$^{\text{1,3}}$}
+\title{Identifying and correcting multiple sources of misspecification\\in GWAS summary statistics for prediction}
+\author{Florian Priv\'e,$^{\text{1,}*}$ Julyan Arbel,$^{\text{2}}$ Timothy S. H. Mak,$^{\text{3}}$ and Bjarni J. Vilhj\'almsson$^{\text{1,4}}$}
 
@@
 \noindent$^{\text{\sf 1}}$National Centre for Register-Based Research, Aarhus University, Aarhus, 8210, Denmark. \\
-\noindent$^{\text{\sf 2}}$Fano Labs, Hong Kong \\
-\noindent$^{\text{\sf 3}}$Bioinformatics Research Centre, Aarhus University, Aarhus, 8000, Denmark. \\
+\noindent$^{\text{\sf 2}}$Univ.\ Grenoble Alpes, Inria, CNRS, Grenoble INP, LJK, Grenoble, 38000, France. \\
+\noindent$^{\text{\sf 3}}$Fano Labs, Hong Kong. \\
+\noindent$^{\text{\sf 4}}$Bioinformatics Research Centre, Aarhus University, Aarhus, 8000, Denmark. \\
 \noindent$^\ast$To whom correspondence should be addressed.\\
@@
 Note that C+T does not use this sample size information.
-For methods that use the per-variant GWAS sample sizes, predictive performance slightly decreases [TODO: REPLACE NUMBERS] from 17.5 to 17.4 for lassosum, from 18.0 to 17.5 for lassosum2, but dramatically decreases for LDpred2 with new values of 11.5 for LDpred2-grid, 5.7 for LDpred2-auto and 6.8 for LDpred2-inf (Figure \ref{fig:simu-misN}).
+When using the total GWAS sample size instead of the per-variant sample sizes, predictive performance slightly decreases [TODO: REPLACE NUMBERS] from 17.5 to 17.4 for lassosum, from 18.0 to 17.5 for lassosum2, but dramatically decreases for LDpred2 with new values of 11.5 for LDpred2-grid, 5.7 for LDpred2-auto and 6.8 for LDpred2-inf (Figure \ref{fig:simu-misN}).
 Therefore, this extreme simulation scenario shows that LDpred2 can be sensitive to GWAS sample size misspecification, whereas lassosum (and lassosum2) seems little affected by this.
+
+\begin{figure}[h]
+	\centerline{\includegraphics[width=0.95\textwidth]{simu-misN}}
+	\caption{Results for the simulations with sample size misspecification.
+The GWAS sample size is ``true'' when providing the true per-variant sample size, ``max'' when providing instead the maximum sample size as a unique value to be used for all variants,  ``imputed'' (see Methods in main text), or ``any'' when the method does not use this information (the case for C+T).}
+	\label{fig:simu-misN}
+\end{figure}
 
@@
 Finally, in these simulations with misspecified sample sizes, it seems highly beneficial to use a small value for the SNP heritability hyper-parameter in LDpred2, e.g.\ a value of 0.02 or even 0.002 when the true value is 0.2 (Figure \ref{fig:ldpred2-misN}).
-We recall that using a small value for this hyper-parameter induces a larger regularization on the effect sizes, and this benefits LDpred2 even when using imputed sample sizes.
+We recall that using a small value for this hyper-parameter induces a larger regularization (shrinkage) on the effect sizes, and this benefits LDpred2 even when using imputed sample sizes.
+
+[NEED TO TALK ABOUT THE LD TRANSFO SOMEWHERE + NEW METHODS]
 
@@
 This is the first correction of summary statistics we try in the simulations below. 
-Note that we recompute INFO scores for the subset of 362,307 European individuals used in this paper as they sometimes differ from the ones reported by the UK Biobank (Figure \ref{fig:compare-info}).
-Instead of using dosages to perform the GWAS, it has been argued that using multiple imputation (MI) would be more appropriate \cite[]{Palmer2016}.
+Note that we recompute INFO scores for the subset of 362,307 European individuals used in this paper as they can differ substantially from the ones reported by the UK Biobank for the whole data (Figure \ref{fig:compare-info}).
+As a second option, instead of using dosages to perform the GWAS, it has been argued that using multiple imputation (MI) would be more appropriate \cite[]{Palmer2016}.
 In simulations, we show that $Z_j^{MI} \approx Z_j^{imp} \cdot \text{INFO}_j$ and $\hat{\beta}_j^{MI} \approx \hat{\beta}_j^{imp} \cdot \text{INFO}_j$ (Figure \ref{fig:compare-mi}).
 This is the second correction of summary statistics we try in the simulations below, along with $n_j = N \cdot \text{INFO}_j$. 
+Finally, we try an in-between solution as a third correction, using $\hat{\beta}_j^{imp} \cdot \text{INFO}_j$, $\text{se}(\hat{\beta}_j)^{imp} \cdot \sqrt{\text{INFO}_j}$, and $N \cdot \text{INFO}_j$.
 
 Using 40,000 variants from chromosome 22, we simulate quantitative phenotypes assuming an heritability of 20\% and 2000 causal variants using the ``true'' dataset (cf.\ Methods section ``Data for simulations'').
-We run GWAS on the dosage dataset and use these summary statistics to run LDpred2 and lassosum2 with either no correction of the summary statistics, or with one of the two described above.
+We run GWAS on the dosage dataset and use these summary statistics to run LDpred2 and lassosum2 with either no correction of the summary statistics, or with one of the three described above.
 The LD reference used by LDpred2 and lassosum2 is computed from the validation set using the ``true'' dataset.
 For lassosum2 and LDpred2 which tune parameters using validation set and can possibly use strong regularization, the two corrections just slighly improve predictive performance in these simulations (Figure \ref{fig:simu-info}).
-Correction ``sqrt\_info'', which is equivalent to postprocessing PGS effects by $\sqrt{\text{INFO}}$ slighly improves predictive performance for all methods, but correction ``info'' (using $\hat{\beta}_j^{imp} \cdot \text{INFO}_j$  and $N \cdot \text{INFO}_j$) provides much better predictive performance for LDpred2-auto, with a mean squared correlation of 15.4\% compared to 12.9\% with no correction and 13.3\% with correction ``sqrt\_info'' (Figure \ref{fig:simu-info}).
+All three correction improve predictive performance (Figure \ref{fig:simu-info}), [TODO: FINISH WHEN INTRODUCED TRANSFO + NEW METHODS].
+In the real data applications hereinafter, we choose to use the first correction, ``sqrt\_info'', which is simple because it is equivalent to postprocessing PGS effects by multiplying them by $\sqrt{\text{INFO}_j}$.
 
@@
 \begin{figure}[h]
-	\centerline{\includegraphics[width=0.95\textwidth]{res_brca}}
+	\centerline{\includegraphics[width=0.95\textwidth]{res-brca}}
 	\caption{AUC (for \textit{all} models) for predicting breast cancer in the UK Biobank when using either the OncoArray or the iCOGS summary statistics.}
@@
 
+\subsection*{New LD reference}
+
+We make three changes to the LD reference. 
+First, when using imputed data, we multiply the correlation between variants $j$ and $k$ by $\sqrt{\text{INFO}_j \cdot \text{INFO}_k}$ (for $j \neq k$) since it should approximate the correlation from non-imputed data ([TODO: FIGURE]).
+Second, we also define nearly independent LD blocks using the new optimal algorithm developed in \cite{prive2021optimal}.
+For different number of blocks and maximum number of variants in each block, we use the split with the minimum cost within the ones reducing the original number of non-zero values to less than 60\% (70\% for chromosome 6).
+Having a correlation matrix with independent blocks prevents the small errors in the algorithm (e.g.\ the Gibbs sampler in LDpred2) to propagate to the whole genome.
+It also makes running LDpred2 (and lassosum2) faster, taking about 60\% of the initial time (since only 60\% of the initial non-zero values of the correlation matrix are kept).
+Finally, we have developed a new ``compact'' format for the SFBMs (sparse matrix on disk).
+Instead of using something similar to the ``compressed sparse column'' format which stores all $\{i,~x(i, j)\}$ for a given column $j$, we only store the first index $i_0$ and all the contiguous non-zero values $\{x(i_0, j),~x(i_0 + 1, j),~\dots\}$.
+This makes this format twice as efficient for both LDpred2 and lassosum2.
+
+\subsection*{LDpred2-low-h2 and LDpred2-auto-rob}
+
+Here we introduce the small changes made to LDpred2 (-grid and -auto) in order to make them more robust.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2021-08-13 paper/paper-misspec-supp.tex
add methods

@@
 		\centering
-		\includegraphics[width=.8\linewidth]{lassosum2-mis-maxN}
+		\includegraphics[width=.8\linewidth]{lassosum2-mis-maxN-corr1}
 	\end{subfigure}\vspace{2em}
@@
 		\centering
-		\includegraphics[width=.8\linewidth]{lassosum2-mis-imputeN}
+		\includegraphics[width=.8\linewidth]{lassosum2-mis-imputeN-corr1}
 	\end{subfigure}
@@
 		\centering
-		\includegraphics[width=.8\linewidth]{ldpred2-mis-maxN}
+		\includegraphics[width=.8\linewidth]{ldpred2-mis-maxN-corr1}
 	\end{subfigure}\vspace{2em}
@@
 		\centering
-		\includegraphics[width=.8\linewidth]{ldpred2-mis-imputeN}
+		\includegraphics[width=.8\linewidth]{ldpred2-mis-imputeN-corr1}
 	\end{subfigure}
@@
 \begin{figure}[p]
-	\centerline{\includegraphics[width=0.8\textwidth]{brca_onco_hist_info}}
-	\caption{Histogram of the imputation INFO scores of the HapMap3 variants from the OncoArray GWAS summary statistics for breast cancer \cite[]{michailidou2017association}. \label{fig:hist-info-brca}}
+	\centerline{\includegraphics[width=0.65\textwidth]{brca_icogs_hist_info}}
+	\caption{Histogram of the imputation INFO scores of the HapMap3 variants from the iCOGS GWAS summary statistics for breast cancer \cite[]{michailidou2013large}. \label{fig:hist-info-brca2}}
 \end{figure}
@@
 \begin{figure}[p]
-\centerline{\includegraphics[width=0.8\textwidth]{brca_onco_hist_info}}
-\caption{Histogram of the imputation INFO scores of the HapMap3 variants from the iCOGS GWAS summary statistics for breast cancer \cite[]{michailidou2013large}. \label{fig:hist-info-brca2}}
+	\centerline{\includegraphics[width=0.65\textwidth]{brca_onco_hist_info}}
+	\caption{Histogram of the imputation INFO scores of the HapMap3 variants from the OncoArray GWAS summary statistics for breast cancer \cite[]{michailidou2017association}.
+[TODO: REEXPORT -- WAS CORRUPTED] \label{fig:hist-info-brca}}
 \end{figure}
+
+\FloatBarrier
 
@@
 \FloatBarrier
+
+\begin{figure}[p]
+	\centerline{\includegraphics[width=0.95\textwidth]{compare-cor}}
+	\caption{Comparing correlations between a subset of HapMap3 variants on chromosome 22, computed in three different ways: 1/ from imputed dosages; 2/ from multiple imputation (i.e.\ generating multiple complete datasets sampled according to imputation probabilities, computing correlations, and averaging results); from imputed dosages with the multiplicative correction of $\sqrt{\text{INFO}_j \cdot \text{INFO}_k}$ (for $j \neq k$).}
+	\label{fig:compare-cor}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\FloatBarrier
 \clearpage

### 2021-08-13 paper/paper-misspec.tex
add methods

@@
 \title{Identifying and correcting multiple sources of misspecification\\in GWAS summary statistics for prediction}
-\author{Florian Priv\'e,$^{\text{1,}*}$ Julyan Arbel,$^{\text{2}}$ Timothy S. H. Mak,$^{\text{3}}$ and Bjarni J. Vilhj\'almsson$^{\text{1,4}}$}
+\author{Florian Priv\'e,$^{\text{1,}*}$ Julyan Arbel,$^{\text{2}}$ Timothy S. H. Mak,$^{\text{3}}$ Hugues Aschard,$^{\text{4,5}}$ and Bjarni J. Vilhj\'almsson$^{\text{1,6}}$}
 
@@
 \noindent$^{\text{\sf 3}}$Fano Labs, Hong Kong. \\
-\noindent$^{\text{\sf 4}}$Bioinformatics Research Centre, Aarhus University, Aarhus, 8000, Denmark. \\
+\noindent$^{\text{\sf 4}}$Department of Computational Biology, Institut Pasteur, Paris, 75015, France. \\
+\noindent$^{\text{\sf 5}}$Program in Genetic Epidemiology and Statistical Genetics, Harvard T.H. Chan School of Public Health, Boston, MA, 02115, USA. \\
+\noindent$^{\text{\sf 6}}$Bioinformatics Research Centre, Aarhus University, Aarhus, 8000, Denmark. \\
 \noindent$^\ast$To whom correspondence should be addressed.\\
@@
 
-\subsection*{GWAS using expected genotypes from imputation}
+\subsection*{When using allele dosages from imputation}
 
@@
 We make three changes to the LD reference. 
-First, when using imputed data, we multiply the correlation between variants $j$ and $k$ by $\sqrt{\text{INFO}_j \cdot \text{INFO}_k}$ (for $j \neq k$) since it should approximate the correlation from non-imputed data ([TODO: FIGURE]).
-Second, we also define nearly independent LD blocks using the new optimal algorithm developed in \cite{prive2021optimal}.
+First, when using imputed data, we multiply the correlation between variants $j$ and $k$ by $\sqrt{\text{INFO}_j \cdot \text{INFO}_k}$ (for $j \neq k$) since it approximates well the correlation from non-imputed data (Figure \ref{fig:compare-cor}).
+Second, we also define nearly independent LD blocks using the optimal algorithm developed in \cite{prive2021optimal}.
 For different number of blocks and maximum number of variants in each block, we use the split with the minimum cost within the ones reducing the original number of non-zero values to less than 60\% (70\% for chromosome 6).
-Having a correlation matrix with independent blocks prevents the small errors in the algorithm (e.g.\ the Gibbs sampler in LDpred2) to propagate to the whole genome.
+Having a correlation matrix with independent blocks prevents the small errors in the algorithm (e.g.\ the Gibbs sampler in LDpred2) from propagating to the whole genome.
 It also makes running LDpred2 (and lassosum2) faster, taking about 60\% of the initial time (since only 60\% of the initial non-zero values of the correlation matrix are kept).
-Finally, we have developed a new ``compact'' format for the SFBMs (sparse matrix on disk).
-Instead of using something similar to the ``compressed sparse column'' format which stores all $\{i,~x(i, j)\}$ for a given column $j$, we only store the first index $i_0$ and all the contiguous non-zero values $\{x(i_0, j),~x(i_0 + 1, j),~\dots\}$.
-This makes this format twice as efficient for both LDpred2 and lassosum2.
+Finally, we have developed a new ``compact'' format for the SFBMs (sparse matrices on disk).
+Instead of using something similar to the standard ``compressed sparse column'' format which stores all $\{i,~x(i, j)\}$ for a given column $j$, we only store the first index $i_0$ and all the contiguous values $\{x(i_0, j),~x(i_0 + 1, j),~\dots\}$ up to the last non-zero value for this column $j$.
+This makes this format about twice as efficient for both LDpred2 and lassosum2.
 
@@
 Here we introduce the small changes made to LDpred2 (-grid and -auto) in order to make them more robust.
+First, LDpred2-low-h2 simply consists in running LDpred2-grid by testing $h^2$ within $\{0.3, 0.7, 1, 1.4\} \cdot h^2_\text{LDSC}$ (note the added $0.3$ compared to \cite{prive2020ldpred2}), where $h^2_\text{LDSC}$ is the heritability estimate from LD score regression.
+Indeed, we show in simulations here that using lower values for $h^2$ may provide higher predictive performance in the case of misspecifications (thanks to more shrinkage of the effects).
+In simulations, because of the large misspecifications, we use a larger grid over $\{0.01, 0.1, 0.3, 0.7, 1, 1.4\} \cdot h^2_\text{LDSC}$.
+
+For LDpred2-auto, we introduce two new parameters.
+The first one, \texttt{shrink\_corr}, allows for shrinking off-diagonal elements of the correlation matrix.
+This is similar to parameter `s' in lassosum, and act as a regularization.
+We use a value of 0.9 in simulations and 0.95 in real data when running ``LDpred2-auto-rob'' (and the default value of 1, without any effect, when running ``LDpred2-auto'').
+The second new parameter, \texttt{allow\_jump\_sign}, controls whether variants can change sign over two consecutive iterations of the Gibbs sampler.
+When setting this parameter to false (in the method we name ``LDpred2-auto-rob'' here), this forces the effects to go through 0 before changing sign.
+This is useful to prevent instability (oscillation and ultimately divergence) of the Gibbs sampler under large misspecifications, and is also useful for accelerating convergence of chains with a large initial value for $p$, the proportion of causal variants.
 

### 2021-08-13 paper/paper-misspec.tex
almost finished with the methods

@@
 
-When using the expected genotypes from imputation (dosages), their standard deviations are usually lower than the expected value under Hardy-Weinberg equilibrium (cf.\ Methods section ``Standard deviations and imputation measures'').
-In simulations (cf.\ Methods section ``Data for simulations''), we show that $\text{sd}(G_j)^{true} \approx \text{sd}(G_j)^{imp} / \sqrt{\text{INFO}_j}$, $\text{se}(\hat{\beta}_j)^{true} \approx \text{se}(\hat{\beta}_j)^{imp} \cdot \sqrt{\text{INFO}_j}$, and $\hat{\beta}_j^{true} \approx \hat{\beta}_j^{imp} \cdot \sqrt{\text{INFO}_j}$ (Figures \ref{fig:compare-sd}, \ref{fig:compare-beta}, and \ref{fig:compare-beta-se}).
+The MACH $\hat{r}_j^2$ measure for variant $j$ is given by $\hat{r}_j^2 = \frac{var(G_j)}{2 \hat{\theta} (1 - \hat{\theta})}$, where $\hat{\theta}$ is the estimated allele frequency \cite[]{marchini2010genotype}.
+This measure is also highly concordant with the IMPUTE INFO measure \cite[]{marchini2010genotype}.
+Therefore, when using the expected genotypes from imputation (dosages), their standard deviations are usually lower than the expected value under Hardy-Weinberg equilibrium because $\text{INFO}_j \approx \frac{var(G_j)}{2 \hat{\theta} (1 - \hat{\theta})}$.
+In simulations (cf.\ Methods section ``Data for simulations''), we show that $\text{sd}(G_j)^{true} \approx \text{sd}(G_j)^{imp} / \sqrt{\text{INFO}_j}$, $\text{se}(\hat{\gamma}_j)^{true} \approx \text{se}(\hat{\gamma}_j)^{imp} \cdot \sqrt{\text{INFO}_j}$, and $\hat{\gamma}_j^{true} \approx \hat{\gamma}_j^{imp} \cdot \sqrt{\text{INFO}_j}$ (Figures \ref{fig:compare-sd}, \ref{fig:compare-beta}, and \ref{fig:compare-beta-se}).
 This is the first correction of summary statistics we try in the simulations below. 
@@
 As a second option, instead of using dosages to perform the GWAS, it has been argued that using multiple imputation (MI) would be more appropriate \cite[]{Palmer2016}.
-In simulations, we show that $Z_j^{MI} \approx Z_j^{imp} \cdot \text{INFO}_j$ and $\hat{\beta}_j^{MI} \approx \hat{\beta}_j^{imp} \cdot \text{INFO}_j$ (Figure \ref{fig:compare-mi}).
+In simulations, we show that $Z_j^{MI} \approx Z_j^{imp} \cdot \text{INFO}_j$ and $\hat{\gamma}_j^{MI} \approx \hat{\gamma}_j^{imp} \cdot \text{INFO}_j$ (Figure \ref{fig:compare-mi}).
 This is the second correction of summary statistics we try in the simulations below, along with $n_j = N \cdot \text{INFO}_j$. 
-Finally, we try an in-between solution as a third correction, using $\hat{\beta}_j^{imp} \cdot \text{INFO}_j$, $\text{se}(\hat{\beta}_j)^{imp} \cdot \sqrt{\text{INFO}_j}$, and $N \cdot \text{INFO}_j$.
+Finally, we try an in-between solution as a third correction, using $\hat{\gamma}_j^{imp} \cdot \text{INFO}_j$, $\text{se}(\hat{\gamma}_j)^{imp} \cdot \sqrt{\text{INFO}_j}$, and $N \cdot \text{INFO}_j$.
 
@@
 	\caption{Results for the simulations using dosage data.
-Correction ``sqrt\_info'' corresponds to using $\hat{\beta}_j^{imp} \cdot \sqrt{\text{INFO}_j}$ and $\text{se}(\hat{\beta}_j)^{imp} \cdot \sqrt{\text{INFO}_j}$.
-	Correction ``info'' corresponds to using $\hat{\beta}_j^{imp} \cdot \text{INFO}_j$  and $N \cdot \text{INFO}_j$.}
+Correction ``sqrt\_info'' corresponds to using $\hat{\gamma}_j^{imp} \cdot \sqrt{\text{INFO}_j}$ and $\text{se}(\hat{\gamma}_j)^{imp} \cdot \sqrt{\text{INFO}_j}$.
+	Correction ``info'' corresponds to using $\hat{\gamma}_j^{imp} \cdot \text{INFO}_j$  and $N \cdot \text{INFO}_j$.}
 	\label{fig:simu-info}
@@
 We then also filter out the two long-range LD regions of chromosome 6 and 8 for the OncoArray summary statistics and remove around 500 variants when filtering on differences of MAFs between summary statistics and the validation dataset.
-As for the correction for the INFO scores, we use $\hat{\beta}_j^{imp} \cdot \text{INFO}_j$ and $N \cdot \text{INFO}_j$ as we have shown to perform well in simulations.
+As for the correction for the INFO scores, we use $\hat{\gamma}_j^{imp} \cdot \text{INFO}_j$ and $N \cdot \text{INFO}_j$ as we have shown to perform well in simulations.
 Although results for both QC used are very similar, correcting for the INFO score really improves prediction, especially for LDpred2-auto [TODO: NUMBERS] (Figure \ref{fig:res_brca}).
@@
 
-Among genetic variants with a minor allele frequency larger than 0.01 and an imputation INFO score larger than 0.4 (as reported in the UK Biobank), we sample 40,000 of them according to the inverse of the INFO score density so that they have varying levels of imputation accuracy (Figure \ref{fig:hist-info}).
+Among genetic variants on chromosome 22 and with a minor allele frequency larger than 0.01 and an imputation INFO score larger than 0.4 (as reported by the UK Biobank), we sample 40,000 of them according to the inverse of the INFO score density so that they have varying levels of imputation accuracy (Figure \ref{fig:hist-info}).
 We read the UK Biobank data into two different datasets using function \texttt{snp\_readBGEN} from R package bigsnpr \cite[]{prive2017efficient}, one by reading the BGEN data at random according to imputation probabilities, and another one reading it as dosages (i.e.\ expected values according to imputation probabilities).
-The first dataset is used as what could be the real genotype calls and the second dataset as what would be its imputed version; this technique was used in \cite{prive2019making}.
+The first dataset is used as what could be the real genotype calls and the second dataset as what would be its imputed version; this design technique was used in \cite{prive2019making}.
 
@@
 We sample 10,000 individuals to form a validation set and use the remaining 352,307 individuals as test set.
-We restrict to the genetic variants to the 1,054,315 variants used in the LD reference provided in \cite{prive2020ldpred2}.
+We restrict to the genetic variants to the 1,054,315 HapMap3 variants used in the LD reference provided in \cite{prive2020ldpred2}.
+
+[TODO: MAYBE THE RIGHT PLACE TO PRESENT THE TWO NEW SETS]
 
@@
 \begin{equation}\label{eq:approx-sd-lin}
-\text{sd}(G_j) \approx \dfrac{\text{sd}(y)}{\sqrt{n_j ~ \text{se}(\hat{\beta}_j)^2 + \hat{\beta}_j^2}} ~,
+\text{sd}(G_j) \approx \dfrac{\text{sd}(y)}{\sqrt{n_j ~ \text{se}(\hat{\gamma}_j)^2 + \hat{\gamma}_j^2}} ~,
 \end{equation}
-where $\hat{\beta}_j$ is the marginal (GWAS) effect of variant $j$, $n_j$ is the GWAS sample size associated with variant $j$, $y$ is the vector of phenotypes and ${G}_j$ is the vector of genotypes for variant $j$.
+where $\hat{\gamma}_j$ is the marginal (GWAS) effect of variant $j$, $n_j$ is the GWAS sample size associated with variant $j$, $y$ is the vector of phenotypes and ${G}_j$ is the vector of genotypes for variant $j$.
 This formula is used in LDpred2 \cite{prive2020ldpred2,prive2021high}. Note that, for a binary trait for which logistic regression is used, we have instead
 \begin{equation}\label{eq:approx-sd-log}
-\text{sd}(G_j) \approx \dfrac{2}{\sqrt{n_j^\text{eff} ~ \text{se}(\hat{\beta}_j)^2 + \hat{\beta}_j^2}} ~,
+\text{sd}(G_j) \approx \dfrac{2}{\sqrt{n_j^\text{eff} ~ \text{se}(\hat{\gamma}_j)^2 + \hat{\gamma}_j^2}} ~,
 \end{equation}
@@
 \begin{equation}\label{eq:imputeN}
-n_j \approx \dfrac{\text{var}(y) / \text{var}(G_j) - \hat{\beta}_j^2}{\text{se}(\hat{\beta}_j)^2} ~.
+n_j \approx \dfrac{\text{var}(y) / \text{var}(G_j) - \hat{\gamma}_j^2}{\text{se}(\hat{\gamma}_j)^2} ~,
+\end{equation}
+and impute $n_j^\text{eff}$ from equation \eqref{eq:approx-sd-log} using
+\begin{equation}\label{eq:imputeN}
+n_j^\text{eff} \approx \dfrac{4 / \text{var}(G_j) - \hat{\gamma}_j^2}{\text{se}(\hat{\gamma}_j)^2} ~.
 \end{equation}
 In practice, we also bound this estimate to be between $0.5 \cdot N$ and $1.1 \cdot N$, where $N$ is the global sample size.
-
-\subsection*{Standard deviations and imputation measures}
-
-The MACH $\hat{r}_j^2$ measure for variant $j$ is given by $\hat{r}_j^2 = \frac{var(G_j)}{2 \hat{\theta} (1 - \hat{\theta})}$, where $\hat{\theta}$ is the estimated allele frequency \cite[]{marchini2010genotype}.
-This measure is also highly concordant with the IMPUTE INFO measure \cite[]{marchini2010genotype}.
-Therefore $\text{INFO}_j \approx \frac{var(G_j)}{2 \hat{\theta} (1 - \hat{\theta})}$.
 
@@
 Following notations from \cite{prive2020ldpred2} and denoting $\widetilde{X} = \frac{1}{\sqrt{n-1}} C_n G S^{-1}$, where $G$ is the genotype matrix, $C_n$ is the centering matrix and $S$ is the diagonal matrix of standard deviations of the columns of $G$.
-Then $\widetilde{X}_j^T \widetilde{X} = R_{j,.} = R_{.,j}^T$, $\widetilde{X}_j^T \widetilde{X}_j = 1$ and
-\[
-u_j^{(t)} = \beta_j^{(t-1)} + \widehat{\beta}_j - R_{.,j}^T \beta^{(t-1)} ~,
-\]
-where $r_j = \widehat{\beta}_j =  \dfrac{\widehat{\beta}_j}{\sqrt{n_j ~ \text{se}(\widehat{\beta}_j)^2 + \widehat{\beta}_j^2}}$ and $\widehat{\beta}_j$ is the GWAS effect of variant $j$ and $n$ is the GWAS sample size \cite[]{mak2017polygenic,prive2021high}.
+Then $\widetilde{X}_j^T \widetilde{X} = R_{j,.} = R_{.,j}^T$ and $\widetilde{X}_j^T \widetilde{X}_j = 1$. Moreover, $u_j^{(t)} = \beta_j^{(t-1)} + \widehat{\beta}_j - R_{.,j}^T \beta^{(t-1)}$,
+where $r_j = \widehat{\beta}_j =  \dfrac{\widehat{\gamma}_j}{\sqrt{n_j ~ \text{se}(\widehat{\gamma}_j)^2 + \widehat{\gamma}_j^2}}$ and $\widehat{\gamma}_j$ is the GWAS effect of variant $j$ and $n$ is the GWAS sample size \cite[]{mak2017polygenic,prive2021high}.
 Then the most time-consuming part is computing $R_{.,j}^T \beta^{(t-1)}$.
@@
 
-In this new implementation of the lassosum model, the input parameters are the correlation matrix $R$, the GWAS summary statistics ($\widehat{\beta}_j$, $\text{se}(\widehat{\beta}_j)$ and $n_j$), and the two hyper-parameters $\lambda$ and $\delta$.
+In this new implementation of the lassosum model, the input parameters are the correlation matrix $R$, the GWAS summary statistics ($\widehat{\gamma}_j$, $\text{se}(\widehat{\gamma}_j)$ and $n_j$), and the two hyper-parameters $\lambda$ and $\delta$.
 Therefore, except for the hyper-parameters, this is the exact same input as for LDpred2 \cite[]{prive2020ldpred2}.
 We try $\delta \in \{0.001, 0.005, 0.02, 0.1, 0.6, 3\}$ by default in lassosum2, instead of $s \in \{0.2, 0.5, 0.8, 1.0\}$ in lassosum.
-For $\lambda$, the default in lassosum uses a sequence of 20 values equally spaced on a log scale between 0.1 and 0.001; we now use a sequence between $\lambda_0$ and $\lambda_0 / 100$ by default in lassosum2, where $\lambda_0 = \max_j \left|\widehat{\beta}_j\right|$ is the minimum $\lambda$ for which no variable enters the model because the L1-regularization is too strong.
+For $\lambda$, while the default in lassosum uses a sequence of 20 values equally spaced on a log scale between 0.1 and 0.001; we now use a sequence between $\lambda_0$ and $\lambda_0 / 100$ by default in lassosum2, where $\lambda_0 = \max_j \left|\widehat{\beta}_j\right|$ is the minimum $\lambda$ for which no variable enters the model because the L1-regularization is too strong.
 Note that we do not provide an ``auto'' version using pseudo-validation (as in \cite{mak2017polygenic}) as we have not found it to be very robust (Figure \ref{fig:pseudoval}).
-Also note that, as in LDpred2, we run lassosum2 genome-wide using a sparse correlation matrix which assumes that variants further away than 3 cM are not correlated, and therefore we do not require splitting the genome into independent LD blocks anymore (as is required in lassosum).
+Also note that, as in LDpred2, we run lassosum2 genome-wide using a sparse correlation matrix which assumes that variants further away than 3 cM are not correlated, and therefore we do not require splitting the genome into independent LD blocks anymore (as done in lassosum).
 
@@
 
-Authors thank GenomeDK and Aarhus University for providing computational resources and support that contributed to these research results.
+Authors thank Matthew Stephens for helpful discussions. 
+Authors also thank GenomeDK and Aarhus University for providing computational resources and support that contributed to these research results.
 This research has been conducted using the UK Biobank Resource under Application Number 58024.

### 2021-08-16 paper/paper-misspec-supp.tex
restart results

@@
 	\caption{Multiple results (squared correlation r2 between polygenic score and phenotype) over a grid of parameters ($\lambda$ and $\delta$) for lassosum2 in one of the simulations with misspecified GWAS sample sizes (N).
-The top subfigure is using the maximum N, whereas the bottom subfigure is using the imputed N (cf.\ figure \ref{fig:simu-misN}).
+The top subfigure is using the maximum N, whereas the bottom subfigure is using the imputed N.
 Missing points represent models that were detected as divergent.}
@@
 	\caption{Multiple results (squared correlation r2 between polygenic score and phenotype) over a grid of parameters (the SNP heritability h2 and the proportion of causal variants p) for LDpred2(-grid) in one of the simulations with misspecified GWAS sample sizes (N).
-The top subfigure is using the maximum N, whereas the bottom subfigure is using the imputed N (cf.\ figure \ref{fig:simu-misN}).
-Note that the small h2 values (0.002, 0.02, 0.06) were not considered in the results of figure \ref{fig:simu-misN}.
-Missing points represent models that were detected as divergent.
-[TODO: REPLACE BY CONTOUR PLOT?
-REMOVE SPARSITY COLORING?]}
+The top subfigure is using the maximum N, whereas the bottom subfigure is using the imputed N.
+Note that the small h2 values (0.002, 0.02, 0.06) were added and are part of the method we call ``LDpred2-low-h2'' in the main text.
+Missing points represent models that were detected as divergent.}
 	\label{fig:ldpred2-misN}
+\end{figure}
+
+\begin{figure}[p]
+	\begin{subfigure}{\textwidth}
+		\centering
+		\includegraphics[width=.8\linewidth]{ldpred2-mis-maxN-corr2}
+	\end{subfigure}\vspace{2em}
+	\begin{subfigure}{\textwidth}
+		\centering
+		\includegraphics[width=.8\linewidth]{ldpred2-mis-imputeN-corr2}
+	\end{subfigure}
+	\caption{Same results as figure \ref{fig:ldpred2-misN}, but using the LD with independent blocks.}
+	\label{fig:ldpred2-misN-corr2}
 \end{figure}

### 2021-08-16 paper/paper-misspec.tex
restart results

@@
 %\usepackage{algorithm2e}
-\usepackage{url}
+\usepackage{xurl}
 \usepackage{bbm}
@@
 
-\vspace*{8em}
+\vspace*{5em}
 
@@
 
-\section*{Introduction}
+\section{Introduction}
 
@@
 
-\section*{Results}
+\section{Results}
 
-\subsection*{Misspecification of GWAS sample sizes}
+\subsection{Misspecification of GWAS sample sizes}
 
 We design simulations where variants have different GWAS sample sizes, which is often the case when meta-analyzing GWAS from multiple cohorts without the same genome coverage.
-Using 40,000 variants from chromosome 22 (Methods), we simulate quantitative phenotypes assuming an heritability of 20\% and 2000 causal variants.
-For half of the variants, we use 100\% of 300,000 individuals for GWAS, but use only 80\% for one quarter of them and 60\% for the remaining quarter. 
-We then run C+T, LDpred2-inf, LDpred2(-grid), LDpred2-auto, lassosum, lassosum2, PRS-CS and SBayesR by using either the true per-variant GWAS sample size, or the total sample size. 
-We do not show results for SBayesR because it always diverged and for PRS-CS because the overlap with the LD reference they provide is too small.
-Averaged over 10 simulations, when providing true per-variant GWAS sample sizes, squared correlations (in \%) between the polygenic scores and the simulated phenotypes are of [TODO: COMPLETE] (Figure \ref{fig:simu-misN}).
+Using 40,000 variants from chromosome 22 (Methods), we simulate quantitative phenotypes with a heritability of 20\% and 2000 causal variants.
+For half of the variants, we use 100\% of 300,000 individuals for GWAS, but use only 80\% for one quarter of variants and 60\% for the remaining quarter. 
+We then run C+T, LDpred2-inf, LDpred2(-grid), LDpred2-auto, lassosum, lassosum2, PRS-CS and SBayesR by using either the true per-variant GWAS sample size, the total sample size, or the imputed sample size.
+However, we do not show results for SBayesR because it always diverged and for PRS-CS because the overlap with the LD reference they provide is too small.
+Averaged over 10 simulations, when providing true per-variant GWAS sample sizes, squared correlations between the polygenic scores and the simulated phenotypes are of 0.123 for C+T, 0.161 for lassosum, 0.169 for lassosum2, 0.159 for LDpred2(-grid), 0.140 for LDpred2-auto, and 0.141 for LDpred2-inf (Figure \ref{fig:simu-misN}).
+Results when using imputed sample sizes are quite similar.
 Note that C+T does not use this sample size information.
-When using the total GWAS sample size instead of the per-variant sample sizes, predictive performance slightly decreases [TODO: REPLACE NUMBERS] from 17.5 to 17.4 for lassosum, from 18.0 to 17.5 for lassosum2, but dramatically decreases for LDpred2 with new values of 11.5 for LDpred2-grid, 5.7 for LDpred2-auto and 6.8 for LDpred2-inf (Figure \ref{fig:simu-misN}).
+When using the total GWAS sample size instead of the per-variant sample sizes, predictive performance slightly decreases to 0.157 for lassosum and to 0.163 for lassosum2, but dramatically decreases for LDpred2 with new values of 0.134 for LDpred2-grid, 0.119 for LDpred2-auto, and 0.123 for LDpred2-inf (Figure \ref{fig:simu-misN}).
 Therefore, this extreme simulation scenario shows that LDpred2 can be sensitive to GWAS sample size misspecification, whereas lassosum (and lassosum2) seems little affected by this.
@@
 	\caption{Results for the simulations with sample size misspecification.
-The GWAS sample size is ``true'' when providing the true per-variant sample size, ``max'' when providing instead the maximum sample size as a unique value to be used for all variants,  ``imputed'' (see Methods in main text), or ``any'' when the method does not use this information (the case for C+T).}
+The GWAS sample size is ``true'' when providing the true per-variant sample size, ``max'' when providing instead the maximum sample size as a unique value to be used for all variants,  ``imputed'' (see Methods in main text), or ``any'' when the method does not use this information (the case for C+T).
+Red bars correspond to using the LD with independent blocks (Methods).}
 	\label{fig:simu-misN}
@@
 First, the reason why results for LDpred2-auto are the same as with LDpred2-inf is because it always converges to an infinitesimal model (p = 1) in these simulations.
+We therefore introduce two new parameters to make LDpred2-auto more robust and call this ``LDpred2-auto-rob'' here (Methods).
 Second, for lassosum2, results for a grid of parameters (over $\lambda$ and $\delta$) and quite smooth compared to LDpred2 (Figures \ref{fig:lassosum2-misN} and \ref{fig:ldpred2-misN}). 
-Results slightly improve globally for both lassosum2 and LDpred2 when using imputed instead of maximum GWAS sample sizes.
-Finally, in these simulations with misspecified sample sizes, it seems highly beneficial to use a small value for the SNP heritability hyper-parameter in LDpred2, e.g.\ a value of 0.02 or even 0.002 when the true value is 0.2 (Figure \ref{fig:ldpred2-misN}).
-We recall that using a small value for this hyper-parameter induces a larger regularization (shrinkage) on the effect sizes, and this benefits LDpred2 even when using imputed sample sizes.
+In these simulations with misspecified sample sizes, it seems highly beneficial to use a small value for the SNP heritability hyper-parameter $h^2$ in LDpred2, e.g.\ a value of 0.02 or even 0.002 when the true value is 0.2 (Figure \ref{fig:ldpred2-misN}).
+We recall that using a small value for this hyper-parameter induces a larger regularization (shrinkage) on the effect sizes.
+When trying LDpred2(-grid) with these low values for $h^2$, we call this ``LDpred2-low-h2'' here.
+Results with LDpred2-low-h2 improves from 0.159 to 0.169 when using true sample sizes and from 0.134 to 0.163 when using the maximum sample size.
+Finally, we introduce a last change for robustness here; we form independent LD blocks in the LD matrix to prevent small errors in the Gibbs sampler to propagate to too many variants (Methods).
+This change seems to solve convergence issues of LDpred2 in these simulations (Figure \ref{fig:ldpred2-misN-corr2}) and further improves predictive performance for all LDpred2 methods (Figure \ref{fig:simu-misN}).
 
-[NEED TO TALK ABOUT THE LD TRANSFO SOMEWHERE + NEW METHODS]
+\subsection{When using allele dosages from imputation}
 
-\subsection*{When using allele dosages from imputation}
-
-The MACH $\hat{r}_j^2$ measure for variant $j$ is given by $\hat{r}_j^2 = \frac{var(G_j)}{2 \hat{\theta} (1 - \hat{\theta})}$, where $\hat{\theta}$ is the estimated allele frequency \cite[]{marchini2010genotype}.
-This measure is also highly concordant with the IMPUTE INFO measure \cite[]{marchini2010genotype}.
-Therefore, when using the expected genotypes from imputation (dosages), their standard deviations are usually lower than the expected value under Hardy-Weinberg equilibrium because $\text{INFO}_j \approx \frac{var(G_j)}{2 \hat{\theta} (1 - \hat{\theta})}$.
-In simulations (cf.\ Methods section ``Data for simulations''), we show that $\text{sd}(G_j)^{true} \approx \text{sd}(G_j)^{imp} / \sqrt{\text{INFO}_j}$, $\text{se}(\hat{\gamma}_j)^{true} \approx \text{se}(\hat{\gamma}_j)^{imp} \cdot \sqrt{\text{INFO}_j}$, and $\hat{\gamma}_j^{true} \approx \hat{\gamma}_j^{imp} \cdot \sqrt{\text{INFO}_j}$ (Figures \ref{fig:compare-sd}, \ref{fig:compare-beta}, and \ref{fig:compare-beta-se}).
+\cite{marchini2010genotype} showed that the IMPUTE INFO measure is highly concordant with the MACH measure $\hat{r}_j^2 = \frac{var(G_j)}{2 \hat{\theta} (1 - \hat{\theta_j})}$, where $\hat{\theta_j}$ is the estimated allele frequency of variant $j$.
+Therefore, when using the expected genotypes from imputation (dosages), their standard deviations are often lower than the expected value under Hardy-Weinberg equilibrium, because $\text{INFO}_j \approx \frac{var(G_j)}{2 \hat{\theta_j} (1 - \hat{\theta_j})}$.
+In simulations (cf.\ Methods section ``Data for simulations''), we verify that $\text{sd}(G_j)^{true} \approx \text{sd}(G_j)^{imp} / \sqrt{\text{INFO}_j}$ (Figures \ref{fig:compare-sd}.
+We also show that GWAS effect sizes $\hat{\gamma}$ computed from imputed dosages are over-estimated: $\hat{\gamma}_j^{true} \approx \hat{\gamma}_j^{imp} \cdot \sqrt{\text{INFO}_j}$ and $\text{se}(\hat{\gamma}_j)^{true} \approx \text{se}(\hat{\gamma}_j)^{imp} \cdot \sqrt{\text{INFO}_j}$ (Figures \ref{fig:compare-beta} and \ref{fig:compare-beta-se}).
 This is the first correction of summary statistics we try in the simulations below. 
 Note that we recompute INFO scores for the subset of 362,307 European individuals used in this paper as they can differ substantially from the ones reported by the UK Biobank for the whole data (Figure \ref{fig:compare-info}).
-As a second option, instead of using dosages to perform the GWAS, it has been argued that using multiple imputation (MI) would be more appropriate \cite[]{Palmer2016}.
-In simulations, we show that $Z_j^{MI} \approx Z_j^{imp} \cdot \text{INFO}_j$ and $\hat{\gamma}_j^{MI} \approx \hat{\gamma}_j^{imp} \cdot \text{INFO}_j$ (Figure \ref{fig:compare-mi}).
+As a second option, instead of using dosages to compute the GWAS summary statistics, it has been argued that using multiple imputation (MI) would be more appropriate \cite[]{Palmer2016}.
+In simulations, we show that $\hat{\gamma}_j^{MI} \approx \hat{\gamma}_j^{imp} \cdot \text{INFO}_j$ and $Z_j^{MI} \approx Z_j^{imp} \cdot \text{INFO}_j$, where $Z = \hat{\gamma} / \text{se}(\hat{\gamma})$ (Figure \ref{fig:compare-mi}).
 This is the second correction of summary statistics we try in the simulations below, along with $n_j = N \cdot \text{INFO}_j$. 
@@
 	\centerline{\includegraphics[width=0.95\textwidth]{simu-info}}
-	\caption{Results for the simulations using dosage data.
+	\caption{Results of predictive performance for the simulations using GWAS summary statistics from imputed dosage data.
 Correction ``sqrt\_info'' corresponds to using $\hat{\gamma}_j^{imp} \cdot \sqrt{\text{INFO}_j}$ and $\text{se}(\hat{\gamma}_j)^{imp} \cdot \sqrt{\text{INFO}_j}$.
-	Correction ``info'' corresponds to using $\hat{\gamma}_j^{imp} \cdot \text{INFO}_j$  and $N \cdot \text{INFO}_j$.}
+	Correction ``info'' corresponds to using $\hat{\gamma}_j^{imp} \cdot \text{INFO}_j$  and $N \cdot \text{INFO}_j$.
+Correction ``in\_between'' corresponds to using $\hat{\gamma}_j^{imp} \cdot \text{INFO}_j$, $\text{se}(\hat{\gamma}_j)^{imp} \cdot \sqrt{\text{INFO}_j}$, and $N \cdot \text{INFO}_j$.}
 	\label{fig:simu-info}
@@
 
-\subsection*{Application to breast cancer summary statistics}
+\subsection{Application to breast cancer summary statistics}
 
@@
 For the iCOGS summary statistics, we also see a clear trend with sd\_ss being lower and lower than sd\_af as INFO decreases, but they do not have issues with PCs  (Figure \ref{fig:brca_icogs_qc}).
-
-[ALSO ADD THE GWAS META-ANALYSIS NEFF \textasciitilde20K?]
 
@@
 
-\section*{Discussion}
+\section{Discussion}
+
+[TODO: TALK ABOUT HARD INFERRING N != INFO]
 
@@
 
-\section*{Materials and Methods}
+\section{Materials and Methods}
 
-\subsection*{Data for simulations}
+\subsection{Data for simulations}
 
@@
 
-\subsection*{Data for real analyses}
+\subsection{Data for real analyses}
 
@@
 
-\subsection*{GWAS sample size imputation}
+\subsection{GWAS sample size imputation}
 
@@
 
-\subsection*{New implementation of lassosum in bigsnpr}
+\subsection{New implementation of lassosum in bigsnpr}
 
@@
 
-\subsection*{New LD reference}
+\subsection{New LD reference}
 
@@
 
-\subsection*{LDpred2-low-h2 and LDpred2-auto-rob}
+\subsection{LDpred2-low-h2 and LDpred2-auto-rob}
 
@@
 
-\section*{Code and results availability}
+\section*{Code availability}
 
-All code used for this paper is available at %\url{https://github.com/privefl/paper-lassosum2/tree/master/code}.
-Latest versions of R package bigsnpr can be installed from GitHub.
-A tutorial on running lassosum2 along with LDpred2 using R package bigsnpr is available at \url{https://privefl.github.io/bigsnpr-extdoc/polygenic-scores-pgs.html}.
+All code used for this paper is available at \url{https://github.com/privefl/paper-misspec/tree/master/code}.
+The latest version of R package bigsnpr can be installed from GitHub.
+Two tutorials on running LDpred2 and lassosum2 using R package bigsnpr are available at \url{https://privefl.github.io/bigsnpr/articles/LDpred2.html} and \url{https://privefl.github.io/bigsnpr-extdoc/polygenic-scores-pgs.html}.
 We have extensively used R packages bigstatsr and bigsnpr \cite[]{prive2017efficient} for analyzing large genetic data, packages from the future framework \cite[]{bengtsson2020unifying} for easy scheduling and parallelization of analyses on the HPC cluster, and packages from the tidyverse suite \cite[]{wickham2019welcome} for shaping and visualizing results.

### 2021-08-16 paper/paper-misspec.tex
continue results

@@
 
-\cite{marchini2010genotype} showed that the IMPUTE INFO measure is highly concordant with the MACH measure $\hat{r}_j^2 = \frac{var(G_j)}{2 \hat{\theta} (1 - \hat{\theta_j})}$, where $\hat{\theta_j}$ is the estimated allele frequency of variant $j$.
+\cite{marchini2010genotype} showed that the IMPUTE INFO measure is highly concordant with the MACH measure $\hat{r}_j^2 = \frac{var(G_j)}{2 \hat{\theta_j} (1 - \hat{\theta_j})}$, where $\hat{\theta_j}$ is the estimated allele frequency of variant $j$.
 Therefore, when using the expected genotypes from imputation (dosages), their standard deviations are often lower than the expected value under Hardy-Weinberg equilibrium, because $\text{INFO}_j \approx \frac{var(G_j)}{2 \hat{\theta_j} (1 - \hat{\theta_j})}$.
@@
 Using 40,000 variants from chromosome 22, we simulate quantitative phenotypes assuming an heritability of 20\% and 2000 causal variants using the ``true'' dataset (cf.\ Methods section ``Data for simulations'').
-We run GWAS on the dosage dataset and use these summary statistics to run LDpred2 and lassosum2 with either no correction of the summary statistics, or with one of the three described above.
-The LD reference used by LDpred2 and lassosum2 is computed from the validation set using the ``true'' dataset.
-For lassosum2 and LDpred2 which tune parameters using validation set and can possibly use strong regularization, the two corrections just slighly improve predictive performance in these simulations (Figure \ref{fig:simu-info}).
-All three correction improve predictive performance (Figure \ref{fig:simu-info}), [TODO: FINISH WHEN INTRODUCED TRANSFO + NEW METHODS].
-In the real data applications hereinafter, we choose to use the first correction, ``sqrt\_info'', which is simple because it is equivalent to postprocessing PGS effects by multiplying them by $\sqrt{\text{INFO}_j}$.
+We compute GWAS summary statistics from the dosage dataset and use these summary statistics to run LDpred2 and lassosum2 with either no correction of the summary statistics, or with one of the three corrections described above.
+The LD reference used by LDpred2 and lassosum2 is computed from the validation set using the dataset with ``true'' genotypes.
+For lassosum2 and LDpred2(-grid) which tune parameters using the validation set and can possibly use strong regularization, correcting for imputation quality slighly improve predictive predictive performance in these simulations (Figure \ref{fig:simu-info}).
+However, correcting for imputation quality can dramatically improve predictive performance for LDpred2-auto. 
+Moreover, new additions for robustness introduced before, namely LDpred2-low-h2, LDpred2-auto-rob, and forming independent blocks in the LD matrix improve predictive performance as well for all corrections (Figure \ref{fig:simu-info}).
+
+In the real data applications hereinafter, we choose to use the first correction, ``sqrt\_info'', which is simple because it is equivalent to post-processing PGS effects by multiplying them by $\sqrt{\text{INFO}}$.
 
@@
 	Correction ``info'' corresponds to using $\hat{\gamma}_j^{imp} \cdot \text{INFO}_j$  and $N \cdot \text{INFO}_j$.
-Correction ``in\_between'' corresponds to using $\hat{\gamma}_j^{imp} \cdot \text{INFO}_j$, $\text{se}(\hat{\gamma}_j)^{imp} \cdot \sqrt{\text{INFO}_j}$, and $N \cdot \text{INFO}_j$.}
+Correction ``in\_between'' corresponds to using $\hat{\gamma}_j^{imp} \cdot \text{INFO}_j$, $\text{se}(\hat{\gamma}_j)^{imp} \cdot \sqrt{\text{INFO}_j}$, and $N \cdot \text{INFO}_j$.
+Red bars correspond to using the LD with independent blocks (Methods).}
 	\label{fig:simu-info}
@@
 
-[TALK ABOUT THE SAMPLE SIZES BEFORE/AFTER QC?]
-
 Breast cancer summary statistics are interesting because they include results from two mega analyses \cite[]{michailidou2013large,michailidou2015genome,michailidou2017association}, which means there is some larger precision in the parameters reported, such as the INFO scores and the sample sizes.
-INFO scores for the OncoArray summary statistics (after having restricted to HapMap3 variants) are globally very good (Figure \ref{fig:hist-info-brca}), but the ones from iCOGS are a bit worse (Figure \ref{fig:hist-info-brca2}), probably because the chip used included around 200K variants only, compared to more than 500K variants for the OncoArray.
-For the OncoArray summary statistics, we first compare the standard deviations inferred from the reported allele frequencies (i.e.\ $\sqrt{2 f (1 - f)}$ where $f$ is the allele frequency, and denoted as sd\_af) versus the ones inferred from the GWAS summary statistics (Equation \eqref{eq:approx-sd-log}, and denoted as sd\_ss).
-When coloring by INFO scores, we see a clear trend with sd\_ss being lower and lower than sd\_af as INFO decreases.
-Using $\text{sd\_ss} / \sqrt{\text{INFO}}$ instead, we get a very good fit with sd\_af, except for some variants of chromosome 6 and 8.
+Imputation INFO scores for the OncoArray summary statistics (after having restricted to HapMap3 variants) are generally very good (Figure \ref{fig:hist-info-brca}) and better than the ones from iCOGS (Figure \ref{fig:hist-info-brca2}), probably because the chip used included around 200K variants only, compared to more than 500K variants for the OncoArray.
+For both summary statistics, we first compare the standard deviations inferred from the reported allele frequencies (i.e.\ $\sqrt{2 f (1 - f)}$ where $f$ is the allele frequency, and denoted as sd\_af) versus the ones inferred from the GWAS summary statistics (Equation \eqref{eq:approx-sd-log}, and denoted as sd\_ss).
+When coloring by INFO scores, we see a clear trend with sd\_ss being lower than sd\_af as INFO decreases; indeed, using $\text{sd\_ss} / \sqrt{\text{INFO}}$ provides a very good fit for sd\_af, except for some variants of chromosome 6 and 8 for the OncoArray summary statistics (Figures \ref{fig:brca_onco_qc} and \ref{fig:brca_icogs_qc}).
 Most of these outlier variants are in regions 25-33 Mbp of chromosome 6 and 8-12 Mbp of chromosome 8 (Figure \ref{fig:hist-bad-pos}), which are two known long-range LD regions \cite[]{price2008long}.
-We hypothesize that this is due to using principal components (PCs) that capture LD structure as covariates in GWAS \cite[]{prive2020efficient}.
+We hypothesize that this is due to using principal components (PCs) as covariates in GWAS that capture LD structure instead of population structure \cite[]{prive2020efficient}.
 To validate this hypothesis, we simulate a phenotype using HapMap3 variants of chromosome 6 for 10,000 individuals from the UK Biobank, then we run GWAS with or without PC19 as covariate.
@@
 In these simulations, the same bias as in figure \ref{fig:hist-bad-pos}B is observed for the variants in this region (Figure \ref{fig:gwas_bad_pc}), therefore confirming our hypothesis.
-For the iCOGS summary statistics, we also see a clear trend with sd\_ss being lower and lower than sd\_af as INFO decreases, but they do not have issues with PCs  (Figure \ref{fig:brca_icogs_qc}).
 
@@
 	\caption{Standard deviations inferred from the OncoArray breast cancer GWAS summary statistics versus the ones inferred from the reported GWAS allele frequencies.
-Only 100,000 variants are represented, at random.}
+Only 100,000 HapMap3 variants are represented, at random.}
 	\label{fig:brca_onco_qc}
@@
 
-Therefore, having the INFO score enables both to correct for it in the QC step to be able to pinpoint possible other problems, and possibly to correct for it and improve prediction as shown in the simulations before.
+Therefore, when an accurate imputation INFO score is provided, this enables both to correct for it in the QC step we propose to possibly uncover problems with the summary statistics, and when building polygenic scores to improve prediction.
 We try this for the two breast cancer summary statistics.
@@
 We then also filter out the two long-range LD regions of chromosome 6 and 8 for the OncoArray summary statistics and remove around 500 variants when filtering on differences of MAFs between summary statistics and the validation dataset.
-As for the correction for the INFO scores, we use $\hat{\gamma}_j^{imp} \cdot \text{INFO}_j$ and $N \cdot \text{INFO}_j$ as we have shown to perform well in simulations.
-Although results for both QC used are very similar, correcting for the INFO score really improves prediction, especially for LDpred2-auto [TODO: NUMBERS] (Figure \ref{fig:res_brca}).
-
-[SHOULD JUST REPORT ONE PREDICTION INSTEAD OF ALL? THEN MAKE A TABLE?]
+As for the correction for the INFO scores, we use the first correction, ``sqrt\_info'', which is simple because it is equivalent to post-processing PGS effects by multiplying them by $\sqrt{\text{INFO}}$.
+Although results for both QC used are very similar, correcting for the INFO score slightly improves prediction when deriving polygenic scores based on iCOGS summary statistics (Figure \ref{fig:res_brca}).
+All other improvements introduced before have little or no effect here [DEPRESSING; SHOULD REWORD?].
 
@@
 	\centerline{\includegraphics[width=0.95\textwidth]{res-brca}}
-	\caption{AUC (for \textit{all} models) for predicting breast cancer in the UK Biobank when using either the OncoArray or the iCOGS summary statistics.}
+	\caption{Raw partial correlations (for \textit{all} models) for predicting breast cancer in the UK Biobank when using either the OncoArray or the iCOGS summary statistics.}
 	\label{fig:res_brca}
 \end{figure}
+
+[TODO: NEED A TABLE WITH THE SUMSTATS AND SAMPLE SIZES AND SO ON]
 

### 2021-08-16 paper/paper-misspec.tex
sumstats table

@@
 
-\title{Identifying and correcting multiple sources of misspecification\\in GWAS summary statistics for prediction}
+\title{Identifying and correcting multiple sources of misspecification\\in GWAS summary statistics for prediction\\---~OR~---\\Reducing misspecifications and improving tagging\\for better polygenic scores}
 \author{Florian Priv\'e,$^{\text{1,}*}$ Julyan Arbel,$^{\text{2}}$ Timothy S. H. Mak,$^{\text{3}}$ Hugues Aschard,$^{\text{4,5}}$ and Bjarni J. Vilhj\'almsson$^{\text{1,6}}$}
@@
 
-[TODO: NEED A TABLE WITH THE SUMSTATS AND SAMPLE SIZES AND SO ON]
+\subsection{}
 
@@
 We restrict to the genetic variants to the 1,054,315 HapMap3 variants used in the LD reference provided in \cite{prive2020ldpred2}.
+We also try two new sets of variants (see next section).
+
+To define phenotypes in the UK Biobank, we first map ICD10 and ICD9 codes (UKBB fields 40001, 40002, 40006, 40013, 41202, 41270 and 41271) to phecodes using R package PheWAS \cite[]{carroll2014r,wu2019mapping}.
+We use external published GWAS summary statistics listed in table \ref{tab:sumstats} to derive polygenic scores.
+
+\begin{table}[h]
+	\centering
+	\begin{tabular}{|l||c|c|c||c|c|}
+		\hline
+		Trait & GWAS citation & \begin{tabular}{@{}c@{}}Effective GWAS\\sample size\end{tabular} & \# GWAS variants & \begin{tabular}{@{}c@{}}\# matched variants\\with INFO > 0.4\end{tabular} & Mean INFO \\
+		\hline
+		Breast cancer (BRCA) [iCOGS] & \cite{michailidou2017association} & ~~87036.6 & 11,792,542 & 1,051,242 & 0.841 \\
+		Breast cancer (BRCA) [OncoArray] & \cite{michailidou2017association} & 104441.6 & 11,792,542 & 1,054,233 & 0.968 \\
+		Type 1 diabetes (T1D) [Affymetrix] & \cite{censin2017childhood} & ~~~~5515.6  & ~~8,996,866 & ~~~934,712 & 0.885 \\	
+		Type 1 diabetes (T1D) [Illumina] & \cite{censin2017childhood} & ~~~~7982.0  & ~~8,996,866 & ~~~949,334 & 0.942 \\
+		Prostate cancer (PRCA) & \cite{schumacher2018association} & 135316.1 & 20,370,946 & ~~~818,400 & 0.969 \\
+		Depression (MDD) [without UKBB] & \cite{wray2018genome} & 110463.7 & ~~9,874,289 & 1,049,455 & 0.968 \\
+		Coronary artery disease (CAD) & \cite{nikpay2015comprehensive} & 129014.3 & ~~9,455,778 & 1,052,200 & 0.982 \\
+		\hline
+	\end{tabular}
+	\caption{Summary of external GWAS summary statistics used. PRCA summary statistics have many variants with a missing INFO score, which we discard.\label{tab:sumstats}}
+\end{table}
+
+\subsection{Two new sets of variants}
 

### 2021-08-17 paper/paper-misspec.tex
desc larger sets

@@
 
-[TODO: MAYBE THE RIGHT PLACE TO PRESENT THE TWO NEW SETS]
+We also design two larger sets of imputed variants to compare against using only HapMap3 variants for prediction. 
+Following \cite{prive2021high}, we first restrict to UKBB variants with MAF > 0.01 and INFO > 0.3.
+We then compile frequencies and imputation INFO scores from other datasets, iPSYCH and summary statistics for breast cancer, prostate cancer, coronary artery disease, type-1 diabetes and depression \cite[]{bybjerg2020ipsych2015,michailidou2017association,schumacher2018association,nikpay2015comprehensive,censin2017childhood,wray2018genome}.
+We restrict to variants with a mean INFO > 0.3 in these other datasets, and also compute the median frequency per variant.
+To exclude potential mismappings in the genotyped data \cite[]{kunert2020allele} that might have propagated to the imputed data, we compare median frequencies in the external data to the ones in the UK Biobank.
+As we expect these potential errors to be localized around errors in the genotype data, we apply a moving-average smoothing on the frequency differences to increase power to detect these errors and also reduce false positives.
+We define the threshold on these smoothed differences based on visual inspection of their histogram (0.03).
+This results in an initial set of 9,394,361 variants.
+
+From this large set of variants, we define two sets.
+One is based on clumping, using a threshold $r^2 = 0.9$ over a radius of 100 Kbp and prioritizing HapMap3 variants and larger INFO scores.
+This results in a set ``clump'' of 2,465,478 variants, among which there are 554,655 of the initial HapMap3 variants.
+For the second set, we aim at maximizing the tagging of the initial 9,394,361 variants, i.e.
+$\sum_{i \in \text{all}} \max_{j \in \text{set}}{|R_{i,j}|}$ (inspired from the alternative sensitivity of \cite{agier2016systematic}).
+We design a greedy algorithm that first select all HapMap3 variants, then add one variant at a time, the one that maximizes the addition to this sum, until no variant can add more than 0.2.
+This results in a set ``maxtag'' of 2,029,086 variants.
 

### 2021-08-17 paper/paper-misspec-supp.tex
continue results
finish later

@@
 
-\begin{figure}[h]
-	\centerline{\includegraphics[width=0.85\textwidth]{gwas_bad_pc}}
+\begin{figure}[p]
+	\centerline{\includegraphics[width=0.95\textwidth]{gwas_bad_pc}}
 	\caption{Standard deviations inferred from the simulated GWAS summary statistics (\textbf{A:} with no covariate; \textbf{B:} with PC19 from the UK Biobank as covariate) versus the ones inferred from the allele frequencies.
@@
 
-\begin{figure}[h]
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\FloatBarrier
+
+\begin{figure}[p]
 	\centerline{\includegraphics[width=0.95\textwidth]{brca_icogs_qc}}
@@
 	\label{fig:brca_icogs_qc}
+\end{figure}
+
+\begin{figure}[p]
+	\centerline{\includegraphics[width=0.95\textwidth]{cad_qc}}
+	\caption{Standard deviations inferred from the CAD GWAS summary statistics versus the ones inferred from the reported GWAS allele frequencies. Only 100,000 variants are represented, at random.}
+	\label{fig:cad_qc}
+\end{figure}
+
+\begin{figure}[p]
+	\centerline{\includegraphics[width=0.95\textwidth]{mdd_qc}}
+	\caption{Standard deviations inferred from the MDD GWAS summary statistics versus the ones inferred from the reported GWAS allele frequencies. Only 100,000 variants are represented, at random.}
+	\label{fig:mdd_qc}
+\end{figure}
+
+\begin{figure}[p]
+	\centerline{\includegraphics[width=0.95\textwidth]{prca_qc}}
+	\caption{Standard deviations inferred from the PRCA GWAS summary statistics versus the ones inferred from the reported GWAS allele frequencies.
+Only 100,000 variants are represented, at random.
+[TODO: REEXPORT FIGURE]}
+	\label{fig:prca_qc}
+\end{figure}
+
+\FloatBarrier
+
+\begin{figure}[p]
+	\centerline{\includegraphics[width=0.95\textwidth]{t1d_affy_qc}}
+	\caption{Standard deviations inferred from the T1D (Affymetrix) GWAS summary statistics versus the ones inferred from the reported GWAS allele frequencies.
+Only 100,000 variants are represented, at random.}
+	\label{fig:t1d_affy_qc}
+\end{figure}
+
+\begin{figure}[p]
+\centerline{\includegraphics[width=0.95\textwidth]{t1d_illu_qc}}
+\caption{Standard deviations inferred from the T1D (Illumina) GWAS summary statistics versus the ones inferred from the reported GWAS allele frequencies.
+Only 100,000 variants are represented, at random.}
+\label{fig:t1d_illu_qc}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\FloatBarrier
+
+\begin{figure}[p]
+	\centerline{\includegraphics[width=0.75\textwidth]{res-cad}}
+	\caption{Raw partial correlations (for \textit{all} models) for predicting CAD in the UK Biobank from external summary statistics.}
+	\label{fig:res_cad}
+\end{figure}
+
+\begin{figure}[p]
+	\centerline{\includegraphics[width=0.75\textwidth]{res-mdd}}
+	\caption{Raw partial correlations (for \textit{all} models) for predicting MDD in the UK Biobank from external summary statistics.}
+	\label{fig:res_mdd}
+\end{figure}
+
+\begin{figure}[p]
+	\centerline{\includegraphics[width=0.75\textwidth]{res-prca}}
+	\caption{Raw partial correlations (for \textit{all} models) for predicting PRCA in the UK Biobank from external summary statistics.}
+	\label{fig:res_prca}
+\end{figure}
+
+\begin{figure}[p]
+	\centerline{\includegraphics[width=0.95\textwidth]{res-t1d}}
+	\caption{Raw partial correlations (for \textit{all} models) for predicting T1D in the UK Biobank from external summary statistics.}
+	\label{fig:res_t1d}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\FloatBarrier
+
+\begin{figure}[p]
+	\centerline{\includegraphics[width=0.95\textwidth]{res-all-brca}}
+	\caption{Raw partial correlations (for \textit{all} models and \textit{all} sets) for predicting BRCA in the UK Biobank from external summary statistics.}
+	\label{fig:res_brca_all}
+\end{figure}
+
+\begin{figure}[p]
+	\centerline{\includegraphics[width=0.9\textwidth]{res-all-cad}}
+	\caption{Raw partial correlations (for \textit{all} models and \textit{all} sets) for predicting CAD in the UK Biobank from external summary statistics.}
+	\label{fig:res_cad_all}
+\end{figure}
+
+\begin{figure}[p]
+	\centerline{\includegraphics[width=0.9\textwidth]{res-all-mdd}}
+	\caption{Raw partial correlations (for \textit{all} models and \textit{all} sets) for predicting MDD in the UK Biobank from external summary statistics.}
+	\label{fig:res_mdd_all}
+\end{figure}
+
+\begin{figure}[p]
+	\centerline{\includegraphics[width=0.9\textwidth]{res-all-prca}}
+	\caption{Raw partial correlations (for \textit{all} models and \textit{all} sets) for predicting PRCA in the UK Biobank from external summary statistics.}
+	\label{fig:res_prca_all}
+\end{figure}
+
+\begin{figure}[p]
+	\centerline{\includegraphics[width=0.95\textwidth]{res-all-t1d}}
+	\caption{Raw partial correlations (for \textit{all} models and \textit{all} sets) for predicting T1D in the UK Biobank from external summary statistics.}
+	\label{fig:res_t1d_all}
 \end{figure}

### 2021-08-17 paper/paper-misspec.tex
finish later

@@
 
-\vspace*{5em}
+\clearpage
 
@@
 
+\vspace*{5em}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\clearpage
+%\clearpage
 
@@
 \section{Introduction}
+
+TODO
 
@@
 For half of the variants, we use 100\% of 300,000 individuals for GWAS, but use only 80\% for one quarter of variants and 60\% for the remaining quarter. 
-We then run C+T, LDpred2-inf, LDpred2(-grid), LDpred2-auto, lassosum, lassosum2, PRS-CS and SBayesR by using either the true per-variant GWAS sample size, the total sample size, or the imputed sample size.
+We then run C+T, LDpred2-inf, LDpred2(-grid), LDpred2-auto, lassosum, lassosum2, PRS-CS and SBayesR by using either the true per-variant GWAS sample sizes, the total sample size, or imputed sample sizes.
 However, we do not show results for SBayesR because it always diverged and for PRS-CS because the overlap with the LD reference they provide is too small.
 Averaged over 10 simulations, when providing true per-variant GWAS sample sizes, squared correlations between the polygenic scores and the simulated phenotypes are of 0.123 for C+T, 0.161 for lassosum, 0.169 for lassosum2, 0.159 for LDpred2(-grid), 0.140 for LDpred2-auto, and 0.141 for LDpred2-inf (Figure \ref{fig:simu-misN}).
-Results when using imputed sample sizes are quite similar.
+Results when using imputed (instead of true) sample sizes are quite similar.
 Note that C+T does not use this sample size information.
@@
 	\caption{Results for the simulations with sample size misspecification.
-The GWAS sample size is ``true'' when providing the true per-variant sample size, ``max'' when providing instead the maximum sample size as a unique value to be used for all variants,  ``imputed'' (see Methods in main text), or ``any'' when the method does not use this information (the case for C+T).
+The GWAS sample size is ``true'' when providing the true per-variant sample size, ``max'' when providing instead the maximum sample size as a unique value to be used for all variants,  ``imputed'' (Methods), or ``any'' when the method does not use this information (the case for C+T).
 Red bars correspond to using the LD with independent blocks (Methods).}
@@
 We therefore introduce two new parameters to make LDpred2-auto more robust and call this ``LDpred2-auto-rob'' here (Methods).
-Second, for lassosum2, results for a grid of parameters (over $\lambda$ and $\delta$) and quite smooth compared to LDpred2 (Figures \ref{fig:lassosum2-misN} and \ref{fig:ldpred2-misN}). 
+Second, for lassosum2, results for a grid of parameters (over $\lambda$ and $\delta$) are quite smooth compared to LDpred2 (Figures \ref{fig:lassosum2-misN} and \ref{fig:ldpred2-misN}). 
 In these simulations with misspecified sample sizes, it seems highly beneficial to use a small value for the SNP heritability hyper-parameter $h^2$ in LDpred2, e.g.\ a value of 0.02 or even 0.002 when the true value is 0.2 (Figure \ref{fig:ldpred2-misN}).
@@
 
-\cite{marchini2010genotype} showed that the IMPUTE INFO measure is highly concordant with the MACH measure $\hat{r}_j^2 = \frac{var(G_j)}{2 \hat{\theta_j} (1 - \hat{\theta_j})}$, where $\hat{\theta_j}$ is the estimated allele frequency of variant $j$.
-Therefore, when using the expected genotypes from imputation (dosages), their standard deviations are often lower than the expected value under Hardy-Weinberg equilibrium, because $\text{INFO}_j \approx \frac{var(G_j)}{2 \hat{\theta_j} (1 - \hat{\theta_j})}$.
-In simulations (cf.\ Methods section ``Data for simulations''), we verify that $\text{sd}(G_j)^{true} \approx \text{sd}(G_j)^{imp} / \sqrt{\text{INFO}_j}$ (Figures \ref{fig:compare-sd}.
+\cite{marchini2010genotype} showed that the IMPUTE INFO measure is highly concordant with the MACH measure $\hat{r}_j^2 = \frac{var(G_j)}{2 \hat{\theta}_j (1 - \hat{\theta}_j)}$, where $\hat{\theta}_j$ is the estimated allele frequency of variant $j$.
+Therefore, when using the expected genotypes from imputation (dosages), their standard deviations are often lower than the expected value under Hardy-Weinberg equilibrium, because $\text{INFO}_j \approx \frac{var(G_j)}{2 \hat{\theta}_j (1 - \hat{\theta}_j)}$.
+In simulations (cf.\ Methods section ``Data for simulations''), we verify that $\text{sd}(G_j)^{true} \approx \text{sd}(G_j)^{imp} / \sqrt{\text{INFO}_j}$ (Figure \ref{fig:compare-sd}).
 We also show that GWAS effect sizes $\hat{\gamma}$ computed from imputed dosages are over-estimated: $\hat{\gamma}_j^{true} \approx \hat{\gamma}_j^{imp} \cdot \sqrt{\text{INFO}_j}$ and $\text{se}(\hat{\gamma}_j)^{true} \approx \text{se}(\hat{\gamma}_j)^{imp} \cdot \sqrt{\text{INFO}_j}$ (Figures \ref{fig:compare-beta} and \ref{fig:compare-beta-se}).
@@
 We compute GWAS summary statistics from the dosage dataset and use these summary statistics to run LDpred2 and lassosum2 with either no correction of the summary statistics, or with one of the three corrections described above.
-The LD reference used by LDpred2 and lassosum2 is computed from the validation set using the dataset with ``true'' genotypes.
-For lassosum2 and LDpred2(-grid) which tune parameters using the validation set and can possibly use strong regularization, correcting for imputation quality slighly improve predictive predictive performance in these simulations (Figure \ref{fig:simu-info}).
-However, correcting for imputation quality can dramatically improve predictive performance for LDpred2-auto. 
-Moreover, new additions for robustness introduced before, namely LDpred2-low-h2, LDpred2-auto-rob, and forming independent blocks in the LD matrix improve predictive performance as well for all corrections (Figure \ref{fig:simu-info}).
+The LD reference used by LDpred2 and lassosum2 is computed from the validation set using the dataset with the ``true'' genotypes.
+For lassosum2 and LDpred2(-grid) which tune parameters using the validation set, correcting for imputation quality slightly improves predictive performance in these simulations (Figure \ref{fig:simu-info}).
+However, correcting for imputation quality can dramatically improves predictive performance for LDpred2-auto. 
+Moreover, new additions for robustness introduced before, namely LDpred2-low-h2, LDpred2-auto-rob, and forming independent blocks in the LD matrix also improve predictive performance for all corrections (Figure \ref{fig:simu-info}).
 
@@
 We first compare the standard QC proposed in \cite{prive2020ldpred2} (which ends up filtering on MAF here, which we call ``qc1'').
-We then also filter out the two long-range LD regions of chromosome 6 and 8 for the OncoArray summary statistics and remove around 500 variants when filtering on differences of MAFs between summary statistics and the validation dataset.
+We then also filter out the two long-range LD regions of chromosome 6 and 8 for the OncoArray summary statistics and remove around 500 variants when filtering on differences of MAFs between summary statistics and the validation dataset (``qc2'').
 As for the correction for the INFO scores, we use the first correction, ``sqrt\_info'', which is simple because it is equivalent to post-processing PGS effects by multiplying them by $\sqrt{\text{INFO}}$.
@@
 
-\subsection{}
+\subsection{Larger sets of variants and other phenotypes}
+
+We use external summary statistics for which INFO scores are reported (Table \ref{tab:sumstats}); they all have a very high mean INFO score (larger than 0.94), except for BRCA-iCOGS (0.841) and T1D-affy (0.885).
+QC plots comparing standard deviations usually look good (after the INFO score correction), except for coronary artery disease (CAD) summary statistics (Figures \ref{fig:brca_icogs_qc}-\ref{fig:t1d_illu_qc}).
+We show previous results as figure \ref{fig:res_brca} for other phenotypes in figures \ref{fig:res_cad}-\ref{fig:res_t1d}; most changes introduced before have little or no impact on the predictions, except for the second quality control performed for CAD when using LDpred2-auto (Figure \ref{fig:res_cad}).
+We then show results with two new sets of variants (Methods) in figures \ref{fig:res_brca_all}-\ref{fig:res_t1d_all}.
+[WHAT TO SAY ABOUT THIS? MAXTAG GOOD FOR SOME TRAITS, CLUMP NOT GREAT. PUT ONE FIGURE IN MAIN TEXT?]
 
@@
 
+TODO
+
 [TODO: TALK ABOUT HARD INFERRING N != INFO]
+
+[TALK ABOUT SUM(N\_EFF) VS N\_EFF(SUM)?]
 
@@
 
-We also use the UK Biobank data, and use the same individuals as described in the previous subsection.
+We also use the UK Biobank data, and use the same individuals as described in the previous section.
 We sample 10,000 individuals to form a validation set and use the remaining 352,307 individuals as test set.
@@
 As we expect these potential errors to be localized around errors in the genotype data, we apply a moving-average smoothing on the frequency differences to increase power to detect these errors and also reduce false positives.
-We define the threshold on these smoothed differences based on visual inspection of their histogram (0.03).
+We define the threshold (of 0.03) on these smoothed differences based on visual inspection of their histogram.
 This results in an initial set of 9,394,361 variants.
 
-From this large set of variants, we define two sets.
+From this large set of variants, we define the two sets.
 One is based on clumping, using a threshold $r^2 = 0.9$ over a radius of 100 Kbp and prioritizing HapMap3 variants and larger INFO scores.
 This results in a set ``clump'' of 2,465,478 variants, among which there are 554,655 of the initial HapMap3 variants.
-For the second set, we aim at maximizing the tagging of the initial 9,394,361 variants, i.e.
+For the second set, we aim at maximizing the tagging of all the initial 9,394,361 variants, i.e.
 $\sum_{i \in \text{all}} \max_{j \in \text{set}}{|R_{i,j}|}$ (inspired from the alternative sensitivity of \cite{agier2016systematic}).
-We design a greedy algorithm that first select all HapMap3 variants, then add one variant at a time, the one that maximizes the addition to this sum, until no variant can add more than 0.2.
+We design a greedy algorithm that first selects all HapMap3 variants, then adds one variant at a time, the one that maximizes the addition to this sum, until no variant can add more than 0.2.
 This results in a set ``maxtag'' of 2,029,086 variants.
@@
 Second, we also define nearly independent LD blocks using the optimal algorithm developed in \cite{prive2021optimal}.
-For different number of blocks and maximum number of variants in each block, we use the split with the minimum cost within the ones reducing the original number of non-zero values to less than 60\% (70\% for chromosome 6).
-Having a correlation matrix with independent blocks prevents the small errors in the algorithm (e.g.\ the Gibbs sampler in LDpred2) from propagating to the whole genome.
+For different numbers of blocks and maximum number of variants in each block, we use the split with the minimum cost within the ones reducing the original number of non-zero values to less than 60\% (70\% for chromosome 6).
+Having a correlation matrix with independent blocks prevents the small errors in the algorithm (e.g.\ the Gibbs sampler in LDpred2) from propagating to too many variants.
 It also makes running LDpred2 (and lassosum2) faster, taking about 60\% of the initial time (since only 60\% of the initial non-zero values of the correlation matrix are kept).

### 2021-08-31 paper/paper-misspec.tex
start discu

@@
 
+[SHOULD ADD THE LDAK-ALPHA IN THIS WORK? MAYBE LATER -> IN A THIRD VERSION]
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
+[WHY CHOOSED TO FOCUS ON LDPRED AND LASSOSUM -> COPY PREVIOUS INTRO + MORE CONTROL ON THE IMPLEMENTATION]
+
+[WHAT I WANTED TO INVESTIGATE]
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 For half of the variants, we use 100\% of 300,000 individuals for GWAS, but use only 80\% for one quarter of variants and 60\% for the remaining quarter. 
-We then run C+T, LDpred2-inf, LDpred2(-grid), LDpred2-auto, lassosum, lassosum2, PRS-CS and SBayesR by using either the true per-variant GWAS sample sizes, the total sample size, or imputed sample sizes.
+We then run C+T, LDpred2-inf, LDpred2(-grid), LDpred2-auto, lassosum, lassosum2 (Methods), PRS-CS and SBayesR \cite[]{prive2019making,prive2020ldpred2,mak2017polygenic,ge2019polygenic,lloyd2019improved} by using either the true per-variant GWAS sample sizes, the total sample size, or imputed sample sizes.
 However, we do not show results for SBayesR because it always diverged and for PRS-CS because the overlap with the LD reference they provide is too small.
@@
 For lassosum2 and LDpred2(-grid) which tune parameters using the validation set, correcting for imputation quality slightly improves predictive performance in these simulations (Figure \ref{fig:simu-info}).
-However, correcting for imputation quality can dramatically improves predictive performance for LDpred2-auto. 
+However, correcting for imputation quality can dramatically improve predictive performance for LDpred2-auto. 
 Moreover, new additions for robustness introduced before, namely LDpred2-low-h2, LDpred2-auto-rob, and forming independent blocks in the LD matrix also improve predictive performance for all corrections (Figure \ref{fig:simu-info}).
@@
 	\centerline{\includegraphics[width=0.95\textwidth]{brca_onco_qc}}
-	\caption{Standard deviations inferred from the OncoArray breast cancer GWAS summary statistics versus the ones inferred from the reported GWAS allele frequencies.
+	\caption{Standard deviations inferred from the OncoArray breast cancer GWAS summary statistics (Equation \ref{eq:approx-sd-log}) versus the ones inferred from the reported GWAS allele frequencies ($\sqrt{2 f (1 - f)}$).
 Only 100,000 HapMap3 variants are represented, at random.}
@@
 Therefore, when an accurate imputation INFO score is provided, this enables both to correct for it in the QC step we propose to possibly uncover problems with the summary statistics, and when building polygenic scores to improve prediction.
-We try this for the two breast cancer summary statistics.
+We apply this correction to the two breast cancer summary statistics.
 We first compare the standard QC proposed in \cite{prive2020ldpred2} (which ends up filtering on MAF here, which we call ``qc1'').
@@
 Although results for both QC used are very similar, correcting for the INFO score slightly improves prediction when deriving polygenic scores based on iCOGS summary statistics (Figure \ref{fig:res_brca}).
-All other improvements introduced before have little or no effect here [DEPRESSING; SHOULD REWORD?].
+All other improvements introduced before have little or no effect here, probably because misspecifications are much smaller than in the simulations.
 
@@
 
-\subsection{Larger sets of variants and other phenotypes}
+\subsection{Other phenotypes and larger sets of variants}
 
@@
 We show previous results as figure \ref{fig:res_brca} for other phenotypes in figures \ref{fig:res_cad}-\ref{fig:res_t1d}; most changes introduced before have little or no impact on the predictions, except for the second quality control performed for CAD when using LDpred2-auto (Figure \ref{fig:res_cad}).
-We then show results with two new sets of variants (Methods) in figures \ref{fig:res_brca_all}-\ref{fig:res_t1d_all}.
-[WHAT TO SAY ABOUT THIS? MAXTAG GOOD FOR SOME TRAITS, CLUMP NOT GREAT. PUT ONE FIGURE IN MAIN TEXT?]
+
+We then introduce two new sets of variants as possible replacement of HapMap3 variants (Methods).
+These sets include more variants (more than 2M), therefore possibly of lower quality on average. 
+We show results with these two new sets of variants (Methods) in figures \ref{fig:res_brca_all}-\ref{fig:res_t1d_all}.
+Using the ``maxtag'' set generally provides larger predictive performance than using HapMap3 variants, especially when predicting prostate cancer (Figure \ref{fig:res_prca_all}).
 
@@
 
-TODO
+Here we have investigated misspecifications in GWAS summary statistics, including when they are derived from imputed data, and when the information on sample size per variant is different for all variants because of meta-analysis.
+Using LDpred2 models \cite[]{prive2020ldpred2} and our reimplementation of the lassosum model \cite[]{mak2017polygenic} for deriving polygenic scores, we have introduced some changes to possibly make these models more robust to these misspecifications.
+Previously, we proposed a quality control based on comparing standard deviations derived from GWAS summary statistics with the ones from a reference panel \cite[]{prive2020ldpred2}.
+Here we show that we can refine this quality control and almost get perfect matches by deriving standard deviations from reported allele frequencies in the summary statistics (instead of from the reference panel, which can now be used to compare allele frequencies), and by also correcting for the INFO score (e.g.\ see figure \ref{fig:brca_onco_qc}).
+Using this refined QC, we were able to identify a potential issue with how principal components were derived in a set of breast cancer summary statistics.
+Fortunately this has practically no effect on the predictive performance of the derived polygenic scores.
+
+Note that, in this study, we use summary statistics for which this information (INFO scores and allele frequencies) is available, but this is far from being the case for all GWAS summary statistics. 
+We acknowledge that, in the case of a meta-analysis from multiple studies, providing a single INFO score may not be possible; would a weighted averaged INFO score work?
+Nevertheless, this quality control could be easily performed within each study before pooling results, to make sure that summary statistics have the best possible quality for follow-up analyses such as deriving polygenic scores.
+Finally, another information, the effect sample size per variant, is often missing from GWAS summary statistics. 
+We recall that if some studies of a meta-analysis have an imbalanced number of cases and controls, the global effective sample size should not be computed from the total numbers of cases and controls overall, but instead from the sum of each effective sample size.
+Indeed, take the extreme example of meta-analyzing two studies, one with 1000 cases and 0 controls, and another one with 0 cases and 1000 controls, then the effective sample size of the meta-analysis is 0, not 2000.
+Again, we note that an overestimated sample size can be detected from the QC plot we propose (when the slope is less than 1) for case-control studies using logistic regression; otherwise the standard deviation of the phenotype is also needed (Equation \ref{eq:approx-sd-lin}).
+
+[TODO: FINISH]
 
@@
 [TALK ABOUT SUM(N\_EFF) VS N\_EFF(SUM)?]
+
+[BOTH DISAPOINTING BUT REASSURING]
 
@@
 and impute $n_j^\text{eff}$ from equation \eqref{eq:approx-sd-log} using
-\begin{equation}\label{eq:imputeN}
+\begin{equation}\label{eq:imputeNeff}
 n_j^\text{eff} \approx \dfrac{4 / \text{var}(G_j) - \hat{\gamma}_j^2}{\text{se}(\hat{\gamma}_j)^2} ~.

### 2021-08-31 paper/paper-misspec-supp.tex
include suggestions

@@
 	\end{subfigure}
-	\caption{Multiple results (squared correlation r2 between polygenic score and phenotype) over a grid of parameters ($\lambda$ and $\delta$) for lassosum2 in one of the simulations with misspecified GWAS sample sizes (N).
-The top subfigure is using the maximum N, whereas the bottom subfigure is using the imputed N.
-Missing points represent models that were detected as divergent.}
+	\caption{Multiple results (squared correlation $r^2$ between polygenic score and phenotype) over a grid of parameters ($\lambda$ and $\delta$) for lassosum2 in one of the simulations with misspecified GWAS sample sizes ($n_j$).
+The top subfigure is using the maximum of $n_j$'s, whereas the bottom subfigure is using the imputed $n_j$'s.
+Missing points represent models that were detected as divergent.
+[TODO: USE GREEK LETTERS + FACETS FOR COMMON LEGEND SCALE]}
 	\label{fig:lassosum2-misN}
@@
 	\end{subfigure}
-	\caption{Multiple results (squared correlation r2 between polygenic score and phenotype) over a grid of parameters (the SNP heritability h2 and the proportion of causal variants p) for LDpred2(-grid) in one of the simulations with misspecified GWAS sample sizes (N).
-The top subfigure is using the maximum N, whereas the bottom subfigure is using the imputed N.
-Note that the small h2 values (0.002, 0.02, 0.06) were added and are part of the method we call ``LDpred2-low-h2'' in the main text.
+	\caption{Multiple results (squared correlation $r^2$ between polygenic score and phenotype) over a grid of parameters (the SNP heritability $h^2$ and the proportion of causal variants $p$) for LDpred2(-grid) in one of the simulations with misspecified GWAS sample sizes ($n_j$).
+The top subfigure is using the maximum of $n_j$'s, whereas the bottom subfigure is using the imputed $n_j$'s.
+Note that the small $h^2$ values (0.002, 0.02, 0.06) were added and are part of the method we call ``LDpred2-low-h2'' in the main text.
 Missing points represent models that were detected as divergent.}
@@
 	\centerline{\includegraphics[width=0.8\textwidth]{compare-sd}}
-	\caption{Standard deviations from ``true'' genotype calls versus from imputed dosages, colored by INFO scores. \label{fig:compare-sd}}
+	\caption{Standard deviations from ``true'' genotype calls versus from imputed dosages, colored by INFO scores. [TODO: ALL, MAKE SIDE BY SIDE WITH A/B + DESCRIBE MORE] \label{fig:compare-sd}}
 \end{figure}
@@
 	\centerline{\includegraphics[width=0.8\textwidth]{compare-info}}
-	\caption{INFO scores reported in the UK Biobank vs.\ INFO scores recomputed from the subset of 362,307 European individuals used in this paper. \label{fig:compare-info}}
+	\caption{INFO scores reported in the UK Biobank vs.\ INFO scores recomputed from the subset of 362,307 European individuals used in this paper.
+[TODO: FOR THE VARIANTS IN SIMU?] \label{fig:compare-info}}
 \end{figure}
@@
 	\caption{GWAS effect sizes and Z-scores from imputed dosages versus from multiple imputation (MI, 10 draws used), ``corrected'' and colored by INFO scores.
-This shows only one tenth of the simulated variants, because MI takes some time to run.
+This shows only one tenth of the simulated variants, because MI can be computationally intensive.
 \label{fig:compare-mi}}
@@
 	\centerline{\includegraphics[width=0.75\textwidth]{res-cad}}
-	\caption{Raw partial correlations (for \textit{all} models) for predicting CAD in the UK Biobank from external summary statistics.}
+	\caption{Raw partial correlations (for \textit{all} models) for predicting CAD in the UK Biobank from external summary statistics.
+These are computed using function \texttt{pcor} of R package bigstatsr where 95\% confidence intervals are obtained through Fisher's Z-transformation.}
 	\label{fig:res_cad}
@@
 	\centerline{\includegraphics[width=0.75\textwidth]{res-mdd}}
-	\caption{Raw partial correlations (for \textit{all} models) for predicting MDD in the UK Biobank from external summary statistics.}
+	\caption{Raw partial correlations (for \textit{all} models) for predicting MDD in the UK Biobank from external summary statistics.
+These are computed using function \texttt{pcor} of R package bigstatsr where 95\% confidence intervals are obtained through Fisher's Z-transformation.}
 	\label{fig:res_mdd}
@@
 	\centerline{\includegraphics[width=0.75\textwidth]{res-prca}}
-	\caption{Raw partial correlations (for \textit{all} models) for predicting PRCA in the UK Biobank from external summary statistics.}
+	\caption{Raw partial correlations (for \textit{all} models) for predicting PRCA in the UK Biobank from external summary statistics.
+These are computed using function \texttt{pcor} of R package bigstatsr where 95\% confidence intervals are obtained through Fisher's Z-transformation.}
 	\label{fig:res_prca}
@@
 	\centerline{\includegraphics[width=0.95\textwidth]{res-t1d}}
-	\caption{Raw partial correlations (for \textit{all} models) for predicting T1D in the UK Biobank from external summary statistics.}
+	\caption{Raw partial correlations (for \textit{all} models) for predicting T1D in the UK Biobank from external summary statistics.
+These are computed using function \texttt{pcor} of R package bigstatsr where 95\% confidence intervals are obtained through Fisher's Z-transformation.}
 	\label{fig:res_t1d}
@@
 	\centerline{\includegraphics[width=0.95\textwidth]{res-all-brca}}
-	\caption{Raw partial correlations (for \textit{all} models and \textit{all} sets) for predicting BRCA in the UK Biobank from external summary statistics.}
+	\caption{Raw partial correlations (for \textit{all} models and \textit{all} sets) for predicting BRCA in the UK Biobank from external summary statistics.
+These are computed using function \texttt{pcor} of R package bigstatsr where 95\% confidence intervals are obtained through Fisher's Z-transformation.}
 	\label{fig:res_brca_all}
@@
 	\centerline{\includegraphics[width=0.9\textwidth]{res-all-cad}}
-	\caption{Raw partial correlations (for \textit{all} models and \textit{all} sets) for predicting CAD in the UK Biobank from external summary statistics.}
+	\caption{Raw partial correlations (for \textit{all} models and \textit{all} sets) for predicting CAD in the UK Biobank from external summary statistics.
+These are computed using function \texttt{pcor} of R package bigstatsr where 95\% confidence intervals are obtained through Fisher's Z-transformation.}
 	\label{fig:res_cad_all}
@@
 	\centerline{\includegraphics[width=0.9\textwidth]{res-all-mdd}}
-	\caption{Raw partial correlations (for \textit{all} models and \textit{all} sets) for predicting MDD in the UK Biobank from external summary statistics.}
+	\caption{Raw partial correlations (for \textit{all} models and \textit{all} sets) for predicting MDD in the UK Biobank from external summary statistics.
+These are computed using function \texttt{pcor} of R package bigstatsr where 95\% confidence intervals are obtained through Fisher's Z-transformation.}
 	\label{fig:res_mdd_all}
@@
 	\centerline{\includegraphics[width=0.9\textwidth]{res-all-prca}}
-	\caption{Raw partial correlations (for \textit{all} models and \textit{all} sets) for predicting PRCA in the UK Biobank from external summary statistics.}
+	\caption{Raw partial correlations (for \textit{all} models and \textit{all} sets) for predicting PRCA in the UK Biobank from external summary statistics.
+These are computed using function \texttt{pcor} of R package bigstatsr where 95\% confidence intervals are obtained through Fisher's Z-transformation.}
 	\label{fig:res_prca_all}
@@
 	\centerline{\includegraphics[width=0.95\textwidth]{res-all-t1d}}
-	\caption{Raw partial correlations (for \textit{all} models and \textit{all} sets) for predicting T1D in the UK Biobank from external summary statistics.}
+	\caption{Raw partial correlations (for \textit{all} models and \textit{all} sets) for predicting T1D in the UK Biobank from external summary statistics.
+These are computed using function \texttt{pcor} of R package bigstatsr where 95\% confidence intervals are obtained through Fisher's Z-transformation.}
 	\label{fig:res_t1d_all}
@@
 	\centerline{\includegraphics[width=0.85\textwidth]{pseudoval}}
-	\caption{For one simulation, scores from validation vs.\ scores from pseudo-validation as described in \cite{mak2017polygenic} using either correlations (the default in lassosum) or p-values when computing local false discovery rates.}
+	\caption{For one simulation, scores from validation vs.\ scores from pseudo-validation as described in \cite{mak2017polygenic} using either \textbf{A:} correlations (the default in lassosum) or \textbf{B:} p-values when computing local false discovery rates.}
 	\label{fig:pseudoval}

### 2021-08-31 paper/paper-misspec.tex
include suggestions

@@
 Using 40,000 variants from chromosome 22 (Methods), we simulate quantitative phenotypes with a heritability of 20\% and 2000 causal variants.
-For half of the variants, we use 100\% of 300,000 individuals for GWAS, but use only 80\% for one quarter of variants and 60\% for the remaining quarter. 
-We then run C+T, LDpred2-inf, LDpred2(-grid), LDpred2-auto, lassosum, lassosum2 (Methods), PRS-CS and SBayesR \cite[]{prive2019making,prive2020ldpred2,mak2017polygenic,ge2019polygenic,lloyd2019improved} by using either the true per-variant GWAS sample sizes, the total sample size, or imputed sample sizes.
-However, we do not show results for SBayesR because it always diverged and for PRS-CS because the overlap with the LD reference they provide is too small.
+We then divide the 40,000 variants into three groups: for half of the variants, we use 100\% of 300,000 individuals for GWAS, but use only 80\% for one quarter of variants and 60\% for the remaining quarter.
+We then run C+T, lassosum, lassosum2 (Methods), LDpred2-inf, LDpred2(-grid), and LDpred2-auto \cite[]{prive2019making,mak2017polygenic,prive2020ldpred2} by using either the true per-variant GWAS sample sizes, the total sample size, or imputed sample sizes.
+Note that we initially included PRS-CS and SBayesR in our comparison \cite[]{ge2019polygenic,lloyd2019improved}.
+However, results for SBayesR always diverged and the overlap with the LD reference provided for PRS-CS was too small.
 Averaged over 10 simulations, when providing true per-variant GWAS sample sizes, squared correlations between the polygenic scores and the simulated phenotypes are of 0.123 for C+T, 0.161 for lassosum, 0.169 for lassosum2, 0.159 for LDpred2(-grid), 0.140 for LDpred2-auto, and 0.141 for LDpred2-inf (Figure \ref{fig:simu-misN}).
@@
 When using the total GWAS sample size instead of the per-variant sample sizes, predictive performance slightly decreases to 0.157 for lassosum and to 0.163 for lassosum2, but dramatically decreases for LDpred2 with new values of 0.134 for LDpred2-grid, 0.119 for LDpred2-auto, and 0.123 for LDpred2-inf (Figure \ref{fig:simu-misN}).
-Therefore, this extreme simulation scenario shows that LDpred2 can be sensitive to GWAS sample size misspecification, whereas lassosum (and lassosum2) seems little affected by this.
+This extreme simulation scenario shows that LDpred2 can be sensitive to GWAS sample size misspecification, whereas lassosum (and lassosum2) seems little affected by this.
 
@@
 	\centerline{\includegraphics[width=0.95\textwidth]{simu-misN}}
-	\caption{Results for the simulations with sample size misspecification.
+	\caption{Results for the simulations with sample size misspecification, averaged over 10 simulations for each scenario.
+Reported 95\% confidence intervals are computed from 10,000 non-parametric bootstrap replicates of the mean.
 The GWAS sample size is ``true'' when providing the true per-variant sample size, ``max'' when providing instead the maximum sample size as a unique value to be used for all variants,  ``imputed'' (Methods), or ``any'' when the method does not use this information (the case for C+T).
@@
 
-We then conduct further investigations to explain results of figure \ref{fig:simu-misN}.
-First, the reason why results for LDpred2-auto are the same as with LDpred2-inf is because it always converges to an infinitesimal model (p = 1) in these simulations.
-We therefore introduce two new parameters to make LDpred2-auto more robust and call this ``LDpred2-auto-rob'' here (Methods).
+We conduct further investigations to explain results of figure \ref{fig:simu-misN}.
+First, the results for LDpred2-auto are the same as with LDpred2-inf because it always converges to an infinitesimal model ($p = 1$) in these simulations.
+To solve this limitation, we introduce two new parameters to make LDpred2-auto more robust, referred to as ``LDpred2-auto-rob'' here (Methods).
 Second, for lassosum2, results for a grid of parameters (over $\lambda$ and $\delta$) are quite smooth compared to LDpred2 (Figures \ref{fig:lassosum2-misN} and \ref{fig:ldpred2-misN}). 
 In these simulations with misspecified sample sizes, it seems highly beneficial to use a small value for the SNP heritability hyper-parameter $h^2$ in LDpred2, e.g.\ a value of 0.02 or even 0.002 when the true value is 0.2 (Figure \ref{fig:ldpred2-misN}).
-We recall that using a small value for this hyper-parameter induces a larger regularization (shrinkage) on the effect sizes.
-When trying LDpred2(-grid) with these low values for $h^2$, we call this ``LDpred2-low-h2'' here.
+Indeed, using a small value for this hyper-parameter induces a larger regularization (shrinkage) on the effect sizes.
+When running LDpred2(-grid) with a grid of hyper-parameters including these low values for $h^2$, we refer to this as ``LDpred2-low-h2'' here.
 Results with LDpred2-low-h2 improves from 0.159 to 0.169 when using true sample sizes and from 0.134 to 0.163 when using the maximum sample size.
-Finally, we introduce a last change for robustness here; we form independent LD blocks in the LD matrix to prevent small errors in the Gibbs sampler to propagate to too many variants (Methods).
+Finally, we introduce a last change for robustness here: we form independent LD blocks in the LD matrix to prevent small errors in the Gibbs sampler to propagate to too many variants (Methods).
 This change seems to solve convergence issues of LDpred2 in these simulations (Figure \ref{fig:ldpred2-misN-corr2}) and further improves predictive performance for all LDpred2 methods (Figure \ref{fig:simu-misN}).
@@
 
-\cite{marchini2010genotype} showed that the IMPUTE INFO measure is highly concordant with the MACH measure $\hat{r}_j^2 = \frac{var(G_j)}{2 \hat{\theta}_j (1 - \hat{\theta}_j)}$, where $\hat{\theta}_j$ is the estimated allele frequency of variant $j$.
-Therefore, when using the expected genotypes from imputation (dosages), their standard deviations are often lower than the expected value under Hardy-Weinberg equilibrium, because $\text{INFO}_j \approx \frac{var(G_j)}{2 \hat{\theta}_j (1 - \hat{\theta}_j)}$.
-In simulations (cf.\ Methods section ``Data for simulations''), we verify that $\text{sd}(G_j)^{true} \approx \text{sd}(G_j)^{imp} / \sqrt{\text{INFO}_j}$ (Figure \ref{fig:compare-sd}).
-We also show that GWAS effect sizes $\hat{\gamma}$ computed from imputed dosages are over-estimated: $\hat{\gamma}_j^{true} \approx \hat{\gamma}_j^{imp} \cdot \sqrt{\text{INFO}_j}$ and $\text{se}(\hat{\gamma}_j)^{true} \approx \text{se}(\hat{\gamma}_j)^{imp} \cdot \sqrt{\text{INFO}_j}$ (Figures \ref{fig:compare-beta} and \ref{fig:compare-beta-se}).
-This is the first correction of summary statistics we try in the simulations below. 
+\cite{marchini2010genotype} showed that the IMPUTE INFO measure is highly concordant with the MACH measure $\hat{r}_j^2 = \frac{\text{var}(G_j)}{2 \hat{\theta}_j (1 - \hat{\theta}_j)}$, where $\hat{\theta}_j$ is the estimated allele frequency of $G_j$, the genotypes for variant $j$.
+Therefore, when using the expected genotypes from imputation (dosages), their standard deviations are often lower than the expected value under Hardy-Weinberg equilibrium, because $\text{INFO}_j \approx \frac{\text{var}(G_j)}{2 \hat{\theta}_j (1 - \hat{\theta}_j)}$.
+In simulations (cf.\ Methods section ``Data for simulations''), we verify that $\text{sd}(G_j)^{\text{true}} \approx \text{sd}(G_j)^{\text{imp}} / \sqrt{\text{INFO}_j}$ (Figure \ref{fig:compare-sd}).
+We also show that GWAS effect sizes $\hat{\gamma}$ computed from imputed dosages are over-estimated: $\hat{\gamma}_j^{\text{true}} \approx \hat{\gamma}_j^{\text{imp}} \cdot \sqrt{\text{INFO}_j}$ and $\text{se}(\hat{\gamma}_j)^{\text{true}} \approx \text{se}(\hat{\gamma}_j)^{\text{imp}} \cdot \sqrt{\text{INFO}_j}$ (Figures \ref{fig:compare-beta} and \ref{fig:compare-beta-se}).
+This is the first correction of summary statistics we consider in the simulations below. 
 Note that we recompute INFO scores for the subset of 362,307 European individuals used in this paper as they can differ substantially from the ones reported by the UK Biobank for the whole data (Figure \ref{fig:compare-info}).
 As a second option, instead of using dosages to compute the GWAS summary statistics, it has been argued that using multiple imputation (MI) would be more appropriate \cite[]{Palmer2016}.
-In simulations, we show that $\hat{\gamma}_j^{MI} \approx \hat{\gamma}_j^{imp} \cdot \text{INFO}_j$ and $Z_j^{MI} \approx Z_j^{imp} \cdot \text{INFO}_j$, where $Z = \hat{\gamma} / \text{se}(\hat{\gamma})$ (Figure \ref{fig:compare-mi}).
-This is the second correction of summary statistics we try in the simulations below, along with $n_j = N \cdot \text{INFO}_j$. 
-Finally, we try an in-between solution as a third correction, using $\hat{\gamma}_j^{imp} \cdot \text{INFO}_j$, $\text{se}(\hat{\gamma}_j)^{imp} \cdot \sqrt{\text{INFO}_j}$, and $N \cdot \text{INFO}_j$.
+In simulations, we show that $\hat{\gamma}_j^{\text{MI}} \approx \hat{\gamma}_j^{\text{imp}} \cdot \text{INFO}_j$ and $Z_j^{\text{MI}} \approx Z_j^{\text{imp}} \cdot \text{INFO}_j$, where $Z = \hat{\gamma} / \text{se}(\hat{\gamma})$ (Figure \ref{fig:compare-mi}).
+This is the second correction of summary statistics we implement in the simulations below, along with $n_j = N \cdot \text{INFO}_j$. 
+Finally, we consider an in-between solution as a third correction, using $\hat{\gamma}_j = \hat{\gamma}_j^{\text{imp}} \cdot \text{INFO}_j$, $\text{se}(\hat{\gamma}_j) = \text{se}(\hat{\gamma}_j)^{\text{imp}} \cdot \sqrt{\text{INFO}_j}$, and $n_j = N \cdot \text{INFO}_j$.
 
-Using 40,000 variants from chromosome 22, we simulate quantitative phenotypes assuming an heritability of 20\% and 2000 causal variants using the ``true'' dataset (cf.\ Methods section ``Data for simulations'').
+Using 40,000 variants from chromosome 22, we simulate quantitative phenotypes assuming a heritability of 20\% and 2000 causal variants using the ``true'' dataset (cf.\ Methods section ``Data for simulations'').
 We compute GWAS summary statistics from the dosage dataset and use these summary statistics to run LDpred2 and lassosum2 with either no correction of the summary statistics, or with one of the three corrections described above.
@@
 
+[SHOULD I USE ``INFO'' INSTEAD? BUT I NEED TO RERUN EVERYTHING BECAUSE I CAN'T JUST POSTPROCESS RESULTS. AND NOT BETTER FOR LDPRED2-LOW-H2 AND LDPRED2-AUTO-ROB]
+
 \begin{figure}[h]
 	\centerline{\includegraphics[width=0.95\textwidth]{simu-info}}
-	\caption{Results of predictive performance for the simulations using GWAS summary statistics from imputed dosage data.
-Correction ``sqrt\_info'' corresponds to using $\hat{\gamma}_j^{imp} \cdot \sqrt{\text{INFO}_j}$ and $\text{se}(\hat{\gamma}_j)^{imp} \cdot \sqrt{\text{INFO}_j}$.
-	Correction ``info'' corresponds to using $\hat{\gamma}_j^{imp} \cdot \text{INFO}_j$  and $N \cdot \text{INFO}_j$.
-Correction ``in\_between'' corresponds to using $\hat{\gamma}_j^{imp} \cdot \text{INFO}_j$, $\text{se}(\hat{\gamma}_j)^{imp} \cdot \sqrt{\text{INFO}_j}$, and $N \cdot \text{INFO}_j$.
+	\caption{Results of predictive performance for the simulations using GWAS summary statistics from imputed dosage data, averaged over 10 simulations for each scenario.
+Reported 95\% confidence intervals are computed from 10,000 non-parametric bootstrap replicates of the mean.
+Correction ``sqrt\_info'' corresponds to using $\hat{\gamma}_j^{\text{imp}} \cdot \sqrt{\text{INFO}_j}$ and $\text{se}(\hat{\gamma}_j)^{\text{imp}} \cdot \sqrt{\text{INFO}_j}$.
+	Correction ``info'' corresponds to using $\hat{\gamma}_j^{\text{imp}} \cdot \text{INFO}_j$  and $N \cdot \text{INFO}_j$.
+Correction ``in\_between'' corresponds to using $\hat{\gamma}_j^{\text{imp}} \cdot \text{INFO}_j$, $\text{se}(\hat{\gamma}_j)^{\text{imp}} \cdot \sqrt{\text{INFO}_j}$, and $N \cdot \text{INFO}_j$.
 Red bars correspond to using the LD with independent blocks (Methods).}
@@
 PC19 from the UK Biobank was previously reported to capture LD structure in region 70-91 Mbp of chromsome 6 \cite[]{prive2020efficient}.
-In these simulations, the same bias as in figure \ref{fig:hist-bad-pos}B is observed for the variants in this region (Figure \ref{fig:gwas_bad_pc}), therefore confirming our hypothesis.
+In these simulations, the same bias as in figure \ref{fig:hist-bad-pos}B is observed for the variants in this region (Figure \ref{fig:gwas_bad_pc}), confirming our hypothesis.
 
@@
 
-Therefore, when an accurate imputation INFO score is provided, this enables both to correct for it in the QC step we propose to possibly uncover problems with the summary statistics, and when building polygenic scores to improve prediction.
+Therefore, providing an accurate imputation INFO score is important for two reasons.
+First, it allows for correcting for a reduced standard deviation when using imputed data in the QC step we propose, in order to better uncover problems with the summary statistics.
+Second, it may lead to an improved prediction when building polygenic scores using one of the proposed corrections.
 We apply this correction to the two breast cancer summary statistics.
@@
 	\centerline{\includegraphics[width=0.95\textwidth]{res-brca}}
-	\caption{Raw partial correlations (for \textit{all} models) for predicting breast cancer in the UK Biobank when using either the OncoArray or the iCOGS summary statistics.}
+	\caption{Raw partial correlations (for \textit{all} models) for predicting breast cancer in the UK Biobank when using either the OncoArray or the iCOGS summary statistics.
+These are computed using function \texttt{pcor} of R package bigstatsr where 95\% confidence intervals are obtained through Fisher's Z-transformation.}
 	\label{fig:res_brca}
@@
 We use external summary statistics for which INFO scores are reported (Table \ref{tab:sumstats}); they all have a very high mean INFO score (larger than 0.94), except for BRCA-iCOGS (0.841) and T1D-affy (0.885).
-QC plots comparing standard deviations usually look good (after the INFO score correction), except for coronary artery disease (CAD) summary statistics (Figures \ref{fig:brca_icogs_qc}-\ref{fig:t1d_illu_qc}).
+QC plots comparing standard deviations usually show little deviation from $x=y$ (after the INFO score correction), except for coronary artery disease (CAD) summary statistics (Figures \ref{fig:brca_icogs_qc}-\ref{fig:t1d_illu_qc}).
 We show previous results as figure \ref{fig:res_brca} for other phenotypes in figures \ref{fig:res_cad}-\ref{fig:res_t1d}; most changes introduced before have little or no impact on the predictions, except for the second quality control performed for CAD when using LDpred2-auto (Figure \ref{fig:res_cad}).
@@
 To define phenotypes in the UK Biobank, we first map ICD10 and ICD9 codes (UKBB fields 40001, 40002, 40006, 40013, 41202, 41270 and 41271) to phecodes using R package PheWAS \cite[]{carroll2014r,wu2019mapping}.
-We use external published GWAS summary statistics listed in table \ref{tab:sumstats} to derive polygenic scores.
+We use published GWAS summary statistics listed in table \ref{tab:sumstats} to derive polygenic scores.
 
@@
 		\hline
-		Breast cancer (BRCA) [iCOGS] & \cite{michailidou2017association} & ~~87036.6 & 11,792,542 & 1,051,242 & 0.841 \\
-		Breast cancer (BRCA) [OncoArray] & \cite{michailidou2017association} & 104441.6 & 11,792,542 & 1,054,233 & 0.968 \\
-		Type 1 diabetes (T1D) [Affymetrix] & \cite{censin2017childhood} & ~~~~5515.6  & ~~8,996,866 & ~~~934,712 & 0.885 \\	
-		Type 1 diabetes (T1D) [Illumina] & \cite{censin2017childhood} & ~~~~7982.0  & ~~8,996,866 & ~~~949,334 & 0.942 \\
-		Prostate cancer (PRCA) & \cite{schumacher2018association} & 135316.1 & 20,370,946 & ~~~818,400 & 0.969 \\
-		Depression (MDD) [without UKBB] & \cite{wray2018genome} & 110463.7 & ~~9,874,289 & 1,049,455 & 0.968 \\
-		Coronary artery disease (CAD) & \cite{nikpay2015comprehensive} & 129014.3 & ~~9,455,778 & 1,052,200 & 0.982 \\
+		Breast cancer (BRCA) [iCOGS] & \cite{michailidou2017association} & ~~87,037 & 11,792,542 & 1,051,242 & 0.841 \\
+		Breast cancer (BRCA) [OncoArray] & \cite{michailidou2017association} & 104,442 & 11,792,542 & 1,054,233 & 0.968 \\
+		Type 1 diabetes (T1D) [Affymetrix] & \cite{censin2017childhood} & ~~~~~5516  & ~~8,996,866 & ~~~934,712 & 0.885 \\	
+		Type 1 diabetes (T1D) [Illumina] & \cite{censin2017childhood} & ~~~~~7982  & ~~8,996,866 & ~~~949,334 & 0.942 \\
+		Prostate cancer (PRCA) & \cite{schumacher2018association} & 135,316 & 20,370,946 & ~~~818,400 & 0.969 \\
+		Depression (MDD) [without UKBB] & \cite{wray2018genome} & 110,463 & ~~9,874,289 & 1,049,455 & 0.968 \\
+		Coronary artery disease (CAD) & \cite{nikpay2015comprehensive} & 129,014 & ~~9,455,778 & 1,052,200 & 0.982 \\
 		\hline
@@
 
-From this large set of variants, we define the two sets.
+We then define the two sets from this large set of variants.
 One is based on clumping, using a threshold $r^2 = 0.9$ over a radius of 100 Kbp and prioritizing HapMap3 variants and larger INFO scores.
@@
 For the second set, we aim at maximizing the tagging of all the initial 9,394,361 variants, i.e.
-$\sum_{i \in \text{all}} \max_{j \in \text{set}}{|R_{i,j}|}$ (inspired from the alternative sensitivity of \cite{agier2016systematic}).
+$\sum_{i \in \text{all}} \max_{j \in \text{set}}{|R_{i,j}|}$, where $R_{i,j}$ is the correlation between variants $i$ and $j$ (inspired from the alternative sensitivity of \cite{agier2016systematic}).
 We design a greedy algorithm that first selects all HapMap3 variants, then adds one variant at a time, the one that maximizes the addition to this sum, until no variant can add more than 0.2.
@@
 
-Instead of using a regularized version of the correlation matrix $R$ parameterized by $s$, $R_s = (1 - s) R + s I$, we use $R_{\delta} = R + \delta I$, which makes it clearer that lassosum is also using L2-regularization (therefore elastic-net).
+Instead of using a regularized version of the correlation matrix $R$ parameterized by $s$, $R_s = (1 - s) R + s I$ (where $0 < s \leq 1$), we use $R_{\delta} = R + \delta I$ (where $\delta > 0$), which makes it clearer that lassosum is also using L2-regularization (therefore elastic-net).
 Then, from \cite{mak2017polygenic}, the solution from lassosum can be obtained by iteratively updating 
@@
 \]
-Following notations from \cite{prive2020ldpred2} and denoting $\widetilde{X} = \frac{1}{\sqrt{n-1}} C_n G S^{-1}$, where $G$ is the genotype matrix, $C_n$ is the centering matrix and $S$ is the diagonal matrix of standard deviations of the columns of $G$.
-Then $\widetilde{X}_j^T \widetilde{X} = R_{j,.} = R_{.,j}^T$ and $\widetilde{X}_j^T \widetilde{X}_j = 1$. Moreover, $u_j^{(t)} = \beta_j^{(t-1)} + \widehat{\beta}_j - R_{.,j}^T \beta^{(t-1)}$,
+Following the notations from \cite{prive2020ldpred2} and denoting $\widetilde{X} = \frac{1}{\sqrt{n-1}} C_n G S^{-1}$, where $G$ is the genotype matrix, $C_n$ is the centering matrix and $S$ is the diagonal matrix of standard deviations of the columns of $G$.
+Then $\widetilde{X}_j^T \widetilde{X} = R_{j,.} = R_{.,j}^T$ and $\widetilde{X}_j^T \widetilde{X}_j = 1$.
+Moreover, using the notations from \cite{prive2020ldpred2}, $u_j^{(t)} = \widehat{\beta}_j - R_{.,j}^T \beta^{(t-1)} + \beta_j^{(t-1)}$,
 where $r_j = \widehat{\beta}_j =  \dfrac{\widehat{\gamma}_j}{\sqrt{n_j ~ \text{se}(\widehat{\gamma}_j)^2 + \widehat{\gamma}_j^2}}$ and $\widehat{\gamma}_j$ is the GWAS effect of variant $j$ and $n$ is the GWAS sample size \cite[]{mak2017polygenic,prive2021high}.
 Then the most time-consuming part is computing $R_{.,j}^T \beta^{(t-1)}$.
-To make this faster, instead of computing $R_{.,j}^T \beta^{(t-1)}$ at each iteration ($j$ and $t$), we can start with a vector with only 0s initially (for all $j$) since $\beta^{(0)} \equiv 0$, and then updating this vector when $\beta_j^{(t)} \neq \beta_j^{(t-1)}$ only.
+To make this faster, instead of computing $R_{.,j}^T \beta^{(t-1)}$ at each iteration ($j$ and $t$), we can start with an initial vector of 0s only (for all $j$) since $\beta^{(0)} \equiv 0$, and then updating this vector when $\beta_j^{(t)} \neq \beta_j^{(t-1)}$ only.
 Note that only positions $k$ for which $R_{k,j} \neq 0$ must be updated in this vector $R_{.,j}^T \beta^{(t-1)}$.
@@
 We try $\delta \in \{0.001, 0.005, 0.02, 0.1, 0.6, 3\}$ by default in lassosum2, instead of $s \in \{0.2, 0.5, 0.8, 1.0\}$ in lassosum.
-For $\lambda$, while the default in lassosum uses a sequence of 20 values equally spaced on a log scale between 0.1 and 0.001; we now use a sequence between $\lambda_0$ and $\lambda_0 / 100$ by default in lassosum2, where $\lambda_0 = \max_j \left|\widehat{\beta}_j\right|$ is the minimum $\lambda$ for which no variable enters the model because the L1-regularization is too strong.
+For $\lambda$, the default in lassosum uses a sequence of 20 values equally spaced on a log scale between 0.1 and 0.001.
+We use instead a sequence between $\lambda_0$ and $\lambda_0 / 100$ by default in lassosum2, where $\lambda_0 = \max_j \left|\widehat{\beta}_j\right|$ is the minimum $\lambda$ for which no variable enters the model because the L1-regularization is too strong.
 Note that we do not provide an ``auto'' version using pseudo-validation (as in \cite{mak2017polygenic}) as we have not found it to be very robust (Figure \ref{fig:pseudoval}).

### 2021-09-02 paper/paper-misspec.tex
finish discu

@@
 First, it allows for correcting for a reduced standard deviation when using imputed data in the QC step we propose, in order to better uncover problems with the summary statistics.
-Second, it may lead to an improved prediction when building polygenic scores using one of the proposed corrections.
+Second, using one of the proposed corrections may lead to an improved prediction when deriving polygenic scores.
 We apply this correction to the two breast cancer summary statistics.
@@
 Here we have investigated misspecifications in GWAS summary statistics, including when they are derived from imputed data, and when the information on sample size per variant is different for all variants because of meta-analysis.
-Using LDpred2 models \cite[]{prive2020ldpred2} and our reimplementation of the lassosum model \cite[]{mak2017polygenic} for deriving polygenic scores, we have introduced some changes to possibly make these models more robust to these misspecifications.
-Previously, we proposed a quality control based on comparing standard deviations derived from GWAS summary statistics with the ones from a reference panel \cite[]{prive2020ldpred2}.
-Here we show that we can refine this quality control and almost get perfect matches by deriving standard deviations from reported allele frequencies in the summary statistics (instead of from the reference panel, which can now be used to compare allele frequencies), and by also correcting for the INFO score (e.g.\ see figure \ref{fig:brca_onco_qc}).
+Previously, we proposed a quality control (QC) based on comparing standard deviations derived from GWAS summary statistics with the ones from a reference panel \cite[]{prive2020ldpred2}.
+Here we show that we can refine this quality control by deriving the second standard deviations directly from the reported allele frequencies in the GWAS summary statistics, and by correcting for the INFO scores of the GWAS (e.g.\ see figure \ref{fig:brca_onco_qc}).
 Using this refined QC, we were able to identify a potential issue with how principal components were derived in a set of breast cancer summary statistics.
 Fortunately this has practically no effect on the predictive performance of the derived polygenic scores.
+Additional QC can also be performed, e.g.\ comparing reported GWAS allele frequencies with the ones from the LD reference panel, e.g.\ to detect genotyping or allelic errors.
+We do perform this additional QC as part of ``qc2'' here, and also when designing the large set of variants from the UK Biobank (Methods). 
+One can also run other QC tools such as DENTIST \cite[]{chen2021improved}.
 
-Note that, in this study, we use summary statistics for which this information (INFO scores and allele frequencies) is available, but this is far from being the case for all GWAS summary statistics. 
-We acknowledge that, in the case of a meta-analysis from multiple studies, providing a single INFO score may not be possible; would a weighted averaged INFO score work?
+Note that, in this study, we use summary statistics that include extended information (e.g.\ INFO scores and allele frequencies), yet most GWAS summary statistics do not. 
+We acknowledge that, in the case of a meta-analysis from multiple studies, providing a single INFO score per variant may not be possible; would a weighted averaged INFO score work?
 Nevertheless, this quality control could be easily performed within each study before pooling results, to make sure that summary statistics have the best possible quality for follow-up analyses such as deriving polygenic scores.
-Finally, another information, the effect sample size per variant, is often missing from GWAS summary statistics. 
+Another information, the effective sample size per variant, is often missing from GWAS summary statistics.
+Even the total effective sample size is sometimes difficult to assess.
+Indeed, often only the total number of cases and controls are reported in the abstract, and where numbers possibly includes sizes before QC; therefore one has to dig into supplementary materials to find this information.
 We recall that if some studies of a meta-analysis have an imbalanced number of cases and controls, the global effective sample size should not be computed from the total numbers of cases and controls overall, but instead from the sum of each effective sample size.
 Indeed, take the extreme example of meta-analyzing two studies, one with 1000 cases and 0 controls, and another one with 0 cases and 1000 controls, then the effective sample size of the meta-analysis is 0, not 2000.
-Again, we note that an overestimated sample size can be detected from the QC plot we propose (when the slope is less than 1) for case-control studies using logistic regression; otherwise the standard deviation of the phenotype is also needed (Equation \ref{eq:approx-sd-lin}).
+Fortunately, an overestimated sample size can be detected from the QC plot we propose (where the slope is then less than 1) for case-control studies using logistic regression; otherwise the standard deviation of the phenotype is also needed (Equation \ref{eq:approx-sd-lin}).
 
-[TODO: FINISH]
-
-[TODO: TALK ABOUT HARD INFERRING N != INFO]
-
-[TALK ABOUT SUM(N\_EFF) VS N\_EFF(SUM)?]
-
-[BOTH DISAPOINTING BUT REASSURING]
+We have assessed the impact of these misspecifications in GWAS summary statistics on the predictive performance of some polygenic score methods.
+Using LDpred2 models \cite[]{prive2020ldpred2} and our reimplementation of the lassosum model \cite[]{mak2017polygenic} for deriving polygenic scores, we have introduced and investigated some changes to possibly make these models more robust to misspecifications.
+Overall, these changes have provided large improvements of predictive performance in simulations with large misspecifications.
+However, they have almost no effect when using the published GWAS summary statistics, except for breast cancer and coronary artery disease summary statistics.
+This is a little disappointing, but more importantly it is also reassuring because it means that the summary statistics we use in this study are usually of good quality for follow-up analyses such as deriving polygenic scores.
 
@@
 		Prostate cancer (PRCA) & \cite{schumacher2018association} & 135,316 & 20,370,946 & ~~~818,400 & 0.969 \\
-		Depression (MDD) [without UKBB] & \cite{wray2018genome} & 110,463 & ~~9,874,289 & 1,049,455 & 0.968 \\
+		Depression (MDD) [without UKBB] & \cite{wray2018genome} & 110,464 & ~~9,874,289 & 1,049,455 & 0.968 \\
 		Coronary artery disease (CAD) & \cite{nikpay2015comprehensive} & 129,014 & ~~9,455,778 & 1,052,200 & 0.982 \\

### 2021-09-02 paper/paper-misspec-supp.tex
better figs

@@
 \begin{figure}[h]
-	\centerline{\includegraphics[width=0.8\textwidth]{simu-qc-plot}}
+	\centerline{\includegraphics[width=0.7\textwidth]{simu-qc-plot}}
 	\caption{Quality control plot, as proposed in \cite{prive2020ldpred2}, for the simulations with sample size misspecification.
@@
 \begin{figure}[p]
-	\begin{subfigure}{\textwidth}
-		\centering
-		\includegraphics[width=.8\linewidth]{lassosum2-mis-maxN-corr1}
-	\end{subfigure}\vspace{2em}
-	\begin{subfigure}{\textwidth}
-		\centering
-		\includegraphics[width=.8\linewidth]{lassosum2-mis-imputeN-corr1}
-	\end{subfigure}
+	\centerline{\includegraphics[width=0.95\textwidth]{lassosum2-misN}}
 	\caption{Multiple results (squared correlation $r^2$ between polygenic score and phenotype) over a grid of parameters ($\lambda$ and $\delta$) for lassosum2 in one of the simulations with misspecified GWAS sample sizes ($n_j$).
-The top subfigure is using the maximum of $n_j$'s, whereas the bottom subfigure is using the imputed $n_j$'s.
-Missing points represent models that were detected as divergent.
-[TODO: USE GREEK LETTERS + FACETS FOR COMMON LEGEND SCALE]}
+The different colors indicate whether we use the maximum of $n_j$'s (``maxN'') or the imputed $n_j$'s (``imputeN''), and the normal LD matrix (``normal'') or the one with independent LD blocks (``with\_indep\_ld'').
+Missing points represent models that were detected as divergent.}
 	\label{fig:lassosum2-misN}
@@
 \begin{figure}[p]
-	\begin{subfigure}{\textwidth}
-		\centering
-		\includegraphics[width=.8\linewidth]{ldpred2-mis-maxN-corr1}
-	\end{subfigure}\vspace{2em}
-	\begin{subfigure}{\textwidth}
-		\centering
-		\includegraphics[width=.8\linewidth]{ldpred2-mis-imputeN-corr1}
-	\end{subfigure}
+	\centerline{\includegraphics[width=0.95\textwidth]{ldpred2-misN}}
 	\caption{Multiple results (squared correlation $r^2$ between polygenic score and phenotype) over a grid of parameters (the SNP heritability $h^2$ and the proportion of causal variants $p$) for LDpred2(-grid) in one of the simulations with misspecified GWAS sample sizes ($n_j$).
-The top subfigure is using the maximum of $n_j$'s, whereas the bottom subfigure is using the imputed $n_j$'s.
+The different colors indicate whether we use the maximum of $n_j$'s (``maxN'') or the imputed $n_j$'s (``imputeN''), and the normal LD matrix (``normal'') or the one with independent LD blocks (``with\_indep\_ld'').
 Note that the small $h^2$ values (0.002, 0.02, 0.06) were added and are part of the method we call ``LDpred2-low-h2'' in the main text.
@@
 	\label{fig:ldpred2-misN}
-\end{figure}
-
-\begin{figure}[p]
-	\begin{subfigure}{\textwidth}
-		\centering
-		\includegraphics[width=.8\linewidth]{ldpred2-mis-maxN-corr2}
-	\end{subfigure}\vspace{2em}
-	\begin{subfigure}{\textwidth}
-		\centering
-		\includegraphics[width=.8\linewidth]{ldpred2-mis-imputeN-corr2}
-	\end{subfigure}
-	\caption{Same results as figure \ref{fig:ldpred2-misN}, but using the LD with independent blocks.}
-	\label{fig:ldpred2-misN-corr2}
 \end{figure}

### 2021-09-02 paper/paper-misspec.tex
better figs

@@
 Finally, we introduce a last change for robustness here: we form independent LD blocks in the LD matrix to prevent small errors in the Gibbs sampler to propagate to too many variants (Methods).
-This change seems to solve convergence issues of LDpred2 in these simulations (Figure \ref{fig:ldpred2-misN-corr2}) and further improves predictive performance for all LDpred2 methods (Figure \ref{fig:simu-misN}).
+This change seems to solve convergence issues of LDpred2 in these simulations (Figure \ref{fig:ldpred2-misN}) and further improves predictive performance for all LDpred2 methods (Figure \ref{fig:simu-misN}).
 

### 2021-09-07 code/greedy-maxtag.cpp
export code

@@ -1,0 +1,123 @@
+/******************************************************************************/
+
+#include <Rcpp.h>
+#include <unordered_set>
+using namespace Rcpp;
+using namespace std;
+
+// Enable C++11 via this plugin (Rcpp 0.10.3 or later)
+// [[Rcpp::plugins(cpp11)]]
+
+/******************************************************************************/
+
+double compute_add(const std::vector<size_t>& p,
+                   const IntegerVector& i,
+                   const NumericVector& x,
+                   const NumericVector& tagged,
+                   int j) {
+
+  double sum_add = 0;
+
+  for (size_t k = p[j]; k < p[j + 1]; k++) {
+
+    int    ind = i[k];
+    double val = x[k];
+
+    double add = val - tagged[ind];
+    if (add > 0) sum_add += add;
+  }
+
+  return sum_add;
+}
+
+/******************************************************************************/
+
+// [[Rcpp::export]]
+ListOf<NumericVector> set_max_tag(const std::vector<size_t>& p,
+                                  const IntegerVector& i,
+                                  const NumericVector& x0,
+                                  const LogicalVector& select,
+                                  const LogicalVector& exclude,
+                                  const NumericVector& sqrt_info,
+                                  double min_add,
+                                  bool remove_diag) {
+
+  NumericVector x = Rcpp::abs(x0);
+  LogicalVector removed = Rcpp::clone(exclude);
+
+  int m = p.size() - 1;
+  int m_remain = Rcpp::sum(!removed);
+
+  NumericVector can_add(m, 0.0);
+  for (int j = 0; j < m; j++) {
+    for (size_t k = p[j]; k < p[j + 1]; k++) {
+      if (remove_diag && i[k] == j) {
+        x[k] = 0;
+      } else {
+        x[k] *= sqrt_info[j];
+        if (!removed[j]) can_add[j] += x[k];
+      }
+    }
+  }
+  can_add[select] = R_PosInf;
+
+  NumericVector tagged(m, 0.0), added(m, NA_REAL);
+
+  std::unordered_set<int> ind_has_changed, ind_could_change;
+
+  do {
+
+    int j_max = which_max(can_add);
+    if (can_add[j_max] == 0) break;
+
+    // Rcout << j_max + 1 << " -> " << can_add[j_max] << std::endl;
+
+    added[j_max] = can_add[j_max];
+    can_add[j_max] = 0;
+    removed[j_max] = true;
[…]

### 2021-09-07 code/investigate-GWAS-from-imputed.R

@@ -1,0 +1,140 @@
+library(bigsnpr)
+library(ggplot2)
+theme_set(theme_bigstatsr())
+
+simu_true <- snp_attach("data/ukbb4simu.rds")
+G_true <- simu_true$genotypes
+INFO0 <- simu_true$map$info   # the ones reported in UKBB
+NCORES <- nb_cores()
+
+pheno <- snp_simuPheno(G_true, h2 = 0.2, M = 2000, ncores = NCORES)
+y <- pheno$pheno
+
+load("data/ukbb4simu_ind.RData")
+gwas_true <- big_univLinReg(G_true, y[ind.gwas], ind.train = ind.gwas, ncores = NCORES)
+
+simu_imp <- snp_attach("data/ukbb4simu_imp.rds")
+INFO <- simu_imp$map$info  # the ones recomputed for the subset
+qplot(INFO, INFO0) +
+  geom_abline(col = "red", lwd = 2) +
+  coord_equal() +
+  labs(x = "INFO scores recomputed from the European subset",
+       y = "INFO scores reported in the UK Biobank")
+# ggsave("figures/compare-info.png", width = 7, height = 7)
+
+G_imp <- simu_imp$genotypes
+gwas_imp <- big_univLinReg(G_imp, y[ind.gwas], ind.train = ind.gwas, ncores = NCORES)
+
+scale_true <- big_scale()(G_true, ind.row = ind.gwas)
+scale_imp <- big_scale()(G_imp, ind.row = ind.gwas)
+# scale_freq <- sqrt(scale_imp$center * (1 - scale_imp$center / 2))
+
+source("https://raw.githubusercontent.com/privefl/paper4-bedpca/master/code/plot_grid2.R")
+plot_grid2(list(
+  qplot(scale_true$scale, scale_imp$scale, color = INFO) +
+    geom_abline(color = "red") +
+    coord_equal() +
+    scale_color_viridis_c() +
+    guides(color = guide_colorbar(barheight = 15, ticks.linewidth = 2)) +
+    labs(y = "SD_imp", x = "SD_true"),
+  qplot(scale_true$scale, scale_imp$scale / sqrt(INFO), color = INFO) +
+    geom_abline(color = "red") +
+    coord_equal() +
+    scale_color_viridis_c() +
+    labs(y = "SD_imp / sqrt(INFO)", x = "SD_true")
[…]

### 2021-09-07 code/investigate-PCwithLD-GWAS.R

@@ -1,0 +1,44 @@
+library(bigsnpr)
+ukb <- snp_attach("data/UKBB_HM3_val.rds")
+G <- ukb$genotypes
+PC19 <- as.matrix(bigreadr::fread2("UKBB/ukb41181.csv", select = "22009-0.19")[ukb$fam$id_csv, 1])
+
+ind_chr6 <- which(ukb$map$chromosome == 6)
+y <- snp_simuPheno(G, h2 = 0.2, M = 2000, ind.possible = ind_chr6)$pheno
+
+POS <- ukb$map$physical.pos[ind_chr6]
+is_in_bad_region <- (POS > 70e6) & (POS < 91e6)
+
+gwas  <- big_univLinReg(G, y, ind.col = ind_chr6, ncores = nb_cores())
+gwas2 <- big_univLinReg(G, y, ind.col = ind_chr6, ncores = nb_cores(), covar.train = PC19)
+
+sd_ss  <- with(gwas,  1 / sqrt(length(y) * std.err^2 + estim^2))
+sd_ss2 <- with(gwas2, 1 / sqrt(length(y) * std.err^2 + estim^2))
+sd_af <- with(ukb$map[ind_chr6, ], sqrt(2 * freq * (1 - freq)))
+INFO <- ukb$map$info[ind_chr6]
+
+library(ggplot2)
+plot_grid(
+  qplot(sd_af, sd_ss / sqrt(INFO), color = is_in_bad_region, alpha = I(0.5)) +
+    theme_bigstatsr(0.7) +
+    coord_equal() +
+    geom_abline(linetype = 2, color = "red") +
+    labs(x = "Standard deviations derived from the allele frequencies",
+         y = "Standard deviations derived from the summary statistics",
+         color = "Is in 70-91 Mbp\nof chromosome 6?") +
+    scale_color_manual(values = c("#E69F00", "#000000")) +
+    theme(legend.position = c(0.28, 0.8)),
+
+  qplot(sd_af, sd_ss2 / sqrt(INFO), color = is_in_bad_region, alpha = I(0.5)) +
+    theme_bigstatsr(0.7) +
+    coord_equal() +
+    geom_abline(linetype = 2, color = "red") +
+    labs(x = "Standard deviations derived from the allele frequencies",
+         y = "Standard deviations derived from the summary statistics",
+         color = "Is in 70-91 Mbp\nof chromosome 6?") +
[…]

### 2021-09-07 code/investigate-corr-imp.R

@@ -1,0 +1,86 @@
+library(bigsnpr)
+ukb <- snp_attach("data/UKBB_HM3_val.rds")
+G <- ukb$genotypes
+CHR <- ukb$map$chromosome
+POS2 <- ukb$map$genetic.dist
+INFO <- ukb$map$info
+
+(NCORES <- nb_cores())
+
+chr <- 22
+ind.chr <- which(CHR == chr)
+
+corr <- runonce::save_run(
+  snp_cor(G, ind.col = ind.chr, infos.pos = POS2[ind.chr],
+          size = 3 / 1000, ncores = NCORES),
+  file = paste0("data/corr/chr", chr, ".rds")
+)
+
+corr2 <- runonce::save_run(
+  snp_cor(G, ind.col = ind.chr, infos.pos = POS2[ind.chr],
+          size = 3 / 1000, info = INFO[ind.chr], ncores = NCORES),
+  file = paste0("data/corr/adj_chr", chr, ".rds")
+)
+
+ind_csv <- ukb$fam$id_csv
+all_id <- bigreadr::fread2("UKBB/ukb41181.csv", select = "eid")[[1]]
+sample <- bigreadr::fread2("UKBB/ukb58024_imp_chr1_v3_s487296.sample")[-1, ]
+ind.indiv <- match(all_id[ind_csv], sample$ID_2)
+
+list_snp_id <- with(ukb$map[ind.chr, ],
+                    split(paste(chromosome, physical.pos, allele1, allele2, sep = "_"),
+                          as.integer(chromosome)))
+
+for (ic in 1:10) {
+
+  print(ic)
+
+  rds <- snp_readBGEN(
+    bgenfiles   = glue::glue("UKBB/bgen/ukb_imp_chr{chr}_v3.bgen"),
+    list_snp_id = list_snp_id,
+    backingfile = tempfile(tmpdir = "tmp-data/sfbm/"),
+    ind_row     = ind.indiv,
+    read_as     = "random"
+  )
+
+  if (ic == 1) {
+    corr3 <- snp_cor(snp_attach(rds)$genotypes, infos.pos = POS2[ind.chr],
+                     size = 3 / 1000, ncores = NCORES)
+  } else {
+    corr3 <- corr3 + snp_cor(snp_attach(rds)$genotypes, infos.pos = POS2[ind.chr],
+                             size = 3 / 1000, ncores = NCORES)
+  }
+
+  file.remove(c(rds, sub("\\.rds$", ".bk", rds)))
+}
+
[…]

### 2021-09-07 code/investigate-misspec-N.R

@@ -1,0 +1,177 @@
+library(dplyr)
+library(ggplot2)
+library(bigsnpr)
+G <- snp_attach("data/ukbb4simu.rds")$genotypes
+
+load("data/ukbb4simu_ind.RData")
+
+# Correlation matrices made in 'code/prepare-corr-simu-chr22.R'
+corr0 <- readRDS("tmp-data/corr0_simu_val.rds")
+corr1 <- readRDS("tmp-data/corr_simu_val.rds")
+corr2 <- readRDS("tmp-data/corr_simu_val_with_blocks.rds")
+dim(corr0)  # 40000 x 40000
+
+NCORES <- nb_cores()
+
+
+#### Run GWAS and prepare sumstats ####
+
+set.seed(1)
+y <- snp_simuPheno(G, h2 = 0.2, M = 2000)$pheno
+
+list_ind <- list(ind.gwas,
+                 sort(sample(ind.gwas, length(ind.gwas) * 0.8)),
+                 sort(sample(ind.gwas, length(ind.gwas) * 0.6)))
+
+# GWAS to get sumstats
+gwas_set <- sample(rep_len(c(1, 1, 2, 3), ncol(G)))
+df_beta <- data.frame(beta = rep(NA, ncol(G)),
+                      beta_se = NA, n_eff = NA, lpval = NA)
+
+for (k in 1:3) {
+  ind_set <- which(gwas_set == k)
+  ind.gwas.sub <- list_ind[[k]]
+  gwas <- big_univLinReg(G, y[ind.gwas.sub], ind.train = ind.gwas.sub,
+                         ind.col = ind_set, ncores = NCORES)
+  df_beta$beta[ind_set]    <- gwas$estim
+  df_beta$beta_se[ind_set] <- gwas$std.err
+  df_beta$n_eff[ind_set]   <- length(ind.gwas.sub)
+  df_beta$lpval[ind_set]   <- -predict(gwas)
+}
+
+df_beta2 <- df_beta; df_beta2$n_eff <- max(df_beta$n_eff)
+
+# Quality control plot from the LDpred2 paper
+sd_val <- sqrt(big_colstats(G, ind.row = ind.val, ncores = NCORES)$var)
+sd_ss <- 1 / with(df_beta2, sqrt(n_eff * beta_se^2 + beta^2))
+sd_y <- sqrt(0.5) / max(sd_ss)
+sd_ss <- sd_y * sd_ss
+
[…]

### 2021-09-07 code/prepare-corr-large-clump.R

@@ -1,0 +1,50 @@
+library(bigsnpr)
+ukb <- snp_attach("data/UKBB_large_val.rds")
+G <- ukb$genotype
+CHR <- ukb$map$chromosome <- as.integer(ukb$map$chromosome)
+POS <- ukb$map$physical.pos
+
+map_hm3 <- snp_attach("data/UKBB_HM3_val.rds")$map
+in_hm3 <- vctrs::vec_in(ukb$map[1:6], map_hm3[1:6])
+sum(in_hm3)  # 1,041,651
+
+ind_keep <- runonce::save_run({
+
+  exclude <- with(ukb$map, info < 0.4 | freq < 0.01 | freq > 0.99 |
+                    paste(allele1, allele2) %in% c("A T", "T A", "C G", "G C"))
+
+  snp_clumping(G, thr.r2 = 0.9, S = ukb$map$mean_info + in_hm3,
+               infos.chr = CHR, infos.pos = POS, size = 100,
+               exclude = which(exclude), ncores = nb_cores())
+}, file = "data/ind_keep_large_clump.rds")
+# 9 min -- 2.5 M variants with thr_r2 = 0.9
+
+sum(in_hm3[ind_keep]) # 554,655
+table(CHR_keep <- CHR[ind_keep])
+#      1      2      3      4      5      6      7      8      9     10     11
+# 190939 204351 170390 164476 151197 150797 146197 132861 112483 123023 119075
+#     12     13     14     15     16     17     18     19     20     21     22
+# 115771  88334  78674  76638  86605  78339  74165  65514  60271  35898  39480
+
+POS2_keep <- snp_asGeneticPos(CHR_keep, POS[ind_keep], dir = "tmp-data",
+                              ncores = nb_cores())
+INFO_keep <- ukb$map$info[ind_keep]
+
+bigassertr::assert_dir("data/corr-large")
+
+library(future.batchtools)
+NCORES <- 15
+plan(batchtools_slurm(resources = list(
+  t = "12:00:00", c = NCORES + 1, mem = "100g",
+  name = basename(rstudioapi::getSourceEditorContext()$path))))
+
+furrr::future_walk(1:22, function(chr) {
+
+  ind.chr <- which(CHR_keep == chr)
+
+  runonce::save_run(
+    snp_cor(G, ind.col = ind_keep[ind.chr], infos.pos = POS2_keep[ind.chr],
[…]

### 2021-09-07 code/prepare-corr-maxtag-with-blocks.R

@@ -1,0 +1,60 @@
+library(future.batchtools)
+plan(batchtools_slurm(resources = list(
+  t = "12:00:00", c = 2, mem = "150g",
+  name = basename(rstudioapi::getSourceEditorContext()$path))))
+
+all_final_grp <- furrr::future_map_dfr(1:22, function(chr) {
+
+  corr0 <- readRDS(paste0("data/corr-large/maxtag_adj_chr", chr, ".rds"))
+  dim(corr0)
+
+  library(bigsnpr)
+  all_splits <- runonce::save_run({
+    SEQ <- round(seq_log(1000 + ncol(corr0) / 50,
+                         5000 + ncol(corr0) / 10,
+                         length.out = 10))
+
+    splits <- purrr::map_dfr(SEQ, function(max_size) {
+      res <- snp_ldsplit(corr0, thr_r2 = 0.02, min_size = 150,
+                         max_size = max_size, max_K = 250)
+      res$max_size <- max_size
+      res
+    })
+  }, file = paste0("tmp-data/split_corr_maxtag_chr", chr, ".rds"))
+
+
+  library(ggplot2)
+  qplot(data = all_splits, perc_kept, cost, color = as.factor(max_size)) +
+    theme_bw(12) +
+    theme(legend.position = "top") + guides(colour = guide_legend(nrow = 1)) +
+    labs(y = "Sum of squared correlations outside blocks (log-scale)",
+         x = "% of non-zero values kept", color = "Maximum block size") +
+    ylim(0, min(quantile(all_splits$cost, 0.6), 200)) +
+    geom_vline(xintercept = 0.6, linetype = 3)
+
+
+  library(dplyr)
+  final_split <- all_splits %>%
+    filter(perc_kept < `if`(chr == 6, 0.7, 0.6)) %>%
+    arrange(cost) %>%
+    slice(1)
+
+  final_grp <- final_split$block_num[[1]]
+
+  corr0T <- as(corr0, "dgTMatrix")
+  corr0T@x <- ifelse(final_grp[corr0T@i + 1L] == final_grp[corr0T@j + 1L], corr0T@x, 0)
+
+  corr0T %>%
+    Matrix::drop0() %>%
+    as("symmetricMatrix") %>%
+    saveRDS(paste0("data/corr-large/maxtag_adj_with_blocks_chr", chr, ".rds"))
+
+  final_split
+})
+
[…]

### 2021-09-07 code/prepare-corr-maxtag.R

@@ -1,0 +1,32 @@
+library(bigsnpr)
+ukb <- snp_attach("data/UKBB_maxtag_val.rds")
+G <- ukb$genotypes
+CHR <- as.integer(ukb$map$chromosome)
+table(CHR)
+#      1      2      3      4      5      6      7      8      9     10     11
+# 163457 168666 141715 133943 123370 128468 116418 110180  92967 103221 100611
+#     12     13     14     15     16     17     18     19     20     21     22
+#  95061  74998  64767  62391  66374  60870  62353  46982  51635  29977  30662
+
+POS2 <- snp_asGeneticPos(CHR, ukb$map$physical.pos, dir = "tmp-data",
+                         ncores = nb_cores())
+INFO <- ukb$map$info
+
+bigassertr::assert_dir("data/corr-large")
+
+library(future.batchtools)
+NCORES <- 15
+plan(batchtools_slurm(resources = list(
+  t = "12:00:00", c = NCORES + 1, mem = "100g",
+  name = basename(rstudioapi::getSourceEditorContext()$path))))
+
+furrr::future_walk(1:22, function(chr) {
+
+  ind.chr <- which(CHR == chr)
+
+  runonce::save_run(
+    snp_cor(G, ind.col = ind.chr, infos.pos = POS2[ind.chr],
+            size = 3 / 1000, info = INFO[ind.chr], ncores = NCORES),
+    file = paste0("data/corr-large/maxtag_adj_chr", chr, ".rds")
+  )
+})

### 2021-09-07 code/prepare-genotypes-simu-chr22.R

@@ -1,0 +1,80 @@
+# Get variant info
+chr <- 22
+info_chr <- bigreadr::fread2(paste0("UKBB/mfi/ukb_mfi_chr", chr, "_v3.txt"))
+info_chr_sub <- subset(info_chr, V6 > 0.01 & V8 > 0.4, -c(V1, V2, V7))
+map <- cbind.data.frame(chr = chr, setNames(info_chr_sub, c("pos", "a1", "a2", "maf", "info")))
+hist(map$info)
+M <- 40e3
+
+{
+  all_info <- map$info
+  kern_smooth <- ks::kde(all_info)
+  set.seed(1)
+  ind_snp <- sort(sample(length(all_info), M, replace = FALSE,
+                         prob = 1 / predict(kern_smooth, x = all_info)))
+  # pdf("figures/simu_hist_info.pdf", width = 7, height = 5)
+  hist(all_info[ind_snp], main = NULL, xlab = "INFO")
+  # dev.off()
+  list_snp_id <- list(with(map[ind_snp, ], paste(chr, pos, a1, a2, sep = "_")))
+}
+
+
+df0 <- bigreadr::fread2(
+  "UKBB/ukb41181.csv",
+  select = c("eid", "22020-0.0", paste0("22009-0.", 1:16)),
+  col.names = c("eid", "used_in_pca", paste0("PC", 1:16))
+)
+
+# sample still in data
+sample <- bigreadr::fread2("UKBB/ukb58024_imp_chr1_v3_s487296.sample")[-1, ]
+ind.indiv <- match(df0$eid, sample$ID_2)
+id_to_rm <- scan("../../UKBiobank/ukb_withdrawn_individuals_8Feb2021.id")
+ind.indiv[df0$eid %in% id_to_rm] <- NA
+sub <- which(!is.na(ind.indiv) & df0$used_in_pca)
+
+# Genetically homogeneous
+dist <- bigutilsr::dist_ogk(as.matrix(df0[sub, -(1:2)]))
+hist(log(dist), "FD")
+sub2 <- sub[log(dist) < 5]
+length(sub2) # 362,307
+
+
+library(bigsnpr)
+NCORES <- nb_cores()
+
+# Read all data as random hard calls
+bgen <- "UKBB/bgen/ukb_imp_chr22_v3.bgen"
+system.time(
+  snp_readBGEN(
+    bgenfiles = bgen,
+    list_snp_id = list_snp_id,
[…]

### 2021-09-07 code/prepare-genotypes.R

@@ -1,0 +1,82 @@
+# file.symlink("../../UKBB/", "UKBB")
+
+map_hapmap3 <- readRDS(url("https://ndownloader.figshare.com/files/25503788"))
+
+library(doParallel)
+cl <- makeCluster(22)
+parallel::clusterExport(cl, "map_hapmap3")
+list_snp_id <- parLapply(cl, 1:22, function(chr) {
+  mfi <- paste0("UKBB/mfi/ukb_mfi_chr", chr, "_v3.txt")
+  infos_chr <- bigreadr::fread2(mfi, showProgress = FALSE)
+  joined <- dplyr::inner_join(cbind(chr = chr, infos_chr), map_hapmap3[1:2],
+                              by = c("chr", "V3" = "pos"))
+  with(joined[!vctrs::vec_duplicate_detect(joined$V3), ],
+       paste(chr, V3, V4, V5, sep = "_"))
+})
+stopCluster(cl)
+
+sum(lengths(list_snp_id))  # 1,054,315
+
+
+df0 <- bigreadr::fread2(
+  "UKBB/ukb41181.csv",
+  select = c("eid", "22020-0.0", paste0("22009-0.", 1:16)),
+  col.names = c("eid", "used_in_pca", paste0("PC", 1:16))
+)
+
+# sample still in data
+sample <- bigreadr::fread2("UKBB/ukb58024_imp_chr1_v3_s487296.sample")[-1, ]
+ind.indiv <- match(df0$eid, sample$ID_2)
+id_to_rm <- scan("../../UKBiobank/ukb_withdrawn_individuals_8Feb2021.id")
+ind.indiv[df0$eid %in% id_to_rm] <- NA
+sub <- which(!is.na(ind.indiv) & df0$used_in_pca)
+
+# Genetically homogeneous
+dist <- bigutilsr::dist_ogk(as.matrix(df0[sub, -(1:2)]))
+hist(log(dist), "FD")
+sub2 <- sub[log(dist) < 5]
+length(sub2) # 362,307
+
+
+# Validation set
+set.seed(1); ind.val <- sort(sample(sub2, 10e3))
+
+library(bigsnpr)
+snp_readBGEN(
+  bgenfiles   = glue::glue("UKBB/bgen/ukb_imp_chr{chr}_v3.bgen", chr = 1:22),
+  list_snp_id = list_snp_id,
+  backingfile = "data/UKBB_HM3_val",
+  ind_row     = ind.indiv[ind.val],
[…]

### 2021-09-07 code/prepare-large-corr.R

@@ -1,0 +1,32 @@
+library(bigsnpr)
+ukb <- snp_attach("data/UKBB_large_val.rds")
+G <- ukb$genotypes
+CHR <- as.integer(ukb$map$chromosome)
+POS <- ukb$map$physical.pos
+INFO <- ukb$map$info
+
+
+#### Compute LD matrices ####
+
+bigassertr::assert_dir("data/corr-large")
+
+library(future.batchtools)
+NCORES <- 30
+plan(batchtools_slurm(resources = list(
+  t = "12:00:00", c = NCORES + 2, mem = "200g",
+  name = basename(rstudioapi::getSourceEditorContext()$path))))
+
+future.apply::future_lapply(1:22, function(chr) {
+
+  ind.chr <- which(CHR == chr)
+  POS2 <- snp_asGeneticPos(CHR[ind.chr], POS[ind.chr], dir = "tmp-data")
+
+  corr <- snp_cor(G, ind.col = ind.chr, infos.pos = POS2, size = 3 / 1000,
+                  info = INFO[ind.chr], thr_r2 = 0.01, ncores = NCORES)
+
+  saveRDS(corr, file = paste0("data/corr-large/chr", chr, ".rds"), version = 2)
+})
+# 10 min for chr 22
+
+sum(sapply(list.files("data/corr-large", full.names = TRUE), file.size)) / 1024^3
+# 41.7 GB

### 2021-09-07 code/prepare-larger-set.R

@@ -1,0 +1,163 @@
+library(bigreadr)
+library(dplyr)
+library(data.table)
+
+# UKBB
+map_ukbb <- do.call("rbind", lapply(1:22, function(chr) {
+  print(chr)
+  mfi <- fread2(paste0("UKBB/mfi/ukb_mfi_chr", chr, "_v3.txt"),
+                select = c(3:6, 8),
+                col.names = c("pos", "a1", "a0", "maf", "info_ukbb"))
+  cbind(chr = chr, subset(mfi, maf > 0.01, info_ukbb > 0.3))
+}))
+str(map_ukbb)
+
+# iPSYCH (2015 + 2012)
+map_ipsych <- bigsnpr::snp_attach("../data_ipsych/dosage_ipsych2015.rds")$map %>%
+  transmute(chr = CHR, pos = POS, info_2012, freq_2012, info_2015, freq_2015)
+
+merged <- merge(as.data.table(map_ukbb), map_ipsych, all.x = TRUE)
+setkey(merged, chr, pos)
+
+wd <- setwd("tmp-data")
+
+#### BRCA ####
+# download.file("ftp://ftp.ebi.ac.uk/pub/databases/gwas/summary_statistics/MichailidouK_29059683_GCST004988/oncoarray_bcac_public_release_oct17%20(1).txt.gz",
+#               destfile = "tmp-data/sumstats_BRCA.txt.gz")
+# R.utils::gunzip("tmp-data/sumstats_BRCA.txt.gz")
+sumstats_brca <- fread2(
+  "sumstats_BRCA.txt", na.strings = "NULL",
+  select = c("chr", "position_b37", "bcac_onco2_r2", "bcac_icogs2_r2", "bcac_gwas_all_eaf_controls"),
+  col.names = c("chr", "pos", "info_onco_brca", "info_icogs_brca", "freq_brca"))
+str(sumstats_brca)
+
+merged <- merge(merged, sumstats_brca, all.x = TRUE)
+
+#### PRCA ####
+# download.file("http://practical.icr.ac.uk/blog/wp-content/uploads/uploadedfiles/oncoarray/MetaSummaryData/meta_v3_onco_euro_overall_ChrAll_1_release.zip",
+#               destfile = "sumstats_PRCA.zip")
+# unzip("sumstats_PRCA.zip"); file.remove("sumstats_PRCA.zip")
+sumstats_prca <- fread2(
+  "meta_v3_onco_euro_overall_ChrAll_1_release.txt",
[…]

### 2021-09-07 code/prepare-sumstats-other-sets.R

@@ -1,0 +1,57 @@
+# dput(list.files("code/prepare-sumstats/"))
+files <- c(
+  "BrCa-icogs.R",
+  "BrCa-onco.R",
+  "CAD.R",
+  "MDD.R",
+  "PrCa.R",
+  "T1D-affy.R",
+  "T1D-illu.R"
+)
+
+
+bigassertr::assert_dir("data/sumstats-maxtag")
+
+for (file in files) {
+
+  rm(list = setdiff(ls(), c("file", "files")))
+
+  code <- readLines(file.path("code/prepare-sumstats", file))
+
+  library(magrittr)
+  new_code <- code %>%
+    sub('"data/UKBB_HM3_val.rds"', '"data/UKBB_maxtag_val.rds"', ., fixed = TRUE) %>%
+    sub('"data/sumstats/', '"data/sumstats-maxtag/', ., fixed = TRUE) %>%
+    sub('# saveRDS(filter(info_snp2', 'saveRDS(filter(info_snp2', ., fixed = TRUE) %>%
+    print()
+
+  stopifnot(sum(new_code != code) == 3)
+
+  eval(parse(text = new_code))
+}
+
+
+bigassertr::assert_dir("data/sumstats-clump")
+
+map <- readRDS("data/UKBB_large_val.rds")$map
+ind_clump <- readRDS("data/ind_keep_large_clump.rds")
+saveRDS(list(map = map[ind_clump, ]), "tmp-data/fake_for_clump.rds")
+
+for (file in files) {
+
+  rm(list = setdiff(ls(), c("file", "files")))
+
+  code <- readLines(file.path("code/prepare-sumstats", file))
+
+  library(magrittr)
+  new_code <- code %>%
+    sub('snp_attach(', 'readRDS(', ., fixed = TRUE) %>%
+    sub('"data/UKBB_HM3_val.rds"', '"tmp-data/fake_for_clump.rds"', ., fixed = TRUE) %>%
+    sub('"data/sumstats/', '"data/sumstats-clump/', ., fixed = TRUE) %>%
+    sub('# saveRDS(filter(info_snp2', 'saveRDS(filter(info_snp2', ., fixed = TRUE) %>%
+    print()
+
+  stopifnot(sum(new_code != code) == 3)
+
+  eval(parse(text = new_code))
+}

### 2021-09-07 code/prepare-sumstats/BrCa-icogs.R

@@ -1,0 +1,119 @@
+library(bigsnpr)
+library(bigreadr)
+library(dplyr)
+library(ggplot2)
+
+map <- transmute(snp_attach("data/UKBB_HM3_val.rds")$map,
+                 chr = as.integer(chromosome), pos = physical.pos, af_val = freq,
+                 a0 = allele1, a1 = allele2)  # reversed somehow..
+
+
+#### Breast cancer (BRCA) ####
+
+# download.file("ftp://ftp.ebi.ac.uk/pub/databases/gwas/summary_statistics/MichailidouK_29059683_GCST004988/oncoarray_bcac_public_release_oct17%20(1).txt.gz",
+#               destfile = "tmp-data/sumstats_BRCA.txt.gz")
+# R.utils::gunzip("tmp-data/sumstats_BRCA.txt.gz")
+
+# iCOGS
+sumstats <- fread2("tmp-data/sumstats_BRCA.txt", na.strings = "NULL",
+                   select = c("chr", "position_b37", "a0", "a1",
+                              "bcac_icogs2_eaf_controls",
+                              "bcac_icogs2_beta",
+                              "bcac_icogs2_se",
+                              "bcac_icogs2_r2"),
+                   col.names = c("chr", "pos", "a0", "a1", "freq",
+                                 "beta", "beta_se", "info")) %>%
+  filter(info > 0.4, beta_se > 0)
+
+Neff1 <- 4 / (1 / 46785  + 1 / 42892)
+Neff2 <- 4 / (1 / 45290  + 1 / 41880)
+Neff1 / Neff2  # 1.03
+sumstats$n_eff <- Neff2
+
+info_snp <- as_tibble(bigsnpr::snp_match(sumstats, map))
+# 11,417,746 variants to be matched.
+# 0 ambiguous SNPs have been removed.
+# 1,051,242 variants have been matched; 0 were flipped and 0 were reversed.
+
+# pdf("figures/brca_icogs_hist_info.pdf", width = 7, height = 6)
+# par(mar = c(4.1, 4.1, 0.1, 0.1))
+# hist(info_snp$info, main = NULL, xlab = "INFO")
+# dev.off()
+mean(info_snp$info)
+
+info_snp2 <- mutate(info_snp,
+                    sd_af = sqrt(2 * freq * (1 - freq)),
+                    sd_ss = 2 / sqrt(n_eff * beta_se^2 + beta^2),
+                    sd_ss2 = sd_ss / sqrt(info))
+
[…]

### 2021-09-07 code/prepare-sumstats/CAD.R

@@ -1,0 +1,149 @@
+library(bigsnpr)
+library(bigreadr)
+library(dplyr)
+library(ggplot2)
+
+map <- transmute(snp_attach("data/UKBB_HM3_val.rds")$map,
+                 chr = as.integer(chromosome), pos = physical.pos, af_val = freq,
+                 a0 = allele1, a1 = allele2)  # reversed somehow..
+
+
+#### Coronary artery disease (CAD) ####
+
+# download.file("ftp://ftp.ebi.ac.uk/pub/databases/gwas/summary_statistics/NikpayM_26343387_GCST003116/cad.add.160614.website.txt",
+#               destfile = "tmp-data/sumstats_CAD.txt")
+# bigreadr::nlines("tmp-data/sumstats_CAD.txt")  # 9,455,779
+sumstats <- fread2("tmp-data/sumstats_CAD.txt",
+                   select = c("chr", "bp_hg19", "noneffect_allele",
+                              "effect_allele", "beta", "se_dgc", "p_dgc",
+                              "effect_allele_freq", "median_info", "n_studies", "het_pvalue"),
+                   col.names = c("chr", "pos", "a0", "a1", "beta", "beta_se",
+                                 "p", "freq", "info", "n_studies", "het_pvalue"))
+
+study_size <- c(
+  # from Supp Table 1
+  "5719 6545","206 259","278 312","505 1021","392 410","1010 3998",
+  "1628 368","2083 2048","1216 653","658 5841","1802 466","2099 2690",
+  "634 1608","1207 1288","1061 1467","1089 1147","877 2187",
+  "2700 2758","361 2778","487 1381","758 3337","2791 3757","2095 503",
+  "933 468","2905 2998","947 1008","1294 1529","843 318","933 468",
+  "119 830","631 334","836 761","426 594","814 5999","322 857",
+  "1926 2938","4651 4452","4380 3929","1535 772","1007 22286","402 448",
+  "745 1389","397 2474","506 5335","259 4202","334 3446","2034 3210","454 8443")
+study_size2 <- fread2(text = study_size, col.names = c("Nca", "Nco"))
+nrow(study_size2)   # 48
+colSums(study_size2)   # Nca: 61289 -- Nco: 126310  (slightly different from 60801 -- 123504)
+(Neff <- sum(4 / rowSums(1 / study_size2)))  # 129014.3
[…]

### 2021-09-07 code/prepare-sumstats/MDD.R

@@ -1,0 +1,130 @@
+library(bigsnpr)
+library(bigreadr)
+library(dplyr)
+library(ggplot2)
+
+map <- transmute(snp_attach("data/UKBB_HM3_val.rds")$map,
+                 chr = as.integer(chromosome), pos = physical.pos, af_val = freq,
+                 a0 = allele1, a1 = allele2)  # reversed somehow..
+
+
+#### MDD ####
+
+gz <- runonce::download_file(
+  "https://pgcdata.med.unc.edu/major_depressive_disorders/daner_pgc_mdd_meta_w2_no23andMe_rmUKBB.gz",
+  dir = "tmp-data")
+sumstats <- fread2(gz,
+                   select = c("CHR", "BP", "A1", "A2", "OR", "SE", "P",
+                              "Nca", "Nco", "INFO", "FRQ_A_45396", "FRQ_U_97250", "Neff_half"),
+                   col.names = c("chr", "pos", "a1", "a0", "or", "beta_se", "p",
+                                 "Nca", "Nco", "info", "freq1", "freq2", "Neff_half")) %>%
+  as_tibble() %>%
+  mutate(beta = log(or), or = NULL, chr = as.integer(chr),
+         n_eff = 2 * Neff_half, Neff_half = NULL,
+         n_eff2 = 4 / (1 / Nca + 1 / Nco),
+         Nca = NULL, Nco = NULL,
+         info = pmin(info, 1)) %>%
+  filter(n_eff > (0.5 * max(n_eff)),
+         info > 0.4) %>%
+  print()
+
+info_snp <- as_tibble(bigsnpr::snp_match(sumstats, map))
+# 9,343,125 variants to be matched.
+# 0 ambiguous SNPs have been removed.
+# 1,049,455 variants have been matched; 0 were flipped and 509,899 were reversed.
+
+hist(info_snp$n_eff, "FD"); abline(v = 4 / (1 / 45396 + 1 / 97250), col = "red")
+hist(info_snp$n_eff2, "FD"); abline(v = 4 / (1 / 45396 + 1 / 97250), col = "red")
+max(info_snp$n_eff)          # 110463.7
+max(info_snp$n_eff2)         # 127996.2
+4 / (1 / 45396 + 1 / 97250)  # 123796.3
+
+mean(info_snp$info)
+hist(info_snp$info, "FD")
+
+info_snp2 <- info_snp %>%
+  mutate(freq = (freq1 * 45396 + freq2 * 97250) / (45396 + 97250),
[…]

### 2021-09-07 code/prepare-sumstats/PrCa.R

@@ -1,0 +1,121 @@
+library(bigsnpr)
+library(bigreadr)
+library(dplyr)
+library(ggplot2)
+
+map <- transmute(snp_attach("data/UKBB_HM3_val.rds")$map,
+                 chr = as.integer(chromosome), pos = physical.pos, af_val = freq,
+                 a0 = allele1, a1 = allele2)  # reversed somehow..
+
+
+#### Prostate cancer (PRCA) ####
+
+# download.file("http://practical.icr.ac.uk/blog/wp-content/uploads/uploadedfiles/oncoarray/MetaSummaryData/meta_v3_onco_euro_overall_ChrAll_1_release.zip",
+#               destfile = "tmp-data/sumstats_PRCA.zip")
+# unzip("tmp-data/sumstats_PRCA.zip", exdir = "tmp-data")
+sumstats <- fread2(
+  "tmp-data/meta_v3_onco_euro_overall_ChrAll_1_release.txt",
+  select = c("Chr", "position", "Allele1", "Allele2", "Effect", "StdErr",
+             "Pvalue", "Freq1", "OncoArray_imputation_r2"),
+  col.names = c("chr", "pos", "a1", "a0", "beta", "beta_se", "p", "freq", "info")
+) %>%
+  mutate(a0 = toupper(a0), a1 = toupper(a1)) %>%
+  filter(info > 0.4, beta_se > 0)
+
+info_snp <- as_tibble(bigsnpr::snp_match(sumstats, map))
+# 13,421,441 variants to be matched.
+# 0 ambiguous SNPs have been removed.
+# 818,400 variants have been matched; 0 were flipped and 398,171 were reversed.
+
+sizes <- bigreadr::fread2(text = c("44825\t27904","20219\t20440","1854\t1894","3650\t3940",
+                                   "474\t482","1458\t512","2068\t2993", "4600\t2941"))
+info_snp$n_eff <- print(sum(4 / rowSums(1 / sizes)))  # 135316.1
+4 / (1 / 79148 + 1 / 61106)  # 137933.1
+
+mean(info_snp$info)
+hist(info_snp$info, "FD")
+
+info_snp2 <- mutate(info_snp,
+                    sd_af = sqrt(2 * freq * (1 - freq)),
+                    sd_ss = 2 / sqrt(n_eff * beta_se^2 + beta^2),
+                    sd_ss2 = sd_ss / sqrt(info))
+
[…]

### 2021-09-07 code/prepare-sumstats/T1D-illu.R

@@ -1,0 +1,126 @@
+library(bigsnpr)
+library(bigreadr)
+library(dplyr)
+library(ggplot2)
+
+map <- transmute(snp_attach("data/UKBB_HM3_val.rds")$map,
+                 chr = as.integer(chromosome), pos = physical.pos, af_val = freq,
+                 a0 = allele1, a1 = allele2)  # reversed somehow..
+
+
+#### Type 1 diabetes (T1D) ####
+
+# To request a download link, go to
+# https://datadryad.org/stash/dataset/doi:10.5061/dryad.ns8q3
+# download.file("http://merritt.cdlib.org/cloudcontainer/mrtstore2/35227889.tar.gz",
+#               destfile = "tmp-data/sumstats_T1D.tar.gz")
+# untar("tmp-data/sumstats_T1D.tar.gz", exdir = "T1D")
+
+# Illumina
+sumstats <- fread2(
+  paste0("tmp-data/meta_chr_", 1:22),
+  select = c("chromosome", "position", "a0", "a1", "info_score.I",
+             "EUR_MAF_1kG", "controls_maf.I", "ill.OR", "ill.se", "qc.check"),
+  col.names = c("chr", "pos", "a0", "a1", "info",
+                "freq_1000G", "freq", "OR", "beta_se", "qc")) %>%
+  mutate(beta = log(OR), OR = NULL) %>%
+  filter(info > 0.4, beta_se > 0)
+# https://journals.plos.org/plosmedicine/article?id=10.1371/journal.pmed.1002362#pmed-1002362-t001
+sumstats$n_eff <- 4 / (1 / 3983 + 1 / 3999)
+
+(info_snp <- as_tibble(bigsnpr::snp_match(sumstats, map)))
+# 8,865,956 variants to be matched.
+# 0 ambiguous SNPs have been removed.
+# 949,334 variants have been matched; 527 were flipped and 1,789 were reversed.
+
+table(info_snp$qc)
+mean(info_snp$info)
+hist(info_snp$info, "FD")
+summary(info_snp$info)
+
+info_snp2 <- mutate(info_snp,
+                    sd_af = sqrt(2 * freq * (1 - freq)),
+                    sd_val = sqrt(2 * af_val * (1 - af_val)),
+                    sd_ss = 2 / sqrt(n_eff * beta_se^2 + beta^2),
+                    sd_ss2 = sd_ss / sqrt(info))
+
+# original filtering applied in LDpred2 paper (but using af from sumstats)
[…]

### 2021-09-07 code/process-PRS-maxtag.R

@@ -1,0 +1,178 @@
+library(bigsnpr)
+ukb <- snp_attach("data/UKBB_maxtag_val.rds")
+G <- ukb$genotypes
+
+library(dplyr)
+df0 <- bigreadr::fread2(
+  "UKBB/ukb41181.csv",
+  select = c("eid", "34-0.0", "52-0.0", "22001-0.0", "21022-0.0", "189-0.0",
+             paste0("22009-0.", 1:16)),
+  col.names = c("eid", "year", "month", "sex", "age", "deprivation_index",
+                paste0("PC", 1:16))
+) %>%
+  mutate(date = (year - 1900) + (month - 0.5) / 12, year = NULL, month = NULL)
+
+ind.val <- ukb$fam$id_csv
+covar <- as.matrix(df0[ind.val, -1])
+
+ALL_METHODS <- c("lassosum2", "LDpred2", "LDpred2-low-h2",
+                 "LDpred2-auto", "LDpred2-auto-rob")
+
+bigassertr::assert_dir("results-maxtag-final")
+
+library(dplyr)
+grid <- tibble::tribble(
+  ~pheno,                               ~phecode,
+  paste0("t1d_", c("affy", "illu")),    "250.1",
+  paste0("brca_", c("onco", "icogs")),  "174.1",
+  "cad",                                "411.4",
+  "prca",                               "185",
+  "mdd",                                "296.22",
+) %>%
+  tidyr::expand_grid(
+    qc = c("qc1", "qc2"),
+    use_info = c(FALSE, TRUE),
+    name_corr = c("data/corr-large/maxtag_adj_chr",
+                  "data/corr-large/maxtag_adj_with_blocks_chr"),
+    method = ALL_METHODS) %>%
+  tidyr::unnest("pheno") %>%
+  print()
+
+
+library(future.apply)
+plan("multisession", workers = 10)
+
+grid$effects <- furrr::future_pmap(grid[1:6], function(
+  pheno, phecode, qc, use_info, name_corr, method) {
+
+  # pheno <- "t1d_illu"
+  # phecode <- "250.1"
+  # qc <- "qc1"
+  # use_info <- TRUE
+  # name_corr <- "data/corr-large/maxtag_adj_with_blocks_chr"
+  # method <- "lassosum2"
+
+  basename <- paste0(pheno, "_", qc)
+  sumstats_file <- paste0("data/sumstats-maxtag/", basename, ".rds")
[…]

### 2021-09-07 code/process-PRS.R

@@ -1,0 +1,168 @@
+library(bigsnpr)
+ukb <- snp_attach("data/UKBB_HM3_val.rds")
+G <- ukb$genotypes
+NCORES <- nb_cores()
+
+library(dplyr)
+df0 <- bigreadr::fread2(
+  "UKBB/ukb41181.csv",
+  select = c("eid", "34-0.0", "52-0.0", "22001-0.0", "21022-0.0", "189-0.0",
+             paste0("22009-0.", 1:16)),
+  col.names = c("eid", "year", "month", "sex", "age", "deprivation_index",
+                paste0("PC", 1:16))
+) %>%
+  mutate(date = (year - 1900) + (month - 0.5) / 12, year = NULL, month = NULL)
+
+ind.val <- ukb$fam$id_csv
+covar <- as.matrix(df0[ind.val, -1])
+
+ALL_METHODS <- c("lassosum2", "LDpred2", "LDpred2-low-h2",
+                 "LDpred2-auto", "LDpred2-auto-rob")
+
+bigassertr::assert_dir("results-final")
+
+library(dplyr)
+grid <- tibble::tribble(
+  ~pheno,                               ~phecode,
+  paste0("t1d_", c("affy", "illu")),    "250.1",
+  paste0("brca_", c("onco", "icogs")),  "174.1",
+  "cad",                                "411.4",
+  "prca",                               "185",
+  "mdd",                                "296.22",
+) %>%
+  tidyr::expand_grid(
+    qc = c("qc1", "qc2"),
+    use_info = c(FALSE, TRUE),
+    name_corr = c("data/corr/chr", "data/corr/adj_with_blocks_chr"),
+    method = ALL_METHODS) %>%
+  tidyr::unnest("pheno") %>%
+  print()
+
+
+library(future.apply)
+plan("multisession", workers = NCORES)
+
+grid$effects <- furrr::future_pmap(grid[1:6], function(
+  pheno, phecode, qc, use_info, name_corr, method) {
+
+  # pheno <- "t1d_illu"
+  # phecode <- "250.1"
+  # qc <- "qc1"
+  # use_info <- TRUE
+  # name_corr <- "data/corr/adj_with_blocks_chr"
+  # method <- "lassosum2"
+
+  basename <- paste0(pheno, "_", qc)
+  sumstats_file <- paste0("data/sumstats/", basename, ".rds")
+  if (grepl("blocks", name_corr, fixed = TRUE))
+    basename <- paste0(basename, "_adj_with_blocks")
[…]

### 2021-09-07 code/run-PRS-clump.R

@@ -1,0 +1,107 @@
+
+bigassertr::assert_dir("results-clump")
+bigassertr::assert_dir("log")
+
+library(dplyr)
+files <- tibble(pheno = c(paste0("t1d_", c("affy", "illu")),
+                          paste0("brca_", c("onco", "icogs")),
+                          "cad", "prca", "mdd")) %>%
+  tidyr::expand_grid(
+    qc = c("qc1", "qc2"),
+    name_corr = c("data/corr-large/clump_adj_chr",
+                  "data/corr-large/clump_adj_with_blocks_chr")) %>%
+  print()
+
+library(future.batchtools)
+NCORES <- 30
+plan(batchtools_slurm(resources = list(
+  t = "24:00:00", c = NCORES + 2, mem = "150g",
+  name = basename(rstudioapi::getSourceEditorContext()$path))))
+
+furrr::future_pwalk(files, function(pheno, qc, name_corr) {
+
+  # pheno <- "t1d_illu"
+  # qc <- "qc1"
+  # name_corr <- "data/corr-large/clump_adj_with_blocks_chr"
+
+  gwas_file <- paste0("data/sumstats-clump/", pheno, "_", qc, ".rds")
+  sumstats <- readRDS(gwas_file)
+  basename <- file.path("results-clump", sub("\\.rds$", "", basename(gwas_file)))
+  if (grepl("blocks", name_corr, fixed = TRUE))
+    basename <- paste0(basename, "_adj_with_blocks")
+
+  library(bigsnpr)
+  CHR <- as.integer(readRDS("tmp-data/fake_for_clump.rds")$map$chromosome)
+
+  tmp <- tempfile(tmpdir = "tmp-data/sfbm/")
+  on.exit(file.remove(paste0(tmp, ".sbk")), add = TRUE)
+
+  for (chr in 1:22) {
+
+    print(chr)
+
+    ## indices in 'sumstats'
+    ind.chr <- which(sumstats$chr == chr)
+    ## indices in 'G'
+    ind.chr2 <- sumstats$`_NUM_ID_`[ind.chr]
+    ## indices in 'corr'
+    ind.chr3 <- match(ind.chr2, which(CHR == chr))
+
+    corr0 <- readRDS(paste0(name_corr, chr, ".rds"))
+    ld_chr <- Matrix::colSums(corr0^2)
[…]

### 2021-09-07 code/run-PRS-maxtag.R

@@ -1,0 +1,107 @@
+
+bigassertr::assert_dir("results-maxtag")
+bigassertr::assert_dir("log")
+
+library(dplyr)
+files <- tibble(pheno = c(paste0("t1d_", c("affy", "illu")),
+                          paste0("brca_", c("onco", "icogs")),
+                          "cad", "prca", "mdd")) %>%
+  tidyr::expand_grid(
+    qc = c("qc1", "qc2"),
+    name_corr = c("data/corr-large/maxtag_adj_chr",
+                  "data/corr-large/maxtag_adj_with_blocks_chr")) %>%
+  print()
+
+library(future.batchtools)
+NCORES <- 30
+plan(batchtools_slurm(resources = list(
+  t = "12:00:00", c = NCORES + 2, mem = "150g",
+  name = basename(rstudioapi::getSourceEditorContext()$path))))
+
+furrr::future_pwalk(files, function(pheno, qc, name_corr) {
+
+  # pheno <- "t1d_illu"
+  # qc <- "qc1"
+  # name_corr <- "data/corr-large/maxtag_adj_with_blocks_chr"
+
+  gwas_file <- paste0("data/sumstats-maxtag/", pheno, "_", qc, ".rds")
+  sumstats <- readRDS(gwas_file)
+  basename <- file.path("results-maxtag", sub("\\.rds$", "", basename(gwas_file)))
+  if (grepl("blocks", name_corr, fixed = TRUE))
+    basename <- paste0(basename, "_adj_with_blocks")
+
+  library(bigsnpr)
+  CHR <- as.integer(snp_attach("data/UKBB_maxtag_val.rds")$map$chromosome)
+
+  tmp <- tempfile(tmpdir = "tmp-data/sfbm/")
+  on.exit(file.remove(paste0(tmp, ".sbk")), add = TRUE)
+
+  for (chr in 1:22) {
+
+    print(chr)
+
+    ## indices in 'sumstats'
+    ind.chr <- which(sumstats$chr == chr)
+    ## indices in 'G'
+    ind.chr2 <- sumstats$`_NUM_ID_`[ind.chr]
+    ## indices in 'corr'
+    ind.chr3 <- match(ind.chr2, which(CHR == chr))
+
+    corr0 <- readRDS(paste0(name_corr, chr, ".rds"))
+    ld_chr <- Matrix::colSums(corr0^2)
[…]

### 2021-09-07 code/run-methods.R

@@ -1,0 +1,250 @@
+
+#### Run LDpred2 ####
+
+run_ldpred2 <- function(df_beta, corr) {
+
+  # LDSc reg
+  corr0 <- readRDS("tmp-data/corr0_simu_val.rds")
+  print(ldsc <- snp_ldsc2(corr0, df_beta))
+  h2_est <- ldsc[["h2"]]
+
+  # LDpred-inf
+  beta_inf <- snp_ldpred2_inf(corr, df_beta, h2 = h2_est)
+  pred_inf <- big_prodVec(G, beta_inf, ind.row = ind.test)
+  print(r21 <- cor(pred_inf, y[ind.test])**2)
+
+  ## LDpred2-grid
+  (h2_seq <- round(h2_est * c(0.7, 1, 1.4), 4))
+  (p_seq <- signif(seq_log(1e-4, 1, length.out = 15), 2))
+  (params <- expand.grid(p = p_seq, h2 = h2_seq, sparse = FALSE))
+
+  beta_grid <- snp_ldpred2_grid(corr, df_beta, params, ncores = NCORES)
+  pred_grid <- big_prodMat(G, beta_grid, ncores = NCORES)
+  score.val <- apply(pred_grid[ind.val, ], 2, cor, y = y[ind.val])
+  pred_gibbs <- pred_grid[ind.test, which.max(score.val)]
+  print(r22 <- cor(pred_gibbs, y[ind.test])**2)
+
+  # LDpred2-grid (including small values for h2)
+  (h2_seq2 <- round(h2_est * c(0.01, 0.1, 0.3, 0.7, 1, 1.4), 4))
+  (params2 <- expand.grid(p = p_seq, h2 = h2_seq2, sparse = FALSE))
+
+  beta_grid2 <- snp_ldpred2_grid(corr, df_beta, params2, ncores = NCORES)
+  pred_grid2 <- big_prodMat(G, beta_grid2, ncores = NCORES)
+  score.val2 <- apply(pred_grid2[ind.val, ], 2, cor, y = y[ind.val])
+  pred_gibbs2 <- pred_grid2[ind.test, which.max(score.val2)]
+  print(r23 <- cor(pred_gibbs2, y[ind.test])**2)
+
+  # LDpred-auto
+  auto <- snp_ldpred2_auto(corr, df_beta, h2_init = h2_est,
+                           shrink_corr = 1, allow_jump_sign = TRUE)
+  print(c(auto[[1]]$h2_est, auto[[1]]$p_est))
[…]

### 2021-09-07 code/test-lassosum2-auto.R

@@ -1,0 +1,77 @@
+library(bigsnpr)
+ukb <- snp_attach("data/ukbb4simu.rds")
+G <- ukb$genotypes
+
+load("data/ukbb4simu_ind.RData")
+
+(NCORES <- nb_cores())
+
+corr0 <- runonce::save_run({
+  POS2 <- snp_asGeneticPos(as.integer(ukb$map$chromosome),
+                           ukb$map$physical.pos, dir = "tmp-data")
+  snp_cor(G, ind.row = ind.val, infos.pos = POS2, size = 3 / 1000, ncores = nb_cores())
+}, file = "tmp-data/corr0_simu_val.rds")
+corr <- runonce::save_run(as_SFBM(corr0, "tmp-data/corr_simu_val"),
+                          file = "tmp-data/corr_simu_val.rds")
+
+y <- snp_simuPheno(G, h2 = 0.2, M = 2000, ncores = NCORES)$pheno
+
+# GWAS to get sumstats
+gwas <- big_univLinReg(G, y[ind.gwas], ind.train = ind.gwas, ncores = NCORES)
+library(dplyr)
+df_beta <- gwas %>%
+  transmute(beta = estim, beta_se = std.err, n_eff = length(ind.gwas))
+
+# lassosum2
+beta_lassosum <- snp_lassosum2(corr, df_beta, ncores = NCORES)
+(params <- attr(beta_lassosum, "grid_param"))
+
+# validation
+pred_lassosum <- big_prodMat(G, beta_lassosum, ncores = NCORES)
+params$score <- apply(pred_lassosum[ind.val, ], 2, cor, y = y[ind.val])
+
+# pseudo-validation
+scale <- with(df_beta, sqrt(n_eff * beta_se^2 + beta^2))
+beta_hat <- df_beta$beta / scale
+
+fdr <- fdrtool::fdrtool(beta_hat, statistic = "correlation", plot = FALSE)
+beta_hat_shrunk <- round(beta_hat * (1 - fdr$lfdr), 16)
+
+params$auto_score <- apply(beta_lassosum, 2, function(beta) {
+  cat(".")
+  beta <- beta / scale
+  bRb <- crossprod(beta, bigsparser::sp_prodVec(corr, beta))
+  crossprod(beta, beta_hat_shrunk) / sqrt(bRb)
+})
+
+library(ggplot2)
+qplot(auto_score, score, color = delta, data = params) +
+  theme_bw(15) +
+  scale_color_viridis_c(trans = "log10") +
[…]

### 2021-09-07 paper/paper-misspec-supp.tex
add intro

@@
 \begin{figure}[p]
-	\centerline{\includegraphics[width=0.8\textwidth]{compare-sd}}
-	\caption{Standard deviations from ``true'' genotype calls versus from imputed dosages, colored by INFO scores. [TODO: ALL, MAKE SIDE BY SIDE WITH A/B + DESCRIBE MORE] \label{fig:compare-sd}}
-\end{figure}
-
-\begin{figure}[p]
-	\centerline{\includegraphics[width=0.8\textwidth]{compare-beta}}
-	\caption{GWAS effect sizes from ``true'' genotype calls versus from imputed dosages, colored by INFO scores. \label{fig:compare-beta}}
-\end{figure}
-
-\begin{figure}[p]
-	\centerline{\includegraphics[width=0.8\textwidth]{compare-beta-se}}
-	\caption{Standard errors of GWAS effect sizes from ``true'' genotype calls versus from imputed dosages, colored by INFO scores. \label{fig:compare-beta-se}}
+	\centerline{\includegraphics[width=0.8\textwidth]{simu_hist_info}}
+	\caption{Histogram of the imputation INFO scores of the 40,000 variants from chromosome 22 of the UK Biobank data used in the simulations. \label{fig:hist-info}}
 \end{figure}
@@
 \begin{figure}[p]
-	\centerline{\includegraphics[width=0.8\textwidth]{compare-info}}
-	\caption{INFO scores reported in the UK Biobank vs.\ INFO scores recomputed from the subset of 362,307 European individuals used in this paper.
-[TODO: FOR THE VARIANTS IN SIMU?] \label{fig:compare-info}}
+	\centerline{\includegraphics[width=0.75\textwidth]{compare-info}}
+	\caption{INFO scores reported in the UK Biobank for variants used in simulations versus INFO scores recomputed from the subset of 362,307 European individuals used in this paper.
+\label{fig:compare-info}}
 \end{figure}
@@
 \begin{figure}[p]
-	\centerline{\includegraphics[width=0.8\textwidth]{compare-mult-imp}}
-	\caption{GWAS effect sizes and Z-scores from imputed dosages versus from multiple imputation (MI, 10 draws used), ``corrected'' and colored by INFO scores.
+	\centerline{\includegraphics[width=0.95\textwidth]{compare-sd}}
+	\caption{\textbf{A:} Raw standard deviations (SDs) from imputed dosages or \textbf{B:} corrected SDs from imputed dosages (dividing them by $\sqrt{\text{INFO}}$) versus SDs from ``true'' genotype calls, colored by INFO scores.
+\label{fig:compare-sd}}
+\end{figure}
+
+\begin{figure}[p]
+	\centerline{\includegraphics[width=0.95\textwidth]{compare-beta-se}}
+	\caption{Standard errors of GWAS effect sizes (SEs) from ``true'' genotype calls versus either \textbf{A:} raw SEs from imputed dosages or \textbf{B:} corrected SEs from imputed dosages (multiplying them by $\sqrt{\text{INFO}}$), colored by INFO scores.
+\label{fig:compare-beta-se}}
+\end{figure}
+
+\begin{figure}[p]
+	\centerline{\includegraphics[width=0.95\textwidth]{compare-beta}}
+	\caption{GWAS effect sizes ($\gamma$) from ``true'' genotype calls versus either \textbf{A:} raw $\gamma$ from imputed dosages or \textbf{B:} corrected $\gamma$ from imputed dosages (multiplying them by $\sqrt{\text{INFO}}$), colored by INFO scores.
+\label{fig:compare-beta}}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{figure}[p]
+	\centerline{\includegraphics[width=0.95\textwidth]{compare-mult-imp}}
+	\caption{Corrected GWAS \textbf{A:} Z-scores and \textbf{B:} effect sizes from imputed dosages (multiplying them by $\text{INFO}$) versus the ones from multiple imputation (MI, 10 draws used), colored by INFO scores.
 This shows only one tenth of the simulated variants, because MI can be computationally intensive.
@@
 	\centerline{\includegraphics[width=0.65\textwidth]{brca_onco_hist_info}}
-	\caption{Histogram of the imputation INFO scores of the HapMap3 variants from the OncoArray GWAS summary statistics for breast cancer \cite[]{michailidou2017association}.
-[TODO: REEXPORT -- WAS CORRUPTED] \label{fig:hist-info-brca}}
+	\caption{Histogram of the imputation INFO scores of the HapMap3 variants from the OncoArray GWAS summary statistics for breast cancer \cite[]{michailidou2017association}. \label{fig:hist-info-brca}}
 \end{figure}
@@
 \begin{figure}[p]
-	\centerline{\includegraphics[width=0.8\textwidth]{hist_bad_pos}}
-	\caption{Histogram of the positions of outlier variants from Figure 2B [TODO: VERIF NUMBER AT THE END] in the main text. \label{fig:hist-bad-pos}}
+	\centerline{\includegraphics[width=0.75\textwidth]{hist_bad_pos}}
+	\caption{Histogram of the positions of outlier variants from Figure 3B in the main text.%TODO: VERIF NUMBER AT THE END
+	\label{fig:hist-bad-pos}}
 \end{figure}
@@
 	\centerline{\includegraphics[width=0.95\textwidth]{brca_icogs_qc}}
-	\caption{Standard deviations inferred from the iCOGS breast cancer GWAS summary statistics versus the ones inferred from the reported GWAS allele frequencies.
+	\caption{Standard deviations inferred from the iCOGS breast cancer GWAS summary statistics (\textbf{A:} Raw or \textbf{B:} dividing them by $\sqrt{\text{INFO}}$) versus the ones inferred from the reported GWAS allele frequencies.
 Only 100,000 variants are represented, at random.}
@@
 	\centerline{\includegraphics[width=0.95\textwidth]{cad_qc}}
-	\caption{Standard deviations inferred from the CAD GWAS summary statistics versus the ones inferred from the reported GWAS allele frequencies. Only 100,000 variants are represented, at random.}
+	\caption{Standard deviations inferred from the CAD GWAS summary statistics (\textbf{A:} Raw or \textbf{B:} dividing them by $\sqrt{\text{INFO}}$) versus the ones inferred from the reported GWAS allele frequencies.
+Only 100,000 variants are represented, at random.}
 	\label{fig:cad_qc}
@@
 	\centerline{\includegraphics[width=0.95\textwidth]{mdd_qc}}
-	\caption{Standard deviations inferred from the MDD GWAS summary statistics versus the ones inferred from the reported GWAS allele frequencies. Only 100,000 variants are represented, at random.}
+	\caption{Standard deviations inferred from the MDD GWAS summary statistics (\textbf{A:} Raw or \textbf{B:} dividing them by $\sqrt{\text{INFO}}$) versus the ones inferred from the reported GWAS allele frequencies.
+Only 100,000 variants are represented, at random.}
 	\label{fig:mdd_qc}
@@
 	\centerline{\includegraphics[width=0.95\textwidth]{prca_qc}}
-	\caption{Standard deviations inferred from the PRCA GWAS summary statistics versus the ones inferred from the reported GWAS allele frequencies.
-Only 100,000 variants are represented, at random.
-[TODO: REEXPORT FIGURE]}
+	\caption{Standard deviations inferred from the PRCA GWAS summary statistics (\textbf{A:} Raw or \textbf{B:} dividing them by $\sqrt{\text{INFO}}$) versus the ones inferred from the reported GWAS allele frequencies.
+Only 100,000 variants are represented, at random.}
 	\label{fig:prca_qc}
@@
 	\centerline{\includegraphics[width=0.95\textwidth]{t1d_affy_qc}}
-	\caption{Standard deviations inferred from the T1D (Affymetrix) GWAS summary statistics versus the ones inferred from the reported GWAS allele frequencies.
+	\caption{Standard deviations inferred from the T1D (Affymetrix) GWAS summary statistics (\textbf{A:} Raw or \textbf{B:} dividing them by $\sqrt{\text{INFO}}$) versus the ones inferred from the reported GWAS allele frequencies.
 Only 100,000 variants are represented, at random.}
@@
 \centerline{\includegraphics[width=0.95\textwidth]{t1d_illu_qc}}
-\caption{Standard deviations inferred from the T1D (Illumina) GWAS summary statistics versus the ones inferred from the reported GWAS allele frequencies.
+\caption{Standard deviations inferred from the T1D (Illumina) GWAS summary statistics (\textbf{A:} Raw or \textbf{B:} dividing them by $\sqrt{\text{INFO}}$) versus the ones inferred from the reported GWAS allele frequencies.
 Only 100,000 variants are represented, at random.}
@@
 \begin{figure}[p]
-	\centerline{\includegraphics[width=0.75\textwidth]{res-cad}}
+	\centerline{\includegraphics[width=0.7\textwidth]{res-cad}}
 	\caption{Raw partial correlations (for \textit{all} models) for predicting CAD in the UK Biobank from external summary statistics.
@@
 \begin{figure}[p]
-	\centerline{\includegraphics[width=0.75\textwidth]{res-mdd}}
+	\centerline{\includegraphics[width=0.7\textwidth]{res-mdd}}
 	\caption{Raw partial correlations (for \textit{all} models) for predicting MDD in the UK Biobank from external summary statistics.
@@
 \begin{figure}[p]
-	\centerline{\includegraphics[width=0.75\textwidth]{res-prca}}
+	\centerline{\includegraphics[width=0.7\textwidth]{res-prca}}
 	\caption{Raw partial correlations (for \textit{all} models) for predicting PRCA in the UK Biobank from external summary statistics.
@@
 
-\FloatBarrier
-
-\begin{figure}[p]
-	\centerline{\includegraphics[width=0.8\textwidth]{simu_hist_info}}
-	\caption{Histogram of the imputation INFO scores of the 40,000 variants from chromosome 22 of the UK Biobank data used in the simulations. \label{fig:hist-info}}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 %\clearpage
@@
 	\centerline{\includegraphics[width=0.85\textwidth]{pseudoval}}
-	\caption{For one simulation, scores from validation vs.\ scores from pseudo-validation as described in \cite{mak2017polygenic} using either \textbf{A:} correlations (the default in lassosum) or \textbf{B:} p-values when computing local false discovery rates.}
+	\caption{For one simulation, scores from validation versus scores from pseudo-validation as described in \cite{mak2017polygenic} using either \textbf{A:} correlations (the default in lassosum) or \textbf{B:} p-values when computing local false discovery rates.}
 	\label{fig:pseudoval}

### 2021-09-07 paper/paper-misspec.tex
add intro

@@
 
-\title{Identifying and correcting multiple sources of misspecification\\in GWAS summary statistics for prediction\\---~OR~---\\Reducing misspecifications and improving tagging\\for better polygenic scores}
-\author{Florian Priv\'e,$^{\text{1,}*}$ Julyan Arbel,$^{\text{2}}$ Timothy S. H. Mak,$^{\text{3}}$ Hugues Aschard,$^{\text{4,5}}$ and Bjarni J. Vilhj\'almsson$^{\text{1,6}}$}
+\title{Identifying and correcting multiple sources of misspecification\\in GWAS summary statistics for polygenic scores}
+\author{Florian Priv\'e,$^{\text{1,}*}$ Julyan Arbel,$^{\text{2}}$ Hugues Aschard,$^{\text{3,4}}$ and Bjarni J. Vilhj\'almsson$^{\text{1,5}}$}
 
@@
 \noindent$^{\text{\sf 2}}$Univ.\ Grenoble Alpes, Inria, CNRS, Grenoble INP, LJK, Grenoble, 38000, France. \\
-\noindent$^{\text{\sf 3}}$Fano Labs, Hong Kong. \\
-\noindent$^{\text{\sf 4}}$Department of Computational Biology, Institut Pasteur, Paris, 75015, France. \\
-\noindent$^{\text{\sf 5}}$Program in Genetic Epidemiology and Statistical Genetics, Harvard T.H. Chan School of Public Health, Boston, MA, 02115, USA. \\
-\noindent$^{\text{\sf 6}}$Bioinformatics Research Centre, Aarhus University, Aarhus, 8000, Denmark. \\
+\noindent$^{\text{\sf 3}}$Department of Computational Biology, Institut Pasteur, Paris, 75015, France. \\
+\noindent$^{\text{\sf 4}}$Program in Genetic Epidemiology and Statistical Genetics, Harvard T.H. Chan School of Public Health, Boston, MA, 02115, USA. \\
+\noindent$^{\text{\sf 5}}$Bioinformatics Research Centre, Aarhus University, Aarhus, 8000, Denmark. \\
 \noindent$^\ast$To whom correspondence should be addressed.\\
@@
 
-\clearpage
+\vspace*{5em}
 
@@
 
-\vspace*{5em}
-
-[SHOULD ADD THE LDAK-ALPHA IN THIS WORK? MAYBE LATER -> IN A THIRD VERSION]
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-%\clearpage
+\clearpage
 
@@
 
-TODO
+Multiple studies have shown that LDpred2 and lassosum are consistently ranking best among methods for polygenic prediction \cite[]{mak2017polygenic,prive2020ldpred2,pain2021evaluation,kulm2021systematic}.
+These two methods can be widely used because they require only summary statistics from genome-wide association studies (GWAS), as opposed to individual-level genotypes and phenotypes.
+However, GWAS summary statistics come with different qualities, such as the imputation qualities of each variant reported.
+They also come with some heterogeneity in the methods used for performing the different GWAS and meta-analyses, as well as the information reported in the summary statistics.
+LDpred2, as many other Bayesian polygenic methods \cite[]{ge2019polygenic,lloyd2019improved}, uses a Gibbs sampler, which is known to be sensitive to model misspecifications [TODO: @JULYAN, GOOD REFS FOR THAT?].
+Previously, to make sure that input parameters passed to LDpred2 (and especially its ``auto'' version) were consistent with what LDpred2 was expecting, we proposed a quality control (QC) based on comparing standard deviations derived from GWAS summary statistics with the ones from a reference panel \cite[]{prive2020ldpred2}.
 
-[WHY CHOOSED TO FOCUS ON LDPRED AND LASSOSUM -> COPY PREVIOUS INTRO + MORE CONTROL ON THE IMPLEMENTATION]
-
-[WHAT I WANTED TO INVESTIGATE]
+Here, we further investigate some of the possible misspecifications that come with the use of GWAS summary statistics. 
+We aim at identifying and correcting multiple sources of misspecifications in the summary statistics in order to improve the predictive performance of polygenic scores.
+We work on three different angles.
+First, based on other information such as the imputation INFO scores and allele frequencies from the GWAS summary statistics, we refine the QC we proposed in \cite{prive2020ldpred2}.
+For instance, we show that standard deviations of imputed genotypes (allele dosages) are lower than the expected values under Hardy-Weinberg equilibrium.
+Second, we investigate possible corrections to apply to the input parameters of polygenic methods, namely the LD (linkage disequilibrium) reference, the GWAS effect sizes, their standard errors and the corresponding sample sizes.
+For example, we show that GWAS effect sizes computed from imputed dosages are larger than if true genotypes were available.
+Third, we introduce small changes to LDpred2 to make it more robust to these misspecifications.
+For further investigations, we also reimplement a new version of lassosum, called lassosum2, that uses the exact same input parameters as LDpred2, and test the QCs and corrections for lassosum2 as well.
+We first perform our investigations in simulations, then we use published summary statistics while restricting to the widely-used HapMap3 variants, finally we investigate two alternative sets with more variants.
 
@@
 Here we show that we can refine this quality control by deriving the second standard deviations directly from the reported allele frequencies in the GWAS summary statistics, and by correcting for the INFO scores of the GWAS (e.g.\ see figure \ref{fig:brca_onco_qc}).
-Using this refined QC, we were able to identify a potential issue with how principal components were derived in a set of breast cancer summary statistics.
-Fortunately this has practically no effect on the predictive performance of the derived polygenic scores.
+Using this refined QC, we are able to identify a potential issue with how principal components were derived in a set of breast cancer summary statistics.
+Fortunately, this has practically no effect on the predictive performance of the derived polygenic scores.
 Additional QC can also be performed, e.g.\ comparing reported GWAS allele frequencies with the ones from the LD reference panel, e.g.\ to detect genotyping or allelic errors.
@@
 Even the total effective sample size is sometimes difficult to assess.
-Indeed, often only the total number of cases and controls are reported in the abstract, and where numbers possibly includes sizes before QC; therefore one has to dig into supplementary materials to find this information.
-We recall that if some studies of a meta-analysis have an imbalanced number of cases and controls, the global effective sample size should not be computed from the total numbers of cases and controls overall, but instead from the sum of each effective sample size.
+Indeed, often only the total number of cases and controls are reported in the abstract of GWAS papers, where reported numbers possibly includes sizes before QC; therefore one has to dig into supplementary materials to find this information.
+We also recall that if some studies of a meta-analysis have an imbalanced number of cases and controls, the global effective sample size should not be computed from the total numbers of cases and controls overall, but instead from the sum of each effective sample size.
 Indeed, take the extreme example of meta-analyzing two studies, one with 1000 cases and 0 controls, and another one with 0 cases and 1000 controls, then the effective sample size of the meta-analysis is 0, not 2000.
-Fortunately, an overestimated sample size can be detected from the QC plot we propose (where the slope is then less than 1) for case-control studies using logistic regression; otherwise the standard deviation of the phenotype is also needed (Equation \ref{eq:approx-sd-lin}).
+Fortunately, an overestimated sample size can be detected from the QC plot we propose (where the slope is then less than 1) for case-control studies using logistic regression; otherwise the standard deviation of the phenotype is also needed (Equation \ref{eq:approx-sd-lin}), but can be estimate \cite[]{prive2020ldpred2}.
 
 We have assessed the impact of these misspecifications in GWAS summary statistics on the predictive performance of some polygenic score methods.
-Using LDpred2 models \cite[]{prive2020ldpred2} and our reimplementation of the lassosum model \cite[]{mak2017polygenic} for deriving polygenic scores, we have introduced and investigated some changes to possibly make these models more robust to misspecifications.
+Using both the Bayesian LDpred2 models \cite[]{prive2020ldpred2} and our reimplementation of the frequentist lassosum model \cite[]{mak2017polygenic} for deriving polygenic scores, we have introduced and investigated some changes to possibly make these models more robust to misspecifications.
 Overall, these changes have provided large improvements of predictive performance in simulations with large misspecifications.
 However, they have almost no effect when using the published GWAS summary statistics, except for breast cancer and coronary artery disease summary statistics.
-This is a little disappointing, but more importantly it is also reassuring because it means that the summary statistics we use in this study are usually of good quality for follow-up analyses such as deriving polygenic scores.
+This is slightly disappointing, but more importantly it is also reassuring because it means that the summary statistics we use in this study are usually of good quality for follow-up analyses such as deriving polygenic scores.
 
@@
 
-Authors thank Matthew Stephens for helpful discussions. 
+Authors thank Timothy Shin Heng Mak, Shing Wan Choi, and Matthew Stephens for helpful discussions. 
 Authors also thank GenomeDK and Aarhus University for providing computational resources and support that contributed to these research results.
@@
 %\clearpage
+\vspace*{3em}
 

### 2021-09-13 paper/paper-misspec.tex
minor text changes

@@
 
-Multiple studies have shown that LDpred2 and lassosum are consistently ranking best among methods for polygenic prediction \cite[]{mak2017polygenic,prive2020ldpred2,pain2021evaluation,kulm2021systematic}.
-These two methods can be widely used because they require only summary statistics from genome-wide association studies (GWAS), as opposed to individual-level genotypes and phenotypes.
-However, GWAS summary statistics come with different qualities, such as the imputation qualities of each variant reported.
-They also come with some heterogeneity in the methods used for performing the different GWAS and meta-analyses, as well as the information reported in the summary statistics.
-LDpred2, as many other Bayesian polygenic methods \cite[]{ge2019polygenic,lloyd2019improved}, uses a Gibbs sampler, which is known to be sensitive to model misspecifications [TODO: @JULYAN, GOOD REFS FOR THAT?].
-Previously, to make sure that input parameters passed to LDpred2 (and especially its ``auto'' version) were consistent with what LDpred2 was expecting, we proposed a quality control (QC) based on comparing standard deviations derived from GWAS summary statistics with the ones from a reference panel \cite[]{prive2020ldpred2}.
+Contrary to individual-level genotypes and phenotypes, summary statistics from genome-wide association studies (GWAS) are widely available.
+Therefore, they have been extensively used e.g.\ to derive polygenic scores (PGS).
+However, GWAS summary statistics come with uneven qualities, such as the imputation qualities of each variant reported. 
+There is also some heterogeneity in the methods used for performing the individual GWAS and their meta-analyses, as well as the information reported in the resulting summary statistics.
+Moreover, many PGS methods such as PRS-CS, SBayesR, and LDpred2 \cite[]{ge2019polygenic,lloyd2019improved,prive2020ldpred2} use Bayesian models, which can be sensitive to model misspecifications \cite[]{WALKER20131621,miller2018robust}.
+Previously, to make sure that input parameters passed to LDpred2 were consistent with its modeling assumptions, we proposed a quality control (QC) based on comparing standard deviations derived from GWAS summary statistics with the ones from a reference panel \cite[]{prive2020ldpred2}.
+This was particularly important for LDpred2-auto, which also estimates key model parameters from the data directly.
 
-Here, we further investigate some of the possible misspecifications that come with the use of GWAS summary statistics. 
-We aim at identifying and correcting multiple sources of misspecifications in the summary statistics in order to improve the predictive performance of polygenic scores.
-We work on three different angles.
-First, based on other information such as the imputation INFO scores and allele frequencies from the GWAS summary statistics, we refine the QC we proposed in \cite{prive2020ldpred2}.
+Here, we investigate some of the possible misspecifications that come with GWAS summary statistics. 
+We aim to identify and correct multiple sources of misspecification in the summary statistics in order to improve the predictive performance of polygenic scores.
+We approach this from three different angles.
+First, based on additional information such as the imputation INFO scores and allele frequencies from the GWAS summary statistics, we refine our previously proposed QC \cite{prive2020ldpred2}.
 For instance, we show that standard deviations of imputed genotypes (allele dosages) are lower than the expected values under Hardy-Weinberg equilibrium.
-Second, we investigate possible corrections to apply to the input parameters of polygenic methods, namely the LD (linkage disequilibrium) reference, the GWAS effect sizes, their standard errors and the corresponding sample sizes.
-For example, we show that GWAS effect sizes computed from imputed dosages are larger than if true genotypes were available.
-Third, we introduce small changes to LDpred2 to make it more robust to these misspecifications.
-For further investigations, we also reimplement a new version of lassosum, called lassosum2, that uses the exact same input parameters as LDpred2, and test the QCs and corrections for lassosum2 as well.
-We first perform our investigations in simulations, then we use published summary statistics while restricting to the widely-used HapMap3 variants, finally we investigate two alternative sets with more variants.
+Second, we investigate possible corrections to apply to the input parameters of polygenic methods, namely the reference LD (linkage disequilibrium) matrix, the GWAS effect sizes, their standard errors and the corresponding sample sizes.
+For example, we show that GWAS effect sizes computed from imputed dosages are larger in magnitude compared to if true genotypes were available.
+Third, we introduce two new optional parameters in LDpred2-auto to make it more robust to these types of misspecification.
+We focus our investigations on LDpred2 and lassosum for two reasons.
+First, multiple studies have shown that LDpred2 and lassosum are consistently ranking best among methods for polygenic prediction \cite[]{mak2017polygenic,prive2020ldpred2,pain2021evaluation,kulm2021systematic}.
+Second, we reimplement and use a new version of lassosum, called lassosum2, that uses the exact same input parameters as LDpred2, which makes it easy for us to test the QCs and corrections presented here.
+We perform our investigations using simulations first, then we use public summary statistics while restricting to the widely-used HapMap3 variants, finally we investigate two alternative sets with more variants.
 
@@
 In the real data applications hereinafter, we choose to use the first correction, ``sqrt\_info'', which is simple because it is equivalent to post-processing PGS effects by multiplying them by $\sqrt{\text{INFO}}$.
-
-[SHOULD I USE ``INFO'' INSTEAD? BUT I NEED TO RERUN EVERYTHING BECAUSE I CAN'T JUST POSTPROCESS RESULTS. AND NOT BETTER FOR LDPRED2-LOW-H2 AND LDPRED2-AUTO-ROB]
 
@@
 
-Here we have investigated misspecifications in GWAS summary statistics, including when they are derived from imputed data, and when the information on sample size per variant is different for all variants because of meta-analysis.
+We have investigated misspecifications in GWAS summary statistics, focusing on the impact of sample size heterogeneity and imputation quality.
 Previously, we proposed a quality control (QC) based on comparing standard deviations derived from GWAS summary statistics with the ones from a reference panel \cite[]{prive2020ldpred2}.
-Here we show that we can refine this quality control by deriving the second standard deviations directly from the reported allele frequencies in the GWAS summary statistics, and by correcting for the INFO scores of the GWAS (e.g.\ see figure \ref{fig:brca_onco_qc}).
+Here we show that we can refine this quality control by deriving the latter directly from the reported allele frequencies in the GWAS summary statistics, and by correcting the former using the GWAS imputation INFO scores (e.g.\ see figure \ref{fig:brca_onco_qc}).
 Using this refined QC, we are able to identify a potential issue with how principal components were derived in a set of breast cancer summary statistics.
@@
 Another information, the effective sample size per variant, is often missing from GWAS summary statistics. 
-Even the total effective sample size is sometimes difficult to assess.
-Indeed, often only the total number of cases and controls are reported in the abstract of GWAS papers, where reported numbers possibly includes sizes before QC; therefore one has to dig into supplementary materials to find this information.
-We also recall that if some studies of a meta-analysis have an imbalanced number of cases and controls, the global effective sample size should not be computed from the total numbers of cases and controls overall, but instead from the sum of each effective sample size.
+Sometimes, it can even be challenging to recover the total effective sample size from large meta-analyses.
+We recall that if some studies of a meta-analysis have an imbalanced number of cases and controls, the global effective sample size should not be computed from the total numbers of cases and controls overall, but instead from the sum of the effective sample sizes of each study.
 Indeed, take the extreme example of meta-analyzing two studies, one with 1000 cases and 0 controls, and another one with 0 cases and 1000 controls, then the effective sample size of the meta-analysis is 0, not 2000.
-Fortunately, an overestimated sample size can be detected from the QC plot we propose (where the slope is then less than 1) for case-control studies using logistic regression; otherwise the standard deviation of the phenotype is also needed (Equation \ref{eq:approx-sd-lin}), but can be estimate \cite[]{prive2020ldpred2}.
+Fortunately, an overestimated sample size can be detected from the QC plot we propose, where the slope is then less than 1 for case-control studies using logistic regression; otherwise the standard deviation of the phenotype is also needed (Equation \ref{eq:approx-sd-lin}), but can be estimated \cite[]{prive2020ldpred2}.
 
@@
 Overall, these changes have provided large improvements of predictive performance in simulations with large misspecifications.
-However, they have almost no effect when using the published GWAS summary statistics, except for breast cancer and coronary artery disease summary statistics.
-This is slightly disappointing, but more importantly it is also reassuring because it means that the summary statistics we use in this study are usually of good quality for follow-up analyses such as deriving polygenic scores.
+However, they have almost no effect when using the real GWAS summary statistics we chose, except for breast cancer and coronary artery disease summary statistics.
+Although these results are somewhat unfortunate, they are reassuring because it means that the summary statistics we use in this study are usually of good quality for follow-up analyses such as deriving polygenic scores.
 

### 2021-09-14 paper/paper-misspec.tex
finished first draft

@@
 
-\vspace*{5em}
+\vspace*{8em}
 
 \abstract{
+	Are genome-wide association studies (GWAS) summary statistics of good enough quality for performing follow-up analyses?
+Can we detect possible misspecifications in GWAS summary statistics and correct for them in order to improve predictive performance of polygenic scores?
 }
@@
 
-Contrary to individual-level genotypes and phenotypes, summary statistics from genome-wide association studies (GWAS) are widely available.
+Contrary to individual-level genotypes and phenotypes, summary statistics resulting from genome-wide association studies (GWAS) are widely available.
 Therefore, they have been extensively used e.g.\ to derive polygenic scores (PGS).
@@
 Moreover, many PGS methods such as PRS-CS, SBayesR, and LDpred2 \cite[]{ge2019polygenic,lloyd2019improved,prive2020ldpred2} use Bayesian models, which can be sensitive to model misspecifications \cite[]{WALKER20131621,miller2018robust}.
-Previously, to make sure that input parameters passed to LDpred2 were consistent with its modeling assumptions, we proposed a quality control (QC) based on comparing standard deviations derived from GWAS summary statistics with the ones from a reference panel \cite[]{prive2020ldpred2}.
-This was particularly important for LDpred2-auto, which also estimates key model parameters from the data directly.
+Previously, to make sure that input parameters passed to LDpred2 were consistent with its modeling assumptions, we proposed a quality control (QC) based on comparing standard deviations inferred from GWAS summary statistics with the ones computed from a reference panel \cite[]{prive2020ldpred2}.
+This was particularly important for LDpred2-auto, which directly estimates key model parameters from the data.
 
@@
 We approach this from three different angles.
-First, based on additional information such as the imputation INFO scores and allele frequencies from the GWAS summary statistics, we refine our previously proposed QC \cite{prive2020ldpred2}.
+First, based on additional summary information such as the imputation INFO scores and allele frequencies from the GWAS summary statistics, we refine our previously proposed QC \cite{prive2020ldpred2}.
 For instance, we show that standard deviations of imputed genotypes (allele dosages) are lower than the expected values under Hardy-Weinberg equilibrium.
@@
 \begin{figure}[h]
-	\centerline{\includegraphics[width=0.95\textwidth]{simu-misN}}
+	\centerline{\includegraphics[width=0.98\textwidth]{simu-misN}}
 	\caption{Results for the simulations with sample size misspecification, averaged over 10 simulations for each scenario.
@@
 \begin{figure}[h]
-	\centerline{\includegraphics[width=0.95\textwidth]{simu-info}}
+	\centerline{\includegraphics[width=0.98\textwidth]{simu-info}}
 	\caption{Results of predictive performance for the simulations using GWAS summary statistics from imputed dosage data, averaged over 10 simulations for each scenario.
@@
 PC19 from the UK Biobank was previously reported to capture LD structure in region 70-91 Mbp of chromsome 6 \cite[]{prive2020efficient}.
-In these simulations, the same bias as in figure \ref{fig:hist-bad-pos}B is observed for the variants in this region (Figure \ref{fig:gwas_bad_pc}), confirming our hypothesis.
+In these simulations, the same bias as in figure \ref{fig:brca_onco_qc}B is observed for the variants in this region (Figure \ref{fig:gwas_bad_pc}), confirming our hypothesis.
 
 \begin{figure}[h]
-	\centerline{\includegraphics[width=0.95\textwidth]{brca_onco_qc}}
+	\centerline{\includegraphics[width=0.98\textwidth]{brca_onco_qc}}
 	\caption{Standard deviations inferred from the OncoArray breast cancer GWAS summary statistics (Equation \ref{eq:approx-sd-log}) versus the ones inferred from the reported GWAS allele frequencies ($\sqrt{2 f (1 - f)}$).
@@
 \begin{figure}[h]
-	\centerline{\includegraphics[width=0.95\textwidth]{res-brca}}
+	\centerline{\includegraphics[width=0.98\textwidth]{res-brca}}
 	\caption{Raw partial correlations (for \textit{all} models) for predicting breast cancer in the UK Biobank when using either the OncoArray or the iCOGS summary statistics.
@@
 We show previous results as figure \ref{fig:res_brca} for other phenotypes in figures \ref{fig:res_cad}-\ref{fig:res_t1d}; most changes introduced before have little or no impact on the predictions, except for the second quality control performed for CAD when using LDpred2-auto (Figure \ref{fig:res_cad}).
-
-We then introduce two new sets of variants as possible replacement of HapMap3 variants (Methods).
+We then introduce two new sets of variants as possible replacement for HapMap3 variants (Methods).
 These sets include more variants (more than 2M), therefore possibly of lower quality on average. 
-We show results with these two new sets of variants (Methods) in figures \ref{fig:res_brca_all}-\ref{fig:res_t1d_all}.
+We present similar results with these two new sets of variants in figures \ref{fig:res_brca_all}-\ref{fig:res_t1d_all}.
 Using the ``maxtag'' set generally provides larger predictive performance than using HapMap3 variants, especially when predicting prostate cancer (Figure \ref{fig:res_prca_all}).
@@
 We have investigated misspecifications in GWAS summary statistics, focusing on the impact of sample size heterogeneity and imputation quality.
-Previously, we proposed a quality control (QC) based on comparing standard deviations derived from GWAS summary statistics with the ones from a reference panel \cite[]{prive2020ldpred2}.
+Previously, we proposed a quality control (QC) based on comparing standard deviations inferred from GWAS summary statistics with the ones computed from a reference panel \cite[]{prive2020ldpred2}.
 Here we show that we can refine this quality control by deriving the latter directly from the reported allele frequencies in the GWAS summary statistics, and by correcting the former using the GWAS imputation INFO scores (e.g.\ see figure \ref{fig:brca_onco_qc}).
@@
 We acknowledge that, in the case of a meta-analysis from multiple studies, providing a single INFO score per variant may not be possible; would a weighted averaged INFO score work?
-Nevertheless, this quality control could be easily performed within each study before pooling results, to make sure that summary statistics have the best possible quality for follow-up analyses such as deriving polygenic scores.
+Nevertheless, this quality control could be performed within each study before pooling results, to make sure that summary statistics have the best possible quality for follow-up analyses such as deriving polygenic scores.
 Another information, the effective sample size per variant, is often missing from GWAS summary statistics. 

### 2021-12-02 code/compute-all-info-eur.R
save code

@@ -1,0 +1,106 @@
+## Get all SNP IDs and split them in blocks
+library(dplyr)
+
+all_snp_id <- runonce::save_run(
+  purrr::map_dfr(1:22, function(chr) {
+
+    print(chr)
+
+    bgifile <- paste0("UKBB/bgen/ukb_imp_chr", chr, "_v3.bgen.bgi")
+    db_con <- RSQLite::dbConnect(RSQLite::SQLite(), bgifile)
+    on.exit(RSQLite::dbDisconnect(db_con), add = TRUE)
+
+    snp_id <- tbl(db_con, "Variant") %>%
+      select(chr = chromosome, pos = position, a1 = allele1, a0 = allele2) %>%
+      collect() %>%
+      with(paste(chr, pos, a1, a0, sep = "_"))
+
+    all_splits <- bigparallelr::split_vec(snp_id, block_len = 50e3)
+
+    tibble(bgifile, snp_id = all_splits)
+  }),
+  file = "tmp-data/split_all_variants.rds"
+)
+
+
+## Loop over all blocks
+grid <- mutate(all_snp_id, id = row_number())  # 1874 parts
+
+library(future.batchtools)
+NCORES <- 15
+plan(batchtools_slurm(resources = list(
+  t = "12:00:00", c = NCORES + 1, mem = "100g",
+  name = basename(rstudioapi::getSourceEditorContext()$path))))
+
+bigassertr::assert_dir("tmp-data/bgen")
+bigassertr::assert_dir("all_info")
+
+furrr::future_pwalk(grid, function(bgifile, snp_id, id) {
+
+  runonce::save_run({
+    tmp <- tempfile(tmpdir = "tmp-data/bgen")
+    on.exit(file.remove(paste0(tmp, c(".bk", ".rds"))), add = TRUE)
+
+    library(bigsnpr)
+    map <- snp_attach(snp_readBGEN(
+      bgenfiles   = sub("\\.bgi$", "", bgifile),
+      list_snp_id = list(snp_id),
+      backingfile = tmp,
+      ind_row     = readRDS("tmp-data/eur_ind_bgen.rds"),
+      ncores      = nb_cores()
+    ))$map
+  }, file = paste0("all_info/part", id, ".rds"))
+})
+
+
+## Look at chr22
+all_info <- purrr::map_dfr(grep("chr22", grid$bgifile), function(id) {
+  readRDS(paste0("all_info/part", id, ".rds"))
[…]

### 2021-12-02 code/prepare-genotypes.R

@@ -36,6 +36,7 @@ dist <- bigutilsr::dist_ogk(as.matrix(df0[sub, -(1:2)]))
 hist(log(dist), "FD")
 sub2 <- sub[log(dist) < 5]
 length(sub2) # 362,307
+# saveRDS(ind.indiv[sub2], "tmp-data/eur_ind_bgen.rds")
 
 
 # Validation set

### 2021-12-02 tmp-code/plot-approx-sd2.R

@@ -1,0 +1,59 @@
+file <- runonce::download_file(
+  "ftp://share.sph.umich.edu/UKBB_SAIGE_HRC/PheCode_191.11_SAIGE_MACge20.txt.vcf.gz",
+  dir = "tmp-data"
+)
+R.utils::gunzip(file)
+sumstats <- bigreadr::fread2("tmp-data/PheCode_191.11_SAIGE_MACge20.txt.vcf")
+sumstats <- dplyr::filter(sumstats, af > 0.005)
+
+hm3_info <- readRDS(url("https://github.com/privefl/bigsnpr/raw/master/data-raw/hm3_variants.rds"))
+
+match_info <- bigsnpr::snp_match(
+  sumstats = dplyr::rename(sumstats, chr = "#CHROM", pos = POS, a1 = ALT, a0 = REF),
+  info_snp = hm3_info
+)
+# 11,904,913 variants to be matched.
+# 1,790,164 ambiguous SNPs have been removed.
+# 1,119,564 variants have been matched; 0 were flipped and 783,941 were reversed.
+
+# https://github.com/privefl/paper-ldpred2/blob/master/paper/paper-ldpred2-supp.pdf
+df <- dplyr::mutate(match_info,
+                    sd_af = sqrt(2 * af * (1 - af)),
+                    n_eff = 4 / (1 / num_cases + 1 / num_controls),
+                    sd_ss = 2 / (sebeta * sqrt(n_eff)))
+
+# Reduce points for plotting (but try to keep outliers)
+hist(ratio <- with(df, sd_ss / sd_af))
+hist(out <- abs(ratio - median(ratio)) / mad(ratio))
+ind <- sample(nrow(df), 50e3, prob = out)
+
+library(ggplot2)
+theme_set(bigstatsr::theme_bigstatsr())
+
+ggplot(df[ind, ]) +
+  geom_point(aes(sd_af, sd_ss)) +
+  geom_abline(color = "red", size = 2)
+# ggsave("tmp-figures/qcplot1.png", width = 7, height = 6)
+
+ggplot(df[ind, ]) +
+  geom_point(aes(sd_af, sd_ss, color = INFO)) +
+  scale_color_viridis_c()
+# ggsave("tmp-figures/qcplot2.png", width = 7, height = 6)
+
+ggplot(df[ind, ]) +
+  geom_point(aes(sd_af, sd_ss / sqrt(INFO), color = INFO)) +
[…]

### 2021-12-02 paper/paper-misspec.tex
start overview table

@@
 %\usepackage[authoryear]{natbib}
-\usepackage{natbib} \bibpunct{(}{)}{;}{author-year}{}{,}
+\usepackage{natbib} %\bibpunct{(}{)}{;}{author-year}{}{,}
 
@@
 \let\tabend\endtabular
-\renewenvironment{tabular}{\begin{adjustbox}{max width=0.9\textwidth}\tabbeg}{\tabend\end{adjustbox}}
+\renewenvironment{tabular}{\begin{adjustbox}{max width=0.95\textwidth}\tabbeg}{\tabend\end{adjustbox}}
 
@@
 
-Here, we investigate some of the possible misspecifications that come with GWAS summary statistics. 
-We aim to identify and correct multiple sources of misspecification in the summary statistics in order to improve the predictive performance of polygenic scores.
+Here, we present an overview of the possible misspecifications that come with GWAS summary statistics (Table \ref{tab:overview}).
+Then, we investigate some of those, and propose corrections to improve the predictive performance of polygenic scores derived from GWAS summary statistics.
 We approach this from three different angles.
@@
 We perform our investigations using simulations first, then we use public summary statistics while restricting to the widely-used HapMap3 variants, finally we investigate two alternative sets with more variants.
+
+\begin{table}[h]
+	\centering
+	\begin{tabular}{|p{\dimexpr0.2\textwidth-2\tabcolsep-\arrayrulewidth\relax}|
+			p{\dimexpr0.5\textwidth-2\tabcolsep-\arrayrulewidth\relax}|
+			p{\dimexpr0.3\textwidth-2\tabcolsep-\arrayrulewidth\relax}|}
+		\hline
+		Misspecification & Possible consequence & Possible correction \\
+		\hline
+		\hline
+		Global sample size misspecification$^{(1)}$ & Bias estimation of SNP heritability \cite[]{gazal2018functional,grotzinger2021pervasive} & Estimation of the effective sample size \\
+		\hline
+		Differences in per-variant sample size & Break some model assumptions & Imputation using equations \ref{eq:imputeN} and \ref{eq:imputeNeff}, and filtering$^{(2)}$   \\
+		\hline
+		Using imputed dosages for GWAS & Standard deviations of dosages are smaller, resulting in overestimated effect sizes and standard errors & See section \ref{sec:imputation} \\
+		\hline
+		Using imputed dosages for LD computation & Overestimated LD [TODO: ALSO VERIFY WITH 1000G] & See correction in section \ref{sec:LD} \\
+		\hline 
+		Ancestry mismatch between summary statistics and LD & Strong mismatch of LD and allele frequencies, often resulting in divergence of models & Check ancestry proportions from GWAS summary statistics \cite[]{prive2021using} \\
+		\hline
+		Error in summary statistics (e.g.\ allele inversions) & Strong violation of model assumptions, which can result in e.g.\ identifying false-positives in fine-mapping \cite[]{susie} & Appropriate quality control \\
+		\hline
+		Any & Violation of model assumptions and bad model convergence & Using more regularization \\
+		\hline
+	\end{tabular}
+	\caption{Overview of possible misspecifications when using GWAS summary statistics. 
+	$^{(1)}$ Examples: using the global number of cases and controls in a meta-analysis of binary traits, an often larger effective sample size when using BOLT-LMM summary statistics \cite[]{loh2018mixed}, and a reduced effective sample size when using SAIGE on binary traits with low case-control imbalance \cite[]{zhou2018efficiently,mbatchou2021computationally}.
+	$^{(2)}$ E.g.\ removeing SNPs with an effective sample size less than 0.67 times the 90th percentile of sample size \cite[]{zheng2017ld}.
+	\label{tab:overview}}
+\end{table}
 
@@
 
-\subsection{When using allele dosages from imputation}
+\subsection{When using allele dosages from imputation}\label{sec:imputation}
 
@@
 
-Note that, in this study, we use summary statistics that include extended information (e.g.\ INFO scores and allele frequencies), yet most GWAS summary statistics do not. 
+Note that, in this study, we use summary statistics that include extended information (e.g.\ INFO scores and allele frequencies), yet most GWAS summary statistics do not \cite[]{macarthur2021workshop}.
 We acknowledge that, in the case of a meta-analysis from multiple studies, providing a single INFO score per variant may not be possible; would a weighted averaged INFO score work?
@@
 Indeed, take the extreme example of meta-analyzing two studies, one with 1000 cases and 0 controls, and another one with 0 cases and 1000 controls, then the effective sample size of the meta-analysis is 0, not 2000.
+This misspecification can lead to serious issues such as misestimating the SNP heritability \cite[]{grotzinger2021pervasive}.
 Fortunately, an overestimated sample size can be detected from the QC plot we propose, where the slope is then less than 1 for case-control studies using logistic regression; otherwise the standard deviation of the phenotype is also needed (Equation \ref{eq:approx-sd-lin}), but can be estimated \cite[]{prive2020ldpred2}.
@@
 
-\subsection{New LD reference}
+\subsection{New LD reference}\label{sec:LD}
 

### 2022-01-25 paper/paper-misspec-supp.tex
start overview table
save

@@
 	\centerline{\includegraphics[width=0.95\textwidth]{compare-cor}}
-	\caption{Comparing correlations between a subset of HapMap3 variants on chromosome 22, computed in three different ways: 1/ from imputed dosages; 2/ from multiple imputation (i.e.\ generating multiple complete datasets sampled according to imputation probabilities, computing correlations, and averaging results); from imputed dosages with the multiplicative correction of $\sqrt{\text{INFO}_j \cdot \text{INFO}_k}$ (for $j \neq k$).}
+	\caption{Comparing correlations between a subset of HapMap3 variants on chromosome 1 with an INFO score lower than 0.85, computed in three different ways: 1/ from imputed dosages using 10,000 individuals from the UK Biobank (UKBB) data; 2/ from multiple imputation (i.e.\ generating multiple complete datasets sampled according to imputation probabilities, computing correlations, and averaging results) also using UKBB; 3/ from 190 individuals from the 1000 Genomes data (GBR and CEU).
+Each point, representing the correlation between two variants, is colored by the geometric mean of the INFO scores of these two variants.}
 	\label{fig:compare-cor}

### 2022-01-25 paper/paper-misspec.tex
save

@@
 		\hline
-		Misspecification & Possible consequence & Possible correction \\
+		Misspecification & Possible consequence(s) & Possible correction \\
 		\hline
@@
 		\hline
-		Using imputed dosages for LD computation & Overestimated LD [TODO: ALSO VERIFY WITH 1000G] & See correction in section \ref{sec:LD} \\
+		Using imputed dosages for LD computation & LD seems correctly estimated (Figure \ref{fig:compare-cor}) & None required \\
 		\hline 
@@
 		\hline
-		Error in summary statistics (e.g.\ allele inversions) & Strong violation of model assumptions, which can result in e.g.\ identifying false-positives in fine-mapping \cite[]{susie} & Appropriate quality control \\
+		Error in summary statistics (e.g.\ allele inversions) & Strong violation of model assumptions, which can result in e.g.\ identifying false-positives in fine-mapping \cite[]{susie} & Quality control \\
 		\hline
-		Any & Violation of model assumptions and bad model convergence & Using more regularization \\
+		Any & Violation of model assumptions resulting in poor model convergence & Using more regularization \\
 		\hline
@@
 	$^{(1)}$ Examples: using the global number of cases and controls in a meta-analysis of binary traits, an often larger effective sample size when using BOLT-LMM summary statistics \cite[]{loh2018mixed}, and a reduced effective sample size when using SAIGE on binary traits with low case-control imbalance \cite[]{zhou2018efficiently,mbatchou2021computationally}.
-	$^{(2)}$ E.g.\ removeing SNPs with an effective sample size less than 0.67 times the 90th percentile of sample size \cite[]{zheng2017ld}.
+	$^{(2)}$ E.g.\ removing SNPs with an effective sample size less than 0.67 times the 90th percentile of sample size \cite[]{zheng2017ld}.
 	\label{tab:overview}}
@@
 
-\subsection{New implementation of lassosum in bigsnpr}
+\subsection{New implementation of lassosum in bigsnpr}\label{lassosum2}
 
@@
 
-We make three changes to the LD reference. 
-First, when using imputed data, we multiply the correlation between variants $j$ and $k$ by $\sqrt{\text{INFO}_j \cdot \text{INFO}_k}$ (for $j \neq k$) since it approximates well the correlation from non-imputed data (Figure \ref{fig:compare-cor}).
-Second, we also define nearly independent LD blocks using the optimal algorithm developed in \cite{prive2021optimal}.
-For different numbers of blocks and maximum number of variants in each block, we use the split with the minimum cost within the ones reducing the original number of non-zero values to less than 60\% (70\% for chromosome 6).
+We define nearly independent LD blocks using the optimal algorithm developed in \cite{prive2021optimal}.
+For different numbers of blocks and maximum number of variants in each block, we use the split with the minimum cost within the ones reducing the original number of non-zero values to less than 60\% (or 70\% for chromosome 6).
 Having a correlation matrix with independent blocks prevents the small errors in the algorithm (e.g.\ the Gibbs sampler in LDpred2) from propagating to too many variants.
 It also makes running LDpred2 (and lassosum2) faster, taking about 60\% of the initial time (since only 60\% of the initial non-zero values of the correlation matrix are kept).
-Finally, we have developed a new ``compact'' format for the SFBMs (sparse matrices on disk).
+
+We have also developed a new ``compact'' format for the SFBMs (sparse matrices on disk).
 Instead of using something similar to the standard ``compressed sparse column'' format which stores all $\{i,~x(i, j)\}$ for a given column $j$, we only store the first index $i_0$ and all the contiguous values $\{x(i_0, j),~x(i_0 + 1, j),~\dots\}$ up to the last non-zero value for this column $j$.
@@
 The first one, \texttt{shrink\_corr}, allows for shrinking off-diagonal elements of the correlation matrix.
-This is similar to parameter `s' in lassosum, and act as a regularization.
+This is similar to parameter `s' in lassosum (Section \ref{lassosum2}), and act as a means of regularization.
 We use a value of 0.9 in simulations and 0.95 in real data when running ``LDpred2-auto-rob'' (and the default value of 1, without any effect, when running ``LDpred2-auto'').

### 2022-01-27 paper/paper-misspec.tex
save

@@
 Moreover, many PGS methods such as PRS-CS, SBayesR, and LDpred2 \cite[]{ge2019polygenic,lloyd2019improved,prive2020ldpred2} use Bayesian models, which can be sensitive to model misspecifications \cite[]{WALKER20131621,miller2018robust}.
-Previously, to make sure that input parameters passed to LDpred2 were consistent with its modeling assumptions, we proposed a quality control (QC) based on comparing standard deviations inferred from GWAS summary statistics with the ones computed from a reference panel \cite[]{prive2020ldpred2}.
+Previously, to make sure that the input parameters used in LDpred2 were consistent with its modeling assumptions, we proposed a quality control (QC) based on comparing standard deviations (of genotypes) inferred from GWAS summary statistics with the ones computed from a reference panel \cite[]{prive2020ldpred2}.
 This was particularly important for LDpred2-auto, which directly estimates key model parameters from the data.
@@
 First, based on additional summary information such as the imputation INFO scores and allele frequencies from the GWAS summary statistics, we refine our previously proposed QC \cite{prive2020ldpred2}.
-For instance, we show that standard deviations of imputed genotypes (allele dosages) are lower than the expected values under Hardy-Weinberg equilibrium.
+For example, we show that standard deviations of imputed genotypes (allele dosages) are lower than the expected values under Hardy-Weinberg equilibrium.
 Second, we investigate possible corrections to apply to the input parameters of polygenic methods, namely the reference LD (linkage disequilibrium) matrix, the GWAS effect sizes, their standard errors and the corresponding sample sizes.
@@
 First, multiple studies have shown that LDpred2 and lassosum are consistently ranking best among methods for polygenic prediction \cite[]{mak2017polygenic,prive2020ldpred2,pain2021evaluation,kulm2021systematic}.
-Second, we reimplement and use a new version of lassosum, called lassosum2, that uses the exact same input parameters as LDpred2, which makes it easy for us to test the QCs and corrections presented here.
+Second, we reimplement and use a new version of lassosum, called lassosum2, that uses the exact same input parameters as LDpred2, which makes it easy for us to test the different QCs and corrections presented here.
 We perform our investigations using simulations first, then we use public summary statistics while restricting to the widely-used HapMap3 variants, finally we investigate two alternative sets with more variants.
@@
 	\caption{Overview of possible misspecifications when using GWAS summary statistics. 
-	$^{(1)}$ Examples: using the global number of cases and controls in a meta-analysis of binary traits, an often larger effective sample size when using BOLT-LMM summary statistics \cite[]{loh2018mixed}, and a reduced effective sample size when using SAIGE on binary traits with low case-control imbalance \cite[]{zhou2018efficiently,mbatchou2021computationally}.
+	$^{(1)}$ Examples: using the global number of cases and controls in a meta-analysis of binary traits, a larger effective sample size when using BOLT-LMM summary statistics \cite[]{loh2018mixed}, and a reduced effective sample size when using SAIGE on binary traits with low case-control imbalance \cite[]{zhou2018efficiently,mbatchou2021computationally}.
 	$^{(2)}$ E.g.\ removing SNPs with an effective sample size less than 0.67 times the 90th percentile of sample size \cite[]{zheng2017ld}.
@@
 
-We design simulations where variants have different GWAS sample sizes, which is often the case when meta-analyzing GWAS from multiple cohorts without the same genome coverage.
+We design simulations where variants have different GWAS sample sizes, which is often the case when meta-analyzing GWAS from multiple cohorts without the same genome coverage \cite[]{wang2021global}.
 Using 40,000 variants from chromosome 22 (Methods), we simulate quantitative phenotypes with a heritability of 20\% and 2000 causal variants.
-We then divide the 40,000 variants into three groups: for half of the variants, we use 100\% of 300,000 individuals for GWAS, but use only 80\% for one quarter of variants and 60\% for the remaining quarter.
+We then divide the 40,000 variants into three groups: for half of the variants, we use 100\% of 300,000 individuals for GWAS, but use only 80\% for one quarter of the variants and 60\% for the remaining quarter.
 We then run C+T, lassosum, lassosum2 (Methods), LDpred2-inf, LDpred2(-grid), and LDpred2-auto \cite[]{prive2019making,mak2017polygenic,prive2020ldpred2} by using either the true per-variant GWAS sample sizes, the total sample size, or imputed sample sizes.
@@
 Averaged over 10 simulations, when providing true per-variant GWAS sample sizes, squared correlations between the polygenic scores and the simulated phenotypes are of 0.123 for C+T, 0.161 for lassosum, 0.169 for lassosum2, 0.159 for LDpred2(-grid), 0.140 for LDpred2-auto, and 0.141 for LDpred2-inf (Figure \ref{fig:simu-misN}).
-Results when using imputed (instead of true) sample sizes are quite similar.
+Results when using imputed sample sizes are quite similar to when using the true ones.
 Note that C+T does not use this sample size information.
@@
 The LD reference used by LDpred2 and lassosum2 is computed from the validation set using the dataset with the ``true'' genotypes.
-For lassosum2 and LDpred2(-grid) which tune parameters using the validation set, correcting for imputation quality slightly improves predictive performance in these simulations (Figure \ref{fig:simu-info}).
+For lassosum2 and LDpred2(-grid), which tune parameters using the validation set, correcting for imputation quality slightly improves predictive performance in these simulations (Figure \ref{fig:simu-info}).
 However, correcting for imputation quality can dramatically improve predictive performance for LDpred2-auto. 
@@
 In the real data applications hereinafter, we choose to use the first correction, ``sqrt\_info'', which is simple because it is equivalent to post-processing PGS effects by multiplying them by $\sqrt{\text{INFO}}$.
+[TODO: TRY THEM ALL AGAIN?
+AND META-ANALYZE (BRCA AND T1D)?]
 
@@
 
-Breast cancer summary statistics are interesting because they include results from two mega analyses \cite[]{michailidou2013large,michailidou2015genome,michailidou2017association}, which means there is some larger precision in the parameters reported, such as the INFO scores and the sample sizes.
+Breast cancer summary statistics are interesting because they include results from two mega analyses \cite[]{michailidou2013large,michailidou2015genome,michailidou2017association}, which means there is some larger precision in the parameters reported in the GWAS summary statistics, such as the INFO scores and the sample sizes.
 Imputation INFO scores for the OncoArray summary statistics (after having restricted to HapMap3 variants) are generally very good (Figure \ref{fig:hist-info-brca}) and better than the ones from iCOGS (Figure \ref{fig:hist-info-brca2}), probably because the chip used included around 200K variants only, compared to more than 500K variants for the OncoArray.
-For both summary statistics, we first compare the standard deviations inferred from the reported allele frequencies (i.e.\ $\sqrt{2 f (1 - f)}$ where $f$ is the allele frequency, and denoted as sd\_af) versus the ones inferred from the GWAS summary statistics (Equation \eqref{eq:approx-sd-log}, and denoted as sd\_ss).
+For both summary statistics, we first compare the standard deviations (of genotypes) inferred from the reported allele frequencies (i.e.\ $\sqrt{2 f (1 - f)}$ where $f$ is the allele frequency, and denoted as sd\_af) versus the ones inferred from the GWAS summary statistics (Equation \eqref{eq:approx-sd-log}, and denoted as sd\_ss).
 When coloring by INFO scores, we see a clear trend with sd\_ss being lower than sd\_af as INFO decreases; indeed, using $\text{sd\_ss} / \sqrt{\text{INFO}}$ provides a very good fit for sd\_af, except for some variants of chromosome 6 and 8 for the OncoArray summary statistics (Figures \ref{fig:brca_onco_qc} and \ref{fig:brca_icogs_qc}).
 Most of these outlier variants are in regions 25-33 Mbp of chromosome 6 and 8-12 Mbp of chromosome 8 (Figure \ref{fig:hist-bad-pos}), which are two known long-range LD regions \cite[]{price2008long}.
-We hypothesize that this is due to using principal components (PCs) as covariates in GWAS that capture LD structure instead of population structure \cite[]{prive2020efficient}.
+We hypothesize that this is due to using (as covariates in the GWAS) principal components (PCs)  that capture LD structure instead of population structure \cite[]{prive2020efficient}.
 To validate this hypothesis, we simulate a phenotype using HapMap3 variants of chromosome 6 for 10,000 individuals from the UK Biobank, then we run GWAS with or without PC19 as covariate.
@@
 First, it allows for correcting for a reduced standard deviation when using imputed data in the QC step we propose, in order to better uncover problems with the summary statistics.
-Second, using one of the proposed corrections may lead to an improved prediction when deriving polygenic scores.
+Second, using one of the proposed corrections based on the INFO scores may lead to an improved prediction when deriving polygenic scores.
 We apply this correction to the two breast cancer summary statistics.
@@
 We have investigated misspecifications in GWAS summary statistics, focusing on the impact of sample size heterogeneity and imputation quality.
-Previously, we proposed a quality control (QC) based on comparing standard deviations inferred from GWAS summary statistics with the ones computed from a reference panel \cite[]{prive2020ldpred2}.
-Here we show that we can refine this quality control by deriving the latter directly from the reported allele frequencies in the GWAS summary statistics, and by correcting the former using the GWAS imputation INFO scores (e.g.\ see figure \ref{fig:brca_onco_qc}).
+Previously, we proposed a quality control (QC) based on comparing standard deviations (of genotypes) inferred from GWAS summary statistics with the ones computed from a reference panel \cite[]{prive2020ldpred2}.
+Here we show that we can refine this QC by deriving the latter directly from the reported allele frequencies in the GWAS summary statistics, and by correcting the former using the GWAS imputation INFO scores (e.g.\ see figure \ref{fig:brca_onco_qc}).
 Using this refined QC, we are able to identify a potential issue with how principal components were derived in a set of breast cancer summary statistics.
@@
 We do perform this additional QC as part of ``qc2'' here, and also when designing the large set of variants from the UK Biobank (Methods). 
-One can also run other QC tools such as DENTIST \cite[]{chen2021improved}.
+One can also run other QC tools such as DENTIST \cite[]{chen2021improved}, and also infer ancestry proportions from summary statistics to make sure these are matching with the LD reference used \cite[]{prive2021using}.
 
-Note that, in this study, we use summary statistics that include extended information (e.g.\ INFO scores and allele frequencies), yet most GWAS summary statistics do not \cite[]{macarthur2021workshop}.
+Note that, in this study, we use summary statistics that include extended information (e.g.\ INFO scores and allele frequencies), yet most GWAS summary statistics do not provide such exhaustive information \cite[]{macarthur2021workshop}.
 We acknowledge that, in the case of a meta-analysis from multiple studies, providing a single INFO score per variant may not be possible; would a weighted averaged INFO score work?
@@
 This results in 362,307 individuals.
-We sample 300,000 individuals to form a training set (e.g.\ to run GWAS), 10,000 individuals to form a validation set (to tune hyper-parameters), and use the remaining 52,307 individuals to form a test set (to evaluate final predictive models).
+We sample 300,000 individuals to form a training set (e.g.\ to run the GWAS), 10,000 individuals to form a validation set (to tune hyper-parameters), and use the remaining 52,307 individuals to form a test set (to evaluate final predictive models).
 
@@
 \end{equation}
-In practice, we also bound this estimate to be between $0.5 \cdot N$ and $1.1 \cdot N$, where $N$ is the global sample size.
+In practice, we also bound this estimate to be between $0.5 \cdot N$ and $1.1 \cdot N$, where $N$ is the global (effective) sample size.
 
@@
 This is similar to parameter `s' in lassosum (Section \ref{lassosum2}), and act as a means of regularization.
-We use a value of 0.9 in simulations and 0.95 in real data when running ``LDpred2-auto-rob'' (and the default value of 1, without any effect, when running ``LDpred2-auto'').
+We use a value of 0.9 in simulations and 0.95 in real data when running ``LDpred2-auto-rob'' (and the default value of 1, with no effect, when running ``LDpred2-auto'').
 The second new parameter, \texttt{allow\_jump\_sign}, controls whether variants can change sign over two consecutive iterations of the Gibbs sampler.
@@
 
-F.P.\ and B.J.V.\ are supported by the Danish National Research Foundation (Niels Bohr Professorship to Prof. John McGrath).
-B.J.V.\ is also supported by a Lundbeck Foundation Fellowship (R335-2019-2339).
+F.P.\ and B.J.V.\ are supported by the Danish National Research Foundation (Niels Bohr Professorship to Prof. John McGrath) and by a Lundbeck Foundation Fellowship (R335-2019-2339 to B.J.V.).
 
@@
 The authors declare no competing interests.
+[TODO: ADD ALLELICA FOR BJARNI?]
 

### 2022-02-02 keep4later/new-sets.tex
rm the two additional sets

--- 2022-01-25 paper/paper-misspec.tex
@@
-%% LyX 1.3 created this file.  For more info, see http://www.lyx.org/.
-%% Do not edit unless you really know what you are doing.
-\documentclass[english, 12pt]{article}
-\usepackage{times}
-%\usepackage{algorithm2e}
-\usepackage{xurl}
-\usepackage{bbm}
-\usepackage[T1]{fontenc}
-\usepackage[latin1]{inputenc}
-\usepackage{geometry}
-\geometry{verbose,letterpaper,tmargin=2cm,bmargin=2cm,lmargin=1.5cm,rmargin=1.5cm}
-\usepackage{rotating}
-\usepackage{color}
-\usepackage{graphicx}
-\usepackage{amsmath, amsthm, amssymb}
-\usepackage{setspace}
-\usepackage{lineno}
-\usepackage{hyperref}
-\usepackage{bbm}
-\usepackage{makecell}
-
-\renewcommand{\arraystretch}{1.3}
-
-\usepackage{xr}
-\externaldocument{paper-misspec-supp}
-
-%\linenumbers
-%\doublespacing
-\onehalfspacing
-%\usepackage[authoryear]{natbib}
-\usepackage{natbib} %\bibpunct{(}{)}{;}{author-year}{}{,}
-
-%Pour les rajouts
-\usepackage{color}
-\definecolor{trustcolor}{rgb}{0,0,1}
-
-\usepackage{dsfont}
-\usepackage[warn]{textcomp}
-\usepackage{adjustbox}
-\usepackage{multirow}
-\usepackage{graphicx}
-\graphicspath{{../figures/}}
-\DeclareMathOperator*{\argmin}{\arg\!\min}
-\usepackage{algorithm}
-\usepackage{algpseudocode}
-
-\let\tabbeg\tabular
-\let\tabend\endtabular
-\renewenvironment{tabular}{\begin{adjustbox}{max width=0.95\textwidth}\tabbeg}{\tabend\end{adjustbox}}
-
-\makeatletter
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% LyX specific LaTeX commands.
-%% Bold symbol macro for standard LaTeX users
-%\newcommand{\boldsymbol}[1]{\mbox{\boldmath $#1$}}
-
-%% Because html converters don't know tabularnewline
-\providecommand{\tabularnewline}{\\}
-
-\usepackage{babel}
-\makeatother
-
-\begin{document}
-
-\title{Identifying and correcting multiple sources of misspecification\\in GWAS summary statistics for polygenic scores}
-\author{Florian Priv\'e,$^{\text{1,}*}$ Julyan Arbel,$^{\text{2}}$ Hugues Aschard,$^{\text{3,4}}$ and Bjarni J. Vilhj\'almsson$^{\text{1,5}}$}
-
-\date{~ }
-\maketitle
-
-\noindent$^{\text{\sf 1}}$National Centre for Register-Based Research, Aarhus University, Aarhus, 8210, Denmark. \\
-\noindent$^{\text{\sf 2}}$Univ.\ Grenoble Alpes, Inria, CNRS, Grenoble INP, LJK, Grenoble, 38000, France. \\
-\noindent$^{\text{\sf 3}}$Department of Computational Biology, Institut Pasteur, Paris, 75015, France. \\
-\noindent$^{\text{\sf 4}}$Program in Genetic Epidemiology and Statistical Genetics, Harvard T.H. Chan School of Public Health, Boston, MA, 02115, USA. \\
-\noindent$^{\text{\sf 5}}$Bioinformatics Research Centre, Aarhus University, Aarhus, 8000, Denmark. \\
-\noindent$^\ast$To whom correspondence should be addressed.\\
-
-\noindent Contact: \url{florian.prive.21@gmail.com}
-
-\vspace*{8em}
-
-\abstract{
-	Are genome-wide association studies (GWAS) summary statistics of good enough quality for performing follow-up analyses?
-Can we detect possible misspecifications in GWAS summary statistics and correct for them in order to improve predictive performance of polygenic scores?
-}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\clearpage
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Introduction}
-
-Contrary to individual-level genotypes and phenotypes, summary statistics resulting from genome-wide association studies (GWAS) are widely available.
-Therefore, they have been extensively used e.g.\ to derive polygenic scores (PGS).
-However, GWAS summary statistics come with uneven qualities, such as the imputation qualities of each variant reported. 
-There is also some heterogeneity in the methods used for performing the individual GWAS and their meta-analyses, as well as the information reported in the resulting summary statistics.
-Moreover, many PGS methods such as PRS-CS, SBayesR, and LDpred2 \cite[]{ge2019polygenic,lloyd2019improved,prive2020ldpred2} use Bayesian models, which can be sensitive to model misspecifications \cite[]{WALKER20131621,miller2018robust}.
-Previously, to make sure that input parameters passed to LDpred2 were consistent with its modeling assumptions, we proposed a quality control (QC) based on comparing standard deviations inferred from GWAS summary statistics with the ones computed from a reference panel \cite[]{prive2020ldpred2}.
-This was particularly important for LDpred2-auto, which directly estimates key model parameters from the data.
-
-Here, we present an overview of the possible misspecifications that come with GWAS summary statistics (Table \ref{tab:overview}).
-Then, we investigate some of those, and propose corrections to improve the predictive performance of polygenic scores derived from GWAS summary statistics.
-We approach this from three different angles.
-First, based on additional summary information such as the imputation INFO scores and allele frequencies from the GWAS summary statistics, we refine our previously proposed QC \cite{prive2020ldpred2}.
-For instance, we show that standard deviations of imputed genotypes (allele dosages) are lower than the expected values under Hardy-Weinberg equilibrium.
-Second, we investigate possible corrections to apply to the input parameters of polygenic methods, namely the reference LD (linkage disequilibrium) matrix, the GWAS effect sizes, their standard errors and the corresponding sample sizes.
-For example, we show that GWAS effect sizes computed from imputed dosages are larger in magnitude compared to if true genotypes were available.
-Third, we introduce two new optional parameters in LDpred2-auto to make it more robust to these types of misspecification.
-We focus our investigations on LDpred2 and lassosum for two reasons.
-First, multiple studies have shown that LDpred2 and lassosum are consistently ranking best among methods for polygenic prediction \cite[]{mak2017polygenic,prive2020ldpred2,pain2021evaluation,kulm2021systematic}.
-Second, we reimplement and use a new version of lassosum, called lassosum2, that uses the exact same input parameters as LDpred2, which makes it easy for us to test the QCs and corrections presented here.
-We perform our investigations using simulations first, then we use public summary statistics while restricting to the widely-used HapMap3 variants, finally we investigate two alternative sets with more variants.
-
-\begin{table}[h]
-	\centering
-	\begin{tabular}{|p{\dimexpr0.2\textwidth-2\tabcolsep-\arrayrulewidth\relax}|
-			p{\dimexpr0.5\textwidth-2\tabcolsep-\arrayrulewidth\relax}|
-			p{\dimexpr0.3\textwidth-2\tabcolsep-\arrayrulewidth\relax}|}
-		\hline
-		Misspecification & Possible consequence(s) & Possible correction \\
-		\hline
-		\hline
-		Global sample size misspecification$^{(1)}$ & Bias estimation of SNP heritability \cite[]{gazal2018functional,grotzinger2021pervasive} & Estimation of the effective sample size \\
-		\hline
-		Differences in per-variant sample size & Break some model assumptions & Imputation using equations \ref{eq:imputeN} and \ref{eq:imputeNeff}, and filtering$^{(2)}$   \\
-		\hline
-		Using imputed dosages for GWAS & Standard deviations of dosages are smaller, resulting in overestimated effect sizes and standard errors & See section \ref{sec:imputation} \\
-		\hline
-		Using imputed dosages for LD computation & LD seems correctly estimated (Figure \ref{fig:compare-cor}) & None required \\
-		\hline 
-		Ancestry mismatch between summary statistics and LD & Strong mismatch of LD and allele frequencies, often resulting in divergence of models & Check ancestry proportions from GWAS summary statistics \cite[]{prive2021using} \\
-		\hline
-		Error in summary statistics (e.g.\ allele inversions) & Strong violation of model assumptions, which can result in e.g.\ identifying false-positives in fine-mapping \cite[]{susie} & Quality control \\
-		\hline
-		Any & Violation of model assumptions resulting in poor model convergence & Using more regularization \\
-		\hline
-	\end{tabular}
-	\caption{Overview of possible misspecifications when using GWAS summary statistics. 
-	$^{(1)}$ Examples: using the global number of cases and controls in a meta-analysis of binary traits, an often larger effective sample size when using BOLT-LMM summary statistics \cite[]{loh2018mixed}, and a reduced effective sample size when using SAIGE on binary traits with low case-control imbalance \cite[]{zhou2018efficiently,mbatchou2021computationally}.
-	$^{(2)}$ E.g.\ removing SNPs with an effective sample size less than 0.67 times the 90th percentile of sample size \cite[]{zheng2017ld}.
-	\label{tab:overview}}
-\end{table}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Results}
-
-\subsection{Misspecification of GWAS sample sizes}
-
-We design simulations where variants have different GWAS sample sizes, which is often the case when meta-analyzing GWAS from multiple cohorts without the same genome coverage.
-Using 40,000 variants from chromosome 22 (Methods), we simulate quantitative phenotypes with a heritability of 20\% and 2000 causal variants.
-We then divide the 40,000 variants into three groups: for half of the variants, we use 100\% of 300,000 individuals for GWAS, but use only 80\% for one quarter of variants and 60\% for the remaining quarter.
-We then run C+T, lassosum, lassosum2 (Methods), LDpred2-inf, LDpred2(-grid), and LDpred2-auto \cite[]{prive2019making,mak2017polygenic,prive2020ldpred2} by using either the true per-variant GWAS sample sizes, the total sample size, or imputed sample sizes.
-Note that we initially included PRS-CS and SBayesR in our comparison \cite[]{ge2019polygenic,lloyd2019improved}.
-However, results for SBayesR always diverged and the overlap with the LD reference provided for PRS-CS was too small.
-Averaged over 10 simulations, when providing true per-variant GWAS sample sizes, squared correlations between the polygenic scores and the simulated phenotypes are of 0.123 for C+T, 0.161 for lassosum, 0.169 for lassosum2, 0.159 for LDpred2(-grid), 0.140 for LDpred2-auto, and 0.141 for LDpred2-inf (Figure \ref{fig:simu-misN}).
-Results when using imputed (instead of true) sample sizes are quite similar.
-Note that C+T does not use this sample size information.
-When using the total GWAS sample size instead of the per-variant sample sizes, predictive performance slightly decreases to 0.157 for lassosum and to 0.163 for lassosum2, but dramatically decreases for LDpred2 with new values of 0.134 for LDpred2-grid, 0.119 for LDpred2-auto, and 0.123 for LDpred2-inf (Figure \ref{fig:simu-misN}).
-This extreme simulation scenario shows that LDpred2 can be sensitive to GWAS sample size misspecification, whereas lassosum (and lassosum2) seems little affected by this.
-
-\begin{figure}[h]
-	\centerline{\includegraphics[width=0.98\textwidth]{simu-misN}}
-	\caption{Results for the simulations with sample size misspecification, averaged over 10 simulations for each scenario.
-Reported 95\% confidence intervals are computed from 10,000 non-parametric bootstrap replicates of the mean.
-The GWAS sample size is ``true'' when providing the true per-variant sample size, ``max'' when providing instead the maximum sample size as a unique value to be used for all variants,  ``imputed'' (Methods), or ``any'' when the method does not use this information (the case for C+T).
-Red bars correspond to using the LD with independent blocks (Methods).}
-	\label{fig:simu-misN}
-\end{figure}
-
-We conduct further investigations to explain results of figure \ref{fig:simu-misN}.
-First, the results for LDpred2-auto are the same as with LDpred2-inf because it always converges to an infinitesimal model ($p = 1$) in these simulations.
-To solve this limitation, we introduce two new parameters to make LDpred2-auto more robust, referred to as ``LDpred2-auto-rob'' here (Methods).
-Second, for lassosum2, results for a grid of parameters (over $\lambda$ and $\delta$) are quite smooth compared to LDpred2 (Figures \ref{fig:lassosum2-misN} and \ref{fig:ldpred2-misN}). 
-In these simulations with misspecified sample sizes, it seems highly beneficial to use a small value for the SNP heritability hyper-parameter $h^2$ in LDpred2, e.g.\ a value of 0.02 or even 0.002 when the true value is 0.2 (Figure \ref{fig:ldpred2-misN}).
-Indeed, using a small value for this hyper-parameter induces a larger regularization (shrinkage) on the effect sizes.
-When running LDpred2(-grid) with a grid of hyper-parameters including these low values for $h^2$, we refer to this as ``LDpred2-low-h2'' here.
-Results with LDpred2-low-h2 improves from 0.159 to 0.169 when using true sample sizes and from 0.134 to 0.163 when using the maximum sample size.
-Finally, we introduce a last change for robustness here: we form independent LD blocks in the LD matrix to prevent small errors in the Gibbs sampler to propagate to too many variants (Methods).
-This change seems to solve convergence issues of LDpred2 in these simulations (Figure \ref{fig:ldpred2-misN}) and further improves predictive performance for all LDpred2 methods (Figure \ref{fig:simu-misN}).
-
-\subsection{When using allele dosages from imputation}\label{sec:imputation}
-
-\cite{marchini2010genotype} showed that the IMPUTE INFO measure is highly concordant with the MACH measure $\hat{r}_j^2 = \frac{\text{var}(G_j)}{2 \hat{\theta}_j (1 - \hat{\theta}_j)}$, where $\hat{\theta}_j$ is the estimated allele frequency of $G_j$, the genotypes for variant $j$.
-Therefore, when using the expected genotypes from imputation (dosages), their standard deviations are often lower than the expected value under Hardy-Weinberg equilibrium, because $\text{INFO}_j \approx \frac{\text{var}(G_j)}{2 \hat{\theta}_j (1 - \hat{\theta}_j)}$.
-In simulations (cf.\ Methods section ``Data for simulations''), we verify that $\text{sd}(G_j)^{\text{true}} \approx \text{sd}(G_j)^{\text{imp}} / \sqrt{\text{INFO}_j}$ (Figure \ref{fig:compare-sd}).
-We also show that GWAS effect sizes $\hat{\gamma}$ computed from imputed dosages are over-estimated: $\hat{\gamma}_j^{\text{true}} \approx \hat{\gamma}_j^{\text{imp}} \cdot \sqrt{\text{INFO}_j}$ and $\text{se}(\hat{\gamma}_j)^{\text{true}} \approx \text{se}(\hat{\gamma}_j)^{\text{imp}} \cdot \sqrt{\text{INFO}_j}$ (Figures \ref{fig:compare-beta} and \ref{fig:compare-beta-se}).
-This is the first correction of summary statistics we consider in the simulations below. 
-Note that we recompute INFO scores for the subset of 362,307 European individuals used in this paper as they can differ substantially from the ones reported by the UK Biobank for the whole data (Figure \ref{fig:compare-info}).
-As a second option, instead of using dosages to compute the GWAS summary statistics, it has been argued that using multiple imputation (MI) would be more appropriate \cite[]{Palmer2016}.
-In simulations, we show that $\hat{\gamma}_j^{\text{MI}} \approx \hat{\gamma}_j^{\text{imp}} \cdot \text{INFO}_j$ and $Z_j^{\text{MI}} \approx Z_j^{\text{imp}} \cdot \text{INFO}_j$, where $Z = \hat{\gamma} / \text{se}(\hat{\gamma})$ (Figure \ref{fig:compare-mi}).
-This is the second correction of summary statistics we implement in the simulations below, along with $n_j = N \cdot \text{INFO}_j$. 
-Finally, we consider an in-between solution as a third correction, using $\hat{\gamma}_j = \hat{\gamma}_j^{\text{imp}} \cdot \text{INFO}_j$, $\text{se}(\hat{\gamma}_j) = \text{se}(\hat{\gamma}_j)^{\text{imp}} \cdot \sqrt{\text{INFO}_j}$, and $n_j = N \cdot \text{INFO}_j$.
-
-Using 40,000 variants from chromosome 22, we simulate quantitative phenotypes assuming a heritability of 20\% and 2000 causal variants using the ``true'' dataset (cf.\ Methods section ``Data for simulations'').
-We compute GWAS summary statistics from the dosage dataset and use these summary statistics to run LDpred2 and lassosum2 with either no correction of the summary statistics, or with one of the three corrections described above.
-The LD reference used by LDpred2 and lassosum2 is computed from the validation set using the dataset with the ``true'' genotypes.
-For lassosum2 and LDpred2(-grid) which tune parameters using the validation set, correcting for imputation quality slightly improves predictive performance in these simulations (Figure \ref{fig:simu-info}).
-However, correcting for imputation quality can dramatically improve predictive performance for LDpred2-auto. 
-Moreover, new additions for robustness introduced before, namely LDpred2-low-h2, LDpred2-auto-rob, and forming independent blocks in the LD matrix also improve predictive performance for all corrections (Figure \ref{fig:simu-info}).
-
-In the real data applications hereinafter, we choose to use the first correction, ``sqrt\_info'', which is simple because it is equivalent to post-processing PGS effects by multiplying them by $\sqrt{\text{INFO}}$.
-
-\begin{figure}[h]
-	\centerline{\includegraphics[width=0.98\textwidth]{simu-info}}
-	\caption{Results of predictive performance for the simulations using GWAS summary statistics from imputed dosage data, averaged over 10 simulations for each scenario.
-Reported 95\% confidence intervals are computed from 10,000 non-parametric bootstrap replicates of the mean.
-Correction ``sqrt\_info'' corresponds to using $\hat{\gamma}_j^{\text{imp}} \cdot \sqrt{\text{INFO}_j}$ and $\text{se}(\hat{\gamma}_j)^{\text{imp}} \cdot \sqrt{\text{INFO}_j}$.
-	Correction ``info'' corresponds to using $\hat{\gamma}_j^{\text{imp}} \cdot \text{INFO}_j$  and $N \cdot \text{INFO}_j$.
-Correction ``in\_between'' corresponds to using $\hat{\gamma}_j^{\text{imp}} \cdot \text{INFO}_j$, $\text{se}(\hat{\gamma}_j)^{\text{imp}} \cdot \sqrt{\text{INFO}_j}$, and $N \cdot \text{INFO}_j$.
-Red bars correspond to using the LD with independent blocks (Methods).}
-	\label{fig:simu-info}
-\end{figure}
-
-\subsection{Application to breast cancer summary statistics}
-
-Breast cancer summary statistics are interesting because they include results from two mega analyses \cite[]{michailidou2013large,michailidou2015genome,michailidou2017association}, which means there is some larger precision in the parameters reported, such as the INFO scores and the sample sizes.
-Imputation INFO scores for the OncoArray summary statistics (after having restricted to HapMap3 variants) are generally very good (Figure \ref{fig:hist-info-brca}) and better than the ones from iCOGS (Figure \ref{fig:hist-info-brca2}), probably because the chip used included around 200K variants only, compared to more than 500K variants for the OncoArray.
-For both summary statistics, we first compare the standard deviations inferred from the reported allele frequencies (i.e.\ $\sqrt{2 f (1 - f)}$ where $f$ is the allele frequency, and denoted as sd\_af) versus the ones inferred from the GWAS summary statistics (Equation \eqref{eq:approx-sd-log}, and denoted as sd\_ss).
-When coloring by INFO scores, we see a clear trend with sd\_ss being lower than sd\_af as INFO decreases; indeed, using $\text{sd\_ss} / \sqrt{\text{INFO}}$ provides a very good fit for sd\_af, except for some variants of chromosome 6 and 8 for the OncoArray summary statistics (Figures \ref{fig:brca_onco_qc} and \ref{fig:brca_icogs_qc}).
-Most of these outlier variants are in regions 25-33 Mbp of chromosome 6 and 8-12 Mbp of chromosome 8 (Figure \ref{fig:hist-bad-pos}), which are two known long-range LD regions \cite[]{price2008long}.
-We hypothesize that this is due to using principal components (PCs) as covariates in GWAS that capture LD structure instead of population structure \cite[]{prive2020efficient}.
-To validate this hypothesis, we simulate a phenotype using HapMap3 variants of chromosome 6 for 10,000 individuals from the UK Biobank, then we run GWAS with or without PC19 as covariate.
-PC19 from the UK Biobank was previously reported to capture LD structure in region 70-91 Mbp of chromsome 6 \cite[]{prive2020efficient}.
-In these simulations, the same bias as in figure \ref{fig:brca_onco_qc}B is observed for the variants in this region (Figure \ref{fig:gwas_bad_pc}), confirming our hypothesis.
-
-\begin{figure}[h]
-	\centerline{\includegraphics[width=0.98\textwidth]{brca_onco_qc}}
-	\caption{Standard deviations inferred from the OncoArray breast cancer GWAS summary statistics (Equation \ref{eq:approx-sd-log}) versus the ones inferred from the reported GWAS allele frequencies ($\sqrt{2 f (1 - f)}$).
-Only 100,000 HapMap3 variants are represented, at random.}
-	\label{fig:brca_onco_qc}
-\end{figure}
-
-Therefore, providing an accurate imputation INFO score is important for two reasons.
-First, it allows for correcting for a reduced standard deviation when using imputed data in the QC step we propose, in order to better uncover problems with the summary statistics.
-Second, using one of the proposed corrections may lead to an improved prediction when deriving polygenic scores.
-We apply this correction to the two breast cancer summary statistics.
-We first compare the standard QC proposed in \cite{prive2020ldpred2} (which ends up filtering on MAF here, which we call ``qc1'').
-We then also filter out the two long-range LD regions of chromosome 6 and 8 for the OncoArray summary statistics and remove around 500 variants when filtering on differences of MAFs between summary statistics and the validation dataset (``qc2'').
-As for the correction for the INFO scores, we use the first correction, ``sqrt\_info'', which is simple because it is equivalent to post-processing PGS effects by multiplying them by $\sqrt{\text{INFO}}$.
-Although results for both QC used are very similar, correcting for the INFO score slightly improves prediction when deriving polygenic scores based on iCOGS summary statistics (Figure \ref{fig:res_brca}).
-All other improvements introduced before have little or no effect here, probably because misspecifications are much smaller than in the simulations.
-
-\begin{figure}[h]
-	\centerline{\includegraphics[width=0.98\textwidth]{res-brca}}
-	\caption{Raw partial correlations (for \textit{all} models) for predicting breast cancer in the UK Biobank when using either the OncoArray or the iCOGS summary statistics.
-These are computed using function \texttt{pcor} of R package bigstatsr where 95\% confidence intervals are obtained through Fisher's Z-transformation.}
-	\label{fig:res_brca}
-\end{figure}
-
-\subsection{Other phenotypes and larger sets of variants}
-
-We use external summary statistics for which INFO scores are reported (Table \ref{tab:sumstats}); they all have a very high mean INFO score (larger than 0.94), except for BRCA-iCOGS (0.841) and T1D-affy (0.885).
-QC plots comparing standard deviations usually show little deviation from $x=y$ (after the INFO score correction), except for coronary artery disease (CAD) summary statistics (Figures \ref{fig:brca_icogs_qc}-\ref{fig:t1d_illu_qc}).
-We show previous results as figure \ref{fig:res_brca} for other phenotypes in figures \ref{fig:res_cad}-\ref{fig:res_t1d}; most changes introduced before have little or no impact on the predictions, except for the second quality control performed for CAD when using LDpred2-auto (Figure \ref{fig:res_cad}).
-We then introduce two new sets of variants as possible replacement for HapMap3 variants (Methods).
-These sets include more variants (more than 2M), therefore possibly of lower quality on average. 
-We present similar results with these two new sets of variants in figures \ref{fig:res_brca_all}-\ref{fig:res_t1d_all}.
-Using the ``maxtag'' set generally provides larger predictive performance than using HapMap3 variants, especially when predicting prostate cancer (Figure \ref{fig:res_prca_all}).
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Discussion}
-
-We have investigated misspecifications in GWAS summary statistics, focusing on the impact of sample size heterogeneity and imputation quality.
-Previously, we proposed a quality control (QC) based on comparing standard deviations inferred from GWAS summary statistics with the ones computed from a reference panel \cite[]{prive2020ldpred2}.
-Here we show that we can refine this quality control by deriving the latter directly from the reported allele frequencies in the GWAS summary statistics, and by correcting the former using the GWAS imputation INFO scores (e.g.\ see figure \ref{fig:brca_onco_qc}).
-Using this refined QC, we are able to identify a potential issue with how principal components were derived in a set of breast cancer summary statistics.
-Fortunately, this has practically no effect on the predictive performance of the derived polygenic scores.
-Additional QC can also be performed, e.g.\ comparing reported GWAS allele frequencies with the ones from the LD reference panel, e.g.\ to detect genotyping or allelic errors.
-We do perform this additional QC as part of ``qc2'' here, and also when designing the large set of variants from the UK Biobank (Methods). 
-One can also run other QC tools such as DENTIST \cite[]{chen2021improved}.
-
-Note that, in this study, we use summary statistics that include extended information (e.g.\ INFO scores and allele frequencies), yet most GWAS summary statistics do not \cite[]{macarthur2021workshop}.
-We acknowledge that, in the case of a meta-analysis from multiple studies, providing a single INFO score per variant may not be possible; would a weighted averaged INFO score work?
-Nevertheless, this quality control could be performed within each study before pooling results, to make sure that summary statistics have the best possible quality for follow-up analyses such as deriving polygenic scores.
-Another information, the effective sample size per variant, is often missing from GWAS summary statistics. 
-Sometimes, it can even be challenging to recover the total effective sample size from large meta-analyses.
-We recall that if some studies of a meta-analysis have an imbalanced number of cases and controls, the global effective sample size should not be computed from the total numbers of cases and controls overall, but instead from the sum of the effective sample sizes of each study.
-Indeed, take the extreme example of meta-analyzing two studies, one with 1000 cases and 0 controls, and another one with 0 cases and 1000 controls, then the effective sample size of the meta-analysis is 0, not 2000.
-This misspecification can lead to serious issues such as misestimating the SNP heritability \cite[]{grotzinger2021pervasive}.
-Fortunately, an overestimated sample size can be detected from the QC plot we propose, where the slope is then less than 1 for case-control studies using logistic regression; otherwise the standard deviation of the phenotype is also needed (Equation \ref{eq:approx-sd-lin}), but can be estimated \cite[]{prive2020ldpred2}.
-
-We have assessed the impact of these misspecifications in GWAS summary statistics on the predictive performance of some polygenic score methods.
-Using both the Bayesian LDpred2 models \cite[]{prive2020ldpred2} and our reimplementation of the frequentist lassosum model \cite[]{mak2017polygenic} for deriving polygenic scores, we have introduced and investigated some changes to possibly make these models more robust to misspecifications.
-Overall, these changes have provided large improvements of predictive performance in simulations with large misspecifications.
-However, they have almost no effect when using the real GWAS summary statistics we chose, except for breast cancer and coronary artery disease summary statistics.
-Although these results are somewhat unfortunate, they are reassuring because it means that the summary statistics we use in this study are usually of good quality for follow-up analyses such as deriving polygenic scores.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Materials and Methods}
-
-\subsection{Data for simulations}
-
-We use the UK Biobank imputed (BGEN) data \cite[]{bycroft2018uk}. 
-We restrict individuals to the ones used for computing the principal components (PCs) in the UK Biobank (Field 22020).
-These individuals are unrelated and have passed some quality control including removing samples with a missing rate on autosomes larger than 0.02, having a mismatch between inferred sex and self-reported sex, and outliers based on heterozygosity (more details can be found in section S3 of \cite{bycroft2018uk}).
-To get a set of genetically homogeneous individuals, we compute a robust Mahalanobis distance based on the first 16 PCs and further restrict individuals to those within a log-distance of 5 \cite[]{prive2020efficient}.
-This results in 362,307 individuals.
-We sample 300,000 individuals to form a training set (e.g.\ to run GWAS), 10,000 individuals to form a validation set (to tune hyper-parameters), and use the remaining 52,307 individuals to form a test set (to evaluate final predictive models).
-
-Among genetic variants on chromosome 22 and with a minor allele frequency larger than 0.01 and an imputation INFO score larger than 0.4 (as reported by the UK Biobank), we sample 40,000 of them according to the inverse of the INFO score density so that they have varying levels of imputation accuracy (Figure \ref{fig:hist-info}).
-We read the UK Biobank data into two different datasets using function \texttt{snp\_readBGEN} from R package bigsnpr \cite[]{prive2017efficient}, one by reading the BGEN data at random according to imputation probabilities, and another one reading it as dosages (i.e.\ expected values according to imputation probabilities).
-The first dataset is used as what could be the real genotype calls and the second dataset as what would be its imputed version; this design technique was used in \cite{prive2019making}.
-
-\subsection{Data for real analyses}
-
-We also use the UK Biobank data, and use the same individuals as described in the previous section.
-We sample 10,000 individuals to form a validation set and use the remaining 352,307 individuals as test set.
-We restrict to the genetic variants to the 1,054,315 HapMap3 variants used in the LD reference provided in \cite{prive2020ldpred2}.
-We also try two new sets of variants (see next section).
-
-To define phenotypes in the UK Biobank, we first map ICD10 and ICD9 codes (UKBB fields 40001, 40002, 40006, 40013, 41202, 41270 and 41271) to phecodes using R package PheWAS \cite[]{carroll2014r,wu2019mapping}.
-We use published GWAS summary statistics listed in table \ref{tab:sumstats} to derive polygenic scores.
-
-\begin{table}[h]
-	\centering
-	\begin{tabular}{|l||c|c|c||c|c|}
-		\hline
-		Trait & GWAS citation & \begin{tabular}{@{}c@{}}Effective GWAS\\sample size\end{tabular} & \# GWAS variants & \begin{tabular}{@{}c@{}}\# matched variants\\with INFO > 0.4\end{tabular} & Mean INFO \\
-		\hline
-		Breast cancer (BRCA) [iCOGS] & \cite{michailidou2017association} & ~~87,037 & 11,792,542 & 1,051,242 & 0.841 \\
-		Breast cancer (BRCA) [OncoArray] & \cite{michailidou2017association} & 104,442 & 11,792,542 & 1,054,233 & 0.968 \\
-		Type 1 diabetes (T1D) [Affymetrix] & \cite{censin2017childhood} & ~~~~~5516  & ~~8,996,866 & ~~~934,712 & 0.885 \\	
-		Type 1 diabetes (T1D) [Illumina] & \cite{censin2017childhood} & ~~~~~7982  & ~~8,996,866 & ~~~949,334 & 0.942 \\
-		Prostate cancer (PRCA) & \cite{schumacher2018association} & 135,316 & 20,370,946 & ~~~818,400 & 0.969 \\
-		Depression (MDD) [without UKBB] & \cite{wray2018genome} & 110,464 & ~~9,874,289 & 1,049,455 & 0.968 \\
-		Coronary artery disease (CAD) & \cite{nikpay2015comprehensive} & 129,014 & ~~9,455,778 & 1,052,200 & 0.982 \\
-		\hline
-	\end{tabular}
-	\caption{Summary of external GWAS summary statistics used. PRCA summary statistics have many variants with a missing INFO score, which we discard.\label{tab:sumstats}}
-\end{table}
-
 \subsection{Two new sets of variants}
@@
 
-\subsection{GWAS sample size imputation}
-
-In this paper, we extensively use the following formula
-\begin{equation}\label{eq:approx-sd-lin}
-\text{sd}(G_j) \approx \dfrac{\text{sd}(y)}{\sqrt{n_j ~ \text{se}(\hat{\gamma}_j)^2 + \hat{\gamma}_j^2}} ~,
-\end{equation}
-where $\hat{\gamma}_j$ is the marginal (GWAS) effect of variant $j$, $n_j$ is the GWAS sample size associated with variant $j$, $y$ is the vector of phenotypes and ${G}_j$ is the vector of genotypes for variant $j$.
-This formula is used in LDpred2 \cite{prive2020ldpred2,prive2021high}. Note that, for a binary trait for which logistic regression is used, we have instead
-\begin{equation}\label{eq:approx-sd-log}
-\text{sd}(G_j) \approx \dfrac{2}{\sqrt{n_j^\text{eff} ~ \text{se}(\hat{\gamma}_j)^2 + \hat{\gamma}_j^2}} ~,
-\end{equation}
-where $n_j^\text{eff} = \frac{4}{1 / n_j^\text{cases} + 1 / n_j^\text{controls}}$.
-
-We can then impute $n_j$ from equation \eqref{eq:approx-sd-lin} using
-\begin{equation}\label{eq:imputeN}
-n_j \approx \dfrac{\text{var}(y) / \text{var}(G_j) - \hat{\gamma}_j^2}{\text{se}(\hat{\gamma}_j)^2} ~,
-\end{equation}
-and impute $n_j^\text{eff}$ from equation \eqref{eq:approx-sd-log} using
-\begin{equation}\label{eq:imputeNeff}
-n_j^\text{eff} \approx \dfrac{4 / \text{var}(G_j) - \hat{\gamma}_j^2}{\text{se}(\hat{\gamma}_j)^2} ~.
-\end{equation}
-In practice, we also bound this estimate to be between $0.5 \cdot N$ and $1.1 \cdot N$, where $N$ is the global sample size.
-
-\subsection{New implementation of lassosum in bigsnpr}\label{lassosum2}
-
-Instead of using a regularized version of the correlation matrix $R$ parameterized by $s$, $R_s = (1 - s) R + s I$ (where $0 < s \leq 1$), we use $R_{\delta} = R + \delta I$ (where $\delta > 0$), which makes it clearer that lassosum is also using L2-regularization (therefore elastic-net).
-Then, from \cite{mak2017polygenic}, the solution from lassosum can be obtained by iteratively updating 
-\[
-\beta_j^{(t)} =
-\begin{cases}
-\text{sign}\left(u_j^{(t)}\right) \left(\left|u_j^{(t)}\right| - \lambda\right) / \left(\widetilde{X}_j^T \widetilde{X}_j + \delta\right) & \text{if } \left|u_j^{(t)}\right| > \lambda ~, \\
-0 & \text{otherwise.}
-\end{cases}
-\]
-where 
-\[
-u_j^{(t)} = r_j - \widetilde{X}_j^T \left( \widetilde{X} \beta^{(t-1)} - \widetilde{X}_j \beta_j^{(t-1)} \right) ~.
-\]
-Following the notations from \cite{prive2020ldpred2} and denoting $\widetilde{X} = \frac{1}{\sqrt{n-1}} C_n G S^{-1}$, where $G$ is the genotype matrix, $C_n$ is the centering matrix and $S$ is the diagonal matrix of standard deviations of the columns of $G$.
-Then $\widetilde{X}_j^T \widetilde{X} = R_{j,.} = R_{.,j}^T$ and $\widetilde{X}_j^T \widetilde{X}_j = 1$.
-Moreover, using the notations from \cite{prive2020ldpred2}, $u_j^{(t)} = \widehat{\beta}_j - R_{.,j}^T \beta^{(t-1)} + \beta_j^{(t-1)}$,
-where $r_j = \widehat{\beta}_j =  \dfrac{\widehat{\gamma}_j}{\sqrt{n_j ~ \text{se}(\widehat{\gamma}_j)^2 + \widehat{\gamma}_j^2}}$ and $\widehat{\gamma}_j$ is the GWAS effect of variant $j$ and $n$ is the GWAS sample size \cite[]{mak2017polygenic,prive2021high}.
-Then the most time-consuming part is computing $R_{.,j}^T \beta^{(t-1)}$.
-To make this faster, instead of computing $R_{.,j}^T \beta^{(t-1)}$ at each iteration ($j$ and $t$), we can start with an initial vector of 0s only (for all $j$) since $\beta^{(0)} \equiv 0$, and then updating this vector when $\beta_j^{(t)} \neq \beta_j^{(t-1)}$ only.
-Note that only positions $k$ for which $R_{k,j} \neq 0$ must be updated in this vector $R_{.,j}^T \beta^{(t-1)}$.
-
-In this new implementation of the lassosum model, the input parameters are the correlation matrix $R$, the GWAS summary statistics ($\widehat{\gamma}_j$, $\text{se}(\widehat{\gamma}_j)$ and $n_j$), and the two hyper-parameters $\lambda$ and $\delta$.
-Therefore, except for the hyper-parameters, this is the exact same input as for LDpred2 \cite[]{prive2020ldpred2}.
-We try $\delta \in \{0.001, 0.005, 0.02, 0.1, 0.6, 3\}$ by default in lassosum2, instead of $s \in \{0.2, 0.5, 0.8, 1.0\}$ in lassosum.
-For $\lambda$, the default in lassosum uses a sequence of 20 values equally spaced on a log scale between 0.1 and 0.001.
-We use instead a sequence between $\lambda_0$ and $\lambda_0 / 100$ by default in lassosum2, where $\lambda_0 = \max_j \left|\widehat{\beta}_j\right|$ is the minimum $\lambda$ for which no variable enters the model because the L1-regularization is too strong.
-Note that we do not provide an ``auto'' version using pseudo-validation (as in \cite{mak2017polygenic}) as we have not found it to be very robust (Figure \ref{fig:pseudoval}).
-Also note that, as in LDpred2, we run lassosum2 genome-wide using a sparse correlation matrix which assumes that variants further away than 3 cM are not correlated, and therefore we do not require splitting the genome into independent LD blocks anymore (as done in lassosum).
-
-\subsection{New LD reference}\label{sec:LD}
-
-We define nearly independent LD blocks using the optimal algorithm developed in \cite{prive2021optimal}.
-For different numbers of blocks and maximum number of variants in each block, we use the split with the minimum cost within the ones reducing the original number of non-zero values to less than 60\% (or 70\% for chromosome 6).
-Having a correlation matrix with independent blocks prevents the small errors in the algorithm (e.g.\ the Gibbs sampler in LDpred2) from propagating to too many variants.
-It also makes running LDpred2 (and lassosum2) faster, taking about 60\% of the initial time (since only 60\% of the initial non-zero values of the correlation matrix are kept).
-
-We have also developed a new ``compact'' format for the SFBMs (sparse matrices on disk).
-Instead of using something similar to the standard ``compressed sparse column'' format which stores all $\{i,~x(i, j)\}$ for a given column $j$, we only store the first index $i_0$ and all the contiguous values $\{x(i_0, j),~x(i_0 + 1, j),~\dots\}$ up to the last non-zero value for this column $j$.
-This makes this format about twice as efficient for both LDpred2 and lassosum2.
-
-\subsection{LDpred2-low-h2 and LDpred2-auto-rob}
-
-Here we introduce the small changes made to LDpred2 (-grid and -auto) in order to make them more robust.
-First, LDpred2-low-h2 simply consists in running LDpred2-grid by testing $h^2$ within $\{0.3, 0.7, 1, 1.4\} \cdot h^2_\text{LDSC}$ (note the added $0.3$ compared to \cite{prive2020ldpred2}), where $h^2_\text{LDSC}$ is the heritability estimate from LD score regression.
-Indeed, we show in simulations here that using lower values for $h^2$ may provide higher predictive performance in the case of misspecifications (thanks to more shrinkage of the effects).
-In simulations, because of the large misspecifications, we use a larger grid over $\{0.01, 0.1, 0.3, 0.7, 1, 1.4\} \cdot h^2_\text{LDSC}$.
-
-For LDpred2-auto, we introduce two new parameters.
-The first one, \texttt{shrink\_corr}, allows for shrinking off-diagonal elements of the correlation matrix.
-This is similar to parameter `s' in lassosum (Section \ref{lassosum2}), and act as a means of regularization.
-We use a value of 0.9 in simulations and 0.95 in real data when running ``LDpred2-auto-rob'' (and the default value of 1, without any effect, when running ``LDpred2-auto'').
-The second new parameter, \texttt{allow\_jump\_sign}, controls whether variants can change sign over two consecutive iterations of the Gibbs sampler.
-When setting this parameter to false (in the method we name ``LDpred2-auto-rob'' here), this forces the effects to go through 0 before changing sign.
-This is useful to prevent instability (oscillation and ultimately divergence) of the Gibbs sampler under large misspecifications, and is also useful for accelerating convergence of chains with a large initial value for $p$, the proportion of causal variants.
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-%\clearpage
-%\vspace*{5em}
+\FloatBarrier
 
-\section*{Code availability}
+\begin{figure}[p]
+	\centerline{\includegraphics[width=0.95\textwidth]{res-all-brca}}
+	\caption{Raw partial correlations (for \textit{all} models and \textit{all} sets) for predicting BRCA in the UK Biobank from external summary statistics.
+These are computed using function \texttt{pcor} of R package bigstatsr where 95\% confidence intervals are obtained through Fisher's Z-transformation.}
+	\label{fig:res_brca_all}
+\end{figure}
 
-All code used for this paper is available at \url{https://github.com/privefl/paper-misspec/tree/master/code}.
-The latest version of R package bigsnpr can be installed from GitHub.
-Two tutorials on running LDpred2 and lassosum2 using R package bigsnpr are available at \url{https://privefl.github.io/bigsnpr/articles/LDpred2.html} and \url{https://privefl.github.io/bigsnpr-extdoc/polygenic-scores-pgs.html}.
-We have extensively used R packages bigstatsr and bigsnpr \cite[]{prive2017efficient} for analyzing large genetic data, packages from the future framework \cite[]{bengtsson2020unifying} for easy scheduling and parallelization of analyses on the HPC cluster, and packages from the tidyverse suite \cite[]{wickham2019welcome} for shaping and visualizing results.
+\begin{figure}[p]
+	\centerline{\includegraphics[width=0.9\textwidth]{res-all-cad}}
+	\caption{Raw partial correlations (for \textit{all} models and \textit{all} sets) for predicting CAD in the UK Biobank from external summary statistics.
+These are computed using function \texttt{pcor} of R package bigstatsr where 95\% confidence intervals are obtained through Fisher's Z-transformation.}
+	\label{fig:res_cad_all}
+\end{figure}
 
-\section*{Acknowledgements}
+\begin{figure}[p]
+	\centerline{\includegraphics[width=0.9\textwidth]{res-all-mdd}}
+	\caption{Raw partial correlations (for \textit{all} models and \textit{all} sets) for predicting MDD in the UK Biobank from external summary statistics.
+These are computed using function \texttt{pcor} of R package bigstatsr where 95\% confidence intervals are obtained through Fisher's Z-transformation.}
+	\label{fig:res_mdd_all}
+\end{figure}
 
-Authors thank Timothy Shin Heng Mak, Shing Wan Choi, and Matthew Stephens for helpful discussions. 
-Authors also thank GenomeDK and Aarhus University for providing computational resources and support that contributed to these research results.
-This research has been conducted using the UK Biobank Resource under Application Number 58024.
+\begin{figure}[p]
+	\centerline{\includegraphics[width=0.9\textwidth]{res-all-prca}}
+	\caption{Raw partial correlations (for \textit{all} models and \textit{all} sets) for predicting PRCA in the UK Biobank from external summary statistics.
+These are computed using function \texttt{pcor} of R package bigstatsr where 95\% confidence intervals are obtained through Fisher's Z-transformation.}
+	\label{fig:res_prca_all}
+\end{figure}
 
-\section*{Funding}
-
-F.P.\ and B.J.V.\ are supported by the Danish National Research Foundation (Niels Bohr Professorship to Prof. John McGrath).
-B.J.V.\ is also supported by a Lundbeck Foundation Fellowship (R335-2019-2339).
-
-\section*{Declaration of Interests}
-
-The authors declare no competing interests.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%\clearpage
-\vspace*{3em}
-
-\bibliographystyle{natbib}
-\bibliography{refs}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\end{document}
+\begin{figure}[p]
+	\centerline{\includegraphics[width=0.95\textwidth]{res-all-t1d}}
+	\caption{Raw partial correlations (for \textit{all} models and \textit{all} sets) for predicting T1D in the UK Biobank from external summary statistics.
+These are computed using function \texttt{pcor} of R package bigstatsr where 95\% confidence intervals are obtained through Fisher's Z-transformation.}
+	\label{fig:res_t1d_all}
+\end{figure}

### 2022-02-02 paper/paper-misspec-supp.tex
rm the two additional sets

@@
 
-\FloatBarrier
-
-\begin{figure}[p]
-	\centerline{\includegraphics[width=0.95\textwidth]{res-all-brca}}
-	\caption{Raw partial correlations (for \textit{all} models and \textit{all} sets) for predicting BRCA in the UK Biobank from external summary statistics.
-These are computed using function \texttt{pcor} of R package bigstatsr where 95\% confidence intervals are obtained through Fisher's Z-transformation.}
-	\label{fig:res_brca_all}
-\end{figure}
-
-\begin{figure}[p]
-	\centerline{\includegraphics[width=0.9\textwidth]{res-all-cad}}
-	\caption{Raw partial correlations (for \textit{all} models and \textit{all} sets) for predicting CAD in the UK Biobank from external summary statistics.
-These are computed using function \texttt{pcor} of R package bigstatsr where 95\% confidence intervals are obtained through Fisher's Z-transformation.}
-	\label{fig:res_cad_all}
-\end{figure}
-
-\begin{figure}[p]
-	\centerline{\includegraphics[width=0.9\textwidth]{res-all-mdd}}
-	\caption{Raw partial correlations (for \textit{all} models and \textit{all} sets) for predicting MDD in the UK Biobank from external summary statistics.
-These are computed using function \texttt{pcor} of R package bigstatsr where 95\% confidence intervals are obtained through Fisher's Z-transformation.}
-	\label{fig:res_mdd_all}
-\end{figure}
-
-\begin{figure}[p]
-	\centerline{\includegraphics[width=0.9\textwidth]{res-all-prca}}
-	\caption{Raw partial correlations (for \textit{all} models and \textit{all} sets) for predicting PRCA in the UK Biobank from external summary statistics.
-These are computed using function \texttt{pcor} of R package bigstatsr where 95\% confidence intervals are obtained through Fisher's Z-transformation.}
-	\label{fig:res_prca_all}
-\end{figure}
-
-\begin{figure}[p]
-	\centerline{\includegraphics[width=0.95\textwidth]{res-all-t1d}}
-	\caption{Raw partial correlations (for \textit{all} models and \textit{all} sets) for predicting T1D in the UK Biobank from external summary statistics.
-These are computed using function \texttt{pcor} of R package bigstatsr where 95\% confidence intervals are obtained through Fisher's Z-transformation.}
-	\label{fig:res_t1d_all}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 %\clearpage

### 2022-02-02 paper/paper-misspec.tex
rm the two additional sets

@@
 
+[TODO: UPDATE AND REMOVE NEW SETS]
+
 We use external summary statistics for which INFO scores are reported (Table \ref{tab:sumstats}); they all have a very high mean INFO score (larger than 0.94), except for BRCA-iCOGS (0.841) and T1D-affy (0.885).
@@
 \end{table}
-
-\subsection{Two new sets of variants}
-
-We also design two larger sets of imputed variants to compare against using only HapMap3 variants for prediction. 
-Following \cite{prive2021high}, we first restrict to UKBB variants with MAF > 0.01 and INFO > 0.3.
-We then compile frequencies and imputation INFO scores from other datasets, iPSYCH and summary statistics for breast cancer, prostate cancer, coronary artery disease, type-1 diabetes and depression \cite[]{bybjerg2020ipsych2015,michailidou2017association,schumacher2018association,nikpay2015comprehensive,censin2017childhood,wray2018genome}.
-We restrict to variants with a mean INFO > 0.3 in these other datasets, and also compute the median frequency per variant.
-To exclude potential mismappings in the genotyped data \cite[]{kunert2020allele} that might have propagated to the imputed data, we compare median frequencies in the external data to the ones in the UK Biobank.
-As we expect these potential errors to be localized around errors in the genotype data, we apply a moving-average smoothing on the frequency differences to increase power to detect these errors and also reduce false positives.
-We define the threshold (of 0.03) on these smoothed differences based on visual inspection of their histogram.
-This results in an initial set of 9,394,361 variants.
-
-We then define the two sets from this large set of variants.
-One is based on clumping, using a threshold $r^2 = 0.9$ over a radius of 100 Kbp and prioritizing HapMap3 variants and larger INFO scores.
-This results in a set ``clump'' of 2,465,478 variants, among which there are 554,655 of the initial HapMap3 variants.
-For the second set, we aim at maximizing the tagging of all the initial 9,394,361 variants, i.e.
-$\sum_{i \in \text{all}} \max_{j \in \text{set}}{|R_{i,j}|}$, where $R_{i,j}$ is the correlation between variants $i$ and $j$ (inspired from the alternative sensitivity of \cite{agier2016systematic}).
-We design a greedy algorithm that first selects all HapMap3 variants, then adds one variant at a time, the one that maximizes the addition to this sum, until no variant can add more than 0.2.
-This results in a set ``maxtag'' of 2,029,086 variants.
 

### 2022-02-04 paper/paper-misspec-supp.tex
update intro

@@
 
+\begin{figure}[p]
+	\centerline{\includegraphics[width=0.98\textwidth]{compare-all-maf-info-chr22}}
+	\caption{Minor allele frequencies (MAF) and INFO scores in the UK Biobank for 50,000 variants on chromosome 22 (chosen at random).
+These are either reported (in UKBB MFI files), recomputed from the whole data or from the subset of 362,307 European individuals used in this paper.
+\label{fig:compare-maf-info}}
+\end{figure}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{figure}[h]
+	\centerline{\includegraphics[width=0.95\textwidth]{simu-rounded}}
+	\caption{Results for the simulations with summary statistics (effect sizes and their standard errors) possibly rounded to two significant digits, averaged over 10 simulations for each scenario.
+Reported 95\% confidence intervals are computed from 10,000 non-parametric bootstrap replicates of the mean.
+Red bars correspond to using the LD with independent blocks (Methods).}
+	\label{fig:simu-rounded}
+\end{figure}
 

### 2022-02-04 paper/paper-misspec.tex
update intro

@@
 
-\title{Identifying and correcting multiple sources of misspecification\\in GWAS summary statistics for polygenic scores}
+\title{Identifying and correcting for misspecifications\\in GWAS summary statistics for polygenic scores}
 \author{Florian Priv\'e,$^{\text{1,}*}$ Julyan Arbel,$^{\text{2}}$ Hugues Aschard,$^{\text{3,4}}$ and Bjarni J. Vilhj\'almsson$^{\text{1,5}}$}
@@
 Can we detect possible misspecifications in GWAS summary statistics and correct for them in order to improve predictive performance of polygenic scores?
+	[TODO: LIST SOME OF THEM (AND POSSIBLE CONSEQUENCES?)]
 }
@@
 
-Contrary to individual-level genotypes and phenotypes, summary statistics resulting from genome-wide association studies (GWAS) are widely available.
-Therefore, they have been extensively used e.g.\ to derive polygenic scores (PGS).
-However, GWAS summary statistics come with uneven qualities, such as the imputation qualities of each variant reported. 
-There is also some heterogeneity in the methods used for performing the individual GWAS and their meta-analyses, as well as the information reported in the resulting summary statistics.
-Moreover, many PGS methods such as PRS-CS, SBayesR, and LDpred2 \cite[]{ge2019polygenic,lloyd2019improved,prive2020ldpred2} use Bayesian models, which can be sensitive to model misspecifications \cite[]{WALKER20131621,miller2018robust}.
-Previously, to make sure that the input parameters used in LDpred2 were consistent with its modeling assumptions, we proposed a quality control (QC) based on comparing standard deviations (of genotypes) inferred from GWAS summary statistics with the ones computed from a reference panel \cite[]{prive2020ldpred2}.
-This was particularly important for LDpred2-auto, which directly estimates key model parameters from the data.
+Contrary to individual-level genotypes and phenotypes, summary statistics resulting from genome-wide association studies (GWAS) are widely available, and one can obtain very large sample sizes through meta-analysis \cite[]{yengo2022saturated}.
+Therefore, they have been extensively used to e.g.\ derive polygenic scores (PGS), perform fine-mapping, and estimate key genetic architecture parameters.
+However, GWAS summary statistics come with uneven quality, such as the imputation accuracy of each variant reported. 
+There is also heterogeneity in the information made available in these summary statistics, e.g.\ per-variant imputation INFO scores, sample sizes and allele frequencies are often missing.
+Moreover, many methods based on summary statistics use Bayesian models, which can be sensitive to model misspecifications \cite[]{WALKER20131621,miller2018robust}.
 
-Here, we present an overview of the possible misspecifications that come with GWAS summary statistics (Table \ref{tab:overview}).
-Then, we investigate some of those, and propose corrections to improve the predictive performance of polygenic scores derived from GWAS summary statistics.
-We approach this from three different angles.
-First, based on additional summary information such as the imputation INFO scores and allele frequencies from the GWAS summary statistics, we refine our previously proposed QC \cite{prive2020ldpred2}.
-For example, we show that standard deviations of imputed genotypes (allele dosages) are lower than the expected values under Hardy-Weinberg equilibrium.
-Second, we investigate possible corrections to apply to the input parameters of polygenic methods, namely the reference LD (linkage disequilibrium) matrix, the GWAS effect sizes, their standard errors and the corresponding sample sizes.
-For example, we show that GWAS effect sizes computed from imputed dosages are larger in magnitude compared to if true genotypes were available.
-Third, we introduce two new optional parameters in LDpred2-auto to make it more robust to these types of misspecification.
-We focus our investigations on LDpred2 and lassosum for two reasons.
-First, multiple studies have shown that LDpred2 and lassosum are consistently ranking best among methods for polygenic prediction \cite[]{mak2017polygenic,prive2020ldpred2,pain2021evaluation,kulm2021systematic}.
-Second, we reimplement and use a new version of lassosum, called lassosum2, that uses the exact same input parameters as LDpred2, which makes it easy for us to test the different QCs and corrections presented here.
-We perform our investigations using simulations first, then we use public summary statistics while restricting to the widely-used HapMap3 variants, finally we investigate two alternative sets with more variants.
+We present an overview of possible misspecifications that come with GWAS summary statistics (Table \ref{tab:overview}). 
+First, the total sample size used can be misestimated, which would result e.g.\ in a biased estimation of the SNP heritability.
+For example, the total effective sample size is overestimated when computing it from the total number of cases and controls from a meta-analysis of binary outcomes \cite[]{grotzinger2021pervasive}; using BOLT-LMM summary statistics can result in an increased effective sample size \cite[]{loh2018mixed,gazal2018functional}; using SAIGE on binary traits with a large prevalence can result in a reduced effective sample size \cite[]{zhou2018efficiently,mbatchou2021computationally}.
+Second, per-variant sample sizes can be very different when meta-analyzing GWAS summary statistics from multiple cohorts with different sets of variants \cite[]{wang2021global}.
+Per-variant sample sizes should be used instead of the overall total sample size to avoid any bias resulting from overestimating the sample size.
+Moreover, using very different per-variant sample sizes could be problematic for models implicitly assuming that summary statistics have all been derived from the same individuals \cite[]{zhu2017bayesian,zhou2021fast}.
+As a QC step, one can e.g.\ remove SNPs with a sample size less than 0.67 times the 90th percentile of all sample sizes \cite[]{zheng2017ld}.
+Third, many genetic variants are imputed, which can cause some bias in summary statistics.
+Fourth, the quality of imputation of each variant, often used in a quality control (QC) step in statistical genetics analyses, can be severely misestimated (often overestimated) when computed for multi-ancestry individuals instead of a more homogeneous subset.
+Fifth, there can be some errors in the summary statistics (e.g.\ allele inversions); QC is particularly important here.
+Sixth, there can be some mismatch between the GWAS summary statistics and the LD reference used.
 
-\begin{table}[h]
+\begin{table}[!htb]
 	\centering
-	\begin{tabular}{|p{\dimexpr0.2\textwidth-2\tabcolsep-\arrayrulewidth\relax}|
-			p{\dimexpr0.5\textwidth-2\tabcolsep-\arrayrulewidth\relax}|
+	\begin{tabular}{|p{\dimexpr0.21\textwidth-2\tabcolsep-\arrayrulewidth\relax}|
+			p{\dimexpr0.49\textwidth-2\tabcolsep-\arrayrulewidth\relax}|
 			p{\dimexpr0.3\textwidth-2\tabcolsep-\arrayrulewidth\relax}|}
@@
 		\hline
-		Global sample size misspecification$^{(1)}$ & Bias estimation of SNP heritability \cite[]{gazal2018functional,grotzinger2021pervasive} & Estimation of the effective sample size \\
+		Misestimation of the total sample size$^{(1)}$ & Bias estimation of SNP heritability \cite[]{gazal2018functional,grotzinger2021pervasive} & Estimation of the effective sample size \\
 		\hline
-		Differences in per-variant sample size & Break some model assumptions & Imputation using equations \ref{eq:imputeN} and \ref{eq:imputeNeff}, and filtering$^{(2)}$   \\
+		Differences in per-variant sample size & Possible violation of model assumptions resulting in poor model convergence & Imputation using equations \ref{eq:imputeN} and \ref{eq:imputeNeff}, and filtering$^{(2)}$   \\
 		\hline
@@
 		\hline
-		Using imputed dosages for LD computation & LD seems correctly estimated (Figure \ref{fig:compare-cor}) & None required \\
+		Using imputed dosages for LD computation &  & None required; LD seems correctly estimated (Figure \ref{fig:compare-cor}) \\
 		\hline 
-		Ancestry mismatch between summary statistics and LD & Strong mismatch of LD and allele frequencies, often resulting in divergence of models & Check ancestry proportions from GWAS summary statistics \cite[]{prive2021using} \\
+		Misestimation of imputation INFO scores & Misuse when filtering or adjusting input parameters based on these & Recomputing for homogeneous subset \\
 		\hline
@@
 		\hline
-		Any & Violation of model assumptions resulting in poor model convergence & Using more regularization \\
+		Rounding of summary statistics (effect sizes and SEs) & Add unnecessary noise to the data & None, but this has almost no impact on the predictive performance (Figure \ref{fig:simu-rounded}) \\
+		\hline 
+		Ancestry mismatch between summary statistics and LD & Strong mismatch of LD and allele frequencies, often resulting in divergence of models & Checking ancestry proportions from GWAS summary statistics \cite[]{prive2021using} \\
+		\hline
+		Any & Possible violation of model assumptions resulting in poor model convergence & Quality control, using more regularization, and constraining LD to blocks \\
 		\hline
@@
 	\caption{Overview of possible misspecifications when using GWAS summary statistics. 
-	$^{(1)}$ Examples: using the global number of cases and controls in a meta-analysis of binary traits, a larger effective sample size when using BOLT-LMM summary statistics \cite[]{loh2018mixed}, and a reduced effective sample size when using SAIGE on binary traits with low case-control imbalance \cite[]{zhou2018efficiently,mbatchou2021computationally}.
+	$^{(1)}$ Examples: using the total number of cases and controls in a meta-analysis of binary traits, a larger effective sample size when using BOLT-LMM summary statistics \cite[]{loh2018mixed}, and a reduced effective sample size when using SAIGE on binary traits with a large prevalence \cite[]{zhou2018efficiently,mbatchou2021computationally}.
 	$^{(2)}$ E.g.\ removing SNPs with an effective sample size less than 0.67 times the 90th percentile of sample size \cite[]{zheng2017ld}.
@@
 \end{table}
+
+Here we investigate some of these misspecifications, and propose adjustments to improve the predictive performance of polygenic scores derived from GWAS summary statistics.
+We approach this from three different angles.
+First, based on additional summary information such as the imputation INFO scores and allele frequencies from the GWAS summary statistics, we refine our previously proposed QC \cite{prive2020ldpred2}.
+This QC consists in comparing standard deviations (of genotypes) inferred from GWAS summary statistics with the ones computed from a reference panel.
+This is useful to check that the input parameters used are consistent with each other. 
+This was particularly important for LDpred2-auto, which directly estimates key model parameters from the data.
+Here, we show e.g.\ that standard deviations of imputed genotypes (allele dosages) are lower than the expected values under Hardy-Weinberg equilibrium.
+Second, we investigate possible adjustments to apply to the input parameters of methods using summary statistics (focusing on PGS methods here), namely the reference LD (linkage disequilibrium) matrix, the GWAS effect sizes, their standard errors and the corresponding sample sizes.
+For example, we show that GWAS effect sizes computed from imputed dosages are larger in magnitude compared to if true genotypes were available.
+Third, we introduce two new optional parameters in LDpred2-auto to make it more robust to these types of misspecification.
+We focus our investigations on LDpred2 and lassosum for two reasons.
+First, multiple studies have shown that LDpred2 and lassosum rank among the best methods for single-trait polygenic prediction \cite[]{mak2017polygenic,prive2020ldpred2,pain2021evaluation,kulm2021systematic}.
+Second, we reimplement and use a new version of lassosum, called lassosum2, that uses the exact same input parameters as LDpred2, which makes it easy for us to test the different QCs and adjustments presented here.
 
@@
 
-We design simulations where variants have different GWAS sample sizes, which is often the case when meta-analyzing GWAS from multiple cohorts without the same genome coverage \cite[]{wang2021global}.
+We design simulations where variants have different GWAS sample sizes, which is often the case when meta-analyzing GWAS summary statistics from multiple cohorts with different sets of variants \cite[]{wang2021global}.
 Using 40,000 variants from chromosome 22 (Methods), we simulate quantitative phenotypes with a heritability of 20\% and 2000 causal variants.
@@
 Note that we initially included PRS-CS and SBayesR in our comparison \cite[]{ge2019polygenic,lloyd2019improved}.
-However, results for SBayesR always diverged and the overlap with the LD reference provided for PRS-CS was too small.
+However, results for SBayesR always diverged (stopped with an error), and the overlap with the LD reference provided for PRS-CS was too small.
 Averaged over 10 simulations, when providing true per-variant GWAS sample sizes, squared correlations between the polygenic scores and the simulated phenotypes are of 0.123 for C+T, 0.161 for lassosum, 0.169 for lassosum2, 0.159 for LDpred2(-grid), 0.140 for LDpred2-auto, and 0.141 for LDpred2-inf (Figure \ref{fig:simu-misN}).
@@
 To solve this limitation, we introduce two new parameters to make LDpred2-auto more robust, referred to as ``LDpred2-auto-rob'' here (Methods).
+The first parameter prevents $p$ from diverging to $1$, while the second shrinks the off-diagonal elements of the LD matrix (a form of regularization). 
 Second, for lassosum2, results for a grid of parameters (over $\lambda$ and $\delta$) are quite smooth compared to LDpred2 (Figures \ref{fig:lassosum2-misN} and \ref{fig:ldpred2-misN}). 
@@
 In simulations (cf.\ Methods section ``Data for simulations''), we verify that $\text{sd}(G_j)^{\text{true}} \approx \text{sd}(G_j)^{\text{imp}} / \sqrt{\text{INFO}_j}$ (Figure \ref{fig:compare-sd}).
-We also show that GWAS effect sizes $\hat{\gamma}$ computed from imputed dosages are over-estimated: $\hat{\gamma}_j^{\text{true}} \approx \hat{\gamma}_j^{\text{imp}} \cdot \sqrt{\text{INFO}_j}$ and $\text{se}(\hat{\gamma}_j)^{\text{true}} \approx \text{se}(\hat{\gamma}_j)^{\text{imp}} \cdot \sqrt{\text{INFO}_j}$ (Figures \ref{fig:compare-beta} and \ref{fig:compare-beta-se}).
+As a direct consequence of the lower standard deviation of dosages, we also show that GWAS effect sizes $\hat{\gamma}$ computed from imputed dosages are overestimated: $\hat{\gamma}_j^{\text{true}} \approx \hat{\gamma}_j^{\text{imp}} \cdot \sqrt{\text{INFO}_j}$ and $\text{se}(\hat{\gamma}_j)^{\text{true}} \approx \text{se}(\hat{\gamma}_j)^{\text{imp}} \cdot \sqrt{\text{INFO}_j}$ (Figures \ref{fig:compare-beta} and \ref{fig:compare-beta-se}).
 This is the first correction of summary statistics we consider in the simulations below. 
-Note that we recompute INFO scores for the subset of 362,307 European individuals used in this paper as they can differ substantially from the ones reported by the UK Biobank for the whole data (Figure \ref{fig:compare-info}).
+Note that we recompute INFO scores for the subset of 362,307 European individuals used in this paper as they can differ substantially from the ones reported by the UK Biobank for the whole data (Figures \ref{fig:compare-info} and \ref{fig:compare-maf-info}).
 As a second option, instead of using dosages to compute the GWAS summary statistics, it has been argued that using multiple imputation (MI) would be more appropriate \cite[]{Palmer2016}.
@@
 For lassosum2 and LDpred2(-grid), which tune parameters using the validation set, correcting for imputation quality slightly improves predictive performance in these simulations (Figure \ref{fig:simu-info}).
-However, correcting for imputation quality can dramatically improve predictive performance for LDpred2-auto. 
+However, correcting for imputation quality can dramatically improve predictive performance for LDpred2-auto, provided the imputation quality is well estimated. 
 Moreover, new additions for robustness introduced before, namely LDpred2-low-h2, LDpred2-auto-rob, and forming independent blocks in the LD matrix also improve predictive performance for all corrections (Figure \ref{fig:simu-info}).
@@
 
-\subsection{Other phenotypes and larger sets of variants}
-
-[TODO: UPDATE AND REMOVE NEW SETS]
+\subsection{Results for other phenotypes}
 
@@
 We show previous results as figure \ref{fig:res_brca} for other phenotypes in figures \ref{fig:res_cad}-\ref{fig:res_t1d}; most changes introduced before have little or no impact on the predictions, except for the second quality control performed for CAD when using LDpred2-auto (Figure \ref{fig:res_cad}).
-We then introduce two new sets of variants as possible replacement for HapMap3 variants (Methods).
-These sets include more variants (more than 2M), therefore possibly of lower quality on average. 
-We present similar results with these two new sets of variants in figures \ref{fig:res_brca_all}-\ref{fig:res_t1d_all}.
-Using the ``maxtag'' set generally provides larger predictive performance than using HapMap3 variants, especially when predicting prostate cancer (Figure \ref{fig:res_prca_all}).
 
@@
 Additional QC can also be performed, e.g.\ comparing reported GWAS allele frequencies with the ones from the LD reference panel, e.g.\ to detect genotyping or allelic errors.
-We do perform this additional QC as part of ``qc2'' here, and also when designing the large set of variants from the UK Biobank (Methods). 
+We do perform this additional QC as part of ``qc2'' here. 
 One can also run other QC tools such as DENTIST \cite[]{chen2021improved}, and also infer ancestry proportions from summary statistics to make sure these are matching with the LD reference used \cite[]{prive2021using}.
@@
 Sometimes, it can even be challenging to recover the total effective sample size from large meta-analyses.
-We recall that if some studies of a meta-analysis have an imbalanced number of cases and controls, the global effective sample size should not be computed from the total numbers of cases and controls overall, but instead from the sum of the effective sample sizes of each study.
+We recall that if some studies of a meta-analysis have an imbalanced number of cases and controls, the total effective sample size should not be computed from the total numbers of cases and controls overall, but instead from the sum of the effective sample sizes of each study.
 Indeed, take the extreme example of meta-analyzing two studies, one with 1000 cases and 0 controls, and another one with 0 cases and 1000 controls, then the effective sample size of the meta-analysis is 0, not 2000.
@@
 Although these results are somewhat unfortunate, they are reassuring because it means that the summary statistics we use in this study are usually of good quality for follow-up analyses such as deriving polygenic scores.
+
+[TODO: PRS-CS REGU + SMALL LD REF UNMATCHED WITH ANCESTRY + LD BLOCKS]
+[LD BLOCKS MAINLY FOR SCALA BUT SHOW THAT USEFUL FOR ROBUSTNESS]
 
@@
 We restrict to the genetic variants to the 1,054,315 HapMap3 variants used in the LD reference provided in \cite{prive2020ldpred2}.
-We also try two new sets of variants (see next section).
 
@@
 \end{equation}
-In practice, we also bound this estimate to be between $0.5 \cdot N$ and $1.1 \cdot N$, where $N$ is the global (effective) sample size.
+In practice, we also bound this estimate to be between $0.5 \cdot N$ and $1.1 \cdot N$, where $N$ is the total (effective) sample size.
+Also note that the estimate of $\text{var}(G_j)$ has to account for the imputation accuracy (Section \ref{sec:imputation}).
 
@@
 All code used for this paper is available at \url{https://github.com/privefl/paper-misspec/tree/master/code}.
+We have recomputed the allele frequencies and imputation INFO scores within the European subset used here and across all variants in the UK Biobank, and made them available at \url{https://doi.org/10.6084/m9.figshare.16635388.v1}.
 The latest version of R package bigsnpr can be installed from GitHub.
@@
 
-%\clearpage
-\vspace*{3em}
+\clearpage
+%\vspace*{3em}
 

### 2022-02-04 paper/paper-misspec.tex
add mismatched LD

@@
 \end{figure}
+
+\subsection{Mismatch between LD reference and GWAS summary statistics}
+
+Here we design simulations to understand the impact of using a mismatched LD reference panel, e.g.\ that comes from a different population compared to the one used to compute the GWAS summary statistics.
+We use the same simulation setup as before (Methods).
+In addition, we design an alternative reference panel based on 10,000 individuals from South Europe by using the ``Italy'' center defined in \cite{prive2021high}.
+Averaged over 10 simulations, when using an ``alternative'' LD reference panel instead of well-matched one, squared correlations between the polygenic scores and the simulated phenotypes drop from 0.174 to 0.167 for lassosum2, from 0.168 to 0.147 for LDpred2(-grid), from 0.174 to 0.169 for LDpred2-low-h2, from 0.143 to 0.139 for LDpred2-auto, from 0.162 to 0.146 for LDpred2-auto-rob, and from 0.144 to 0.140 for LDpred2-inf (Figure \ref{fig:simu-altpop}).
+Defining LD blocks in these two LD matrices always helps, but do not improve performance by much when using the alternative LD.
 

### 2022-02-04 paper/paper-misspec.tex
extend discu

@@
 In the real data applications hereinafter, we choose to use the first correction, ``sqrt\_info'', which is simple because it is equivalent to post-processing PGS effects by multiplying them by $\sqrt{\text{INFO}}$.
-[TODO: TRY THEM ALL AGAIN?
-AND META-ANALYZE (BRCA AND T1D)?]
 
@@
 
-We have investigated misspecifications in GWAS summary statistics, focusing on the impact of sample size heterogeneity and imputation quality.
+We have investigated misspecifications in GWAS summary statistics, focusing particularly on the impact of sample size heterogeneity and imputation quality, and the application to polygenic scores (PGS) methods.
 Previously, we proposed a quality control (QC) based on comparing standard deviations (of genotypes) inferred from GWAS summary statistics with the ones computed from a reference panel \cite[]{prive2020ldpred2}.
@@
 Fortunately, this has practically no effect on the predictive performance of the derived polygenic scores.
-Additional QC can also be performed, e.g.\ comparing reported GWAS allele frequencies with the ones from the LD reference panel, e.g.\ to detect genotyping or allelic errors.
+Additional QC can also be performed, e.g.\ comparing reported GWAS allele frequencies with the ones from the LD reference panel, e.g.\ to detect genotyping errors or allele inversions.
 We do perform this additional QC as part of ``qc2'' here. 
@@
 
-[TODO: PRS-CS REGU + SMALL LD REF UNMATCHED WITH ANCESTRY + LD BLOCKS]
-[LD BLOCKS MAINLY FOR SCALA BUT SHOW THAT USEFUL FOR ROBUSTNESS]
+Anyway, we recommend to adopt these changes, i.e.\ performing the (refined) QC proposed here, defining LD blocks in the LD matrix, and using more regularization when needed (i.e.\ also testing smaller values for the heritability parameter in LDpred2-grid, and using the two new parameters in LDpred2-auto (Section \ref{sec:robust}).
+Note that LD blocks are widely used already by several methods (e.g.\ lassosum and PRS-CS) because they allow for processing smaller matrices at once \cite[]{mak2017polygenic,ge2019polygenic}; here we have shown that these blocks are also useful to make methods more robust.
+PRS-CS is one of the most robust PGS methods out there; for example, it can use the (small) 1000 Genomes data as LD reference \cite[]{ge2019polygenic}, and can even use an European LD reference panel with multi-ancestry GWAS summary statistics \cite[]{wang2021global}.
+We think this is made possible by the use of a (very) strong regularization in PRS-CS ($\phi^{-1} \psi_j^{-1} \ge 1$, which would approximately corresponds to using $s=0.5$ in lassosum, $\delta = 1$ in lassosum2, and $\texttt{shrink\_corr}=0.5$ in LDpred2-auto).
+Using enough regularization is good for robustness, but note that using too much regularization can also damage predictive performance, and should not be a replacement for a proper QC and choice of ancestry-matched LD [TODO: A BIT AGGRESSIVE?].
+As a future direction, we are interested in using ancestry-matched LD matrices to improve polygenic prediction in all ancestries.
+Also it would be useful to investigate the impact of the QC and adjustments we propose here on other (non-PGS) methods, such as the ones for inference, where robustness is likely to be very important as well.
 
@@
 
-\subsection{LDpred2-low-h2 and LDpred2-auto-rob}
+\subsection{LDpred2-low-h2 and LDpred2-auto-rob}\label{sec:robust}
 

### 2022-02-10 paper/paper-misspec.tex
text minor

@@
 
-Contrary to individual-level genotypes and phenotypes, summary statistics resulting from genome-wide association studies (GWAS) are widely available, and one can obtain very large sample sizes through meta-analysis \cite[]{yengo2022saturated}.
+Contrary to individual-level genotypes and phenotypes, summary statistics resulting from genome-wide association studies (GWAS) are widely available, and very large sample sizes can be obtained through meta-analysis \cite[]{yengo2022saturated}.
 Therefore, they have been extensively used to e.g.\ derive polygenic scores (PGS), perform fine-mapping, and estimate key genetic architecture parameters.
@@
 First, the total sample size used can be misestimated, which would result e.g.\ in a biased estimation of the SNP heritability.
-For example, the total effective sample size is overestimated when computing it from the total number of cases and controls from a meta-analysis of binary outcomes \cite[]{grotzinger2021pervasive}; using BOLT-LMM summary statistics can result in an increased effective sample size \cite[]{loh2018mixed,gazal2018functional}; using SAIGE on binary traits with a large prevalence can result in a reduced effective sample size \cite[]{zhou2018efficiently,mbatchou2021computationally}.
+For example, the total effective sample size is overestimated when computed from the total number of cases and controls from a meta-analysis of binary outcomes \cite[]{grotzinger2021pervasive}; using BOLT-LMM summary statistics can result in an increased effective sample size \cite[]{loh2018mixed,gazal2018functional}; using SAIGE on binary traits with a large prevalence can result in a reduced effective sample size \cite[]{zhou2018efficiently,mbatchou2021computationally}.
 Second, per-variant sample sizes can be very different when meta-analyzing GWAS summary statistics from multiple cohorts with different sets of variants \cite[]{wang2021global}.
@@
 Moreover, using very different per-variant sample sizes could be problematic for models implicitly assuming that summary statistics have all been derived from the same individuals \cite[]{zhu2017bayesian,zhou2021fast}.
-As a QC step, one can e.g.\ remove SNPs with a sample size less than 0.67 times the 90th percentile of all sample sizes \cite[]{zheng2017ld}.
+As a quality control (QC) step, one can e.g.\ remove SNPs with a sample size less than 0.67 times the 90th percentile of all sample sizes \cite[]{zheng2017ld}.
 Third, many genetic variants are imputed, which can cause some bias in summary statistics.
-Fourth, the quality of imputation of each variant, often used in a quality control (QC) step in statistical genetics analyses, can be severely misestimated (often overestimated) when computed for multi-ancestry individuals instead of a more homogeneous subset.
-Fifth, there can be some errors in the summary statistics (e.g.\ allele inversions); QC is particularly important here.
+Fourth, the quality of imputation of each variant, often used in a QC step in statistical genetics analyses, can be severely misestimated (often overestimated) when computed for multi-ancestry individuals instead of a more homogeneous subset.
+Fifth, errors such as allele inversions can be present in the summary statistics; QC is particularly important here.
 Sixth, there can be some mismatch between the GWAS summary statistics and the LD reference used.
@@
 		\hline
-		Using imputed dosages for GWAS & Standard deviations of dosages are smaller, resulting in overestimated effect sizes and standard errors & See section \ref{sec:imputation} \\
+		Using imputed dosages for GWAS & Standard deviations of dosages are too small, resulting in overestimated effect sizes and standard errors & See section \ref{sec:imputation} \\
 		\hline
@@
 	\end{tabular}
-	\caption{Overview of possible misspecifications when using GWAS summary statistics. 
-	$^{(1)}$ Examples: using the total number of cases and controls in a meta-analysis of binary traits, a larger effective sample size when using BOLT-LMM summary statistics \cite[]{loh2018mixed}, and a reduced effective sample size when using SAIGE on binary traits with a large prevalence \cite[]{zhou2018efficiently,mbatchou2021computationally}.
+	\caption{Overview of possible misspecifications when using GWAS summary statistics, along with possible consequences and corrections.\\
+	$^{(1)}$ Examples: using the total number of cases and controls in a meta-analysis of binary traits, a larger effective sample size when using BOLT-LMM summary statistics \cite[]{loh2018mixed}, and a reduced effective sample size when using SAIGE on binary traits with a large prevalence \cite[]{zhou2018efficiently,mbatchou2021computationally}.\\
 	$^{(2)}$ E.g.\ removing SNPs with an effective sample size less than 0.67 times the 90th percentile of sample size \cite[]{zheng2017ld}.
@@
 We approach this from three different angles.
-First, based on additional summary information such as the imputation INFO scores and allele frequencies from the GWAS summary statistics, we refine our previously proposed QC \cite{prive2020ldpred2}.
+First, based on additional summary information such as the imputation INFO scores and allele frequencies from the GWAS summary statistics, we refine our previously proposed QC \cite[]{prive2020ldpred2}.
 This QC consists in comparing standard deviations (of genotypes) inferred from GWAS summary statistics with the ones computed from a reference panel.
@@
 This was particularly important for LDpred2-auto, which directly estimates key model parameters from the data.
-Here, we show e.g.\ that standard deviations of imputed genotypes (allele dosages) are lower than the expected values under Hardy-Weinberg equilibrium.
+Here, we show e.g.\ that standard deviations of imputed genotypes (allele dosages) are lower than the expected values under Hardy--Weinberg equilibrium.
 Second, we investigate possible adjustments to apply to the input parameters of methods using summary statistics (focusing on PGS methods here), namely the reference LD (linkage disequilibrium) matrix, the GWAS effect sizes, their standard errors and the corresponding sample sizes.
@@
 First, the results for LDpred2-auto are the same as with LDpred2-inf because it always converges to an infinitesimal model ($p = 1$) in these simulations.
-To solve this limitation, we introduce two new parameters to make LDpred2-auto more robust, referred to as ``LDpred2-auto-rob'' here (Methods).
+To overcome this limitation, we introduce two new parameters to make LDpred2-auto more robust, referred to as ``LDpred2-auto-rob'' here (Methods).
 The first parameter prevents $p$ from diverging to $1$, while the second shrinks the off-diagonal elements of the LD matrix (a form of regularization). 
@@
 \cite{marchini2010genotype} showed that the IMPUTE INFO measure is highly concordant with the MACH measure $\hat{r}_j^2 = \frac{\text{var}(G_j)}{2 \hat{\theta}_j (1 - \hat{\theta}_j)}$, where $\hat{\theta}_j$ is the estimated allele frequency of $G_j$, the genotypes for variant $j$.
-Therefore, when using the expected genotypes from imputation (dosages), their standard deviations are often lower than the expected value under Hardy-Weinberg equilibrium, because $\text{INFO}_j \approx \frac{\text{var}(G_j)}{2 \hat{\theta}_j (1 - \hat{\theta}_j)}$.
+Therefore, when using the expected genotypes from imputation (dosages), their standard deviations are often lower than the expected value under Hardy--Weinberg equilibrium, because $\text{INFO}_j \approx \frac{\text{var}(G_j)}{2 \hat{\theta}_j (1 - \hat{\theta}_j)}$.
 In simulations (cf.\ Methods section ``Data for simulations''), we verify that $\text{sd}(G_j)^{\text{true}} \approx \text{sd}(G_j)^{\text{imp}} / \sqrt{\text{INFO}_j}$ (Figure \ref{fig:compare-sd}).
@@
 In simulations, we show that $\hat{\gamma}_j^{\text{MI}} \approx \hat{\gamma}_j^{\text{imp}} \cdot \text{INFO}_j$ and $Z_j^{\text{MI}} \approx Z_j^{\text{imp}} \cdot \text{INFO}_j$, where $Z = \hat{\gamma} / \text{se}(\hat{\gamma})$ (Figure \ref{fig:compare-mi}).
-This is the second correction of summary statistics we implement in the simulations below, along with $n_j = N \cdot \text{INFO}_j$. 
+This is the second correction of summary statistics we implement in the simulations below, along with $n_j = N \cdot \text{INFO}_j$ where $n_j$ and $N$ are the per-variant and total sample sizes. 
 Finally, we consider an in-between solution as a third correction, using $\hat{\gamma}_j = \hat{\gamma}_j^{\text{imp}} \cdot \text{INFO}_j$, $\text{se}(\hat{\gamma}_j) = \text{se}(\hat{\gamma}_j)^{\text{imp}} \cdot \sqrt{\text{INFO}_j}$, and $n_j = N \cdot \text{INFO}_j$.
@@
 In addition, we design an alternative reference panel based on 10,000 individuals from South Europe by using the ``Italy'' center defined in \cite{prive2021high}.
-Averaged over 10 simulations, when using an ``alternative'' LD reference panel instead of well-matched one, squared correlations between the polygenic scores and the simulated phenotypes drop from 0.174 to 0.167 for lassosum2, from 0.168 to 0.147 for LDpred2(-grid), from 0.174 to 0.169 for LDpred2-low-h2, from 0.143 to 0.139 for LDpred2-auto, from 0.162 to 0.146 for LDpred2-auto-rob, and from 0.144 to 0.140 for LDpred2-inf (Figure \ref{fig:simu-altpop}).
+Averaged over 10 simulations, when using an ``alternative'' LD reference panel instead of a well-matched one, squared correlations between the polygenic scores and the simulated phenotypes drop from 0.174 to 0.167 for lassosum2, from 0.168 to 0.147 for LDpred2(-grid), from 0.174 to 0.169 for LDpred2-low-h2, from 0.143 to 0.139 for LDpred2-auto, from 0.162 to 0.146 for LDpred2-auto-rob, and from 0.144 to 0.140 for LDpred2-inf (Figure \ref{fig:simu-altpop}).
 Defining LD blocks in these two LD matrices always helps, but do not improve performance by much when using the alternative LD.
@@
 
-Breast cancer summary statistics are interesting because they include results from two mega analyses \cite[]{michailidou2013large,michailidou2015genome,michailidou2017association}, which means there is some larger precision in the parameters reported in the GWAS summary statistics, such as the INFO scores and the sample sizes.
-Imputation INFO scores for the OncoArray summary statistics (after having restricted to HapMap3 variants) are generally very good (Figure \ref{fig:hist-info-brca}) and better than the ones from iCOGS (Figure \ref{fig:hist-info-brca2}), probably because the chip used included around 200K variants only, compared to more than 500K variants for the OncoArray.
+Breast cancer summary statistics are interesting because they include results from two mega analyses \cite[]{michailidou2013large,michailidou2015genome,michailidou2017association}, which means that parameters reported in the GWAS summary statistics, such as the INFO scores and the sample sizes, are estimated with high precision.
+Imputation INFO scores for the OncoArray summary statistics (after having restricted to HapMap3 variants) are generally very good (Figure \ref{fig:hist-info-brca}) and better than the ones from iCOGS (Figure \ref{fig:hist-info-brca2}), probably because the iCOGS chip included around 200K variants only, compared to more than 500K variants for the OncoArray.
 For both summary statistics, we first compare the standard deviations (of genotypes) inferred from the reported allele frequencies (i.e.\ $\sqrt{2 f (1 - f)}$ where $f$ is the allele frequency, and denoted as sd\_af) versus the ones inferred from the GWAS summary statistics (Equation \eqref{eq:approx-sd-log}, and denoted as sd\_ss).
@@
 	\centerline{\includegraphics[width=0.98\textwidth]{brca_onco_qc}}
-	\caption{Standard deviations inferred from the OncoArray breast cancer GWAS summary statistics (Equation \ref{eq:approx-sd-log}) versus the ones inferred from the reported GWAS allele frequencies ($\sqrt{2 f (1 - f)}$).
+	\caption{Standard deviations inferred from the OncoArray breast cancer GWAS summary statistics using equation \ref{eq:approx-sd-log} (\textbf{A:} Raw or \textbf{B:} dividing them by $\sqrt{\text{INFO}}$) versus the ones inferred from the reported GWAS allele frequencies $f_j$ (using $\sqrt{2 f_j (1 - f_j)}$).
 Only 100,000 HapMap3 variants are represented, at random.}
@@
 
-Therefore, providing an accurate imputation INFO score is important for two reasons.
+Therefore, providing an accurate imputation INFO score is useful for two reasons.
 First, it allows for correcting for a reduced standard deviation when using imputed data in the QC step we propose, in order to better uncover problems with the summary statistics.
@@
 Although results for both QC used are very similar, correcting for the INFO score slightly improves prediction when deriving polygenic scores based on iCOGS summary statistics (Figure \ref{fig:res_brca}).
-All other improvements introduced before have little or no effect here, probably because misspecifications are much smaller than in the simulations.
+All other improvements introduced before have little to no effect here, probably because misspecifications are much smaller than in the simulations.
 
@@
 We use external summary statistics for which INFO scores are reported (Table \ref{tab:sumstats}); they all have a very high mean INFO score (larger than 0.94), except for BRCA-iCOGS (0.841) and T1D-affy (0.885).
-QC plots comparing standard deviations usually show little deviation from $x=y$ (after the INFO score correction), except for coronary artery disease (CAD) summary statistics (Figures \ref{fig:brca_icogs_qc}-\ref{fig:t1d_illu_qc}).
-We show previous results as figure \ref{fig:res_brca} for other phenotypes in figures \ref{fig:res_cad}-\ref{fig:res_t1d}; most changes introduced before have little or no impact on the predictions, except for the second quality control performed for CAD when using LDpred2-auto (Figure \ref{fig:res_cad}).
+QC plots comparing standard deviations usually show little deviation from the identity line (after the INFO score correction), except for coronary artery disease (CAD) summary statistics (Figures \ref{fig:brca_icogs_qc}-\ref{fig:t1d_illu_qc}).
+We show previous results as figure \ref{fig:res_brca} for other phenotypes in figures \ref{fig:res_cad}-\ref{fig:res_t1d}; most changes introduced before have little to no impact on the predictions, except for the second quality control performed for CAD when using LDpred2-auto (Figure \ref{fig:res_cad}).
 
@@
 
-Anyway, we recommend to adopt these changes, i.e.\ performing the (refined) QC proposed here, defining LD blocks in the LD matrix, and using more regularization when needed (i.e.\ also testing smaller values for the heritability parameter in LDpred2-grid, and using the two new parameters in LDpred2-auto (Section \ref{sec:robust}).
-Note that LD blocks are widely used already by several methods (e.g.\ lassosum and PRS-CS) because they allow for processing smaller matrices at once \cite[]{mak2017polygenic,ge2019polygenic}; here we have shown that these blocks are also useful to make methods more robust.
-PRS-CS is one of the most robust PGS methods out there; for example, it can use the (small) 1000 Genomes data as LD reference \cite[]{ge2019polygenic}, and can even use an European LD reference panel with multi-ancestry GWAS summary statistics \cite[]{wang2021global}.
-We think this is made possible by the use of a (very) strong regularization in PRS-CS ($\phi^{-1} \psi_j^{-1} \ge 1$, which would approximately corresponds to using $s=0.5$ in lassosum, $\delta = 1$ in lassosum2, and $\texttt{shrink\_corr}=0.5$ in LDpred2-auto).
+Nevertheless, we recommend to adopt these changes, i.e.\ performing the (refined) QC proposed here, defining LD blocks in the LD matrix, and using more regularization when needed (i.e.\ also testing smaller values for the heritability parameter in LDpred2-grid, and using the two new parameters in LDpred2-auto (Section \ref{sec:robust}).
+Note that LD blocks are already widely used by several methods (e.g.\ lassosum and PRS-CS) because they allow for processing smaller matrices at once \cite[]{mak2017polygenic,ge2019polygenic}; here we have shown that these blocks are also useful to make methods more robust.
+PRS-CS is currently one of the most robust PGS methods; for example, it can use the (small) 1000 Genomes data as LD reference \cite[]{ge2019polygenic}, and can even use a European LD reference panel with multi-ancestry GWAS summary statistics \cite[]{wang2021global}.
+We think this is made possible by the use of a (very) strong regularization in PRS-CS ($\phi^{-1} \psi_j^{-1} \ge 1$, which would approximately correspond to using $s=0.5$ in lassosum, $\delta = 1$ in lassosum2, and $\texttt{shrink\_corr}=0.5$ in LDpred2-auto).
 Using enough regularization is good for robustness, but note that using too much regularization can also damage predictive performance, and should not be a replacement for a proper QC and choice of ancestry-matched LD [TODO: A BIT AGGRESSIVE?].
-As a future direction, we are interested in using ancestry-matched LD matrices to improve polygenic prediction in all ancestries.
+As a future research direction, we are interested in using ancestry-matched LD matrices to improve polygenic prediction in all ancestries.
 Also it would be useful to investigate the impact of the QC and adjustments we propose here on other (non-PGS) methods, such as the ones for inference, where robustness is likely to be very important as well.
@@
 This results in 362,307 individuals.
-We sample 300,000 individuals to form a training set (e.g.\ to run the GWAS), 10,000 individuals to form a validation set (to tune hyper-parameters), and use the remaining 52,307 individuals to form a test set (to evaluate final predictive models).
+We randomly sample 300,000 individuals to form a training set (e.g.\ to run the GWAS), 10,000 individuals to form a validation set (to tune hyper-parameters), and use the remaining 52,307 individuals to form a test set (to evaluate final predictive models).
 
@@
 \begin{cases}
-\text{sign}\left(u_j^{(t)}\right) \left(\left|u_j^{(t)}\right| - \lambda\right) / \left(\widetilde{X}_j^T \widetilde{X}_j + \delta\right) & \text{if } \left|u_j^{(t)}\right| > \lambda ~, \\
+\dfrac{\text{sign}\big(u_j^{(t)}\big) \big( \big|u_j^{(t)}\big| - \lambda \big)}{\widetilde{X}_j^T \widetilde{X}_j + \delta} & \text{if } \big|u_j^{(t)}\big| > \lambda ~, \\
 0 & \text{otherwise.}
@@
 \[
-u_j^{(t)} = r_j - \widetilde{X}_j^T \left( \widetilde{X} \beta^{(t-1)} - \widetilde{X}_j \beta_j^{(t-1)} \right) ~.
+u_j^{(t)} = r_j - \widetilde{X}_j^T \big( \widetilde{X} \beta^{(t-1)} - \widetilde{X}_j \beta_j^{(t-1)} \big) ~.
 \]
-Following the notations from \cite{prive2020ldpred2} and denoting $\widetilde{X} = \frac{1}{\sqrt{n-1}} C_n G S^{-1}$, where $G$ is the genotype matrix, $C_n$ is the centering matrix and $S$ is the diagonal matrix of standard deviations of the columns of $G$.
+Following the notations from \cite{prive2020ldpred2}, denote $\widetilde{X} = \frac{1}{\sqrt{n-1}} C_n G S^{-1}$, where $G$ is the genotype matrix, $C_n$ is the centering matrix and $S$ is the diagonal matrix of standard deviations of the columns of $G$.
 Then $\widetilde{X}_j^T \widetilde{X} = R_{j,.} = R_{.,j}^T$ and $\widetilde{X}_j^T \widetilde{X}_j = 1$.
@@
 where $r_j = \widehat{\beta}_j =  \dfrac{\widehat{\gamma}_j}{\sqrt{n_j ~ \text{se}(\widehat{\gamma}_j)^2 + \widehat{\gamma}_j^2}}$ and $\widehat{\gamma}_j$ is the GWAS effect of variant $j$ and $n$ is the GWAS sample size \cite[]{mak2017polygenic,prive2021high}.
-Then the most time-consuming part is computing $R_{.,j}^T \beta^{(t-1)}$.
+Then computing $R_{.,j}^T \beta^{(t-1)}$ is the most time-consuming part of each iteration.
 To make this faster, instead of computing $R_{.,j}^T \beta^{(t-1)}$ at each iteration ($j$ and $t$), we can start with an initial vector of 0s only (for all $j$) since $\beta^{(0)} \equiv 0$, and then updating this vector when $\beta_j^{(t)} \neq \beta_j^{(t-1)}$ only.
@@
 For $\lambda$, the default in lassosum uses a sequence of 20 values equally spaced on a log scale between 0.1 and 0.001.
-We use instead a sequence between $\lambda_0$ and $\lambda_0 / 100$ by default in lassosum2, where $\lambda_0 = \max_j \left|\widehat{\beta}_j\right|$ is the minimum $\lambda$ for which no variable enters the model because the L1-regularization is too strong.
-Note that we do not provide an ``auto'' version using pseudo-validation (as in \cite{mak2017polygenic}) as we have not found it to be very robust (Figure \ref{fig:pseudoval}).
+By default in lassosum2, we use a similar sequence, but between $\lambda_0$ and $\lambda_0 / 100$ instead, where $\lambda_0 = \max_j \big|\widehat{\beta}_j\big|$ is the minimum $\lambda$ for which no variable enters the model because the L1-regularization is too strong.
+Note that we do not provide an ``auto'' version using pseudo-validation (as in \cite{mak2017polygenic}) as we have not found it to be robust enough (Figure \ref{fig:pseudoval}).
 Also note that, as in LDpred2, we run lassosum2 genome-wide using a sparse correlation matrix which assumes that variants further away than 3 cM are not correlated, and therefore we do not require splitting the genome into independent LD blocks anymore (as done in lassosum).
@@
 For different numbers of blocks and maximum number of variants in each block, we use the split with the minimum cost within the ones reducing the original number of non-zero values to less than 60\% (or 70\% for chromosome 6).
-Having a correlation matrix with independent blocks prevents the small errors in the algorithm (e.g.\ the Gibbs sampler in LDpred2) from propagating to too many variants.
+Having a correlation matrix with independent blocks prevents small errors in the algorithm (e.g.\ the Gibbs sampler in LDpred2) from propagating to too many variants.
 It also makes running LDpred2 (and lassosum2) faster, taking about 60\% of the initial time (since only 60\% of the initial non-zero values of the correlation matrix are kept).
@@
 The first one, \texttt{shrink\_corr}, allows for shrinking off-diagonal elements of the correlation matrix.
-This is similar to parameter `s' in lassosum (Section \ref{lassosum2}), and act as a means of regularization.
+This is similar to parameter `s' in lassosum (Section \ref{lassosum2}), and acts as a means of regularization.
 We use a value of 0.9 in simulations and 0.95 in real data when running ``LDpred2-auto-rob'' (and the default value of 1, with no effect, when running ``LDpred2-auto'').
@@
 
-%\clearpage
-%\vspace*{5em}
+\clearpage
+%\vspace*{3em}
 

### 2022-02-18 paper/paper-misspec-supp.tex
add mismatched LD
save

@@
 	\centerline{\includegraphics[width=0.95\textwidth]{compare-mult-imp}}
-	\caption{Corrected GWAS \textbf{A:} Z-scores and \textbf{B:} effect sizes from imputed dosages (multiplying them by $\text{INFO}$) versus the ones from multiple imputation (MI, 10 draws used), colored by INFO scores.
+	\caption{(Over)corrected GWAS \textbf{A:} Z-scores and \textbf{B:} effect sizes from imputed dosages (multiplying them by $\text{INFO}$) versus the ones obtained from multiple imputation (MI, 20 draws used, \cite{Palmer2016}), colored by INFO scores.
 This shows only one tenth of the simulated variants, because MI can be computationally intensive.
@@
 
-\begin{figure}[h]
-	\centerline{\includegraphics[width=0.95\textwidth]{simu-rounded}}
+\FloatBarrier
+
+\begin{figure}[p]
+	\centerline{\includegraphics[width=0.9\textwidth]{simu-rounded}}
 	\caption{Results for the simulations with summary statistics (effect sizes and their standard errors) possibly rounded to two significant digits, averaged over 10 simulations for each scenario.
@@
 	\label{fig:simu-rounded}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{figure}[p]
+	\centerline{\includegraphics[width=0.9\textwidth]{simu-altpop}}
+	\caption{Results for the simulations with summary statistics with LD matrices based on two different populations.
+One comes from the same population used for computing the GWAS summary statistics (North-West Europe), while the other one comes from South Europe (alternative LD reference).
+Reported 95\% confidence intervals are computed from 10,000 non-parametric bootstrap replicates of the mean.
+Red bars correspond to using the LD with independent blocks (Methods).}
+	\label{fig:simu-altpop}
 \end{figure}

### 2022-02-18 paper/paper-misspec.tex
save

@@
 Contrary to individual-level genotypes and phenotypes, summary statistics resulting from genome-wide association studies (GWAS) are widely available, and very large sample sizes can be obtained through meta-analysis \cite[]{yengo2022saturated}.
-Therefore, they have been extensively used to e.g.\ derive polygenic scores (PGS), perform fine-mapping, and estimate key genetic architecture parameters.
-However, GWAS summary statistics come with uneven quality, such as the imputation accuracy of each variant reported. 
-There is also heterogeneity in the information made available in these summary statistics, e.g.\ per-variant imputation INFO scores, sample sizes and allele frequencies are often missing.
+GWAS summary statistics have been extensively used to derive polygenic scores (PGS), perform fine-mapping, and estimate a range of key genetic architecture parameters.
+However, GWAS summary statistics come with uneven imputation accuracy.
+There is also heterogeneity in the information made available in these summary statistics; for example, per-variant imputation INFO scores, sample sizes and allele frequencies are often missing.
 Moreover, many methods based on summary statistics use Bayesian models, which can be sensitive to model misspecifications \cite[]{WALKER20131621,miller2018robust}.
@@
 We present an overview of possible misspecifications that come with GWAS summary statistics (Table \ref{tab:overview}). 
-First, the total sample size used can be misestimated, which would result e.g.\ in a biased estimation of the SNP heritability.
+First, the total sample size used can be misestimated, which would result in a biased estimation of the SNP heritability.
 For example, the total effective sample size is overestimated when computed from the total number of cases and controls from a meta-analysis of binary outcomes \cite[]{grotzinger2021pervasive}; using BOLT-LMM summary statistics can result in an increased effective sample size \cite[]{loh2018mixed,gazal2018functional}; using SAIGE on binary traits with a large prevalence can result in a reduced effective sample size \cite[]{zhou2018efficiently,mbatchou2021computationally}.
-Second, per-variant sample sizes can be very different when meta-analyzing GWAS summary statistics from multiple cohorts with different sets of variants \cite[]{wang2021global}.
-Per-variant sample sizes should be used instead of the overall total sample size to avoid any bias resulting from overestimating the sample size.
-Moreover, using very different per-variant sample sizes could be problematic for models implicitly assuming that summary statistics have all been derived from the same individuals \cite[]{zhu2017bayesian,zhou2021fast}.
+Second, per-variant sample sizes can vary substantially and be much smaller than the total sample size when meta-analyzing GWAS summary statistics from multiple cohorts with different sets of variants \cite[]{wang2021global}.
+Using very different per-variant sample sizes can be problematic for models implicitly assuming that summary statistics have all been derived from the same individuals \cite[]{zhu2017bayesian,zhou2021fast}.
 As a quality control (QC) step, one can e.g.\ remove SNPs with a sample size less than 0.67 times the 90th percentile of all sample sizes \cite[]{zheng2017ld}.
 Third, many genetic variants are imputed, which can cause some bias in summary statistics.
-Fourth, the quality of imputation of each variant, often used in a QC step in statistical genetics analyses, can be severely misestimated (often overestimated) when computed for multi-ancestry individuals instead of a more homogeneous subset.
+Fourth, the imputation quality of each variant (e.g.\ INFO scores), often used in a QC step in statistical genetics analyses, can be severely misestimated (often overestimated) when computed for multi-ancestry individuals instead of a more homogeneous subset.
 Fifth, errors such as allele inversions can be present in the summary statistics; QC is particularly important here.
@@
 This was particularly important for LDpred2-auto, which directly estimates key model parameters from the data.
-Here, we show e.g.\ that standard deviations of imputed genotypes (allele dosages) are lower than the expected values under Hardy--Weinberg equilibrium.
+Here we further show that standard deviations of imputed genotypes (allele dosages) are lower than the expected values under Hardy--Weinberg equilibrium.
 Second, we investigate possible adjustments to apply to the input parameters of methods using summary statistics (focusing on PGS methods here), namely the reference LD (linkage disequilibrium) matrix, the GWAS effect sizes, their standard errors and the corresponding sample sizes.
@@
 
-We design simulations where variants have different GWAS sample sizes, which is often the case when meta-analyzing GWAS summary statistics from multiple cohorts with different sets of variants \cite[]{wang2021global}.
+We design GWAS simulations where variants have different sample sizes, which is often the case when meta-analyzing GWAS summary statistics from multiple cohorts with different sets of variants \cite[]{wang2021global}.
 Using 40,000 variants from chromosome 22 (Methods), we simulate quantitative phenotypes with a heritability of 20\% and 2000 causal variants.
@@
 However, results for SBayesR always diverged (stopped with an error), and the overlap with the LD reference provided for PRS-CS was too small.
-Averaged over 10 simulations, when providing true per-variant GWAS sample sizes, squared correlations between the polygenic scores and the simulated phenotypes are of 0.123 for C+T, 0.161 for lassosum, 0.169 for lassosum2, 0.159 for LDpred2(-grid), 0.140 for LDpred2-auto, and 0.141 for LDpred2-inf (Figure \ref{fig:simu-misN}).
+When providing true per-variant GWAS sample sizes, squared correlations between the polygenic scores and the simulated phenotypes are of 0.123 for C+T, 0.161 for lassosum, 0.169 for lassosum2, 0.159 for LDpred2(-grid), 0.140 for LDpred2-auto, and 0.141 for LDpred2-inf (Figure \ref{fig:simu-misN}, averaging over 10 simulations).
 Results when using imputed sample sizes are quite similar to when using the true ones.
@@
 In addition, we design an alternative reference panel based on 10,000 individuals from South Europe by using the ``Italy'' center defined in \cite{prive2021high}.
-Averaged over 10 simulations, when using an ``alternative'' LD reference panel instead of a well-matched one, squared correlations between the polygenic scores and the simulated phenotypes drop from 0.174 to 0.167 for lassosum2, from 0.168 to 0.147 for LDpred2(-grid), from 0.174 to 0.169 for LDpred2-low-h2, from 0.143 to 0.139 for LDpred2-auto, from 0.162 to 0.146 for LDpred2-auto-rob, and from 0.144 to 0.140 for LDpred2-inf (Figure \ref{fig:simu-altpop}).
+When using an ``alternative'' LD reference panel instead of a well-matched one, squared correlations between the polygenic scores and the simulated phenotypes drop from 0.174 to 0.167 for lassosum2, from 0.168 to 0.147 for LDpred2(-grid), from 0.174 to 0.169 for LDpred2-low-h2, from 0.143 to 0.139 for LDpred2-auto, from 0.162 to 0.146 for LDpred2-auto-rob, and from 0.144 to 0.140 for LDpred2-inf (Figure \ref{fig:simu-altpop}, averaging over 10 simulations).
 Defining LD blocks in these two LD matrices always helps, but do not improve performance by much when using the alternative LD.
@@
 Imputation INFO scores for the OncoArray summary statistics (after having restricted to HapMap3 variants) are generally very good (Figure \ref{fig:hist-info-brca}) and better than the ones from iCOGS (Figure \ref{fig:hist-info-brca2}), probably because the iCOGS chip included around 200K variants only, compared to more than 500K variants for the OncoArray.
-For both summary statistics, we first compare the standard deviations (of genotypes) inferred from the reported allele frequencies (i.e.\ $\sqrt{2 f (1 - f)}$ where $f$ is the allele frequency, and denoted as sd\_af) versus the ones inferred from the GWAS summary statistics (Equation \eqref{eq:approx-sd-log}, and denoted as sd\_ss).
-When coloring by INFO scores, we see a clear trend with sd\_ss being lower than sd\_af as INFO decreases; indeed, using $\text{sd\_ss} / \sqrt{\text{INFO}}$ provides a very good fit for sd\_af, except for some variants of chromosome 6 and 8 for the OncoArray summary statistics (Figures \ref{fig:brca_onco_qc} and \ref{fig:brca_icogs_qc}).
+For both summary statistics, we compare the standard deviations (of genotypes) inferred from the reported allele frequencies (i.e.\ $\sqrt{2 f (1 - f)}$ where $f$ is the allele frequency, and denoted as sd\_af) versus the ones inferred from the GWAS summary statistics (Equation \eqref{eq:approx-sd-log}, and denoted as sd\_ss).
+As shown in figures \ref{fig:brca_onco_qc} and \ref{fig:brca_icogs_qc}, there is a clear trend with sd\_ss being lower than sd\_af as INFO decreases; indeed, using $\text{sd\_ss} / \sqrt{\text{INFO}}$ provides a very good fit for sd\_af, except for some variants of chromosome 6 and 8 for the OncoArray summary statistics.
 Most of these outlier variants are in regions 25-33 Mbp of chromosome 6 and 8-12 Mbp of chromosome 8 (Figure \ref{fig:hist-bad-pos}), which are two known long-range LD regions \cite[]{price2008long}.
@@
 We then also filter out the two long-range LD regions of chromosome 6 and 8 for the OncoArray summary statistics and remove around 500 variants when filtering on differences of MAFs between summary statistics and the validation dataset (``qc2'').
-As for the correction for the INFO scores, we use the first correction, ``sqrt\_info'', which is simple because it is equivalent to post-processing PGS effects by multiplying them by $\sqrt{\text{INFO}}$.
-Although results for both QC used are very similar, correcting for the INFO score slightly improves prediction when deriving polygenic scores based on iCOGS summary statistics (Figure \ref{fig:res_brca}).
+As for the correction using the INFO scores, we use the first correction, ``sqrt\_info'', which is simple because it is equivalent to post-processing PGS effects by multiplying them by $\sqrt{\text{INFO}}$.
+Although results for both QC used are very similar, correcting using the INFO scores slightly improves prediction when deriving polygenic scores based on iCOGS summary statistics (Figure \ref{fig:res_brca}).
 All other improvements introduced before have little to no effect here, probably because misspecifications are much smaller than in the simulations.
@@
 Note that, in this study, we use summary statistics that include extended information (e.g.\ INFO scores and allele frequencies), yet most GWAS summary statistics do not provide such exhaustive information \cite[]{macarthur2021workshop}.
-We acknowledge that, in the case of a meta-analysis from multiple studies, providing a single INFO score per variant may not be possible; would a weighted averaged INFO score work?
+We acknowledge that, in the case of a meta-analysis from multiple studies, providing a single INFO score per variant may not be possible.
+Solutions such as using a weighted averaged INFO score might be worth exploring in future studies.
 Nevertheless, this quality control could be performed within each study before pooling results, to make sure that summary statistics have the best possible quality for follow-up analyses such as deriving polygenic scores.
@@
 We think this is made possible by the use of a (very) strong regularization in PRS-CS ($\phi^{-1} \psi_j^{-1} \ge 1$, which would approximately correspond to using $s=0.5$ in lassosum, $\delta = 1$ in lassosum2, and $\texttt{shrink\_corr}=0.5$ in LDpred2-auto).
-Using enough regularization is good for robustness, but note that using too much regularization can also damage predictive performance, and should not be a replacement for a proper QC and choice of ancestry-matched LD [TODO: A BIT AGGRESSIVE?].
+Using enough regularization is good for robustness, but note that using too much regularization can also damage predictive performance.
+To address this limitation, we recommend to rely on a proper QC and choice of ancestry-matched LD instead of on too much regularization.
 As a future research direction, we are interested in using ancestry-matched LD matrices to improve polygenic prediction in all ancestries.
-Also it would be useful to investigate the impact of the QC and adjustments we propose here on other (non-PGS) methods, such as the ones for inference, where robustness is likely to be very important as well.
+Also it would be useful to investigate the impact of the QC and adjustments we propose here on other (non-PGS) methods, where robustness is likely to be very important as well.
 

### 2022-02-23 paper/paper-misspec-supp.tex
reexport code reread paper

@@
 	\caption{Minor allele frequencies (MAF) and INFO scores in the UK Biobank for 50,000 variants on chromosome 22 (chosen at random).
-These are either reported (in UKBB MFI files), recomputed from the whole data or from the subset of 362,307 European individuals used in this paper.
+These are either reported (in MFI files from the UK Biobank), recomputed from the whole data or from the subset of 362,307 European individuals used in this paper.
+MAF are represented on a log scale.
 \label{fig:compare-maf-info}}
@@
 	\centerline{\includegraphics[width=0.95\textwidth]{compare-beta}}
-	\caption{GWAS effect sizes ($\gamma$) from ``true'' genotype calls versus either \textbf{A:} raw $\gamma$ from imputed dosages or \textbf{B:} corrected $\gamma$ from imputed dosages (multiplying them by $\sqrt{\text{INFO}}$), colored by INFO scores.
+	\caption{GWAS effect sizes ($\hat{\gamma}$) from ``true'' genotype calls versus either \textbf{A:} raw $\hat{\gamma}$ from imputed dosages or \textbf{B:} corrected $\hat{\gamma}$ from imputed dosages (multiplying them by $\sqrt{\text{INFO}}$), colored by INFO scores.
 \label{fig:compare-beta}}

### 2022-02-23 paper/paper-misspec.tex
reexport code reread paper

@@
 
-\title{Identifying and correcting for misspecifications\\in GWAS summary statistics for polygenic scores}
+\title{Identifying and correcting for misspecifications\\in GWAS summary statistics and polygenic scores}
 \author{Florian Priv\'e,$^{\text{1,}*}$ Julyan Arbel,$^{\text{2}}$ Hugues Aschard,$^{\text{3,4}}$ and Bjarni J. Vilhj\'almsson$^{\text{1,5}}$}
@@
 Can we detect possible misspecifications in GWAS summary statistics and correct for them in order to improve predictive performance of polygenic scores?
-	[TODO: LIST SOME OF THEM (AND POSSIBLE CONSEQUENCES?)]
 }
@@
 Contrary to individual-level genotypes and phenotypes, summary statistics resulting from genome-wide association studies (GWAS) are widely available, and very large sample sizes can be obtained through meta-analysis \cite[]{yengo2022saturated}.
-GWAS summary statistics have been extensively used to derive polygenic scores (PGS), perform fine-mapping, and estimate a range of key genetic architecture parameters.
+GWAS summary statistics have been extensively used to derive polygenic scores (PGS), perform fine-mapping, and estimate a range of key genetic architecture parameters \cite[]{prive2021finding}.
 However, GWAS summary statistics come with uneven imputation accuracy.
@@
 Third, many genetic variants are imputed, which can cause some bias in summary statistics.
-Fourth, the imputation quality of each variant (e.g.\ INFO scores), often used in a QC step in statistical genetics analyses, can be severely misestimated (often overestimated) when computed for multi-ancestry individuals instead of a more homogeneous subset.
+Fourth, the imputation quality of each variant (e.g.\ INFO scores), often used in a QC step in statistical genetics analyses, can be severely misestimated (often overestimated) when computed from multi-ancestry individuals instead of a more homogeneous subset.
 Fifth, errors such as allele inversions can be present in the summary statistics; QC is particularly important here.
@@
 		\hline
-		Differences in per-variant sample size & Possible violation of model assumptions resulting in poor model convergence & Imputation using equations \ref{eq:imputeN} and \ref{eq:imputeNeff}, and filtering$^{(2)}$   \\
+		Differences in per-variant sample sizes & Possible violation of model assumptions resulting in poor model convergence & Imputation using equations \ref{eq:imputeN} and \ref{eq:imputeNeff}, and filtering$^{(2)}$   \\
 		\hline
@@
 		\hline 
-		Misestimation of imputation INFO scores & Misuse when filtering or adjusting input parameters based on these & Recomputing for homogeneous subset \\
+		Misestimation of imputation INFO scores & Misuse when filtering or adjusting input parameters based on these & Recomputing from a homogeneous subset \\
 		\hline
@@
 		\hline
-		Rounding of summary statistics (effect sizes and SEs) & Add unnecessary noise to the data & None, but this has almost no impact on the predictive performance (Figure \ref{fig:simu-rounded}) \\
+		Rounding of summary statistics (e.g.\ effect sizes and SEs) & Add unnecessary noise to the data & None, but this has almost no impact on the predictive performance (Figure \ref{fig:simu-rounded}) \\
 		\hline 
@@
 This is useful to check that the input parameters used are consistent with each other. 
-This was particularly important for LDpred2-auto, which directly estimates key model parameters from the data.
+This was particularly important for LDpred2-auto, which directly estimates key model parameters from the data \cite[]{prive2020ldpred2}.
 Here we further show that standard deviations of imputed genotypes (allele dosages) are lower than the expected values under Hardy--Weinberg equilibrium.
-Second, we investigate possible adjustments to apply to the input parameters of methods using summary statistics (focusing on PGS methods here), namely the reference LD (linkage disequilibrium) matrix, the GWAS effect sizes, their standard errors and the corresponding sample sizes.
-For example, we show that GWAS effect sizes computed from imputed dosages are larger in magnitude compared to if true genotypes were available.
+Second, we investigate possible adjustments to apply to the input parameters of methods using summary statistics (focusing on PGS methods here), namely the reference LD (linkage disequilibrium) matrix, the GWAS effect sizes, their standard errors and corresponding sample sizes.
+For example, we show that GWAS effect sizes computed from imputed dosages are larger in magnitude compared to if computed from true genotypes.
 Third, we introduce two new optional parameters in LDpred2-auto to make it more robust to these types of misspecification.
@@
 
-\subsection{Misspecification of GWAS sample sizes}
+\subsection{Misspecification of per-variant GWAS sample sizes}
 
@@
 We then divide the 40,000 variants into three groups: for half of the variants, we use 100\% of 300,000 individuals for GWAS, but use only 80\% for one quarter of the variants and 60\% for the remaining quarter.
-We then run C+T, lassosum, lassosum2 (Methods), LDpred2-inf, LDpred2(-grid), and LDpred2-auto \cite[]{prive2019making,mak2017polygenic,prive2020ldpred2} by using either the true per-variant GWAS sample sizes, the total sample size, or imputed sample sizes.
+We then run C+T, lassosum, lassosum2 (Methods), LDpred2-inf, LDpred2(-grid), and LDpred2-auto \cite[]{prive2019making,mak2017polygenic,prive2020ldpred2} by using either the true per-variant GWAS sample sizes, the total sample size for all variants, or imputed (per-variant) sample sizes.
 Note that we initially included PRS-CS and SBayesR in our comparison \cite[]{ge2019polygenic,lloyd2019improved}.
-However, results for SBayesR always diverged (stopped with an error), and the overlap with the LD reference provided for PRS-CS was too small.
+However, results for SBayesR always diverged (stopping with an error), and the overlap with the LD reference provided for PRS-CS was too small.
 When providing true per-variant GWAS sample sizes, squared correlations between the polygenic scores and the simulated phenotypes are of 0.123 for C+T, 0.161 for lassosum, 0.169 for lassosum2, 0.159 for LDpred2(-grid), 0.140 for LDpred2-auto, and 0.141 for LDpred2-inf (Figure \ref{fig:simu-misN}, averaging over 10 simulations).
@@
 When using the total GWAS sample size instead of the per-variant sample sizes, predictive performance slightly decreases to 0.157 for lassosum and to 0.163 for lassosum2, but dramatically decreases for LDpred2 with new values of 0.134 for LDpred2-grid, 0.119 for LDpred2-auto, and 0.123 for LDpred2-inf (Figure \ref{fig:simu-misN}).
-This extreme simulation scenario shows that LDpred2 can be sensitive to GWAS sample size misspecification, whereas lassosum (and lassosum2) seems little affected by this.
+This extreme simulation scenario shows that LDpred2 can be sensitive to the misspecification of per-variant GWAS sample sizes, whereas lassosum (and lassosum2) seems little affected by this.
 
@@
 Reported 95\% confidence intervals are computed from 10,000 non-parametric bootstrap replicates of the mean.
-The GWAS sample size is ``true'' when providing the true per-variant sample size, ``max'' when providing instead the maximum sample size as a unique value to be used for all variants,  ``imputed'' (Methods), or ``any'' when the method does not use this information (the case for C+T).
+The GWAS sample size is either ``true'' when providing the true per-variant sample size, ``max'' when providing instead the maximum sample size as a unique value to be used for all variants, ``imputed'' (Methods), or ``any'' when the method does not use this information (the case for C+T).
 Red bars correspond to using the LD with independent blocks (Methods).}
@@
 
-We conduct further investigations to explain results of figure \ref{fig:simu-misN}.
-First, the results for LDpred2-auto are the same as with LDpred2-inf because it always converges to an infinitesimal model ($p = 1$) in these simulations.
-To overcome this limitation, we introduce two new parameters to make LDpred2-auto more robust, referred to as ``LDpred2-auto-rob'' here (Methods).
+We conduct further investigations to explain the results of figure \ref{fig:simu-misN}.
+First, the results for LDpred2-auto are similar to LDpred2-inf because it always converges to an infinitesimal model ($p = 1$) in these simulations.
+To overcome this limitation, we introduce two new parameters to make LDpred2-auto more robust, and refer to this as ``LDpred2-auto-rob'' here (Methods).
 The first parameter prevents $p$ from diverging to $1$, while the second shrinks the off-diagonal elements of the LD matrix (a form of regularization). 
 Second, for lassosum2, results for a grid of parameters (over $\lambda$ and $\delta$) are quite smooth compared to LDpred2 (Figures \ref{fig:lassosum2-misN} and \ref{fig:ldpred2-misN}). 
-In these simulations with misspecified sample sizes, it seems highly beneficial to use a small value for the SNP heritability hyper-parameter $h^2$ in LDpred2, e.g.\ a value of 0.02 or even 0.002 when the true value is 0.2 (Figure \ref{fig:ldpred2-misN}).
+In these simulations with misspecified per-variant sample sizes, it seems highly beneficial to use a small value for the SNP heritability hyper-parameter $h^2$ in LDpred2, e.g.\ a value of 0.02 or even 0.002 when the true value is 0.2 (Figure \ref{fig:ldpred2-misN}).
 Indeed, using a small value for this hyper-parameter induces a larger regularization (shrinkage) on the effect sizes.
-When running LDpred2(-grid) with a grid of hyper-parameters including these low values for $h^2$, we refer to this as ``LDpred2-low-h2'' here.
+Here we call ``LDpred2-low-h2'' when running LDpred2(-grid) with a grid of hyper-parameters including these low values for $h^2$.
 Results with LDpred2-low-h2 improves from 0.159 to 0.169 when using true sample sizes and from 0.134 to 0.163 when using the maximum sample size.
@@
 \cite{marchini2010genotype} showed that the IMPUTE INFO measure is highly concordant with the MACH measure $\hat{r}_j^2 = \frac{\text{var}(G_j)}{2 \hat{\theta}_j (1 - \hat{\theta}_j)}$, where $\hat{\theta}_j$ is the estimated allele frequency of $G_j$, the genotypes for variant $j$.
-Therefore, when using the expected genotypes from imputation (dosages), their standard deviations are often lower than the expected value under Hardy--Weinberg equilibrium, because $\text{INFO}_j \approx \frac{\text{var}(G_j)}{2 \hat{\theta}_j (1 - \hat{\theta}_j)}$.
+Therefore, when using the expected genotypes from imputation (allele dosages), their standard deviations are often lower than the expected value under Hardy--Weinberg equilibrium, because $\text{INFO}_j \approx \frac{\text{var}(G_j)}{2 \hat{\theta}_j (1 - \hat{\theta}_j)}$.
 In simulations (cf.\ Methods section ``Data for simulations''), we verify that $\text{sd}(G_j)^{\text{true}} \approx \text{sd}(G_j)^{\text{imp}} / \sqrt{\text{INFO}_j}$ (Figure \ref{fig:compare-sd}).
@@
 This is the first correction of summary statistics we consider in the simulations below. 
-Note that we recompute INFO scores for the subset of 362,307 European individuals used in this paper as they can differ substantially from the ones reported by the UK Biobank for the whole data (Figures \ref{fig:compare-info} and \ref{fig:compare-maf-info}).
+Note that we recompute INFO scores for the subset of 362,307 European individuals used in this paper since they can differ substantially from the ones reported by the UK Biobank for the whole data (Figures \ref{fig:compare-info} and \ref{fig:compare-maf-info}).
 As a second option, instead of using dosages to compute the GWAS summary statistics, it has been argued that using multiple imputation (MI) would be more appropriate \cite[]{Palmer2016}.
 In simulations, we show that $\hat{\gamma}_j^{\text{MI}} \approx \hat{\gamma}_j^{\text{imp}} \cdot \text{INFO}_j$ and $Z_j^{\text{MI}} \approx Z_j^{\text{imp}} \cdot \text{INFO}_j$, where $Z = \hat{\gamma} / \text{se}(\hat{\gamma})$ (Figure \ref{fig:compare-mi}).
-This is the second correction of summary statistics we implement in the simulations below, along with $n_j = N \cdot \text{INFO}_j$ where $n_j$ and $N$ are the per-variant and total sample sizes. 
-Finally, we consider an in-between solution as a third correction, using $\hat{\gamma}_j = \hat{\gamma}_j^{\text{imp}} \cdot \text{INFO}_j$, $\text{se}(\hat{\gamma}_j) = \text{se}(\hat{\gamma}_j)^{\text{imp}} \cdot \sqrt{\text{INFO}_j}$, and $n_j = N \cdot \text{INFO}_j$.
+This is the second correction of summary statistics we implement in the simulations below, along with $n_j \cdot \text{INFO}_j$ as the new per-variant sample sizes. 
+Finally, we consider an in-between solution as a third correction, using $\hat{\gamma}_j = \hat{\gamma}_j^{\text{imp}} \cdot \text{INFO}_j$, $\text{se}(\hat{\gamma}_j) = \text{se}(\hat{\gamma}_j)^{\text{imp}} \cdot \sqrt{\text{INFO}_j}$, and $n_j \cdot \text{INFO}_j$.
 
@@
 Correction ``sqrt\_info'' corresponds to using $\hat{\gamma}_j^{\text{imp}} \cdot \sqrt{\text{INFO}_j}$ and $\text{se}(\hat{\gamma}_j)^{\text{imp}} \cdot \sqrt{\text{INFO}_j}$.
-	Correction ``info'' corresponds to using $\hat{\gamma}_j^{\text{imp}} \cdot \text{INFO}_j$  and $N \cdot \text{INFO}_j$.
-Correction ``in\_between'' corresponds to using $\hat{\gamma}_j^{\text{imp}} \cdot \text{INFO}_j$, $\text{se}(\hat{\gamma}_j)^{\text{imp}} \cdot \sqrt{\text{INFO}_j}$, and $N \cdot \text{INFO}_j$.
+	Correction ``info'' corresponds to using $\hat{\gamma}_j^{\text{imp}} \cdot \text{INFO}_j$  and $n_j \cdot \text{INFO}_j$.
+Correction ``in\_between'' corresponds to using $\hat{\gamma}_j^{\text{imp}} \cdot \text{INFO}_j$, $\text{se}(\hat{\gamma}_j)^{\text{imp}} \cdot \sqrt{\text{INFO}_j}$, and $n_j \cdot \text{INFO}_j$.
 Red bars correspond to using the LD with independent blocks (Methods).}
@@
 In addition, we design an alternative reference panel based on 10,000 individuals from South Europe by using the ``Italy'' center defined in \cite{prive2021high}.
-When using an ``alternative'' LD reference panel instead of a well-matched one, squared correlations between the polygenic scores and the simulated phenotypes drop from 0.174 to 0.167 for lassosum2, from 0.168 to 0.147 for LDpred2(-grid), from 0.174 to 0.169 for LDpred2-low-h2, from 0.143 to 0.139 for LDpred2-auto, from 0.162 to 0.146 for LDpred2-auto-rob, and from 0.144 to 0.140 for LDpred2-inf (Figure \ref{fig:simu-altpop}, averaging over 10 simulations).
-Defining LD blocks in these two LD matrices always helps, but do not improve performance by much when using the alternative LD.
+When using this alternative LD reference panel instead of a well-matched one as in the previous sections, squared correlations between the polygenic scores and the simulated phenotypes drop from 0.174 to 0.167 for lassosum2, from 0.168 to 0.147 for LDpred2(-grid), from 0.174 to 0.169 for LDpred2-low-h2, from 0.143 to 0.139 for LDpred2-auto, from 0.162 to 0.146 for LDpred2-auto-rob, and from 0.144 to 0.140 for LDpred2-inf (Figure \ref{fig:simu-altpop}, averaging over 10 simulations).
+Forming independent LD blocks in these two LD matrices always helps, but do not improve performance by much when using the alternative LD.
 
@@
 
-Breast cancer summary statistics are interesting because they include results from two mega analyses \cite[]{michailidou2013large,michailidou2015genome,michailidou2017association}, which means that parameters reported in the GWAS summary statistics, such as the INFO scores and the sample sizes, are estimated with high precision.
-Imputation INFO scores for the OncoArray summary statistics (after having restricted to HapMap3 variants) are generally very good (Figure \ref{fig:hist-info-brca}) and better than the ones from iCOGS (Figure \ref{fig:hist-info-brca2}), probably because the iCOGS chip included around 200K variants only, compared to more than 500K variants for the OncoArray.
+Breast cancer summary statistics are interesting because they include results from two mega analyses \cite[]{michailidou2013large,michailidou2015genome,michailidou2017association}, which means that parameters reported in these GWAS summary statistics, such as INFO scores and sample sizes, are estimated with high precision.
+Imputation INFO scores for the OncoArray summary statistics are generally very good (mean of 0.968 after having restricted to HapMap3 variants, Figure \ref{fig:hist-info-brca}) and better than the ones from iCOGS (mean of 0.841, Figure \ref{fig:hist-info-brca2}), probably because the iCOGS chip included around 200K variants only, compared to more than 500K variants for the OncoArray.
 For both summary statistics, we compare the standard deviations (of genotypes) inferred from the reported allele frequencies (i.e.\ $\sqrt{2 f (1 - f)}$ where $f$ is the allele frequency, and denoted as sd\_af) versus the ones inferred from the GWAS summary statistics (Equation \eqref{eq:approx-sd-log}, and denoted as sd\_ss).
 As shown in figures \ref{fig:brca_onco_qc} and \ref{fig:brca_icogs_qc}, there is a clear trend with sd\_ss being lower than sd\_af as INFO decreases; indeed, using $\text{sd\_ss} / \sqrt{\text{INFO}}$ provides a very good fit for sd\_af, except for some variants of chromosome 6 and 8 for the OncoArray summary statistics.
-Most of these outlier variants are in regions 25-33 Mbp of chromosome 6 and 8-12 Mbp of chromosome 8 (Figure \ref{fig:hist-bad-pos}), which are two known long-range LD regions \cite[]{price2008long}.
-We hypothesize that this is due to using (as covariates in the GWAS) principal components (PCs)  that capture LD structure instead of population structure \cite[]{prive2020efficient}.
+Most of these outlier variants are either in region 25-33 Mbp of chromosome 6 or in 8-12 Mbp of chromosome 8 (Figure \ref{fig:hist-bad-pos}), which are two known long-range LD regions \cite[]{price2008long}.
+We hypothesize that this is due to using principal components (PCs) that capture LD structure instead of population structure, as covariates in the GWAS \cite[]{prive2020efficient}.
 To validate this hypothesis, we simulate a phenotype using HapMap3 variants of chromosome 6 for 10,000 individuals from the UK Biobank, then we run GWAS with or without PC19 as covariate.
@@
 First, it allows for correcting for a reduced standard deviation when using imputed data in the QC step we propose, in order to better uncover problems with the summary statistics.
-Second, using one of the proposed corrections based on the INFO scores may lead to an improved prediction when deriving polygenic scores.
-We apply this correction to the two breast cancer summary statistics.
-We first compare the standard QC proposed in \cite{prive2020ldpred2} (which ends up filtering on MAF here, which we call ``qc1'').
+Second, using one of the proposed corrections based on INFO scores may lead to an improved prediction when deriving polygenic scores.
+We apply these corrections to the two breast cancer summary statistics. 
+In the comparison, we use the QC proposed in \cite{prive2020ldpred2} first (which ends up filtering on MAF here, which we call ``qc1'').
 We then also filter out the two long-range LD regions of chromosome 6 and 8 for the OncoArray summary statistics and remove around 500 variants when filtering on differences of MAFs between summary statistics and the validation dataset (``qc2'').
-As for the correction using the INFO scores, we use the first correction, ``sqrt\_info'', which is simple because it is equivalent to post-processing PGS effects by multiplying them by $\sqrt{\text{INFO}}$.
-Although results for both QC used are very similar, correcting using the INFO scores slightly improves prediction when deriving polygenic scores based on iCOGS summary statistics (Figure \ref{fig:res_brca}).
+As for the correction using INFO scores, we use the first correction, ``sqrt\_info'', which is simple because it is equivalent to post-processing PGS effects by multiplying them by $\sqrt{\text{INFO}}$.
+Although results for both QC used are very similar, correcting using INFO scores slightly improves prediction when deriving polygenic scores based on iCOGS summary statistics (Figure \ref{fig:res_brca}).
 All other improvements introduced before have little to no effect here, probably because misspecifications are much smaller than in the simulations.
@@
 
-We use external summary statistics for which INFO scores are reported (Table \ref{tab:sumstats}); they all have a very high mean INFO score (larger than 0.94), except for BRCA-iCOGS (0.841) and T1D-affy (0.885).
+We use other external GWAS summary statistics for which INFO scores are reported (Table \ref{tab:sumstats}); they all have a very high mean INFO score (larger than 0.94), except for T1D-affy (0.885).
 QC plots comparing standard deviations usually show little deviation from the identity line (after the INFO score correction), except for coronary artery disease (CAD) summary statistics (Figures \ref{fig:brca_icogs_qc}-\ref{fig:t1d_illu_qc}).
-We show previous results as figure \ref{fig:res_brca} for other phenotypes in figures \ref{fig:res_cad}-\ref{fig:res_t1d}; most changes introduced before have little to no impact on the predictions, except for the second quality control performed for CAD when using LDpred2-auto (Figure \ref{fig:res_cad}).
+In figures \ref{fig:res_cad}-\ref{fig:res_t1d}, we then provide similar results as figure \ref{fig:res_brca} for other phenotypes.
+Most changes introduced before have little to no impact on the predictions, except for the second quality control performed for CAD when using LDpred2-auto (Figure \ref{fig:res_cad}).
 
@@
 Previously, we proposed a quality control (QC) based on comparing standard deviations (of genotypes) inferred from GWAS summary statistics with the ones computed from a reference panel \cite[]{prive2020ldpred2}.
-Here we show that we can refine this QC by deriving the latter directly from the reported allele frequencies in the GWAS summary statistics, and by correcting the former using the GWAS imputation INFO scores (e.g.\ see figure \ref{fig:brca_onco_qc}).
-Using this refined QC, we are able to identify a potential issue with how principal components were derived in a set of breast cancer summary statistics.
+Here we show that we can refine this QC by deriving the latter directly from the reported allele frequencies in the GWAS summary statistics, and by correcting the former using imputation INFO scores.
+Using this refined QC, we are able to identify a potential issue with how principal components were derived in a GWAS of breast cancer.
 Fortunately, this has practically no effect on the predictive performance of the derived polygenic scores.
 Additional QC can also be performed, e.g.\ comparing reported GWAS allele frequencies with the ones from the LD reference panel, e.g.\ to detect genotyping errors or allele inversions.
-We do perform this additional QC as part of ``qc2'' here. 
+We perform this additional QC as part of ``qc2'' here. 
 One can also run other QC tools such as DENTIST \cite[]{chen2021improved}, and also infer ancestry proportions from summary statistics to make sure these are matching with the LD reference used \cite[]{prive2021using}.
 
-Note that, in this study, we use summary statistics that include extended information (e.g.\ INFO scores and allele frequencies), yet most GWAS summary statistics do not provide such exhaustive information \cite[]{macarthur2021workshop}.
+Note that, in this study, we use GWAS summary statistics that include extended information (e.g.\ INFO scores and allele frequencies), yet most GWAS summary statistics do not provide such exhaustive information \cite[]{macarthur2021workshop}.
 We acknowledge that, in the case of a meta-analysis from multiple studies, providing a single INFO score per variant may not be possible.
 Solutions such as using a weighted averaged INFO score might be worth exploring in future studies.
-Nevertheless, this quality control could be performed within each study before pooling results, to make sure that summary statistics have the best possible quality for follow-up analyses such as deriving polygenic scores.
+Nevertheless, this quality control could be performed within each study before meta-analyzing results, to make sure that resulting summary statistics have the best possible quality for follow-up analyses such as deriving polygenic scores.
 Another information, the effective sample size per variant, is often missing from GWAS summary statistics. 
@@
 Indeed, take the extreme example of meta-analyzing two studies, one with 1000 cases and 0 controls, and another one with 0 cases and 1000 controls, then the effective sample size of the meta-analysis is 0, not 2000.
-This misspecification can lead to serious issues such as misestimating the SNP heritability \cite[]{grotzinger2021pervasive}.
+Misspecifying the GWAS sample size can lead to serious issues such as misestimating the SNP heritability \cite[]{grotzinger2021pervasive}.
 Fortunately, an overestimated sample size can be detected from the QC plot we propose, where the slope is then less than 1 for case-control studies using logistic regression; otherwise the standard deviation of the phenotype is also needed (Equation \ref{eq:approx-sd-lin}), but can be estimated \cite[]{prive2020ldpred2}.
@@
 Using both the Bayesian LDpred2 models \cite[]{prive2020ldpred2} and our reimplementation of the frequentist lassosum model \cite[]{mak2017polygenic} for deriving polygenic scores, we have introduced and investigated some changes to possibly make these models more robust to misspecifications.
-Overall, these changes have provided large improvements of predictive performance in simulations with large misspecifications.
-However, they have almost no effect when using the real GWAS summary statistics we chose, except for breast cancer and coronary artery disease summary statistics.
+Overall, these changes provided large improvements of predictive performance in the simulations with large misspecifications.
+However, they had little to no effect when using the real GWAS summary statistics we chose, except for breast cancer and coronary artery disease summary statistics.
 Although these results are somewhat unfortunate, they are reassuring because it means that the summary statistics we use in this study are usually of good quality for follow-up analyses such as deriving polygenic scores.
 
-Nevertheless, we recommend to adopt these changes, i.e.\ performing the (refined) QC proposed here, defining LD blocks in the LD matrix, and using more regularization when needed (i.e.\ also testing smaller values for the heritability parameter in LDpred2-grid, and using the two new parameters in LDpred2-auto (Section \ref{sec:robust}).
-Note that LD blocks are already widely used by several methods (e.g.\ lassosum and PRS-CS) because they allow for processing smaller matrices at once \cite[]{mak2017polygenic,ge2019polygenic}; here we have shown that these blocks are also useful to make methods more robust.
+Nevertheless, we recommend to adopt these changes, i.e.\ performing the (refined) QC proposed here, forming independent LD blocks in the LD matrix, and using more regularization when needed. 
+More regularization can be achieved by testing additional smaller values for the heritability parameter in LDpred2-grid, and using the two new parameters introduced here in LDpred2-auto (Section \ref{sec:robust}).
+Note that LD blocks are already widely used by several methods (such as lassosum and PRS-CS) because they allow for processing smaller matrices at once \cite[]{mak2017polygenic,ge2019polygenic}; here we have shown that these blocks are also useful to make methods more robust.
 PRS-CS is currently one of the most robust PGS methods; for example, it can use the (small) 1000 Genomes data as LD reference \cite[]{ge2019polygenic}, and can even use a European LD reference panel with multi-ancestry GWAS summary statistics \cite[]{wang2021global}.
@@
 To address this limitation, we recommend to rely on a proper QC and choice of ancestry-matched LD instead of on too much regularization.
-As a future research direction, we are interested in using ancestry-matched LD matrices to improve polygenic prediction in all ancestries.
-Also it would be useful to investigate the impact of the QC and adjustments we propose here on other (non-PGS) methods, where robustness is likely to be very important as well.
+As a future research direction, we are interested in using multi-ancestry-matched LD matrices to improve polygenic prediction in all ancestries.
+It would also be useful to investigate the impact of the QC and adjustments we propose here on other (non-PGS) methods, for which consistency and robustness are likely to be very important as well \cite[]{chen2021improved}.
 
@@
 We sample 10,000 individuals to form a validation set and use the remaining 352,307 individuals as test set.
-We restrict to the genetic variants to the 1,054,315 HapMap3 variants used in the LD reference provided in \cite{prive2020ldpred2}.
+We restrict to the 1,054,315 HapMap3 variants used in the LD reference provided in \cite{prive2020ldpred2}.
 
@@
 	\end{tabular}
-	\caption{Summary of external GWAS summary statistics used. PRCA summary statistics have many variants with a missing INFO score, which we discard.\label{tab:sumstats}}
+	\caption{Summary of external GWAS summary statistics used.
+PRCA summary statistics have many variants with a missing INFO score, which we discard.
+We also restrict to variants with an INFO score larger than 0.4.\label{tab:sumstats}}
 \end{table}
@@
 \end{equation}
-where $\hat{\gamma}_j$ is the marginal (GWAS) effect of variant $j$, $n_j$ is the GWAS sample size associated with variant $j$, $y$ is the vector of phenotypes and ${G}_j$ is the vector of genotypes for variant $j$.
+where $\hat{\gamma}_j$ is the marginal GWAS effect size of variant $j$, $n_j$ is the GWAS sample size associated with variant $j$, $y$ is the vector of phenotypes and ${G}_j$ is the vector of genotypes for variant $j$.
 This formula is used in LDpred2 \cite{prive2020ldpred2,prive2021high}. Note that, for a binary trait for which logistic regression is used, we have instead
@@
 Then computing $R_{.,j}^T \beta^{(t-1)}$ is the most time-consuming part of each iteration.
-To make this faster, instead of computing $R_{.,j}^T \beta^{(t-1)}$ at each iteration ($j$ and $t$), we can start with an initial vector of 0s only (for all $j$) since $\beta^{(0)} \equiv 0$, and then updating this vector when $\beta_j^{(t)} \neq \beta_j^{(t-1)}$ only.
+To make this faster, instead of computing $R_{.,j}^T \beta^{(t-1)}$ at each iteration ($j$ and $t$), we can start with an initial vector of 0s only (for all $j$) since $\beta^{(0)} \equiv 0$, then update this vector when $\beta_j^{(t)} \neq \beta_j^{(t-1)}$ only.
 Note that only positions $k$ for which $R_{k,j} \neq 0$ must be updated in this vector $R_{.,j}^T \beta^{(t-1)}$.
 
-In this new implementation of the lassosum model, the input parameters are the correlation matrix $R$, the GWAS summary statistics ($\widehat{\gamma}_j$, $\text{se}(\widehat{\gamma}_j)$ and $n_j$), and the two hyper-parameters $\lambda$ and $\delta$.
-Therefore, except for the hyper-parameters, this is the exact same input as for LDpred2 \cite[]{prive2020ldpred2}.
+In this new implementation of the lassosum model, which we call lassosum2, the input parameters are the correlation matrix $R$, the GWAS summary statistics ($\widehat{\gamma}_j$, $\text{se}(\widehat{\gamma}_j)$ and $n_j$), and the two hyper-parameters $\lambda$ and $\delta$.
+Therefore, except for the two hyper-parameters, lassosum2 uses the exact same input parameters as LDpred2 \cite[]{prive2020ldpred2}.
 We try $\delta \in \{0.001, 0.005, 0.02, 0.1, 0.6, 3\}$ by default in lassosum2, instead of $s \in \{0.2, 0.5, 0.8, 1.0\}$ in lassosum.
@@
 Note that we do not provide an ``auto'' version using pseudo-validation (as in \cite{mak2017polygenic}) as we have not found it to be robust enough (Figure \ref{fig:pseudoval}).
-Also note that, as in LDpred2, we run lassosum2 genome-wide using a sparse correlation matrix which assumes that variants further away than 3 cM are not correlated, and therefore we do not require splitting the genome into independent LD blocks anymore (as done in lassosum).
+Also note that, as in LDpred2, we run lassosum2 genome-wide using a sparse correlation matrix which assumes that variants further away than 3 cM are not correlated.
+Although we do not require splitting the genome into independent LD blocks anymore (as done in lassosum), we recommend to do so for robustness in LDpred2, and for extra speed gain (cf.\ next section).
 
@@
 
-We define nearly independent LD blocks using the optimal algorithm developed in \cite{prive2021optimal}.
-For different numbers of blocks and maximum number of variants in each block, we use the split with the minimum cost within the ones reducing the original number of non-zero values to less than 60\% (or 70\% for chromosome 6).
+We form nearly independent LD blocks using the optimal algorithm developed in \cite{prive2021optimal}.
+For different numbers of blocks and maximum number of variants in each block, we use the split with the minimum cost within the ones reducing the original number of non-zero values to less than 60\% (70\% for chromosome 6).
 Having a correlation matrix with independent blocks prevents small errors in the algorithm (e.g.\ the Gibbs sampler in LDpred2) from propagating to too many variants.
@@
 Instead of using something similar to the standard ``compressed sparse column'' format which stores all $\{i,~x(i, j)\}$ for a given column $j$, we only store the first index $i_0$ and all the contiguous values $\{x(i_0, j),~x(i_0 + 1, j),~\dots\}$ up to the last non-zero value for this column $j$.
-This makes this format about twice as efficient for both LDpred2 and lassosum2.
+This makes this format about twice as efficient for both LDpred2 and lassosum2 (in terms of both memory and speed).
+Therefore, using both this new format and the LD blocks should divide computation times by 3.
 
@@
 We use a value of 0.9 in simulations and 0.95 in real data when running ``LDpred2-auto-rob'' (and the default value of 1, with no effect, when running ``LDpred2-auto'').
-The second new parameter, \texttt{allow\_jump\_sign}, controls whether variants can change sign over two consecutive iterations of the Gibbs sampler.
-When setting this parameter to false (in the method we name ``LDpred2-auto-rob'' here), this forces the effects to go through 0 before changing sign.
-This is useful to prevent instability (oscillation and ultimately divergence) of the Gibbs sampler under large misspecifications, and is also useful for accelerating convergence of chains with a large initial value for $p$, the proportion of causal variants.
+The second new parameter, \texttt{allow\_jump\_sign}, controls whether variant effect sizes can change sign over two consecutive iterations of the Gibbs sampler.
+When setting this parameter to false (in the method we refer to as ``LDpred2-auto-rob'' here), this forces the effects to go through 0 before changing sign.
+This is useful to prevent instability (oscillation and ultimately divergence) of the Gibbs sampler under large misspecifications, and is also useful for accelerating convergence of chains with a large initial value for $p$ (the proportion of causal variants).
 
@@
 
-\section*{Code availability}
+\section*{Code and data availability}
 
 All code used for this paper is available at \url{https://github.com/privefl/paper-misspec/tree/master/code}.
-We have recomputed the allele frequencies and imputation INFO scores within the European subset used here and across all variants in the UK Biobank, and made them available at \url{https://doi.org/10.6084/m9.figshare.16635388.v1}.
-The latest version of R package bigsnpr can be installed from GitHub.
+We have extensively used R packages bigstatsr and bigsnpr \cite[]{prive2017efficient} for analyzing large genetic data, packages from the future framework \cite[]{bengtsson2020unifying} for easy scheduling and parallelization of analyses on the HPC cluster, and packages from the tidyverse suite \cite[]{wickham2019welcome} for shaping and visualizing results.
+
+The latest version of R package bigsnpr can be installed from GitHub, and a recent enough version from CRAN.
 Two tutorials on running LDpred2 and lassosum2 using R package bigsnpr are available at \url{https://privefl.github.io/bigsnpr/articles/LDpred2.html} and \url{https://privefl.github.io/bigsnpr-extdoc/polygenic-scores-pgs.html}.
-We have extensively used R packages bigstatsr and bigsnpr \cite[]{prive2017efficient} for analyzing large genetic data, packages from the future framework \cite[]{bengtsson2020unifying} for easy scheduling and parallelization of analyses on the HPC cluster, and packages from the tidyverse suite \cite[]{wickham2019welcome} for shaping and visualizing results.
+
+We have recomputed the allele frequencies and imputation INFO scores within the European subset used here and across all imputed variants in the UK Biobank, and made them available at \url{https://doi.org/10.6084/m9.figshare.16635388}.
+We have formed independent LD blocks within the European LD references provided in \cite{prive2020ldpred2}, and made these updated versions available at \url{https://doi.org/10.6084/m9.figshare.19213299}.
 
@@
 
-The authors declare no competing interests.
-[TODO: ADD ALLELICA FOR BJARNI?]
+B.J.V.\ is on Allelica's international advisory board.
+The other authors declare no competing interests.
 

### 2022-02-23 code/compute-all-info-eur.R

@@ -62,28 +62,61 @@ mfi <- bigreadr::fread2("UKBB/mfi/ukb_mfi_chr22_v3.txt")
 merged <- inner_join(all_info, mfi,
                      by = c(rsid = "V2", allele1 = "V4", allele2 = "V5"))
 
-sampled <- slice_sample(merged, n = 50e3)
-library(ggplot2)
-ggplot(sampled, aes(pmin(freq, (1 - freq)), V6)) +
-  theme_bw(14) +
-  geom_point(alpha = 0.1) +
-  labs(x = "MAF recomputed", y = "MAF reported") +
-  coord_equal() +
-  geom_abline(color = "red") +
-  scale_x_log10() + scale_y_log10()
-# ggsave("figures/compare-all-maf-chr22.png", width = 8, height = 7)
+set.seed(1); sampled <- slice_sample(merged, n = 50e3)
 
-ggplot(sampled, aes(info, V8, color = V6 / pmin(freq, (1 - freq)))) +
-  theme_bw(14) +
-  geom_point(alpha = 0.2) +
-  labs(x = "INFO recomputed", y = "INFO reported",
-       color = "MAF_reported / MAF_recomputed") +
-  coord_equal() +
-  geom_abline(color = "red") +
-  scale_color_viridis_c(trans = "log", breaks = 10^(-3:4)) +
-  theme(legend.position = "top") +
-  guides(color = guide_colorbar(barwidth = 12, ticks.linewidth = 2))
-# ggsave("figures/compare-all-info-chr22.png", width = 7.5, height = 8)
+library(bigsnpr)
+map_subset <- snp_attach(snp_readBGEN(
+  bgenfiles   = "UKBB/bgen/ukb_imp_chr22_v3.bgen",
+  list_snp_id = list(with(sampled, paste(chromosome, physical.pos, allele1, allele2, sep = "_"))),
+  backingfile = tempfile(),
+  ncores      = nb_cores()
+))$map
+
+library(ggplot2)
+plot_grid(
+  ggplot(aes(pmin(map_subset$freq, (1 - map_subset$freq)), sampled$V6), data = NULL) +
+    theme_bw(14) +
+    geom_point(alpha = 0.1) +
+    labs(x = "MAF recomputed using all UKBB individuals", y = "MAF reported in MFI files") +
+    coord_equal() +
+    geom_abline(color = "red") +
+    scale_x_log10(breaks = c(0.1, 0.001, 1e-5, 1e-7)) +
[…]

### 2022-02-23 code/investigate-GWAS-from-imputed.R

@@ -36,58 +36,63 @@ plot_grid2(list(
     coord_equal() +
     scale_color_viridis_c() +
     guides(color = guide_colorbar(barheight = 15, ticks.linewidth = 2)) +
-    labs(y = "SD_imp", x = "SD_true"),
+    labs(x = "Standard deviations of \"true\" genotypes",
+         y = "Standard deviations of imputed dosages"),
   qplot(scale_true$scale, scale_imp$scale / sqrt(INFO), color = INFO) +
     geom_abline(color = "red") +
     coord_equal() +
     scale_color_viridis_c() +
-    labs(y = "SD_imp / sqrt(INFO)", x = "SD_true")
-), scale = 0.95, nrow = 1, labels = c("A", "B"), label_size = 16,
+    labs(x = "Standard deviations of \"true\" genotypes",
+         y = "Corrected standard deviations of imputed dosages")
+), scale = 0.95, nrow = 1, labels = c("A", "B"), label_size = 18,
 title_ratio = 0, legend_ratio = 0.1)
-# ggsave("figures/compare-sd.png", width = 12.5, height = 5.5)
+# ggsave("figures/compare-sd.png", width = 12.5, height = 5.5, scale = 1.3)
 
 plot_grid2(list(
   qplot(gwas_imp$estim, gwas_true$estim, color = INFO) +
     geom_abline(color = "red") +
     scale_color_viridis_c() +
     guides(color = guide_colorbar(barheight = 15, ticks.linewidth = 2)) +
-    labs(x = "beta_imp", y = "beta_true"),
+    labs(x = "GWAS effect sizes from imputed dosages",
+         y = "GWAS effect sizes from \"true\" genotypes"),
   qplot(gwas_imp$estim * sqrt(INFO), gwas_true$estim, color = INFO) +
     geom_abline(color = "red") +
     scale_color_viridis_c() +
-    labs(x = "beta_imp * sqrt(INFO)", y = "beta_true")
-), scale = 0.95, nrow = 1, labels = c("A", "B"), label_size = 16,
+    labs(x = "Corrected GWAS effect sizes from imputed dosages",
+         y = "GWAS effect sizes from \"true\" genotypes")
+), scale = 0.95, nrow = 1, labels = c("A", "B"), label_size = 18,
 title_ratio = 0, legend_ratio = 0.1)
[…]

### 2022-02-23 code/investigate-PCwithLD-GWAS.R

@@ -7,7 +7,7 @@ ind_chr6 <- which(ukb$map$chromosome == 6)
 y <- snp_simuPheno(G, h2 = 0.2, M = 2000, ind.possible = ind_chr6)$pheno
 
 POS <- ukb$map$physical.pos[ind_chr6]
-is_in_bad_region <- (POS > 70e6) & (POS < 91e6)
+is_in_bad_region <- ifelse((POS > 70e6) & (POS < 91e6), "Yes", "No")
 
 gwas  <- big_univLinReg(G, y, ind.col = ind_chr6, ncores = nb_cores())
 gwas2 <- big_univLinReg(G, y, ind.col = ind_chr6, ncores = nb_cores(), covar.train = PC19)

### 2022-02-23 code/investigate-corr-imp.R

@@ -4,34 +4,43 @@ G <- ukb$genotypes
 CHR <- ukb$map$chromosome
 POS2 <- ukb$map$genetic.dist
 INFO <- ukb$map$info
-
 (NCORES <- nb_cores())
 
-chr <- 22
-ind.chr <- which(CHR == chr)
+chr <- 1
+length(ind.chr <- which(CHR == chr & INFO < 0.85))  # 214
 
-corr <- runonce::save_run(
-  snp_cor(G, ind.col = ind.chr, infos.pos = POS2[ind.chr],
-          size = 3 / 1000, ncores = NCORES),
-  file = paste0("data/corr/chr", chr, ".rds")
-)
 
-corr2 <- runonce::save_run(
-  snp_cor(G, ind.col = ind.chr, infos.pos = POS2[ind.chr],
-          size = 3 / 1000, info = INFO[ind.chr], ncores = NCORES),
-  file = paste0("data/corr/adj_chr", chr, ".rds")
-)
+## From 190 individuals in 1000G
+obj.bed <- bed(download_1000G("tmp-data"))
+fam2 <- bigreadr::fread2(paste0(obj.bed$prefix, ".fam2"))
+ind_row <- which(fam2$Population %in% c("CEU", "GBR"))  # only 190 individuals
+matched <- snp_match(setNames(obj.bed$map[c(1, 4:6, 3)], c("chr", "pos", "a1", "a0", "beta")),
+                     setNames(cbind(ukb$map[ind.chr, c(1, 4:6)], POS2[ind.chr]),
+                              c("chr", "pos", "a1", "a0", "pos2")))
+ind_col <- matched$`_NUM_ID_.ss`
 
+corr_1kg <- bed_cor(obj.bed, ind.row = ind_row, ind.col = ind_col, fill.diag = FALSE,
+                    infos.pos = matched$pos2, size = 3 / 1000, ncores = NCORES)
+
+
+## From the UKBB dosage data
+ind.chr2 <- ind.chr[matched$`_NUM_ID_`]
+hist(INFO[ind.chr2])
+
+corr_dosage <- snp_cor(G, ind.col = ind.chr2, infos.pos = POS2[ind.chr2],
+                       size = 3 / 1000, ncores = NCORES)
+
+## From multiple imputation
 ind_csv <- ukb$fam$id_csv
 all_id <- bigreadr::fread2("UKBB/ukb41181.csv", select = "eid")[[1]]
[…]

### 2022-02-23 code/investigate-lassosum2-auto.R

@@ -6,13 +6,8 @@ load("data/ukbb4simu_ind.RData")
 
 (NCORES <- nb_cores())
 
-corr0 <- runonce::save_run({
-  POS2 <- snp_asGeneticPos(as.integer(ukb$map$chromosome),
-                           ukb$map$physical.pos, dir = "tmp-data")
-  snp_cor(G, ind.row = ind.val, infos.pos = POS2, size = 3 / 1000, ncores = nb_cores())
-}, file = "tmp-data/corr0_simu_val.rds")
-corr <- runonce::save_run(as_SFBM(corr0, "tmp-data/corr_simu_val"),
-                          file = "tmp-data/corr_simu_val.rds")
+# Correlation matrix made in 'code/prepare-corr-simu-chr22.R'
+corr <- readRDS("tmp-data/corr_simu_val.rds")
 
 y <- snp_simuPheno(G, h2 = 0.2, M = 2000, ncores = NCORES)$pheno
 
@@ -62,16 +57,19 @@ params$auto_score2 <- apply(beta_lassosum, 2, function(beta) {
   crossprod(beta, beta_hat_shrunk2) / sqrt(bRb)
 })
 
-source("https://raw.githubusercontent.com/privefl/paper4-bedpca/master/code/plot_grid2.R")
-plot_grid2(list(
+plot_grid(
   qplot(auto_score, score, color = as.factor(delta), data = params) +
     theme_bw(15) +
+    geom_abline(linetype = 2, color = "red") +
     labs(x = "Score from pseudo-validation (using correlations)",
-         y = "Score from validation", color = "delta"),
+         y = "Score from validation", color = expression(delta)) +
+    theme(legend.position = c(0.15, 0.7)),
   qplot(auto_score2, score, color = as.factor(delta), data = params) +
     theme_bw(15) +
+    geom_abline(linetype = 2, color = "red") +
     labs(x = "Score from pseudo-validation (using p-values)",
-         y = "Score from validation")
-), scale = 0.95, labels = c("A", "B"), label_size = 16, ncol = 1,
-title_ratio = 0, legend_ratio = 0.15)
-# ggsave("figures/pseudoval.pdf", width = 8, height = 10)
+         y = "Score from validation", color = expression(delta)) +
+    theme(legend.position = c(0.8, 0.3)),
+  scale = 0.95, labels = c("A", "B"), label_size = 16, ncol = 1
+)
+# ggsave("figures/pseudoval.pdf", width = 7, height = 10)

### 2022-02-23 code/prepare-corr-altpop.R

@@ -1,0 +1,108 @@
+library(dplyr)
+library(ggplot2)
+library(bigsnpr)
+NCORES <- nb_cores()
+
+snp_id <- with(snp_attach("data/ukbb4simu.rds")$map,
+               paste(chromosome, physical.pos, allele1, allele2, sep = "_"))
+
+df0 <- bigreadr::fread2(
+  "UKBB/ukb41181.csv",
+  select    = c("eid", paste0("22009-0.", 1:16)),
+  col.names = c("eid", paste0("PC", 1:16))
+)
+
+# still in data
+sample <- bigreadr::fread2("UKBB/ukb58024_imp_chr1_v3_s487296.sample")[-1, ]
+ind.indiv <- match(df0$eid, sample$ID_2)
+id_to_rm <- scan("../../UKBiobank/ukb_withdrawn_individuals_8Feb2021.id")
+ind.indiv[df0$eid %in% id_to_rm] <- NA
+sub <- which(!is.na(ind.indiv))
+
+# filter for relatedness
+rel <- bigreadr::fread2("UKBB/ukb58024_rel_s488264.dat")
+rel2 <- dplyr::filter(rel, Kinship > 2^-3.5)
+ind.indiv[df0$eid %in% rel2$ID2] <- NA
+sub <- which(!is.na(ind.indiv))
+
+# Closer to South Europe (than UK)
+PC_UKBB <- select(df0, PC1:PC16)[sub, ]
+all_centers <- read.csv(
+  "https://raw.githubusercontent.com/privefl/UKBB-PGS/main/pop_centers.csv",
+  stringsAsFactors = FALSE)
+center_italy <- unlist(all_centers[4, -1])
+all_sq_dist <- rowSums(sweep(PC_UKBB, 2, center_italy, '-')^2)
+closest_ind <- head(order(all_sq_dist), 10e3)
+hist(log(all_sq_dist), "FD", xlim = c(6, 9)); abline(v = log(all_sq_dist[tail(closest_ind, 1)]), col = "red")
+sub2 <- sub[closest_ind]
+length(sub2) # 10,000
+eid_UK <- df0$eid[sub2]
+
+# Read the data
+snp_readBGEN(
+  bgenfiles = "UKBB/bgen/ukb_imp_chr22_v3.bgen",
+  list_snp_id = list(snp_id),
+  backingfile = "data/ukbb4simu_altpop",
[…]

### 2022-02-23 code/prepare-corr-simu-chr22.R

@@ -13,7 +13,7 @@ corr0 <- runonce::save_run(
   snp_cor(G, ind.row = ind.val, infos.pos = POS2, size = 3 / 1000, ncores = nb_cores()),
   file = "tmp-data/corr0_simu_val.rds")
 
-corr <- runonce::save_run(as_SFBM(corr0, "tmp-data/corr_simu_val"),
+corr <- runonce::save_run(as_SFBM(corr0, "tmp-data/corr_simu_val", compact = TRUE),
                           file = "tmp-data/corr_simu_val.rds")
 
 
@@ -26,39 +26,28 @@ splits <- purrr::map_dfr(3:8 * 1000, function(max_size) {
   res
 })
 
-corr0T <- as(corr0, "dgTMatrix")
-corr0_i <- corr0T@i + 1L
-corr0_j <- corr0T@j + 1L
-
-library(future.apply)
-plan("multisession", workers = nb_cores())
-options(future.globals.maxSize = 2e9)
-splits$cost_size <- future_sapply(splits$block_num, function(grp) {
-  sum(grp[corr0_i] == grp[corr0_j])
-})
-plan("sequential")
-
 library(ggplot2)
-qplot(data = splits, cost_size / length(corr0T@x), cost, color = as.factor(max_size)) +
+qplot(data = splits, perc_kept, cost, color = as.factor(max_size)) +
   theme_bw(12) +
   theme(legend.position = "top") + guides(colour = guide_legend(nrow = 1)) +
   labs(y = "Sum of squared correlations outside blocks (log-scale)",
        x = "% of non-zero values kept", color = "Maximum block size") +
   ylim(0, median(splits$cost)) +
   geom_vline(xintercept = 0.6, linetype = 3)
-# ggsave("figures/cost_split_chr22.pdf", width = 7, height = 6)
+# ggsave("figures/cost_split_chr22.pdf", width = 8, height = 6)
 
 library(dplyr)
 final_grp <- splits %>%
-  filter(cost_size < (0.6 * length(corr0T@x))) %>%
+  filter(perc_kept < 0.6) %>%
   arrange(cost) %>%
   slice(1) %>%
   print() %>%
   pull(block_num) %>%
   unlist()
 
-corr0T@x <- ifelse(final_grp[corr0_i] == final_grp[corr0_j], corr0T@x, 0)
+corr0T <- as(corr0, "dgTMatrix")
[…]

### 2022-02-23 code/prepare-corr.R

@@ -3,7 +3,6 @@ ukb <- snp_attach("data/UKBB_HM3_val.rds")
 G <- ukb$genotypes
 CHR <- ukb$map$chromosome
 POS2 <- ukb$map$genetic.dist
-INFO <- ukb$map$info
 
 (NCORES <- nb_cores())
 
@@ -17,11 +16,5 @@ for (chr in 1:22) {
     snp_cor(G, ind.col = ind.chr, infos.pos = POS2[ind.chr],
             size = 3 / 1000, ncores = NCORES),
     file = paste0("data/corr/chr", chr, ".rds")
-  )
-
-  corr2 <- runonce::save_run(
-    snp_cor(G, ind.col = ind.chr, infos.pos = POS2[ind.chr],
-            size = 3 / 1000, info = INFO[ind.chr], ncores = NCORES),
-    file = paste0("data/corr/adj_chr", chr, ".rds")
   )
 }

### 2022-02-23 code/prepare-ldref-with-blocks.R

@@ -1,26 +1,33 @@
 library(future.batchtools)
 plan(batchtools_slurm(resources = list(
-  t = "12:00:00", c = 2, mem = "50g",
+  t = "12:00:00", c = 6, mem = "125g",
   name = basename(rstudioapi::getSourceEditorContext()$path))))
+
+bigassertr::assert_dir("ldref")
 
 all_final_grp <- furrr::future_map_dfr(1:22, function(chr) {
 
-  corr0 <- readRDS(paste0("data/corr/chr", chr, ".rds"))
+  corr0 <- readRDS(paste0("../paper-ldpred2/ld-ref/LD_chr", chr, ".rds"))
   dim(corr0)
 
   library(bigsnpr)
+  library(furrr)
   all_splits <- runonce::save_run({
+
+    plan("multisession", workers = 5)
+    options(future.globals.maxSize = 10e9)
+
     SEQ <- round(seq_log(1000 + ncol(corr0) / 50,
                          5000 + ncol(corr0) / 10,
                          length.out = 10))
 
-    splits <- purrr::map_dfr(SEQ, function(max_size) {
+    splits <- future_map_dfr(SEQ, function(max_size) {
       res <- snp_ldsplit(corr0, thr_r2 = 0.02, min_size = 100,
                          max_size = max_size, max_K = 200)
       res$max_size <- max_size
       res
     })
-  }, file = paste0("tmp-data/split_corr_chr", chr, ".rds"))
+  }, file = paste0("tmp-data/split_ldref_chr", chr, ".rds"))
 
 
   library(ggplot2)
@@ -45,17 +52,18 @@ all_final_grp <- furrr::future_map_dfr(1:22, function(chr) {
   corr0T <- as(corr0, "dgTMatrix")
   corr0T@x <- ifelse(final_grp[corr0T@i + 1L] == final_grp[corr0T@j + 1L], corr0T@x, 0)
 
-  corr0T %>%
-    Matrix::drop0() %>%
-    as("symmetricMatrix") %>%
-    saveRDS(paste0("data/corr/adj_with_blocks_chr", chr, ".rds"))
+  new_corr0 <- runonce::save_run(
+    as(Matrix::drop0(corr0T), "symmetricMatrix"),
+    file = paste0("ldref/LD_with_blocks_chr", chr, ".rds")
+  )
 
   final_split
 })
+# <20 min after chr6
 
[…]

### 2022-02-23 code/process-PRS.R

@@ -144,11 +144,12 @@ all_res0 <- grid %>%
 
 for (PHENO in c("t1d", "brca", "mdd", "prca", "cad")) {
 
-  all_res <- filter(all_res0, grepl(PHENO, pheno))
+  all_res <- filter(all_res0, grepl(PHENO, pheno)) %>%
+    mutate(use_info = ifelse(use_info, "Yes", "No"))
 
   library(ggplot2)
   ggplot(filter(all_res, !use_blocks),
-         aes(method, pcor_1, fill = paste(qc, use_info))) +
+         aes(method, pcor_1, fill = paste(qc, use_info, sep = " - "))) +
     facet_wrap(~ pheno, nrow = 1) +
     bigstatsr::theme_bigstatsr(0.8) +
     scale_fill_manual(values =  c("#999999", "#E69F00", "#56B4E9", "#009E73")) +
@@ -157,12 +158,14 @@ for (PHENO in c("t1d", "brca", "mdd", "prca", "cad")) {
                   position = position_dodge(width = 0.9),
                   color = "black", width = 0.2, size = 1) +
     labs(x = "Method", y = "Partial correlation between PGS and phenotype",
-         fill = "Which QC? Correction using INFO?") +
-    theme(legend.position = "top") +
+         fill = "Which QC? - Correction using INFO?") +
+    theme(legend.position = "top",
+          legend.text = element_text(margin = margin(r = 10, unit = "pt")),
+          legend.title = element_text(margin = margin(r = 10, unit = "pt"))) +
     geom_col(data = filter(all_res, use_blocks),
              position = position_dodge(), color = "red",
              alpha = 0, show.legend = FALSE)
 
   ggsave(paste0("figures/res-", PHENO, ".pdf"),
-         width = `if`(nrow(all_res) < 50, 9, 13), height = 6.5)
+         width = `if`(nrow(all_res) < 50, 9.5, 13.5), height = 6.5)
 }

### 2022-02-23 code/run-PRS.R

@@ -14,7 +14,7 @@ files <- tibble(pheno = c(paste0("t1d_", c("affy", "illu")),
 library(future.batchtools)
 NCORES <- 30
 plan(batchtools_slurm(resources = list(
-  t = "12:00:00", c = NCORES + 2, mem = "80g",
+  t = "12:00:00", c = NCORES + 2, mem = "120g",
   name = basename(rstudioapi::getSourceEditorContext()$path))))
 
 furrr::future_pwalk(files, function(pheno, qc, name_corr) {

### 2022-02-23 code/simu-altpop.R

@@ -1,0 +1,128 @@
+library(dplyr)
+library(bigsnpr)
+ukb <- snp_attach("data/ukbb4simu.rds")
+G <- ukb$genotypes
+
+
+library(future.batchtools)
+NCORES <- 15
+plan(batchtools_slurm(resources = list(
+  t = "12:00:00", c = NCORES + 1, mem = "100g",
+  name = basename(rstudioapi::getSourceEditorContext()$path))))
+
+bigassertr::assert_dir("results-simu-altpop")
+bigassertr::assert_dir("log")
+
+
+furrr::future_walk(1:10, function(ic) {
+
+  res_file <- paste0("results-simu-altpop/res_", ic, ".rds")
+  if (file.exists(res_file)) return(NULL)
+
+  #### Run GWAS and prepare sumstats ####
+
+  load("data/ukbb4simu_ind.RData")
+
+  y <- snp_simuPheno(G, h2 = 0.2, M = 2000, ncores = NCORES)$pheno
+
+  # GWAS to get sumstats
+  gwas <- big_univLinReg(G, y[ind.gwas], ind.train = ind.gwas, ncores = NCORES)
+  df_beta <- transmute(gwas, beta = estim, beta_se = std.err, n_eff = length(ind.gwas))
+
+  with(df_beta, min(pchisq((beta / beta_se)^2, df = 1, lower.tail = FALSE)))
+
+  #### Run them all ####
+
+  source("code/run-methods.R", local = TRUE)
+
+  # Correlation matrices made in 'code/prepare-corr-simu-chr22.R'
+  corr1 <- readRDS("tmp-data/corr_simu_val.rds")
+  corr2 <- readRDS("tmp-data/corr_simu_val_with_blocks.rds")
+  # Correlation matrices made in 'code/prepare-corr-altpop.R'
+  corr3 <- readRDS("tmp-data/corr_simu_altpop.rds")
+  corr4 <- readRDS("tmp-data/corr_simu_altpop_with_blocks.rds")
+
+  ALL_COR <- list(corr1, corr2, corr3, corr4)
+  ALL_LDPRED <- c("LDpred2-inf", "LDpred2", "LDpred2-low-h2",
+                  "LDpred2-auto", "LDpred2-auto-rob")
+
+  all_comb <- tibble::tribble(
+    ~ Method,    ~ fun,         ~ corr,
+    ALL_LDPRED,  run_ldpred2,   ALL_COR,
[…]

### 2022-02-23 code/simu-rounded.R

@@ -1,0 +1,141 @@
+library(dplyr)
+library(bigsnpr)
+ukb <- snp_attach("data/ukbb4simu.rds")
+G <- ukb$genotypes
+
+
+library(future.batchtools)
+NCORES <- 15
+plan(batchtools_slurm(resources = list(
+  t = "12:00:00", c = NCORES + 1, mem = "100g",
+  name = basename(rstudioapi::getSourceEditorContext()$path))))
+
+bigassertr::assert_dir("results-simu-rounded")
+bigassertr::assert_dir("log")
+
+
+furrr::future_walk(1:10, function(ic) {
+
+  res_file <- paste0("results-simu-rounded/res_", ic, ".rds")
+  if (file.exists(res_file)) return(NULL)
+
+  #### Run GWAS and prepare sumstats ####
+
+  load("data/ukbb4simu_ind.RData")
+
+  y <- snp_simuPheno(G, h2 = 0.2, M = 2000, ncores = NCORES)$pheno
+
+  # GWAS to get sumstats
+  gwas <- big_univLinReg(G, y[ind.gwas], ind.train = ind.gwas, ncores = NCORES)
+  df_beta <- transmute(gwas, beta = estim, beta_se = std.err, n_eff = length(ind.gwas))
+
+  with(df_beta, min(pchisq((beta / beta_se)^2, df = 1, lower.tail = FALSE)))
+
+  # library(ggplot2)
+  # maf <- snp_MAF(G, ind.row = ind.gwas, ncores = NCORES)
+  # qplot(sqrt(2 * maf * (1 - maf)), 1 / sqrt(n_eff * beta_se^2 + beta^2), alpha = I(0.5),
+  #       data = mutate_at(df_beta, c("beta", "beta_se"), ~ signif(., 2))) +
+  #   theme_bigstatsr() +
+  #   coord_equal() +
+  #   scale_color_viridis_d(direction = -1) +
+  #   geom_abline(linetype = 2, color = "red") +
+  #   labs(x = "Standard deviations derived from allele frequencies",
+  #        y = "Standard deviations derived from the summary statistics",
+  #        color = "Removed?")
+
+
+  #### Run them all ####
+
+  source("code/run-methods.R", local = TRUE)
+
+  # Correlation matrices made in 'code/prepare-corr-simu-chr22.R'
+  corr1 <- readRDS("tmp-data/corr_simu_val.rds")
[…]

### 2022-02-23 tmp-code/investigate-T2D-rounded.R

@@ -1,0 +1,52 @@
+library(bigsnpr)
+library(bigreadr)
+library(dplyr)
+library(ggplot2)
+
+map <- transmute(snp_attach("data/UKBB_HM3_val.rds")$map,
+                 chr = as.integer(chromosome), pos = physical.pos, af_val = freq,
+                 a0 = allele1, a1 = allele2)  # reversed somehow..
+
+
+#### Type 2 diabetes (T2D) ####
+
+# Download from http://diagram-consortium.org/downloads.html
+# DIAGRAM 1000G GWAS meta-analysis Stage 1 Summary statistics
+# Published in Scott et al (2017)
+# unzip("tmp-data/METAANALYSIS_DIAGRAM_SE1.zip", exdir = "tmp-data")
+sumstats <- fread2("tmp-data/METAANALYSIS_DIAGRAM_SE1.txt")
+sumstats <- tidyr::separate(sumstats, "Chr:Position", c("chr", "pos"), convert = TRUE)
+names(sumstats) <- c("chr", "pos", "a1", "a0", "beta", "beta_se", "p", "N")
+head(sumstats$beta)     # -0.0130 -0.0710  0.0140  0.1500 -0.0085 -0.0060
+head(sumstats$beta_se)  # 0.026 0.170 0.012 0.130 0.013 0.016
+
+table_text <- "cases/408/347\ncontrols/3,098/3,911\ncases/97/35\ncontrols/297/158\ncases/181/74\ncontrols/801/846\ncases/3,982/3,357\ncontrols/35,946/47,103\ncases/413/266\ncontrols/281/416\ncases/530/493\ncontrols/541/533\ncases/41/39\ncontrols/866/902\ncases/161/228\ncontrols/2,754/3,259\ncases/2,229/2,395\ncontrols/1,673/2,995\ncases/386/287\ncontrols/3,441/4,219\ncases/653/508\ncontrols/574/600\ncases/1,773/1,525\ncontrols/1,376/1,267\ncases/-/1,124\ncontrols/-/1,298\ncases/201/146\ncontrols/764/786\ncases/-/1,467\ncontrols/-/1,754\ncases/65/46\ncontrols/409/429\ncases/251/403\ncontrols/2,111/3,108\ncases/166/0\ncontrols/953/0\ncases/1,118/806\ncontrols/1,446/1,492"
+sizes <- bigreadr::fread2(text = gsub(",", "", table_text, fixed = TRUE),
[…]

### 2022-03-24 paper/paper-misspec.tex
longer abstract

@@
 
-\vspace*{8em}
+\vspace*{7em}
 
 \abstract{
-	Are genome-wide association studies (GWAS) summary statistics of good enough quality for performing follow-up analyses?
-Can we detect possible misspecifications in GWAS summary statistics and correct for them in order to improve predictive performance of polygenic scores?
+	Genome-wide association studies (GWAS) summary statistics come with uneven quality.
+Are they of good enough quality for performing follow-up analyses?
+First, we present an overview of possible misspecifications that come with GWAS summary statistics.
+Can we detect these possible misspecifications in GWAS summary statistics and correct for them to e.g.
+improve predictive performance of polygenic scores?
+Then, we show that additional information such as imputation INFO scores, allele frequencies, and per-variant sample sizes are very useful information to report in GWAS summary statistics because they can be used to detect possible issues and correct for misspecifications in the GWAS summary statistics.
+Unfortunately, due to the lack of reporting standards for GWAS summary statistics, this additional information is not systematically reported.
+Finally, we discuss how to make polygenic score methods, such as lassosum and LDpred2, more robust to these misspecifications to improve their predictive power.
 }
@@
 As for the correction using INFO scores, we use the first correction, ``sqrt\_info'', which is simple because it is equivalent to post-processing PGS effects by multiplying them by $\sqrt{\text{INFO}}$.
-Although results for both QC used are very similar, correcting using INFO scores slightly improves prediction when deriving polygenic scores based on iCOGS summary statistics (Figure \ref{fig:res_brca}).
+Although results for both QC used are very similar, correcting using INFO scores slightly improves predictive performance when deriving polygenic scores based on iCOGS summary statistics (Figure \ref{fig:res_brca}).
 All other improvements introduced before have little to no effect here, probably because misspecifications are much smaller than in the simulations.
@@
 In figures \ref{fig:res_cad}-\ref{fig:res_t1d}, we then provide similar results as figure \ref{fig:res_brca} for other phenotypes.
-Most changes introduced before have little to no impact on the predictions, except for the second quality control performed for CAD when using LDpred2-auto (Figure \ref{fig:res_cad}).
+Most changes introduced before have little to no impact on predictive performance, except for the second quality control performed for CAD when using LDpred2-auto (Figure \ref{fig:res_cad}).
 

### 2022-03-31 paper/paper-misspec-supp.tex
update text

@@
 \begin{figure}[p]
-	\centerline{\includegraphics[width=0.9\textwidth]{simu-altpop}}
-	\caption{Results for the simulations with summary statistics with LD matrices based on two different populations.
-One comes from the same population used for computing the GWAS summary statistics (North-West Europe), while the other one comes from South Europe (alternative LD reference).
-Reported 95\% confidence intervals are computed from 10,000 non-parametric bootstrap replicates of the mean.
-Red bars correspond to using the LD with independent blocks (Methods).}
-	\label{fig:simu-altpop}
+	\centerline{\includegraphics[width=0.95\textwidth]{res-BBJ}}
+	\caption{Results for PGS derived from four Biobank Japan GWAS summary statistics and using three different LD reference.
+Partial correlations are computed using function \texttt{pcor} of R package bigstatsr where 95\% confidence intervals are obtained through Fisher's Z-transformation, then all values are squared to report the phenotypic variance explained by PGS.}
+	\label{fig:bbj}
 \end{figure}
@@
 	\centerline{\includegraphics[width=0.75\textwidth]{hist_bad_pos}}
-	\caption{Histogram of the positions of outlier variants from Figure 3B in the main text.%TODO: VERIF NUMBER AT THE END
+	\caption{Histogram of the positions of outlier variants from Figure 5B in the main text.%TODO: VERIF NUMBER AT THE END
 	\label{fig:hist-bad-pos}}
@@
 \begin{figure}[p]
-	\centerline{\includegraphics[width=0.7\textwidth]{res-cad}}
-	\caption{Raw partial correlations (for \textit{all} models) for predicting CAD in the UK Biobank from external summary statistics.
-These are computed using function \texttt{pcor} of R package bigstatsr where 95\% confidence intervals are obtained through Fisher's Z-transformation.}
-	\label{fig:res_cad}
+	\centerline{\includegraphics[width=0.7\textwidth]{res-brca_icogs}}
+	\caption{Variance explained of BRCA in the UK Biobank by PGS derived from external summary statistics (iCOGS).
+These are computed using function \texttt{pcor} of R package bigstatsr where 95\% confidence intervals are obtained through Fisher's Z-transformation; these values are then squared.}
+	\label{fig:res_brca_icogs}
+\end{figure}
+
+\begin{figure}[p]
+	\centerline{\includegraphics[width=0.7\textwidth]{res-brca_onco}}
+	\caption{Variance explained of BRCA in the UK Biobank by PGS derived from external summary statistics (OncoArray).
+These are computed using function \texttt{pcor} of R package bigstatsr where 95\% confidence intervals are obtained through Fisher's Z-transformation; these values are then squared.}
+	\label{fig:res_brca_onco}
 \end{figure}
@@
 	\centerline{\includegraphics[width=0.7\textwidth]{res-mdd}}
-	\caption{Raw partial correlations (for \textit{all} models) for predicting MDD in the UK Biobank from external summary statistics.
-These are computed using function \texttt{pcor} of R package bigstatsr where 95\% confidence intervals are obtained through Fisher's Z-transformation.}
+	\caption{Variance explained of MDD in the UK Biobank by PGS derived from external summary statistics.
+These are computed using function \texttt{pcor} of R package bigstatsr where 95\% confidence intervals are obtained through Fisher's Z-transformation; these values are then squared.}
 	\label{fig:res_mdd}
@@
 	\centerline{\includegraphics[width=0.7\textwidth]{res-prca}}
-	\caption{Raw partial correlations (for \textit{all} models) for predicting PRCA in the UK Biobank from external summary statistics.
-These are computed using function \texttt{pcor} of R package bigstatsr where 95\% confidence intervals are obtained through Fisher's Z-transformation.}
+	\caption{Variance explained of PRCA in the UK Biobank by PGS derived from external summary statistics.
+These are computed using function \texttt{pcor} of R package bigstatsr where 95\% confidence intervals are obtained through Fisher's Z-transformation; these values are then squared.}
 	\label{fig:res_prca}
@@
 \begin{figure}[p]
-	\centerline{\includegraphics[width=0.95\textwidth]{res-t1d}}
-	\caption{Raw partial correlations (for \textit{all} models) for predicting T1D in the UK Biobank from external summary statistics.
-These are computed using function \texttt{pcor} of R package bigstatsr where 95\% confidence intervals are obtained through Fisher's Z-transformation.}
-	\label{fig:res_t1d}
+	\centerline{\includegraphics[width=0.7\textwidth]{res-t1d_illu}}
+	\caption{Variance explained of T1D in the UK Biobank by PGS derived from external summary statistics (Illumina).
+These are computed using function \texttt{pcor} of R package bigstatsr where 95\% confidence intervals are obtained through Fisher's Z-transformation; these values are then squared.}
+	\label{fig:res_t1d_illu}
+\end{figure}
+
+\begin{figure}[p]
+	\centerline{\includegraphics[width=0.7\textwidth]{res-t1d_affy}}
+	\caption{Variance explained of T1D in the UK Biobank by PGS derived from external summary statistics (Affymetrix).
+These are computed using function \texttt{pcor} of R package bigstatsr where 95\% confidence intervals are obtained through Fisher's Z-transformation; these values are then squared.}
+	\label{fig:res_t1d_aff}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%\clearpage
+
+\FloatBarrier
+
+\begin{figure}[p]
+\centerline{\includegraphics[width=0.8\textwidth]{vitaminD_qc}}
+\caption{Standard deviations inferred from the vitamin D GWAS summary statistics versus the ones inferred from the allele frequencies of the LD reference.
+Only 100,000 variants are represented, at random.}
+\label{fig:vitaminD_qc}
+\end{figure}
+
+\begin{figure}[p]
+\centerline{\includegraphics[width=0.75\textwidth]{hist_N_vitaminD}}
+\caption{Histogram of the per-variant sample sizes in the vitamin D GWAS summary statistics. \label{fig:hist-N-vitaminD}}
 \end{figure}
@@
 	\centerline{\includegraphics[width=0.85\textwidth]{pseudoval}}
-	\caption{For one simulation, scores from validation versus scores from pseudo-validation as described in \cite{mak2017polygenic} using either \textbf{A:} correlations (the default in lassosum) or \textbf{B:} p-values when computing local false discovery rates.}
+	\caption{For one simulation, scores from validation versus scores from pseudo-validation as described in \cite{mak2017polygenic} using either \textbf{A:} correlations (the default in lassosum) or \textbf{B:} p-values, when computing local false discovery rates.}
 	\label{fig:pseudoval}

### 2022-03-31 paper/paper-misspec.tex
update text

@@
 Unfortunately, due to the lack of reporting standards for GWAS summary statistics, this additional information is not systematically reported.
+We also show how important it is to use well-matched LD references.
 Finally, we discuss how to make polygenic score methods, such as lassosum and LDpred2, more robust to these misspecifications to improve their predictive power.
@@
 
-\begin{figure}[h]
+\begin{figure}[!h]
 	\centerline{\includegraphics[width=0.98\textwidth]{simu-misN}}
@@
 
-\begin{figure}[h]
+\begin{figure}[!h]
 	\centerline{\includegraphics[width=0.98\textwidth]{simu-info}}
@@
 
+\begin{figure}[!h]
+	\centerline{\includegraphics[width=0.9\textwidth]{simu-altpop}}
+	\caption{Results for the simulations with summary statistics with LD matrices based on two different populations.
+One comes from the same population used for computing the GWAS summary statistics (North-West Europe), while the other one comes from South Europe (alternative LD reference).
+Reported 95\% confidence intervals are computed from 10,000 non-parametric bootstrap replicates of the mean.
+Red bars correspond to using the LD with independent blocks (Methods).}
+	\label{fig:simu-altpop}
+\end{figure}
+
+\subsection{Application to FinnGen and Biobank Japan summary statistics}
+
+We download and use GWAS summary statistics from five disease endpoints from FinnGen (release 6, \cite{kurki2022finngen}), namely breast and prostate cancers (BrCa and PrCa), coronary artery disease (CAD), type 1 and type 2 diabetes (T1D and T2D).
+These have been derived using SAIGE \cite[]{zhou2018efficiently}.
+First, for each phenotype, we estimate a global effective sample using the 80th percentile of imputed sample sizes from equation \eqref{eq:imputeNeff}.
+This estimation is only 84.7\% for BrCa, 79.1\% for PrCa, 73.1\% for T1D, 64.5\% for T2D, and 61.3\% for CAD, when compared to the effective sample computed from the reported numbers of cases and controls.
+We believe this reduction in effective sample size is due to both having related individuals as well as using SAIGE, as noted in the introduction.
+We then compare three different LD reference panels to use with these GWAS summary statistics (Methods section \ref{sec:altLD}).
+Using the small Finnish LD reference panel composed of 503 individuals only seems to consistently provide more predictive polygenic scores than using the other two LD references, including the large UK one composed of 10,000 individuals (Figure \ref{fig:finngen}).
+Note that the PGS are validated in people of UK-like ancestry, so we would expect larger improvements if validating in a Finnish population.
+
+\begin{figure}[!h]
+	\centerline{\includegraphics[width=0.85\textwidth]{res-FIN}}
+	\caption{Results for PGS derived from five FinnGen GWAS summary statistics and using three different LD reference.
+Partial correlations are computed using function \texttt{pcor} of R package bigstatsr where 95\% confidence intervals are obtained through Fisher's Z-transformation, then all values are squared to report the phenotypic variance explained by PGS.}
+	\label{fig:finngen}
+\end{figure}
+
+We also download and use GWAS summary statistics from four continuous outcomes from Biobank Japan \cite[]{sakaue2021cross}.
+These have been derived using BOLT-LMM-inf \cite[]{loh2018mixed}.
+First, for each phenotype, we estimate a global power improvement provided by BOLT-LMM by comparing $\chi^2$-statistics (the boost of BOLT-LMM-inf versus linear regression) across genome-wide significant variants, as recommended in \cite{loh2018mixed}.
+This estimated power ratio (increase in effective sample size) is 1.143 for height, 1.039 for HDL cholesterol, 1.025 for BMI, and 1.013 for systolic blood pressure.
+We then compare three different LD reference panels to use with these GWAS summary statistics (Methods section \ref{sec:altLD}).
+Interesting, when looking at height and to some extent at HDL cholesterol where we could get the best predictive performance, using the smaller UKBB Japanese LD reference seems to provide more predictive PGS than using the other two LD references, including the one using a larger set of East Asian individuals from the UKBB (Figure \ref{fig:bbj}).
+Note that the PGS are validated in people of broad East Asian ancestry, so we would expect larger improvements if validating in a Japanese population.
+
 \subsection{Application to breast cancer summary statistics}
@@
 
-\begin{figure}[h]
+\begin{figure}[!h]
 	\centerline{\includegraphics[width=0.98\textwidth]{brca_onco_qc}}
@@
 Therefore, providing an accurate imputation INFO score is useful for two reasons.
-First, it allows for correcting for a reduced standard deviation when using imputed data in the QC step we propose, in order to better uncover problems with the summary statistics.
+First, it allows for correcting for a reduced standard deviation when using imputed data in the QC step we propose, in order to better uncover possible problems with the GWAS summary statistics.
 Second, using one of the proposed corrections based on INFO scores may lead to an improved prediction when deriving polygenic scores.
@@
 In the comparison, we use the QC proposed in \cite{prive2020ldpred2} first (which ends up filtering on MAF here, which we call ``qc1'').
-We then also filter out the two long-range LD regions of chromosome 6 and 8 for the OncoArray summary statistics and remove around 500 variants when filtering on differences of MAFs between summary statistics and the validation dataset (``qc2'').
+We then also filter out the two long-range LD regions of chromosome 6 and 8 for the OncoArray summary statistics and remove around 500 variants when filtering on differences of AFs between summary statistics and the validation dataset (``qc2'').
 As for the correction using INFO scores, we use the first correction, ``sqrt\_info'', which is simple because it is equivalent to post-processing PGS effects by multiplying them by $\sqrt{\text{INFO}}$.
-Although results for both QC used are very similar, correcting using INFO scores slightly improves predictive performance when deriving polygenic scores based on iCOGS summary statistics (Figure \ref{fig:res_brca}).
+Although results for both QC used are very similar, correcting using INFO scores slightly improves predictive performance when deriving polygenic scores based on iCOGS summary statistics (Figure \ref{fig:res_brca_icogs}).
 All other improvements introduced before have little to no effect here, probably because misspecifications are much smaller than in the simulations.
-
-\begin{figure}[h]
-	\centerline{\includegraphics[width=0.98\textwidth]{res-brca}}
-	\caption{Raw partial correlations (for \textit{all} models) for predicting breast cancer in the UK Biobank when using either the OncoArray or the iCOGS summary statistics.
-These are computed using function \texttt{pcor} of R package bigstatsr where 95\% confidence intervals are obtained through Fisher's Z-transformation.}
-	\label{fig:res_brca}
-\end{figure}
 
@@
 QC plots comparing standard deviations usually show little deviation from the identity line (after the INFO score correction), except for coronary artery disease (CAD) summary statistics (Figures \ref{fig:brca_icogs_qc}-\ref{fig:t1d_illu_qc}).
-In figures \ref{fig:res_cad}-\ref{fig:res_t1d}, we then provide similar results as figure \ref{fig:res_brca} for other phenotypes.
-Most changes introduced before have little to no impact on predictive performance, except for the second quality control performed for CAD when using LDpred2-auto (Figure \ref{fig:res_cad}).
+In figures \ref{fig:res_cad} and \ref{fig:res_brca_onco}-\ref{fig:res_t1d_aff}, we then provide similar results as figure \ref{fig:res_brca_icogs} for other phenotypes.
+For MDD and PRCA, most changes introduced before have little to no impact on predictive performance.
+For CAD and T1D, the QC proposed in \cite{prive2020ldpred2} and the additional QC proposed here provides much better predictive performance, especially for LDpred2-auto (and LDpred2-auto-rob) than when using no quality control, showing how important this preliminary step is.
+
+\begin{figure}[p]
+	\centerline{\includegraphics[width=0.7\textwidth]{res-cad}}
+	\caption{Variance explained of CAD in the UK Biobank by PGS derived from external summary statistics.
+These are computed using function \texttt{pcor} of R package bigstatsr where 95\% confidence intervals are obtained through Fisher's Z-transformation; these values are then squared.}
+	\label{fig:res_cad}
+\end{figure}
+
+We also investigate results for vitamin D GWAS summary statistics, which do not report INFO scores nor allele frequencies, as opposed to previous ones.
+The QC procedure is then less precise and uses allele frequencies from the LD reference (Figure \ref{fig:vitaminD_qc});
+However, these summary statistics do report per-variant sample sizes, which cover a wide range of different values (Figure \ref{fig:hist-N-vitaminD}).
+Here we test using either the maximum sample size or the true per-variant sample sizes when deriving polygenic scores, and an additional QC step ``qc2'' which refers to removing all variants with a per-variant sample size less than 70\% of the maximum one.
+When the true per-variant sample sizes are used, the additional ``qc2'' does not seem to be necessary (Figure \ref{fig:res_vitaminD}), which is a bit surprising to us.
+When the maximum sample size is used, this ``qc2'' is very important, especially for LDpred2-auto.
+Note that, when using the maximum sample size to derive ``sd\_ss'' in ``qc1'' (not the case here; we used the per-variant sample sizes), ``qc1'' would remove variants with underestimated standard deviations due to overestimated sample sizes, which would effectively remove variants with low sample sizes, similar to the ``qc2'' used here.
+
+\begin{figure}[p]
+	\centerline{\includegraphics[width=0.7\textwidth]{res-vitaminD}}
+	\caption{Variance explained of vitamin D in the UK Biobank by PGS derived from external summary statistics.
+These are computed using function \texttt{pcor} of R package bigstatsr where 95\% confidence intervals are obtained through Fisher's Z-transformation; these values are then squared.}
+	\label{fig:res_vitaminD}
+\end{figure}
 
@@
 
-Note that, in this study, we use GWAS summary statistics that include extended information (e.g.\ INFO scores and allele frequencies), yet most GWAS summary statistics do not provide such exhaustive information \cite[]{macarthur2021workshop}.
+Note that, in this study, we mostly use GWAS summary statistics that include extended information (e.g.\ INFO scores and allele frequencies), yet most GWAS summary statistics do not provide such exhaustive information \cite[]{macarthur2021workshop}.
 We acknowledge that, in the case of a meta-analysis from multiple studies, providing a single INFO score per variant may not be possible.
@@
 Fortunately, an overestimated sample size can be detected from the QC plot we propose, where the slope is then less than 1 for case-control studies using logistic regression; otherwise the standard deviation of the phenotype is also needed (Equation \ref{eq:approx-sd-lin}), but can be estimated \cite[]{prive2020ldpred2}.
+Here we have used this strategy to estimate reduced effective sample sizes in FinnGen GWAS summary statistics.
 
@@
 Nevertheless, we recommend to adopt these changes, i.e.\ performing the (refined) QC proposed here, forming independent LD blocks in the LD matrix, and using more regularization when needed. 
+We have shown other useful results, such as that filtering small sample sizes is very useful, as well as using a well-matched LD reference panel. 
 More regularization can be achieved by testing additional smaller values for the heritability parameter in LDpred2-grid, and using the two new parameters introduced here in LDpred2-auto (Section \ref{sec:robust}).
@@
 To address this limitation, we recommend to rely on a proper QC and choice of ancestry-matched LD instead of on too much regularization.
+We would like to invite large biobanks, such as FinnGen and Biobank Japan, to provide LD reference matrices matching the large GWAS summary statistics they provide, ideally based on the same large number of individuals.
 As a future research direction, we are interested in using multi-ancestry-matched LD matrices to improve polygenic prediction in all ancestries.
-It would also be useful to investigate the impact of the QC and adjustments we propose here on other (non-PGS) methods, for which consistency and robustness are likely to be very important as well \cite[]{chen2021improved}.
+It would also be useful to investigate the impact of the QC, well-matched LD reference panels and other adjustments we propose here on other (non-PGS) methods, for which consistency and robustness are likely to be very important as well \cite[]{chen2021improved}.
 
@@
 To define phenotypes in the UK Biobank, we first map ICD10 and ICD9 codes (UKBB fields 40001, 40002, 40006, 40013, 41202, 41270 and 41271) to phecodes using R package PheWAS \cite[]{carroll2014r,wu2019mapping}.
+We also use some continuous phenotypes, namely vitamin D (Data-Field 30890), height (50), BMI (21001), systolic blood pressure (4080), and HDL cholesterol (30760).
+
 We use published GWAS summary statistics listed in table \ref{tab:sumstats} to derive polygenic scores.
+We also use GWAS summary statistics for five disease endpoints from FinnGen (release 6, \cite{kurki2022finngen}) and for four continuous outcomes from Biobank Japan \cite[]{sakaue2021cross}.
 
@@
 		Coronary artery disease (CAD) & \cite{nikpay2015comprehensive} & 129,014 & ~~9,455,778 & 1,052,200 & 0.982 \\
+		Vitamin D & \cite{jiang2018genome} & ~~79,366 & ~~2,579,296 & 1,016,935 & NA \\
 		\hline
@@
 PRCA summary statistics have many variants with a missing INFO score, which we discard.
-We also restrict to variants with an INFO score larger than 0.4.\label{tab:sumstats}}
+We also restrict to variants with an INFO score larger than 0.4.
+Vitamin D summary statistics do not report INFO scores.
+\label{tab:sumstats}}
 \end{table}
@@
 
-\subsection{New LD reference}\label{sec:LD}
-
-We form nearly independent LD blocks using the optimal algorithm developed in \cite{prive2021optimal}.
-For different numbers of blocks and maximum number of variants in each block, we use the split with the minimum cost within the ones reducing the original number of non-zero values to less than 60\% (70\% for chromosome 6).
-Having a correlation matrix with independent blocks prevents small errors in the algorithm (e.g.\ the Gibbs sampler in LDpred2) from propagating to too many variants.
-It also makes running LDpred2 (and lassosum2) faster, taking about 60\% of the initial time (since only 60\% of the initial non-zero values of the correlation matrix are kept).
-
-We have also developed a new ``compact'' format for the SFBMs (sparse matrices on disk).
-Instead of using something similar to the standard ``compressed sparse column'' format which stores all $\{i,~x(i, j)\}$ for a given column $j$, we only store the first index $i_0$ and all the contiguous values $\{x(i_0, j),~x(i_0 + 1, j),~\dots\}$ up to the last non-zero value for this column $j$.
-This makes this format about twice as efficient for both LDpred2 and lassosum2 (in terms of both memory and speed).
-Therefore, using both this new format and the LD blocks should divide computation times by 3.
-
 \subsection{LDpred2-low-h2 and LDpred2-auto-rob}\label{sec:robust}
@@
 This is useful to prevent instability (oscillation and ultimately divergence) of the Gibbs sampler under large misspecifications, and is also useful for accelerating convergence of chains with a large initial value for $p$ (the proportion of causal variants).
+
+\subsection{New LD reference}\label{sec:LD}
+
+We form nearly independent LD blocks using the optimal algorithm developed in \cite{prive2021optimal}.
+For different numbers of blocks and maximum number of variants in each block, we use the split with the minimum cost within the ones reducing the original number of non-zero values to less than 60\% (70\% for chromosome 6).
+Having a correlation matrix with independent blocks prevents small errors in the algorithm (e.g.\ the Gibbs sampler in LDpred2) from propagating to too many variants.
+It also makes running LDpred2 (and lassosum2) faster, taking about 60\% of the initial time (since only 60\% of the initial non-zero values of the correlation matrix are kept).
+
+We have also developed a new ``compact'' format for the SFBMs (sparse matrices on disk).
+Instead of using something similar to the standard ``compressed sparse column'' format which stores all $\{i,~x(i, j)\}$ for a given column $j$, we only store the first index $i_0$ and all the contiguous values $\{x(i_0, j),~x(i_0 + 1, j),~\dots\}$ up to the last non-zero value for this column $j$.
+This makes this format about twice as efficient for both LDpred2 and lassosum2 (in terms of both memory and speed).
+Therefore, using both this new format and the LD blocks should divide computation times by 3.
+
+\subsection{Alternative LD reference for FinnGen and Biobank Japan}\label{sec:altLD}
+
+To be used with GWAS summary statistics from FinnGen, we investigate three different LD reference panels.
+To define homogeneous ancestry groups in the UK Biobank, we include all individuals within a specific distance to a population center in the PCA space \cite[]{prive2021high,prive2021using}.
+We use either the 503 European (including 99 Finnish) individuals from the 1000 Genomes (1000G) data \cite[]{10002015global}, 404 Finnish-like UKBB + the 99 Finnish 1000G individuals (also 503 in total), or the 10,000 UK individuals from the validation set we use in this paper.
+
+To be used with GWAS summary statistics from Biobank Japan, we also investigate three different LD reference panels.
+We use either the 504 East Asian (including 104 Japanese) individuals from the 1000G, 400 Japanese-like UKBB + the 104 Japanese 1000G individuals (also 504 in total), or 2041 East Asian UKBB individuals (including the 400 previous Japanese individuals).
+We also construct versions of these LD references with independent LD blocks, as described in the previous section.
 

### 2022-04-05 paper/paper-misspec.tex
minor text update

@@
 \usepackage{geometry}
-\geometry{verbose,letterpaper,tmargin=2cm,bmargin=2cm,lmargin=1.5cm,rmargin=1.5cm}
+\geometry{verbose,letterpaper,tmargin=2cm,bmargin=2cm,lmargin=2cm,rmargin=2cm}
 \usepackage{rotating}
@@
 \usepackage{makecell}
+%\usepackage[font=small]{caption}
 
@@
 \let\tabend\endtabular
-\renewenvironment{tabular}{\begin{adjustbox}{max width=0.95\textwidth}\tabbeg}{\tabend\end{adjustbox}}
+\renewenvironment{tabular}{\begin{adjustbox}{max width=\textwidth}\tabbeg}{\tabend\end{adjustbox}}
 
@@
 
-\vspace*{7em}
+\vspace*{6em}
 
-\abstract{
-	Genome-wide association studies (GWAS) summary statistics come with uneven quality.
-Are they of good enough quality for performing follow-up analyses?
+\begin{abstract}
+Genome-wide association studies (GWAS) summary statistics come with uneven quality; are they of good enough quality for performing follow-up analyses?
 First, we present an overview of possible misspecifications that come with GWAS summary statistics.
-Can we detect these possible misspecifications in GWAS summary statistics and correct for them to e.g.
-improve predictive performance of polygenic scores?
-Then, we show that additional information such as imputation INFO scores, allele frequencies, and per-variant sample sizes are very useful information to report in GWAS summary statistics because they can be used to detect possible issues and correct for misspecifications in the GWAS summary statistics.
+Can we detect these possible misspecifications in GWAS summary statistics and correct for them to e.g.\ improve predictive performance of polygenic scores?
+Then, we show that additional information such as imputation INFO scores, allele frequencies, and per-variant sample sizes in GWAS summary statistics can be used to detect possible issues and correct for misspecifications in the GWAS summary statistics.
 Unfortunately, due to the lack of reporting standards for GWAS summary statistics, this additional information is not systematically reported.
-We also show how important it is to use well-matched LD references.
+We also show that using well-matched LD references can improve model fit and translate into more accurate prediction.
 Finally, we discuss how to make polygenic score methods, such as lassosum and LDpred2, more robust to these misspecifications to improve their predictive power.
-}
+\end{abstract}
 
@@
 Contrary to individual-level genotypes and phenotypes, summary statistics resulting from genome-wide association studies (GWAS) are widely available, and very large sample sizes can be obtained through meta-analysis \cite[]{yengo2022saturated}.
-GWAS summary statistics have been extensively used to derive polygenic scores (PGS), perform fine-mapping, and estimate a range of key genetic architecture parameters \cite[]{prive2021finding}.
+GWAS summary statistics have been extensively used to derive polygenic scores (PGS), perform fine-mapping, and estimate a range of key genetic architecture parameters \cite[]{pasaniuc2017dissecting,prive2021finding,chen2021improved}.
 However, GWAS summary statistics come with uneven imputation accuracy.
 There is also heterogeneity in the information made available in these summary statistics; for example, per-variant imputation INFO scores, sample sizes and allele frequencies are often missing.
-Moreover, many methods based on summary statistics use Bayesian models, which can be sensitive to model misspecifications \cite[]{WALKER20131621,miller2018robust}.
+Moreover, many methods based on summary statistics use Bayesian models and iterative algorithms, which can be sensitive to model misspecifications \cite[]{WALKER20131621,miller2018robust}.
 
-We present an overview of possible misspecifications that come with GWAS summary statistics (Table \ref{tab:overview}). 
+We present an overview of possible misspecifications that come with GWAS summary statistics in Table \ref{tab:overview}. 
 First, the total sample size used can be misestimated, which would result in a biased estimation of the SNP heritability.
@@
 This is useful to check that the input parameters used are consistent with each other. 
-This was particularly important for LDpred2-auto, which directly estimates key model parameters from the data \cite[]{prive2020ldpred2}.
+This was particularly important for LDpred2-auto, which directly estimates two key model parameters, the SNP heritability and polygenicity, from the data \cite[]{prive2020ldpred2}.
 Here we further show that standard deviations of imputed genotypes (allele dosages) are lower than the expected values under Hardy--Weinberg equilibrium.
@@
 We then divide the 40,000 variants into three groups: for half of the variants, we use 100\% of 300,000 individuals for GWAS, but use only 80\% for one quarter of the variants and 60\% for the remaining quarter.
-We then run C+T, lassosum, lassosum2 (Methods), LDpred2-inf, LDpred2(-grid), and LDpred2-auto \cite[]{prive2019making,mak2017polygenic,prive2020ldpred2} by using either the true per-variant GWAS sample sizes, the total sample size for all variants, or imputed (per-variant) sample sizes.
+We then run C+T, lassosum, lassosum2 (Methods section \ref{lassosum2}), LDpred2-inf, LDpred2(-grid), and LDpred2-auto \cite[]{prive2019making,mak2017polygenic,prive2020ldpred2} by using either the true per-variant GWAS sample sizes, the total sample size for all variants, or imputed (per-variant) sample sizes.
 Note that we initially included PRS-CS and SBayesR in our comparison \cite[]{ge2019polygenic,lloyd2019improved}.
@@
 \begin{figure}[!h]
-	\centerline{\includegraphics[width=0.98\textwidth]{simu-misN}}
+	\centerline{\includegraphics[width=\textwidth]{simu-misN}}
 	\caption{Results for the simulations with sample size misspecification, averaged over 10 simulations for each scenario.
 Reported 95\% confidence intervals are computed from 10,000 non-parametric bootstrap replicates of the mean.
-The GWAS sample size is either ``true'' when providing the true per-variant sample size, ``max'' when providing instead the maximum sample size as a unique value to be used for all variants, ``imputed'' (Methods), or ``any'' when the method does not use this information (the case for C+T).
-Red bars correspond to using the LD with independent blocks (Methods).}
+The GWAS sample size is either ``true'' when providing the true per-variant sample size, ``max'' when providing instead the maximum sample size as a unique value to be used for all variants, ``imputed'' (cf.\ equation \eqref{eq:imputeN}), or ``any'' when the method does not use this information (the case for C+T).
+Red bars correspond to using the LD with independent blocks (Methods section \ref{sec:LD}).}
 	\label{fig:simu-misN}
@@
 First, the results for LDpred2-auto are similar to LDpred2-inf because it always converges to an infinitesimal model ($p = 1$) in these simulations.
-To overcome this limitation, we introduce two new parameters to make LDpred2-auto more robust, and refer to this as ``LDpred2-auto-rob'' here (Methods).
+To overcome this limitation, we introduce two new parameters to make LDpred2-auto more robust, and refer to this as ``LDpred2-auto-rob'' here (Methods section \ref{sec:robust}).
 The first parameter prevents $p$ from diverging to $1$, while the second shrinks the off-diagonal elements of the LD matrix (a form of regularization). 
@@
 Results with LDpred2-low-h2 improves from 0.159 to 0.169 when using true sample sizes and from 0.134 to 0.163 when using the maximum sample size.
-Finally, we introduce a last change for robustness here: we form independent LD blocks in the LD matrix to prevent small errors in the Gibbs sampler to propagate to too many variants (Methods).
+Finally, we introduce a last change for robustness here: we form independent LD blocks in the LD matrix to prevent small errors in the Gibbs sampler to propagate to too many variants (Methods section \ref{sec:LD}).
 This change seems to solve convergence issues of LDpred2 in these simulations (Figure \ref{fig:ldpred2-misN}) and further improves predictive performance for all LDpred2 methods (Figure \ref{fig:simu-misN}).
@@
 \begin{figure}[!h]
-	\centerline{\includegraphics[width=0.98\textwidth]{simu-info}}
+	\centerline{\includegraphics[width=\textwidth]{simu-info}}
 	\caption{Results of predictive performance for the simulations using GWAS summary statistics from imputed dosage data, averaged over 10 simulations for each scenario.
@@
 Correction ``in\_between'' corresponds to using $\hat{\gamma}_j^{\text{imp}} \cdot \text{INFO}_j$, $\text{se}(\hat{\gamma}_j)^{\text{imp}} \cdot \sqrt{\text{INFO}_j}$, and $n_j \cdot \text{INFO}_j$.
-Red bars correspond to using the LD with independent blocks (Methods).}
+Red bars correspond to using the LD with independent blocks (Methods section \ref{sec:LD}).}
 	\label{fig:simu-info}
@@
 \begin{figure}[!h]
-	\centerline{\includegraphics[width=0.9\textwidth]{simu-altpop}}
+	\centerline{\includegraphics[width=\textwidth]{simu-altpop}}
 	\caption{Results for the simulations with summary statistics with LD matrices based on two different populations.
@@
 Reported 95\% confidence intervals are computed from 10,000 non-parametric bootstrap replicates of the mean.
-Red bars correspond to using the LD with independent blocks (Methods).}
+Red bars correspond to using the LD with independent blocks (Methods section \ref{sec:LD}).}
 	\label{fig:simu-altpop}
@@
 
-We download and use GWAS summary statistics from five disease endpoints from FinnGen (release 6, \cite{kurki2022finngen}), namely breast and prostate cancers (BrCa and PrCa), coronary artery disease (CAD), type 1 and type 2 diabetes (T1D and T2D).
-These have been derived using SAIGE \cite[]{zhou2018efficiently}.
+In this section, we transition to using real data and investigate the use of different LD reference panels with GWAS summary statistics from two large biobanks of isolated populations.
+We download and use GWAS summary statistics for five disease endpoints from FinnGen (release 6, \cite{kurki2022finngen}), namely breast and prostate cancers (BrCa and PrCa), coronary artery disease (CAD), type 1 and type 2 diabetes (T1D and T2D).
+These were derived using SAIGE \cite[]{zhou2018efficiently}.
 First, for each phenotype, we estimate a global effective sample using the 80th percentile of imputed sample sizes from equation \eqref{eq:imputeNeff}.
 This estimation is only 84.7\% for BrCa, 79.1\% for PrCa, 73.1\% for T1D, 64.5\% for T2D, and 61.3\% for CAD, when compared to the effective sample computed from the reported numbers of cases and controls.
-We believe this reduction in effective sample size is due to both having related individuals as well as using SAIGE, as noted in the introduction.
+We believe this reduction in effective sample size is due to both having related individuals included in the analyses, as well as using SAIGE, as noted in the introduction.
 We then compare three different LD reference panels to use with these GWAS summary statistics (Methods section \ref{sec:altLD}).
-Using the small Finnish LD reference panel composed of 503 individuals only seems to consistently provide more predictive polygenic scores than using the other two LD references, including the large UK one composed of 10,000 individuals (Figure \ref{fig:finngen}).
-Note that the PGS are validated in people of UK-like ancestry, so we would expect larger improvements if validating in a Finnish population.
+Interestingly, using the small Finnish LD reference panel composed of 503 individuals only we have defined here seems to consistently provide more predictive polygenic scores than using the large UK one composed of 10,000 individuals or the widely used European subset of the 1000 Genomes (1000G) data (Figure \ref{fig:finngen}).
+Note that the PGS here are validated in people of UK-like ancestry in the UK biobank to obtain sufficient sample sizes.
 
@@
 	\centerline{\includegraphics[width=0.85\textwidth]{res-FIN}}
-	\caption{Results for PGS derived from five FinnGen GWAS summary statistics and using three different LD reference.
+	\caption{Results for PGS derived from five FinnGen GWAS summary statistics and using three different LD references.
 Partial correlations are computed using function \texttt{pcor} of R package bigstatsr where 95\% confidence intervals are obtained through Fisher's Z-transformation, then all values are squared to report the phenotypic variance explained by PGS.}
@@
 
-We also download and use GWAS summary statistics from four continuous outcomes from Biobank Japan \cite[]{sakaue2021cross}.
-These have been derived using BOLT-LMM-inf \cite[]{loh2018mixed}.
+We also download and use GWAS summary statistics for four continuous outcomes from Biobank Japan \cite[]{sakaue2021cross}.
+These were derived using BOLT-LMM-inf \cite[]{loh2018mixed}.
 First, for each phenotype, we estimate a global power improvement provided by BOLT-LMM by comparing $\chi^2$-statistics (the boost of BOLT-LMM-inf versus linear regression) across genome-wide significant variants, as recommended in \cite{loh2018mixed}.
-This estimated power ratio (increase in effective sample size) is 1.143 for height, 1.039 for HDL cholesterol, 1.025 for BMI, and 1.013 for systolic blood pressure.
+This estimated power ratio, which corresponds to the increase in effective sample size, is 1.143 for height, 1.039 for HDL cholesterol, 1.025 for BMI, and 1.013 for systolic blood pressure.
 We then compare three different LD reference panels to use with these GWAS summary statistics (Methods section \ref{sec:altLD}).
-Interesting, when looking at height and to some extent at HDL cholesterol where we could get the best predictive performance, using the smaller UKBB Japanese LD reference seems to provide more predictive PGS than using the other two LD references, including the one using a larger set of East Asian individuals from the UKBB (Figure \ref{fig:bbj}).
-Note that the PGS are validated in people of broad East Asian ancestry, so we would expect larger improvements if validating in a Japanese population.
+Interestingly, when looking at height and to some extent at HDL cholesterol where we could get the best predictive performance, using the smaller Japanese LD reference we have defined here seems to provide more predictive PGS than using the one using a larger set of East Asian individuals from the UKBB or the widely used East Asian subset of the 1000 Genomes (1000G) data (Figure \ref{fig:bbj}).
+Note that the PGS here are validated in people of broad East Asian ancestry in the UK biobank and for continuous outcomes  to obtain sufficient sample sizes.
 
@@
 
-Breast cancer summary statistics are interesting because they include results from two mega analyses \cite[]{michailidou2013large,michailidou2015genome,michailidou2017association}, which means that parameters reported in these GWAS summary statistics, such as INFO scores and sample sizes, are estimated with high precision.
+Breast cancer GWAS summary statistics are interesting because they include results from two mega analyses \cite[]{michailidou2013large,michailidou2015genome,michailidou2017association}, which means that parameters reported in these GWAS summary statistics, such as INFO scores and sample sizes, are estimated with high precision.
 Imputation INFO scores for the OncoArray summary statistics are generally very good (mean of 0.968 after having restricted to HapMap3 variants, Figure \ref{fig:hist-info-brca}) and better than the ones from iCOGS (mean of 0.841, Figure \ref{fig:hist-info-brca2}), probably because the iCOGS chip included around 200K variants only, compared to more than 500K variants for the OncoArray.
@@
 To validate this hypothesis, we simulate a phenotype using HapMap3 variants of chromosome 6 for 10,000 individuals from the UK Biobank, then we run GWAS with or without PC19 as covariate.
-PC19 from the UK Biobank was previously reported to capture LD structure in region 70-91 Mbp of chromsome 6 \cite[]{prive2020efficient}.
+PC19 from the UK Biobank was previously reported to capture LD structure in region 70-91 Mbp of chromosome 6 \cite[]{prive2020efficient}.
 In these simulations, the same bias as in figure \ref{fig:brca_onco_qc}B is observed for the variants in this region (Figure \ref{fig:gwas_bad_pc}), confirming our hypothesis.
@@
 \begin{figure}[!h]
-	\centerline{\includegraphics[width=0.98\textwidth]{brca_onco_qc}}
+	\centerline{\includegraphics[width=\textwidth]{brca_onco_qc}}
 	\caption{Standard deviations inferred from the OncoArray breast cancer GWAS summary statistics using equation \ref{eq:approx-sd-log} (\textbf{A:} Raw or \textbf{B:} dividing them by $\sqrt{\text{INFO}}$) versus the ones inferred from the reported GWAS allele frequencies $f_j$ (using $\sqrt{2 f_j (1 - f_j)}$).
@@
 
-We use other external GWAS summary statistics for which INFO scores are reported (Table \ref{tab:sumstats}); they all have a very high mean INFO score (larger than 0.94), except for T1D-affy (0.885).
+We use other external GWAS summary statistics for which INFO scores are reported (see Table \ref{tab:sumstats}); they all have a very high mean INFO score (larger than 0.94), except for T1D-affy (0.885).
 QC plots comparing standard deviations usually show little deviation from the identity line (after the INFO score correction), except for coronary artery disease (CAD) summary statistics (Figures \ref{fig:brca_icogs_qc}-\ref{fig:t1d_illu_qc}).
@@
 
-We also investigate results for vitamin D GWAS summary statistics, which do not report INFO scores nor allele frequencies, as opposed to previous ones.
+Lastly, we also investigate results for vitamin D GWAS summary statistics, which do not report INFO scores nor allele frequencies, as opposed to previous ones.
 The QC procedure is then less precise and uses allele frequencies from the LD reference (Figure \ref{fig:vitaminD_qc});
 However, these summary statistics do report per-variant sample sizes, which cover a wide range of different values (Figure \ref{fig:hist-N-vitaminD}).
-Here we test using either the maximum sample size or the true per-variant sample sizes when deriving polygenic scores, and an additional QC step ``qc2'' which refers to removing all variants with a per-variant sample size less than 70\% of the maximum one.
+Here we compare using either the maximum sample size or the true per-variant sample sizes when deriving polygenic scores, and an additional QC step ``qc2'' which refers to removing all variants with a per-variant sample size less than 70\% of the maximum one.
 When the true per-variant sample sizes are used, the additional ``qc2'' does not seem to be necessary (Figure \ref{fig:res_vitaminD}), which is a bit surprising to us.
@@
 
-We have investigated misspecifications in GWAS summary statistics, focusing particularly on the impact of sample size heterogeneity and imputation quality, and the application to polygenic scores (PGS) methods.
+Here we have investigated misspecifications in GWAS summary statistics, focusing particularly on the impact of sample size heterogeneity and imputation quality, and the application to polygenic scores (PGS) methods.
 Previously, we proposed a quality control (QC) based on comparing standard deviations (of genotypes) inferred from GWAS summary statistics with the ones computed from a reference panel \cite[]{prive2020ldpred2}.
@@
 Sometimes, it can even be challenging to recover the total effective sample size from large meta-analyses.
-We recall that if some studies of a meta-analysis have an imbalanced number of cases and controls, the total effective sample size should not be computed from the total numbers of cases and controls overall, but instead from the sum of the effective sample sizes of each study.
+We recall that, when some studies have an imbalanced number of cases and controls, the total effective sample size of their meta-analysis should not be computed from the total numbers of cases and controls overall, but instead from the sum of the effective sample sizes of each study \cite[]{grotzinger2021pervasive}.
 Indeed, take the extreme example of meta-analyzing two studies, one with 1000 cases and 0 controls, and another one with 0 cases and 1000 controls, then the effective sample size of the meta-analysis is 0, not 2000.
@@
 Overall, these changes provided large improvements of predictive performance in the simulations with large misspecifications.
-However, they had little to no effect when using the real GWAS summary statistics we chose, except for breast cancer and coronary artery disease summary statistics.
-Although these results are somewhat unfortunate, they are reassuring because it means that the summary statistics we use in this study are usually of good quality for follow-up analyses such as deriving polygenic scores.
+The proposed quality controls on GWAS summary statistics also provided better predictive performance for CAD, T1D, and vitamin D.
+However, these changes had limited effect when applied to some other real GWAS summary statistics, which is both unfortunate but also reassuring because it means that these GWAS summary statistics are of particularly good quality for follow-up analyses such as deriving polygenic scores.
 
-Nevertheless, we recommend to adopt these changes, i.e.\ performing the (refined) QC proposed here, forming independent LD blocks in the LD matrix, and using more regularization when needed. 
-We have shown other useful results, such as that filtering small sample sizes is very useful, as well as using a well-matched LD reference panel. 
-More regularization can be achieved by testing additional smaller values for the heritability parameter in LDpred2-grid, and using the two new parameters introduced here in LDpred2-auto (Section \ref{sec:robust}).
-Note that LD blocks are already widely used by several methods (such as lassosum and PRS-CS) because they allow for processing smaller matrices at once \cite[]{mak2017polygenic,ge2019polygenic}; here we have shown that these blocks are also useful to make methods more robust.
+In conclusion, we recommend adopting these changes, i.e.\ performing the (refined) QC proposed here, forming independent LD blocks in the LD matrix, and using more regularization when needed. 
+We also recommend using well-matched LD reference panels. 
+More regularization can be achieved by testing additional smaller values for the heritability parameter in LDpred2-grid, and using the two new parameters introduced here in LDpred2-auto (Methods section \ref{sec:robust}).
+Note that LD blocks are already widely used by several methods, such as lassosum and PRS-CS, because they allow for processing smaller matrices at once \cite[]{mak2017polygenic,ge2019polygenic}.
+Imposing independent blocks on the LD matrix could result in further misspecifications \cite[]{zhou2021fast}, but here we have shown that well defined blocks can actually make PGS methods more robust.
 PRS-CS is currently one of the most robust PGS methods; for example, it can use the (small) 1000 Genomes data as LD reference \cite[]{ge2019polygenic}, and can even use a European LD reference panel with multi-ancestry GWAS summary statistics \cite[]{wang2021global}.
-We think this is made possible by the use of a (very) strong regularization in PRS-CS ($\phi^{-1} \psi_j^{-1} \ge 1$, which would approximately correspond to using $s=0.5$ in lassosum, $\delta = 1$ in lassosum2, and $\texttt{shrink\_corr}=0.5$ in LDpred2-auto).
+We believe this is made possible by the use of a strong regularization in PRS-CS ($\phi^{-1} \psi_j^{-1} \ge 1$, which would approximately correspond to using $s=0.5$ in lassosum, $\delta = 1$ in lassosum2, and the new parameter $\texttt{shrink\_corr}=0.5$ in LDpred2-auto).
 Using enough regularization is good for robustness, but note that using too much regularization can also damage predictive performance.
 To address this limitation, we recommend to rely on a proper QC and choice of ancestry-matched LD instead of on too much regularization.
-We would like to invite large biobanks, such as FinnGen and Biobank Japan, to provide LD reference matrices matching the large GWAS summary statistics they provide, ideally based on the same large number of individuals.
-As a future research direction, we are interested in using multi-ancestry-matched LD matrices to improve polygenic prediction in all ancestries.
+We would like to encourage large biobanks, such as FinnGen and Biobank Japan, to provide LD reference matrices matching the large GWAS summary statistics they provide, ideally based on the same large number of individuals.
+As a future research direction, we are interested in using multi-ancestry-matched LD matrices to use with multi-ancestry GWAS summary statistics in order to improve polygenic prediction in all ancestries.
 It would also be useful to investigate the impact of the QC, well-matched LD reference panels and other adjustments we propose here on other (non-PGS) methods, for which consistency and robustness are likely to be very important as well \cite[]{chen2021improved}.
@@
 
-We also use the UK Biobank data, and use the same individuals as described in the previous section.
+We also use the UK Biobank data for validation/testing in real data analyses, and use the same individuals as described in the previous section.
 We sample 10,000 individuals to form a validation set and use the remaining 352,307 individuals as test set.
@@
 In practice, we also bound this estimate to be between $0.5 \cdot N$ and $1.1 \cdot N$, where $N$ is the total (effective) sample size.
-Also note that the estimate of $\text{var}(G_j)$ has to account for the imputation accuracy (Section \ref{sec:imputation}).
+Also note that the estimate of $\text{var}(G_j)$ has to account for the imputation accuracy, i.e.\ use $2 \cdot f_j \cdot (1 - f_j) \cdot \text{INFO}_j$ (Section \ref{sec:imputation}).
 
@@
 Also note that, as in LDpred2, we run lassosum2 genome-wide using a sparse correlation matrix which assumes that variants further away than 3 cM are not correlated.
-Although we do not require splitting the genome into independent LD blocks anymore (as done in lassosum), we recommend to do so for robustness in LDpred2, and for extra speed gain (cf.\ next section).
+Although we do not require splitting the genome into independent LD blocks anymore (as done in lassosum), we recommend to do so for robustness in LDpred2, and for extra speed gain (cf.\ Methods section \ref{sec:LD}).
 
@@
 Here we introduce the small changes made to LDpred2 (-grid and -auto) in order to make them more robust.
-First, LDpred2-low-h2 simply consists in running LDpred2-grid by testing $h^2$ within $\{0.3, 0.7, 1, 1.4\} \cdot h^2_\text{LDSC}$ (note the added $0.3$ compared to \cite{prive2020ldpred2}), where $h^2_\text{LDSC}$ is the heritability estimate from LD score regression.
+First, LDpred2-low-h2 simply consists in running LDpred2-grid while testing $h^2$ within $\{0.3, 0.7, 1, 1.4\} \cdot h^2_\text{LDSC}$ (note the added $0.3$ compared to \cite{prive2020ldpred2}), where $h^2_\text{LDSC}$ is the heritability estimate from LD score regression.
 Indeed, we show in simulations here that using lower values for $h^2$ may provide higher predictive performance in the case of misspecifications (thanks to more shrinkage of the effects).
@@
 The first one, \texttt{shrink\_corr}, allows for shrinking off-diagonal elements of the correlation matrix.
-This is similar to parameter `s' in lassosum (Section \ref{lassosum2}), and acts as a means of regularization.
-We use a value of 0.9 in simulations and 0.95 in real data when running ``LDpred2-auto-rob'' (and the default value of 1, with no effect, when running ``LDpred2-auto'').
+This is similar to parameter `s' in lassosum (Methods section \ref{lassosum2}), and acts as a means of regularization.
+We use a value of 0.9 in simulations and 0.95 in real data when running ``LDpred2-auto-rob'' here, and the default value of 1 (with no effect) when running ``LDpred2-auto''.
 The second new parameter, \texttt{allow\_jump\_sign}, controls whether variant effect sizes can change sign over two consecutive iterations of the Gibbs sampler.
@@
 
-The latest version of R package bigsnpr can be installed from GitHub, and a recent enough version from CRAN.
-Two tutorials on running LDpred2 and lassosum2 using R package bigsnpr are available at \url{https://privefl.github.io/bigsnpr/articles/LDpred2.html} and \url{https://privefl.github.io/bigsnpr-extdoc/polygenic-scores-pgs.html}.
+The latest version of R package bigsnpr can be installed from GitHub, and a recent enough version can be installed from CRAN.
+A tutorial on running LDpred2 and lassosum2 using R package bigsnpr is available at \url{https://privefl.github.io/bigsnpr/articles/LDpred2.html}.
 
 We have recomputed the allele frequencies and imputation INFO scores within the European subset used here and across all imputed variants in the UK Biobank, and made them available at \url{https://doi.org/10.6084/m9.figshare.16635388}.
-We have formed independent LD blocks within the European LD references provided in \cite{prive2020ldpred2}, and made these updated versions available at \url{https://doi.org/10.6084/m9.figshare.19213299}.
+We have formed independent LD blocks within the Northwestern European LD references provided in \cite{prive2020ldpred2}, and made these updated versions available at \url{https://doi.org/10.6084/m9.figshare.19213299}.
 

### 2022-04-06 paper/paper-misspec-supp.tex
minor text update
update text

@@
 
+\FloatBarrier
+
+\begin{figure}[p]
+	\centerline{\includegraphics[width=0.95\textwidth]{compare-sd}}
+	\caption{\textbf{A:} Raw standard deviations (SDs) from imputed dosages or \textbf{B:} corrected SDs from imputed dosages (dividing them by $\sqrt{\text{INFO}}$) versus SDs from ``true'' genotype calls, colored by INFO scores.
+\label{fig:compare-sd}}
+\end{figure}
+
+\begin{figure}[p]
+	\centerline{\includegraphics[width=0.95\textwidth]{compare-beta}}
+	\caption{GWAS effect sizes ($\hat{\gamma}$) from ``true'' genotype calls versus either \textbf{A:} raw $\hat{\gamma}$ from imputed dosages or \textbf{B:} corrected $\hat{\gamma}$ from imputed dosages (multiplying them by $\sqrt{\text{INFO}}$), colored by INFO scores.
+\label{fig:compare-beta}}
+\end{figure}
+
+\begin{figure}[p]
+\centerline{\includegraphics[width=0.95\textwidth]{compare-beta-se}}
+\caption{Standard errors of GWAS effect sizes (SEs) from ``true'' genotype calls versus either \textbf{A:} raw SEs from imputed dosages or \textbf{B:} corrected SEs from imputed dosages (multiplying them by $\sqrt{\text{INFO}}$), colored by INFO scores.
+\label{fig:compare-beta-se}}
+\end{figure}
+
+\begin{figure}[p]
+\centerline{\includegraphics[width=0.95\textwidth]{compare-mult-imp}}
+\caption{(Over)corrected GWAS \textbf{A:} Z-scores and \textbf{B:} effect sizes from imputed dosages (multiplying them by $\text{INFO}$) versus the ones obtained from multiple imputation (MI, 20 draws used, \cite{Palmer2016}), colored by INFO scores.
+This shows only one tenth of the simulated variants, because MI can be computationally intensive.
+\label{fig:compare-mi}}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\FloatBarrier
+
 \begin{figure}[p]
@@
 \begin{figure}[p]
-	\centerline{\includegraphics[width=0.95\textwidth]{compare-sd}}
-	\caption{\textbf{A:} Raw standard deviations (SDs) from imputed dosages or \textbf{B:} corrected SDs from imputed dosages (dividing them by $\sqrt{\text{INFO}}$) versus SDs from ``true'' genotype calls, colored by INFO scores.
-\label{fig:compare-sd}}
-\end{figure}
-
-\begin{figure}[p]
-	\centerline{\includegraphics[width=0.95\textwidth]{compare-beta-se}}
-	\caption{Standard errors of GWAS effect sizes (SEs) from ``true'' genotype calls versus either \textbf{A:} raw SEs from imputed dosages or \textbf{B:} corrected SEs from imputed dosages (multiplying them by $\sqrt{\text{INFO}}$), colored by INFO scores.
-\label{fig:compare-beta-se}}
-\end{figure}
-
-\begin{figure}[p]
-	\centerline{\includegraphics[width=0.95\textwidth]{compare-beta}}
-	\caption{GWAS effect sizes ($\hat{\gamma}$) from ``true'' genotype calls versus either \textbf{A:} raw $\hat{\gamma}$ from imputed dosages or \textbf{B:} corrected $\hat{\gamma}$ from imputed dosages (multiplying them by $\sqrt{\text{INFO}}$), colored by INFO scores.
-\label{fig:compare-beta}}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{figure}[p]
-	\centerline{\includegraphics[width=0.95\textwidth]{compare-mult-imp}}
-	\caption{(Over)corrected GWAS \textbf{A:} Z-scores and \textbf{B:} effect sizes from imputed dosages (multiplying them by $\text{INFO}$) versus the ones obtained from multiple imputation (MI, 20 draws used, \cite{Palmer2016}), colored by INFO scores.
-This shows only one tenth of the simulated variants, because MI can be computationally intensive.
-\label{fig:compare-mi}}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\FloatBarrier
-
-\begin{figure}[p]
 	\centerline{\includegraphics[width=0.9\textwidth]{simu-rounded}}
@@
 	\label{fig:simu-rounded}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{figure}[p]
-	\centerline{\includegraphics[width=0.95\textwidth]{res-BBJ}}
-	\caption{Results for PGS derived from four Biobank Japan GWAS summary statistics and using three different LD reference.
-Partial correlations are computed using function \texttt{pcor} of R package bigstatsr where 95\% confidence intervals are obtained through Fisher's Z-transformation, then all values are squared to report the phenotypic variance explained by PGS.}
-	\label{fig:bbj}
 \end{figure}
@@
 	\caption{Variance explained of T1D in the UK Biobank by PGS derived from external summary statistics (Affymetrix).
-These are computed using function \texttt{pcor} of R package bigstatsr where 95\% confidence intervals are obtained through Fisher's Z-transformation; these values are then squared.}
+These are computed using function \texttt{pcor} of R package bigstatsr where 95\% confidence intervals are obtained through Fisher's Z-transformation; these values are then squared (while keeping the sign).}
 	\label{fig:res_t1d_aff}
@@
 
-%\clearpage
+\FloatBarrier
+
+\begin{figure}[p]
+	\centerline{\includegraphics[width=0.95\textwidth]{res-BBJ}}
+	\caption{Results for PGS derived from four Biobank Japan GWAS summary statistics and using three different LD references.
+Partial correlations are computed using function \texttt{pcor} of R package bigstatsr where 95\% confidence intervals are obtained through Fisher's Z-transformation, then all values are squared to report the phenotypic variance explained by PGS.}
+	\label{fig:bbj}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 

### 2022-04-06 paper/paper-misspec.tex
update text

@@
 \begin{abstract}
-Genome-wide association studies (GWAS) summary statistics come with uneven quality; are they of good enough quality for performing follow-up analyses?
+Publicly available genome-wide association studies (GWAS) summary statistics exhibit uneven quality, which can impact the validity of follow-up analyses.
 First, we present an overview of possible misspecifications that come with GWAS summary statistics.
-Can we detect these possible misspecifications in GWAS summary statistics and correct for them to e.g.\ improve predictive performance of polygenic scores?
-Then, we show that additional information such as imputation INFO scores, allele frequencies, and per-variant sample sizes in GWAS summary statistics can be used to detect possible issues and correct for misspecifications in the GWAS summary statistics.
+Then, in both simulations and real data analyses, we show that additional information such as imputation INFO scores, allele frequencies, and per-variant sample sizes in GWAS summary statistics can be used to detect possible issues and correct for misspecifications in the GWAS summary statistics.
+One important motivation for us is to improve the predictive performance of polygenic scores built from these summary statistics.
 Unfortunately, due to the lack of reporting standards for GWAS summary statistics, this additional information is not systematically reported.
@@
 First, the total sample size used can be misestimated, which would result in a biased estimation of the SNP heritability.
-For example, the total effective sample size is overestimated when computed from the total number of cases and controls from a meta-analysis of binary outcomes \cite[]{grotzinger2021pervasive}; using BOLT-LMM summary statistics can result in an increased effective sample size \cite[]{loh2018mixed,gazal2018functional}; using SAIGE on binary traits with a large prevalence can result in a reduced effective sample size \cite[]{zhou2018efficiently,mbatchou2021computationally}.
+For example, the total effective sample size is overestimated when computed from the total number of cases and controls from a meta-analysis of binary outcomes \cite[]{grotzinger2021pervasive}; using BOLT-LMM summary statistics can result in an increased effective sample size \cite[]{loh2018mixed,gazal2018functional}; and using SAIGE on binary traits with a large prevalence can result in a reduced effective sample size \cite[]{zhou2018efficiently,mbatchou2021computationally}.
 Second, per-variant sample sizes can vary substantially and be much smaller than the total sample size when meta-analyzing GWAS summary statistics from multiple cohorts with different sets of variants \cite[]{wang2021global}.
 Using very different per-variant sample sizes can be problematic for models implicitly assuming that summary statistics have all been derived from the same individuals \cite[]{zhu2017bayesian,zhou2021fast}.
-As a quality control (QC) step, one can e.g.\ remove SNPs with a sample size less than 0.67 times the 90th percentile of all sample sizes \cite[]{zheng2017ld}.
-Third, many genetic variants are imputed, which can cause some bias in summary statistics.
-Fourth, the imputation quality of each variant (e.g.\ INFO scores), often used in a QC step in statistical genetics analyses, can be severely misestimated (often overestimated) when computed from multi-ancestry individuals instead of a more homogeneous subset.
+Third, many genetic variants are imputed, with association statistics being reported for the allele dosages instead of the true alleles, which can lead to some bias in the summary statistics.
+Fourth, the imputation quality of each variant (e.g.\ INFO scores), often used in a quality control (QC) step in statistical genetics analyses, can be severely misestimated (often overestimated) when computed from multi-ancestry individuals instead of a more homogeneous subset.
 Fifth, errors such as allele inversions can be present in the summary statistics; QC is particularly important here.
-Sixth, there can be some mismatch between the GWAS summary statistics and the LD reference used.
+Sixth, many follow-up analyses require using linkage disequilibrium (LD) from a reference panel.
+There can be some mismatch between the GWAS summary statistics and the LD reference used, which can e.g.\ lead to suboptimal predictive performance for polygenic scores.
 
@@
 Here we further show that standard deviations of imputed genotypes (allele dosages) are lower than the expected values under Hardy--Weinberg equilibrium.
-Second, we investigate possible adjustments to apply to the input parameters of methods using summary statistics (focusing on PGS methods here), namely the reference LD (linkage disequilibrium) matrix, the GWAS effect sizes, their standard errors and corresponding sample sizes.
+Second, we investigate possible adjustments to apply to the input parameters of PGS methods using summary statistics, namely the reference LD (linkage disequilibrium) matrix, the GWAS effect sizes, their standard errors and corresponding sample sizes.
 For example, we show that GWAS effect sizes computed from imputed dosages are larger in magnitude compared to if computed from true genotypes.
@@
 We then divide the 40,000 variants into three groups: for half of the variants, we use 100\% of 300,000 individuals for GWAS, but use only 80\% for one quarter of the variants and 60\% for the remaining quarter.
-We then run C+T, lassosum, lassosum2 (Methods section \ref{lassosum2}), LDpred2-inf, LDpred2(-grid), and LDpred2-auto \cite[]{prive2019making,mak2017polygenic,prive2020ldpred2} by using either the true per-variant GWAS sample sizes, the total sample size for all variants, or imputed (per-variant) sample sizes.
+We then run C+T, lassosum, lassosum2 (Methods section \ref{lassosum2}), LDpred2-inf, LDpred2(-grid), and LDpred2-auto \cite[]{prive2019making,mak2017polygenic,prive2020ldpred2} by using either the true per-variant GWAS sample sizes, the total sample size for all variants, or per-variant sample sizes imputed using equation \eqref{eq:imputeN}.
 Note that we initially included PRS-CS and SBayesR in our comparison \cite[]{ge2019polygenic,lloyd2019improved}.
@@
 This is the first correction of summary statistics we consider in the simulations below.
-Note that we recompute INFO scores for the subset of 362,307 European individuals used in this paper since they can differ substantially from the ones reported by the UK Biobank for the whole data (Figures \ref{fig:compare-info} and \ref{fig:compare-maf-info}).
 As a second option, instead of using dosages to compute the GWAS summary statistics, it has been argued that using multiple imputation (MI) would be more appropriate \cite[]{Palmer2016}.
@@
 This is the second correction of summary statistics we implement in the simulations below, along with $n_j \cdot \text{INFO}_j$ as the new per-variant sample sizes. 
-Finally, we consider an in-between solution as a third correction, using $\hat{\gamma}_j = \hat{\gamma}_j^{\text{imp}} \cdot \text{INFO}_j$, $\text{se}(\hat{\gamma}_j) = \text{se}(\hat{\gamma}_j)^{\text{imp}} \cdot \sqrt{\text{INFO}_j}$, and $n_j \cdot \text{INFO}_j$.
+Finally, we consider an in-between solution as a third correction, using $\hat{\gamma}_j = \hat{\gamma}_j^{\text{imp}} \cdot \text{INFO}_j$, $\text{se}(\hat{\gamma}_j) = \text{se}(\hat{\gamma}_j)^{\text{imp}} \cdot \sqrt{\text{INFO}_j}$, and $n_j \cdot \text{INFO}_j$ as the per-variant sample sizes.
+Note that we have recomputed INFO scores for the subset of 362,307 European individuals used in this paper since they can differ substantially from the ones reported by the UK Biobank for the whole data (Figures \ref{fig:compare-info} and \ref{fig:compare-maf-info}).
 
-Using 40,000 variants from chromosome 22, we simulate quantitative phenotypes assuming a heritability of 20\% and 2000 causal variants using the ``true'' dataset (cf.\ Methods section ``Data for simulations'').
+To compare these three corrections, we conduct a simulation using 40,000 variants from chromosome 22, where we simulate quantitative phenotypes assuming a heritability of 20\% and 2000 causal variants using the ``true'' dataset (cf.\ Methods section ``Data for simulations'').
 We compute GWAS summary statistics from the dosage dataset and use these summary statistics to run LDpred2 and lassosum2 with either no correction of the summary statistics, or with one of the three corrections described above.
@@
 
-\subsection{Application to FinnGen and Biobank Japan summary statistics}
-
-In this section, we transition to using real data and investigate the use of different LD reference panels with GWAS summary statistics from two large biobanks of isolated populations.
-We download and use GWAS summary statistics for five disease endpoints from FinnGen (release 6, \cite{kurki2022finngen}), namely breast and prostate cancers (BrCa and PrCa), coronary artery disease (CAD), type 1 and type 2 diabetes (T1D and T2D).
-These were derived using SAIGE \cite[]{zhou2018efficiently}.
-First, for each phenotype, we estimate a global effective sample using the 80th percentile of imputed sample sizes from equation \eqref{eq:imputeNeff}.
-This estimation is only 84.7\% for BrCa, 79.1\% for PrCa, 73.1\% for T1D, 64.5\% for T2D, and 61.3\% for CAD, when compared to the effective sample computed from the reported numbers of cases and controls.
-We believe this reduction in effective sample size is due to both having related individuals included in the analyses, as well as using SAIGE, as noted in the introduction.
-We then compare three different LD reference panels to use with these GWAS summary statistics (Methods section \ref{sec:altLD}).
-Interestingly, using the small Finnish LD reference panel composed of 503 individuals only we have defined here seems to consistently provide more predictive polygenic scores than using the large UK one composed of 10,000 individuals or the widely used European subset of the 1000 Genomes (1000G) data (Figure \ref{fig:finngen}).
-Note that the PGS here are validated in people of UK-like ancestry in the UK biobank to obtain sufficient sample sizes.
-
-\begin{figure}[!h]
-	\centerline{\includegraphics[width=0.85\textwidth]{res-FIN}}
-	\caption{Results for PGS derived from five FinnGen GWAS summary statistics and using three different LD references.
-Partial correlations are computed using function \texttt{pcor} of R package bigstatsr where 95\% confidence intervals are obtained through Fisher's Z-transformation, then all values are squared to report the phenotypic variance explained by PGS.}
-	\label{fig:finngen}
-\end{figure}
-
-We also download and use GWAS summary statistics for four continuous outcomes from Biobank Japan \cite[]{sakaue2021cross}.
-These were derived using BOLT-LMM-inf \cite[]{loh2018mixed}.
-First, for each phenotype, we estimate a global power improvement provided by BOLT-LMM by comparing $\chi^2$-statistics (the boost of BOLT-LMM-inf versus linear regression) across genome-wide significant variants, as recommended in \cite{loh2018mixed}.
-This estimated power ratio, which corresponds to the increase in effective sample size, is 1.143 for height, 1.039 for HDL cholesterol, 1.025 for BMI, and 1.013 for systolic blood pressure.
-We then compare three different LD reference panels to use with these GWAS summary statistics (Methods section \ref{sec:altLD}).
-Interestingly, when looking at height and to some extent at HDL cholesterol where we could get the best predictive performance, using the smaller Japanese LD reference we have defined here seems to provide more predictive PGS than using the one using a larger set of East Asian individuals from the UKBB or the widely used East Asian subset of the 1000 Genomes (1000G) data (Figure \ref{fig:bbj}).
-Note that the PGS here are validated in people of broad East Asian ancestry in the UK biobank and for continuous outcomes  to obtain sufficient sample sizes.
-
 \subsection{Application to breast cancer summary statistics}
 
+In this section, we transition to using real data.
 Breast cancer GWAS summary statistics are interesting because they include results from two mega analyses \cite[]{michailidou2013large,michailidou2015genome,michailidou2017association}, which means that parameters reported in these GWAS summary statistics, such as INFO scores and sample sizes, are estimated with high precision.
@@
 
-\begin{figure}[p]
+\begin{figure}[!h]
 	\centerline{\includegraphics[width=0.7\textwidth]{res-cad}}
@@
 
-Lastly, we also investigate results for vitamin D GWAS summary statistics, which do not report INFO scores nor allele frequencies, as opposed to previous ones.
+We also investigate results for vitamin D GWAS summary statistics, which do not report INFO scores nor allele frequencies, as opposed to previous ones.
 The QC procedure is then less precise and uses allele frequencies from the LD reference (Figure \ref{fig:vitaminD_qc});
@@
 When the maximum sample size is used, this ``qc2'' is very important, especially for LDpred2-auto.
-Note that, when using the maximum sample size to derive ``sd\_ss'' in ``qc1'' (not the case here; we used the per-variant sample sizes), ``qc1'' would remove variants with underestimated standard deviations due to overestimated sample sizes, which would effectively remove variants with low sample sizes, similar to the ``qc2'' used here.
+Note that, when using the maximum sample size to derive ``sd\_ss'' in ``qc1'' (not the case here; we used the per-variant sample sizes), ``qc1'' would remove variants with underestimated standard deviations due to overestimated sample sizes, which would effectively remove variants with low sample sizes (Figure \ref{fig:qcplot}), similar to the ``qc2'' used here.
 
-\begin{figure}[p]
+\begin{figure}[!h]
 	\centerline{\includegraphics[width=0.7\textwidth]{res-vitaminD}}
@@
 \end{figure}
+
+\subsection{Application to FinnGen and Biobank Japan summary statistics}
+
+Here we investigate the use of different LD reference panels with GWAS summary statistics from two large biobanks of isolated populations.
+We download and use GWAS summary statistics for five disease endpoints from FinnGen (release 6, \cite{kurki2022finngen}), namely breast and prostate cancers (BrCa and PrCa), coronary artery disease (CAD), type 1 and type 2 diabetes (T1D and T2D).
+These were derived using SAIGE \cite[]{zhou2018efficiently}.
+First, for each phenotype, we estimate a global effective sample using the 80th percentile of imputed sample sizes from equation \eqref{eq:imputeNeff}.
+This estimation is only 84.7\% for BrCa, 79.1\% for PrCa, 73.1\% for T1D, 64.5\% for T2D, and 61.3\% for CAD, when compared to the effective sample computed from the reported numbers of cases and controls.
+We believe this reduction in effective sample size is due to both having related individuals included in the analyses, as well as using SAIGE, as noted in the introduction.
+We then compare three different LD reference panels to use with these GWAS summary statistics (Methods section \ref{sec:altLD}).
+Interestingly, using the small Finnish LD reference panel composed of 503 individuals only we have defined here seems to consistently provide more predictive polygenic scores than using the large UK one composed of 10,000 individuals or the widely used European subset of the 1000 Genomes (1000G) data (Figure \ref{fig:finngen}).
+Note that the PGS here are validated in people of UK-like ancestry in the UK biobank to obtain sufficient sample sizes.
+
+\begin{figure}[!h]
+	\centerline{\includegraphics[width=0.85\textwidth]{res-FIN}}
+	\caption{Results for PGS derived from five FinnGen GWAS summary statistics and using three different LD references.
+Partial correlations are computed using function \texttt{pcor} of R package bigstatsr where 95\% confidence intervals are obtained through Fisher's Z-transformation, then all values are squared to report the phenotypic variance explained by PGS.}
+	\label{fig:finngen}
+\end{figure}
+
+We also download and use GWAS summary statistics for four continuous outcomes from Biobank Japan \cite[]{sakaue2021cross}.
+These were derived using BOLT-LMM-inf \cite[]{loh2018mixed}.
+First, for each phenotype, we estimate a global power improvement provided by BOLT-LMM by comparing $\chi^2$-statistics (the boost of BOLT-LMM-inf versus linear regression) across genome-wide significant variants, as recommended in \cite{loh2018mixed}.
+This estimated power ratio, which corresponds to the increase in effective sample size, is 1.143 for height, 1.039 for HDL cholesterol, 1.025 for BMI, and 1.013 for systolic blood pressure.
+We then compare three different LD reference panels to use with these GWAS summary statistics (Methods section \ref{sec:altLD}).
+Interestingly, when looking at height and to some extent at HDL cholesterol where we could get the best predictive performance, using the smaller Japanese LD reference we have defined here seems to provide more predictive PGS than using the one using a larger set of East Asian individuals from the UKBB or the widely used East Asian subset of the 1000 Genomes (1000G) data (Figure \ref{fig:bbj}).
+Note that the PGS here are validated in people of broad East Asian ancestry in the UK biobank and for continuous outcomes  to obtain sufficient sample sizes.
 

### 2022-04-08 paper/paper-misspec-supp.tex
update text

@@
 \begin{figure}[p]
-\centerline{\includegraphics[width=0.75\textwidth]{hist_N_vitaminD}}
-\caption{Histogram of the per-variant sample sizes in the vitamin D GWAS summary statistics. \label{fig:hist-N-vitaminD}}
+\centerline{\includegraphics[width=0.95\textwidth]{hist_N_vitaminD}}
+\caption{Histogram of the per-variant sample sizes in the vitamin D GWAS summary statistics.
+The vertical red line corresponds to 70\% of the maximum sample size, the threshold used in ``qc2''.
+\label{fig:hist-N-vitaminD}}
 \end{figure}
@@
 	\centerline{\includegraphics[width=0.95\textwidth]{compare-cor}}
-	\caption{Comparing correlations between a subset of HapMap3 variants on chromosome 1 with an INFO score lower than 0.85, computed in three different ways: 1/ from imputed dosages using 10,000 individuals from the UK Biobank (UKBB) data; 2/ from multiple imputation (i.e.\ generating multiple complete datasets sampled according to imputation probabilities, computing correlations, and averaging results) also using UKBB; 3/ from 190 individuals from the 1000 Genomes data (GBR and CEU).
+	\caption{Comparing pairwise correlations between a subset of HapMap3 variants on chromosome 1 with an INFO score lower than 0.85, computed in three different ways: 1/ (left y-axis) from imputed dosages using 10,000 individuals from the UK Biobank (UKBB) data; 2/ (right y-axis) from multiple imputation (i.e.\ generating multiple complete datasets sampled according to imputation probabilities, computing correlations, and averaging results) also using UKBB; 3/ (common x-axis) from 190 individuals from the 1000 Genomes data (GBR and CEU).
 Each point, representing the correlation between two variants, is colored by the geometric mean of the INFO scores of these two variants.}

### 2022-04-08 paper/paper-misspec.tex
update text

@@
 We also show that using well-matched LD references can improve model fit and translate into more accurate prediction.
-Finally, we discuss how to make polygenic score methods, such as lassosum and LDpred2, more robust to these misspecifications to improve their predictive power.
+Finally, we discuss how to make polygenic score methods such as lassosum and LDpred2 more robust to these misspecifications to improve their predictive power.
 \end{abstract}
@@
 
-Contrary to individual-level genotypes and phenotypes, summary statistics resulting from genome-wide association studies (GWAS) are widely available, and very large sample sizes can be obtained through meta-analysis \cite[]{yengo2022saturated}.
+Contrary to individual-level genotypes and phenotypes, summary statistics resulting from genome-wide association studies (GWAS) are widely available, and very large sample sizes can be obtained through meta-analyses \cite[]{yengo2022saturated}.
 GWAS summary statistics have been extensively used to derive polygenic scores (PGS), perform fine-mapping, and estimate a range of key genetic architecture parameters \cite[]{pasaniuc2017dissecting,prive2021finding,chen2021improved}.
@@
 		\hline
-		Misspecification & Possible consequence(s) & Possible correction \\
+		\textbf{Misspecification} & \textbf{Possible consequence(s)} & \textbf{Possible correction} \\
 		\hline
@@
 		\hline
-		Differences in per-variant sample sizes & Possible violation of model assumptions resulting in poor model convergence & Imputation using equations \ref{eq:imputeN} and \ref{eq:imputeNeff}, and filtering$^{(2)}$   \\
+		Differences in per-variant sample sizes & Possible violation of model assumptions resulting in poor model convergence & Imputation using equations \eqref{eq:imputeN} and \eqref{eq:imputeNeff}, and filtering$^{(2)}$   \\
 		\hline
@@
 This QC consists in comparing standard deviations (of genotypes) inferred from GWAS summary statistics with the ones computed from a reference panel.
-This is useful to check that the input parameters used are consistent with each other. 
+This is useful to check that the input parameters used are consistent with one another. 
 This was particularly important for LDpred2-auto, which directly estimates two key model parameters, the SNP heritability and polygenicity, from the data \cite[]{prive2020ldpred2}.
@@
 The GWAS sample size is either ``true'' when providing the true per-variant sample size, ``max'' when providing instead the maximum sample size as a unique value to be used for all variants, ``imputed'' (cf.\ equation \eqref{eq:imputeN}), or ``any'' when the method does not use this information (the case for C+T).
-Red bars correspond to using the LD with independent blocks (Methods section \ref{sec:LD}).}
+Red bars correspond to using the LD with independent blocks (Methods section \ref{sec:LD}); this strategy cannot be tested straightforwardly for C+T and lassosum.}
 	\label{fig:simu-misN}
@@
 
-\subsection{When using allele dosages from imputation}\label{sec:imputation}
+\subsection{Misspecification when using imputed allele dosages}\label{sec:imputation}
 
@@
 	\centerline{\includegraphics[width=\textwidth]{brca_onco_qc}}
-	\caption{Standard deviations inferred from the OncoArray breast cancer GWAS summary statistics using equation \ref{eq:approx-sd-log} (\textbf{A:} Raw or \textbf{B:} dividing them by $\sqrt{\text{INFO}}$) versus the ones inferred from the reported GWAS allele frequencies $f_j$ (using $\sqrt{2 f_j (1 - f_j)}$).
+	\caption{Standard deviations inferred from the OncoArray breast cancer GWAS summary statistics using equation \eqref{eq:approx-sd-log} (\textbf{A:} Raw or \textbf{B:} dividing them by $\sqrt{\text{INFO}}$) versus the ones inferred from the reported GWAS allele frequencies $f_j$ (using $\sqrt{2 f_j (1 - f_j)}$).
 Only 100,000 HapMap3 variants are represented, at random.}
@@
 \begin{figure}[!h]
-	\centerline{\includegraphics[width=0.7\textwidth]{res-cad}}
+	\centerline{\includegraphics[width=0.67\textwidth]{res-cad}}
 	\caption{Variance explained of CAD in the UK Biobank by PGS derived from external summary statistics.
-These are computed using function \texttt{pcor} of R package bigstatsr where 95\% confidence intervals are obtained through Fisher's Z-transformation; these values are then squared.}
+These are computed using function \texttt{pcor} of R package bigstatsr where 95\% confidence intervals are obtained through Fisher's Z-transformation; these values are then squared.
+Red bars correspond to using the LD with independent blocks (Methods).}
 	\label{fig:res_cad}
@@
 We also investigate results for vitamin D GWAS summary statistics, which do not report INFO scores nor allele frequencies, as opposed to previous ones.
-The QC procedure is then less precise and uses allele frequencies from the LD reference (Figure \ref{fig:vitaminD_qc});
+The QC procedure is then less precise and uses allele frequencies from the LD reference (Figure \ref{fig:vitaminD_qc}).
 However, these summary statistics do report per-variant sample sizes, which cover a wide range of different values (Figure \ref{fig:hist-N-vitaminD}).
@@
 \begin{figure}[!h]
-	\centerline{\includegraphics[width=0.7\textwidth]{res-vitaminD}}
+	\centerline{\includegraphics[width=0.67\textwidth]{res-vitaminD}}
 	\caption{Variance explained of vitamin D in the UK Biobank by PGS derived from external summary statistics.
-These are computed using function \texttt{pcor} of R package bigstatsr where 95\% confidence intervals are obtained through Fisher's Z-transformation; these values are then squared.}
+These are computed using function \texttt{pcor} of R package bigstatsr where 95\% confidence intervals are obtained through Fisher's Z-transformation; these values are then squared.
+Red bars correspond to using the LD with independent blocks (Methods).}
 	\label{fig:res_vitaminD}
@@
 Here we investigate the use of different LD reference panels with GWAS summary statistics from two large biobanks of isolated populations.
-We download and use GWAS summary statistics for five disease endpoints from FinnGen (release 6, \cite{kurki2022finngen}), namely breast and prostate cancers (BrCa and PrCa), coronary artery disease (CAD), type 1 and type 2 diabetes (T1D and T2D).
+We use GWAS summary statistics for five disease endpoints from FinnGen (release 6, \cite{kurki2022finngen}), namely breast and prostate cancers (BrCa and PrCa), coronary artery disease (CAD), type 1 and type 2 diabetes (T1D and T2D).
 These were derived using SAIGE \cite[]{zhou2018efficiently}.
@@
 	\caption{Results for PGS derived from five FinnGen GWAS summary statistics and using three different LD references.
-Partial correlations are computed using function \texttt{pcor} of R package bigstatsr where 95\% confidence intervals are obtained through Fisher's Z-transformation, then all values are squared to report the phenotypic variance explained by PGS.}
+Partial correlations are computed using function \texttt{pcor} of R package bigstatsr where 95\% confidence intervals are obtained through Fisher's Z-transformation, then all values are squared to report the phenotypic variance explained by PGS.
+Red bars correspond to using the LD with independent blocks (Methods).}
 	\label{fig:finngen}
@@
 
-We also download and use GWAS summary statistics for four continuous outcomes from Biobank Japan \cite[]{sakaue2021cross}.
+We also use GWAS summary statistics for four continuous outcomes from Biobank Japan \cite[]{sakaue2021cross}.
 These were derived using BOLT-LMM-inf \cite[]{loh2018mixed}.
@@
 Misspecifying the GWAS sample size can lead to serious issues such as misestimating the SNP heritability \cite[]{grotzinger2021pervasive}.
-Fortunately, an overestimated sample size can be detected from the QC plot we propose, where the slope is then less than 1 for case-control studies using logistic regression; otherwise the standard deviation of the phenotype is also needed (Equation \ref{eq:approx-sd-lin}), but can be estimated \cite[]{prive2020ldpred2}.
+Fortunately, an overestimated sample size can be detected from the QC plot we propose, where the slope is then less than 1 for case-control studies using logistic regression; otherwise the standard deviation of the phenotype is also needed (Equation \eqref{eq:approx-sd-lin}), but can be estimated \cite[]{prive2020ldpred2}.
 Here we have used this strategy to estimate reduced effective sample sizes in FinnGen GWAS summary statistics.

### 2022-04-13 code/compute-all-info-eur.R
reexport

@@ -107,14 +107,15 @@ plot_grid(
     theme_bw(14) +
     geom_point(alpha = 0.2) +
     labs(x = "INFO recomputed in European subset", y = "INFO reported in MFI files",
-         color = "MAF_reported / MAF_recomputed") +
+         color = expression(frac(MAF_reported, MAF_recomputed))) +
     coord_equal() +
     geom_abline(color = "red") +
     scale_color_viridis_c(trans = "log", breaks = 10^(-3:4)) +
-    theme(legend.position = "top") +
-    guides(color = guide_colorbar(barwidth = 12, ticks.linewidth = 2)),
+    theme(legend.position = c(0.75, 0.15), legend.title.align = 0.5) +
+    guides(color = guide_colorbar(barwidth = 12, ticks.linewidth = 1.5,
+                                  direction = "horizontal", title.position = "top")),
 
-  scale = 0.98, axis = "bl"
+  scale = 0.98
 )
 # ggsave("figures/compare-all-maf-info-chr22.png", width = 13, height = 12)
 

### 2022-04-13 code/investigate-lassosum2-auto.R

@@ -23,7 +23,7 @@ beta_lassosum <- snp_lassosum2(corr, df_beta, ncores = NCORES)
 
 # validation
 pred_lassosum <- big_prodMat(G, beta_lassosum, ncores = NCORES)
-params$score <- apply(pred_lassosum[ind.val, ], 2, cor, y = y[ind.val])
+params$score <- apply(pred_lassosum[ind.test, ], 2, cor, y = y[ind.test])
 
 # pseudo-validation
 scale <- with(df_beta, sqrt(n_eff * beta_se^2 + beta^2))

### 2022-04-13 code/investigate-misspec-N.R

@@ -39,21 +39,23 @@ for (k in 1:3) {
   df_beta$lpval[ind_set]   <- -predict(gwas)
 }
 
-df_beta2 <- df_beta; df_beta2$n_eff <- max(df_beta$n_eff)
+df_beta2 <- df_beta; df_beta2$n_eff <- max(df_beta$n_eff)  # using max(N)
 
 # Quality control plot from the LDpred2 paper
 sd_val <- sqrt(big_colstats(G, ind.row = ind.val, ncores = NCORES)$var)
-sd_ss <- 1 / with(df_beta2, sqrt(n_eff * beta_se^2 + beta^2))
+sd_ss <- 1 / with(df_beta2, sqrt(n_eff * beta_se^2 + beta^2))  # using max(N)
 sd_y <- sqrt(0.5) / max(sd_ss)
 sd_ss <- sd_y * sd_ss
 
-qplot(sd_val, sd_ss, alpha = I(0.6), color = as.factor(df_beta$n_eff)) +
+qplot(sd_val, sd_ss, alpha = I(0.6),
+      color = factor(df_beta$n_eff, levels = sort(unique(df_beta$n_eff), decreasing = TRUE))) +
   theme_bigstatsr(0.9) +
   coord_equal() +
   geom_abline(linetype = 2, color = "red") +
-  labs(x = "Standard deviations in the validation set", color = "N",
-       y = "Standard deviations derived from the summary statistics") +
-  theme(legend.position = c(0.25, 0.8))
+  labs(x = "Standard deviations in the validation set", color = "True per-variant\nsample size",
+       y = "Standard deviations derived from the summary statistics (using maxN)") +
+  theme(legend.position = c(0.25, 0.8)) +
+  guides(colour = guide_legend(override.aes = list(alpha = 1, size = 2)))
 # ggsave("figures/simu-qc-plot.png", width = 7, height = 7)
 
 # Imputation of N

### 2022-04-13 code/prepare-corr-1000G-EAS.R

@@ -1,0 +1,85 @@
+library(dplyr)
+library(bigreadr)
+library(bigsnpr)
+
+obj.1000G <- snp_attach("data/1000G_HM3.rds")
+G <- obj.1000G$genotypes
+ind_EAS <- which(obj.1000G$fam$`Super Population` == "EAS")
+
+
+#### Prepare correlations ####
+
+CHR <- obj.1000G$map$chromosome
+POS2 <- snp_attach("data/UKBB_HM3_val.rds")$map$genetic.dist
+
+library(future.batchtools)
+plan(batchtools_slurm(resources = list(
+  t = "12:00:00", c = 15, mem = "125g",
+  name = basename(rstudioapi::getSourceEditorContext()$path))))
+
+bigassertr::assert_dir("data/corr_1000G_EAS")
+
+all_final_grp <- furrr::future_map_dfr(1:22, function(chr) {
+
+  ind.chr <- which(CHR == chr)
+
+  library(bigsnpr)
+  corr0 <- runonce::save_run({
+    corr0 <- snp_cor(G, ind.row = ind_EAS, ind.col = ind.chr, infos.pos = POS2[ind.chr],
+                     size = 3 / 1000, ncores = nb_cores())
+    mean(is.na(corr0@x))  # 0.001 for chr 22
+    corr0@x <- ifelse(is.na(corr0@x) | is.infinite(corr0@x), 0, corr0@x)
+    corr0
+  }, file = paste0("data/corr_1000G_EAS/chr", chr, ".rds"))
+
+  all_splits <- runonce::save_run({
+
+    SEQ <- round(seq_log(1000 + ncol(corr0) / 50,
+                         5000 + ncol(corr0) / 10,
+                         length.out = 10))
+
+    library(furrr)
+    plan("multisession", workers = 5)
+    options(future.globals.maxSize = 10e9)
+
+    splits <- future_map_dfr(SEQ, function(max_size) {
+      res <- snp_ldsplit(corr0, thr_r2 = 0.05, min_size = 100,
+                         max_size = max_size, max_K = 200)
+      res$max_size <- max_size
+      res
+    })
+  }, file = paste0("tmp-data/split_corr_1000G_EAS_chr", chr, ".rds"))
+
+
+  library(ggplot2)
+  qplot(data = all_splits, perc_kept, cost, color = as.factor(max_size)) +
+    theme_bw(12) +
[…]

### 2022-04-13 code/prepare-corr-1000G-EUR.R

@@ -1,0 +1,84 @@
+library(dplyr)
+library(bigreadr)
+library(bigsnpr)
+
+obj.1000G <- snp_attach("data/1000G_HM3.rds")
+G <- obj.1000G$genotypes
+ind_EUR <- which(obj.1000G$fam$`Super Population` == "EUR")
+
+
+#### Prepare correlations ####
+
+CHR <- obj.1000G$map$chromosome
+POS2 <- snp_attach("data/UKBB_HM3_val.rds")$map$genetic.dist
+
+library(future.batchtools)
+plan(batchtools_slurm(resources = list(
+  t = "12:00:00", c = 15, mem = "125g",
+  name = basename(rstudioapi::getSourceEditorContext()$path))))
+
+bigassertr::assert_dir("data/corr_1000G_EUR")
+
+all_final_grp <- furrr::future_map_dfr(1:22, function(chr) {
+
+  ind.chr <- which(CHR == chr)
+
+  library(bigsnpr)
+  corr0 <- runonce::save_run({
+    corr0 <- snp_cor(G, ind.row = ind_EUR, ind.col = ind.chr, infos.pos = POS2[ind.chr],
+                     size = 3 / 1000, ncores = nb_cores())
+    mean(is.na(corr0@x))  # 0.001 for chr 22
+    corr0@x <- ifelse(is.na(corr0@x) | is.infinite(corr0@x), 0, corr0@x)
+    corr0
+  }, file = paste0("data/corr_1000G_EUR/chr", chr, ".rds"))
+
+  library(furrr)
+  plan("multisession", workers = 5)
+  options(future.globals.maxSize = 10e9)
+
+  all_splits <- runonce::save_run({
+    SEQ <- round(seq_log(1000 + ncol(corr0) / 50,
+                         5000 + ncol(corr0) / 10,
+                         length.out = 10))
+
+    splits <- future_map_dfr(SEQ, function(max_size) {
+      res <- snp_ldsplit(corr0, thr_r2 = 0.05, min_size = 100,
+                         max_size = max_size, max_K = 200)
+      res$max_size <- max_size
+      res
+    })
+  }, file = paste0("tmp-data/split_corr_1000G_EUR_chr", chr, ".rds"))
+
+
+  library(ggplot2)
+  qplot(data = all_splits, perc_kept, cost, color = as.factor(max_size)) +
+    theme_bw(12) +
[…]

### 2022-04-13 code/prepare-corr-FIN.R

@@ -1,0 +1,153 @@
+library(dplyr)
+library(bigreadr)
+
+#### Get 503 "Finnish" individuals from the UKBB + 1000G ####
+
+df_UKBB <- runonce::save_run({
+
+  # Country of birth
+  code_country   <- filter(fread2("UKBB/coding89.tsv"), selectable == "Y")
+
+  eid_imputed <- fread2("UKBB/ukb58024_imp_chr1_v3_s487296.sample")$ID_1[-1]
+
+  fread2(
+    "UKBB/ukb41181.csv",
+    select = c("eid", "20115-0.0", paste0("22009-0.", 1:16)),
+    col.names = c("eid", "country", paste0("PC", 1:16))
+  ) %>%
+    filter(eid %in% eid_imputed) %>%
+    mutate(country = factor(country, levels = code_country$coding,
+                            labels = code_country$meaning))
+
+}, file = "tmp-data/info-UKBB.rds")
+
+center_FIN <- df_UKBB %>%
+  filter(country == "Finland") %>%
+  select(PC1:PC16) %>%
+  bigutilsr::geometric_median()
+
+ldist_to_FIN <- df_UKBB %>%
+  select(PC1:PC16) %>%
+  { sweep(., 2, center_FIN, '-')^2 } %>%
+  rowSums() %>%
+  sqrt() %>%
+  log()
+
+hist(ldist_to_FIN, "FD")
+
+print_table <- function(x, min_count = 5) {
+  tab <- sort(table(x, exclude = NULL), decreasing = TRUE)
+  tab <- tab[tab >= min_count]
+  cat(paste0(names(tab), ": ", tab), sep = " || ")
+}
+
+# Remove withdrawn and related individuals
+eid_withdrawn <- scan("UKBB/w58024_20220222.csv")
+rel <- fread2("UKBB/ukb58024_rel_s488264.dat")
+ldist_to_FIN[df_UKBB$eid %in% c(eid_withdrawn, filter(rel, Kinship > 2^-2.5)$ID2)] <- Inf
+
+ind <- head(order(ldist_to_FIN), 503 - 99)
+print_table(df_UKBB$country[ind])
+# Finland: 150 || NA: 135 || Russia: 39 || Sweden: 20 || Germany: 19 ||
+# Poland: 8 || Estonia: 7 || USA: 7 || Norway: 5
+
+
+
+#### Read in + merge with the 99 from 1000G ####
+
+library(bigsnpr)
[…]

### 2022-04-13 code/prepare-corr-JPN.R

@@ -1,0 +1,152 @@
+library(dplyr)
+library(bigreadr)
+
+#### Get 504 "Japanese" individuals from the UKBB + 1000G ####
+
+df_UKBB <- runonce::save_run({
+
+  # Country of birth
+  code_country   <- filter(fread2("UKBB/coding89.tsv"), selectable == "Y")
+
+  eid_imputed <- fread2("UKBB/ukb58024_imp_chr1_v3_s487296.sample")$ID_1[-1]
+
+  fread2(
+    "UKBB/ukb41181.csv",
+    select = c("eid", "20115-0.0", paste0("22009-0.", 1:16)),
+    col.names = c("eid", "country", paste0("PC", 1:16))
+  ) %>%
+    filter(eid %in% eid_imputed) %>%
+    mutate(country = factor(country, levels = code_country$coding,
+                            labels = code_country$meaning))
+
+}, file = "tmp-data/info-UKBB.rds")
+
+center_JPN <- df_UKBB %>%
+  filter(country == "Japan") %>%
+  select(PC1:PC16) %>%
+  bigutilsr::geometric_median()
+
+ldist_to_JPN <- df_UKBB %>%
+  select(PC1:PC16) %>%
+  { sweep(., 2, center_JPN, '-')^2 } %>%
+  rowSums() %>%
+  sqrt() %>%
+  log()
+
+hist(ldist_to_JPN, "FD")
+
+print_table <- function(x, min_count = 5) {
+  tab <- sort(table(x, exclude = NULL), decreasing = TRUE)
+  tab <- tab[tab >= min_count]
+  cat(paste0(names(tab), ": ", tab), sep = " || ")
+}
+
+# Remove withdrawn and related individuals
+eid_withdrawn <- scan("UKBB/w58024_20220222.csv")
+rel <- fread2("UKBB/ukb58024_rel_s488264.dat")
+ldist_to_JPN[df_UKBB$eid %in% c(eid_withdrawn, filter(rel, Kinship > 2^-2.5)$ID2)] <- Inf
+
+ind <- head(order(ldist_to_JPN), 504 - 104)
+print_table(df_UKBB$country[ind])
+# Japan: 241 || China: 100 || South Korea: 26 || NA: 9 || North Korea: 6 || Hong Kong: 5 || Brazil: 5
+
+
+#### Read in + merge with the 104 from 1000G ####
+
+library(bigsnpr)
[…]

### 2022-04-13 code/prepare-corr-UKBB-EAS.R

@@ -1,0 +1,160 @@
+library(dplyr)
+library(bigreadr)
+
+#### Get East Asian individuals close to "Japanese" individuals from the UKBB ####
+
+df_UKBB <- runonce::save_run({
+
+  # Country of birth
+  code_country   <- filter(fread2("UKBB/coding89.tsv"), selectable == "Y")
+
+  eid_imputed <- fread2("UKBB/ukb58024_imp_chr1_v3_s487296.sample")$ID_1[-1]
+
+  fread2(
+    "UKBB/ukb41181.csv",
+    select = c("eid", "20115-0.0", paste0("22009-0.", 1:16)),
+    col.names = c("eid", "country", paste0("PC", 1:16))
+  ) %>%
+    filter(eid %in% eid_imputed) %>%
+    mutate(country = factor(country, levels = code_country$coding,
+                            labels = code_country$meaning))
+
+}, file = "tmp-data/info-UKBB.rds")
+
+center_JPN <- df_UKBB %>%
+  filter(country == "Japan") %>%
+  select(PC1:PC16) %>%
+  bigutilsr::geometric_median()
+
+ldist_to_JPN <- df_UKBB %>%
+  select(PC1:PC16) %>%
+  { sweep(., 2, center_JPN, '-')^2 } %>%
+  rowSums() %>%
+  sqrt() %>%
+  log()
+
+hist(ldist_to_JPN, "FD")
+
+print_table <- function(x, min_count = 5) {
+  tab <- sort(table(x, exclude = NULL), decreasing = TRUE)
+  tab <- tab[tab >= min_count]
+  cat(paste0(names(tab), ": ", tab), sep = " || ")
+}
+
+# Remove withdrawn and related individuals
+eid_withdrawn <- scan("UKBB/w58024_20220222.csv")
+rel <- fread2("UKBB/ukb58024_rel_s488264.dat")
+ldist_to_JPN[df_UKBB$eid %in% c(eid_withdrawn, filter(rel, Kinship > 2^-2.5)$ID2)] <- Inf
+
+hist(ldist_to_JPN)
+length(ind <- which(ldist_to_JPN < 4.5))  # 2041
+print_table(df_UKBB$country[ind])
+# Hong Kong: 450 || China: 373 || Malaysia: 284 || Japan: 241 || NA: 130 || Nepal: 116 ||
[…]

### 2022-04-13 code/prepare-genotypes-1000G.R

@@ -1,0 +1,70 @@
+# https://www.cog-genomics.org/plink/2.0/resources#1kg_phase3
+pgen_zst <- runonce::download_file(
+  "https://www.dropbox.com/s/afvvf1e15gqzsqo/all_phase3.pgen.zst?dl=1",
+  dir = "tmp-data")
+pvar_zst <- runonce::download_file(
+  "https://www.dropbox.com/s/op9osq6luy3pjg8/all_phase3.pvar.zst?dl=1",
+  dir = "tmp-data")
+psam <- runonce::download_file(
+  "https://www.dropbox.com/s/yozrzsdrwqej63q/phase3_corrected.psam?dl=1",
+  dir = "tmp-data", fname = "all_phase3.psam")
+
+library(bigsnpr)
+plink2 <- download_plink2("tmp-data")
+
+sysglue <- function(...) system(glue::glue(...))
+pgen <- sub('\\.zst', '', pgen_zst)
+if (!file.exists(pgen)) sysglue("{plink2} --zst-decompress {pgen_zst} > {pgen}")
+pvar <- sub('\\.zst', '', pvar_zst)
+if (!file.exists(pvar)) sysglue("{plink2} --zst-decompress {pvar_zst} > {pvar}")
+
+# https://dx.doi.org/10.1016%2Fj.ajhg.2017.03.004
+outliers <- c("NA20314", "HG00731", "HG00732", "HG01880", "HG01882",
+              "HG01944", "HG02497", "NA20320", "NA20321")
+bigsnpr:::write.table2(cbind.data.frame(0, outliers), (tmp <- tempfile()))
+
+sysglue("{plink2} --pfile tmp-data/all_phase3",
+        " --remove {tmp} --keep-founders",
+        " --max-alleles 2 --mac 10 --autosome",
+        " --make-bed --out tmp-data/all_phase3",
+        " --threads {nb_cores()[[1]]} --memory 50e3")
+
+obj.bed <- bed("tmp-data/all_phase3.bed")
+
+snp_info <- snp_attach("data/UKBB_HM3_val.rds")$map
+
+library(dplyr)
+matched <- snp_match(
+  transmute(obj.bed$map, chr = as.integer(chromosome), pos = physical.pos,
+            a0 = allele2, a1 = allele1, beta = 1),
+  transmute(snp_info, chr = chromosome, pos = physical.pos, a0 = allele2, a1 = allele1)
+) %>%
[…]

### 2022-04-13 code/prepare-genotypes-EAS.R

@@ -1,0 +1,78 @@
+library(dplyr)
+library(bigreadr)
+
+#### Get some East Asian individuals from the UKBB ####
+
+df_UKBB <- runonce::save_run({
+
+  # Country of birth
+  code_country   <- filter(fread2("UKBB/coding89.tsv"), selectable == "Y")
+
+  eid_imputed <- fread2("UKBB/ukb58024_imp_chr1_v3_s487296.sample")$ID_1[-1]
+
+  fread2(
+    "UKBB/ukb41181.csv",
+    select = c("eid", "20115-0.0", paste0("22009-0.", 1:16)),
+    col.names = c("eid", "country", paste0("PC", 1:16))
+  ) %>%
+    filter(eid %in% eid_imputed) %>%
+    mutate(country = factor(country, levels = code_country$coding,
+                            labels = code_country$meaning))
+
+}, file = "tmp-data/info-UKBB.rds")
+
+all_centers <- read.csv(
+  "https://raw.githubusercontent.com/privefl/UKBB-PGS/main/pop_centers.csv",
+  stringsAsFactors = FALSE)
+
+center_EAS <- all_centers %>%
+  filter(Ancestry == "China") %>%
+  select(-Ancestry) %>%
+  unlist()
+
+ldist <- df_UKBB %>%
+  select(PC1:PC16) %>%
+  { sweep(., 2, center_EAS, '-')^2 } %>%
+  rowSums() %>%
+  sqrt() %>%
+  log()
+
+hist(ldist[ldist < 5], "FD")
+
+print_table <- function(x, min_count = 5) {
+  tab <- sort(table(x, exclude = NULL), decreasing = TRUE)
+  tab <- tab[tab >= min_count]
+  cat(paste0(names(tab), ": ", tab), sep = " || ")
+}
+
+# Remove withdrawn and related individuals
+eid_withdrawn <- scan("UKBB/w58024_20220222.csv")
+rel <- fread2("UKBB/ukb58024_rel_s488264.dat")
+ldist[df_UKBB$eid %in% c(eid_withdrawn, filter(rel, Kinship > 2^-2.5)$ID2)] <- Inf
+
+ind <- which(ldist < 3.8)  # 1777 indivs
+print_table(df_UKBB$country[ind])
[…]

### 2022-04-13 code/prepare-genotypes.R

@@ -1,19 +1,24 @@
 # file.symlink("../../UKBB/", "UKBB")
 
-map_hapmap3 <- readRDS(url("https://ndownloader.figshare.com/files/25503788"))
+list_snp_id <- runonce::save_run({
 
-library(doParallel)
-cl <- makeCluster(22)
-parallel::clusterExport(cl, "map_hapmap3")
-list_snp_id <- parLapply(cl, 1:22, function(chr) {
-  mfi <- paste0("UKBB/mfi/ukb_mfi_chr", chr, "_v3.txt")
-  infos_chr <- bigreadr::fread2(mfi, showProgress = FALSE)
-  joined <- dplyr::inner_join(cbind(chr = chr, infos_chr), map_hapmap3[1:2],
-                              by = c("chr", "V3" = "pos"))
-  with(joined[!vctrs::vec_duplicate_detect(joined$V3), ],
-       paste(chr, V3, V4, V5, sep = "_"))
-})
-stopCluster(cl)
+  map_hapmap3 <- readRDS(url("https://ndownloader.figshare.com/files/25503788"))
+
+  library(doParallel)
+  cl <- makeCluster(22)
+  parallel::clusterExport(cl, "map_hapmap3")
+  list_snp_id <- parLapply(cl, 1:22, function(chr) {
+    mfi <- paste0("UKBB/mfi/ukb_mfi_chr", chr, "_v3.txt")
+    infos_chr <- bigreadr::fread2(mfi, showProgress = FALSE)
+    joined <- dplyr::inner_join(cbind(chr = chr, infos_chr), map_hapmap3[1:2],
+                                by = c("chr", "V3" = "pos"))
+    with(joined[!vctrs::vec_duplicate_detect(joined$V3), ],
+         paste(chr, V3, V4, V5, sep = "_"))
+  })
+  stopCluster(cl)
+
+  list_snp_id
+}, file = "data/list_snp_id_HM3.rds")
 
 sum(lengths(list_snp_id))  # 1,054,315
 

### 2022-04-13 code/prepare-phecodes.R

@@ -69,8 +69,6 @@ phecodes <- createPhenotypes(
   full.population.ids = seq_along(sex)
 )
 
-saveRDS(phecodes[colSums(phecodes, na.rm = TRUE) > 500],
-        "data/all_phecodes.rds")
 
 # some verifs
 table(pheno = phecodes$`185`, sex, exclude = NULL)
@@ -82,3 +80,39 @@ table(pheno = phecodes$`185`, sex, exclude = NULL)
 table(pheno = phecodes$`174.1`, sex, exclude = NULL)
 table(pheno = phecodes$`654.2`, sex, exclude = NULL)
 table(pheno = phecodes$`296.2`, sex, exclude = NULL)
+
+
+phecodes_sub <- phecodes[colSums(phecodes, na.rm = TRUE) > 500]
+
+
+#### Add a few continuous outcomes ####
+
+# Vitamin D (30890)
+y <- fread2(csv, select = "30890-0.0")[[1]]
+hist(y)
+y[y > 150] <- NA
+phecodes_sub$vitaminD <- y
+
+# Standing height (50)
+y <- fread2(csv, select = "50-0.0")[[1]]
+hist(y)
+y[y < 130] <- NA
+phecodes_sub$height <- y
+
+# Body mass index (BMI) (21001)
+y <- fread2(csv, select = "21001-0.0")[[1]]
+hist(y)
+phecodes_sub$bmi <- y
+
+# Systolic blood pressure, automated reading (4080)
+y <- fread2(csv, select = "4080-0.0")[[1]]
+hist(y)
+phecodes_sub$systolic_bp <- y
+
+# HDL cholesterol (30760)
+y <- fread2(csv, select = "30760-0.0")[[1]]
+hist(y)
+phecodes_sub$hdl_cholesterol <- y
+
+
+saveRDS(phecodes_sub, "data/all_phecodes.rds")

### 2022-04-13 code/prepare-sumstats-bbj/BMI.R

@@ -1,0 +1,73 @@
+library(bigsnpr)
+library(bigreadr)
+library(dplyr)
+library(ggplot2)
+
+map <- transmute(snp_attach("data/UKBB_HM3_val.rds")$map,
+                 chr = as.integer(chromosome), pos = physical.pos, rsid,
+                 a0 = allele1, a1 = allele2)  # reversed somehow..
+
+zip <- runonce::download_file(
+  "https://pheweb.jp/download/BMI",
+  dir = "tmp-data", fname = "sumstats_bmi_BBJ.zip")
+unzip(zip, exdir = "tmp-data", overwrite = FALSE)
+
+sumstats <- bigreadr::fread2(
+  "tmp-data/hum0197.v3.BBJ.BMI.v1/GWASsummary_BMI_Japanese_SakaueKanai2020.auto.txt.gz",
+  select = c("CHR", "BP", "SNP", "ALLELE0", "ALLELE1", "A1FREQ", "INFO",
+             "BETA", "SE", "CHISQ_LINREG", "CHISQ_BOLT_LMM_INF"),
+  col.names = c("chr", "pos", "rsid", "a0", "a1", "freq", "info",
+                "beta", "beta_se", "CHISQ_LINREG", "CHISQ_BOLT_LMM_INF")
+) %>%
+  filter(chr %in% 1:22, pmin(freq, 1 - freq) > 0.01) %>%
+  mutate(chr = as.integer(chr)) %>%
+  print()
+
+# Estimate power gain using the random effects
+qplot(CHISQ_BOLT_LMM_INF / CHISQ_LINREG, color = as.factor(chr), geom = "density",
+      data = filter(sumstats, CHISQ_LINREG > 30))
+
+power <-
+  with(filter(sumstats, CHISQ_LINREG > 30),
+       matrixStats::weightedMedian(CHISQ_BOLT_LMM_INF / CHISQ_LINREG,
+                                   w = CHISQ_BOLT_LMM_INF))
+# 1.025
+
+info_snp <- as_tibble(bigsnpr::snp_match(sumstats, map, join_by_pos = FALSE))
+# 7,444,325 variants to be matched.
+# 0 ambiguous SNPs have been removed.
+# 870,005 variants have been matched; 0 were flipped and 870,005 were reversed.
+
+hist(info_snp$info)
+mean(info_snp$info)  # 98.6%
+
+info_snp2 <- info_snp %>%
+  mutate(n_eff = 163835 * power,
[…]

### 2022-04-13 code/prepare-sumstats-bbj/HDL_cholesterol.R

@@ -1,0 +1,73 @@
+library(bigsnpr)
+library(bigreadr)
+library(dplyr)
+library(ggplot2)
+
+map <- transmute(snp_attach("data/UKBB_HM3_val.rds")$map,
+                 chr = as.integer(chromosome), pos = physical.pos, rsid,
+                 a0 = allele1, a1 = allele2)  # reversed somehow..
+
+zip <- runonce::download_file(
+  "https://pheweb.jp/download/HDLC",
+  dir = "tmp-data", fname = "sumstats_hdlc_BBJ.zip")
+unzip(zip, exdir = "tmp-data", overwrite = FALSE)
+
+sumstats <- bigreadr::fread2(
+  "tmp-data/hum0197.v3.BBJ.HDLC.v1/GWASsummary_HDLC_Japanese_SakaueKanai2020.auto.txt.gz",
+  select = c("CHR", "BP", "SNP", "ALLELE0", "ALLELE1", "A1FREQ", "INFO",
+             "BETA", "SE", "CHISQ_LINREG", "CHISQ_BOLT_LMM_INF"),
+  col.names = c("chr", "pos", "rsid", "a0", "a1", "freq", "info",
+                "beta", "beta_se", "CHISQ_LINREG", "CHISQ_BOLT_LMM_INF")
+) %>%
+  filter(chr %in% 1:22, pmin(freq, 1 - freq) > 0.01) %>%
+  mutate(chr = as.integer(chr)) %>%
+  print()
+
+# Estimate power gain using the random effects
+qplot(CHISQ_BOLT_LMM_INF / CHISQ_LINREG, color = as.factor(chr), geom = "density",
+      data = filter(sumstats, CHISQ_LINREG > 30))
+
+power <-
+  with(filter(sumstats, CHISQ_LINREG > 30),
+       matrixStats::weightedMedian(CHISQ_BOLT_LMM_INF / CHISQ_LINREG,
+                                   w = CHISQ_BOLT_LMM_INF))
+# 1.039
+
+info_snp <- as_tibble(bigsnpr::snp_match(sumstats, map, join_by_pos = FALSE))
+# 7,443,885 variants to be matched.
+# 0 ambiguous SNPs have been removed.
+# 870,022 variants have been matched; 0 were flipped and 870,022 were reversed.
+
+hist(info_snp$info)
+mean(info_snp$info)  # 98.6%
+
+info_snp2 <- info_snp %>%
+  mutate(n_eff = 74970 * power,
[…]

### 2022-04-13 code/prepare-sumstats-bbj/height.R

@@ -1,0 +1,73 @@
+library(bigsnpr)
+library(bigreadr)
+library(dplyr)
+library(ggplot2)
+
+map <- transmute(snp_attach("data/UKBB_HM3_val.rds")$map,
+                 chr = as.integer(chromosome), pos = physical.pos, rsid,
+                 a0 = allele1, a1 = allele2)  # reversed somehow..
+
+zip <- runonce::download_file(
+  "https://pheweb.jp/download/Height",
+  dir = "tmp-data", fname = "sumstats_height_BBJ.zip")
+unzip(zip, exdir = "tmp-data", overwrite = FALSE)
+
+sumstats <- bigreadr::fread2(
+  "tmp-data/hum0197.v3.BBJ.Hei.v1/GWASsummary_Height_Japanese_SakaueKanai2020.auto.txt.gz",
+  select = c("CHR", "BP", "SNP", "ALLELE0", "ALLELE1", "A1FREQ", "INFO",
+             "BETA", "SE", "CHISQ_LINREG", "CHISQ_BOLT_LMM_INF"),
+  col.names = c("chr", "pos", "rsid", "a0", "a1", "freq", "info",
+                "beta", "beta_se", "CHISQ_LINREG", "CHISQ_BOLT_LMM_INF")
+) %>%
+  filter(chr %in% 1:22, pmin(freq, 1 - freq) > 0.01) %>%
+  mutate(chr = as.integer(chr)) %>%
+  print()
+
+# Estimate power gain using the random effects
+qplot(CHISQ_BOLT_LMM_INF / CHISQ_LINREG, color = as.factor(chr), geom = "density",
+      data = filter(sumstats, CHISQ_LINREG > 30))
+
+power <-
+  with(filter(sumstats, CHISQ_LINREG > 30),
+       matrixStats::weightedMedian(CHISQ_BOLT_LMM_INF / CHISQ_LINREG,
+                                   w = CHISQ_BOLT_LMM_INF))
+# 1.143
+
+info_snp <- as_tibble(bigsnpr::snp_match(sumstats, map, join_by_pos = FALSE))
+# 7,444,297 variants to be matched.
+# 0 ambiguous SNPs have been removed.
+# 870,000 variants have been matched; 0 were flipped and 870,000 were reversed.
+
+hist(info_snp$info)
+mean(info_snp$info)  # 98.6%
+
+info_snp2 <- info_snp %>%
+  mutate(n_eff = 165056 * power,
+         sd_af = sqrt(2 * freq * (1 - freq)),
[…]

### 2022-04-13 code/prepare-sumstats-bbj/systolic_BP.R

@@ -1,0 +1,73 @@
+library(bigsnpr)
+library(bigreadr)
+library(dplyr)
+library(ggplot2)
+
+map <- transmute(snp_attach("data/UKBB_HM3_val.rds")$map,
+                 chr = as.integer(chromosome), pos = physical.pos, rsid,
+                 a0 = allele1, a1 = allele2)  # reversed somehow..
+
+zip <- runonce::download_file(
+  "https://pheweb.jp/download/SBP",
+  dir = "tmp-data", fname = "sumstats_sbp_BBJ.zip")
+unzip(zip, exdir = "tmp-data", overwrite = FALSE)
+
+sumstats <- bigreadr::fread2(
+  "tmp-data/hum0197.v3.BBJ.SBP.v1/GWASsummary_SBP_Japanese_SakaueKanai2020.auto.txt.gz",
+  select = c("CHR", "BP", "SNP", "ALLELE0", "ALLELE1", "A1FREQ", "INFO",
+             "BETA", "SE", "CHISQ_LINREG", "CHISQ_BOLT_LMM_INF"),
+  col.names = c("chr", "pos", "rsid", "a0", "a1", "freq", "info",
+                "beta", "beta_se", "CHISQ_LINREG", "CHISQ_BOLT_LMM_INF")
+) %>%
+  filter(chr %in% 1:22, pmin(freq, 1 - freq) > 0.01) %>%
+  mutate(chr = as.integer(chr)) %>%
+  print()
+
+# Estimate power gain using the random effects
+qplot(CHISQ_BOLT_LMM_INF / CHISQ_LINREG, color = as.factor(chr), geom = "density",
+      data = filter(sumstats, CHISQ_LINREG > 30))
+
+power <-
+  with(filter(sumstats, CHISQ_LINREG > 30),
+       matrixStats::weightedMedian(CHISQ_BOLT_LMM_INF / CHISQ_LINREG,
+                                   w = CHISQ_BOLT_LMM_INF))
+# 1.013
+
+info_snp <- as_tibble(bigsnpr::snp_match(sumstats, map, join_by_pos = FALSE))
+# 7,444,410 variants to be matched.
+# 0 ambiguous SNPs have been removed.
+# 869,991 variants have been matched; 0 were flipped and 869,991 were reversed.
+
+hist(info_snp$info)
+mean(info_snp$info)  # 98.6%
+
+info_snp2 <- info_snp %>%
+  mutate(n_eff = 145505 * power,
[…]

### 2022-04-13 code/prepare-sumstats-finngen/BrCa.R

@@ -1,0 +1,73 @@
+library(bigsnpr)
+library(bigreadr)
+library(dplyr)
+library(ggplot2)
+
+map <- transmute(snp_attach("data/UKBB_HM3_val.rds")$map,
+                 chr = as.integer(chromosome), pos = physical.pos, rsid,
+                 a0 = allele1, a1 = allele2)  # reversed somehow..
+
+# https://www.finngen.fi/en/access_results
+file <- R.utils::gunzip("tmp-data/summary_stats_finngen_R6_C3_BREAST.gz",
+                        skip = TRUE, remove = FALSE)
+writeLines(readLines(file, n = 2))
+
+sumstats <- bigreadr::fread2(
+  file,
+  select = c("#chrom", "pos", "rsids", "ref", "alt", "af_alt", "beta", "sebeta",
+             "n_hom_cases", "n_hom_ref_cases", "n_het_cases",
+             "n_hom_controls", "n_hom_ref_controls", "n_het_controls"),
+  col.names = c("chr", "pos", "rsid", "a0", "a1", "freq", "beta", "beta_se",
+                "n_hom_cases", "n_hom_ref_cases", "n_het_cases",
+                "n_hom_controls", "n_hom_ref_controls", "n_het_controls"))
+
+info_snp <- as_tibble(bigsnpr::snp_match(sumstats, map, join_by_pos = FALSE))
+# 16,354,723 variants to be matched.
+# 246 ambiguous SNPs have been removed.
+# 1,042,134 variants have been matched; 690 were flipped and 2,804 were reversed.
+
+info_snp2 <- info_snp %>%
+  filter(chr %in% 1:22, pmin(freq, 1 - freq) > 0.01) %>%
+  mutate(n_case = round(n_hom_cases + n_hom_ref_cases + n_het_cases),
+         n_control = round(n_hom_controls + n_hom_ref_controls + n_het_controls))
+
+table(info_snp2$n_case)    # all 11,573
+table(info_snp2$n_control) # all 135,488
+
+info_snp3 <- info_snp2 %>%
+  select(!starts_with("n_")) %>%
+  mutate(n_eff = 4 / (1 / 11573 + 1 / 135488),
+         sd_af = sqrt(2 * freq * (1 - freq)),
+         sd_ss = 2 / sqrt(n_eff * beta_se^2 + beta^2))
+
+qplot(sd_af, sd_ss, alpha = I(0.2),
[…]

### 2022-04-13 code/prepare-sumstats-finngen/CAD.R

@@ -1,0 +1,75 @@
+library(bigsnpr)
+library(bigreadr)
+library(dplyr)
+library(ggplot2)
+
+map <- transmute(snp_attach("data/UKBB_HM3_val.rds")$map,
+                 chr = as.integer(chromosome), pos = physical.pos, rsid,
+                 a0 = allele1, a1 = allele2)  # reversed somehow..
+
+# https://www.finngen.fi/en/access_results
+file <- R.utils::gunzip("tmp-data/summary_stats_finngen_R6_I9_CORATHER.gz",
+                        skip = TRUE, remove = FALSE)
+
+
+writeLines(readLines(file, n = 2))
+
+sumstats <- bigreadr::fread2(
+  file,
+  select = c("#chrom", "pos", "rsids", "ref", "alt", "af_alt", "beta", "sebeta",
+             "n_hom_cases", "n_hom_ref_cases", "n_het_cases",
+             "n_hom_controls", "n_hom_ref_controls", "n_het_controls"),
+  col.names = c("chr", "pos", "rsid", "a0", "a1", "freq", "beta", "beta_se",
+                "n_hom_cases", "n_hom_ref_cases", "n_het_cases",
+                "n_hom_controls", "n_hom_ref_controls", "n_het_controls"))
+
+info_snp <- as_tibble(bigsnpr::snp_match(sumstats, map, join_by_pos = FALSE))
+# 16,355,268 variants to be matched.
+# 246 ambiguous SNPs have been removed.
+# 1,042,134 variants have been matched; 690 were flipped and 2,804 were reversed.
+
+info_snp2 <- info_snp %>%
+  filter(chr %in% 1:22, pmin(freq, 1 - freq) > 0.01) %>%
+  mutate(n_case = round(n_hom_cases + n_hom_ref_cases + n_het_cases),
+         n_control = round(n_hom_controls + n_hom_ref_controls + n_het_controls))
+
+table(info_snp2$n_case)    # all 28,598
+table(info_snp2$n_control) # all 222,551
+
+info_snp3 <- info_snp2 %>%
+  select(!starts_with("n_")) %>%
+  mutate(n_eff = 4 / (1 / 28598 + 1 / 222551),
+         sd_af = sqrt(2 * freq * (1 - freq)),
+         sd_ss = 2 / sqrt(n_eff * beta_se^2 + beta^2))
+
+qplot(sd_af, sd_ss, alpha = I(0.2),
[…]

### 2022-04-13 code/prepare-sumstats-finngen/PrCa.R

@@ -1,0 +1,73 @@
+library(bigsnpr)
+library(bigreadr)
+library(dplyr)
+library(ggplot2)
+
+map <- transmute(snp_attach("data/UKBB_HM3_val.rds")$map,
+                 chr = as.integer(chromosome), pos = physical.pos, rsid,
+                 a0 = allele1, a1 = allele2)  # reversed somehow..
+
+# https://www.finngen.fi/en/access_results
+file <- R.utils::gunzip("tmp-data/summary_stats_finngen_R6_C3_PROSTATE.gz",
+                        skip = TRUE, remove = FALSE)
+writeLines(readLines(file, n = 2))
+
+sumstats <- bigreadr::fread2(
+  file,
+  select = c("#chrom", "pos", "rsids", "ref", "alt", "af_alt", "beta", "sebeta",
+             "n_hom_cases", "n_hom_ref_cases", "n_het_cases",
+             "n_hom_controls", "n_hom_ref_controls", "n_het_controls"),
+  col.names = c("chr", "pos", "rsid", "a0", "a1", "freq", "beta", "beta_se",
+                "n_hom_cases", "n_hom_ref_cases", "n_het_cases",
+                "n_hom_controls", "n_hom_ref_controls", "n_het_controls"))
+
+info_snp <- as_tibble(bigsnpr::snp_match(sumstats, map, join_by_pos = FALSE))
+# 16,354,079 variants to be matched.
+# 246 ambiguous SNPs have been removed.
+# 1,042,134 variants have been matched; 690 were flipped and 2,804 were reversed.
+
+info_snp2 <- info_snp %>%
+  filter(chr %in% 1:22, pmin(freq, 1 - freq) > 0.01) %>%
+  mutate(n_case = round(n_hom_cases + n_hom_ref_cases + n_het_cases),
+         n_control = round(n_hom_controls + n_hom_ref_controls + n_het_controls))
+
+table(info_snp2$n_case)    # all 8709
+table(info_snp2$n_control) # all 104,635
+
+info_snp3 <- info_snp2 %>%
+  select(!starts_with("n_")) %>%
+  mutate(n_eff = 4 / (1 / 8709 + 1 / 104635),
+         sd_af = sqrt(2 * freq * (1 - freq)),
+         sd_ss = 2 / sqrt(n_eff * beta_se^2 + beta^2))
+
+qplot(sd_af, sd_ss, alpha = I(0.2),
[…]

### 2022-04-13 code/prepare-sumstats-finngen/T1D.R

@@ -1,0 +1,74 @@
+library(bigsnpr)
+library(bigreadr)
+library(dplyr)
+library(ggplot2)
+
+map <- transmute(snp_attach("data/UKBB_HM3_val.rds")$map,
+                 chr = as.integer(chromosome), pos = physical.pos, rsid,
+                 a0 = allele1, a1 = allele2)  # reversed somehow..
+
+# https://www.finngen.fi/en/access_results
+file <- R.utils::gunzip("tmp-data/summary_stats_finngen_R6_E4_DM1.gz",
+                        skip = TRUE, remove = FALSE)
+
+writeLines(readLines(file, n = 2))
+
+sumstats <- bigreadr::fread2(
+  file,
+  select = c("#chrom", "pos", "rsids", "ref", "alt", "af_alt", "beta", "sebeta",
+             "n_hom_cases", "n_hom_ref_cases", "n_het_cases",
+             "n_hom_controls", "n_hom_ref_controls", "n_het_controls"),
+  col.names = c("chr", "pos", "rsid", "a0", "a1", "freq", "beta", "beta_se",
+                "n_hom_cases", "n_hom_ref_cases", "n_het_cases",
+                "n_hom_controls", "n_hom_ref_controls", "n_het_controls"))
+
+info_snp <- as_tibble(bigsnpr::snp_match(sumstats, map, join_by_pos = FALSE))
+# 16,355,179 variants to be matched.
+# 246 ambiguous SNPs have been removed.
+# 1,042,134 variants have been matched; 690 were flipped and 2,804 were reversed.
+
+info_snp2 <- info_snp %>%
+  filter(chr %in% 1:22, pmin(freq, 1 - freq) > 0.01) %>%
+  mutate(n_case = round(n_hom_cases + n_hom_ref_cases + n_het_cases),
+         n_control = round(n_hom_controls + n_hom_ref_controls + n_het_controls))
+
+table(info_snp2$n_case)    # all 7609
+table(info_snp2$n_control) # all 215,160
+
+info_snp3 <- info_snp2 %>%
+  select(!starts_with("n_")) %>%
+  mutate(n_eff = 4 / (1 / 7609 + 1 / 215160),
+         sd_af = sqrt(2 * freq * (1 - freq)),
+         sd_ss = 2 / sqrt(n_eff * beta_se^2 + beta^2))
+
+qplot(sd_af, sd_ss, alpha = I(0.2),
[…]

### 2022-04-13 code/prepare-sumstats-finngen/T2D.R

@@ -1,0 +1,74 @@
+library(bigsnpr)
+library(bigreadr)
+library(dplyr)
+library(ggplot2)
+
+map <- transmute(snp_attach("data/UKBB_HM3_val.rds")$map,
+                 chr = as.integer(chromosome), pos = physical.pos, rsid,
+                 a0 = allele1, a1 = allele2)  # reversed somehow..
+
+# https://www.finngen.fi/en/access_results
+file <- R.utils::gunzip("tmp-data/summary_stats_finngen_R6_E4_DM2.gz",
+                        skip = TRUE, remove = FALSE)
+
+writeLines(readLines(file, n = 2))
+
+sumstats <- bigreadr::fread2(
+  file,
+  select = c("#chrom", "pos", "rsids", "ref", "alt", "af_alt", "beta", "sebeta",
+             "n_hom_cases", "n_hom_ref_cases", "n_het_cases",
+             "n_hom_controls", "n_hom_ref_controls", "n_het_controls"),
+  col.names = c("chr", "pos", "rsid", "a0", "a1", "freq", "beta", "beta_se",
+                "n_hom_cases", "n_hom_ref_cases", "n_het_cases",
+                "n_hom_controls", "n_hom_ref_controls", "n_het_controls"))
+
+info_snp <- as_tibble(bigsnpr::snp_match(sumstats, map, join_by_pos = FALSE))
+# 16,355,271 variants to be matched.
+# 246 ambiguous SNPs have been removed.
+# 1,042,134 variants have been matched; 690 were flipped and 2,804 were reversed.
+
+info_snp2 <- info_snp %>%
+  filter(chr %in% 1:22, pmin(freq, 1 - freq) > 0.01) %>%
+  mutate(n_case = round(n_hom_cases + n_hom_ref_cases + n_het_cases),
+         n_control = round(n_hom_controls + n_hom_ref_controls + n_het_controls))
+
+table(info_snp2$n_case)    # all 41,245
+table(info_snp2$n_control) # all 215,160
+
+info_snp3 <- info_snp2 %>%
+  select(!starts_with("n_")) %>%
+  mutate(n_eff = 4 / (1 / 41245 + 1 / 215160),
+         sd_af = sqrt(2 * freq * (1 - freq)),
+         sd_ss = 2 / sqrt(n_eff * beta_se^2 + beta^2))
+
+qplot(sd_af, sd_ss, alpha = I(0.2),
[…]

### 2022-04-13 code/prepare-sumstats/BrCa-icogs.R

@@ -115,5 +115,6 @@ sum(info_snp2$is_bad3 <- with(info_snp2, abs(af_val - freq2) > 0.1))  # 459
 with(info_snp2, table(is_bad2, is_bad3))
 with(info_snp2, table(is_bad, is_bad2 | is_bad3))
 
+# saveRDS(info_snp2, "data/sumstats/brca_icogs_noqc.rds")
 # saveRDS(filter(info_snp2, !is_bad),            "data/sumstats/brca_icogs_qc1.rds")
 # saveRDS(filter(info_snp2, !is_bad2, !is_bad3), "data/sumstats/brca_icogs_qc2.rds")

### 2022-04-13 code/prepare-sumstats/BrCa-onco.R

@@ -146,5 +146,6 @@ sum(info_snp2$is_bad3 <- with(info_snp2, abs(af_val - freq2) > 0.1))  # 21
 with(info_snp2, table(is_bad2, is_bad3))  # none in common
 with(info_snp2, table(is_bad, is_bad2 | is_bad3))
 
+# saveRDS(info_snp2, "data/sumstats/brca_onco_noqc.rds")
 # saveRDS(filter(info_snp2, !is_bad),            "data/sumstats/brca_onco_qc1.rds")
 # saveRDS(filter(info_snp2, !is_bad2, !is_bad3), "data/sumstats/brca_onco_qc2.rds")

### 2022-04-13 code/prepare-sumstats/CAD.R

@@ -145,5 +145,6 @@ with(info_snp2, table(is_bad2, is_bad3))            # almost none in common
 
 info_snp2$n_eff <- with(info_snp2, Neff * n_studies / max(n_studies))
 
+# saveRDS(info_snp2, "data/sumstats/cad_noqc.rds")
 # saveRDS(filter(info_snp2, !is_bad),            "data/sumstats/cad_qc1.rds")
 # saveRDS(filter(info_snp2, !is_bad2, !is_bad3), "data/sumstats/cad_qc2.rds")

### 2022-04-13 code/prepare-sumstats/MDD.R

@@ -126,5 +126,6 @@ info_snp2$is_bad3 <- with(info_snp2, abs(af_val - freq2) > 0.1)
 with(info_snp2, table(is_bad, is_bad2 | is_bad3))
 with(info_snp2, table(is_bad2, is_bad3))
 
+# saveRDS(info_snp2, "data/sumstats/mdd_noqc.rds")
 # saveRDS(filter(info_snp2, !is_bad),            "data/sumstats/mdd_qc1.rds")
 # saveRDS(filter(info_snp2, !is_bad2, !is_bad3), "data/sumstats/mdd_qc2.rds")

### 2022-04-13 code/prepare-sumstats/PrCa.R

@@ -117,5 +117,6 @@ info_snp2$is_bad3 <- with(info_snp2, abs(af_val - freq2) > 0.1)
 with(info_snp2, table(is_bad, is_bad2 | is_bad3))
 with(info_snp2, table(is_bad2, is_bad3))
 
+# saveRDS(info_snp2, "data/sumstats/prca_noqc.rds")
 # saveRDS(filter(info_snp2, !is_bad),            "data/sumstats/prca_qc1.rds")
 # saveRDS(filter(info_snp2, !is_bad2, !is_bad3), "data/sumstats/prca_qc2.rds")

### 2022-04-13 code/prepare-sumstats/T1D-affy.R

@@ -121,5 +121,6 @@ qplot(sd_val, sd_ss2, color = is_bad3, alpha = I(0.5),
 with(info_snp2, table(is_bad2, is_bad3))
 with(info_snp2, table(is_bad, is_bad2 | is_bad3))
 
+# saveRDS(info_snp2, "data/sumstats/t1d_affy_noqc.rds")
 # saveRDS(filter(info_snp2, !is_bad),            "data/sumstats/t1d_affy_qc1.rds")
 # saveRDS(filter(info_snp2, !is_bad2, !is_bad3), "data/sumstats/t1d_affy_qc2.rds")

### 2022-04-13 code/prepare-sumstats/T1D-illu.R

@@ -122,5 +122,6 @@ qplot(sd_val, sd_ss2, color = is_bad3, alpha = I(0.5),
 with(info_snp2, table(is_bad2, is_bad3))
 with(info_snp2, table(is_bad, is_bad2 | is_bad3))
 
+# saveRDS(info_snp2, "data/sumstats/t1d_illu_noqc.rds")
 # saveRDS(filter(info_snp2, !is_bad),            "data/sumstats/t1d_illu_qc1.rds")
 # saveRDS(filter(info_snp2, !is_bad2, !is_bad3), "data/sumstats/t1d_illu_qc2.rds")

### 2022-04-13 code/prepare-sumstats/vitaminD.R

@@ -1,0 +1,62 @@
+library(bigsnpr)
+library(bigreadr)
+library(dplyr)
+library(ggplot2)
+
+
+map <- transmute(snp_attach("data/UKBB_HM3_val.rds")$map,
+                 chr = as.integer(chromosome), pos = physical.pos, rsid, af_val = freq,
+                 a0 = allele1, a1 = allele2)  # reversed somehow..
+
+
+# See google drive link in https://doi.org/10.1038/s41467-017-02662-2
+# Should have used Allele1 -> a1 and Allele2 -> a0
+sumstats <- fread2("tmp-data/25HydroxyVitaminD_QC.METAANALYSIS1.txt",
+                   select = c("Chr", "Pos", "MarkerName", "Allele1", "Allele2",
+                              "Effect", "StdErr", "SAMPLESIZE", "HetPVal"),
+                   col.names = c("chr", "pos", "rsid", "a0", "a1", "beta", "beta_se", "N", "het_pval")) %>%
+  mutate_at(c("a0", "a1"), toupper)
+head(sumstats$pos)
+# 127246000  33550000  24178800  96896300 160770000  59208100  -> useless for matching..
+
+
+info_snp <- as_tibble(snp_match(sumstats, map, join_by_pos = FALSE))
+# 2,579,296 variants to be matched.
+# 2 ambiguous SNPs have been removed.
+# 1,016,935 variants have been matched; 6,394 were flipped and 522,741 were reversed.
+
+hist(-log10(info_snp$het_pval), "FD")
+
+info_snp2 <- info_snp %>%
+  filter(het_pval > 1e-5) %>%
+  mutate(n_eff = N,
+         sd_af = sqrt(2 * af_val * (1 - af_val)),
+         sd_ss = 1 / sqrt(n_eff * beta_se^2 + beta^2),
+         sd_ss = sd_ss / quantile(sd_ss, 0.99) * sqrt(0.5))
+
+
+info_snp2$is_bad <- with(info_snp2,
+                         sd_ss < (0.7 * sd_af) | sd_ss > (sd_af + 0.1) |
+                           sd_ss < 0.1 | sd_af < 0.05)
+
+qplot(sd_af, sd_ss, color = ifelse(is_bad, "Yes", "No"), alpha = I(0.5),
+      data = slice_sample(info_snp2, n = 100e3)) +
+  theme_bigstatsr(0.8) +
+  coord_equal() +
[…]

### 2022-04-13 code/process-PRS-BBJ.R

@@ -1,0 +1,163 @@
+library(bigsnpr)
+ukb <- snp_attach("data/EAS_HM3.rds")
+G <- ukb$genotypes
+NCORES <- nb_cores()
+
+library(dplyr)
+df0 <- bigreadr::fread2(
+  "UKBB/ukb41181.csv",
+  select = c("eid", "34-0.0", "52-0.0", "22001-0.0", "21022-0.0", "189-0.0",
+             paste0("22009-0.", 1:16)),
+  col.names = c("eid", "year", "month", "sex", "age", "deprivation_index",
+                paste0("PC", 1:16))
+) %>%
+  mutate(date = (year - 1900) + (month - 0.5) / 12, year = NULL, month = NULL)
+
+id_csv <- ukb$fam$id_csv
+ind.val <- which(ukb$fam$set == "val")
+covar.val <- as.matrix(df0[id_csv[ind.val], -1])
+
+ALL_METHODS <- c("lassosum2", "LDpred2", "LDpred2-low-h2",
+                 "LDpred2-auto", "LDpred2-auto-rob")
+
+ALL_CORR <- c(
+  corr_JPN               = "data/corr_JPN/chr",
+  corr_JPN_with_blocks   = "data/corr_JPN/adj_with_blocks_chr",
+  corr_1000G             = "data/corr_1000G_EAS/chr",
+  corr_1000G_with_blocks = "data/corr_1000G_EAS/adj_with_blocks_chr",
+  corr_UKBB              = "data/corr_UKBB_EAS/chr",
+  corr_UKBB_with_blocks  = "data/corr_UKBB_EAS/adj_with_blocks_chr"
+)
+
+bigassertr::assert_dir("results-final_BBJ")
+
+library(dplyr)
+grid <- tibble(pheno = c("height", "systolic_bp", "hdl_cholesterol", "bmi")) %>%
+  tidyr::expand_grid(name_corr = names(ALL_CORR), method = ALL_METHODS) %>%
+  print()
+
+
+library(future.apply)
+plan("multisession", workers = NCORES)
+
+grid$effects <- furrr::future_pmap(grid[1:3], function(pheno, name_corr, method) {
+
+  # pheno <- "height"
+  # name_corr <- "corr_1000G"
+  # method <- "lassosum2"
+
+  res_file <- paste0("results_BBJ/", pheno, "_", name_corr, "_", method, ".rds")
+  if (!file.exists(res_file)) {
[…]

### 2022-04-13 code/process-PRS-FIN.R

@@ -1,0 +1,175 @@
+library(bigsnpr)
+ukb <- snp_attach("data/UKBB_HM3_val.rds")
+G <- ukb$genotypes
+NCORES <- nb_cores()
+
+library(dplyr)
+df0 <- bigreadr::fread2(
+  "UKBB/ukb41181.csv",
+  select = c("eid", "34-0.0", "52-0.0", "22001-0.0", "21022-0.0", "189-0.0",
+             paste0("22009-0.", 1:16)),
+  col.names = c("eid", "year", "month", "sex", "age", "deprivation_index",
+                paste0("PC", 1:16))
+) %>%
+  mutate(date = (year - 1900) + (month - 0.5) / 12, year = NULL, month = NULL)
+
+ind.val <- ukb$fam$id_csv
+covar <- as.matrix(df0[ind.val, -1])
+
+ALL_METHODS <- c("lassosum2", "LDpred2", "LDpred2-low-h2",
+                 "LDpred2-auto", "LDpred2-auto-rob")
+ALL_CORR <- c(
+  corr_UK             = "data/corr/chr",
+  corr_UK_with_blocks = "data/corr/adj_with_blocks_chr",
+  corr_FIN             = "data/corr_FIN/chr",
+  corr_FIN_with_blocks = "data/corr_FIN/adj_with_blocks_chr",
+  corr_1000G             = "data/corr_1000G_EUR/chr",
+  corr_1000G_with_blocks = "data/corr_1000G_EUR/adj_with_blocks_chr"
+)
+
+bigassertr::assert_dir("results-final_FIN")
+
+library(dplyr)
+grid <- tibble::tribble(
+  ~pheno,   ~phecode,
+   "t1d",    "250.1",
+   "t2d",    "250.2",
+  "brca",    "174.1",
+   "cad",    "411.4",
+  "prca",      "185",
+) %>%
+  tidyr::expand_grid(name_corr = names(ALL_CORR), method = ALL_METHODS) %>%
+  print()
+
+
+library(future.apply)
+plan("multisession", workers = NCORES)
+
+grid$effects <- furrr::future_pmap(grid[1:4], function(pheno, phecode, name_corr, method) {
+
+  # pheno <- "t2d"
+  # phecode <- "250.2"
+  # name_corr <- "corr_1000G"
+  # method <- "lassosum2"
+
+  res_file <- paste0("results_FIN/", pheno, "_", name_corr, "_", method, ".rds")
[…]

### 2022-04-13 code/process-PRS.R

@@ -30,20 +30,23 @@ grid <- tibble::tribble(
   "prca",                               "185",
   "mdd",                                "296.22",
 ) %>%
+  tidyr::unnest("pheno") %>%
+  mutate(whichN = NA) %>%
+  tidyr::expand_grid(use_info = c(FALSE, TRUE)) %>%
+  add_row(pheno = "vitaminD", phecode = "vitaminD", whichN = c("trueN", "maxN"), use_info = FALSE) %>%
   tidyr::expand_grid(
-    qc = c("qc1", "qc2"),
-    use_info = c(FALSE, TRUE),
+    qc = c("noqc", "qc1", "qc2"),
     name_corr = c("data/corr/chr", "data/corr/adj_with_blocks_chr"),
     method = ALL_METHODS) %>%
-  tidyr::unnest("pheno") %>%
   print()
+# 480 different PGS models
 
 
 library(future.apply)
 plan("multisession", workers = NCORES)
 
-grid$effects <- furrr::future_pmap(grid[1:6], function(
-  pheno, phecode, qc, use_info, name_corr, method) {
+grid$effects <- furrr::future_pmap(grid[1:7], function(
+  pheno, phecode, qc, use_info, whichN, name_corr, method) {
 
   # pheno <- "t1d_illu"
   # phecode <- "250.1"
@@ -52,11 +55,14 @@ grid$effects <- furrr::future_pmap(grid[1:6], function(
   # name_corr <- "data/corr/adj_with_blocks_chr"
   # method <- "lassosum2"
 
+
   basename <- paste0(pheno, "_", qc)
   sumstats_file <- paste0("data/sumstats/", basename, ".rds")
+
+  suffix_N <- `if`(is.na(whichN), "", paste0("_", whichN))
+  basename <- paste0("results/", basename, suffix_N)
   if (grepl("blocks", name_corr, fixed = TRUE))
     basename <- paste0(basename, "_adj_with_blocks")
-  basename <- paste0("results/", basename)
 
   res_file <- paste0(basename, "_", method, ".rds")
   if (!file.exists(res_file)) {
@@ -72,10 +78,9 @@ grid$effects <- furrr::future_pmap(grid[1:6], function(
     res <- readRDS(res_file)
     prs_effects <- rep(0, ncol(G))
 
-    sumstats <- readRDS(sumstats_file)[c("_NUM_ID_", "info")]
-
-    num_id <- sumstats[[1]]
[…]

### 2022-04-13 code/run-PRS-BBJ.R

@@ -1,16 +1,21 @@
 
-bigassertr::assert_dir("results")
+bigassertr::assert_dir("results_BBJ")
 bigassertr::assert_dir("log")
 
+ALL_CORR <- c(
+  corr_JPN               = "data/corr_JPN/chr",
+  corr_JPN_with_blocks   = "data/corr_JPN/adj_with_blocks_chr",
+  corr_1000G             = "data/corr_1000G_EAS/chr",
+  corr_1000G_with_blocks = "data/corr_1000G_EAS/adj_with_blocks_chr",
+  corr_UKBB              = "data/corr_UKBB_EAS/chr",
+  corr_UKBB_with_blocks  = "data/corr_UKBB_EAS/adj_with_blocks_chr"
+)
+
 library(dplyr)
-files <- tibble(pheno = c(paste0("t1d_", c("affy", "illu")),
-                          paste0("brca_", c("onco", "icogs")),
-                          "cad", "prca", "mdd")) %>%
-  mutate(whichN = NA) %>%
-  add_row(pheno = "vitaminD", whichN = c("trueN", "maxN")) %>%
-  tidyr::expand_grid(
-    qc = c("noqc", "qc1", "qc2"),
-    name_corr = c("data/corr/chr", "data/corr/adj_with_blocks_chr")) %>%
+files <- tibble(pheno = c("height", "systolic_bp", "hdl_cholesterol", "bmi")) %>%
+  tidyr::expand_grid(name_corr = names(ALL_CORR)) %>%
+  filter(!file.exists(paste0("results_BBJ/", pheno, "_", name_corr,
+                             "_LDpred2-auto-rob.rds"))) %>%
   print()
 
 library(future.batchtools)
@@ -19,25 +24,13 @@ plan(batchtools_slurm(resources = list(
   t = "12:00:00", c = NCORES + 2, mem = "120g",
   name = basename(rstudioapi::getSourceEditorContext()$path))))
 
-furrr::future_pwalk(files, function(pheno, whichN, qc, name_corr) {
+furrr::future_pwalk(files, function(pheno, qc, name_corr) {
 
-  # pheno <- "vitaminD"
-  # whichN <- "trueN"
-  # qc <- "qc1"
-  # name_corr <- "data/corr/adj_with_blocks_chr"
+  # pheno <- "height"
+  # name_corr <- "corr_JPN"
 
[…]

### 2022-04-13 code/run-PRS-FIN.R

@@ -1,16 +1,21 @@
 
-bigassertr::assert_dir("results")
+bigassertr::assert_dir("results_FIN")
 bigassertr::assert_dir("log")
 
+ALL_CORR <- c(
+  corr_UK             = "data/corr/chr",
+  corr_UK_with_blocks = "data/corr/adj_with_blocks_chr",
+  corr_FIN             = "data/corr_FIN/chr",
+  corr_FIN_with_blocks = "data/corr_FIN/adj_with_blocks_chr",
+  corr_1000G             = "data/corr_1000G_EUR/chr",
+  corr_1000G_with_blocks = "data/corr_1000G_EUR/adj_with_blocks_chr"
+)
+
 library(dplyr)
-files <- tibble(pheno = c(paste0("t1d_", c("affy", "illu")),
-                          paste0("brca_", c("onco", "icogs")),
-                          "cad", "prca", "mdd")) %>%
-  mutate(whichN = NA) %>%
-  add_row(pheno = "vitaminD", whichN = c("trueN", "maxN")) %>%
-  tidyr::expand_grid(
-    qc = c("noqc", "qc1", "qc2"),
-    name_corr = c("data/corr/chr", "data/corr/adj_with_blocks_chr")) %>%
+files <- tibble(pheno = c("brca", "prca", "cad", "t2d", "t1d")) %>%
+  tidyr::expand_grid(name_corr = names(ALL_CORR)) %>%
+  filter(!file.exists(paste0("results_FIN/", pheno, "_", name_corr,
+                             "_LDpred2-auto-rob.rds"))) %>%
   print()
 
 library(future.batchtools)
@@ -19,25 +24,13 @@ plan(batchtools_slurm(resources = list(
   t = "12:00:00", c = NCORES + 2, mem = "120g",
   name = basename(rstudioapi::getSourceEditorContext()$path))))
 
-furrr::future_pwalk(files, function(pheno, whichN, qc, name_corr) {
+furrr::future_pwalk(files, function(pheno, qc, name_corr) {
 
-  # pheno <- "vitaminD"
-  # whichN <- "trueN"
-  # qc <- "qc1"
-  # name_corr <- "data/corr/adj_with_blocks_chr"
+  # pheno <- "brca"
+  # name_corr <- "corr_FIN"
 
-  gwas_file <- paste0("data/sumstats/", pheno, "_", qc, ".rds")
[…]

### 2022-04-13 code/simu-rounded.R

@@ -80,7 +80,8 @@ furrr::future_walk(1:10, function(ic) {
 library(dplyr)
 all_res <- list.files("results-simu-rounded", full.names = TRUE) %>%
   purrr::map_dfr(readRDS) %>%
-  mutate(with_blocks = rep(c(FALSE, FALSE, TRUE, TRUE), 18)) %>%
+  mutate(with_blocks = rep(c(FALSE, FALSE, TRUE, TRUE), 18),
+         rounded = ifelse(rounded, "Yes", "No")) %>%
   tidyr::unnest(c(Method, r2)) %>%
   group_by(Method, with_blocks, rounded) %>%
   summarise(r2 = {
@@ -125,7 +126,7 @@ ggplot(all_res %>% filter(!with_blocks) %>%
          mutate(Method = factor(Method, levels = ALL_METHODS)),
        aes(Method, mean, fill = rounded)) +
   bigstatsr::theme_bigstatsr(0.8) +
-  scale_fill_manual(values =  c("#E69F00", "#56B4E9", "#009E73")) +
+  scale_fill_manual(values =  c("#E69F00", "#56B4E9")) +
   geom_col(position = position_dodge(), alpha = 0.6, color = "black", size = 1) +
   geom_errorbar(aes(ymin = inf, ymax = sup),
                 position = position_dodge(width = 0.9),

### 2022-04-21 paper/paper-misspec.tex
reexport
formatting

@@
 
-%\linenumbers
-%\doublespacing
-\onehalfspacing
+\linenumbers
+\doublespacing
+%\onehalfspacing
 %\usepackage[authoryear]{natbib}
-\usepackage{natbib} %\bibpunct{(}{)}{;}{author-year}{}{,}
+\usepackage[numbers,super]{natbib} %\bibpunct{(}{)}{;}{author-year}{}{,}
 
@@
 
-\vspace*{6em}
+%\vspace*{6em}
+\clearpage
 
@@
 
+\section{Materials and Methods}
+
+\subsection{Data for simulations}
+
+We use the UK Biobank imputed (BGEN) data \cite[]{bycroft2018uk}. 
+We restrict individuals to the ones used for computing the principal components (PCs) in the UK Biobank (Field 22020).
+These individuals are unrelated and have passed some quality control including removing samples with a missing rate on autosomes larger than 0.02, having a mismatch between inferred sex and self-reported sex, and outliers based on heterozygosity (more details can be found in section S3 of \citet{bycroft2018uk}).
+To get a set of genetically homogeneous individuals, we compute a robust Mahalanobis distance based on the first 16 PCs and further restrict individuals to those within a log-distance of 5 \cite[]{prive2020efficient}.
+This results in 362,307 individuals.
+We randomly sample 300,000 individuals to form a training set (e.g.\ to run the GWAS), 10,000 individuals to form a validation set (to tune hyper-parameters), and use the remaining 52,307 individuals to form a test set (to evaluate final predictive models).
+
+Among genetic variants on chromosome 22 and with a minor allele frequency larger than 0.01 and an imputation INFO score larger than 0.4 (as reported by the UK Biobank), we sample 40,000 of them according to the inverse of the INFO score density so that they have varying levels of imputation accuracy (Figure \ref{fig:hist-info}).
+We read the UK Biobank data into two different datasets using function \texttt{snp\_readBGEN} from R package bigsnpr \cite[]{prive2017efficient}, one by reading the BGEN data at random according to imputation probabilities, and another one reading it as dosages (i.e.\ expected values according to imputation probabilities).
+The first dataset is used as what could be the real genotype calls and the second dataset as what would be its imputed version; this design technique was used in \citet{prive2019making}.
+
+\subsection{Data for real analyses}
+
+We also use the UK Biobank data for validation/testing in real data analyses, and use the same individuals as described in the previous section.
+We sample 10,000 individuals to form a validation set and use the remaining 352,307 individuals as test set.
+We restrict to the 1,054,315 HapMap3 variants used in the LD reference provided in \citet{prive2020ldpred2}.
+
+To define phenotypes in the UK Biobank, we first map ICD10 and ICD9 codes (UKBB fields 40001, 40002, 40006, 40013, 41202, 41270 and 41271) to phecodes using R package PheWAS \cite[]{carroll2014r,wu2019mapping}.
+We also use some continuous phenotypes, namely vitamin D (Data-Field 30890), height (50), BMI (21001), systolic blood pressure (4080), and HDL cholesterol (30760).
+
+We use published GWAS summary statistics listed in table \ref{tab:sumstats} to derive polygenic scores.
+We also use GWAS summary statistics for five disease endpoints from FinnGen\cite{kurki2022finngen} (release 6) and for four continuous outcomes from Biobank Japan \cite[]{sakaue2021cross}.
+
+\begin{table}[h]
+	\centering
+	\begin{tabular}{|l||l|c|c||c|c|}
+		\hline
+		Trait & GWAS citation & \begin{tabular}{@{}c@{}}Effective GWAS\\sample size\end{tabular} & \# GWAS variants & \begin{tabular}{@{}c@{}}\# matched variants\\with INFO > 0.4\end{tabular} & Mean INFO \\
+		\hline
+		Breast cancer (BRCA) [iCOGS] & \citet{michailidou2017association} & ~~87,037 & 11,792,542 & 1,051,242 & 0.841 \\
+		Breast cancer (BRCA) [OncoArray] & \citet{michailidou2017association} & 104,442 & 11,792,542 & 1,054,233 & 0.968 \\
+		Type 1 diabetes (T1D) [Affymetrix] & \citet{censin2017childhood} & ~~~~~5516  & ~~8,996,866 & ~~~934,712 & 0.885 \\	
+		Type 1 diabetes (T1D) [Illumina] & \citet{censin2017childhood} & ~~~~~7982  & ~~8,996,866 & ~~~949,334 & 0.942 \\
+		Prostate cancer (PRCA) & \citet{schumacher2018association} & 135,316 & 20,370,946 & ~~~818,400 & 0.969 \\
+		Depression (MDD) [without UKBB] & \citet{wray2018genome} & 110,464 & ~~9,874,289 & 1,049,455 & 0.968 \\
+		Coronary artery disease (CAD) & \citet{nikpay2015comprehensive} & 129,014 & ~~9,455,778 & 1,052,200 & 0.982 \\
+		Vitamin D & \citet{jiang2018genome} & ~~79,366 & ~~2,579,296 & 1,016,935 & NA \\
+		\hline
+	\end{tabular}
+	\caption{Summary of external GWAS summary statistics used.
+PRCA summary statistics have many variants with a missing INFO score, which we discard.
+We also restrict to variants with an INFO score larger than 0.4.
+Vitamin D summary statistics do not report INFO scores.
+\label{tab:sumstats}}
+\end{table}
+
+\subsection{GWAS sample size imputation}
+
+In this paper, we extensively use the following formula
+\begin{equation}\label{eq:approx-sd-lin}
+\text{sd}(G_j) \approx \dfrac{\text{sd}(y)}{\sqrt{n_j ~ \text{se}(\hat{\gamma}_j)^2 + \hat{\gamma}_j^2}} ~,
+\end{equation}
+where $\hat{\gamma}_j$ is the marginal GWAS effect size of variant $j$, $n_j$ is the GWAS sample size associated with variant $j$, $y$ is the vector of phenotypes and ${G}_j$ is the vector of genotypes for variant $j$.
+This formula is used in LDpred2 \cite[]{prive2020ldpred2,prive2021high}. Note that, for a binary trait for which logistic regression is used, we have instead
+\begin{equation}\label{eq:approx-sd-log}
+\text{sd}(G_j) \approx \dfrac{2}{\sqrt{n_j^\text{eff} ~ \text{se}(\hat{\gamma}_j)^2 + \hat{\gamma}_j^2}} ~,
+\end{equation}
+where $n_j^\text{eff} = \frac{4}{1 / n_j^\text{cases} + 1 / n_j^\text{controls}}$.
+
+We can then impute $n_j$ from equation \eqref{eq:approx-sd-lin} using
+\begin{equation}\label{eq:imputeN}
+n_j \approx \dfrac{\text{var}(y) / \text{var}(G_j) - \hat{\gamma}_j^2}{\text{se}(\hat{\gamma}_j)^2} ~,
+\end{equation}
+and impute $n_j^\text{eff}$ from equation \eqref{eq:approx-sd-log} using
+\begin{equation}\label{eq:imputeNeff}
+n_j^\text{eff} \approx \dfrac{4 / \text{var}(G_j) - \hat{\gamma}_j^2}{\text{se}(\hat{\gamma}_j)^2} ~.
+\end{equation}
+In practice, we also bound this estimate to be between $0.5 \cdot N$ and $1.1 \cdot N$, where $N$ is the total (effective) sample size.
+Also note that the estimate of $\text{var}(G_j)$ has to account for the imputation accuracy, i.e.\ use $2 \cdot f_j \cdot (1 - f_j) \cdot \text{INFO}_j$ (Section \ref{sec:imputation}).
+
+\subsection{New implementation of lassosum in bigsnpr}\label{lassosum2}
+
+Instead of using a regularized version of the correlation matrix $R$ parameterized by $s$, $R_s = (1 - s) R + s I$ (where $0 < s \leq 1$), we use $R_{\delta} = R + \delta I$ (where $\delta > 0$), which makes it clearer that lassosum is also using L2-regularization (therefore elastic-net).
+Then, from \citet{mak2017polygenic}, the solution from lassosum can be obtained by iteratively updating 
+\[
+\beta_j^{(t)} =
+\begin{cases}
+\dfrac{\text{sign}\big(u_j^{(t)}\big) \big( \big|u_j^{(t)}\big| - \lambda \big)}{\widetilde{X}_j^T \widetilde{X}_j + \delta} & \text{if } \big|u_j^{(t)}\big| > \lambda ~, \\
+0 & \text{otherwise.}
+\end{cases}
+\]
+where 
+\[
+u_j^{(t)} = r_j - \widetilde{X}_j^T \big( \widetilde{X} \beta^{(t-1)} - \widetilde{X}_j \beta_j^{(t-1)} \big) ~.
+\]
+Following the notations from \citet{prive2020ldpred2}, denote $\widetilde{X} = \frac{1}{\sqrt{n-1}} C_n G S^{-1}$, where $G$ is the genotype matrix, $C_n$ is the centering matrix and $S$ is the diagonal matrix of standard deviations of the columns of $G$.
+Then $\widetilde{X}_j^T \widetilde{X} = R_{j,.} = R_{.,j}^T$ and $\widetilde{X}_j^T \widetilde{X}_j = 1$.
+Moreover, using the notations from \citet{prive2020ldpred2}, $u_j^{(t)} = \widehat{\beta}_j - R_{.,j}^T \beta^{(t-1)} + \beta_j^{(t-1)}$,
+where $r_j = \widehat{\beta}_j =  \dfrac{\widehat{\gamma}_j}{\sqrt{n_j ~ \text{se}(\widehat{\gamma}_j)^2 + \widehat{\gamma}_j^2}}$ and $\widehat{\gamma}_j$ is the GWAS effect of variant $j$ and $n$ is the GWAS sample size \cite[]{mak2017polygenic,prive2021high}.
+Then computing $R_{.,j}^T \beta^{(t-1)}$ is the most time-consuming part of each iteration.
+To make this faster, instead of computing $R_{.,j}^T \beta^{(t-1)}$ at each iteration ($j$ and $t$), we can start with an initial vector of 0s only (for all $j$) since $\beta^{(0)} \equiv 0$, then update this vector when $\beta_j^{(t)} \neq \beta_j^{(t-1)}$ only.
+Note that only positions $k$ for which $R_{k,j} \neq 0$ must be updated in this vector $R_{.,j}^T \beta^{(t-1)}$.
+
+In this new implementation of the lassosum model, which we call lassosum2, the input parameters are the correlation matrix $R$, the GWAS summary statistics ($\widehat{\gamma}_j$, $\text{se}(\widehat{\gamma}_j)$ and $n_j$), and the two hyper-parameters $\lambda$ and $\delta$.
+Therefore, except for the two hyper-parameters, lassosum2 uses the exact same input parameters as LDpred2 \cite[]{prive2020ldpred2}.
+We try $\delta \in \{0.001, 0.005, 0.02, 0.1, 0.6, 3\}$ by default in lassosum2, instead of $s \in \{0.2, 0.5, 0.8, 1.0\}$ in lassosum.
+For $\lambda$, the default in lassosum uses a sequence of 20 values equally spaced on a log scale between 0.1 and 0.001.
+By default in lassosum2, we use a similar sequence, but between $\lambda_0$ and $\lambda_0 / 100$ instead, where $\lambda_0 = \max_j \big|\widehat{\beta}_j\big|$ is the minimum $\lambda$ for which no variable enters the model because the L1-regularization is too strong.
+Note that we do not provide an ``auto'' version using pseudo-validation (as in \citet{mak2017polygenic}) as we have not found it to be robust enough (Figure \ref{fig:pseudoval}).
+Also note that, as in LDpred2, we run lassosum2 genome-wide using a sparse correlation matrix which assumes that variants further away than 3 cM are not correlated.
+Although we do not require splitting the genome into independent LD blocks anymore (as done in lassosum), we recommend to do so for robustness in LDpred2, and for extra speed gain (cf.\ Methods section \ref{sec:LD}).
+
+\subsection{LDpred2-low-h2 and LDpred2-auto-rob}\label{sec:robust}
+
+Here we introduce the small changes made to LDpred2 (-grid and -auto) in order to make them more robust.
+First, LDpred2-low-h2 simply consists in running LDpred2-grid while testing $h^2$ within $\{0.3, 0.7, 1, 1.4\} \cdot h^2_\text{LDSC}$ (note the added $0.3$ compared to \citet{prive2020ldpred2}), where $h^2_\text{LDSC}$ is the heritability estimate from LD score regression.
+Indeed, we show in simulations here that using lower values for $h^2$ may provide higher predictive performance in the case of misspecifications (thanks to more shrinkage of the effects).
+In simulations, because of the large misspecifications, we use a larger grid over $\{0.01, 0.1, 0.3, 0.7, 1, 1.4\} \cdot h^2_\text{LDSC}$.
+
+For LDpred2-auto, we introduce two new parameters.
+The first one, \texttt{shrink\_corr}, allows for shrinking off-diagonal elements of the correlation matrix.
+This is similar to parameter `s' in lassosum (Methods section \ref{lassosum2}), and acts as a means of regularization.
+We use a value of 0.9 in simulations and 0.95 in real data when running ``LDpred2-auto-rob'' here, and the default value of 1 (with no effect) when running ``LDpred2-auto''.
+The second new parameter, \texttt{allow\_jump\_sign}, controls whether variant effect sizes can change sign over two consecutive iterations of the Gibbs sampler.
+When setting this parameter to false (in the method we refer to as ``LDpred2-auto-rob'' here), this forces the effects to go through 0 before changing sign.
+This is useful to prevent instability (oscillation and ultimately divergence) of the Gibbs sampler under large misspecifications, and is also useful for accelerating convergence of chains with a large initial value for $p$ (the proportion of causal variants).
+
+\subsection{New LD reference}\label{sec:LD}
+
+We form nearly independent LD blocks using the optimal algorithm developed in \citet{prive2021optimal}.
+For different numbers of blocks and maximum number of variants in each block, we use the split with the minimum cost within the ones reducing the original number of non-zero values to less than 60\% (70\% for chromosome 6).
+Having a correlation matrix with independent blocks prevents small errors in the algorithm (e.g.\ the Gibbs sampler in LDpred2) from propagating to too many variants.
+It also makes running LDpred2 (and lassosum2) faster, taking about 60\% of the initial time (since only 60\% of the initial non-zero values of the correlation matrix are kept).
+
+We have also developed a new ``compact'' format for the SFBMs (sparse matrices on disk).
+Instead of using something similar to the standard ``compressed sparse column'' format which stores all $\{i,~x(i, j)\}$ for a given column $j$, we only store the first index $i_0$ and all the contiguous values $\{x(i_0, j),~x(i_0 + 1, j),~\dots\}$ up to the last non-zero value for this column $j$.
+This makes this format about twice as efficient for both LDpred2 and lassosum2 (in terms of both memory and speed).
+Therefore, using both this new format and the LD blocks should divide computation times by 3.
+
+\subsection{Alternative LD reference for FinnGen and Biobank Japan}\label{sec:altLD}
+
+To be used with GWAS summary statistics from FinnGen, we investigate three different LD reference panels.
+To define homogeneous ancestry groups in the UK Biobank, we include all individuals within a specific distance to a population center in the PCA space \cite[]{prive2021high,prive2021using}.
+We use either the 503 European (including 99 Finnish) individuals from the 1000 Genomes (1000G) data \cite[]{10002015global}, 404 Finnish-like UKBB + the 99 Finnish 1000G individuals (also 503 in total), or the 10,000 UK individuals from the validation set we use in this paper.
+
+To be used with GWAS summary statistics from Biobank Japan, we also investigate three different LD reference panels.
+We use either the 504 East Asian (including 104 Japanese) individuals from the 1000G, 400 Japanese-like UKBB + the 104 Japanese 1000G individuals (also 504 in total), or 2041 East Asian UKBB individuals (including the 400 previous Japanese individuals).
+We also construct versions of these LD references with independent LD blocks, as described in the previous section.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \section{Results}
@@
 \begin{figure}[!h]
-	\centerline{\includegraphics[width=\textwidth]{simu-misN}}
+	\centerline{\includegraphics[width=0.9\textwidth]{simu-misN}}
 	\caption{Results for the simulations with sample size misspecification, averaged over 10 simulations for each scenario.
@@
 
-\cite{marchini2010genotype} showed that the IMPUTE INFO measure is highly concordant with the MACH measure $\hat{r}_j^2 = \frac{\text{var}(G_j)}{2 \hat{\theta}_j (1 - \hat{\theta}_j)}$, where $\hat{\theta}_j$ is the estimated allele frequency of $G_j$, the genotypes for variant $j$.
+\citet{marchini2010genotype} showed that the IMPUTE INFO measure is highly concordant with the MACH measure $\hat{r}_j^2 = \frac{\text{var}(G_j)}{2 \hat{\theta}_j (1 - \hat{\theta}_j)}$, where $\hat{\theta}_j$ is the estimated allele frequency of $G_j$, the genotypes for variant $j$.
 Therefore, when using the expected genotypes from imputation (allele dosages), their standard deviations are often lower than the expected value under Hardy--Weinberg equilibrium, because $\text{INFO}_j \approx \frac{\text{var}(G_j)}{2 \hat{\theta}_j (1 - \hat{\theta}_j)}$.
@@
 \begin{figure}[!h]
-	\centerline{\includegraphics[width=\textwidth]{simu-info}}
+	\centerline{\includegraphics[width=0.9\textwidth]{simu-info}}
 	\caption{Results of predictive performance for the simulations using GWAS summary statistics from imputed dosage data, averaged over 10 simulations for each scenario.
@@
 We use the same simulation setup as before (Methods).
-In addition, we design an alternative reference panel based on 10,000 individuals from South Europe by using the ``Italy'' center defined in \cite{prive2021high}.
+In addition, we design an alternative reference panel based on 10,000 individuals from South Europe by using the ``Italy'' center defined in \citet{prive2021high}.
 When using this alternative LD reference panel instead of a well-matched one as in the previous sections, squared correlations between the polygenic scores and the simulated phenotypes drop from 0.174 to 0.167 for lassosum2, from 0.168 to 0.147 for LDpred2(-grid), from 0.174 to 0.169 for LDpred2-low-h2, from 0.143 to 0.139 for LDpred2-auto, from 0.162 to 0.146 for LDpred2-auto-rob, and from 0.144 to 0.140 for LDpred2-inf (Figure \ref{fig:simu-altpop}, averaging over 10 simulations).
@@
 \begin{figure}[!h]
-	\centerline{\includegraphics[width=\textwidth]{simu-altpop}}
+	\centerline{\includegraphics[width=0.9\textwidth]{simu-altpop}}
 	\caption{Results for the simulations with summary statistics with LD matrices based on two different populations.
@@
 \begin{figure}[!h]
-	\centerline{\includegraphics[width=\textwidth]{brca_onco_qc}}
+	\centerline{\includegraphics[width=0.9\textwidth]{brca_onco_qc}}
 	\caption{Standard deviations inferred from the OncoArray breast cancer GWAS summary statistics using equation \eqref{eq:approx-sd-log} (\textbf{A:} Raw or \textbf{B:} dividing them by $\sqrt{\text{INFO}}$) versus the ones inferred from the reported GWAS allele frequencies $f_j$ (using $\sqrt{2 f_j (1 - f_j)}$).
@@
 We apply these corrections to the two breast cancer summary statistics. 
-In the comparison, we use the QC proposed in \cite{prive2020ldpred2} first (which ends up filtering on MAF here, which we call ``qc1'').
+In the comparison, we use the QC proposed in \citet{prive2020ldpred2} first (which ends up filtering on MAF here, which we call ``qc1'').
 We then also filter out the two long-range LD regions of chromosome 6 and 8 for the OncoArray summary statistics and remove around 500 variants when filtering on differences of AFs between summary statistics and the validation dataset (``qc2'').
@@
 For MDD and PRCA, most changes introduced before have little to no impact on predictive performance.
-For CAD and T1D, the QC proposed in \cite{prive2020ldpred2} and the additional QC proposed here provides much better predictive performance, especially for LDpred2-auto (and LDpred2-auto-rob) than when using no quality control, showing how important this preliminary step is.
+For CAD and T1D, the QC proposed in \citet{prive2020ldpred2} and the additional QC proposed here provides much better predictive performance, especially for LDpred2-auto (and LDpred2-auto-rob) than when using no quality control, showing how important this preliminary step is.
 
 \begin{figure}[!h]
-	\centerline{\includegraphics[width=0.67\textwidth]{res-cad}}
+	\centerline{\includegraphics[width=0.85\textwidth]{res-cad}}
 	\caption{Variance explained of CAD in the UK Biobank by PGS derived from external summary statistics.
@@
 \begin{figure}[!h]
-	\centerline{\includegraphics[width=0.67\textwidth]{res-vitaminD}}
+	\centerline{\includegraphics[width=0.85\textwidth]{res-vitaminD}}
 	\caption{Variance explained of vitamin D in the UK Biobank by PGS derived from external summary statistics.
@@
 Here we investigate the use of different LD reference panels with GWAS summary statistics from two large biobanks of isolated populations.
-We use GWAS summary statistics for five disease endpoints from FinnGen (release 6, \cite{kurki2022finngen}), namely breast and prostate cancers (BrCa and PrCa), coronary artery disease (CAD), type 1 and type 2 diabetes (T1D and T2D).
+We use GWAS summary statistics for five disease endpoints from FinnGen\cite{kurki2022finngen} (release 6), namely breast and prostate cancers (BrCa and PrCa), coronary artery disease (CAD), type 1 and type 2 diabetes (T1D and T2D).
 These were derived using SAIGE \cite[]{zhou2018efficiently}.
@@
 These were derived using BOLT-LMM-inf \cite[]{loh2018mixed}.
-First, for each phenotype, we estimate a global power improvement provided by BOLT-LMM by comparing $\chi^2$-statistics (the boost of BOLT-LMM-inf versus linear regression) across genome-wide significant variants, as recommended in \cite{loh2018mixed}.
+First, for each phenotype, we estimate a global power improvement provided by BOLT-LMM by comparing $\chi^2$-statistics (the boost of BOLT-LMM-inf versus linear regression) across genome-wide significant variants, as recommended in \citet{loh2018mixed}.
 This estimated power ratio, which corresponds to the increase in effective sample size, is 1.143 for height, 1.039 for HDL cholesterol, 1.025 for BMI, and 1.013 for systolic blood pressure.
@@
 
-\section{Materials and Methods}
-
-\subsection{Data for simulations}
-
-We use the UK Biobank imputed (BGEN) data \cite[]{bycroft2018uk}. 
-We restrict individuals to the ones used for computing the principal components (PCs) in the UK Biobank (Field 22020).
-These individuals are unrelated and have passed some quality control including removing samples with a missing rate on autosomes larger than 0.02, having a mismatch between inferred sex and self-reported sex, and outliers based on heterozygosity (more details can be found in section S3 of \cite{bycroft2018uk}).
-To get a set of genetically homogeneous individuals, we compute a robust Mahalanobis distance based on the first 16 PCs and further restrict individuals to those within a log-distance of 5 \cite[]{prive2020efficient}.
-This results in 362,307 individuals.
-We randomly sample 300,000 individuals to form a training set (e.g.\ to run the GWAS), 10,000 individuals to form a validation set (to tune hyper-parameters), and use the remaining 52,307 individuals to form a test set (to evaluate final predictive models).
-
-Among genetic variants on chromosome 22 and with a minor allele frequency larger than 0.01 and an imputation INFO score larger than 0.4 (as reported by the UK Biobank), we sample 40,000 of them according to the inverse of the INFO score density so that they have varying levels of imputation accuracy (Figure \ref{fig:hist-info}).
-We read the UK Biobank data into two different datasets using function \texttt{snp\_readBGEN} from R package bigsnpr \cite[]{prive2017efficient}, one by reading the BGEN data at random according to imputation probabilities, and another one reading it as dosages (i.e.\ expected values according to imputation probabilities).
-The first dataset is used as what could be the real genotype calls and the second dataset as what would be its imputed version; this design technique was used in \cite{prive2019making}.
-
-\subsection{Data for real analyses}
-
-We also use the UK Biobank data for validation/testing in real data analyses, and use the same individuals as described in the previous section.
-We sample 10,000 individuals to form a validation set and use the remaining 352,307 individuals as test set.
-We restrict to the 1,054,315 HapMap3 variants used in the LD reference provided in \cite{prive2020ldpred2}.
-
-To define phenotypes in the UK Biobank, we first map ICD10 and ICD9 codes (UKBB fields 40001, 40002, 40006, 40013, 41202, 41270 and 41271) to phecodes using R package PheWAS \cite[]{carroll2014r,wu2019mapping}.
-We also use some continuous phenotypes, namely vitamin D (Data-Field 30890), height (50), BMI (21001), systolic blood pressure (4080), and HDL cholesterol (30760).
-
-We use published GWAS summary statistics listed in table \ref{tab:sumstats} to derive polygenic scores.
-We also use GWAS summary statistics for five disease endpoints from FinnGen (release 6, \cite{kurki2022finngen}) and for four continuous outcomes from Biobank Japan \cite[]{sakaue2021cross}.
-
-\begin{table}[h]
-	\centering
-	\begin{tabular}{|l||c|c|c||c|c|}
-		\hline
-		Trait & GWAS citation & \begin{tabular}{@{}c@{}}Effective GWAS\\sample size\end{tabular} & \# GWAS variants & \begin{tabular}{@{}c@{}}\# matched variants\\with INFO > 0.4\end{tabular} & Mean INFO \\
-		\hline
-		Breast cancer (BRCA) [iCOGS] & \cite{michailidou2017association} & ~~87,037 & 11,792,542 & 1,051,242 & 0.841 \\
-		Breast cancer (BRCA) [OncoArray] & \cite{michailidou2017association} & 104,442 & 11,792,542 & 1,054,233 & 0.968 \\
-		Type 1 diabetes (T1D) [Affymetrix] & \cite{censin2017childhood} & ~~~~~5516  & ~~8,996,866 & ~~~934,712 & 0.885 \\	
-		Type 1 diabetes (T1D) [Illumina] & \cite{censin2017childhood} & ~~~~~7982  & ~~8,996,866 & ~~~949,334 & 0.942 \\
-		Prostate cancer (PRCA) & \cite{schumacher2018association} & 135,316 & 20,370,946 & ~~~818,400 & 0.969 \\
-		Depression (MDD) [without UKBB] & \cite{wray2018genome} & 110,464 & ~~9,874,289 & 1,049,455 & 0.968 \\
-		Coronary artery disease (CAD) & \cite{nikpay2015comprehensive} & 129,014 & ~~9,455,778 & 1,052,200 & 0.982 \\
-		Vitamin D & \cite{jiang2018genome} & ~~79,366 & ~~2,579,296 & 1,016,935 & NA \\
-		\hline
-	\end{tabular}
-	\caption{Summary of external GWAS summary statistics used.
-PRCA summary statistics have many variants with a missing INFO score, which we discard.
-We also restrict to variants with an INFO score larger than 0.4.
-Vitamin D summary statistics do not report INFO scores.
-\label{tab:sumstats}}
-\end{table}
-
-\subsection{GWAS sample size imputation}
-
-In this paper, we extensively use the following formula
-\begin{equation}\label{eq:approx-sd-lin}
-\text{sd}(G_j) \approx \dfrac{\text{sd}(y)}{\sqrt{n_j ~ \text{se}(\hat{\gamma}_j)^2 + \hat{\gamma}_j^2}} ~,
-\end{equation}
-where $\hat{\gamma}_j$ is the marginal GWAS effect size of variant $j$, $n_j$ is the GWAS sample size associated with variant $j$, $y$ is the vector of phenotypes and ${G}_j$ is the vector of genotypes for variant $j$.
-This formula is used in LDpred2 \cite{prive2020ldpred2,prive2021high}. Note that, for a binary trait for which logistic regression is used, we have instead
-\begin{equation}\label{eq:approx-sd-log}
-\text{sd}(G_j) \approx \dfrac{2}{\sqrt{n_j^\text{eff} ~ \text{se}(\hat{\gamma}_j)^2 + \hat{\gamma}_j^2}} ~,
-\end{equation}
-where $n_j^\text{eff} = \frac{4}{1 / n_j^\text{cases} + 1 / n_j^\text{controls}}$.
-
-We can then impute $n_j$ from equation \eqref{eq:approx-sd-lin} using
-\begin{equation}\label{eq:imputeN}
-n_j \approx \dfrac{\text{var}(y) / \text{var}(G_j) - \hat{\gamma}_j^2}{\text{se}(\hat{\gamma}_j)^2} ~,
-\end{equation}
-and impute $n_j^\text{eff}$ from equation \eqref{eq:approx-sd-log} using
-\begin{equation}\label{eq:imputeNeff}
-n_j^\text{eff} \approx \dfrac{4 / \text{var}(G_j) - \hat{\gamma}_j^2}{\text{se}(\hat{\gamma}_j)^2} ~.
-\end{equation}
-In practice, we also bound this estimate to be between $0.5 \cdot N$ and $1.1 \cdot N$, where $N$ is the total (effective) sample size.
-Also note that the estimate of $\text{var}(G_j)$ has to account for the imputation accuracy, i.e.\ use $2 \cdot f_j \cdot (1 - f_j) \cdot \text{INFO}_j$ (Section \ref{sec:imputation}).
-
-\subsection{New implementation of lassosum in bigsnpr}\label{lassosum2}
-
-Instead of using a regularized version of the correlation matrix $R$ parameterized by $s$, $R_s = (1 - s) R + s I$ (where $0 < s \leq 1$), we use $R_{\delta} = R + \delta I$ (where $\delta > 0$), which makes it clearer that lassosum is also using L2-regularization (therefore elastic-net).
-Then, from \cite{mak2017polygenic}, the solution from lassosum can be obtained by iteratively updating 
-\[
-\beta_j^{(t)} =
-\begin{cases}
-\dfrac{\text{sign}\big(u_j^{(t)}\big) \big( \big|u_j^{(t)}\big| - \lambda \big)}{\widetilde{X}_j^T \widetilde{X}_j + \delta} & \text{if } \big|u_j^{(t)}\big| > \lambda ~, \\
-0 & \text{otherwise.}
-\end{cases}
-\]
-where 
-\[
-u_j^{(t)} = r_j - \widetilde{X}_j^T \big( \widetilde{X} \beta^{(t-1)} - \widetilde{X}_j \beta_j^{(t-1)} \big) ~.
-\]
-Following the notations from \cite{prive2020ldpred2}, denote $\widetilde{X} = \frac{1}{\sqrt{n-1}} C_n G S^{-1}$, where $G$ is the genotype matrix, $C_n$ is the centering matrix and $S$ is the diagonal matrix of standard deviations of the columns of $G$.
-Then $\widetilde{X}_j^T \widetilde{X} = R_{j,.} = R_{.,j}^T$ and $\widetilde{X}_j^T \widetilde{X}_j = 1$.
-Moreover, using the notations from \cite{prive2020ldpred2}, $u_j^{(t)} = \widehat{\beta}_j - R_{.,j}^T \beta^{(t-1)} + \beta_j^{(t-1)}$,
-where $r_j = \widehat{\beta}_j =  \dfrac{\widehat{\gamma}_j}{\sqrt{n_j ~ \text{se}(\widehat{\gamma}_j)^2 + \widehat{\gamma}_j^2}}$ and $\widehat{\gamma}_j$ is the GWAS effect of variant $j$ and $n$ is the GWAS sample size \cite[]{mak2017polygenic,prive2021high}.
-Then computing $R_{.,j}^T \beta^{(t-1)}$ is the most time-consuming part of each iteration.
-To make this faster, instead of computing $R_{.,j}^T \beta^{(t-1)}$ at each iteration ($j$ and $t$), we can start with an initial vector of 0s only (for all $j$) since $\beta^{(0)} \equiv 0$, then update this vector when $\beta_j^{(t)} \neq \beta_j^{(t-1)}$ only.
-Note that only positions $k$ for which $R_{k,j} \neq 0$ must be updated in this vector $R_{.,j}^T \beta^{(t-1)}$.
-
-In this new implementation of the lassosum model, which we call lassosum2, the input parameters are the correlation matrix $R$, the GWAS summary statistics ($\widehat{\gamma}_j$, $\text{se}(\widehat{\gamma}_j)$ and $n_j$), and the two hyper-parameters $\lambda$ and $\delta$.
-Therefore, except for the two hyper-parameters, lassosum2 uses the exact same input parameters as LDpred2 \cite[]{prive2020ldpred2}.
-We try $\delta \in \{0.001, 0.005, 0.02, 0.1, 0.6, 3\}$ by default in lassosum2, instead of $s \in \{0.2, 0.5, 0.8, 1.0\}$ in lassosum.
-For $\lambda$, the default in lassosum uses a sequence of 20 values equally spaced on a log scale between 0.1 and 0.001.
-By default in lassosum2, we use a similar sequence, but between $\lambda_0$ and $\lambda_0 / 100$ instead, where $\lambda_0 = \max_j \big|\widehat{\beta}_j\big|$ is the minimum $\lambda$ for which no variable enters the model because the L1-regularization is too strong.
-Note that we do not provide an ``auto'' version using pseudo-validation (as in \cite{mak2017polygenic}) as we have not found it to be robust enough (Figure \ref{fig:pseudoval}).
-Also note that, as in LDpred2, we run lassosum2 genome-wide using a sparse correlation matrix which assumes that variants further away than 3 cM are not correlated.
-Although we do not require splitting the genome into independent LD blocks anymore (as done in lassosum), we recommend to do so for robustness in LDpred2, and for extra speed gain (cf.\ Methods section \ref{sec:LD}).
-
-\subsection{LDpred2-low-h2 and LDpred2-auto-rob}\label{sec:robust}
-
-Here we introduce the small changes made to LDpred2 (-grid and -auto) in order to make them more robust.
-First, LDpred2-low-h2 simply consists in running LDpred2-grid while testing $h^2$ within $\{0.3, 0.7, 1, 1.4\} \cdot h^2_\text{LDSC}$ (note the added $0.3$ compared to \cite{prive2020ldpred2}), where $h^2_\text{LDSC}$ is the heritability estimate from LD score regression.
-Indeed, we show in simulations here that using lower values for $h^2$ may provide higher predictive performance in the case of misspecifications (thanks to more shrinkage of the effects).
-In simulations, because of the large misspecifications, we use a larger grid over $\{0.01, 0.1, 0.3, 0.7, 1, 1.4\} \cdot h^2_\text{LDSC}$.
-
-For LDpred2-auto, we introduce two new parameters.
-The first one, \texttt{shrink\_corr}, allows for shrinking off-diagonal elements of the correlation matrix.
-This is similar to parameter `s' in lassosum (Methods section \ref{lassosum2}), and acts as a means of regularization.
-We use a value of 0.9 in simulations and 0.95 in real data when running ``LDpred2-auto-rob'' here, and the default value of 1 (with no effect) when running ``LDpred2-auto''.
-The second new parameter, \texttt{allow\_jump\_sign}, controls whether variant effect sizes can change sign over two consecutive iterations of the Gibbs sampler.
-When setting this parameter to false (in the method we refer to as ``LDpred2-auto-rob'' here), this forces the effects to go through 0 before changing sign.
-This is useful to prevent instability (oscillation and ultimately divergence) of the Gibbs sampler under large misspecifications, and is also useful for accelerating convergence of chains with a large initial value for $p$ (the proportion of causal variants).
-
-\subsection{New LD reference}\label{sec:LD}
-
-We form nearly independent LD blocks using the optimal algorithm developed in \cite{prive2021optimal}.
-For different numbers of blocks and maximum number of variants in each block, we use the split with the minimum cost within the ones reducing the original number of non-zero values to less than 60\% (70\% for chromosome 6).
-Having a correlation matrix with independent blocks prevents small errors in the algorithm (e.g.\ the Gibbs sampler in LDpred2) from propagating to too many variants.
-It also makes running LDpred2 (and lassosum2) faster, taking about 60\% of the initial time (since only 60\% of the initial non-zero values of the correlation matrix are kept).
-
-We have also developed a new ``compact'' format for the SFBMs (sparse matrices on disk).
-Instead of using something similar to the standard ``compressed sparse column'' format which stores all $\{i,~x(i, j)\}$ for a given column $j$, we only store the first index $i_0$ and all the contiguous values $\{x(i_0, j),~x(i_0 + 1, j),~\dots\}$ up to the last non-zero value for this column $j$.
-This makes this format about twice as efficient for both LDpred2 and lassosum2 (in terms of both memory and speed).
-Therefore, using both this new format and the LD blocks should divide computation times by 3.
-
-\subsection{Alternative LD reference for FinnGen and Biobank Japan}\label{sec:altLD}
-
-To be used with GWAS summary statistics from FinnGen, we investigate three different LD reference panels.
-To define homogeneous ancestry groups in the UK Biobank, we include all individuals within a specific distance to a population center in the PCA space \cite[]{prive2021high,prive2021using}.
-We use either the 503 European (including 99 Finnish) individuals from the 1000 Genomes (1000G) data \cite[]{10002015global}, 404 Finnish-like UKBB + the 99 Finnish 1000G individuals (also 503 in total), or the 10,000 UK individuals from the validation set we use in this paper.
-
-To be used with GWAS summary statistics from Biobank Japan, we also investigate three different LD reference panels.
-We use either the 504 East Asian (including 104 Japanese) individuals from the 1000G, 400 Japanese-like UKBB + the 104 Japanese 1000G individuals (also 504 in total), or 2041 East Asian UKBB individuals (including the 400 previous Japanese individuals).
-We also construct versions of these LD references with independent LD blocks, as described in the previous section.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 \clearpage
@@
 We have recomputed the allele frequencies and imputation INFO scores within the European subset used here and across all imputed variants in the UK Biobank, and made them available at \url{https://doi.org/10.6084/m9.figshare.16635388}.
-We have formed independent LD blocks within the Northwestern European LD references provided in \cite{prive2020ldpred2}, and made these updated versions available at \url{https://doi.org/10.6084/m9.figshare.19213299}.
+We have formed independent LD blocks within the Northwestern European LD references provided in \citet{prive2020ldpred2}, and made these updated versions available at \url{https://doi.org/10.6084/m9.figshare.19213299}.
 
@@
 B.J.V.\ is on Allelica's international advisory board.
-The other authors declare no competing interests.
+The other authors have no competing interests to declare.
 
@@
 
-\clearpage
-%\vspace*{3em}
+%\clearpage
+\vspace*{3em}
 

### 2022-06-16 paper/paper-misspec.tex
first text changes

@@
 		\hline
-		Differences in per-variant sample sizes & Possible violation of model assumptions resulting in poor model convergence & Imputation using equations \eqref{eq:imputeN} and \eqref{eq:imputeNeff}, and filtering$^{(2)}$   \\
+		Differences in per-variant sample sizes & Possible violation of model assumptions resulting in poor predictive performance & Imputation using equations \eqref{eq:imputeN} and \eqref{eq:imputeNeff}, and filtering$^{(2)}$   \\
 		\hline
@@
 		\hline
-		Any & Possible violation of model assumptions resulting in poor model convergence & Quality control, using more regularization, and constraining LD to blocks \\
+		Any & Possible violation of model assumptions resulting in poor predictive performance & Quality control, using more regularization, and constraining LD to blocks \\
 		\hline
@@
 Second, we investigate possible adjustments to apply to the input parameters of PGS methods using summary statistics, namely the reference LD (linkage disequilibrium) matrix, the GWAS effect sizes, their standard errors and corresponding sample sizes.
-For example, we show that GWAS effect sizes computed from imputed dosages are larger in magnitude compared to if computed from true genotypes.
+For example, we show that GWAS effect sizes computed from imputed dosages are larger in magnitude than when computed from true genotypes.
 Third, we introduce two new optional parameters in LDpred2-auto to make it more robust to these types of misspecification.
+Note that, in this paper, we use ``robust'' to mean ``similar predictive performance''. 
 We focus our investigations on LDpred2 and lassosum for two reasons.
@@
 Second, we reimplement and use a new version of lassosum, called lassosum2, that uses the exact same input parameters as LDpred2, which makes it easy for us to test the different QCs and adjustments presented here.
+We also investigate PRS-CS and SBayesR, two competitive PGS methods \cite[]{ge2019polygenic,lloyd2019improved}, for which we develop some code to convert between the different formats for the LD matrices.
 
@@
 To get a set of genetically homogeneous individuals, we compute a robust Mahalanobis distance based on the first 16 PCs and further restrict individuals to those within a log-distance of 5 \cite[]{prive2020efficient}.
-This results in 362,307 individuals.
+This results in 362,307 individuals of Northwestern European ancestry.
 We randomly sample 300,000 individuals to form a training set (e.g.\ to run the GWAS), 10,000 individuals to form a validation set (to tune hyper-parameters), and use the remaining 52,307 individuals to form a test set (to evaluate final predictive models).
@@
 
-\subsection{GWAS sample size imputation}
+\subsection{GWAS sample size imputation}\label{sec:imputeN}
 
@@
 \end{equation}
-In practice, we also bound this estimate to be between $0.5 \cdot N$ and $1.1 \cdot N$, where $N$ is the total (effective) sample size.
+In practice, we also bound these estimates to be between $0.5 \cdot N$ and $1.1 \cdot N$, where $N$ is the total (effective) sample size.
 Also note that the estimate of $\text{var}(G_j)$ has to account for the imputation accuracy, i.e.\ use $2 \cdot f_j \cdot (1 - f_j) \cdot \text{INFO}_j$ (Section \ref{sec:imputation}).
+In equations \eqref{eq:approx-sd-lin} and \eqref{eq:imputeN}, we estimate $\text{var}(y)$ by the first percentile (robust minimum) of $0.5 \left(N ~ \text{se}(\hat{\gamma}_j)^2 + \hat{\gamma}_j^2\right)$ \cite[]{prive2020ldpred2}.
 
@@
 Note that only positions $k$ for which $R_{k,j} \neq 0$ must be updated in this vector $R_{.,j}^T \beta^{(t-1)}$.
+Since bigsnpr v1.10.4, we now also use this updating strategy in LDpred2 (-grid and -auto), which makes it much faster, especially when the polygenicity is small.
 
@@
 By default in lassosum2, we use a similar sequence, but between $\lambda_0$ and $\lambda_0 / 100$ instead, where $\lambda_0 = \max_j \big|\widehat{\beta}_j\big|$ is the minimum $\lambda$ for which no variable enters the model because the L1-regularization is too strong.
-Note that we do not provide an ``auto'' version using pseudo-validation (as in \citet{mak2017polygenic}) as we have not found it to be robust enough (Figure \ref{fig:pseudoval}).
+Note that we do not provide an ``auto'' version using pseudo-validation (as in \citet{mak2017polygenic}) since we have not found the pseudo-validation scores to be consistent enough with the predictive performance (Figure \ref{fig:pseudoval}).
 Also note that, as in LDpred2, we run lassosum2 genome-wide using a sparse correlation matrix which assumes that variants further away than 3 cM are not correlated.
@@
 We conduct further investigations to explain the results of figure \ref{fig:simu-misN}.
-First, the results for LDpred2-auto are similar to LDpred2-inf because it always converges to an infinitesimal model ($p = 1$) in these simulations.
+First, the results for LDpred2-auto are similar to LDpred2-inf in these simulations because the internal $p$ estimate of LDpred2-auto always continuously increases up to $1$, which makes LDpred2-auto behave as an infinitesimal model ($p = 1$).
 To overcome this limitation, we introduce two new parameters to make LDpred2-auto more robust, and refer to this as ``LDpred2-auto-rob'' here (Methods section \ref{sec:robust}).
@@
 
-\citet{marchini2010genotype} showed that the IMPUTE INFO measure is highly concordant with the MACH measure $\hat{r}_j^2 = \frac{\text{var}(G_j)}{2 \hat{\theta}_j (1 - \hat{\theta}_j)}$, where $\hat{\theta}_j$ is the estimated allele frequency of $G_j$, the genotypes for variant $j$.
+\citet{marchini2010genotype} showed that the IMPUTE INFO measure is highly concordant with the MACH measure $\frac{\text{var}(G_j)}{2 \hat{\theta}_j (1 - \hat{\theta}_j)}$, where $\hat{\theta}_j$ is the estimated allele frequency of $G_j$, the genotypes for variant $j$.
 Therefore, when using the expected genotypes from imputation (allele dosages), their standard deviations are often lower than the expected value under Hardy--Weinberg equilibrium, because $\text{INFO}_j \approx \frac{\text{var}(G_j)}{2 \hat{\theta}_j (1 - \hat{\theta}_j)}$.
@@
 As a second option, instead of using dosages to compute the GWAS summary statistics, it has been argued that using multiple imputation (MI) would be more appropriate \cite[]{Palmer2016}.
+Here MI consists in forming multiple (e.g.\ 20) datasets of hard-called genotypes by sampling them according to the imputation probabilities stored in the BGEN files, performing a GWAS on each of them, and pooling the GWAS estimates.
 In simulations, we show that $\hat{\gamma}_j^{\text{MI}} \approx \hat{\gamma}_j^{\text{imp}} \cdot \text{INFO}_j$ and $Z_j^{\text{MI}} \approx Z_j^{\text{imp}} \cdot \text{INFO}_j$, where $Z = \hat{\gamma} / \text{se}(\hat{\gamma})$ (Figure \ref{fig:compare-mi}).
@@
 Here we design simulations to understand the impact of using a mismatched LD reference panel, e.g.\ that comes from a different population compared to the one used to compute the GWAS summary statistics.
-We use the same simulation setup as before (Methods).
-In addition, we design an alternative reference panel based on 10,000 individuals from South Europe by using the ``Italy'' center defined in \citet{prive2021high}.
+We use the same simulation setup as before (Methods), i.e.\ using individuals of Northwestern European ancestry from the UK Biobank for training (GWAS), validation (tuning of hyper-parameters), and testing (the final models).
+In addition to the LD reference panel of Northwestern European ancestry, we design an alternative reference panel based on 10,000 individuals from South Europe by using the ``Italy'' center defined in \citet{prive2021high}.
 When using this alternative LD reference panel instead of a well-matched one as in the previous sections, squared correlations between the polygenic scores and the simulated phenotypes drop from 0.174 to 0.167 for lassosum2, from 0.168 to 0.147 for LDpred2(-grid), from 0.174 to 0.169 for LDpred2-low-h2, from 0.143 to 0.139 for LDpred2-auto, from 0.162 to 0.146 for LDpred2-auto-rob, and from 0.144 to 0.140 for LDpred2-inf (Figure \ref{fig:simu-altpop}, averaging over 10 simulations).
@@
 Misspecifying the GWAS sample size can lead to serious issues such as misestimating the SNP heritability \cite[]{grotzinger2021pervasive}.
-Fortunately, an overestimated sample size can be detected from the QC plot we propose, where the slope is then less than 1 for case-control studies using logistic regression; otherwise the standard deviation of the phenotype is also needed (Equation \eqref{eq:approx-sd-lin}), but can be estimated \cite[]{prive2020ldpred2}.
+Fortunately, an overestimated sample size can be detected from the QC plot we propose, where the slope is then less than 1 for case-control studies using logistic regression.
 Here we have used this strategy to estimate reduced effective sample sizes in FinnGen GWAS summary statistics.
@@
 It would also be useful to investigate the impact of the QC, well-matched LD reference panels and other adjustments we propose here on other (non-PGS) methods, for which consistency and robustness are likely to be very important as well \cite[]{chen2021improved}.
+For example, we are interested in assessing the impact of these misspecifications on the inference of disease architecture parameters in some future work.
 

### 2022-06-28 paper/paper-misspec-supp.tex
start updating text

@@
 	\caption{Multiple results (squared correlation $r^2$ between polygenic score and phenotype) over a grid of parameters ($\lambda$ and $\delta$) for lassosum2 in one of the simulations with misspecified GWAS sample sizes ($n_j$).
-The different colors indicate whether we use the maximum of $n_j$'s (``maxN'') or the imputed $n_j$'s (``imputeN''), and the normal LD matrix (``normal'') or the one with independent LD blocks (``with\_indep\_ld'').
+The different colors indicate whether we use the maximum of $n_j$'s (``maxN'') or the imputed $n_j$'s (``imputeN''), and the normal LD matrix (``normal'') or the one with independent LD blocks (``with LD blocks'').
 Missing points represent models that were detected as divergent.}
@@
 	\caption{Multiple results (squared correlation $r^2$ between polygenic score and phenotype) over a grid of parameters (the SNP heritability $h^2$ and the proportion of causal variants $p$) for LDpred2(-grid) in one of the simulations with misspecified GWAS sample sizes ($n_j$).
-The different colors indicate whether we use the maximum of $n_j$'s (``maxN'') or the imputed $n_j$'s (``imputeN''), and the normal LD matrix (``normal'') or the one with independent LD blocks (``with\_indep\_ld'').
+The different colors indicate whether we use the maximum of $n_j$'s (``maxN'') or the imputed $n_j$'s (``imputeN''), and the normal LD matrix (``normal'') or the one with independent LD blocks (``with LD blocks'').
 Note that the small $h^2$ values (0.002, 0.02, 0.06) were added and are part of the method we call ``LDpred2-low-h2'' in the main text.
@@
 	\label{fig:ldpred2-misN}
+\end{figure}
+
+\begin{figure}[!h]
+	\centerline{\includegraphics[width=0.95\textwidth]{simu-misN-smallh2}}
+	\caption{Results for the simulations with sample size misspecification and a heritability of 4\%, averaged over 10 simulations for each scenario.
+Reported 95\% confidence intervals are computed from 10,000 non-parametric bootstrap replicates of the mean.
+The GWAS sample size is either ``true'' when providing the true per-variant sample size, ``max'' when providing instead the maximum sample size as a unique value to be used for all variants, ``imputed'', or ``any'' when the method does not use this information (the case for C+T).
+Red bars correspond to using the LD with independent blocks, which is a requirement in lassosum and PRS-CS.}
+	\label{fig:simu-misN-smallh2}
+\end{figure}
+
+\begin{figure}[!h]
+	\centerline{\includegraphics[width=\textwidth]{simu-misN-shrunk}}
+	\caption{Results for the simulations with sample size misspecification and a heritability of 20\% (bottom panels) or 4\% (top panels), averaged over 10 simulations for each scenario.
+Reported 95\% confidence intervals are computed from 10,000 non-parametric bootstrap replicates of the mean.
+The GWAS sample size is either ``true'' when providing the true per-variant sample size, ``max'' when providing instead the maximum sample size as a unique value to be used for all variants, or ``imputed''.
+Red bars correspond to using the LD with independent blocks.
+Bars filled in blue correspond to using the shrunk LD matrix computed from GCTB \cite[]{lloyd2019improved}, otherwise the windowed LD matrix from LDpred2 is used.}
+	\label{fig:simu-misN-shrunk}
 \end{figure}

### 2022-06-28 paper/paper-misspec.tex
start updating text

--- 2022-04-08 paper/paper-misspec.tex
@@
 There is also heterogeneity in the information made available in these summary statistics; for example, per-variant imputation INFO scores, sample sizes and allele frequencies are often missing.
-Moreover, many methods based on summary statistics use Bayesian models and iterative algorithms, which can be sensitive to model misspecifications \cite[]{WALKER20131621,miller2018robust}.
+Moreover, many methods based on summary statistics use Bayesian models and iterative algorithms, which can be particularly sensitive to model misspecifications \cite[]{WALKER20131621,miller2018robust}.
 
@@
 		\hline
-		Differences in per-variant sample sizes & Possible violation of model assumptions resulting in poor model convergence & Imputation using equations \eqref{eq:imputeN} and \eqref{eq:imputeNeff}, and filtering$^{(2)}$   \\
+		Differences in per-variant sample sizes & Possible violation of model assumptions (e.g.\ equation \eqref{eq:model}) resulting in poor predictive performance & Imputation using equations \eqref{eq:imputeN} and \eqref{eq:imputeNeff}, and filtering$^{(2)}$   \\
 		\hline
@@
 Second, we investigate possible adjustments to apply to the input parameters of PGS methods using summary statistics, namely the reference LD (linkage disequilibrium) matrix, the GWAS effect sizes, their standard errors and corresponding sample sizes.
-For example, we show that GWAS effect sizes computed from imputed dosages are larger in magnitude compared to if computed from true genotypes.
+For example, we show that GWAS effect sizes computed from imputed dosages are larger in magnitude than when computed from true genotypes.
 Third, we introduce two new optional parameters in LDpred2-auto to make it more robust to these types of misspecification.
-We focus our investigations on LDpred2 and lassosum for two reasons.
+Note that, in this paper, we use ``robust'' to mean ``similar predictive performance''. 
+We also reimplement and use a new version of lassosum, called lassosum2, in which we better handle when per-variant sample sizes are different.
+We focus our investigations on LDpred2 and lassosum2 for two reasons.
 First, multiple studies have shown that LDpred2 and lassosum rank among the best methods for single-trait polygenic prediction \cite[]{mak2017polygenic,prive2020ldpred2,pain2021evaluation,kulm2021systematic}.
-Second, we reimplement and use a new version of lassosum, called lassosum2, that uses the exact same input parameters as LDpred2, which makes it easy for us to test the different QCs and adjustments presented here.
+Second, lassosum2 now uses the exact same input parameters as LDpred2, which makes it easy for us to test the different QCs and adjustments presented here.
+We additionally investigate PRS-CS and SBayesR, two competitive PGS methods \cite[]{ge2019polygenic,lloyd2019improved}, for which we develop some code to convert between the different formats required for the LD matrices and input GWAS summary statistics.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Materials and Methods}
+
+\subsection{Data for simulations}
+
+We use the UK Biobank imputed (BGEN) data \cite[]{bycroft2018uk}. 
+We restrict individuals to the ones used for computing the principal components (PCs) in the UK Biobank (Field 22020).
+These individuals are unrelated and have passed some quality control including removing samples with a missing rate on autosomes larger than 0.02, having a mismatch between inferred sex and self-reported sex, and outliers based on heterozygosity (more details can be found in section S3 of \cite{bycroft2018uk}).
+To get a set of genetically homogeneous individuals, we compute a robust Mahalanobis distance based on the first 16 PCs and further restrict individuals to those within a log-distance of 5 \cite[]{prive2020efficient}.
+This results in 362,307 individuals of Northwestern European ancestry.
+We randomly sample 300,000 individuals to form a training set (e.g.\ to run the GWAS), 10,000 individuals to form a validation set (to tune hyper-parameters), and use the remaining 52,307 individuals to form a test set (to evaluate final predictive models).
+
+Among genetic variants on chromosome 22 and with a minor allele frequency larger than 0.01 and an imputation INFO score larger than 0.4 (as reported by the UK Biobank), we sample 40,000 of them according to the inverse of the INFO score density so that they have varying levels of imputation accuracy (Figure \ref{fig:hist-info}).
+We read the UK Biobank data into two different datasets using function \texttt{snp\_readBGEN} from R package bigsnpr \cite[]{prive2017efficient}, one by reading the BGEN data as hard-called genotypes at random according to the imputation probabilities, and another one by reading it as dosages (i.e.\ expected values according to the imputation probabilities).
+The first dataset is used as what could be the real genotype calls and the second dataset as what would be its imputed version; this design technique was used in \cite{prive2019making}.
+
+\subsection{Data for real analyses}
+
+We also use the UK Biobank data for validation/testing in real data analyses, and use the same individuals of Northwestern European ancestry as described in the previous section.
+We sample 10,000 individuals to form a validation set and use the remaining 352,307 individuals as test set.
+We restrict to the 1,054,315 HapMap3 variants used in the LD reference provided in \cite{prive2020ldpred2}.
+
+To define phenotypes in the UK Biobank, we first map ICD10 and ICD9 codes (UKBB fields 40001, 40002, 40006, 40013, 41202, 41270 and 41271) to phecodes using R package PheWAS \cite[]{carroll2014r,wu2019mapping}.
+We also use some continuous phenotypes, namely vitamin D (Data-Field 30890), height (50), BMI (21001), systolic blood pressure (4080), and HDL cholesterol (30760).
+
+To derive polygenic scores, we use published GWAS summary statistics listed in table \ref{tab:sumstats}.
+We also use GWAS summary statistics for five disease endpoints from FinnGen\cite{kurki2022finngen} (release 6) and for four continuous outcomes from Biobank Japan \cite[]{sakaue2021cross}.
+
+\begin{table}[h]
+	\centering
+	\begin{tabular}{|l||c|c|c||c|c|}
+		\hline
+		Trait & GWAS citation & \begin{tabular}{@{}c@{}}Effective GWAS\\sample size\end{tabular} & \# GWAS variants & \begin{tabular}{@{}c@{}}\# matched variants\\with INFO > 0.4\end{tabular} & Mean INFO \\
+		\hline
+		Breast cancer (BRCA) [iCOGS] & \cite{michailidou2017association} & ~~87,037 & 11,792,542 & 1,051,242 & 0.841 \\
+		Breast cancer (BRCA) [OncoArray] & \cite{michailidou2017association} & 104,442 & 11,792,542 & 1,054,233 & 0.968 \\
+		Type 1 diabetes (T1D) [Affymetrix] & \cite{censin2017childhood} & ~~~~~5516  & ~~8,996,866 & ~~~934,712 & 0.885 \\	
+		Type 1 diabetes (T1D) [Illumina] & \cite{censin2017childhood} & ~~~~~7982  & ~~8,996,866 & ~~~949,334 & 0.942 \\
+		Prostate cancer (PRCA) & \cite{schumacher2018association} & 135,316 & 20,370,946 & ~~~818,400 & 0.969 \\
+		Depression (MDD) [without UKBB] & \cite{wray2018genome} & 110,464 & ~~9,874,289 & 1,049,455 & 0.968 \\
+		Coronary artery disease (CAD) & \cite{nikpay2015comprehensive} & 129,014 & ~~9,455,778 & 1,052,200 & 0.982 \\
+		Vitamin D & \cite{jiang2018genome} & ~~79,366 & ~~2,579,296 & 1,016,935 & NA \\
+		\hline
+	\end{tabular}
+	\caption{Summary of external GWAS summary statistics used.
+PRCA summary statistics have many variants with a missing INFO score, which we discard.
+We also restrict to variants with an INFO score larger than 0.4.
+Note that vitamin D summary statistics do not report INFO scores.
+\label{tab:sumstats}}
+\end{table}
+
+\subsection{Model for summary statistics}
+
+In this paper, we extensively use the following formula
+\begin{equation}\label{eq:approx-sd-lin}
+S_j = \text{sd}(G_j) \approx \dfrac{\text{sd}(y)}{\sqrt{n_j ~ \text{se}(\hat{\gamma}_j)^2 + \hat{\gamma}_j^2}} ~,
+\end{equation}
+where $\hat{\gamma}_j$ is the marginal GWAS effect size of variant $j$, $n_j$ is the GWAS sample size associated with variant $j$, $y$ is the vector of phenotypes and ${G}_j$ is the vector of genotypes for variant $j$.
+This formula is used in LDpred2 \cite[]{prive2020ldpred2,prive2021high}. Note that, for a binary trait for which logistic regression is used, we have instead
+\begin{equation}\label{eq:approx-sd-log}
+\text{sd}(G_j) \approx \dfrac{2}{\sqrt{n_j^\text{eff} ~ \text{se}(\hat{\gamma}_j)^2 + \hat{\gamma}_j^2}} ~,
+\end{equation}
+where $n_j^\text{eff} = \frac{4}{1 / n_j^\text{cases} + 1 / n_j^\text{controls}}$.
+
+It is assumed that the marginal (GWAS) effects follow
+\begin{align}
+\begin{split}\label{eq:model}
+ \hat{\gamma} | S,R,\gamma & \sim \mathcal{N}(S^{-1} R S \gamma, S^{-1} R S^{-1}) ~,\\
+ S \hat{\gamma} | S,R,\gamma & \sim \mathcal{N}(R S \gamma, R) ~.
+\end{split}
+\end{align}
+where $R$ is the correlation matrix of genotypes \cite[]{zhu2017bayesian}. 
+Then PGS methods aim at inferring $\gamma$, the true causal (per-allele) effects, from $S$, $R$, and $\hat{\gamma}$.
+However, in practice, we only have estimates for $S$ and $R$, which causes some model misspecifications.
+
+\subsection{GWAS sample size imputation}\label{sec:imputeN}
+
+We can impute $n_j$ from equation \eqref{eq:approx-sd-lin} using
+\begin{equation}\label{eq:imputeN}
+n_j \approx \dfrac{\text{var}(y) / \text{var}(G_j) - \hat{\gamma}_j^2}{\text{se}(\hat{\gamma}_j)^2} ~,
+\end{equation}
+and impute $n_j^\text{eff}$ from equation \eqref{eq:approx-sd-log} using
+\begin{equation}\label{eq:imputeNeff}
+n_j^\text{eff} \approx \dfrac{4 / \text{var}(G_j) - \hat{\gamma}_j^2}{\text{se}(\hat{\gamma}_j)^2} ~.
+\end{equation}
+In practice, we also bound these estimates to be between $0.5 \cdot N$ and $1.1 \cdot N$, where $N$ is the total (effective) sample size.
+Also note that the estimate of $\text{var}(G_j)$ has to account for the imputation accuracy, i.e.\ use $2 \cdot f_j \cdot (1 - f_j) \cdot \text{INFO}_j$ (Section \ref{sec:imputation}).
+In equations \eqref{eq:approx-sd-lin} and \eqref{eq:imputeN}, we estimate $\text{var}(y)$ by the first percentile (robust minimum) of $0.5 \left(N ~ \text{se}(\hat{\gamma}_j)^2 + \hat{\gamma}_j^2\right)$.
+
+\subsection{New implementation of lassosum in bigsnpr}\label{lassosum2}
+
+Instead of using a regularized version of the correlation matrix $R$ parameterized by $s$, $R_s = (1 - s) R + s I$ (where $0 < s \leq 1$), we use $R_{\delta} = R + \delta I$ (where $\delta > 0$), which makes it clearer that lassosum is also using L2-regularization (therefore elastic-net).
+Then, from \cite{mak2017polygenic}, the solution from lassosum can be obtained by iteratively updating 
+\[
+\beta_j^{(t)} =
+\begin{cases}
+\dfrac{\text{sign}\big(u_j^{(t)}\big) \big( \big|u_j^{(t)}\big| - \lambda \big)}{\widetilde{X}_j^T \widetilde{X}_j + \delta} & \text{if } \big|u_j^{(t)}\big| > \lambda ~, \\
+0 & \text{otherwise.}
+\end{cases}
+\]
+where 
+\[
+u_j^{(t)} = r_j - \widetilde{X}_j^T \big( \widetilde{X} \beta^{(t-1)} - \widetilde{X}_j \beta_j^{(t-1)} \big) ~.
+\]
+Following the notations from \cite{prive2020ldpred2}, denote $\widetilde{X} = \frac{1}{\sqrt{n-1}} C_n G S^{-1}$, where $G$ is the genotype matrix, $C_n$ is the centering matrix and $S$ is the diagonal matrix of standard deviations of the columns of $G$.
+Then $\widetilde{X}_j^T \widetilde{X} = R_{j,.} = R_{.,j}^T$ and $\widetilde{X}_j^T \widetilde{X}_j = 1$.
+Moreover, using the notations from \cite{prive2020ldpred2}, $u_j^{(t)} = \widehat{\beta}_j - R_{.,j}^T \beta^{(t-1)} + \beta_j^{(t-1)}$,
+where $r_j = \widehat{\beta}_j =  \dfrac{\widehat{\gamma}_j}{\sqrt{n_j ~ \text{se}(\widehat{\gamma}_j)^2 + \widehat{\gamma}_j^2}}$ and $\widehat{\gamma}_j$ is the GWAS effect of variant $j$ and $n$ is the GWAS sample size \cite[]{mak2017polygenic,prive2021high}.
+Then computing $R_{.,j}^T \beta^{(t-1)}$ is the most time-consuming part of each iteration.
+To make this faster, instead of computing $R_{.,j}^T \beta^{(t-1)}$ at each iteration ($j$ and $t$), we can start with an initial vector of 0s only (for all $j$) since $\beta^{(0)} \equiv 0$, then update this vector when $\beta_j^{(t)} \neq \beta_j^{(t-1)}$ only.
+Note that only positions $k$ for which $R_{k,j} \neq 0$ must be updated in this vector $R_{.,j}^T \beta^{(t-1)}$.
+Since bigsnpr v1.10.4, we now also use this updating strategy in LDpred2 (-grid and -auto), which makes it much faster, especially when the polygenicity is small.
+
+In this new implementation of the lassosum model, which we call lassosum2, the input parameters are the correlation matrix $R$, the GWAS summary statistics ($\widehat{\gamma}_j$, $\text{se}(\widehat{\gamma}_j)$ and $n_j$), and the two hyper-parameters $\lambda$ and $\delta$.
+Therefore, except for the two hyper-parameters, lassosum2 uses the exact same input parameters as LDpred2 \cite[]{prive2020ldpred2}.
+We try $\delta \in \{0.001, 0.01, 0.1, 1\}$ by default in lassosum2, instead of $s \in \{0.2, 0.5, 0.8, 1.0\}$ in lassosum.
+For $\lambda$, the default in lassosum uses a sequence of 20 values equally spaced on a log scale between 0.1 and 0.001.
+By default in lassosum2, we use a similar sequence of 30 values, but between $\lambda_0$ and $\lambda_0 / 100$ instead, where $\lambda_0 = \max_j \big|\widehat{\beta}_j\big|$ is the minimum $\lambda$ for which no variable enters the model because the L1-regularization is too strong.
+To make lassosum2 more robust to different per-variant sample sizes $n_j$, we also introduce per-variant penalty factors $\sqrt{\max_k(n_k) / n_j}$, by which we internally multiply $\lambda$ and $\delta$ to penalize variants with smaller GWAS sample sizes more.
+Note that we do not provide an ``auto'' version using pseudo-validation (as in lassosum \cite[]{mak2017polygenic}) since we have not found the pseudo-validation scores to be consistent enough with the predictive performance (Figure \ref{fig:pseudoval}).
+Also note that, as in LDpred2, we run lassosum2 genome-wide using a sparse correlation matrix which assumes that variants further away than 3 cM are not correlated.
+Contrary to for lassosum, we do not require splitting the genome into independent LD blocks anymore, yet we recommend to do so for robustness in LDpred2, and for extra speed gain (cf.\ Methods section \ref{sec:LD}).
+
+\subsection{LDpred2-low-h2 and LDpred2-auto-rob}\label{sec:robust}
+
+Here we introduce the small changes made to LDpred2 (-grid and -auto) in order to make them more robust.
+First, LDpred2-low-h2 simply consists in running LDpred2-grid while testing $h^2$ within $\{0.3, 0.7, 1, 1.4\} \cdot h^2_\text{LDSC}$ (note the added $0.3$ compared to \cite{prive2020ldpred2}), where $h^2_\text{LDSC}$ is the heritability estimate from LD score regression.
+Indeed, we show in simulations here that using lower values for $h^2$ may provide higher predictive performance in the case of misspecifications (thanks to more shrinkage of the effects).
+In simulations, because of the large misspecifications, we use a larger grid over $\{0.01, 0.1, 0.3, 0.7, 1, 1.4\} \cdot h^2_\text{LDSC}$.
+
+For LDpred2-auto, we introduce two new parameters.
+The first one, \texttt{shrink\_corr}, allows for shrinking off-diagonal elements of the correlation matrix.
+This is similar to parameter `s' in lassosum (Methods section \ref{lassosum2}), and acts as a means of regularization.
+We use a value of 0.9 in simulations and 0.95 in real data when running ``LDpred2-auto-rob'' here, and the default value of 1 (with no effect) when running ``LDpred2-auto''.
+The second new parameter, \texttt{allow\_jump\_sign}, controls whether variant effect sizes can change sign over two consecutive iterations of the Gibbs sampler.
+When setting this parameter to false (in the method we refer to as ``LDpred2-auto-rob'' here), this forces the effects to go through 0 before changing sign.
+This is useful to prevent instability (oscillation and ultimately divergence) of the Gibbs sampler under large misspecifications, and is also useful for accelerating convergence of chains with a large initial value for $p$ (the proportion of causal variants).
+
+\subsection{New LD reference}\label{sec:LD}
+
+[TODO: ADD FIG EX SPLIT CHR22]
+
+We form nearly independent LD blocks using the optimal algorithm developed in \cite{prive2021optimal}.
+Note that a new parameter \texttt{max\_r2} has been added to this method since, which controls the maximum single squared correlation allowed outside blocks, which offers an extra guarantee that the splitting is good and makes the function much faster by discarding lots of possible splits.
+For different numbers of blocks and maximum number of variants in each block, we use the split that minimizes $\big(C2 \cdot \sqrt{5 + C1}\big)$, where $C1$ is the sum of squared correlations outside the blocks, and $C2$ is the sum of squared sizes of the blocks.
+Having a correlation matrix with independent blocks prevents small errors in the algorithm (e.g.\ the Gibbs sampler in LDpred2) from propagating to too many variants.
+It also makes running LDpred2 (and lassosum2) faster, taking e.g.\ 60\% of the initial time (since only 60\% of the initial non-zero values of the correlation matrix are kept).
+Note that we also compute the correlations within the same blocks and from the same data for PRS-CS.
+
+We have also developed a new ``compact'' format for the SFBMs (sparse matrices on disk).
+Instead of using something similar to the standard ``compressed sparse column'' format which stores all $\{i,~x(i, j)\}$ for a given column $j$, we only store the first index $i_0$ and all the contiguous values $\{x(i_0, j),~x(i_0 + 1, j),~\dots\}$ up to the last non-zero value for this column $j$.
+This makes this format about twice as efficient for both LDpred2 and lassosum2 (in terms of both memory and speed).
+Therefore, using both this new format and the LD blocks should divide computation times by 3 (without counting the faster updating strategy of residuals).
+
+\subsection{Alternative LD reference for FinnGen and Biobank Japan}\label{sec:altLD}
+
+To be used with GWAS summary statistics from FinnGen, we investigate three different LD reference panels.
+To define homogeneous ancestry groups in the UK Biobank, we include all individuals within a specific distance to a population center in the PCA space \cite[]{prive2021high,prive2021using}.
+We use either the 503 European (including 99 Finnish) individuals from the 1000 Genomes (1000G) data \cite[]{10002015global}, 404 Finnish-like UKBB + the 99 Finnish 1000G individuals (also 503 in total), or the 10,000 UK individuals from the validation set we use in this paper.
+
+To be used with GWAS summary statistics from Biobank Japan, we also investigate three different LD reference panels.
+We use either the 504 East Asian (including 104 Japanese) individuals from the 1000G, 400 Japanese-like UKBB + the 104 Japanese 1000G individuals (also 504 in total), or 2041 East Asian UKBB individuals (including the 400 previous Japanese individuals).
+For the small LD references (based on \textasciitilde 500 individuals), we restrict to variants with a MAF > 0.02.
+For the one based on 2041 individuals, we restrict to MAF > 0.01.
+We also construct versions of these LD references with independent LD blocks, as described in the previous section.
 
@@
 Using 40,000 variants from chromosome 22 (Methods), we simulate quantitative phenotypes with a heritability of 20\% and 2000 causal variants.
-We then divide the 40,000 variants into three groups: for half of the variants, we use 100\% of 300,000 individuals for GWAS, but use only 80\% for one quarter of the variants and 60\% for the remaining quarter.
-We then run C+T, lassosum, lassosum2 (Methods section \ref{lassosum2}), LDpred2-inf, LDpred2(-grid), and LDpred2-auto \cite[]{prive2019making,mak2017polygenic,prive2020ldpred2} by using either the true per-variant GWAS sample sizes, the total sample size for all variants, or per-variant sample sizes imputed using equation \eqref{eq:imputeN}.
-Note that we initially included PRS-CS and SBayesR in our comparison \cite[]{ge2019polygenic,lloyd2019improved}.
-However, results for SBayesR always diverged (stopping with an error), and the overlap with the LD reference provided for PRS-CS was too small.
-When providing true per-variant GWAS sample sizes, squared correlations between the polygenic scores and the simulated phenotypes are of 0.123 for C+T, 0.161 for lassosum, 0.169 for lassosum2, 0.159 for LDpred2(-grid), 0.140 for LDpred2-auto, and 0.141 for LDpred2-inf (Figure \ref{fig:simu-misN}, averaging over 10 simulations).
-Results when using imputed sample sizes are quite similar to when using the true ones.
-Note that C+T does not use this sample size information.
-When using the total GWAS sample size instead of the per-variant sample sizes, predictive performance slightly decreases to 0.157 for lassosum and to 0.163 for lassosum2, but dramatically decreases for LDpred2 with new values of 0.134 for LDpred2-grid, 0.119 for LDpred2-auto, and 0.123 for LDpred2-inf (Figure \ref{fig:simu-misN}).
+We then randomly divide the 40,000 variants into three groups: for half of the variants, we use 100\% of 300,000 individuals for GWAS, but use only 80\% for one quarter of the variants and 60\% for the remaining quarter.
+We then run C+T, lassosum, lassosum2 (Methods section \ref{lassosum2}), LDpred2-inf, LDpred2(-grid), LDpred2-auto, SBayesR, and PRS-CS-auto \cite[]{prive2019making,mak2017polygenic,prive2020ldpred2,lloyd2019improved,ge2019polygenic} by using either the true per-variant GWAS sample sizes, the total sample size for all variants, or per-variant sample sizes imputed using equation \eqref{eq:imputeN}.
+When providing true per-variant GWAS sample sizes, squared correlations between the polygenic scores and the simulated phenotypes are of 0.123 for C+T, 0.160 for lassosum, 0.169 for lassosum2, 0.159 for LDpred2(-grid), 0.140 for LDpred2-auto, 0.141 for LDpred2-inf, 0.138 for SBayesR, and 0.159 for PRS-CS-auto (Figure \ref{fig:simu-misN}, averaging over 10 simulations).
+Results when using imputed sample sizes are very similar to when using the true ones.
+Note that C+T does not use this sample size information, and that PRS-CS can only be provided with a single value (we use the maximum one).
+When using the total GWAS sample size instead of the per-variant sample sizes, predictive performance slightly decreases to 0.158 for lassosum and to 0.164 for lassosum2, but dramatically decreases for LDpred2 with new values of 0.134 for LDpred2-grid, 0.122 for LDpred2-auto, 0.125 for LDpred2-inf (Figure \ref{fig:simu-misN}).
 This extreme simulation scenario shows that LDpred2 can be sensitive to the misspecification of per-variant GWAS sample sizes, whereas lassosum (and lassosum2) seems little affected by this.
+With 0.141 for SBayesR, predictive performance is actually larger than when using the true per-variant sample sizes.
 
 \begin{figure}[!h]
-	\centerline{\includegraphics[width=\textwidth]{simu-misN}}
+	\centerline{\includegraphics[width=0.95\textwidth]{simu-misN}}
 	\caption{Results for the simulations with sample size misspecification, averaged over 10 simulations for each scenario.
@@
 The GWAS sample size is either ``true'' when providing the true per-variant sample size, ``max'' when providing instead the maximum sample size as a unique value to be used for all variants, ``imputed'' (cf.\ equation \eqref{eq:imputeN}), or ``any'' when the method does not use this information (the case for C+T).
-Red bars correspond to using the LD with independent blocks (Methods section \ref{sec:LD}); this strategy cannot be tested straightforwardly for C+T and lassosum.}
+Red bars correspond to using the LD with independent blocks (Methods section \ref{sec:LD}), which is a requirement in lassosum and PRS-CS.}
 	\label{fig:simu-misN}
@@
 We conduct further investigations to explain the results of figure \ref{fig:simu-misN}.
-First, the results for LDpred2-auto are similar to LDpred2-inf because it always converges to an infinitesimal model ($p = 1$) in these simulations.
-To overcome this limitation, we introduce two new parameters to make LDpred2-auto more robust, and refer to this as ``LDpred2-auto-rob'' here (Methods section \ref{sec:robust}).
+First, the results for LDpred2-auto and SBayesR are similar to LDpred2-inf in these simulations because the internal $p$ estimate of LDpred2-auto (and similarly the proportion of null variants in SBayesR) always continuously increases up to $1$, which makes them behave as an infinitesimal model ($p = 1$).
+To overcome this limitation in LDpred2-auto, we introduce two new parameters to make it more robust, and refer to this as ``LDpred2-auto-rob'' here (Methods section \ref{sec:robust}).
 The first parameter prevents $p$ from diverging to $1$, while the second shrinks the off-diagonal elements of the LD matrix (a form of regularization). 
@@
 Here we call ``LDpred2-low-h2'' when running LDpred2(-grid) with a grid of hyper-parameters including these low values for $h^2$.
-Results with LDpred2-low-h2 improves from 0.159 to 0.169 when using true sample sizes and from 0.134 to 0.163 when using the maximum sample size.
+Results with LDpred2-low-h2 improves from 0.159 to 0.169 when using true sample sizes and from 0.134 to 0.164 when using the maximum sample size.
 Finally, we introduce a last change for robustness here: we form independent LD blocks in the LD matrix to prevent small errors in the Gibbs sampler to propagate to too many variants (Methods section \ref{sec:LD}).
@@
 
+As secondary analysis, we rerun this simulation with a smaller heritability of 4\% instead.
+While predictive performance are overall much lower, relative performance are less attenuated when using the maximum sample size, otherwise results are highly consistent as before (Figure \ref{fig:simu-misN-smallh2}).
+We also try using the shrunk LD matrix from GCTB that is used e.g.\ for SBayesR \cite[]{lloyd2019improved}; in these simulations, higher predictive performance are obtained from LDpred2-auto with this shrunk LD matrix, whereas SBayesR benefit from using LDpred2's windowed LD matrix (Figure \ref{fig:simu-misN-shrunk}).
+
 \subsection{Misspecification when using imputed allele dosages}\label{sec:imputation}
 
-\cite{marchini2010genotype} showed that the IMPUTE INFO measure is highly concordant with the MACH measure $\hat{r}_j^2 = \frac{\text{var}(G_j)}{2 \hat{\theta}_j (1 - \hat{\theta}_j)}$, where $\hat{\theta}_j$ is the estimated allele frequency of $G_j$, the genotypes for variant $j$.
+\cite{marchini2010genotype} showed that the IMPUTE INFO measure is highly concordant with the MACH measure $\frac{\text{var}(G_j)}{2 \hat{\theta}_j (1 - \hat{\theta}_j)}$, where $\hat{\theta}_j$ is the estimated allele frequency of $G_j$, the genotypes for variant $j$.
 Therefore, when using the expected genotypes from imputation (allele dosages), their standard deviations are often lower than the expected value under Hardy--Weinberg equilibrium, because $\text{INFO}_j \approx \frac{\text{var}(G_j)}{2 \hat{\theta}_j (1 - \hat{\theta}_j)}$.
@@
 
-To compare these three corrections, we conduct a simulation using 40,000 variants from chromosome 22, where we simulate quantitative phenotypes assuming a heritability of 20\% and 2000 causal variants using the ``true'' dataset (cf.\ Methods section ``Data for simulations'').
+To compare these three corrections, we conduct a simulation using the same 40,000 variants from chromosome 22 as before, where we simulate quantitative phenotypes assuming a heritability of 20\% and 2000 causal variants using the ``true'' dataset (cf.\ Methods section ``Data for simulations'').
 We compute GWAS summary statistics from the dosage dataset and use these summary statistics to run LDpred2 and lassosum2 with either no correction of the summary statistics, or with one of the three corrections described above.
@@
 \begin{figure}[!h]
-	\centerline{\includegraphics[width=\textwidth]{simu-info}}
+	\centerline{\includegraphics[width=0.95\textwidth]{simu-info}}
 	\caption{Results of predictive performance for the simulations using GWAS summary statistics from imputed dosage data, averaged over 10 simulations for each scenario.
@@
 Here we design simulations to understand the impact of using a mismatched LD reference panel, e.g.\ that comes from a different population compared to the one used to compute the GWAS summary statistics.
-We use the same simulation setup as before (Methods).
-In addition, we design an alternative reference panel based on 10,000 individuals from South Europe by using the ``Italy'' center defined in \cite{prive2021high}.
+We use the same simulation setup as before (Methods), i.e.\ using individuals of Northwestern European ancestry from the UK Biobank for training (GWAS), validation (tuning of hyper-parameters), and testing (the final models).
+In addition to the LD reference panel of Northwestern European ancestry, we design an alternative reference panel based on 10,000 individuals from South Europe by using the ``Italy'' center defined in \cite{prive2021high}.
 When using this alternative LD reference panel instead of a well-matched one as in the previous sections, squared correlations between the polygenic scores and the simulated phenotypes drop from 0.174 to 0.167 for lassosum2, from 0.168 to 0.147 for LDpred2(-grid), from 0.174 to 0.169 for LDpred2-low-h2, from 0.143 to 0.139 for LDpred2-auto, from 0.162 to 0.146 for LDpred2-auto-rob, and from 0.144 to 0.140 for LDpred2-inf (Figure \ref{fig:simu-altpop}, averaging over 10 simulations).
@@
 \begin{figure}[!h]
-	\centerline{\includegraphics[width=0.67\textwidth]{res-cad}}
+	\centerline{\includegraphics[width=0.85\textwidth]{res-cad}}
 	\caption{Variance explained of CAD in the UK Biobank by PGS derived from external summary statistics.
@@
 It would also be useful to investigate the impact of the QC, well-matched LD reference panels and other adjustments we propose here on other (non-PGS) methods, for which consistency and robustness are likely to be very important as well \cite[]{chen2021improved}.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Materials and Methods}
-
-\subsection{Data for simulations}
-
-We use the UK Biobank imputed (BGEN) data \cite[]{bycroft2018uk}. 
-We restrict individuals to the ones used for computing the principal components (PCs) in the UK Biobank (Field 22020).
-These individuals are unrelated and have passed some quality control including removing samples with a missing rate on autosomes larger than 0.02, having a mismatch between inferred sex and self-reported sex, and outliers based on heterozygosity (more details can be found in section S3 of \cite{bycroft2018uk}).
-To get a set of genetically homogeneous individuals, we compute a robust Mahalanobis distance based on the first 16 PCs and further restrict individuals to those within a log-distance of 5 \cite[]{prive2020efficient}.
-This results in 362,307 individuals.
-We randomly sample 300,000 individuals to form a training set (e.g.\ to run the GWAS), 10,000 individuals to form a validation set (to tune hyper-parameters), and use the remaining 52,307 individuals to form a test set (to evaluate final predictive models).
-
-Among genetic variants on chromosome 22 and with a minor allele frequency larger than 0.01 and an imputation INFO score larger than 0.4 (as reported by the UK Biobank), we sample 40,000 of them according to the inverse of the INFO score density so that they have varying levels of imputation accuracy (Figure \ref{fig:hist-info}).
-We read the UK Biobank data into two different datasets using function \texttt{snp\_readBGEN} from R package bigsnpr \cite[]{prive2017efficient}, one by reading the BGEN data at random according to imputation probabilities, and another one reading it as dosages (i.e.\ expected values according to imputation probabilities).
-The first dataset is used as what could be the real genotype calls and the second dataset as what would be its imputed version; this design technique was used in \cite{prive2019making}.
-
-\subsection{Data for real analyses}
-
-We also use the UK Biobank data for validation/testing in real data analyses, and use the same individuals as described in the previous section.
-We sample 10,000 individuals to form a validation set and use the remaining 352,307 individuals as test set.
-We restrict to the 1,054,315 HapMap3 variants used in the LD reference provided in \cite{prive2020ldpred2}.
-
-To define phenotypes in the UK Biobank, we first map ICD10 and ICD9 codes (UKBB fields 40001, 40002, 40006, 40013, 41202, 41270 and 41271) to phecodes using R package PheWAS \cite[]{carroll2014r,wu2019mapping}.
-We also use some continuous phenotypes, namely vitamin D (Data-Field 30890), height (50), BMI (21001), systolic blood pressure (4080), and HDL cholesterol (30760).
-
-We use published GWAS summary statistics listed in table \ref{tab:sumstats} to derive polygenic scores.
-We also use GWAS summary statistics for five disease endpoints from FinnGen (release 6, \cite{kurki2022finngen}) and for four continuous outcomes from Biobank Japan \cite[]{sakaue2021cross}.
-
-\begin{table}[h]
-	\centering
-	\begin{tabular}{|l||c|c|c||c|c|}
-		\hline
-		Trait & GWAS citation & \begin{tabular}{@{}c@{}}Effective GWAS\\sample size\end{tabular} & \# GWAS variants & \begin{tabular}{@{}c@{}}\# matched variants\\with INFO > 0.4\end{tabular} & Mean INFO \\
-		\hline
-		Breast cancer (BRCA) [iCOGS] & \cite{michailidou2017association} & ~~87,037 & 11,792,542 & 1,051,242 & 0.841 \\
-		Breast cancer (BRCA) [OncoArray] & \cite{michailidou2017association} & 104,442 & 11,792,542 & 1,054,233 & 0.968 \\
-		Type 1 diabetes (T1D) [Affymetrix] & \cite{censin2017childhood} & ~~~~~5516  & ~~8,996,866 & ~~~934,712 & 0.885 \\	
-		Type 1 diabetes (T1D) [Illumina] & \cite{censin2017childhood} & ~~~~~7982  & ~~8,996,866 & ~~~949,334 & 0.942 \\
-		Prostate cancer (PRCA) & \cite{schumacher2018association} & 135,316 & 20,370,946 & ~~~818,400 & 0.969 \\
-		Depression (MDD) [without UKBB] & \cite{wray2018genome} & 110,464 & ~~9,874,289 & 1,049,455 & 0.968 \\
-		Coronary artery disease (CAD) & \cite{nikpay2015comprehensive} & 129,014 & ~~9,455,778 & 1,052,200 & 0.982 \\
-		Vitamin D & \cite{jiang2018genome} & ~~79,366 & ~~2,579,296 & 1,016,935 & NA \\
-		\hline
-	\end{tabular}
-	\caption{Summary of external GWAS summary statistics used.
-PRCA summary statistics have many variants with a missing INFO score, which we discard.
-We also restrict to variants with an INFO score larger than 0.4.
-Vitamin D summary statistics do not report INFO scores.
-\label{tab:sumstats}}
-\end{table}
-
-\subsection{GWAS sample size imputation}
-
-In this paper, we extensively use the following formula
-\begin{equation}\label{eq:approx-sd-lin}
-\text{sd}(G_j) \approx \dfrac{\text{sd}(y)}{\sqrt{n_j ~ \text{se}(\hat{\gamma}_j)^2 + \hat{\gamma}_j^2}} ~,
-\end{equation}
-where $\hat{\gamma}_j$ is the marginal GWAS effect size of variant $j$, $n_j$ is the GWAS sample size associated with variant $j$, $y$ is the vector of phenotypes and ${G}_j$ is the vector of genotypes for variant $j$.
-This formula is used in LDpred2 \cite{prive2020ldpred2,prive2021high}. Note that, for a binary trait for which logistic regression is used, we have instead
-\begin{equation}\label{eq:approx-sd-log}
-\text{sd}(G_j) \approx \dfrac{2}{\sqrt{n_j^\text{eff} ~ \text{se}(\hat{\gamma}_j)^2 + \hat{\gamma}_j^2}} ~,
-\end{equation}
-where $n_j^\text{eff} = \frac{4}{1 / n_j^\text{cases} + 1 / n_j^\text{controls}}$.
-
-We can then impute $n_j$ from equation \eqref{eq:approx-sd-lin} using
-\begin{equation}\label{eq:imputeN}
-n_j \approx \dfrac{\text{var}(y) / \text{var}(G_j) - \hat{\gamma}_j^2}{\text{se}(\hat{\gamma}_j)^2} ~,
-\end{equation}
-and impute $n_j^\text{eff}$ from equation \eqref{eq:approx-sd-log} using
-\begin{equation}\label{eq:imputeNeff}
-n_j^\text{eff} \approx \dfrac{4 / \text{var}(G_j) - \hat{\gamma}_j^2}{\text{se}(\hat{\gamma}_j)^2} ~.
-\end{equation}
-In practice, we also bound this estimate to be between $0.5 \cdot N$ and $1.1 \cdot N$, where $N$ is the total (effective) sample size.
-Also note that the estimate of $\text{var}(G_j)$ has to account for the imputation accuracy, i.e.\ use $2 \cdot f_j \cdot (1 - f_j) \cdot \text{INFO}_j$ (Section \ref{sec:imputation}).
-
-\subsection{New implementation of lassosum in bigsnpr}\label{lassosum2}
-
-Instead of using a regularized version of the correlation matrix $R$ parameterized by $s$, $R_s = (1 - s) R + s I$ (where $0 < s \leq 1$), we use $R_{\delta} = R + \delta I$ (where $\delta > 0$), which makes it clearer that lassosum is also using L2-regularization (therefore elastic-net).
-Then, from \cite{mak2017polygenic}, the solution from lassosum can be obtained by iteratively updating 
-\[
-\beta_j^{(t)} =
-\begin{cases}
-\dfrac{\text{sign}\big(u_j^{(t)}\big) \big( \big|u_j^{(t)}\big| - \lambda \big)}{\widetilde{X}_j^T \widetilde{X}_j + \delta} & \text{if } \big|u_j^{(t)}\big| > \lambda ~, \\
-0 & \text{otherwise.}
-\end{cases}
-\]
-where 
-\[
-u_j^{(t)} = r_j - \widetilde{X}_j^T \big( \widetilde{X} \beta^{(t-1)} - \widetilde{X}_j \beta_j^{(t-1)} \big) ~.
-\]
-Following the notations from \cite{prive2020ldpred2}, denote $\widetilde{X} = \frac{1}{\sqrt{n-1}} C_n G S^{-1}$, where $G$ is the genotype matrix, $C_n$ is the centering matrix and $S$ is the diagonal matrix of standard deviations of the columns of $G$.
-Then $\widetilde{X}_j^T \widetilde{X} = R_{j,.} = R_{.,j}^T$ and $\widetilde{X}_j^T \widetilde{X}_j = 1$.
-Moreover, using the notations from \cite{prive2020ldpred2}, $u_j^{(t)} = \widehat{\beta}_j - R_{.,j}^T \beta^{(t-1)} + \beta_j^{(t-1)}$,
-where $r_j = \widehat{\beta}_j =  \dfrac{\widehat{\gamma}_j}{\sqrt{n_j ~ \text{se}(\widehat{\gamma}_j)^2 + \widehat{\gamma}_j^2}}$ and $\widehat{\gamma}_j$ is the GWAS effect of variant $j$ and $n$ is the GWAS sample size \cite[]{mak2017polygenic,prive2021high}.
-Then computing $R_{.,j}^T \beta^{(t-1)}$ is the most time-consuming part of each iteration.
-To make this faster, instead of computing $R_{.,j}^T \beta^{(t-1)}$ at each iteration ($j$ and $t$), we can start with an initial vector of 0s only (for all $j$) since $\beta^{(0)} \equiv 0$, then update this vector when $\beta_j^{(t)} \neq \beta_j^{(t-1)}$ only.
-Note that only positions $k$ for which $R_{k,j} \neq 0$ must be updated in this vector $R_{.,j}^T \beta^{(t-1)}$.
-
-In this new implementation of the lassosum model, which we call lassosum2, the input parameters are the correlation matrix $R$, the GWAS summary statistics ($\widehat{\gamma}_j$, $\text{se}(\widehat{\gamma}_j)$ and $n_j$), and the two hyper-parameters $\lambda$ and $\delta$.
-Therefore, except for the two hyper-parameters, lassosum2 uses the exact same input parameters as LDpred2 \cite[]{prive2020ldpred2}.
-We try $\delta \in \{0.001, 0.005, 0.02, 0.1, 0.6, 3\}$ by default in lassosum2, instead of $s \in \{0.2, 0.5, 0.8, 1.0\}$ in lassosum.
-For $\lambda$, the default in lassosum uses a sequence of 20 values equally spaced on a log scale between 0.1 and 0.001.
-By default in lassosum2, we use a similar sequence, but between $\lambda_0$ and $\lambda_0 / 100$ instead, where $\lambda_0 = \max_j \big|\widehat{\beta}_j\big|$ is the minimum $\lambda$ for which no variable enters the model because the L1-regularization is too strong.
-Note that we do not provide an ``auto'' version using pseudo-validation (as in \cite{mak2017polygenic}) as we have not found it to be robust enough (Figure \ref{fig:pseudoval}).
-Also note that, as in LDpred2, we run lassosum2 genome-wide using a sparse correlation matrix which assumes that variants further away than 3 cM are not correlated.
-Although we do not require splitting the genome into independent LD blocks anymore (as done in lassosum), we recommend to do so for robustness in LDpred2, and for extra speed gain (cf.\ Methods section \ref{sec:LD}).
-
-\subsection{LDpred2-low-h2 and LDpred2-auto-rob}\label{sec:robust}
-
-Here we introduce the small changes made to LDpred2 (-grid and -auto) in order to make them more robust.
-First, LDpred2-low-h2 simply consists in running LDpred2-grid while testing $h^2$ within $\{0.3, 0.7, 1, 1.4\} \cdot h^2_\text{LDSC}$ (note the added $0.3$ compared to \cite{prive2020ldpred2}), where $h^2_\text{LDSC}$ is the heritability estimate from LD score regression.
-Indeed, we show in simulations here that using lower values for $h^2$ may provide higher predictive performance in the case of misspecifications (thanks to more shrinkage of the effects).
-In simulations, because of the large misspecifications, we use a larger grid over $\{0.01, 0.1, 0.3, 0.7, 1, 1.4\} \cdot h^2_\text{LDSC}$.
-
-For LDpred2-auto, we introduce two new parameters.
-The first one, \texttt{shrink\_corr}, allows for shrinking off-diagonal elements of the correlation matrix.
-This is similar to parameter `s' in lassosum (Methods section \ref{lassosum2}), and acts as a means of regularization.
-We use a value of 0.9 in simulations and 0.95 in real data when running ``LDpred2-auto-rob'' here, and the default value of 1 (with no effect) when running ``LDpred2-auto''.
-The second new parameter, \texttt{allow\_jump\_sign}, controls whether variant effect sizes can change sign over two consecutive iterations of the Gibbs sampler.
-When setting this parameter to false (in the method we refer to as ``LDpred2-auto-rob'' here), this forces the effects to go through 0 before changing sign.
-This is useful to prevent instability (oscillation and ultimately divergence) of the Gibbs sampler under large misspecifications, and is also useful for accelerating convergence of chains with a large initial value for $p$ (the proportion of causal variants).
-
-\subsection{New LD reference}\label{sec:LD}
-
-We form nearly independent LD blocks using the optimal algorithm developed in \cite{prive2021optimal}.
-For different numbers of blocks and maximum number of variants in each block, we use the split with the minimum cost within the ones reducing the original number of non-zero values to less than 60\% (70\% for chromosome 6).
-Having a correlation matrix with independent blocks prevents small errors in the algorithm (e.g.\ the Gibbs sampler in LDpred2) from propagating to too many variants.
-It also makes running LDpred2 (and lassosum2) faster, taking about 60\% of the initial time (since only 60\% of the initial non-zero values of the correlation matrix are kept).
-
-We have also developed a new ``compact'' format for the SFBMs (sparse matrices on disk).
-Instead of using something similar to the standard ``compressed sparse column'' format which stores all $\{i,~x(i, j)\}$ for a given column $j$, we only store the first index $i_0$ and all the contiguous values $\{x(i_0, j),~x(i_0 + 1, j),~\dots\}$ up to the last non-zero value for this column $j$.
-This makes this format about twice as efficient for both LDpred2 and lassosum2 (in terms of both memory and speed).
-Therefore, using both this new format and the LD blocks should divide computation times by 3.
-
-\subsection{Alternative LD reference for FinnGen and Biobank Japan}\label{sec:altLD}
-
-To be used with GWAS summary statistics from FinnGen, we investigate three different LD reference panels.
-To define homogeneous ancestry groups in the UK Biobank, we include all individuals within a specific distance to a population center in the PCA space \cite[]{prive2021high,prive2021using}.
-We use either the 503 European (including 99 Finnish) individuals from the 1000 Genomes (1000G) data \cite[]{10002015global}, 404 Finnish-like UKBB + the 99 Finnish 1000G individuals (also 503 in total), or the 10,000 UK individuals from the validation set we use in this paper.
-
-To be used with GWAS summary statistics from Biobank Japan, we also investigate three different LD reference panels.
-We use either the 504 East Asian (including 104 Japanese) individuals from the 1000G, 400 Japanese-like UKBB + the 104 Japanese 1000G individuals (also 504 in total), or 2041 East Asian UKBB individuals (including the 400 previous Japanese individuals).
-We also construct versions of these LD references with independent LD blocks, as described in the previous section.
+For example, we are interested in assessing the impact of these misspecifications on the inference of disease architecture parameters in some future work.
 
@@
 
-\clearpage
-%\vspace*{3em}
+%\clearpage
+\vspace*{3em}
 
-\bibliographystyle{natbib}
+\bibliographystyle{ajhg}
 \bibliography{refs}

### 2022-06-29 paper/paper-misspec-supp.tex
continue text editing

@@
 
+\begin{figure}[p]
+	\centerline{\includegraphics[width=0.95\textwidth]{cost_split_chr22}}
+	\caption{Results of different block splits of the LD matrix of chromosome 22 used in simulations, for different numbers of blocks and maximum number of variants in each block.
+The black point corresponds to the final split chosen.}
+	\label{fig:cost-split}
+\end{figure}
+
 \FloatBarrier
@@
 
-\begin{figure}[!h]
+\begin{figure}[p]
 	\centerline{\includegraphics[width=0.95\textwidth]{simu-misN-smallh2}}
@@
 
-\begin{figure}[!h]
+\begin{figure}[p]
 	\centerline{\includegraphics[width=\textwidth]{simu-misN-shrunk}}
@@
 \label{fig:compare-maf-info}}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{figure}[p]
+	\centerline{\includegraphics[width=0.95\textwidth]{simu-info-smallh2}}
+	\caption{Results of predictive performance for the simulations (with a heritability of 4\%) using GWAS summary statistics from imputed dosage data, averaged over 10 simulations for each scenario.
+Reported 95\% confidence intervals are computed from 10,000 non-parametric bootstrap replicates of the mean.
+Correction ``sqrt\_info'' corresponds to using $\hat{\gamma}_j^{\text{imp}} \cdot \sqrt{\text{INFO}_j}$ and $\text{se}(\hat{\gamma}_j)^{\text{imp}} \cdot \sqrt{\text{INFO}_j}$.
+	Correction ``info'' corresponds to using $\hat{\gamma}_j^{\text{imp}} \cdot \text{INFO}_j$  and $n_j \cdot \text{INFO}_j$.
+Correction ``in\_between'' corresponds to using $\hat{\gamma}_j^{\text{imp}} \cdot \text{INFO}_j$, $\text{se}(\hat{\gamma}_j)^{\text{imp}} \cdot \sqrt{\text{INFO}_j}$, and $n_j \cdot \text{INFO}_j$.
+Red bars correspond to using the LD with independent blocks.}
+	\label{fig:simu-info-smallh2}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\FloatBarrier
+
+\begin{figure}[p]
+	\begin{subfigure}{0.5\textwidth}
+		\centering
+		\includegraphics[width=.95\linewidth]{af_alt_pop}
+	\end{subfigure}\vspace{1em}
+	\begin{subfigure}{0.5\textwidth}
+		\centering
+		\includegraphics[width=.95\linewidth]{corr_alt_pop}
+	\end{subfigure}
+	\caption{Comparison of allele frequencies and pairwise correlations between the two LD reference panels used in the simulations, one from North-West Europe and the alternative one from South Europe.}
+	\label{fig:compare-altpop}
+\end{figure}
+
+\begin{figure}[p]
+	\centerline{\includegraphics[width=0.95\textwidth]{simu-altpop-smallh2}}
+	\caption{Results for the simulations with summary statistics with LD matrices based on two different populations, using a heritability of 4\% (instead of 20\%).
+One comes from the same ancestry used for computing the GWAS summary statistics (North-West Europe), while the other one comes from South Europe (alternative LD reference).
+Reported 95\% confidence intervals are computed from 10,000 non-parametric bootstrap replicates of the mean.
+Red bars correspond to using the LD with independent blocks (Methods section \ref{sec:LD}), which is a requirement for PRS-CS.}
+	\label{fig:simu-altpop-smallh2}
 \end{figure}
@@
 
+\begin{figure}[p]
+	\centerline{\includegraphics[width=0.95\textwidth]{res-cad-shrinkage}}
+	\caption{Variance explained of CAD in the UK Biobank by PGS derived from external summary statistics.
+These are computed using function \texttt{pcor} of R package bigstatsr where 95\% confidence intervals are obtained through Fisher's Z-transformation; these values are then squared.
+Red bars correspond to using the LD with independent blocks (Methods).
+Parameter \texttt{shrinkage} corresponds to the new parameter of LDpred2-auto.}
+	\label{fig:res_cad_shrinkage}
+\end{figure}
+
 \FloatBarrier

### 2022-06-29 paper/paper-misspec.tex
continue text editing

@@
 
-[TODO: ADD FIG EX SPLIT CHR22]
-
 We form nearly independent LD blocks using the optimal algorithm developed in \cite{prive2021optimal}.
 Note that a new parameter \texttt{max\_r2} has been added to this method since, which controls the maximum single squared correlation allowed outside blocks, which offers an extra guarantee that the splitting is good and makes the function much faster by discarding lots of possible splits.
-For different numbers of blocks and maximum number of variants in each block, we use the split that minimizes $\big(C2 \cdot \sqrt{5 + C1}\big)$, where $C1$ is the sum of squared correlations outside the blocks, and $C2$ is the sum of squared sizes of the blocks.
+For different numbers of blocks and maximum number of variants in each block, we use the split that minimizes $\big(C2 \cdot \sqrt{5 + C1}\big)$, where $C1$ is the sum of squared correlations outside the blocks, and $C2$ is the sum of squared sizes of the blocks (Figure \ref{fig:cost-split}).
 Having a correlation matrix with independent blocks prevents small errors in the algorithm (e.g.\ the Gibbs sampler in LDpred2) from propagating to too many variants.
@@
 The GWAS sample size is either ``true'' when providing the true per-variant sample size, ``max'' when providing instead the maximum sample size as a unique value to be used for all variants, ``imputed'' (cf.\ equation \eqref{eq:imputeN}), or ``any'' when the method does not use this information (the case for C+T).
-Red bars correspond to using the LD with independent blocks (Methods section \ref{sec:LD}), which is a requirement in lassosum and PRS-CS.}
+Red bars correspond to using the LD with independent blocks (Methods section \ref{sec:LD}), which is a requirement for lassosum and PRS-CS.}
 	\label{fig:simu-misN}
@@
 To compare these three corrections, we conduct a simulation using the same 40,000 variants from chromosome 22 as before, where we simulate quantitative phenotypes assuming a heritability of 20\% and 2000 causal variants using the ``true'' dataset (cf.\ Methods section ``Data for simulations'').
-We compute GWAS summary statistics from the dosage dataset and use these summary statistics to run LDpred2 and lassosum2 with either no correction of the summary statistics, or with one of the three corrections described above.
-The LD reference used by LDpred2 and lassosum2 is computed from the validation set using the dataset with the ``true'' genotypes.
+We compute GWAS summary statistics from the dosage dataset and use these summary statistics to run LDpred2, lassosum2 and PRS-CS with either no correction of the summary statistics, or with one of the three corrections described above.
+The LD references are computed from the validation set using the dataset with the ``true'' genotypes (i.e.\ same as before).
 For lassosum2 and LDpred2(-grid), which tune parameters using the validation set, correcting for imputation quality slightly improves predictive performance in these simulations (Figure \ref{fig:simu-info}).
@@
 Moreover, new additions for robustness introduced before, namely LDpred2-low-h2, LDpred2-auto-rob, and forming independent blocks in the LD matrix also improve predictive performance for all corrections (Figure \ref{fig:simu-info}).
+However, these corrections do not prove beneficial for PRS-CS-auto (Figure \ref{fig:simu-info}).
 
-In the real data applications hereinafter, we choose to use the first correction, ``sqrt\_info'', which is simple because it is equivalent to post-processing PGS effects by multiplying them by $\sqrt{\text{INFO}}$.
+When performing similar simulations with a heritability of 4\% (instead of 20\%), correction ``sqrt\_info'' still provides slightly higher predictive performance for all methods (compared to no correction), yet the other two corrections sometimes provide lower predictive performance (Figure \ref{fig:simu-info-smallh2}).
+In the real data applications hereinafter, we therefore choose to use the first correction, ``sqrt\_info'', which seems to more consistently provide better results, and is simple because it is somewhat equivalent to post-processing PGS effects by multiplying them by $\sqrt{\text{INFO}}$.
 
@@
 Correction ``in\_between'' corresponds to using $\hat{\gamma}_j^{\text{imp}} \cdot \text{INFO}_j$, $\text{se}(\hat{\gamma}_j)^{\text{imp}} \cdot \sqrt{\text{INFO}_j}$, and $n_j \cdot \text{INFO}_j$.
-Red bars correspond to using the LD with independent blocks (Methods section \ref{sec:LD}).}
+Red bars correspond to using the LD with independent blocks (Methods section \ref{sec:LD}), which is a requirement for PRS-CS.}
 	\label{fig:simu-info}
@@
 In addition to the LD reference panel of Northwestern European ancestry, we design an alternative reference panel based on 10,000 individuals from South Europe by using the ``Italy'' center defined in \cite{prive2021high}.
-When using this alternative LD reference panel instead of a well-matched one as in the previous sections, squared correlations between the polygenic scores and the simulated phenotypes drop from 0.174 to 0.167 for lassosum2, from 0.168 to 0.147 for LDpred2(-grid), from 0.174 to 0.169 for LDpred2-low-h2, from 0.143 to 0.139 for LDpred2-auto, from 0.162 to 0.146 for LDpred2-auto-rob, and from 0.144 to 0.140 for LDpred2-inf (Figure \ref{fig:simu-altpop}, averaging over 10 simulations).
-Forming independent LD blocks in these two LD matrices always helps, but do not improve performance by much when using the alternative LD.
+Allele frequencies and pairwise correlations are quite similar between these two LD reference panels (Figure \ref{fig:compare-altpop}).
+When using this alternative LD reference panel instead of a well-matched one as in the previous sections, squared correlations between the polygenic scores and the simulated phenotypes drop from 0.173 to 0.166 for lassosum2, from 0.168 to 0.159 for LDpred2(-grid), from 0.174 to 0.169 for LDpred2-low-h2, from 0.142 to 0.136 for LDpred2-auto, from 0.162 to 0.147 for LDpred2-auto-rob, from 0.143 to 0.138 for LDpred2-inf, and from  0.163 to 0.155 for PRS-CS-auto (Figure \ref{fig:simu-altpop}, averaging over 10 simulations).
+Forming independent LD blocks in these two LD matrices always improve predictive performance, yet again.
+We also compute and try using a shrunk LD matrix (from GCTB) for this alternative LD reference panel; again, using this type of LD matrix seems beneficial for LDpred2-auto.
+Finally, under the same simulation scenario, but with a smaller simulated heritability of 4\% (instead of 20\%), there remains a small drop in predictive performance when using the alternative LD reference (Figure \ref{fig:simu-altpop-smallh2}).
 
 \begin{figure}[!h]
-	\centerline{\includegraphics[width=0.9\textwidth]{simu-altpop}}
+	\centerline{\includegraphics[width=0.95\textwidth]{simu-altpop}}
 	\caption{Results for the simulations with summary statistics with LD matrices based on two different populations.
-One comes from the same population used for computing the GWAS summary statistics (North-West Europe), while the other one comes from South Europe (alternative LD reference).
+One comes from the same ancestry used for computing the GWAS summary statistics (North-West Europe), while the other one comes from South Europe (alternative LD reference).
 Reported 95\% confidence intervals are computed from 10,000 non-parametric bootstrap replicates of the mean.
-Red bars correspond to using the LD with independent blocks (Methods section \ref{sec:LD}).}
+Red bars correspond to using the LD with independent blocks (Methods section \ref{sec:LD}), which is a requirement for PRS-CS.}
 	\label{fig:simu-altpop}
@@
 QC plots comparing standard deviations usually show little deviation from the identity line (after the INFO score correction), except for coronary artery disease (CAD) summary statistics (Figures \ref{fig:brca_icogs_qc}-\ref{fig:t1d_illu_qc}).
+We recall that the popular CAD summary statistics used here come from a multi-ancestry GWAS meta-analysis with more than 20\% non-European samples \cite[]{nikpay2015comprehensive,prive2021using}.
 In figures \ref{fig:res_cad} and \ref{fig:res_brca_onco}-\ref{fig:res_t1d_aff}, we then provide similar results as figure \ref{fig:res_brca_icogs} for other phenotypes.
@@
 These are computed using function \texttt{pcor} of R package bigstatsr where 95\% confidence intervals are obtained through Fisher's Z-transformation; these values are then squared.
-Red bars correspond to using the LD with independent blocks (Methods).}
+Red bars correspond to using the LD with independent blocks (Methods), which is a requirement for PRS-CS.}
 	\label{fig:res_cad}
@@
 These are computed using function \texttt{pcor} of R package bigstatsr where 95\% confidence intervals are obtained through Fisher's Z-transformation; these values are then squared.
-Red bars correspond to using the LD with independent blocks (Methods).}
+Red bars correspond to using the LD with independent blocks (Methods), which is a requirement for PRS-CS.}
 	\label{fig:res_vitaminD}
 \end{figure}
+
+For all phenotypes, PRS-CS-auto is very robust, whatever the QC procedure used (e.g.\ see the results for CAD, Figure \ref{fig:res_cad}).
+We believe this is because of the strong regularization used internally.
+We then run LDpred2-auto-rob for CAD with different shrinkage multiplicative coefficients for the off-diagonal elements of the LD matrix, from 1 (LD matrix unchanged) to 0 (using the identity matrix instead).
+When using a strong shrinkage of 0.3 or 0.4, results for LDpred2-auto-rob are similar whatever the QC used (Figure \ref{fig:res_cad_shrinkage}).
+Results are always best when using ``qc2'', the most stringent QC; then no shrinkage (1) seems needed, while a strong shrinkage make predictive performance drop.
+For phenotypes with relatively large genetic effects (e.g.\ vitamin D, Figure \ref{fig:res_vitaminD}), PRS-CS-auto perform much worse, possibly due to using too much regularization.
 
@@
 Partial correlations are computed using function \texttt{pcor} of R package bigstatsr where 95\% confidence intervals are obtained through Fisher's Z-transformation, then all values are squared to report the phenotypic variance explained by PGS.
-Red bars correspond to using the LD with independent blocks (Methods).}
+Red bars correspond to using the LD with independent blocks (Methods), which is a requirement for PRS-CS.}
 	\label{fig:finngen}

### 2022-06-30 paper/paper-misspec.tex
almost finished with text changes

@@
 There can be some mismatch between the GWAS summary statistics and the LD reference used, which can e.g.\ lead to suboptimal predictive performance for polygenic scores.
-
-\begin{table}[!htb]
-	\centering
-	\begin{tabular}{|p{\dimexpr0.21\textwidth-2\tabcolsep-\arrayrulewidth\relax}|
-			p{\dimexpr0.49\textwidth-2\tabcolsep-\arrayrulewidth\relax}|
-			p{\dimexpr0.3\textwidth-2\tabcolsep-\arrayrulewidth\relax}|}
-		\hline
-		\textbf{Misspecification} & \textbf{Possible consequence(s)} & \textbf{Possible correction} \\
-		\hline
-		\hline
-		Misestimation of the total sample size$^{(1)}$ & Bias estimation of SNP heritability \cite[]{gazal2018functional,grotzinger2021pervasive} & Estimation of the effective sample size \\
-		\hline
-		Differences in per-variant sample sizes & Possible violation of model assumptions (e.g.\ equation \eqref{eq:model}) resulting in poor predictive performance & Imputation using equations \eqref{eq:imputeN} and \eqref{eq:imputeNeff}, and filtering$^{(2)}$   \\
-		\hline
-		Using imputed dosages for GWAS & Standard deviations of dosages are too small, resulting in overestimated effect sizes and standard errors & See section \ref{sec:imputation} \\
-		\hline
-		Using imputed dosages for LD computation &  & None required; LD seems correctly estimated (Figure \ref{fig:compare-cor}) \\
-		\hline 
-		Misestimation of imputation INFO scores & Misuse when filtering or adjusting input parameters based on these & Recomputing from a homogeneous subset \\
-		\hline
-		Error in summary statistics (e.g.\ allele inversions) & Strong violation of model assumptions, which can result in e.g.\ identifying false-positives in fine-mapping \cite[]{susie} & Quality control \\
-		\hline
-		Rounding of summary statistics (e.g.\ effect sizes and SEs) & Add unnecessary noise to the data & None, but this has almost no impact on the predictive performance (Figure \ref{fig:simu-rounded}) \\
-		\hline 
-		Ancestry mismatch between summary statistics and LD & Strong mismatch of LD and allele frequencies, often resulting in divergence of models & Checking ancestry proportions from GWAS summary statistics \cite[]{prive2021using} \\
-		\hline
-		Any & Possible violation of model assumptions resulting in poor predictive performance & Quality control, using more regularization, and constraining LD to blocks \\
-		\hline
-	\end{tabular}
-	\caption{Overview of possible misspecifications when using GWAS summary statistics, along with possible consequences and corrections.\\
-	$^{(1)}$ Examples: using the total number of cases and controls in a meta-analysis of binary traits, a larger effective sample size when using BOLT-LMM summary statistics \cite[]{loh2018mixed}, and a reduced effective sample size when using SAIGE on binary traits with a large prevalence \cite[]{zhou2018efficiently,mbatchou2021computationally}.\\
-	$^{(2)}$ E.g.\ removing SNPs with an effective sample size less than 0.67 times the 90th percentile of sample size \cite[]{zheng2017ld}.
-	\label{tab:overview}}
-\end{table}
 
@@
 Having a correlation matrix with independent blocks prevents small errors in the algorithm (e.g.\ the Gibbs sampler in LDpred2) from propagating to too many variants.
-It also makes running LDpred2 (and lassosum2) faster, taking e.g.\ 60\% of the initial time (since only 60\% of the initial non-zero values of the correlation matrix are kept).
+It also makes running LDpred2 (and lassosum2) faster, taking e.g.\ 70\% of the initial time (since only 70\% of the initial non-zero values of the correlation matrix are kept).
 Note that we also compute the correlations within the same blocks and from the same data for PRS-CS.
@@
 \begin{figure}[!h]
-	\centerline{\includegraphics[width=0.9\textwidth]{brca_onco_qc}}
+	\centerline{\includegraphics[width=0.95\textwidth]{brca_onco_qc}}
 	\caption{Standard deviations inferred from the OncoArray breast cancer GWAS summary statistics using equation \eqref{eq:approx-sd-log} (\textbf{A:} Raw or \textbf{B:} dividing them by $\sqrt{\text{INFO}}$) versus the ones inferred from the reported GWAS allele frequencies $f_j$ (using $\sqrt{2 f_j (1 - f_j)}$).
@@
 \begin{figure}[!h]
-	\centerline{\includegraphics[width=0.85\textwidth]{res-cad}}
+	\centerline{\includegraphics[width=0.9\textwidth]{res-cad}}
 	\caption{Variance explained of CAD in the UK Biobank by PGS derived from external summary statistics.
@@
 \begin{figure}[!h]
-	\centerline{\includegraphics[width=0.85\textwidth]{res-vitaminD}}
+	\centerline{\includegraphics[width=0.9\textwidth]{res-vitaminD}}
 	\caption{Variance explained of vitamin D in the UK Biobank by PGS derived from external summary statistics.
@@
 We then run LDpred2-auto-rob for CAD with different shrinkage multiplicative coefficients for the off-diagonal elements of the LD matrix, from 1 (LD matrix unchanged) to 0 (using the identity matrix instead).
-When using a strong shrinkage of 0.3 or 0.4, results for LDpred2-auto-rob are similar whatever the QC used (Figure \ref{fig:res_cad_shrinkage}).
-Results are always best when using ``qc2'', the most stringent QC; then no shrinkage (1) seems needed, while a strong shrinkage make predictive performance drop.
+When using a strong shrinkage of 0.3 or 0.4, results for LDpred2-auto-rob are similar whatever the QC used (Figure \ref{fig:res_cad_shrinkage}), as previously observed for PRS-CS-auto.
+Results with LDpred2-auto-rob are always best when using ``qc2'', the most stringent QC; then no shrinkage (1) seems needed, while a strong shrinkage make predictive performance drop.
 For phenotypes with relatively large genetic effects (e.g.\ vitamin D, Figure \ref{fig:res_vitaminD}), PRS-CS-auto perform much worse, possibly due to using too much regularization.
@@
 We then compare three different LD reference panels to use with these GWAS summary statistics (Methods section \ref{sec:altLD}).
-Interestingly, using the small Finnish LD reference panel composed of 503 individuals only we have defined here seems to consistently provide more predictive polygenic scores than using the large UK one composed of 10,000 individuals or the widely used European subset of the 1000 Genomes (1000G) data (Figure \ref{fig:finngen}).
+Interestingly, using the small Finnish LD reference panel we have defined here, composed of 503 individuals only, seems to almost always provide more predictive polygenic scores than using the large UK one composed of 10,000 individuals or the widely used European subset of the 1000 Genomes (1000G) data (Figure \ref{fig:finngen}).
+For PRS-CS-auto, no LD reference seems to consistently outperform the two others across all phenotypes considered.
 Note that the PGS here are validated in people of UK-like ancestry in the UK biobank to obtain sufficient sample sizes.
@@
 \begin{figure}[!h]
-	\centerline{\includegraphics[width=0.85\textwidth]{res-FIN}}
+	\centerline{\includegraphics[width=0.95\textwidth]{res-FIN}}
 	\caption{Results for PGS derived from five FinnGen GWAS summary statistics and using three different LD references.
@@
 We then compare three different LD reference panels to use with these GWAS summary statistics (Methods section \ref{sec:altLD}).
-Interestingly, when looking at height and to some extent at HDL cholesterol where we could get the best predictive performance, using the smaller Japanese LD reference we have defined here seems to provide more predictive PGS than using the one using a larger set of East Asian individuals from the UKBB or the widely used East Asian subset of the 1000 Genomes (1000G) data (Figure \ref{fig:bbj}).
-Note that the PGS here are validated in people of broad East Asian ancestry in the UK biobank and for continuous outcomes  to obtain sufficient sample sizes.
+Interestingly, when looking at height and HDL cholesterol, where we could get the best predictive performance, using the smaller Japanese LD reference we have defined here seems to provide more predictive PGS than using the one using a larger set of East Asian individuals from the UKBB or the widely used East Asian subset of the 1000 Genomes (1000G) data (Figure \ref{fig:bbj}), except when using PRS-CS-auto to predict height.
+Note that the PGS here are validated in people of broad East Asian ancestry in the UK biobank and for continuous outcomes to obtain moderate sample sizes for comparison.
+
+\subsection{Recommendations}
+
+Based on previous results from simulations and real-data analyses, we use this section to suggest some recommendations on how to handle misspecifications in GWAS summary statistics and to get the most of polygenic scores.
+We recall that an overview of these misspecifications, along with possible consequences, and possible corrections is summarized in table \ref{tab:overview}.
+First, standard quality control should always be performed, e.g.\ variants with low frequency or low imputation quality should be removed; just make sure to recompute MAF and INFO scores for the homogeneous subset of individuals you are interested in.
+Second, comparison of allele frequencies can be used to e.g.\ detect problems with signs of effects, while comparing standard deviations can be used to detect other issues such as low per-variant GWAS sample sizes.
+Third, the LD reference panel should be as close as possible (in terms of genetic ancestry) as the data used to derive the GWAS.
+In case of any doubt, ancestry proportions can be estimated from GWAS allele frequencies using the method derived in \cite{prive2021using}.
+Fourth, if some validation data is available for tuning hyper-parameters, one can use methods such as lassosum2 and LDpred2(-grid).
+The grid of parameters used in lassosum2 includes a broad range of regularization, which makes it very robust.
+For LDpred2, we recommend to also include a smaller value for the heritability in the grid of parameters, to allow for more regularization when needed.
+As these two methods use the same data and the same principle (i.e.\ hyper-parameter tuning), one could run both methods and should the best one when tuning hyper-parameters.
+Fifth, if no validation/tuning dataset is available, we recommend using LDpred2-auto with the two new parameters introduced here, or PRS-CS-auto if there are strong misspecifications and the phenotype of interest is known to have small effects only.
+Sixth, we recommend to form independent LD blocks in the 
+LD matrix as it proved to make the methods more robust (and slightly faster).
+
+\begin{table}[!htb]
+	\centering
+	\begin{tabular}{|p{\dimexpr0.21\textwidth-2\tabcolsep-\arrayrulewidth\relax}|
+			p{\dimexpr0.49\textwidth-2\tabcolsep-\arrayrulewidth\relax}|
+			p{\dimexpr0.3\textwidth-2\tabcolsep-\arrayrulewidth\relax}|}
+		\hline
+		\textbf{Misspecification} & \textbf{Possible consequence(s)} & \textbf{Possible correction} \\
+		\hline
+		\hline
+		Misestimation of the total sample size$^{(1)}$ & Bias estimation of SNP heritability \cite[]{gazal2018functional,grotzinger2021pervasive} & Estimation of the effective sample size \\
+		\hline
+		Differences in per-variant sample sizes & Possible violation of model assumptions (e.g.\ equation \eqref{eq:model}) resulting in poor predictive performance & Imputation using equations \eqref{eq:imputeN} and \eqref{eq:imputeNeff}, and filtering$^{(2)}$   \\
+		\hline
+		Using imputed dosages for GWAS & Standard deviations of dosages are too small, resulting in overestimated effect sizes and standard errors & See section \ref{sec:imputation} \\
+		\hline
+		Using imputed dosages for LD computation &  & None required; LD seems correctly estimated (Figure \ref{fig:compare-cor}) \\
+		\hline 
+		Misestimation of imputation INFO scores & Misuse when filtering or adjusting input parameters based on these & Recomputing from a homogeneous subset \\
+		\hline
+		Error in summary statistics (e.g.\ allele inversions) & Strong violation of model assumptions, which can result in e.g.\ identifying false-positives in fine-mapping \cite[]{susie} & Quality control \\
+		\hline
+		Rounding of summary statistics (e.g.\ effect sizes and SEs) & Add unnecessary noise to the data & None, but this has almost no impact on the predictive performance (Figure \ref{fig:simu-rounded}) \\
+		\hline 
+		Ancestry mismatch between summary statistics and LD & Strong mismatch of LD and allele frequencies, often resulting in divergence of models & Checking ancestry proportions from GWAS summary statistics \cite[]{prive2021using} \\
+		\hline
+		Any & Possible violation of model assumptions resulting in poor predictive performance & Quality control, using more regularization, and constraining LD to blocks \\
+		\hline
+	\end{tabular}
+	\caption{Overview of possible misspecifications when using GWAS summary statistics, along with possible consequences and corrections.\\
+		$^{(1)}$ Examples: using the total number of cases and controls in a meta-analysis of binary traits, a larger effective sample size when using BOLT-LMM summary statistics \cite[]{loh2018mixed}, and a reduced effective sample size when using SAIGE on binary traits with a large prevalence \cite[]{zhou2018efficiently,mbatchou2021computationally}.\\
+		$^{(2)}$ E.g.\ removing SNPs with an effective sample size less than 0.67 times the 90th percentile of sample size \cite[]{zheng2017ld}.
+		\label{tab:overview}}
+\end{table}
 
@@
 Imposing independent blocks on the LD matrix could result in further misspecifications \cite[]{zhou2021fast}, but here we have shown that well defined blocks can actually make PGS methods more robust.
-PRS-CS is currently one of the most robust PGS methods; for example, it can use the (small) 1000 Genomes data as LD reference \cite[]{ge2019polygenic}, and can even use a European LD reference panel with multi-ancestry GWAS summary statistics \cite[]{wang2021global}.
+PRS-CS is currently one of the most robust PGS methods; for example, it can use the (small) 1000 Genomes data as LD reference \cite[]{ge2019polygenic}, and can even use a European LD reference panel with multi-ancestry GWAS summary statistics \cite[]{wang2021global}, also shown here.
 We believe this is made possible by the use of a strong regularization in PRS-CS ($\phi^{-1} \psi_j^{-1} \ge 1$, which would approximately correspond to using $s=0.5$ in lassosum, $\delta = 1$ in lassosum2, and the new parameter $\texttt{shrink\_corr}=0.5$ in LDpred2-auto).

### 2022-06-30 paper/paper-misspec.tex
red markup

@@
 Third, we introduce two new optional parameters in LDpred2-auto to make it more robust to these types of misspecification.
-Note that, in this paper, we use ``robust'' to mean ``similar predictive performance''. 
-We also reimplement and use a new version of lassosum, called lassosum2, in which we better handle when per-variant sample sizes are different.
+{\color{red}Note that, in this paper, we use ``robust'' to mean ``similar predictive performance''. 
+We also reimplement and use a new version of lassosum, called lassosum2, in which we better handle when per-variant sample sizes are different.}
 We focus our investigations on LDpred2 and lassosum2 for two reasons.
@@
 Second, lassosum2 now uses the exact same input parameters as LDpred2, which makes it easy for us to test the different QCs and adjustments presented here.
-We additionally investigate PRS-CS and SBayesR, two competitive PGS methods \cite[]{ge2019polygenic,lloyd2019improved}, for which we develop some code to convert between the different formats required for the LD matrices and input GWAS summary statistics.
+{\color{red}We additionally investigate PRS-CS and SBayesR, two competitive PGS methods \cite[]{ge2019polygenic,lloyd2019improved}, for which we develop some code to convert between the different formats required for the LD matrices and input GWAS summary statistics.}
 
@@
 To get a set of genetically homogeneous individuals, we compute a robust Mahalanobis distance based on the first 16 PCs and further restrict individuals to those within a log-distance of 5 \cite[]{prive2020efficient}.
-This results in 362,307 individuals of Northwestern European ancestry.
+This results in 362,307 individuals {\color{red}of Northwestern European ancestry}.
 We randomly sample 300,000 individuals to form a training set (e.g.\ to run the GWAS), 10,000 individuals to form a validation set (to tune hyper-parameters), and use the remaining 52,307 individuals to form a test set (to evaluate final predictive models).
@@
 
-We also use the UK Biobank data for validation/testing in real data analyses, and use the same individuals of Northwestern European ancestry as described in the previous section.
+We also use the UK Biobank data for validation/testing in real data analyses, and use the same individuals {\color{red}of Northwestern European ancestry} as described in the previous section.
 We sample 10,000 individuals to form a validation set and use the remaining 352,307 individuals as test set.
@@
 
-\subsection{Model for summary statistics}
+{\color{red}\subsection{Model for summary statistics}}
 
@@
 
-It is assumed that the marginal (GWAS) effects follow
+{\color{red}It is assumed that the marginal (GWAS) effects follow
 \begin{align}
@@
 However, in practice, we only have estimates for $S$ and $R$, which causes some model misspecifications.
+Moreover, ubiquitous correlations between the large number of variants may cause a Gibbs sampler or any iterative approach used in PGS methods to fail properly estimating genetic effect sizes \cite[]{roberts1997updating}.}
 
@@
 Also note that the estimate of $\text{var}(G_j)$ has to account for the imputation accuracy, i.e.\ use $2 \cdot f_j \cdot (1 - f_j) \cdot \text{INFO}_j$ (Section \ref{sec:imputation}).
-In equations \eqref{eq:approx-sd-lin} and \eqref{eq:imputeN}, we estimate $\text{var}(y)$ by the first percentile (robust minimum) of $0.5 \left(N ~ \text{se}(\hat{\gamma}_j)^2 + \hat{\gamma}_j^2\right)$.
+{\color{red}In equations \eqref{eq:approx-sd-lin} and \eqref{eq:imputeN}, we estimate $\text{var}(y)$ by the first percentile (robust minimum) of $0.5 \left(N ~ \text{se}(\hat{\gamma}_j)^2 + \hat{\gamma}_j^2\right)$.}
 
@@
 Note that only positions $k$ for which $R_{k,j} \neq 0$ must be updated in this vector $R_{.,j}^T \beta^{(t-1)}$.
-Since bigsnpr v1.10.4, we now also use this updating strategy in LDpred2 (-grid and -auto), which makes it much faster, especially when the polygenicity is small.
+{\color{red}Since bigsnpr v1.10.4, we now also use this updating strategy in LDpred2 (-grid and -auto), which makes it much faster, especially when the polygenicity is small.}
 
@@
 By default in lassosum2, we use a similar sequence of 30 values, but between $\lambda_0$ and $\lambda_0 / 100$ instead, where $\lambda_0 = \max_j \big|\widehat{\beta}_j\big|$ is the minimum $\lambda$ for which no variable enters the model because the L1-regularization is too strong.
-To make lassosum2 more robust to different per-variant sample sizes $n_j$, we also introduce per-variant penalty factors $\sqrt{\max_k(n_k) / n_j}$, by which we internally multiply $\lambda$ and $\delta$ to penalize variants with smaller GWAS sample sizes more.
+{\color{red}To make lassosum2 more robust to different per-variant sample sizes $n_j$, we also introduce per-variant penalty factors $\sqrt{\max_k(n_k) / n_j}$, by which we internally multiply $\lambda$ and $\delta$ to penalize variants with smaller GWAS sample sizes more.}
 Note that we do not provide an ``auto'' version using pseudo-validation (as in lassosum \cite[]{mak2017polygenic}) since we have not found the pseudo-validation scores to be consistent enough with the predictive performance (Figure \ref{fig:pseudoval}).
@@
 We form nearly independent LD blocks using the optimal algorithm developed in \cite{prive2021optimal}.
-Note that a new parameter \texttt{max\_r2} has been added to this method since, which controls the maximum single squared correlation allowed outside blocks, which offers an extra guarantee that the splitting is good and makes the function much faster by discarding lots of possible splits.
-For different numbers of blocks and maximum number of variants in each block, we use the split that minimizes $\big(C2 \cdot \sqrt{5 + C1}\big)$, where $C1$ is the sum of squared correlations outside the blocks, and $C2$ is the sum of squared sizes of the blocks (Figure \ref{fig:cost-split}).
+{\color{red}Note that a new parameter \texttt{max\_r2} has been added to this method since, which controls the maximum single squared correlation allowed outside blocks, which offers an extra guarantee that the splitting is good and makes the function much faster by discarding lots of possible splits.
+For different numbers of blocks and maximum number of variants in each block, we use the split that minimizes $\big(C2 \cdot \sqrt{5 + C1}\big)$, where $C1$ is the sum of squared correlations outside the blocks, and $C2$ is the sum of squared sizes of the blocks (Figure \ref{fig:cost-split}).}
 Having a correlation matrix with independent blocks prevents small errors in the algorithm (e.g.\ the Gibbs sampler in LDpred2) from propagating to too many variants.
 It also makes running LDpred2 (and lassosum2) faster, taking e.g.\ 70\% of the initial time (since only 70\% of the initial non-zero values of the correlation matrix are kept).
-Note that we also compute the correlations within the same blocks and from the same data for PRS-CS.
+{\color{red}Note that we also compute the correlations within the same blocks and from the same data for PRS-CS.}
 
@@
 Results when using imputed sample sizes are very similar to when using the true ones.
-Note that C+T does not use this sample size information, and that PRS-CS can only be provided with a single value (we use the maximum one).
+Note that C+T does not use this sample size information{\color{red}, and that PRS-CS can only be provided with a single value (we use the maximum one)}.
 When using the total GWAS sample size instead of the per-variant sample sizes, predictive performance slightly decreases to 0.158 for lassosum and to 0.164 for lassosum2, but dramatically decreases for LDpred2 with new values of 0.134 for LDpred2-grid, 0.122 for LDpred2-auto, 0.125 for LDpred2-inf (Figure \ref{fig:simu-misN}).
 This extreme simulation scenario shows that LDpred2 can be sensitive to the misspecification of per-variant GWAS sample sizes, whereas lassosum (and lassosum2) seems little affected by this.
-With 0.141 for SBayesR, predictive performance is actually larger than when using the true per-variant sample sizes.
+{\color{red}With 0.141 for SBayesR, predictive performance is actually larger than when using the true per-variant sample sizes.}
 
@@
 The GWAS sample size is either ``true'' when providing the true per-variant sample size, ``max'' when providing instead the maximum sample size as a unique value to be used for all variants, ``imputed'' (cf.\ equation \eqref{eq:imputeN}), or ``any'' when the method does not use this information (the case for C+T).
-Red bars correspond to using the LD with independent blocks (Methods section \ref{sec:LD}), which is a requirement for lassosum and PRS-CS.}
+Red bars correspond to using the LD with independent blocks (Methods section \ref{sec:LD}){\color{red}, which is a requirement for lassosum and PRS-CS}.}
 	\label{fig:simu-misN}
@@
 We conduct further investigations to explain the results of figure \ref{fig:simu-misN}.
-First, the results for LDpred2-auto and SBayesR are similar to LDpred2-inf in these simulations because the internal $p$ estimate of LDpred2-auto (and similarly the proportion of null variants in SBayesR) always continuously increases up to $1$, which makes them behave as an infinitesimal model ($p = 1$).
+{\color{red}First, the results for LDpred2-auto and SBayesR are similar to LDpred2-inf in these simulations because the internal $p$ estimate of LDpred2-auto (and similarly the proportion of null variants in SBayesR) always continuously increases up to $1$, which makes them behave as an infinitesimal model ($p = 1$).}
 To overcome this limitation in LDpred2-auto, we introduce two new parameters to make it more robust, and refer to this as ``LDpred2-auto-rob'' here (Methods section \ref{sec:robust}).
@@
 
-As secondary analysis, we rerun this simulation with a smaller heritability of 4\% instead.
+{\color{red}As secondary analysis, we rerun these simulations with a smaller heritability of 4\% instead.
 While predictive performance are overall much lower, relative performance are less attenuated when using the maximum sample size, otherwise results are highly consistent as before (Figure \ref{fig:simu-misN-smallh2}).
-We also try using the shrunk LD matrix from GCTB that is used e.g.\ for SBayesR \cite[]{lloyd2019improved}; in these simulations, higher predictive performance are obtained from LDpred2-auto with this shrunk LD matrix, whereas SBayesR benefit from using LDpred2's windowed LD matrix (Figure \ref{fig:simu-misN-shrunk}).
+We also try using the shrunk LD matrix from GCTB that is used e.g.\ for SBayesR \cite[]{lloyd2019improved}; in these simulations, higher predictive performance are obtained from LDpred2-auto with this shrunk LD matrix, whereas SBayesR benefit from using LDpred2's windowed LD matrix (Figure \ref{fig:simu-misN-shrunk}).}
 
@@
 As a second option, instead of using dosages to compute the GWAS summary statistics, it has been argued that using multiple imputation (MI) would be more appropriate \cite[]{Palmer2016}.
-Here MI consists in forming multiple (e.g.\ 20) datasets of hard-called genotypes by sampling them according to the imputation probabilities stored in the BGEN files, performing a GWAS on each of them, and pooling the GWAS estimates.
+{\color{red}Here MI consists in forming multiple (e.g.\ 20) datasets of hard-called genotypes by sampling them according to the imputation probabilities stored in the BGEN files, performing a GWAS on each of them, and pooling the GWAS estimates.}
 In simulations, we show that $\hat{\gamma}_j^{\text{MI}} \approx \hat{\gamma}_j^{\text{imp}} \cdot \text{INFO}_j$ and $Z_j^{\text{MI}} \approx Z_j^{\text{imp}} \cdot \text{INFO}_j$, where $Z = \hat{\gamma} / \text{se}(\hat{\gamma})$ (Figure \ref{fig:compare-mi}).
@@
 Moreover, new additions for robustness introduced before, namely LDpred2-low-h2, LDpred2-auto-rob, and forming independent blocks in the LD matrix also improve predictive performance for all corrections (Figure \ref{fig:simu-info}).
-However, these corrections do not prove beneficial for PRS-CS-auto (Figure \ref{fig:simu-info}).
+{\color{red}However, these corrections do not prove beneficial for PRS-CS-auto (Figure \ref{fig:simu-info}).
 
-When performing similar simulations with a heritability of 4\% (instead of 20\%), correction ``sqrt\_info'' still provides slightly higher predictive performance for all methods (compared to no correction), yet the other two corrections sometimes provide lower predictive performance (Figure \ref{fig:simu-info-smallh2}).
+When performing similar simulations with a heritability of 4\% (instead of 20\%), correction ``sqrt\_info'' still provides slightly higher predictive performance for all methods (compared to no correction), yet the other two corrections sometimes provide lower predictive performance (Figure \ref{fig:simu-info-smallh2}).}
 In the real data applications hereinafter, we therefore choose to use the first correction, ``sqrt\_info'', which seems to more consistently provide better results, and is simple because it is somewhat equivalent to post-processing PGS effects by multiplying them by $\sqrt{\text{INFO}}$.
@@
 Here we design simulations to understand the impact of using a mismatched LD reference panel, e.g.\ that comes from a different population compared to the one used to compute the GWAS summary statistics.
-We use the same simulation setup as before (Methods), i.e.\ using individuals of Northwestern European ancestry from the UK Biobank for training (GWAS), validation (tuning of hyper-parameters), and testing (the final models).
-In addition to the LD reference panel of Northwestern European ancestry, we design an alternative reference panel based on 10,000 individuals from South Europe by using the ``Italy'' center defined in \cite{prive2021high}.
-Allele frequencies and pairwise correlations are quite similar between these two LD reference panels (Figure \ref{fig:compare-altpop}).
+We use the same simulation setup as before (Methods){\color{red}, i.e.\ using individuals of Northwestern European ancestry from the UK Biobank for training (GWAS), validation (tuning of hyper-parameters), and testing (the final models)}.
+In addition {\color{red}to the LD reference panel of Northwestern European ancestry}, we design an alternative reference panel based on 10,000 individuals from South Europe by using the ``Italy'' center defined in \cite{prive2021high}.
+{\color{red}Allele frequencies and pairwise correlations are quite similar between these two LD reference panels (Figure \ref{fig:compare-altpop}).}
 When using this alternative LD reference panel instead of a well-matched one as in the previous sections, squared correlations between the polygenic scores and the simulated phenotypes drop from 0.173 to 0.166 for lassosum2, from 0.168 to 0.159 for LDpred2(-grid), from 0.174 to 0.169 for LDpred2-low-h2, from 0.142 to 0.136 for LDpred2-auto, from 0.162 to 0.147 for LDpred2-auto-rob, from 0.143 to 0.138 for LDpred2-inf, and from  0.163 to 0.155 for PRS-CS-auto (Figure \ref{fig:simu-altpop}, averaging over 10 simulations).
 Forming independent LD blocks in these two LD matrices always improve predictive performance, yet again.
-We also compute and try using a shrunk LD matrix (from GCTB) for this alternative LD reference panel; again, using this type of LD matrix seems beneficial for LDpred2-auto.
-Finally, under the same simulation scenario, but with a smaller simulated heritability of 4\% (instead of 20\%), there remains a small drop in predictive performance when using the alternative LD reference (Figure \ref{fig:simu-altpop-smallh2}).
+{\color{red}We also compute and test using a shrunk LD matrix (from GCTB) for this alternative LD reference panel; again, using this type of LD matrix seems beneficial for LDpred2-auto.
+Finally, under the same simulation scenario, but with a smaller simulated heritability of 4\% (instead of 20\%), there remains a small drop in predictive performance when using the alternative LD reference (Figure \ref{fig:simu-altpop-smallh2}).}
 
@@
 QC plots comparing standard deviations usually show little deviation from the identity line (after the INFO score correction), except for coronary artery disease (CAD) summary statistics (Figures \ref{fig:brca_icogs_qc}-\ref{fig:t1d_illu_qc}).
-We recall that the popular CAD summary statistics used here come from a multi-ancestry GWAS meta-analysis with more than 20\% non-European samples \cite[]{nikpay2015comprehensive,prive2021using}.
+{\color{red}Note that the popular CAD summary statistics used here come from a multi-ancestry GWAS meta-analysis with more than 20\% non-European samples \cite[]{nikpay2015comprehensive,prive2021using}.}
 In figures \ref{fig:res_cad} and \ref{fig:res_brca_onco}-\ref{fig:res_t1d_aff}, we then provide similar results as figure \ref{fig:res_brca_icogs} for other phenotypes.
@@
 
-For all phenotypes, PRS-CS-auto is very robust, whatever the QC procedure used (e.g.\ see the results for CAD, Figure \ref{fig:res_cad}).
+{\color{red}For all phenotypes, PRS-CS-auto is very robust, whatever the QC procedure used (e.g.\ see the results for CAD, Figure \ref{fig:res_cad}).
 We believe this is because of the strong regularization used internally.
@@
 Results with LDpred2-auto-rob are always best when using ``qc2'', the most stringent QC; then no shrinkage (1) seems needed, while a strong shrinkage make predictive performance drop.
-For phenotypes with relatively large genetic effects (e.g.\ vitamin D, Figure \ref{fig:res_vitaminD}), PRS-CS-auto perform much worse, possibly due to using too much regularization.
+For phenotypes with relatively large genetic effects (e.g.\ vitamin D, Figure \ref{fig:res_vitaminD}), PRS-CS-auto perform much worse, possibly due to using too much regularization.}
 
@@
 Interestingly, using the small Finnish LD reference panel we have defined here, composed of 503 individuals only, seems to almost always provide more predictive polygenic scores than using the large UK one composed of 10,000 individuals or the widely used European subset of the 1000 Genomes (1000G) data (Figure \ref{fig:finngen}).
-For PRS-CS-auto, no LD reference seems to consistently outperform the two others across all phenotypes considered.
+{\color{red}For PRS-CS-auto, no LD reference seems to consistently outperform the two others across all phenotypes considered.}
 Note that the PGS here are validated in people of UK-like ancestry in the UK biobank to obtain sufficient sample sizes.
@@
 We then compare three different LD reference panels to use with these GWAS summary statistics (Methods section \ref{sec:altLD}).
-Interestingly, when looking at height and HDL cholesterol, where we could get the best predictive performance, using the smaller Japanese LD reference we have defined here seems to provide more predictive PGS than using the one using a larger set of East Asian individuals from the UKBB or the widely used East Asian subset of the 1000 Genomes (1000G) data (Figure \ref{fig:bbj}), except when using PRS-CS-auto to predict height.
+Interestingly, when looking at height and HDL cholesterol, where we could get the best predictive performance, using the smaller Japanese LD reference we have defined here seems to provide more predictive PGS than using the one using a larger set of East Asian individuals from the UKBB or the widely used East Asian subset of the 1000 Genomes (1000G) data (Figure \ref{fig:bbj}){\color{red}, except when using PRS-CS-auto to predict height}.
 Note that the PGS here are validated in people of broad East Asian ancestry in the UK biobank and for continuous outcomes to obtain moderate sample sizes for comparison.
 
-\subsection{Recommendations}
+{\color{red}\subsection{Recommendations}
 
@@
 Sixth, we recommend to form independent LD blocks in the 
-LD matrix as it proved to make the methods more robust (and slightly faster).
+LD matrix as it proved to make the methods more robust (and slightly faster).}
 
@@
 		\hline
-		Differences in per-variant sample sizes & Possible violation of model assumptions (e.g.\ equation \eqref{eq:model}) resulting in poor predictive performance & Imputation using equations \eqref{eq:imputeN} and \eqref{eq:imputeNeff}, and filtering$^{(2)}$   \\
+		Differences in per-variant sample sizes & Possible violation of model assumptions {\color{red}(e.g.\ equation \eqref{eq:model}) resulting in poor predictive performance} & Imputation using equations \eqref{eq:imputeN} and \eqref{eq:imputeNeff}, and filtering$^{(2)}$   \\
 		\hline
@@
 It would also be useful to investigate the impact of the QC, well-matched LD reference panels and other adjustments we propose here on other (non-PGS) methods, for which consistency and robustness are likely to be very important as well \cite[]{chen2021improved}.
-For example, we are interested in assessing the impact of these misspecifications on the inference of disease architecture parameters in some future work.
+{\color{red}For example, we are interested in assessing the impact of these misspecifications on the inference of disease architecture parameters in some future work.}
 

### 2022-06-30 paper/paper-misspec-supp.tex
almost finished with text changes
ordering of supp figs

@@
 
-\vspace*{5em}
+\vspace*{7em}
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{figure}[!h]
+	\centerline{\includegraphics[width=0.8\textwidth]{simu_hist_info}}
+	\caption{Histogram of the imputation INFO scores of the 40,000 variants from chromosome 22 of the UK Biobank data used in the simulations. \label{fig:hist-info}}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\FloatBarrier
+
+\begin{figure}[p]
+	\centerline{\includegraphics[width=0.85\textwidth]{pseudoval}}
+	\caption{For one simulation, scores from validation versus scores from pseudo-validation as described in \cite{mak2017polygenic} using either \textbf{A:} correlations (the default in lassosum) or \textbf{B:} p-values, when computing local false discovery rates.}
+	\label{fig:pseudoval}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\FloatBarrier
 
@@
 	\label{fig:cost-split}
-\end{figure}
-
-\FloatBarrier
-
-\begin{figure}[h]
-	\centerline{\includegraphics[width=0.7\textwidth]{simu-qc-plot}}
-	\caption{Quality control plot, as proposed in \cite{prive2020ldpred2}, for the simulations with sample size misspecification.
-The standard deviations are derived from the summary statistics assuming the same global GWAS sample size for all variants (300,000).}
-	\label{fig:qcplot}
 \end{figure}
@@
 	\label{fig:simu-misN-shrunk}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\FloatBarrier
-
-\begin{figure}[p]
-	\centerline{\includegraphics[width=0.8\textwidth]{simu_hist_info}}
-	\caption{Histogram of the imputation INFO scores of the 40,000 variants from chromosome 22 of the UK Biobank data used in the simulations. \label{fig:hist-info}}
 \end{figure}
@@
 \begin{figure}[p]
-	\centerline{\includegraphics[width=0.9\textwidth]{simu-rounded}}
-	\caption{Results for the simulations with summary statistics (effect sizes and their standard errors) possibly rounded to two significant digits, averaged over 10 simulations for each scenario.
-Reported 95\% confidence intervals are computed from 10,000 non-parametric bootstrap replicates of the mean.
-Red bars correspond to using the LD with independent blocks (Methods).}
-	\label{fig:simu-rounded}
+	\centerline{\includegraphics[width=0.65\textwidth]{brca_onco_hist_info}}
+	\caption{Histogram of the imputation INFO scores of the HapMap3 variants from the OncoArray GWAS summary statistics for breast cancer \cite[]{michailidou2017association}. \label{fig:hist-info-brca}}
 \end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\FloatBarrier
 
@@
 
-\begin{figure}[p]
-	\centerline{\includegraphics[width=0.65\textwidth]{brca_onco_hist_info}}
-	\caption{Histogram of the imputation INFO scores of the HapMap3 variants from the OncoArray GWAS summary statistics for breast cancer \cite[]{michailidou2017association}. \label{fig:hist-info-brca}}
-\end{figure}
-
 \FloatBarrier
@@
 \begin{figure}[p]
+	\centerline{\includegraphics[width=0.95\textwidth]{brca_icogs_qc}}
+	\caption{Standard deviations inferred from the iCOGS breast cancer GWAS summary statistics (\textbf{A:} Raw or \textbf{B:} dividing them by $\sqrt{\text{INFO}}$) versus the ones inferred from the reported GWAS allele frequencies.
+Only 100,000 variants are represented, at random.}
+	\label{fig:brca_icogs_qc}
+\end{figure}
+
+\begin{figure}[p]
 	\centerline{\includegraphics[width=0.75\textwidth]{hist_bad_pos}}
-	\caption{Histogram of the positions of outlier variants from Figure 5B in the main text.%TODO: VERIF NUMBER AT THE END
+	\caption{Histogram of the positions of outlier variants from Figure 4B in the main text.%TODO: VERIF NUMBER AT THE END
 	\label{fig:hist-bad-pos}}
@@
 \begin{figure}[p]
-	\centerline{\includegraphics[width=0.95\textwidth]{brca_icogs_qc}}
-	\caption{Standard deviations inferred from the iCOGS breast cancer GWAS summary statistics (\textbf{A:} Raw or \textbf{B:} dividing them by $\sqrt{\text{INFO}}$) versus the ones inferred from the reported GWAS allele frequencies.
-Only 100,000 variants are represented, at random.}
-	\label{fig:brca_icogs_qc}
+	\centerline{\includegraphics[width=0.7\textwidth]{res-brca_icogs}}
+	\caption{Variance explained of BRCA in the UK Biobank by PGS derived from external summary statistics (iCOGS).
+These are computed using function \texttt{pcor} of R package bigstatsr where 95\% confidence intervals are obtained through Fisher's Z-transformation; these values are then squared.}
+	\label{fig:res_brca_icogs}
 \end{figure}
+
+\FloatBarrier
 
@@
 \FloatBarrier
-
-\begin{figure}[p]
-	\centerline{\includegraphics[width=0.7\textwidth]{res-brca_icogs}}
-	\caption{Variance explained of BRCA in the UK Biobank by PGS derived from external summary statistics (iCOGS).
-These are computed using function \texttt{pcor} of R package bigstatsr where 95\% confidence intervals are obtained through Fisher's Z-transformation; these values are then squared.}
-	\label{fig:res_brca_icogs}
-\end{figure}
 
@@
 
+\FloatBarrier
+
+\begin{figure}[p]
+	\centerline{\includegraphics[width=0.7\textwidth]{simu-qc-plot}}
+	\caption{Quality control plot, as proposed in \cite{prive2020ldpred2}, for the simulations with sample size misspecification.
+The standard deviations are derived from the summary statistics assuming the same global GWAS sample size for all variants (300,000).}
+	\label{fig:qcplot}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \begin{figure}[p]
@@
 Red bars correspond to using the LD with independent blocks (Methods).
-Parameter \texttt{shrinkage} corresponds to the new parameter of LDpred2-auto.}
+The shrinkage corresponds to the new parameter \texttt{shrink\_corr} of LDpred2-auto.}
 	\label{fig:res_cad_shrinkage}
@@
 \begin{figure}[p]
-	\centerline{\includegraphics[width=0.85\textwidth]{pseudoval}}
-	\caption{For one simulation, scores from validation versus scores from pseudo-validation as described in \cite{mak2017polygenic} using either \textbf{A:} correlations (the default in lassosum) or \textbf{B:} p-values, when computing local false discovery rates.}
-	\label{fig:pseudoval}
+	\centerline{\includegraphics[width=0.95\textwidth]{compare-cor}}
+	\caption{Comparing pairwise correlations between a subset of HapMap3 variants on chromosome 1 with an INFO score lower than 0.85, computed in three different ways: 1/ (left y-axis) from imputed dosages using 10,000 individuals from the UK Biobank (UKBB) data; 2/ (right y-axis) from multiple imputation (i.e.\ generating multiple complete datasets sampled according to imputation probabilities, computing correlations, and averaging results) also using UKBB; 3/ (common x-axis) from 190 individuals from the 1000 Genomes data (GBR and CEU).
+Each point, representing the correlation between two variants, is colored by the geometric mean of the INFO scores of these two variants.}
+	\label{fig:compare-cor}
 \end{figure}
@@
 \begin{figure}[p]
-	\centerline{\includegraphics[width=0.95\textwidth]{compare-cor}}
-	\caption{Comparing pairwise correlations between a subset of HapMap3 variants on chromosome 1 with an INFO score lower than 0.85, computed in three different ways: 1/ (left y-axis) from imputed dosages using 10,000 individuals from the UK Biobank (UKBB) data; 2/ (right y-axis) from multiple imputation (i.e.\ generating multiple complete datasets sampled according to imputation probabilities, computing correlations, and averaging results) also using UKBB; 3/ (common x-axis) from 190 individuals from the 1000 Genomes data (GBR and CEU).
-Each point, representing the correlation between two variants, is colored by the geometric mean of the INFO scores of these two variants.}
-	\label{fig:compare-cor}
+	\centerline{\includegraphics[width=0.9\textwidth]{simu-rounded}}
+	\caption{Results for the simulations with summary statistics (effect sizes and their standard errors) possibly rounded to two significant digits, averaged over 10 simulations for each scenario.
+Reported 95\% confidence intervals are computed from 10,000 non-parametric bootstrap replicates of the mean.
+Red bars correspond to using the LD with independent blocks (Methods).}
+	\label{fig:simu-rounded}
 \end{figure}

### 2022-06-30 paper/paper-misspec.tex
ordering of supp figs

@@
 We use other external GWAS summary statistics for which INFO scores are reported (see Table \ref{tab:sumstats}); they all have a very high mean INFO score (larger than 0.94), except for T1D-affy (0.885).
-QC plots comparing standard deviations usually show little deviation from the identity line (after the INFO score correction), except for coronary artery disease (CAD) summary statistics (Figures \ref{fig:brca_icogs_qc}-\ref{fig:t1d_illu_qc}).
+QC plots comparing standard deviations usually show little deviation from the identity line (after the INFO score correction), except for coronary artery disease (CAD) summary statistics (Figures \ref{fig:cad_qc}-\ref{fig:t1d_illu_qc}).
 {\color{red}Note that the popular CAD summary statistics used here come from a multi-ancestry GWAS meta-analysis with more than 20\% non-European samples \cite[]{nikpay2015comprehensive,prive2021using}.}

### 2022-07-01 paper/paper-misspec-supp.tex
use \citet{}

@@
 \onehalfspacing
-%\usepackage[authoryear]{natbib}
-\usepackage{natbib} %\bibpunct{(}{)}{;}{author-year}{}{,}
+\usepackage[numbers,super]{natbib}
 
@@
 	\centerline{\includegraphics[width=0.85\textwidth]{pseudoval}}
-	\caption{For one simulation, scores from validation versus scores from pseudo-validation as described in \cite{mak2017polygenic} using either \textbf{A:} correlations (the default in lassosum) or \textbf{B:} p-values, when computing local false discovery rates.}
+	\caption{For one simulation, scores from validation versus scores from pseudo-validation as described in \citet{mak2017polygenic} using either \textbf{A:} correlations (the default in lassosum) or \textbf{B:} p-values, when computing local false discovery rates.}
 	\label{fig:pseudoval}
@@
 \centerline{\includegraphics[width=0.95\textwidth]{compare-mult-imp}}
-\caption{(Over)corrected GWAS \textbf{A:} Z-scores and \textbf{B:} effect sizes from imputed dosages (multiplying them by $\text{INFO}$) versus the ones obtained from multiple imputation (MI, 20 draws used, \cite{Palmer2016}), colored by INFO scores.
+\caption{(Over)corrected GWAS \textbf{A:} Z-scores and \textbf{B:} effect sizes from imputed dosages (multiplying them by $\text{INFO}$) versus the ones obtained from multiple imputation (MI, 20 draws used, \citet{Palmer2016}), colored by INFO scores.
 This shows only one tenth of the simulated variants, because MI can be computationally intensive.
@@
 	\centerline{\includegraphics[width=0.7\textwidth]{simu-qc-plot}}
-	\caption{Quality control plot, as proposed in \cite{prive2020ldpred2}, for the simulations with sample size misspecification.
+	\caption{Quality control plot, as proposed in \citet{prive2020ldpred2}, for the simulations with sample size misspecification.
 The standard deviations are derived from the summary statistics assuming the same global GWAS sample size for all variants (300,000).}
@@
 
-\bibliographystyle{natbib}
+\bibliographystyle{unsrtnat}
 \bibliography{refs}

### 2022-07-01 paper/paper-misspec.tex
use \citet{}

@@
 We restrict individuals to the ones used for computing the principal components (PCs) in the UK Biobank (Field 22020).
-These individuals are unrelated and have passed some quality control including removing samples with a missing rate on autosomes larger than 0.02, having a mismatch between inferred sex and self-reported sex, and outliers based on heterozygosity (more details can be found in section S3 of \cite{bycroft2018uk}).
+These individuals are unrelated and have passed some quality control including removing samples with a missing rate on autosomes larger than 0.02, having a mismatch between inferred sex and self-reported sex, and outliers based on heterozygosity (more details can be found in section S3 of \citet{bycroft2018uk}).
 To get a set of genetically homogeneous individuals, we compute a robust Mahalanobis distance based on the first 16 PCs and further restrict individuals to those within a log-distance of 5 \cite[]{prive2020efficient}.
@@
 We read the UK Biobank data into two different datasets using function \texttt{snp\_readBGEN} from R package bigsnpr \cite[]{prive2017efficient}, one by reading the BGEN data as hard-called genotypes at random according to the imputation probabilities, and another one by reading it as dosages (i.e.\ expected values according to the imputation probabilities).
-The first dataset is used as what could be the real genotype calls and the second dataset as what would be its imputed version; this design technique was used in \cite{prive2019making}.
+The first dataset is used as what could be the real genotype calls and the second dataset as what would be its imputed version; this design technique was used in \citet{prive2019making}.
 
@@
 We sample 10,000 individuals to form a validation set and use the remaining 352,307 individuals as test set.
-We restrict to the 1,054,315 HapMap3 variants used in the LD reference provided in \cite{prive2020ldpred2}.
+We restrict to the 1,054,315 HapMap3 variants used in the LD reference provided in \citet{prive2020ldpred2}.
 
@@
 To derive polygenic scores, we use published GWAS summary statistics listed in table \ref{tab:sumstats}.
-We also use GWAS summary statistics for five disease endpoints from FinnGen\cite{kurki2022finngen} (release 6) and for four continuous outcomes from Biobank Japan \cite[]{sakaue2021cross}.
+We also use GWAS summary statistics for five disease endpoints from FinnGen \cite[]{kurki2022finngen} (release 6) and for four continuous outcomes from Biobank Japan \cite[]{sakaue2021cross}.
 
@@
 		\hline
-		Breast cancer (BRCA) [iCOGS] & \cite{michailidou2017association} & ~~87,037 & 11,792,542 & 1,051,242 & 0.841 \\
-		Breast cancer (BRCA) [OncoArray] & \cite{michailidou2017association} & 104,442 & 11,792,542 & 1,054,233 & 0.968 \\
-		Type 1 diabetes (T1D) [Affymetrix] & \cite{censin2017childhood} & ~~~~~5516  & ~~8,996,866 & ~~~934,712 & 0.885 \\	
-		Type 1 diabetes (T1D) [Illumina] & \cite{censin2017childhood} & ~~~~~7982  & ~~8,996,866 & ~~~949,334 & 0.942 \\
-		Prostate cancer (PRCA) & \cite{schumacher2018association} & 135,316 & 20,370,946 & ~~~818,400 & 0.969 \\
-		Depression (MDD) [without UKBB] & \cite{wray2018genome} & 110,464 & ~~9,874,289 & 1,049,455 & 0.968 \\
-		Coronary artery disease (CAD) & \cite{nikpay2015comprehensive} & 129,014 & ~~9,455,778 & 1,052,200 & 0.982 \\
-		Vitamin D & \cite{jiang2018genome} & ~~79,366 & ~~2,579,296 & 1,016,935 & NA \\
+		Breast cancer (BRCA) [iCOGS] & \citet{michailidou2017association} & ~~87,037 & 11,792,542 & 1,051,242 & 0.841 \\
+		Breast cancer (BRCA) [OncoArray] & \citet{michailidou2017association} & 104,442 & 11,792,542 & 1,054,233 & 0.968 \\
+		Type 1 diabetes (T1D) [Affymetrix] & \citet{censin2017childhood} & ~~~~~5516  & ~~8,996,866 & ~~~934,712 & 0.885 \\	
+		Type 1 diabetes (T1D) [Illumina] & \citet{censin2017childhood} & ~~~~~7982  & ~~8,996,866 & ~~~949,334 & 0.942 \\
+		Prostate cancer (PRCA) & \citet{schumacher2018association} & 135,316 & 20,370,946 & ~~~818,400 & 0.969 \\
+		Depression (MDD) [without UKBB] & \citet{wray2018genome} & 110,464 & ~~9,874,289 & 1,049,455 & 0.968 \\
+		Coronary artery disease (CAD) & \citet{nikpay2015comprehensive} & 129,014 & ~~9,455,778 & 1,052,200 & 0.982 \\
+		Vitamin D & \citet{jiang2018genome} & ~~79,366 & ~~2,579,296 & 1,016,935 & NA \\
 		\hline
@@
 Instead of using a regularized version of the correlation matrix $R$ parameterized by $s$, $R_s = (1 - s) R + s I$ (where $0 < s \leq 1$), we use $R_{\delta} = R + \delta I$ (where $\delta > 0$), which makes it clearer that lassosum is also using L2-regularization (therefore elastic-net).
-Then, from \cite{mak2017polygenic}, the solution from lassosum can be obtained by iteratively updating 
+Then, from \citet{mak2017polygenic}, the solution from lassosum can be obtained by iteratively updating 
 \[
@@
 \]
-Following the notations from \cite{prive2020ldpred2}, denote $\widetilde{X} = \frac{1}{\sqrt{n-1}} C_n G S^{-1}$, where $G$ is the genotype matrix, $C_n$ is the centering matrix and $S$ is the diagonal matrix of standard deviations of the columns of $G$.
+Following the notations from \citet{prive2020ldpred2}, denote $\widetilde{X} = \frac{1}{\sqrt{n-1}} C_n G S^{-1}$, where $G$ is the genotype matrix, $C_n$ is the centering matrix and $S$ is the diagonal matrix of standard deviations of the columns of $G$.
 Then $\widetilde{X}_j^T \widetilde{X} = R_{j,.} = R_{.,j}^T$ and $\widetilde{X}_j^T \widetilde{X}_j = 1$.
-Moreover, using the notations from \cite{prive2020ldpred2}, $u_j^{(t)} = \widehat{\beta}_j - R_{.,j}^T \beta^{(t-1)} + \beta_j^{(t-1)}$,
+Moreover, using the notations from \citet{prive2020ldpred2}, $u_j^{(t)} = \widehat{\beta}_j - R_{.,j}^T \beta^{(t-1)} + \beta_j^{(t-1)}$,
 where $r_j = \widehat{\beta}_j =  \dfrac{\widehat{\gamma}_j}{\sqrt{n_j ~ \text{se}(\widehat{\gamma}_j)^2 + \widehat{\gamma}_j^2}}$ and $\widehat{\gamma}_j$ is the GWAS effect of variant $j$ and $n$ is the GWAS sample size \cite[]{mak2017polygenic,prive2021high}.
@@
 Here we introduce the small changes made to LDpred2 (-grid and -auto) in order to make them more robust.
-First, LDpred2-low-h2 simply consists in running LDpred2-grid while testing $h^2$ within $\{0.3, 0.7, 1, 1.4\} \cdot h^2_\text{LDSC}$ (note the added $0.3$ compared to \cite{prive2020ldpred2}), where $h^2_\text{LDSC}$ is the heritability estimate from LD score regression.
+First, LDpred2-low-h2 simply consists in running LDpred2-grid while testing $h^2$ within $\{0.3, 0.7, 1, 1.4\} \cdot h^2_\text{LDSC}$ (note the added $0.3$ compared to \citet{prive2020ldpred2}), where $h^2_\text{LDSC}$ is the heritability estimate from LD score regression.
 Indeed, we show in simulations here that using lower values for $h^2$ may provide higher predictive performance in the case of misspecifications (thanks to more shrinkage of the effects).
@@
 
-We form nearly independent LD blocks using the optimal algorithm developed in \cite{prive2021optimal}.
+We form nearly independent LD blocks using the optimal algorithm developed in \citet{prive2021optimal}.
 {\color{red}Note that a new parameter \texttt{max\_r2} has been added to this method since, which controls the maximum single squared correlation allowed outside blocks, which offers an extra guarantee that the splitting is good and makes the function much faster by discarding lots of possible splits.
@@
 
-\cite{marchini2010genotype} showed that the IMPUTE INFO measure is highly concordant with the MACH measure $\frac{\text{var}(G_j)}{2 \hat{\theta}_j (1 - \hat{\theta}_j)}$, where $\hat{\theta}_j$ is the estimated allele frequency of $G_j$, the genotypes for variant $j$.
+\citet{marchini2010genotype} showed that the IMPUTE INFO measure is highly concordant with the MACH measure $\frac{\text{var}(G_j)}{2 \hat{\theta}_j (1 - \hat{\theta}_j)}$, where $\hat{\theta}_j$ is the estimated allele frequency of $G_j$, the genotypes for variant $j$.
 Therefore, when using the expected genotypes from imputation (allele dosages), their standard deviations are often lower than the expected value under Hardy--Weinberg equilibrium, because $\text{INFO}_j \approx \frac{\text{var}(G_j)}{2 \hat{\theta}_j (1 - \hat{\theta}_j)}$.
@@
 We use the same simulation setup as before (Methods){\color{red}, i.e.\ using individuals of Northwestern European ancestry from the UK Biobank for training (GWAS), validation (tuning of hyper-parameters), and testing (the final models)}.
-In addition {\color{red}to the LD reference panel of Northwestern European ancestry}, we design an alternative reference panel based on 10,000 individuals from South Europe by using the ``Italy'' center defined in \cite{prive2021high}.
+In addition {\color{red}to the LD reference panel of Northwestern European ancestry}, we design an alternative reference panel based on 10,000 individuals from South Europe by using the ``Italy'' center defined in \citet{prive2021high}.
 {\color{red}Allele frequencies and pairwise correlations are quite similar between these two LD reference panels (Figure \ref{fig:compare-altpop}).}
@@
 For MDD and PRCA, most changes introduced before have little to no impact on predictive performance.
-For CAD and T1D, the QC proposed in \cite{prive2020ldpred2} and the additional QC proposed here provides much better predictive performance, especially for LDpred2-auto (and LDpred2-auto-rob) than when using no quality control, showing how important this preliminary step is.
+For CAD and T1D, the QC proposed in \citet{prive2020ldpred2} and the additional QC proposed here provides much better predictive performance, especially for LDpred2-auto (and LDpred2-auto-rob) than when using no quality control, showing how important this preliminary step is.
 
@@
 Here we investigate the use of different LD reference panels with GWAS summary statistics from two large biobanks of isolated populations.
-We use GWAS summary statistics for five disease endpoints from FinnGen\cite{kurki2022finngen} (release 6), namely breast and prostate cancers (BrCa and PrCa), coronary artery disease (CAD), type 1 and type 2 diabetes (T1D and T2D).
+We use GWAS summary statistics for five disease endpoints from FinnGen \cite[]{kurki2022finngen} (release 6), namely breast and prostate cancers (BrCa and PrCa), coronary artery disease (CAD), type 1 and type 2 diabetes (T1D and T2D).
 These were derived using SAIGE \cite[]{zhou2018efficiently}.
@@
 Third, the LD reference panel should be as close as possible (in terms of genetic ancestry) as the data used to derive the GWAS.
-In case of any doubt, ancestry proportions can be estimated from GWAS allele frequencies using the method derived in \cite{prive2021using}.
+In case of any doubt, ancestry proportions can be estimated from GWAS allele frequencies using the method derived in \citet{prive2021using}.
 Fourth, if some validation data is available for tuning hyper-parameters, one can use methods such as lassosum2 and LDpred2(-grid).
@@
 
-\bibliographystyle{ajhg}
+\bibliographystyle{unsrtnat}
 \bibliography{refs}

### 2022-07-01 paper/paper-misspec.tex
better explain methods

@@
 We design GWAS simulations where variants have different sample sizes, which is often the case when meta-analyzing GWAS summary statistics from multiple cohorts with different sets of variants \cite[]{wang2021global}.
-Using 40,000 variants from chromosome 22 (Methods), we simulate quantitative phenotypes with a heritability of 20\% and 2000 causal variants.
+Using 40,000 variants from chromosome 22 (Methods), we simulate quantitative phenotypes {\color{red}by picking 2000 causal variants at random, sampling their genetic effects (after scaling of the genotypes) from a normal distribution and adding some Gaussian noise to the genetic component of the phenotype so that it has a heritability of 20\%.
+This is implemented in function \texttt{snp\_simuPheno} of R package bigsnpr \cite[]{prive2017efficient}.}
 We then randomly divide the 40,000 variants into three groups: for half of the variants, we use 100\% of 300,000 individuals for GWAS, but use only 80\% for one quarter of the variants and 60\% for the remaining quarter.
+{\color{red}We use a simple linear regression for GWAS, as implemented in function \texttt{big\_univLinReg} of R package bigstatsr \cite[]{prive2017efficient}.}
 We then run C+T, lassosum, lassosum2 (Methods section \ref{lassosum2}), LDpred2-inf, LDpred2(-grid), LDpred2-auto, SBayesR, and PRS-CS-auto \cite[]{prive2019making,mak2017polygenic,prive2020ldpred2,lloyd2019improved,ge2019polygenic} by using either the true per-variant GWAS sample sizes, the total sample size for all variants, or per-variant sample sizes imputed using equation \eqref{eq:imputeN}.
@@
 In the comparison, we use the QC proposed in \citet{prive2020ldpred2} first (which ends up filtering on MAF here, which we call ``qc1'').
-We then also filter out the two long-range LD regions of chromosome 6 and 8 for the OncoArray summary statistics and remove around 500 variants when filtering on differences of AFs between summary statistics and the validation dataset (``qc2'').
+We then also filter out the two long-range LD regions of chromosome 6 and 8 for the OncoArray summary statistics and remove around 500 variants when filtering on differences of AFs {\color{red}(> 0.1)} between summary statistics and the validation dataset (``qc2'').
 As for the correction using INFO scores, we use the first correction, ``sqrt\_info'', which is simple because it is equivalent to post-processing PGS effects by multiplying them by $\sqrt{\text{INFO}}$.

### 2022-07-11 paper/paper-misspec.tex
text update

@@
 {\color{red}We use a simple linear regression for GWAS, as implemented in function \texttt{big\_univLinReg} of R package bigstatsr \cite[]{prive2017efficient}.}
-We then run C+T, lassosum, lassosum2 (Methods section \ref{lassosum2}), LDpred2-inf, LDpred2(-grid), LDpred2-auto, SBayesR, and PRS-CS-auto \cite[]{prive2019making,mak2017polygenic,prive2020ldpred2,lloyd2019improved,ge2019polygenic} by using either the true per-variant GWAS sample sizes, the total sample size for all variants, or per-variant sample sizes imputed using equation \eqref{eq:imputeN}.
+We then run C+T, lassosum, lassosum2 (Methods section \ref{lassosum2}), LDpred2-inf, LDpred2(-grid), LDpred2-auto{\color{red}, SBayesR (GCTB v2.03 with the automatic \texttt{robust} option), and PRS-CS-auto} \cite[]{prive2019making,mak2017polygenic,prive2020ldpred2,lloyd2019improved,ge2019polygenic} by using either the true per-variant GWAS sample sizes, the total sample size for all variants, or per-variant sample sizes imputed using equation \eqref{eq:imputeN}.
 When providing true per-variant GWAS sample sizes, squared correlations between the polygenic scores and the simulated phenotypes are of 0.123 for C+T, 0.160 for lassosum, 0.169 for lassosum2, 0.159 for LDpred2(-grid), 0.140 for LDpred2-auto, 0.141 for LDpred2-inf, 0.138 for SBayesR, and 0.159 for PRS-CS-auto (Figure \ref{fig:simu-misN}, averaging over 10 simulations).
@@
 While predictive performance are overall much lower, relative performance are less attenuated when using the maximum sample size, otherwise results are highly consistent as before (Figure \ref{fig:simu-misN-smallh2}).
-We also try using the shrunk LD matrix from GCTB that is used e.g.\ for SBayesR \cite[]{lloyd2019improved}; in these simulations, higher predictive performance are obtained from LDpred2-auto with this shrunk LD matrix, whereas SBayesR benefit from using LDpred2's windowed LD matrix (Figure \ref{fig:simu-misN-shrunk}).}
+We also try using the shrunk LD matrix from GCTB that is used e.g.\ for SBayesR \cite[]{lloyd2019improved}; in these simulations, higher predictive performance are obtained for LDpred2-auto with this shrunk LD matrix, whereas SBayesR benefit from using LDpred2's windowed LD matrix (Figure \ref{fig:simu-misN-shrunk}).}
 
@@
 Moreover, new additions for robustness introduced before, namely LDpred2-low-h2, LDpred2-auto-rob, and forming independent blocks in the LD matrix also improve predictive performance for all corrections (Figure \ref{fig:simu-info}).
-{\color{red}However, these corrections do not prove beneficial for PRS-CS-auto (Figure \ref{fig:simu-info}).
+{\color{red}However, these corrections do not prove beneficial for PRS-CS-auto (Figure \ref{fig:simu-info}).}
 
-When performing similar simulations with a heritability of 4\% (instead of 20\%), correction ``sqrt\_info'' still provides slightly higher predictive performance for all methods (compared to no correction), yet the other two corrections sometimes provide lower predictive performance (Figure \ref{fig:simu-info-smallh2}).}
+{\color{red}When performing similar simulations with a heritability of 4\% (instead of 20\%), correction ``sqrt\_info'' still provides slightly higher predictive performance for all methods (compared to no correction), yet the other two corrections sometimes provide lower predictive performance (Figure \ref{fig:simu-info-smallh2}).}
 In the real data applications hereinafter, we therefore choose to use the first correction, ``sqrt\_info'', which seems to more consistently provide better results, and is simple because it is somewhat equivalent to post-processing PGS effects by multiplying them by $\sqrt{\text{INFO}}$.
@@
 When using a strong shrinkage of 0.3 or 0.4, results for LDpred2-auto-rob are similar whatever the QC used (Figure \ref{fig:res_cad_shrinkage}), as previously observed for PRS-CS-auto.
-Results with LDpred2-auto-rob are always best when using ``qc2'', the most stringent QC; then no shrinkage (1) seems needed, while a strong shrinkage make predictive performance drop.
+Results with LDpred2-auto-rob are always best when using ``qc2'', the most stringent QC; then no shrinkage (1) seems needed, whereas a strong shrinkage make predictive performance drop.
 For phenotypes with relatively large genetic effects (e.g.\ vitamin D, Figure \ref{fig:res_vitaminD}), PRS-CS-auto perform much worse, possibly due to using too much regularization.}
@@
 Sixth, we recommend to form independent LD blocks in the 
-LD matrix as it proved to make the methods more robust (and slightly faster).}
+LD matrix as it proved to make the methods more robust (and slightly faster).
+Most of these new recommendations are included in the LDpred2 tutorial.}
 
@@
 
-\section*{Code and data availability}
-
-All code used for this paper is available at \url{https://github.com/privefl/paper-misspec/tree/master/code}.
-We have extensively used R packages bigstatsr and bigsnpr \cite[]{prive2017efficient} for analyzing large genetic data, packages from the future framework \cite[]{bengtsson2020unifying} for easy scheduling and parallelization of analyses on the HPC cluster, and packages from the tidyverse suite \cite[]{wickham2019welcome} for shaping and visualizing results.
-
-The latest version of R package bigsnpr can be installed from GitHub, and a recent enough version can be installed from CRAN.
-A tutorial on running LDpred2 and lassosum2 using R package bigsnpr is available at \url{https://privefl.github.io/bigsnpr/articles/LDpred2.html}.
-
-We have recomputed the allele frequencies and imputation INFO scores within the European subset used here and across all imputed variants in the UK Biobank, and made them available at \url{https://doi.org/10.6084/m9.figshare.16635388}.
-We have formed independent LD blocks within the Northwestern European LD references provided in \citet{prive2020ldpred2}, and made these updated versions available at \url{https://doi.org/10.6084/m9.figshare.19213299}.
-
 \section*{Acknowledgements}
@@
 
+\section*{Code and data availability}
+
+The UK Biobank data is available through a procedure described at \url{https://www.ukbiobank.ac.uk/using-the-resource/}. 
+All code used for this paper is available at \url{https://github.com/privefl/paper-misspec/tree/master/code}.
+We have extensively used R packages bigstatsr and bigsnpr \cite[]{prive2017efficient} for analyzing large genetic data, packages from the future framework \cite[]{bengtsson2020unifying} for easy scheduling and parallelization of analyses on the HPC cluster, and packages from the tidyverse suite \cite[]{wickham2019welcome} for shaping and visualizing results.
+The latest version of R package bigsnpr can be installed from GitHub, and a recent enough version can be installed from CRAN.
+
+\section*{Web resources}
+
+\noindent A tutorial on running LDpred2 and lassosum2 using R package bigsnpr is available at \url{https://privefl.github.io/bigsnpr/articles/LDpred2.html}.
+
+\noindent We have recomputed the allele frequencies and imputation INFO scores within the European subset used here and across all imputed variants in the UK Biobank, and made them available at \url{https://doi.org/10.6084/m9.figshare.16635388}.
+
+\noindent We have formed independent LD blocks within the Northwestern European LD references provided in \citet{prive2020ldpred2}, and made these updated versions available at \url{https://doi.org/10.6084/m9.figshare.19213299}.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2022-07-13 paper/paper-misspec.tex
minor text changes

@@
 Third, we introduce two new optional parameters in LDpred2-auto to make it more robust to these types of misspecification.
-{\color{red}Note that, in this paper, we use ``robust'' to mean ``similar predictive performance''. 
+{\color{red}Note that, in this paper, we use ``robust'' to mean that we obtain similar predictive performance to when there is no misspecification. 
 We also reimplement and use a new version of lassosum, called lassosum2, in which we better handle when per-variant sample sizes are different.}
@@
 Among genetic variants on chromosome 22 and with a minor allele frequency larger than 0.01 and an imputation INFO score larger than 0.4 (as reported by the UK Biobank), we sample 40,000 of them according to the inverse of the INFO score density so that they have varying levels of imputation accuracy (Figure \ref{fig:hist-info}).
-We read the UK Biobank data into two different datasets using function \texttt{snp\_readBGEN} from R package bigsnpr \cite[]{prive2017efficient}, one by reading the BGEN data as hard-called genotypes at random according to the imputation probabilities, and another one by reading it as dosages (i.e.\ expected values according to the imputation probabilities).
+We read the UK Biobank data into two different datasets using function \texttt{snp\_readBGEN} from R package bigsnpr \cite[]{prive2017efficient}, one by reading the BGEN data as hard-called genotypes by randomly sampling according to the imputation probabilities (of being 0, 1, or 2), and another one by reading it as dosages (i.e.\ expected values according to the imputation probabilities).
 The first dataset is used as what could be the real genotype calls and the second dataset as what would be its imputed version; this design technique was used in \citet{prive2019making}.
@@
 \begin{split}\label{eq:model}
- \hat{\gamma} | S,R,\gamma & \sim \mathcal{N}(S^{-1} R S \gamma, S^{-1} R S^{-1}) ~,\\
- S \hat{\gamma} | S,R,\gamma & \sim \mathcal{N}(R S \gamma, R) ~.
+ \hat{\gamma} | S,R,\gamma & \sim \mathcal{N}(S^{-1} R S \gamma, S^{-1} R S^{-1}) ~,\text{ or}, \\
+ S \hat{\gamma} | S,R,\gamma & \sim \mathcal{N}(R S \gamma, R) ~,
 \end{split}
@@
 Also note that the estimate of $\text{var}(G_j)$ has to account for the imputation accuracy, i.e.\ use $2 \cdot f_j \cdot (1 - f_j) \cdot \text{INFO}_j$ (Section \ref{sec:imputation}).
-{\color{red}In equations \eqref{eq:approx-sd-lin} and \eqref{eq:imputeN}, we estimate $\text{var}(y)$ by the first percentile (robust minimum) of $0.5 \left(N ~ \text{se}(\hat{\gamma}_j)^2 + \hat{\gamma}_j^2\right)$.}
+{\color{red}In equations \eqref{eq:approx-sd-lin} and \eqref{eq:imputeN}, we estimate $\text{var}(y)$ by the first percentile of $0.5 \left(N ~ \text{se}(\hat{\gamma}_j)^2 + \hat{\gamma}_j^2\right)$, where the first percentile approximates the minimum and is robust to outliers.}
 
@@
 Also note that, as in LDpred2, we run lassosum2 genome-wide using a sparse correlation matrix which assumes that variants further away than 3 cM are not correlated.
-Contrary to for lassosum, we do not require splitting the genome into independent LD blocks anymore, yet we recommend to do so for robustness in LDpred2, and for extra speed gain (cf.\ Methods section \ref{sec:LD}).
+Contrary to lassosum, we do not require splitting the genome into independent LD blocks anymore, yet we recommend to do so for robustness in LDpred2, and for extra speed gain (cf.\ Methods section \ref{sec:LD}).
 
@@
 We conduct further investigations to explain the results of figure \ref{fig:simu-misN}.
-{\color{red}First, the results for LDpred2-auto and SBayesR are similar to LDpred2-inf in these simulations because the internal $p$ estimate of LDpred2-auto (and similarly the proportion of null variants in SBayesR) always continuously increases up to $1$, which makes them behave as an infinitesimal model ($p = 1$).}
+{\color{red}First, the results for LDpred2-auto and SBayesR are similar to LDpred2-inf in these simulations because the internal polygenicity estimate $p$ (proportion of causal variants) of LDpred2-auto, and similarly the proportion of non-null variants in SBayesR, always continuously increases up to $1$ in the Gibbs sampler, which makes them behave as an infinitesimal model ($p = 1$).}
 To overcome this limitation in LDpred2-auto, we introduce two new parameters to make it more robust, and refer to this as ``LDpred2-auto-rob'' here (Methods section \ref{sec:robust}).
@@
 Here we call ``LDpred2-low-h2'' when running LDpred2(-grid) with a grid of hyper-parameters including these low values for $h^2$.
-Results with LDpred2-low-h2 improves from 0.159 to 0.169 when using true sample sizes and from 0.134 to 0.164 when using the maximum sample size.
+Results with LDpred2-low-h2 improves from 0.159 to 0.169 when using true sample sizes and from 0.134 to 0.164 when using the maximum sample size (Figure \ref{fig:simu-misN}).
 Finally, we introduce a last change for robustness here: we form independent LD blocks in the LD matrix to prevent small errors in the Gibbs sampler to propagate to too many variants (Methods section \ref{sec:LD}).
@@
 Interestingly, using the small Finnish LD reference panel we have defined here, composed of 503 individuals only, seems to almost always provide more predictive polygenic scores than using the large UK one composed of 10,000 individuals or the widely used European subset of the 1000 Genomes (1000G) data (Figure \ref{fig:finngen}).
-{\color{red}For PRS-CS-auto, no LD reference seems to consistently outperform the two others across all phenotypes considered.}
+{\color{red}For PRS-CS-auto, none of the three LD references seems to consistently outperform the other two across all phenotypes considered.}
 Note that the PGS here are validated in people of UK-like ancestry in the UK biobank to obtain sufficient sample sizes.
@@
 It would also be useful to investigate the impact of the QC, well-matched LD reference panels and other adjustments we propose here on other (non-PGS) methods, for which consistency and robustness are likely to be very important as well \cite[]{chen2021improved}.
-{\color{red}For example, we are interested in assessing the impact of these misspecifications on the inference of disease architecture parameters in some future work.}
+{\color{red}For example, we are interested in assessing the impact of these misspecifications on the inference of disease architecture parameters in future work.}
 

### 2022-07-15 paper/paper-misspec.tex
add info on LD blocks

@@
 \noindent$^{\text{\sf 2}}$Univ.\ Grenoble Alpes, Inria, CNRS, Grenoble INP, LJK, Grenoble, 38000, France. \\
-\noindent$^{\text{\sf 3}}$Department of Computational Biology, Institut Pasteur, Paris, 75015, France. \\
+\noindent$^{\text{\sf 3}}$Department of Computational Biology, Institut Pasteur,  Universit\'e Paris Cit\'e, Paris, 75015, France. \\
 \noindent$^{\text{\sf 4}}$Program in Genetic Epidemiology and Statistical Genetics, Harvard T.H. Chan School of Public Health, Boston, MA, 02115, USA. \\
@@
 {\color{red}Note that a new parameter \texttt{max\_r2} has been added to this method since, which controls the maximum single squared correlation allowed outside blocks, which offers an extra guarantee that the splitting is good and makes the function much faster by discarding lots of possible splits.
-For different numbers of blocks and maximum number of variants in each block, we use the split that minimizes $\big(C2 \cdot \sqrt{5 + C1}\big)$, where $C1$ is the sum of squared correlations outside the blocks, and $C2$ is the sum of squared sizes of the blocks (Figure \ref{fig:cost-split}).}
+For different numbers of blocks and maximum number of variants in each block, we use the split that minimizes $\big(C2 \cdot \sqrt{5 + C1}\big)$, where $C1$ is the sum of squared correlations outside the blocks, and $C2$ is the sum of squared sizes of the blocks (Figure \ref{fig:cost-split}).
+Twelve blocks from chromosome 22 are used in the simulations, and between 731 and 2527 in the real data analyses (cf.\ next section).}
 Having a correlation matrix with independent blocks prevents small errors in the algorithm (e.g.\ the Gibbs sampler in LDpred2) from propagating to too many variants.
@@
 We use either the 503 European (including 99 Finnish) individuals from the 1000 Genomes (1000G) data \cite[]{10002015global}, 404 Finnish-like UKBB + the 99 Finnish 1000G individuals (also 503 in total), or the 10,000 UK individuals from the validation set we use in this paper.
+{\color{red}1454, 731, and 1165 independent LD blocks are identified for these three LD references, respectively.}
 
@@
 We also construct versions of these LD references with independent LD blocks, as described in the previous section.
+{\color{red}2527, 1769, and 2196 independent LD blocks are identified for these three LD references, respectively.}
 

### 2022-07-20 code/investigate-CAD.R
reexport code

@@ -1,0 +1,136 @@
+library(bigsnpr)
+ukb_test <- snp_attach("data/UKBB_HM3_test.rds")
+G_test <- ukb_test$genotypes
+ind.test <- ukb_test$fam$id_csv
+
+library(dplyr)
+grid <- tidyr::expand_grid(
+  qc = c("noqc", "qc1", "qc2"),
+  name_corr = c("data/corr/chr", "data/corr/adj_with_blocks_chr")
+) %>%
+  print()
+
+library(future.batchtools)
+NCORES <- 15
+plan(batchtools_slurm(resources = list(
+  t = "12:00:00", c = NCORES + 1, mem = "100g",
+  name = basename(rstudioapi::getSourceEditorContext()$path))))
+
+bigassertr::assert_dir("results/cad")
+
+grid$res <- furrr::future_pmap(grid[1:2], function(qc, name_corr) {
+
+  gwas_file <- paste0("data/sumstats/cad_", qc, ".rds")
+  sumstats <- readRDS(gwas_file)
+  num_id <- sumstats[["_NUM_ID_"]]
+
+  basename <- file.path("results/cad", sub("\\.rds$", "", basename(gwas_file)))
+  if (grepl("blocks", name_corr, fixed = TRUE))
+    basename <- paste0(basename, "_adj_with_blocks")
+
+  library(bigsnpr)
+  CHR <- snp_attach("data/UKBB_HM3_val.rds")$map$chromosome
+
+  tmp <- tempfile(tmpdir = "tmp-data/sfbm/")
+  on.exit(file.remove(paste0(tmp, ".sbk")), add = TRUE)
+
+  for (chr in 1:22) {
+
+    print(chr)
+
+    ## indices in 'sumstats'
+    ind.chr <- which(sumstats$chr == chr)
+    ## indices in 'G'
+    ind.chr2 <- sumstats$`_NUM_ID_`[ind.chr]
+    ## indices in 'corr'
+    ind.chr3 <- match(ind.chr2, which(CHR == chr))
+
+    corr0 <- readRDS(paste0(name_corr, chr, ".rds"))
+    ld_chr <- Matrix::colSums(corr0^2)
+    df_beta_chr <- sumstats[ind.chr, c("beta", "beta_se", "n_eff")]
+
+    if (chr == 1) {
+      df_beta <- df_beta_chr
+      ld <- ld_chr[ind.chr3]
[…]

### 2022-07-20 code/investigate-lassosum2-auto.R

@@ -60,13 +60,13 @@ params$auto_score2 <- apply(beta_lassosum, 2, function(beta) {
 plot_grid(
   qplot(auto_score, score, color = as.factor(delta), data = params) +
     theme_bw(15) +
-    geom_abline(linetype = 2, color = "red") +
+    geom_abline(linetype = 2, color = "grey") +
     labs(x = "Score from pseudo-validation (using correlations)",
          y = "Score from validation", color = expression(delta)) +
     theme(legend.position = c(0.15, 0.7)),
   qplot(auto_score2, score, color = as.factor(delta), data = params) +
     theme_bw(15) +
-    geom_abline(linetype = 2, color = "red") +
+    geom_abline(linetype = 2, color = "grey") +
     labs(x = "Score from pseudo-validation (using p-values)",
          y = "Score from validation", color = expression(delta)) +
     theme(legend.position = c(0.8, 0.3)),

### 2022-07-20 code/investigate-misspec-N.R

@@ -69,12 +69,6 @@ abline(0, 1, col = "red", lwd = 3)
 
 #### Run lassosum2 and LDpred2-grid with different parameters ####
 
-library(future.batchtools)
-NCORES <- 15
-plan(batchtools_slurm(resources = list(
-  t = "12:00:00", c = NCORES + 1, mem = "100g",
-  name = basename(rstudioapi::getSourceEditorContext()$path))))
-
 bigassertr::assert_dir("investigate")
 
 grid <- tidyr::expand_grid(
@@ -83,7 +77,9 @@ grid <- tidyr::expand_grid(
   corr_num = 1:2
 )
 
-grid$res <- furrr::future_pmap(grid, function(method, whichN, corr_num) {
+grid$res <- purrr::pmap(grid, function(method, whichN, corr_num) {
+
+  cat(".")
 
   df <- list(maxN = df_beta2, imputeN = df_beta3)[[whichN]]
   corr <- list(corr1, corr2)[[corr_num]]
@@ -124,35 +120,33 @@ grid$res <- furrr::future_pmap(grid, function(method, whichN, corr_num) {
 
 
 filter(grid, method == "lassosum2") %>%
-  mutate(corr = c("normal", "with_indep_ld")[corr_num]) %>%
+  mutate(corr = c("normal", "with LD blocks")[corr_num]) %>%
   tidyr::unnest(res) %>%
   ggplot(aes(x = lambda, y = r2, color = paste(corr, whichN, sep = " - "))) +
   facet_wrap(~ delta, labeller = label_bquote(cols = delta: .(delta))) +
-  theme_bigstatsr(0.8) +
+  theme_bigstatsr(0.7) +
   geom_point(size = 2) +
-  geom_line(size = 1, alpha = 0.8) +
+  geom_line(size = 1, alpha = 0.5) +
   scale_x_log10(breaks = 10^(-5:0)) +
   scale_color_manual(values = c("#56B4E9", "#999999", "#009E73", "#E69F00")) +
   labs(x = expression(lambda), y = expression(r^2), color = "LD - N") +
   theme(legend.position = "top", legend.key.width = unit(2, "line"))
-
-# ggsave("figures/lassosum2-misN.pdf", width = 10, height = 6)
+# ggsave("figures/lassosum2-misN.pdf", width = 9, height = 6)
 
 
 filter(grid, method == "ldpred2") %>%
-  mutate(corr = c("normal", "with_indep_ld")[corr_num]) %>%
[…]

### 2022-07-20 code/prepare-corr-1000G-EAS.R

@@ -4,82 +4,94 @@ library(bigsnpr)
 
 obj.1000G <- snp_attach("data/1000G_HM3.rds")
 G <- obj.1000G$genotypes
+CHR <- obj.1000G$map$chromosome
 ind_EAS <- which(obj.1000G$fam$`Super Population` == "EAS")
 
 
 #### Prepare correlations ####
 
-CHR <- obj.1000G$map$chromosome
-POS2 <- snp_attach("data/UKBB_HM3_val.rds")$map$genetic.dist
-
 library(future.batchtools)
 plan(batchtools_slurm(resources = list(
-  t = "12:00:00", c = 15, mem = "125g",
+  t = "12:00:00", c = 15, mem = "60g",
   name = basename(rstudioapi::getSourceEditorContext()$path))))
 
 bigassertr::assert_dir("data/corr_1000G_EAS")
+bigassertr::assert_dir("data/corr_1000G_EAS/1kg_for_prscs")
 
 all_final_grp <- furrr::future_map_dfr(1:22, function(chr) {
 
   ind.chr <- which(CHR == chr)
+  MAF <- snp_MAF(G, ind.row = ind_EAS, ind.col = ind.chr, ncores = nb_cores())
+  keep <- (MAF > 0.02)
+  ind.chr2 <- ind.chr[keep]
 
   library(bigsnpr)
   corr0 <- runonce::save_run({
-    corr0 <- snp_cor(G, ind.row = ind_EAS, ind.col = ind.chr, infos.pos = POS2[ind.chr],
-                     size = 3 / 1000, ncores = nb_cores())
-    mean(is.na(corr0@x))  # 0.001 for chr 22
-    corr0@x <- ifelse(is.na(corr0@x) | is.infinite(corr0@x), 0, corr0@x)
-    corr0
+    POS2 <- snp_attach("data/UKBB_HM3_val.rds")$map$genetic.dist
+    snp_cor(G, ind.row = ind_EAS, ind.col = ind.chr2, infos.pos = POS2[ind.chr2],
+            size = 3 / 1000, ncores = nb_cores())
   }, file = paste0("data/corr_1000G_EAS/chr", chr, ".rds"))
 
-  all_splits <- runonce::save_run({
 
-    SEQ <- round(seq_log(1000 + ncol(corr0) / 50,
-                         5000 + ncol(corr0) / 10,
-                         length.out = 10))
+  # find nearly independent LD blocks
+  m <- length(ind.chr2)
[…]

### 2022-07-20 code/prepare-corr-1000G-EUR.R

@@ -4,81 +4,94 @@ library(bigsnpr)
 
 obj.1000G <- snp_attach("data/1000G_HM3.rds")
 G <- obj.1000G$genotypes
+CHR <- obj.1000G$map$chromosome
 ind_EUR <- which(obj.1000G$fam$`Super Population` == "EUR")
 
 
 #### Prepare correlations ####
 
-CHR <- obj.1000G$map$chromosome
-POS2 <- snp_attach("data/UKBB_HM3_val.rds")$map$genetic.dist
-
 library(future.batchtools)
 plan(batchtools_slurm(resources = list(
-  t = "12:00:00", c = 15, mem = "125g",
+  t = "12:00:00", c = 15, mem = "60g",
   name = basename(rstudioapi::getSourceEditorContext()$path))))
 
 bigassertr::assert_dir("data/corr_1000G_EUR")
+bigassertr::assert_dir("data/corr_1000G_EUR/1kg_for_prscs")
 
 all_final_grp <- furrr::future_map_dfr(1:22, function(chr) {
 
   ind.chr <- which(CHR == chr)
+  MAF <- snp_MAF(G, ind.row = ind_EUR, ind.col = ind.chr, ncores = nb_cores())
+  keep <- (MAF > 0.02)
+  ind.chr2 <- ind.chr[keep]
 
   library(bigsnpr)
   corr0 <- runonce::save_run({
-    corr0 <- snp_cor(G, ind.row = ind_EUR, ind.col = ind.chr, infos.pos = POS2[ind.chr],
-                     size = 3 / 1000, ncores = nb_cores())
-    mean(is.na(corr0@x))  # 0.001 for chr 22
-    corr0@x <- ifelse(is.na(corr0@x) | is.infinite(corr0@x), 0, corr0@x)
-    corr0
+    POS2 <- snp_attach("data/UKBB_HM3_val.rds")$map$genetic.dist
+    snp_cor(G, ind.row = ind_EUR, ind.col = ind.chr2, infos.pos = POS2[ind.chr2],
+            size = 3 / 1000, ncores = nb_cores())
   }, file = paste0("data/corr_1000G_EUR/chr", chr, ".rds"))
 
-  library(furrr)
-  plan("multisession", workers = 5)
-  options(future.globals.maxSize = 10e9)
 
-  all_splits <- runonce::save_run({
-    SEQ <- round(seq_log(1000 + ncol(corr0) / 50,
-                         5000 + ncol(corr0) / 10,
[…]

### 2022-07-20 code/prepare-corr-FIN.R

@@ -79,75 +79,90 @@ plot(af_1000G, af_UKBB)  # ok
 #### Prepare correlations ####
 
 CHR <- as.integer(obj.bigsnp$map$chromosome)
-POS2 <- snp_attach("data/UKBB_HM3_val.rds")$map$genetic.dist
 
 library(future.batchtools)
 plan(batchtools_slurm(resources = list(
-  t = "12:00:00", c = 15, mem = "125g",
+  t = "12:00:00", c = 15, mem = "60g",
   name = basename(rstudioapi::getSourceEditorContext()$path))))
 
 bigassertr::assert_dir("data/corr_FIN")
+bigassertr::assert_dir("data/corr_FIN/ukbb_for_prscs")
 
 all_final_grp <- furrr::future_map_dfr(1:22, function(chr) {
 
   ind.chr <- which(CHR == chr)
+  MAF <- snp_MAF(G, ind.col = ind.chr, ncores = nb_cores())
+  keep <- (MAF > 0.02)
+  ind.chr2 <- ind.chr[keep]
 
   library(bigsnpr)
   corr0 <- runonce::save_run({
-    corr0 <- snp_cor(G, ind.col = ind.chr, infos.pos = POS2[ind.chr],
-                     size = 3 / 1000, ncores = nb_cores())
-    mean(is.na(corr0@x))  # 0.001 for chr 22
-    corr0@x <- ifelse(is.na(corr0@x) | is.infinite(corr0@x), 0, corr0@x)
-    corr0
+    POS2 <- snp_attach("data/UKBB_HM3_val.rds")$map$genetic.dist
+    snp_cor(G, ind.col = ind.chr2, infos.pos = POS2[ind.chr2],
+            size = 3 / 1000, ncores = nb_cores())
   }, file = paste0("data/corr_FIN/chr", chr, ".rds"))
 
-  library(furrr)
-  plan("multisession", workers = 5)
-  options(future.globals.maxSize = 10e9)
 
-  all_splits <- runonce::save_run({
-    SEQ <- round(seq_log(1000 + ncol(corr0) / 50,
-                         5000 + ncol(corr0) / 10,
-                         length.out = 10))
-
-    splits <- future_map_dfr(SEQ, function(max_size) {
-      res <- snp_ldsplit(corr0, thr_r2 = 0.05, min_size = 100,
-                         max_size = max_size, max_K = 200)
[…]

### 2022-07-20 code/prepare-corr-JPN.R

@@ -77,76 +77,90 @@ plot(af_1000G, af_UKBB)  # ok
 #### Prepare correlations ####
 
 CHR <- as.integer(obj.bigsnp$map$chromosome)
-POS2 <- snp_attach("data/UKBB_HM3_val.rds")$map$genetic.dist
 
 library(future.batchtools)
 plan(batchtools_slurm(resources = list(
-  t = "12:00:00", c = 15, mem = "125g",
+  t = "12:00:00", c = 15, mem = "60g",
   name = basename(rstudioapi::getSourceEditorContext()$path))))
 
 bigassertr::assert_dir("data/corr_JPN")
+bigassertr::assert_dir("data/corr_JPN/ukbb_for_prscs")
 
 all_final_grp <- furrr::future_map_dfr(1:22, function(chr) {
 
   ind.chr <- which(CHR == chr)
+  MAF <- snp_MAF(G, ind.col = ind.chr, ncores = nb_cores())
+  keep <- (MAF > 0.02)
+  ind.chr2 <- ind.chr[keep]
 
   library(bigsnpr)
   corr0 <- runonce::save_run({
-    corr0 <- snp_cor(G, ind.col = ind.chr, infos.pos = POS2[ind.chr],
-                     size = 3 / 1000, ncores = nb_cores())
-    mean(is.na(corr0@x))  # 0.001 for chr 22
-    corr0@x <- ifelse(is.na(corr0@x) | is.infinite(corr0@x), 0, corr0@x)
-    corr0
+    POS2 <- snp_attach("data/UKBB_HM3_val.rds")$map$genetic.dist
+    snp_cor(G, ind.col = ind.chr2, infos.pos = POS2[ind.chr2],
+            size = 3 / 1000, ncores = nb_cores())
   }, file = paste0("data/corr_JPN/chr", chr, ".rds"))
 
-  all_splits <- runonce::save_run({
 
-    SEQ <- round(seq_log(1000 + ncol(corr0) / 50,
-                         5000 + ncol(corr0) / 10,
-                         length.out = 10))
-
-    library(furrr)
-    plan("multisession", workers = 5)
-    options(future.globals.maxSize = 10e9)
-
-    splits <- future_map_dfr(SEQ, function(max_size) {
-      res <- snp_ldsplit(corr0, thr_r2 = 0.05, min_size = 100,
-                         max_size = max_size, max_K = 200)
[…]

### 2022-07-20 code/prepare-corr-UKBB-EAS.R

@@ -85,76 +85,90 @@ plot(af_EAS, af_JPN)  # ok
 #### Prepare correlations ####
 
 CHR <- as.integer(obj.bigsnp$map$chromosome)
-POS2 <- snp_attach("data/UKBB_HM3_val.rds")$map$genetic.dist
 
 library(future.batchtools)
 plan(batchtools_slurm(resources = list(
-  t = "12:00:00", c = 15, mem = "125g",
+  t = "12:00:00", c = 15, mem = "60g",
   name = basename(rstudioapi::getSourceEditorContext()$path))))
 
 bigassertr::assert_dir("data/corr_UKBB_EAS")
+bigassertr::assert_dir("data/corr_UKBB_EAS/ukbb_for_prscs")
 
 all_final_grp <- furrr::future_map_dfr(1:22, function(chr) {
 
   ind.chr <- which(CHR == chr)
+  MAF <- snp_MAF(G, ind.col = ind.chr, ncores = nb_cores())
+  keep <- (MAF > 0.01)
+  ind.chr2 <- ind.chr[keep]
 
   library(bigsnpr)
   corr0 <- runonce::save_run({
-    corr0 <- snp_cor(G, ind.col = ind.chr, infos.pos = POS2[ind.chr],
-                     size = 3 / 1000, ncores = nb_cores())
-    mean(is.na(corr0@x))  # 0.001 for chr 22
-    corr0@x <- ifelse(is.na(corr0@x) | is.infinite(corr0@x), 0, corr0@x)
-    corr0
+    POS2 <- snp_attach("data/UKBB_HM3_val.rds")$map$genetic.dist
+    snp_cor(G, ind.col = ind.chr2, infos.pos = POS2[ind.chr2],
+            size = 3 / 1000, ncores = nb_cores())
   }, file = paste0("data/corr_UKBB_EAS/chr", chr, ".rds"))
 
-  all_splits <- runonce::save_run({
 
-    SEQ <- round(seq_log(1000 + ncol(corr0) / 50,
-                         5000 + ncol(corr0) / 10,
-                         length.out = 10))
-
-    library(furrr)
-    plan("multisession", workers = 5)
-    options(future.globals.maxSize = 10e9)
-
-    splits <- future_map_dfr(SEQ, function(max_size) {
-      res <- snp_ldsplit(corr0, thr_r2 = 0.05, min_size = 100,
[…]

### 2022-07-20 code/prepare-corr-altpop.R

@@ -1,108 +1,0 @@
-library(dplyr)
-library(ggplot2)
-library(bigsnpr)
-NCORES <- nb_cores()
-
-snp_id <- with(snp_attach("data/ukbb4simu.rds")$map,
-               paste(chromosome, physical.pos, allele1, allele2, sep = "_"))
-
-df0 <- bigreadr::fread2(
-  "UKBB/ukb41181.csv",
-  select    = c("eid", paste0("22009-0.", 1:16)),
-  col.names = c("eid", paste0("PC", 1:16))
-)
-
-# still in data
-sample <- bigreadr::fread2("UKBB/ukb58024_imp_chr1_v3_s487296.sample")[-1, ]
-ind.indiv <- match(df0$eid, sample$ID_2)
-id_to_rm <- scan("../../UKBiobank/ukb_withdrawn_individuals_8Feb2021.id")
-ind.indiv[df0$eid %in% id_to_rm] <- NA
-sub <- which(!is.na(ind.indiv))
-
-# filter for relatedness
-rel <- bigreadr::fread2("UKBB/ukb58024_rel_s488264.dat")
-rel2 <- dplyr::filter(rel, Kinship > 2^-3.5)
-ind.indiv[df0$eid %in% rel2$ID2] <- NA
-sub <- which(!is.na(ind.indiv))
-
-# Closer to South Europe (than UK)
-PC_UKBB <- select(df0, PC1:PC16)[sub, ]
-all_centers <- read.csv(
-  "https://raw.githubusercontent.com/privefl/UKBB-PGS/main/pop_centers.csv",
-  stringsAsFactors = FALSE)
-center_italy <- unlist(all_centers[4, -1])
-all_sq_dist <- rowSums(sweep(PC_UKBB, 2, center_italy, '-')^2)
-closest_ind <- head(order(all_sq_dist), 10e3)
-hist(log(all_sq_dist), "FD", xlim = c(6, 9)); abline(v = log(all_sq_dist[tail(closest_ind, 1)]), col = "red")
-sub2 <- sub[closest_ind]
-length(sub2) # 10,000
-eid_UK <- df0$eid[sub2]
-
-# Read the data
-snp_readBGEN(
-  bgenfiles = "UKBB/bgen/ukb_imp_chr22_v3.bgen",
-  list_snp_id = list(snp_id),
-  backingfile = "data/ukbb4simu_altpop",
[…]

### 2022-07-20 code/prepare-corr-simu-altpop.R

@@ -1,0 +1,211 @@
+library(dplyr)
+library(ggplot2)
+library(bigsnpr)
+library(bigreadr)
+NCORES <- nb_cores()
+
+snp_id <- with(snp_attach("data/ukbb4simu.rds")$map,
+               paste(chromosome, physical.pos, allele1, allele2, sep = "_"))
+
+code_country <- filter(fread2("UKBB/coding89.tsv"), selectable == "Y")
+df0 <- fread2(
+  "UKBB/ukb41181.csv",
+  select    = c("eid", "20115-0.0", paste0("22009-0.", 1:16)),
+  col.names = c("eid", "country",   paste0("PC", 1:16))
+) %>%
+  mutate(country = factor(country, levels = code_country$coding,
+                          labels = code_country$meaning))
+
+# still in data
+sample <- fread2("UKBB/ukb58024_imp_chr1_v3_s487296.sample")[-1, ]
+ind.indiv <- match(df0$eid, sample$ID_2)
+id_to_rm <- scan("../../UKBiobank/ukb_withdrawn_individuals_8Feb2021.id")
+ind.indiv[df0$eid %in% id_to_rm] <- NA
+sub <- which(!is.na(ind.indiv))
+
+# filter for relatedness
+rel <- fread2("UKBB/ukb58024_rel_s488264.dat")
+rel2 <- dplyr::filter(rel, Kinship > 2^-3.5)
+ind.indiv[df0$eid %in% rel2$ID2] <- NA
+sub <- which(!is.na(ind.indiv))
+
+# Closer to South Europe (than UK)
+PC_UKBB <- select(df0, PC1:PC16)[sub, ]
+all_centers <- read.csv(
+  "https://raw.githubusercontent.com/privefl/UKBB-PGS/main/pop_centers.csv",
+  stringsAsFactors = FALSE)
+center_italy <- unlist(all_centers[all_centers$Ancestry == "Italy", -1])
+all_sq_dist <- rowSums(sweep(PC_UKBB, 2, center_italy, '-')^2)
+closest_ind <- head(order(all_sq_dist), 10e3)
+hist(log(all_sq_dist), "FD", xlim = c(6, 9)); abline(v = log(all_sq_dist[tail(closest_ind, 1)]), col = "red")
+sub2 <- sub[closest_ind]
+length(sub2) # 10,000
+eid_UK <- df0$eid[sub2]
+
[…]

### 2022-07-20 code/prepare-corr-simu-chr22.R

@@ -12,42 +12,162 @@ load("data/ukbb4simu_ind.RData")
 corr0 <- runonce::save_run(
   snp_cor(G, ind.row = ind.val, infos.pos = POS2, size = 3 / 1000, ncores = nb_cores()),
   file = "tmp-data/corr0_simu_val.rds")
+print(range(corr0@x), digits = 22)
 
-corr <- runonce::save_run(as_SFBM(corr0, "tmp-data/corr_simu_val", compact = TRUE),
-                          file = "tmp-data/corr_simu_val.rds")
+corr <- as_SFBM(corr0, "tmp-data/corr_simu_val", compact = TRUE)$save()
 
 
 #### Further restrict to nearly-independent blocks ####
 
-splits <- purrr::map_dfr(3:8 * 1000, function(max_size) {
-  res <- snp_ldsplit(corr0, thr_r2 = 0.02, min_size = 50,
-                     max_size = max_size, max_K = 50)
-  res$max_size <- max_size
-  res
-})
+# /!\ need v1.10.1
+splits <- snp_ldsplit(corr0, thr_r2 = 0.02, min_size = 50, max_size = 3:10 * 1000,
+                      max_K = 50, max_r2 = 0.2, max_cost = 100)
 
-library(ggplot2)
-qplot(data = splits, perc_kept, cost, color = as.factor(max_size)) +
-  theme_bw(12) +
-  theme(legend.position = "top") + guides(colour = guide_legend(nrow = 1)) +
-  labs(y = "Sum of squared correlations outside blocks (log-scale)",
-       x = "% of non-zero values kept", color = "Maximum block size") +
-  ylim(0, median(splits$cost)) +
-  geom_vline(xintercept = 0.6, linetype = 3)
-# ggsave("figures/cost_split_chr22.pdf", width = 8, height = 6)
+splits$cost2 <- sapply(splits$all_size, function(sizes) sum(sizes^2))
 
 library(dplyr)
-final_grp <- splits %>%
-  filter(perc_kept < 0.6) %>%
-  arrange(cost) %>%
-  slice(1) %>%
+best_split <- splits %>%
+  arrange(cost2 * sqrt(10 + cost)) %>%
   print() %>%
-  pull(block_num) %>%
-  unlist()
+  slice(1) %>%
+  print()
[…]

### 2022-07-20 code/prepare-corr-with-blocks.R

@@ -1,61 +1,0 @@
-library(future.batchtools)
-plan(batchtools_slurm(resources = list(
-  t = "12:00:00", c = 2, mem = "50g",
-  name = basename(rstudioapi::getSourceEditorContext()$path))))
-
-all_final_grp <- furrr::future_map_dfr(1:22, function(chr) {
-
-  corr0 <- readRDS(paste0("data/corr/chr", chr, ".rds"))
-  dim(corr0)
-
-  library(bigsnpr)
-  all_splits <- runonce::save_run({
-    SEQ <- round(seq_log(1000 + ncol(corr0) / 50,
-                         5000 + ncol(corr0) / 10,
-                         length.out = 10))
-
-    splits <- purrr::map_dfr(SEQ, function(max_size) {
-      res <- snp_ldsplit(corr0, thr_r2 = 0.02, min_size = 100,
-                         max_size = max_size, max_K = 200)
-      res$max_size <- max_size
-      res
-    })
-  }, file = paste0("tmp-data/split_corr_chr", chr, ".rds"))
-
-
-  library(ggplot2)
-  qplot(data = all_splits, perc_kept, cost, color = as.factor(max_size)) +
-    theme_bw(12) +
-    theme(legend.position = "top") + guides(colour = guide_legend(nrow = 1)) +
-    labs(y = "Sum of squared correlations outside blocks (log-scale)",
-         x = "% of non-zero values kept", color = "Maximum block size") +
-    ylim(0, min(quantile(all_splits$cost, 0.6), 200)) +
-    geom_vline(xintercept = 0.6, linetype = 3) +
-    scale_x_continuous(breaks = seq(0, 1, by = 0.2))
-
-
-  library(dplyr)
-  final_split <- all_splits %>%
-    filter(perc_kept < `if`(chr == 6, 0.7, 0.6)) %>%
-    arrange(cost) %>%
-    slice(1)
-
-  final_grp <- final_split$block_num[[1]]
-
-  corr0T <- as(corr0, "dgTMatrix")
-  corr0T@x <- ifelse(final_grp[corr0T@i + 1L] == final_grp[corr0T@j + 1L], corr0T@x, 0)
-
-  corr0T %>%
-    Matrix::drop0() %>%
-    as("symmetricMatrix") %>%
-    saveRDS(paste0("data/corr/adj_with_blocks_chr", chr, ".rds"))
-
-  final_split
-})
-
[…]

### 2022-07-20 code/prepare-corr.R

@@ -1,20 +1,109 @@
 library(bigsnpr)
+library(dplyr)
 ukb <- snp_attach("data/UKBB_HM3_val.rds")
-G <- ukb$genotypes
-CHR <- ukb$map$chromosome
-POS2 <- ukb$map$genetic.dist
+G     <- ukb$genotypes
+CHR   <- ukb$map$chromosome
+POS2  <- ukb$map$genetic.dist
 
-(NCORES <- nb_cores())
+library(future.batchtools)
+NCORES <- 14
+plan(batchtools_slurm(resources = list(
+  t = "12:00:00", c = NCORES + 2, mem = "50g",
+  name = basename(rstudioapi::getSourceEditorContext()$path))))
 
 bigassertr::assert_dir("data/corr")
+bigassertr::assert_dir("data/corr/ukbb_for_prscs")
 
-for (chr in 1:22) {
+all_final_grp <- furrr::future_map_dfr(1:22, function(chr) {
 
   ind.chr <- which(CHR == chr)
 
-  corr <- runonce::save_run(
+
+  # LD for LDpred2 and lassosum2
+  corr0 <- runonce::save_run(
     snp_cor(G, ind.col = ind.chr, infos.pos = POS2[ind.chr],
             size = 3 / 1000, ncores = NCORES),
     file = paste0("data/corr/chr", chr, ".rds")
   )
-}
+
+
+  # find nearly independent LD blocks
+  m <- ncol(corr0)
+  (SEQ <- round(seq_log(m / 30, m / 5, length.out = 20)))
+  splits <- snp_ldsplit(corr0, thr_r2 = 0.02, min_size = 50, max_size = SEQ, max_r2 = 0.1)
+  splits$cost2 <- sapply(splits$all_size, function(sizes) sum(sizes^2))
+
+  best_split <- splits %>%
+    arrange(cost2 * sqrt(10 + cost)) %>%
+    print() %>%
+    slice(1) %>%
+    print()
+
+  library(ggplot2)
+  plot_grid(
+    qplot(data = splits, perc_kept, cost, color = as.factor(max_size)) +
+      geom_point(data = best_split, size = 2, color = "black") +
+      theme_bw(12) +
+      theme(legend.position = "top") + guides(colour = guide_legend(nrow = 1)) +
+      scale_x_continuous(limits = c(0, NA), breaks = seq(0, 1, by = 0.1)) +
+      labs(y = "Sum of squared correlations outside blocks",
[…]

### 2022-07-20 code/prepare-ldref-with-blocks.R

@@ -1,69 +1,59 @@
-library(future.batchtools)
-plan(batchtools_slurm(resources = list(
-  t = "12:00:00", c = 6, mem = "125g",
-  name = basename(rstudioapi::getSourceEditorContext()$path))))
+library(bigsnpr)
+library(furrr)
+plan("multisession", workers = 6)
 
 bigassertr::assert_dir("ldref")
 
-all_final_grp <- furrr::future_map_dfr(1:22, function(chr) {
+all_final_grp <- future_map_dfr(1:22, function(chr) {
 
   corr0 <- readRDS(paste0("../paper-ldpred2/ld-ref/LD_chr", chr, ".rds"))
   dim(corr0)
 
-  library(bigsnpr)
-  library(furrr)
-  all_splits <- runonce::save_run({
-
-    plan("multisession", workers = 5)
-    options(future.globals.maxSize = 10e9)
-
-    SEQ <- round(seq_log(1000 + ncol(corr0) / 50,
-                         5000 + ncol(corr0) / 10,
-                         length.out = 10))
-
-    splits <- future_map_dfr(SEQ, function(max_size) {
-      res <- snp_ldsplit(corr0, thr_r2 = 0.02, min_size = 100,
-                         max_size = max_size, max_K = 200)
-      res$max_size <- max_size
-      res
-    })
-  }, file = paste0("tmp-data/split_ldref_chr", chr, ".rds"))
-
-
-  library(ggplot2)
-  qplot(data = all_splits, perc_kept, cost, color = as.factor(max_size)) +
-    theme_bw(12) +
-    theme(legend.position = "top") + guides(colour = guide_legend(nrow = 1)) +
-    labs(y = "Sum of squared correlations outside blocks (log-scale)",
-         x = "% of non-zero values kept", color = "Maximum block size") +
-    ylim(0, min(quantile(all_splits$cost, 0.6), 200)) +
-    geom_vline(xintercept = 0.6, linetype = 3) +
-    scale_x_continuous(breaks = seq(0, 1, by = 0.2))
-
+  # find nearly independent LD blocks
+  m <- ncol(corr0)
+  (SEQ <- round(seq_log(m / 30, m / 5, length.out = 20)))
[…]

### 2022-07-20 code/process-PRS-BBJ.R

@@ -18,7 +18,7 @@ ind.val <- which(ukb$fam$set == "val")
 covar.val <- as.matrix(df0[id_csv[ind.val], -1])
 
 ALL_METHODS <- c("lassosum2", "LDpred2", "LDpred2-low-h2",
-                 "LDpred2-auto", "LDpred2-auto-rob")
+                 "LDpred2-auto", "LDpred2-auto-rob", "PRS-CS-auto")
 
 ALL_CORR <- c(
   corr_JPN               = "data/corr_JPN/chr",
@@ -29,7 +29,7 @@ ALL_CORR <- c(
   corr_UKBB_with_blocks  = "data/corr_UKBB_EAS/adj_with_blocks_chr"
 )
 
-bigassertr::assert_dir("results-final_BBJ")
+bigassertr::assert_dir("results-final/sumstats_BBJ")
 
 library(dplyr)
 grid <- tibble(pheno = c("height", "systolic_bp", "hdl_cholesterol", "bmi")) %>%
@@ -46,31 +46,41 @@ grid$effects <- furrr::future_pmap(grid[1:3], function(pheno, name_corr, method)
   # name_corr <- "corr_1000G"
   # method <- "lassosum2"
 
-  res_file <- paste0("results_BBJ/", pheno, "_", name_corr, "_", method, ".rds")
-  if (!file.exists(res_file)) {
-    print(glue::glue("/!\\ '{res_file}' IS MISSING /!\\"))
-    return(rep(NA, ncol(G)))
+  if (method != "PRS-CS-auto") {
+    res_file <- paste0("results/sumstats_BBJ/", pheno, "_", name_corr, "_", method, ".rds")
+    if (!file.exists(res_file)) {
+      print(glue::glue("/!\\ '{res_file}' IS MISSING /!\\"))
+      return(rep(NA, ncol(G)))
+    }
+    res <- readRDS(res_file)
+  } else {
+    res_files <- paste0("results_prscs/sumstats_BBJ/", pheno, "_",
+                        sub("_with_blocks", "", name_corr), "_chr", 1:22, ".rds")
+    for (res_file in res_files) {
+      if (!file.exists(res_file)) {
+        print(glue::glue("/!\\ '{res_file}' IS MISSING /!\\"))
+        return(rep(NA, ncol(G)))
+      }
+    }
+    res <- unlist(lapply(res_files, function(res_file) readRDS(res_file)$V6))
   }
 
-  res_file2 <- paste0("results-final_BBJ/", basename(res_file))
[…]

### 2022-07-20 code/process-PRS-FIN.R

@@ -17,7 +17,7 @@ ind.val <- ukb$fam$id_csv
 covar <- as.matrix(df0[ind.val, -1])
 
 ALL_METHODS <- c("lassosum2", "LDpred2", "LDpred2-low-h2",
-                 "LDpred2-auto", "LDpred2-auto-rob")
+                 "LDpred2-auto", "LDpred2-auto-rob", "PRS-CS-auto")
 ALL_CORR <- c(
   corr_UK             = "data/corr/chr",
   corr_UK_with_blocks = "data/corr/adj_with_blocks_chr",
@@ -27,7 +27,7 @@ ALL_CORR <- c(
   corr_1000G_with_blocks = "data/corr_1000G_EUR/adj_with_blocks_chr"
 )
 
-bigassertr::assert_dir("results-final_FIN")
+bigassertr::assert_dir("results-final/sumstats_FIN")
 
 library(dplyr)
 grid <- tibble::tribble(
@@ -50,33 +50,43 @@ grid$effects <- furrr::future_pmap(grid[1:4], function(pheno, phecode, name_corr
   # pheno <- "t2d"
   # phecode <- "250.2"
   # name_corr <- "corr_1000G"
-  # method <- "lassosum2"
+  # method <- "PRS-CS-auto"
 
-  res_file <- paste0("results_FIN/", pheno, "_", name_corr, "_", method, ".rds")
-  if (!file.exists(res_file)) {
-    print(glue::glue("/!\\ '{res_file}' IS MISSING /!\\"))
-    return(rep(NA, ncol(G)))
+  if (method != "PRS-CS-auto") {
+    res_file <- paste0("results/sumstats_FIN/", pheno, "_", name_corr, "_", method, ".rds")
+    if (!file.exists(res_file)) {
+      print(glue::glue("/!\\ '{res_file}' IS MISSING /!\\"))
+      return(rep(NA, ncol(G)))
+    }
+    res <- readRDS(res_file)
+  } else {
+    res_files <- paste0("results_prscs/sumstats_FIN/", pheno, "_",
+                        sub("_with_blocks", "", name_corr), "_chr", 1:22, ".rds")
+    for (res_file in res_files) {
+      if (!file.exists(res_file)) {
+        print(glue::glue("/!\\ '{res_file}' IS MISSING /!\\"))
+        return(rep(NA, ncol(G)))
+      }
+    }
+    res <- unlist(lapply(res_files, function(res_file) readRDS(res_file)$V6))
   }
 
[…]

### 2022-07-20 code/process-PRS.R

@@ -17,9 +17,10 @@ ind.val <- ukb$fam$id_csv
 covar <- as.matrix(df0[ind.val, -1])
 
 ALL_METHODS <- c("lassosum2", "LDpred2", "LDpred2-low-h2",
-                 "LDpred2-auto", "LDpred2-auto-rob")
+                 "LDpred2-auto", "LDpred2-auto-rob", "PRS-CS-auto")
 
 bigassertr::assert_dir("results-final")
+bigassertr::assert_dir("results-final/sumstats")
 
 library(dplyr)
 grid <- tibble::tribble(
@@ -38,8 +39,10 @@ grid <- tibble::tribble(
     qc = c("noqc", "qc1", "qc2"),
     name_corr = c("data/corr/chr", "data/corr/adj_with_blocks_chr"),
     method = ALL_METHODS) %>%
+  mutate(name_corr = ifelse(method == "PRS-CS-auto", NA_character_, name_corr)) %>%
+  unique() %>%
   print()
-# 480 different PGS models
+# 528 different PGS models
 
 
 library(future.apply)
@@ -53,43 +56,51 @@ grid$effects <- furrr::future_pmap(grid[1:7], function(
   # qc <- "qc1"
   # use_info <- TRUE
   # name_corr <- "data/corr/adj_with_blocks_chr"
-  # method <- "lassosum2"
-
+  # method <- "PRS-CS-auto"
 
   basename <- paste0(pheno, "_", qc)
   sumstats_file <- paste0("data/sumstats/", basename, ".rds")
 
-  suffix_N <- `if`(is.na(whichN), "", paste0("_", whichN))
-  basename <- paste0("results/", basename, suffix_N)
-  if (grepl("blocks", name_corr, fixed = TRUE))
-    basename <- paste0(basename, "_adj_with_blocks")
+  if (method != "PRS-CS-auto") {
 
-  res_file <- paste0(basename, "_", method, ".rds")
-  if (!file.exists(res_file)) {
-    print(glue::glue("/!\\ '{res_file}' IS MISSING /!\\"))
-    return(rep(NA, ncol(G)))
+    suffix_N <- `if`(is.na(whichN), "", paste0("_", whichN))
+    basename <- paste0("results/sumstats/", basename, suffix_N)
+    if (grepl("blocks", name_corr, fixed = TRUE))
+      basename <- paste0(basename, "_adj_with_blocks")
+
+    res_file <- paste0(basename, "_", method, ".rds")
[…]

### 2022-07-20 code/run-PRS-BBJ.R

@@ -1,7 +1,3 @@
-
-bigassertr::assert_dir("results_BBJ")
-bigassertr::assert_dir("log")
-
 ALL_CORR <- c(
   corr_JPN               = "data/corr_JPN/chr",
   corr_JPN_with_blocks   = "data/corr_JPN/adj_with_blocks_chr",
@@ -12,28 +8,35 @@ ALL_CORR <- c(
 )
 
 library(dplyr)
-files <- tibble(pheno = c("height", "systolic_bp", "hdl_cholesterol", "bmi")) %>%
+grid <- tibble(pheno = c("height", "systolic_bp", "hdl_cholesterol", "bmi")) %>%
   tidyr::expand_grid(name_corr = names(ALL_CORR)) %>%
-  filter(!file.exists(paste0("results_BBJ/", pheno, "_", name_corr,
+  filter(!file.exists(paste0("results/sumstats_BBJ/", pheno, "_", name_corr,
                              "_LDpred2-auto-rob.rds"))) %>%
   print()
 
 library(future.batchtools)
-NCORES <- 30
+NCORES <- 14
 plan(batchtools_slurm(resources = list(
-  t = "12:00:00", c = NCORES + 2, mem = "120g",
+  t = "12:00:00", c = NCORES + 2, mem = "60g",
   name = basename(rstudioapi::getSourceEditorContext()$path))))
 
-furrr::future_pwalk(files, function(pheno, qc, name_corr) {
+bigassertr::assert_dir("results/sumstats_BBJ")
 
-  # pheno <- "height"
-  # name_corr <- "corr_JPN"
-
-  sumstats <- readRDS(paste0("data/sumstats_BBJ/", pheno, ".rds"))
-  basename <- paste0("results_BBJ/", pheno, "_", name_corr)
+furrr::future_pwalk(grid, function(pheno, qc, name_corr) {
 
   library(bigsnpr)
-  CHR <- snp_attach("data/UKBB_HM3_val.rds")$map$chromosome
+  corr_path <- ALL_CORR[[name_corr]]
+  grp <- readRDS(file.path(dirname(corr_path), "all_final_grp.rds"))
+  if (is.null(grp$ind)) {
+    CHR <- snp_attach("data/UKBB_HM3_val.rds")$map$chromosome
+    ind_keep <- split(seq_along(CHR), CHR)
+  } else {
[…]

### 2022-07-20 code/run-PRS-FIN.R

@@ -1,7 +1,3 @@
-
-bigassertr::assert_dir("results_FIN")
-bigassertr::assert_dir("log")
-
 ALL_CORR <- c(
   corr_UK             = "data/corr/chr",
   corr_UK_with_blocks = "data/corr/adj_with_blocks_chr",
@@ -12,28 +8,39 @@ ALL_CORR <- c(
 )
 
 library(dplyr)
-files <- tibble(pheno = c("brca", "prca", "cad", "t2d", "t1d")) %>%
+grid <- tibble(pheno = c("brca", "prca", "cad", "t2d", "t1d")) %>%
   tidyr::expand_grid(name_corr = names(ALL_CORR)) %>%
-  filter(!file.exists(paste0("results_FIN/", pheno, "_", name_corr,
+  filter(!file.exists(paste0("results/sumstats_FIN/", pheno, "_", name_corr,
                              "_LDpred2-auto-rob.rds"))) %>%
   print()
 
 library(future.batchtools)
-NCORES <- 30
+NCORES <- 14
 plan(batchtools_slurm(resources = list(
-  t = "12:00:00", c = NCORES + 2, mem = "120g",
+  t = "12:00:00", c = NCORES + 2, mem = "60g",
   name = basename(rstudioapi::getSourceEditorContext()$path))))
 
-furrr::future_pwalk(files, function(pheno, qc, name_corr) {
+bigassertr::assert_dir("results/sumstats_FIN")
+
+furrr::future_pwalk(grid, function(pheno, qc, name_corr) {
 
   # pheno <- "brca"
   # name_corr <- "corr_FIN"
-
-  sumstats <- readRDS(paste0("data/sumstats_FIN/", pheno, ".rds"))
-  basename <- paste0("results_FIN/", pheno, "_", name_corr)
+  # qc <- "qc1"
 
   library(bigsnpr)
-  CHR <- snp_attach("data/UKBB_HM3_val.rds")$map$chromosome
+  corr_path <- ALL_CORR[[name_corr]]
+  grp <- readRDS(file.path(dirname(corr_path), "all_final_grp.rds"))
+  if (is.null(grp$ind)) {
+    CHR <- snp_attach("data/UKBB_HM3_val.rds")$map$chromosome
+    ind_keep <- split(seq_along(CHR), CHR)
+  } else {
+    ind_keep <- grp$ind
+  }
+
[…]

### 2022-07-20 code/run-PRS.R

@@ -1,11 +1,7 @@
-
-bigassertr::assert_dir("results")
-bigassertr::assert_dir("log")
-
 library(dplyr)
-files <- tibble(pheno = c(paste0("t1d_", c("affy", "illu")),
-                          paste0("brca_", c("onco", "icogs")),
-                          "cad", "prca", "mdd")) %>%
+grid <- tibble(pheno = c(paste0("t1d_", c("affy", "illu")),
+                         paste0("brca_", c("onco", "icogs")),
+                         "cad", "prca", "mdd")) %>%
   mutate(whichN = NA) %>%
   add_row(pheno = "vitaminD", whichN = c("trueN", "maxN")) %>%
   tidyr::expand_grid(
@@ -19,7 +15,10 @@ plan(batchtools_slurm(resources = list(
   t = "12:00:00", c = NCORES + 2, mem = "120g",
   name = basename(rstudioapi::getSourceEditorContext()$path))))
 
-furrr::future_pwalk(files, function(pheno, whichN, qc, name_corr) {
+bigassertr::assert_dir("results")
+bigassertr::assert_dir("results/sumstats")
+
+furrr::future_pwalk(grid, function(pheno, whichN, qc, name_corr) {
 
   # pheno <- "vitaminD"
   # whichN <- "trueN"
@@ -32,7 +31,7 @@ furrr::future_pwalk(files, function(pheno, whichN, qc, name_corr) {
     sumstats$n_eff <- max(sumstats$n_eff)
 
   suffix_N <- `if`(is.na(whichN), "", paste0("_", whichN))
-  basename <- file.path("results", sub("\\.rds$", suffix_N, basename(gwas_file)))
+  basename <- file.path("results/sumstats", sub("\\.rds$", suffix_N, basename(gwas_file)))
   if (grepl("blocks", name_corr, fixed = TRUE))
     basename <- paste0(basename, "_adj_with_blocks")
 
@@ -113,4 +112,62 @@ furrr::future_pwalk(files, function(pheno, whichN, qc, name_corr) {
                      vec_p_init = seq_log(1e-4, 0.5, 30),
                      allow_jump_sign = FALSE, shrink_corr = 0.95),
     file = paste0(basename, "_LDpred2-auto-rob.rds"))
+})
+
+
+#### Run PRS-CS ####
+
+library(dplyr)
[…]

### 2022-07-20 code/run-methods.R

@@ -60,7 +60,7 @@ run_ldpred2 <- function(df_beta, corr) {
 run_lassosum2 <- function(df_beta, corr) {
   beta_lassosum <- snp_lassosum2(corr, df_beta, ncores = NCORES)
   pred_grid <- big_prodMat(G, beta_lassosum, ncores = NCORES)
-  score.val <- apply(pred_grid[ind.val, ], 2, cor, y = y[ind.val])
+  print(score.val <- apply(pred_grid[ind.val, ], 2, cor, y = y[ind.val]))
   pred_lassosum <- pred_grid[ind.test, which.max(score.val)]
   print(cor(pred_lassosum, y[ind.test])**2)
 }
@@ -73,8 +73,8 @@ run_lassosum2 <- function(df_beta, corr) {
 
 run_CT <- function(df_beta) {
 
-  CHR <- as.integer(data$map$chromosome)
-  POS <- data$map$physical.pos
+  CHR <- as.integer(ukb$map$chromosome)
+  POS <- ukb$map$physical.pos
   all_keep <- snp_grid_clumping(G, CHR, POS, ind.row = ind.val,
                                 grid.base.size = 200,
                                 grid.thr.r2 = c(0.05, 0.2, 0.8),
@@ -114,11 +114,11 @@ run_lassosum <- function(df_beta) {
   system.time(
     out <- lassosum.pipeline(
       cor = t / sqrt(n - 2 + t^2),
-      snp = data$map$marker.ID,
-      A1 = data$map$allele1,
-      A2 = data$map$allele2,
+      snp = ukb$map$marker.ID,
+      A1 = ukb$map$allele1,
+      A2 = ukb$map$allele2,
       exclude.ambiguous = FALSE,
-      test.bfile = "tmp-data/simu_chr22",
+      test.bfile = "tmp-data/for_gctb_simu",
       LDblocks = "EUR.hg19",
       cluster = cl,
       destandardize = TRUE
@@ -138,113 +138,134 @@ run_lassosum <- function(df_beta) {
 
 #### Run SBayesR -> do not run because always diverged ####
 
-# gctb <- "../paper-ldpred2/tmp-data/gctb_2.02_Linux/gctb"
-#
-# # Compute LD
[…]

### 2022-07-20 code/simu-altpop.R

@@ -5,25 +5,23 @@ G <- ukb$genotypes
 
 
 library(future.batchtools)
-NCORES <- 15
+NCORES <- 14
 plan(batchtools_slurm(resources = list(
-  t = "12:00:00", c = NCORES + 1, mem = "100g",
+  t = "12:00:00", c = NCORES + 2, mem = "100g",
   name = basename(rstudioapi::getSourceEditorContext()$path))))
 
 bigassertr::assert_dir("results-simu-altpop")
 bigassertr::assert_dir("log")
 
+res_files <- paste0("results-simu-altpop/res_", 1:20, ".rds")
 
-furrr::future_walk(1:10, function(ic) {
-
-  res_file <- paste0("results-simu-altpop/res_", ic, ".rds")
-  if (file.exists(res_file)) return(NULL)
+furrr::future_walk(which(!file.exists(res_files)), function(ic) {
 
   #### Run GWAS and prepare sumstats ####
 
   load("data/ukbb4simu_ind.RData")
 
-  y <- snp_simuPheno(G, h2 = 0.2, M = 2000, ncores = NCORES)$pheno
+  y <- snp_simuPheno(G, h2 = `if`(ic > 10, 0.04, 0.2), M = 2000, ncores = NCORES)$pheno
 
   # GWAS to get sumstats
   gwas <- big_univLinReg(G, y[ind.gwas], ind.train = ind.gwas, ncores = NCORES)
@@ -35,21 +33,26 @@ furrr::future_walk(1:10, function(ic) {
 
   source("code/run-methods.R", local = TRUE)
 
+  ukb <- ukb
+
   # Correlation matrices made in 'code/prepare-corr-simu-chr22.R'
   corr1 <- readRDS("tmp-data/corr_simu_val.rds")
   corr2 <- readRDS("tmp-data/corr_simu_val_with_blocks.rds")
   # Correlation matrices made in 'code/prepare-corr-altpop.R'
   corr3 <- readRDS("tmp-data/corr_simu_altpop.rds")
   corr4 <- readRDS("tmp-data/corr_simu_altpop_with_blocks.rds")
+  corr5 <- readRDS("tmp-data/for_gctb_simu_altpop.rds")
+  corr6 <- readRDS("tmp-data/for_gctb_simu_altpop_with_blocks.rds")
 
-  ALL_COR <- list(corr1, corr2, corr3, corr4)
+  ALL_COR <- list(corr1, corr2, corr3, corr4, corr5, corr6)
[…]

### 2022-07-20 code/simu-dosage-info.R

@@ -1,24 +1,22 @@
 library(future.batchtools)
-NCORES <- 15
+NCORES <- 14
 plan(batchtools_slurm(resources = list(
-  t = "12:00:00", c = NCORES + 1, mem = "100g",
+  t = "12:00:00", c = NCORES + 2, mem = "100g",
   name = basename(rstudioapi::getSourceEditorContext()$path))))
 
 bigassertr::assert_dir("results-simu-info")
 bigassertr::assert_dir("log")
 
+res_files <- paste0("results-simu-info/res_", 1:20, ".rds")
 
-furrr::future_walk(1:10, function(ic) {
-
-  res_file <- paste0("results-simu-info/res_", ic, ".rds")
-  if (file.exists(res_file)) return(NULL)
+furrr::future_walk(which(!file.exists(res_files)), function(ic) {
 
   library(dplyr)
   library(bigsnpr)
-  simu_true <- snp_attach("data/ukbb4simu.rds")
+  ukb <- simu_true <- snp_attach("data/ukbb4simu.rds")
   G <- simu_true$genotypes
 
-  pheno <- snp_simuPheno(G, h2 = 0.2, M = 2000, ncores = NCORES)
+  pheno <- snp_simuPheno(G, h2 = `if`(ic > 10, 0.04, 0.2), M = 2000, ncores = NCORES)
   y <- pheno$pheno
 
   simu_imp <- snp_attach("data/ukbb4simu_imp.rds")
@@ -49,21 +47,20 @@ furrr::future_walk(1:10, function(ic) {
   source("code/run-methods.R", local = TRUE)
 
   # Correlation matrices made in 'code/prepare-corr-simu-chr22.R'
-  corr0 <- readRDS("tmp-data/corr0_simu_val.rds")
   corr1 <- readRDS("tmp-data/corr_simu_val.rds")
   corr2 <- readRDS("tmp-data/corr_simu_val_with_blocks.rds")
-  dim(corr0)  # 40000 x 40000
 
   ALL_LDPRED <- c("LDpred2-inf", "LDpred2", "LDpred2-low-h2",
                   "LDpred2-auto", "LDpred2-auto-rob")
 
   all_comb <- tibble::tribble(
-    ~ Method,     ~ fun,
-    ALL_LDPRED,   run_ldpred2,
-    "lassosum2",  run_lassosum2,
+    ~ Method,      ~ fun,         ~ corr,
[…]

### 2022-07-20 code/simu-misspec-N.R

@@ -3,31 +3,19 @@ library(bigsnpr)
 ukb <- snp_attach("data/ukbb4simu.rds")
 G <- ukb$genotypes
 
-load("data/ukbb4simu_ind.RData")
 
-# prepare data in bed format
-data <- snp_fake(n = nrow(G), m = 1)
-data$map <- ukb$map
-data$map$genetic.dist <- snp_asGeneticPos(as.integer(ukb$map$chromosome),
-                                          ukb$map$physical.pos, dir = "tmp-data")
-data$map$marker.ID <- paste0("SNP", rows_along(data$map))
-data$genotypes <- G
-# snp_writeBed(data, ind.row = ind.val, bedfile = "tmp-data/simu_chr22.bed")
-
+NCORES <- 14
 library(future.batchtools)
-NCORES <- 15
 plan(batchtools_slurm(resources = list(
-  t = "12:00:00", c = NCORES + 1, mem = "100g",
+  t = "12:00:00", c = NCORES + 2, mem = "100g",
   name = basename(rstudioapi::getSourceEditorContext()$path))))
 
 bigassertr::assert_dir("results-simu-misN")
 bigassertr::assert_dir("log")
 
+res_files <- paste0("results-simu-misN/res_", 1:20, ".rds")
 
-furrr::future_walk(1:10, function(ic) {
-
-  res_file <- paste0("results-simu-misN/res_", ic, ".rds")
-  if (file.exists(res_file)) return(NULL)
+furrr::future_walk(which(!file.exists(res_files)), function(ic) {
 
   #### Run GWAS and prepare sumstats ####
 
@@ -37,7 +25,7 @@ furrr::future_walk(1:10, function(ic) {
                    sort(sample(ind.gwas, length(ind.gwas) * 0.8)),
                    sort(sample(ind.gwas, length(ind.gwas) * 0.6)))
 
-  y <- snp_simuPheno(G, h2 = 0.2, M = 2000, ncores = NCORES)$pheno
+  y <- snp_simuPheno(G, h2 = `if`(ic > 10, 0.04, 0.2), M = 2000, ncores = NCORES)$pheno
 
   # GWAS to get sumstats
   gwas_set <- sample(rep_len(c(1, 1, 2, 3), ncol(G)))
@@ -71,24 +59,35 @@ furrr::future_walk(1:10, function(ic) {
 
   source("code/run-methods.R", local = TRUE)
 
-  data <- data
+  ukb <- ukb
 
[…]

### 2022-07-20 paper/paper-misspec.tex
last minor

@@
 We form nearly independent LD blocks using the optimal algorithm developed in \citet{prive2021optimal}.
-{\color{red}Note that a new parameter \texttt{max\_r2} has been added to this method since, which controls the maximum single squared correlation allowed outside blocks, which offers an extra guarantee that the splitting is good and makes the function much faster by discarding lots of possible splits.
+{\color{red}Note that a new parameter \texttt{max\_r2} has now been added to this method, which controls the maximum single squared correlation allowed outside blocks, which offers an extra guarantee that the splitting is good and makes the function much faster by discarding lots of possible splits.
 For different numbers of blocks and maximum number of variants in each block, we use the split that minimizes $\big(C2 \cdot \sqrt{5 + C1}\big)$, where $C1$ is the sum of squared correlations outside the blocks, and $C2$ is the sum of squared sizes of the blocks (Figure \ref{fig:cost-split}).
@@
 As a second option, instead of using dosages to compute the GWAS summary statistics, it has been argued that using multiple imputation (MI) would be more appropriate \cite[]{Palmer2016}.
-{\color{red}Here MI consists in forming multiple (e.g.\ 20) datasets of hard-called genotypes by sampling them according to the imputation probabilities stored in the BGEN files, performing a GWAS on each of them, and pooling the GWAS estimates.}
+{\color{red}Here MI consists in forming multiple (e.g.\ 20) datasets of hard-called genotypes by sampling them according to the imputation probabilities stored in the BGEN files, performing a GWAS on each of these datasets, and pooling the GWAS estimates.}
 In simulations, we show that $\hat{\gamma}_j^{\text{MI}} \approx \hat{\gamma}_j^{\text{imp}} \cdot \text{INFO}_j$ and $Z_j^{\text{MI}} \approx Z_j^{\text{imp}} \cdot \text{INFO}_j$, where $Z = \hat{\gamma} / \text{se}(\hat{\gamma})$ (Figure \ref{fig:compare-mi}).
@@
 For LDpred2, we recommend to also include a smaller value for the heritability in the grid of parameters, to allow for more regularization when needed.
-As these two methods use the same data and the same principle (i.e.\ hyper-parameter tuning), one could run both methods and should the best one when tuning hyper-parameters.
+As these two methods use the same data and the same principle (i.e.\ hyper-parameter tuning), one could run both methods and choose the best one when tuning hyper-parameters.
 Fifth, if no validation/tuning dataset is available, we recommend using LDpred2-auto with the two new parameters introduced here, or PRS-CS-auto if there are strong misspecifications and the phenotype of interest is known to have small effects only.

### 2022-08-04 paper/paper-misspec.tex
rm markup

@@
 Third, we introduce two new optional parameters in LDpred2-auto to make it more robust to these types of misspecification.
-{\color{red}Note that, in this paper, we use ``robust'' to mean that we obtain similar predictive performance to when there is no misspecification. 
-We also reimplement and use a new version of lassosum, called lassosum2, in which we better handle when per-variant sample sizes are different.}
+Note that, in this paper, we use ``robust'' to mean that we obtain similar predictive performance to when there is no misspecification. 
+We also reimplement and use a new version of lassosum, called lassosum2, in which we better handle when per-variant sample sizes are different.
 We focus our investigations on LDpred2 and lassosum2 for two reasons.
@@
 Second, lassosum2 now uses the exact same input parameters as LDpred2, which makes it easy for us to test the different QCs and adjustments presented here.
-{\color{red}We additionally investigate PRS-CS and SBayesR, two competitive PGS methods \cite[]{ge2019polygenic,lloyd2019improved}, for which we develop some code to convert between the different formats required for the LD matrices and input GWAS summary statistics.}
+We additionally investigate PRS-CS and SBayesR, two competitive PGS methods \cite[]{ge2019polygenic,lloyd2019improved}, for which we develop some code to convert between the different formats required for the LD matrices and input GWAS summary statistics.
 
@@
 To get a set of genetically homogeneous individuals, we compute a robust Mahalanobis distance based on the first 16 PCs and further restrict individuals to those within a log-distance of 5 \cite[]{prive2020efficient}.
-This results in 362,307 individuals {\color{red}of Northwestern European ancestry}.
+This results in 362,307 individuals of Northwestern European ancestry.
 We randomly sample 300,000 individuals to form a training set (e.g.\ to run the GWAS), 10,000 individuals to form a validation set (to tune hyper-parameters), and use the remaining 52,307 individuals to form a test set (to evaluate final predictive models).
@@
 
-We also use the UK Biobank data for validation/testing in real data analyses, and use the same individuals {\color{red}of Northwestern European ancestry} as described in the previous section.
+We also use the UK Biobank data for validation/testing in real data analyses, and use the same individuals of Northwestern European ancestry as described in the previous section.
 We sample 10,000 individuals to form a validation set and use the remaining 352,307 individuals as test set.
@@
 
-{\color{red}\subsection{Model for summary statistics}}
+\subsection{Model for summary statistics}
 
@@
 
-{\color{red}It is assumed that the marginal (GWAS) effects follow
+It is assumed that the marginal (GWAS) effects follow
 \begin{align}
@@
 However, in practice, we only have estimates for $S$ and $R$, which causes some model misspecifications.
-Moreover, ubiquitous correlations between the large number of variants may cause a Gibbs sampler or any iterative approach used in PGS methods to fail properly estimating genetic effect sizes \cite[]{roberts1997updating}.}
+Moreover, ubiquitous correlations between the large number of variants may cause a Gibbs sampler or any iterative approach used in PGS methods to fail properly estimating genetic effect sizes \cite[]{roberts1997updating}.
 
@@
 Also note that the estimate of $\text{var}(G_j)$ has to account for the imputation accuracy, i.e.\ use $2 \cdot f_j \cdot (1 - f_j) \cdot \text{INFO}_j$ (Section \ref{sec:imputation}).
-{\color{red}In equations \eqref{eq:approx-sd-lin} and \eqref{eq:imputeN}, we estimate $\text{var}(y)$ by the first percentile of $0.5 \left(N ~ \text{se}(\hat{\gamma}_j)^2 + \hat{\gamma}_j^2\right)$, where the first percentile approximates the minimum and is robust to outliers.}
+In equations \eqref{eq:approx-sd-lin} and \eqref{eq:imputeN}, we estimate $\text{var}(y)$ by the first percentile of $0.5 \left(N ~ \text{se}(\hat{\gamma}_j)^2 + \hat{\gamma}_j^2\right)$, where the first percentile approximates the minimum and is robust to outliers.
 
@@
 Note that only positions $k$ for which $R_{k,j} \neq 0$ must be updated in this vector $R_{.,j}^T \beta^{(t-1)}$.
-{\color{red}Since bigsnpr v1.10.4, we now also use this updating strategy in LDpred2 (-grid and -auto), which makes it much faster, especially when the polygenicity is small.}
+Since bigsnpr v1.10.4, we now also use this updating strategy in LDpred2 (-grid and -auto), which makes it much faster, especially when the polygenicity is small.
 
@@
 By default in lassosum2, we use a similar sequence of 30 values, but between $\lambda_0$ and $\lambda_0 / 100$ instead, where $\lambda_0 = \max_j \big|\widehat{\beta}_j\big|$ is the minimum $\lambda$ for which no variable enters the model because the L1-regularization is too strong.
-{\color{red}To make lassosum2 more robust to different per-variant sample sizes $n_j$, we also introduce per-variant penalty factors $\sqrt{\max_k(n_k) / n_j}$, by which we internally multiply $\lambda$ and $\delta$ to penalize variants with smaller GWAS sample sizes more.}
+To make lassosum2 more robust to different per-variant sample sizes $n_j$, we also introduce per-variant penalty factors $\sqrt{\max_k(n_k) / n_j}$, by which we internally multiply $\lambda$ and $\delta$ to penalize variants with smaller GWAS sample sizes more.
 Note that we do not provide an ``auto'' version using pseudo-validation (as in lassosum \cite[]{mak2017polygenic}) since we have not found the pseudo-validation scores to be consistent enough with the predictive performance (Figure \ref{fig:pseudoval}).
@@
 We form nearly independent LD blocks using the optimal algorithm developed in \citet{prive2021optimal}.
-{\color{red}Note that a new parameter \texttt{max\_r2} has now been added to this method, which controls the maximum single squared correlation allowed outside blocks, which offers an extra guarantee that the splitting is good and makes the function much faster by discarding lots of possible splits.
+Note that a new parameter \texttt{max\_r2} has now been added to this method, which controls the maximum single squared correlation allowed outside blocks, which offers an extra guarantee that the splitting is good and makes the function much faster by discarding lots of possible splits.
 For different numbers of blocks and maximum number of variants in each block, we use the split that minimizes $\big(C2 \cdot \sqrt{5 + C1}\big)$, where $C1$ is the sum of squared correlations outside the blocks, and $C2$ is the sum of squared sizes of the blocks (Figure \ref{fig:cost-split}).
-Twelve blocks from chromosome 22 are used in the simulations, and between 731 and 2527 in the real data analyses (cf.\ next section).}
+Twelve blocks from chromosome 22 are used in the simulations, and between 731 and 2527 in the real data analyses (cf.\ next section).
 Having a correlation matrix with independent blocks prevents small errors in the algorithm (e.g.\ the Gibbs sampler in LDpred2) from propagating to too many variants.
 It also makes running LDpred2 (and lassosum2) faster, taking e.g.\ 70\% of the initial time (since only 70\% of the initial non-zero values of the correlation matrix are kept).
-{\color{red}Note that we also compute the correlations within the same blocks and from the same data for PRS-CS.}
+Note that we also compute the correlations within the same blocks and from the same data for PRS-CS.
 
@@
 We use either the 503 European (including 99 Finnish) individuals from the 1000 Genomes (1000G) data \cite[]{10002015global}, 404 Finnish-like UKBB + the 99 Finnish 1000G individuals (also 503 in total), or the 10,000 UK individuals from the validation set we use in this paper.
-{\color{red}1454, 731, and 1165 independent LD blocks are identified for these three LD references, respectively.}
+1454, 731, and 1165 independent LD blocks are identified for these three LD references, respectively.
 
@@
 We also construct versions of these LD references with independent LD blocks, as described in the previous section.
-{\color{red}2527, 1769, and 2196 independent LD blocks are identified for these three LD references, respectively.}
+2527, 1769, and 2196 independent LD blocks are identified for these three LD references, respectively.
 
@@
 We design GWAS simulations where variants have different sample sizes, which is often the case when meta-analyzing GWAS summary statistics from multiple cohorts with different sets of variants \cite[]{wang2021global}.
-Using 40,000 variants from chromosome 22 (Methods), we simulate quantitative phenotypes {\color{red}by picking 2000 causal variants at random, sampling their genetic effects (after scaling of the genotypes) from a normal distribution and adding some Gaussian noise to the genetic component of the phenotype so that it has a heritability of 20\%.
-This is implemented in function \texttt{snp\_simuPheno} of R package bigsnpr \cite[]{prive2017efficient}.}
+Using 40,000 variants from chromosome 22 (Methods), we simulate quantitative phenotypes by picking 2000 causal variants at random, sampling their genetic effects (after scaling of the genotypes) from a normal distribution and adding some Gaussian noise to the genetic component of the phenotype so that it has a heritability of 20\%.
+This is implemented in function \texttt{snp\_simuPheno} of R package bigsnpr \cite[]{prive2017efficient}.
 We then randomly divide the 40,000 variants into three groups: for half of the variants, we use 100\% of 300,000 individuals for GWAS, but use only 80\% for one quarter of the variants and 60\% for the remaining quarter.
-{\color{red}We use a simple linear regression for GWAS, as implemented in function \texttt{big\_univLinReg} of R package bigstatsr \cite[]{prive2017efficient}.}
-We then run C+T, lassosum, lassosum2 (Methods section \ref{lassosum2}), LDpred2-inf, LDpred2(-grid), LDpred2-auto{\color{red}, SBayesR (GCTB v2.03 with the automatic \texttt{robust} option), and PRS-CS-auto} \cite[]{prive2019making,mak2017polygenic,prive2020ldpred2,lloyd2019improved,ge2019polygenic} by using either the true per-variant GWAS sample sizes, the total sample size for all variants, or per-variant sample sizes imputed using equation \eqref{eq:imputeN}.
+We use a simple linear regression for GWAS, as implemented in function \texttt{big\_univLinReg} of R package bigstatsr \cite[]{prive2017efficient}.
+We then run C+T, lassosum, lassosum2 (Methods section \ref{lassosum2}), LDpred2-inf, LDpred2(-grid), LDpred2-auto, SBayesR (GCTB v2.03 with the automatic \texttt{robust} option), and PRS-CS-auto \cite[]{prive2019making,mak2017polygenic,prive2020ldpred2,lloyd2019improved,ge2019polygenic} by using either the true per-variant GWAS sample sizes, the total sample size for all variants, or per-variant sample sizes imputed using equation \eqref{eq:imputeN}.
 When providing true per-variant GWAS sample sizes, squared correlations between the polygenic scores and the simulated phenotypes are of 0.123 for C+T, 0.160 for lassosum, 0.169 for lassosum2, 0.159 for LDpred2(-grid), 0.140 for LDpred2-auto, 0.141 for LDpred2-inf, 0.138 for SBayesR, and 0.159 for PRS-CS-auto (Figure \ref{fig:simu-misN}, averaging over 10 simulations).
 Results when using imputed sample sizes are very similar to when using the true ones.
-Note that C+T does not use this sample size information{\color{red}, and that PRS-CS can only be provided with a single value (we use the maximum one)}.
+Note that C+T does not use this sample size information, and that PRS-CS can only be provided with a single value (we use the maximum one).
 When using the total GWAS sample size instead of the per-variant sample sizes, predictive performance slightly decreases to 0.158 for lassosum and to 0.164 for lassosum2, but dramatically decreases for LDpred2 with new values of 0.134 for LDpred2-grid, 0.122 for LDpred2-auto, 0.125 for LDpred2-inf (Figure \ref{fig:simu-misN}).
 This extreme simulation scenario shows that LDpred2 can be sensitive to the misspecification of per-variant GWAS sample sizes, whereas lassosum (and lassosum2) seems little affected by this.
-{\color{red}With 0.141 for SBayesR, predictive performance is actually larger than when using the true per-variant sample sizes.}
+With 0.141 for SBayesR, predictive performance is actually larger than when using the true per-variant sample sizes.
 
@@
 The GWAS sample size is either ``true'' when providing the true per-variant sample size, ``max'' when providing instead the maximum sample size as a unique value to be used for all variants, ``imputed'' (cf.\ equation \eqref{eq:imputeN}), or ``any'' when the method does not use this information (the case for C+T).
-Red bars correspond to using the LD with independent blocks (Methods section \ref{sec:LD}){\color{red}, which is a requirement for lassosum and PRS-CS}.}
+Red bars correspond to using the LD with independent blocks (Methods section \ref{sec:LD}), which is a requirement for lassosum and PRS-CS}.
 	\label{fig:simu-misN}
@@
 We conduct further investigations to explain the results of figure \ref{fig:simu-misN}.
-{\color{red}First, the results for LDpred2-auto and SBayesR are similar to LDpred2-inf in these simulations because the internal polygenicity estimate $p$ (proportion of causal variants) of LDpred2-auto, and similarly the proportion of non-null variants in SBayesR, always continuously increases up to $1$ in the Gibbs sampler, which makes them behave as an infinitesimal model ($p = 1$).}
+First, the results for LDpred2-auto and SBayesR are similar to LDpred2-inf in these simulations because the internal polygenicity estimate $p$ (proportion of causal variants) of LDpred2-auto, and similarly the proportion of non-null variants in SBayesR, always continuously increases up to $1$ in the Gibbs sampler, which makes them behave as an infinitesimal model ($p = 1$).
 To overcome this limitation in LDpred2-auto, we introduce two new parameters to make it more robust, and refer to this as ``LDpred2-auto-rob'' here (Methods section \ref{sec:robust}).
@@
 
-{\color{red}As secondary analysis, we rerun these simulations with a smaller heritability of 4\% instead.
+As secondary analysis, we rerun these simulations with a smaller heritability of 4\% instead.
 While predictive performance are overall much lower, relative performance are less attenuated when using the maximum sample size, otherwise results are highly consistent as before (Figure \ref{fig:simu-misN-smallh2}).
-We also try using the shrunk LD matrix from GCTB that is used e.g.\ for SBayesR \cite[]{lloyd2019improved}; in these simulations, higher predictive performance are obtained for LDpred2-auto with this shrunk LD matrix, whereas SBayesR benefit from using LDpred2's windowed LD matrix (Figure \ref{fig:simu-misN-shrunk}).}
+We also try using the shrunk LD matrix from GCTB that is used e.g.\ for SBayesR \cite[]{lloyd2019improved}; in these simulations, higher predictive performance are obtained for LDpred2-auto with this shrunk LD matrix, whereas SBayesR benefit from using LDpred2's windowed LD matrix (Figure \ref{fig:simu-misN-shrunk}).
 
@@
 As a second option, instead of using dosages to compute the GWAS summary statistics, it has been argued that using multiple imputation (MI) would be more appropriate \cite[]{Palmer2016}.
-{\color{red}Here MI consists in forming multiple (e.g.\ 20) datasets of hard-called genotypes by sampling them according to the imputation probabilities stored in the BGEN files, performing a GWAS on each of these datasets, and pooling the GWAS estimates.}
+Here MI consists in forming multiple (e.g.\ 20) datasets of hard-called genotypes by sampling them according to the imputation probabilities stored in the BGEN files, performing a GWAS on each of these datasets, and pooling the GWAS estimates.
 In simulations, we show that $\hat{\gamma}_j^{\text{MI}} \approx \hat{\gamma}_j^{\text{imp}} \cdot \text{INFO}_j$ and $Z_j^{\text{MI}} \approx Z_j^{\text{imp}} \cdot \text{INFO}_j$, where $Z = \hat{\gamma} / \text{se}(\hat{\gamma})$ (Figure \ref{fig:compare-mi}).
@@
 Moreover, new additions for robustness introduced before, namely LDpred2-low-h2, LDpred2-auto-rob, and forming independent blocks in the LD matrix also improve predictive performance for all corrections (Figure \ref{fig:simu-info}).
-{\color{red}However, these corrections do not prove beneficial for PRS-CS-auto (Figure \ref{fig:simu-info}).}
+However, these corrections do not prove beneficial for PRS-CS-auto (Figure \ref{fig:simu-info}).
 
-{\color{red}When performing similar simulations with a heritability of 4\% (instead of 20\%), correction ``sqrt\_info'' still provides slightly higher predictive performance for all methods (compared to no correction), yet the other two corrections sometimes provide lower predictive performance (Figure \ref{fig:simu-info-smallh2}).}
+When performing similar simulations with a heritability of 4\% (instead of 20\%), correction ``sqrt\_info'' still provides slightly higher predictive performance for all methods (compared to no correction), yet the other two corrections sometimes provide lower predictive performance (Figure \ref{fig:simu-info-smallh2}).
 In the real data applications hereinafter, we therefore choose to use the first correction, ``sqrt\_info'', which seems to more consistently provide better results, and is simple because it is somewhat equivalent to post-processing PGS effects by multiplying them by $\sqrt{\text{INFO}}$.
@@
 Here we design simulations to understand the impact of using a mismatched LD reference panel, e.g.\ that comes from a different population compared to the one used to compute the GWAS summary statistics.
-We use the same simulation setup as before (Methods){\color{red}, i.e.\ using individuals of Northwestern European ancestry from the UK Biobank for training (GWAS), validation (tuning of hyper-parameters), and testing (the final models)}.
-In addition {\color{red}to the LD reference panel of Northwestern European ancestry}, we design an alternative reference panel based on 10,000 individuals from South Europe by using the ``Italy'' center defined in \citet{prive2021high}.
-{\color{red}Allele frequencies and pairwise correlations are quite similar between these two LD reference panels (Figure \ref{fig:compare-altpop}).}
+We use the same simulation setup as before (Methods), i.e.\ using individuals of Northwestern European ancestry from the UK Biobank for training (GWAS), validation (tuning of hyper-parameters), and testing (the final models).
+In addition to the LD reference panel of Northwestern European ancestry, we design an alternative reference panel based on 10,000 individuals from South Europe by using the ``Italy'' center defined in \citet{prive2021high}.
+Allele frequencies and pairwise correlations are quite similar between these two LD reference panels (Figure \ref{fig:compare-altpop}).
 When using this alternative LD reference panel instead of a well-matched one as in the previous sections, squared correlations between the polygenic scores and the simulated phenotypes drop from 0.173 to 0.166 for lassosum2, from 0.168 to 0.159 for LDpred2(-grid), from 0.174 to 0.169 for LDpred2-low-h2, from 0.142 to 0.136 for LDpred2-auto, from 0.162 to 0.147 for LDpred2-auto-rob, from 0.143 to 0.138 for LDpred2-inf, and from  0.163 to 0.155 for PRS-CS-auto (Figure \ref{fig:simu-altpop}, averaging over 10 simulations).
 Forming independent LD blocks in these two LD matrices always improve predictive performance, yet again.
-{\color{red}We also compute and test using a shrunk LD matrix (from GCTB) for this alternative LD reference panel; again, using this type of LD matrix seems beneficial for LDpred2-auto.
-Finally, under the same simulation scenario, but with a smaller simulated heritability of 4\% (instead of 20\%), there remains a small drop in predictive performance when using the alternative LD reference (Figure \ref{fig:simu-altpop-smallh2}).}
+We also compute and test using a shrunk LD matrix (from GCTB) for this alternative LD reference panel; again, using this type of LD matrix seems beneficial for LDpred2-auto.
+Finally, under the same simulation scenario, but with a smaller simulated heritability of 4\% (instead of 20\%), there remains a small drop in predictive performance when using the alternative LD reference (Figure \ref{fig:simu-altpop-smallh2}).
 
@@
 In the comparison, we use the QC proposed in \citet{prive2020ldpred2} first (which ends up filtering on MAF here, which we call ``qc1'').
-We then also filter out the two long-range LD regions of chromosome 6 and 8 for the OncoArray summary statistics and remove around 500 variants when filtering on differences of AFs {\color{red}(> 0.1)} between summary statistics and the validation dataset (``qc2'').
+We then also filter out the two long-range LD regions of chromosome 6 and 8 for the OncoArray summary statistics and remove around 500 variants when filtering on differences of AFs (> 0.1) between summary statistics and the validation dataset (``qc2'').
 As for the correction using INFO scores, we use the first correction, ``sqrt\_info'', which is simple because it is equivalent to post-processing PGS effects by multiplying them by $\sqrt{\text{INFO}}$.
@@
 QC plots comparing standard deviations usually show little deviation from the identity line (after the INFO score correction), except for coronary artery disease (CAD) summary statistics (Figures \ref{fig:cad_qc}-\ref{fig:t1d_illu_qc}).
-{\color{red}Note that the popular CAD summary statistics used here come from a multi-ancestry GWAS meta-analysis with more than 20\% non-European samples \cite[]{nikpay2015comprehensive,prive2021using}.}
+Note that the popular CAD summary statistics used here come from a multi-ancestry GWAS meta-analysis with more than 20\% non-European samples \cite[]{nikpay2015comprehensive,prive2021using}.
 In figures \ref{fig:res_cad} and \ref{fig:res_brca_onco}-\ref{fig:res_t1d_aff}, we then provide similar results as figure \ref{fig:res_brca_icogs} for other phenotypes.
@@
 
-{\color{red}For all phenotypes, PRS-CS-auto is very robust, whatever the QC procedure used (e.g.\ see the results for CAD, Figure \ref{fig:res_cad}).
+For all phenotypes, PRS-CS-auto is very robust, whatever the QC procedure used (e.g.\ see the results for CAD, Figure \ref{fig:res_cad}).
 We believe this is because of the strong regularization used internally.
@@
 Results with LDpred2-auto-rob are always best when using ``qc2'', the most stringent QC; then no shrinkage (1) seems needed, whereas a strong shrinkage make predictive performance drop.
-For phenotypes with relatively large genetic effects (e.g.\ vitamin D, Figure \ref{fig:res_vitaminD}), PRS-CS-auto perform much worse, possibly due to using too much regularization.}
+For phenotypes with relatively large genetic effects (e.g.\ vitamin D, Figure \ref{fig:res_vitaminD}), PRS-CS-auto perform much worse, possibly due to using too much regularization.
 
@@
 Interestingly, using the small Finnish LD reference panel we have defined here, composed of 503 individuals only, seems to almost always provide more predictive polygenic scores than using the large UK one composed of 10,000 individuals or the widely used European subset of the 1000 Genomes (1000G) data (Figure \ref{fig:finngen}).
-{\color{red}For PRS-CS-auto, none of the three LD references seems to consistently outperform the other two across all phenotypes considered.}
+For PRS-CS-auto, none of the three LD references seems to consistently outperform the other two across all phenotypes considered.
 Note that the PGS here are validated in people of UK-like ancestry in the UK biobank to obtain sufficient sample sizes.
@@
 We then compare three different LD reference panels to use with these GWAS summary statistics (Methods section \ref{sec:altLD}).
-Interestingly, when looking at height and HDL cholesterol, where we could get the best predictive performance, using the smaller Japanese LD reference we have defined here seems to provide more predictive PGS than using the one using a larger set of East Asian individuals from the UKBB or the widely used East Asian subset of the 1000 Genomes (1000G) data (Figure \ref{fig:bbj}){\color{red}, except when using PRS-CS-auto to predict height}.
+Interestingly, when looking at height and HDL cholesterol, where we could get the best predictive performance, using the smaller Japanese LD reference we have defined here seems to provide more predictive PGS than using the one using a larger set of East Asian individuals from the UKBB or the widely used East Asian subset of the 1000 Genomes (1000G) data (Figure \ref{fig:bbj}), except when using PRS-CS-auto to predict height.
 Note that the PGS here are validated in people of broad East Asian ancestry in the UK biobank and for continuous outcomes to obtain moderate sample sizes for comparison.
 
-{\color{red}\subsection{Recommendations}
+\subsection{Recommendations}
 
@@
 LD matrix as it proved to make the methods more robust (and slightly faster).
-Most of these new recommendations are included in the LDpred2 tutorial.}
+Most of these new recommendations are included in the LDpred2 tutorial.
 
@@
 		\hline
-		Differences in per-variant sample sizes & Possible violation of model assumptions {\color{red}(e.g.\ equation \eqref{eq:model}) resulting in poor predictive performance} & Imputation using equations \eqref{eq:imputeN} and \eqref{eq:imputeNeff}, and filtering$^{(2)}$   \\
+		Differences in per-variant sample sizes & Possible violation of model assumptions (e.g.\ equation \eqref{eq:model}) resulting in poor predictive performance & Imputation using equations \eqref{eq:imputeN} and \eqref{eq:imputeNeff}, and filtering$^{(2)}$   \\
 		\hline
@@
 It would also be useful to investigate the impact of the QC, well-matched LD reference panels and other adjustments we propose here on other (non-PGS) methods, for which consistency and robustness are likely to be very important as well \cite[]{chen2021improved}.
-{\color{red}For example, we are interested in assessing the impact of these misspecifications on the inference of disease architecture parameters in future work.}
+For example, we are interested in assessing the impact of these misspecifications on the inference of disease architecture parameters in future work.
 

### 2022-08-05 paper/paper-misspec.tex
rm links to suppmat

@@
 
-\usepackage{xr}
-\externaldocument{paper-misspec-supp}
+%\usepackage{xr}
+%\externaldocument{paper-misspec-supp}
 
@@
 
-Among genetic variants on chromosome 22 and with a minor allele frequency larger than 0.01 and an imputation INFO score larger than 0.4 (as reported by the UK Biobank), we sample 40,000 of them according to the inverse of the INFO score density so that they have varying levels of imputation accuracy (Figure \ref{fig:hist-info}).
+Among genetic variants on chromosome 22 and with a minor allele frequency larger than 0.01 and an imputation INFO score larger than 0.4 (as reported by the UK Biobank), we sample 40,000 of them according to the inverse of the INFO score density so that they have varying levels of imputation accuracy (Figure S1).
 We read the UK Biobank data into two different datasets using function \texttt{snp\_readBGEN} from R package bigsnpr \cite[]{prive2017efficient}, one by reading the BGEN data as hard-called genotypes by randomly sampling according to the imputation probabilities (of being 0, 1, or 2), and another one by reading it as dosages (i.e.\ expected values according to the imputation probabilities).
@@
 To make lassosum2 more robust to different per-variant sample sizes $n_j$, we also introduce per-variant penalty factors $\sqrt{\max_k(n_k) / n_j}$, by which we internally multiply $\lambda$ and $\delta$ to penalize variants with smaller GWAS sample sizes more.
-Note that we do not provide an ``auto'' version using pseudo-validation (as in lassosum \cite[]{mak2017polygenic}) since we have not found the pseudo-validation scores to be consistent enough with the predictive performance (Figure \ref{fig:pseudoval}).
+Note that we do not provide an ``auto'' version using pseudo-validation (as in lassosum \cite[]{mak2017polygenic}) since we have not found the pseudo-validation scores to be consistent enough with the predictive performance (Figure S2).
 Also note that, as in LDpred2, we run lassosum2 genome-wide using a sparse correlation matrix which assumes that variants further away than 3 cM are not correlated.
@@
 Note that a new parameter \texttt{max\_r2} has now been added to this method, which controls the maximum single squared correlation allowed outside blocks, which offers an extra guarantee that the splitting is good and makes the function much faster by discarding lots of possible splits.
-For different numbers of blocks and maximum number of variants in each block, we use the split that minimizes $\big(C2 \cdot \sqrt{5 + C1}\big)$, where $C1$ is the sum of squared correlations outside the blocks, and $C2$ is the sum of squared sizes of the blocks (Figure \ref{fig:cost-split}).
+For different numbers of blocks and maximum number of variants in each block, we use the split that minimizes $\big(C2 \cdot \sqrt{5 + C1}\big)$, where $C1$ is the sum of squared correlations outside the blocks, and $C2$ is the sum of squared sizes of the blocks (Figure S3).
 Twelve blocks from chromosome 22 are used in the simulations, and between 731 and 2527 in the real data analyses (cf.\ next section).
@@
 The first parameter prevents $p$ from diverging to $1$, while the second shrinks the off-diagonal elements of the LD matrix (a form of regularization). 
-Second, for lassosum2, results for a grid of parameters (over $\lambda$ and $\delta$) are quite smooth compared to LDpred2 (Figures \ref{fig:lassosum2-misN} and \ref{fig:ldpred2-misN}). 
-In these simulations with misspecified per-variant sample sizes, it seems highly beneficial to use a small value for the SNP heritability hyper-parameter $h^2$ in LDpred2, e.g.\ a value of 0.02 or even 0.002 when the true value is 0.2 (Figure \ref{fig:ldpred2-misN}).
+Second, for lassosum2, results for a grid of parameters (over $\lambda$ and $\delta$) are quite smooth compared to LDpred2 (Figures S4 and S5). 
+In these simulations with misspecified per-variant sample sizes, it seems highly beneficial to use a small value for the SNP heritability hyper-parameter $h^2$ in LDpred2, e.g.\ a value of 0.02 or even 0.002 when the true value is 0.2 (Figure S5).
 Indeed, using a small value for this hyper-parameter induces a larger regularization (shrinkage) on the effect sizes.
@@
 Finally, we introduce a last change for robustness here: we form independent LD blocks in the LD matrix to prevent small errors in the Gibbs sampler to propagate to too many variants (Methods section \ref{sec:LD}).
-This change seems to solve convergence issues of LDpred2 in these simulations (Figure \ref{fig:ldpred2-misN}) and further improves predictive performance for all LDpred2 methods (Figure \ref{fig:simu-misN}).
+This change seems to solve convergence issues of LDpred2 in these simulations (Figure S5) and further improves predictive performance for all LDpred2 methods (Figure \ref{fig:simu-misN}).
 
 As secondary analysis, we rerun these simulations with a smaller heritability of 4\% instead.
-While predictive performance are overall much lower, relative performance are less attenuated when using the maximum sample size, otherwise results are highly consistent as before (Figure \ref{fig:simu-misN-smallh2}).
-We also try using the shrunk LD matrix from GCTB that is used e.g.\ for SBayesR \cite[]{lloyd2019improved}; in these simulations, higher predictive performance are obtained for LDpred2-auto with this shrunk LD matrix, whereas SBayesR benefit from using LDpred2's windowed LD matrix (Figure \ref{fig:simu-misN-shrunk}).
+While predictive performance are overall much lower, relative performance are less attenuated when using the maximum sample size, otherwise results are highly consistent as before (Figure S6).
+We also try using the shrunk LD matrix from GCTB that is used e.g.\ for SBayesR \cite[]{lloyd2019improved}; in these simulations, higher predictive performance are obtained for LDpred2-auto with this shrunk LD matrix, whereas SBayesR benefit from using LDpred2's windowed LD matrix (Figure S7).
 
@@
 Therefore, when using the expected genotypes from imputation (allele dosages), their standard deviations are often lower than the expected value under Hardy--Weinberg equilibrium, because $\text{INFO}_j \approx \frac{\text{var}(G_j)}{2 \hat{\theta}_j (1 - \hat{\theta}_j)}$.
-In simulations (cf.\ Methods section ``Data for simulations''), we verify that $\text{sd}(G_j)^{\text{true}} \approx \text{sd}(G_j)^{\text{imp}} / \sqrt{\text{INFO}_j}$ (Figure \ref{fig:compare-sd}).
-As a direct consequence of the lower standard deviation of dosages, we also show that GWAS effect sizes $\hat{\gamma}$ computed from imputed dosages are overestimated: $\hat{\gamma}_j^{\text{true}} \approx \hat{\gamma}_j^{\text{imp}} \cdot \sqrt{\text{INFO}_j}$ and $\text{se}(\hat{\gamma}_j)^{\text{true}} \approx \text{se}(\hat{\gamma}_j)^{\text{imp}} \cdot \sqrt{\text{INFO}_j}$ (Figures \ref{fig:compare-beta} and \ref{fig:compare-beta-se}).
+In simulations (cf.\ Methods section ``Data for simulations''), we verify that $\text{sd}(G_j)^{\text{true}} \approx \text{sd}(G_j)^{\text{imp}} / \sqrt{\text{INFO}_j}$ (Figure S8).
+As a direct consequence of the lower standard deviation of dosages, we also show that GWAS effect sizes $\hat{\gamma}$ computed from imputed dosages are overestimated: $\hat{\gamma}_j^{\text{true}} \approx \hat{\gamma}_j^{\text{imp}} \cdot \sqrt{\text{INFO}_j}$ and $\text{se}(\hat{\gamma}_j)^{\text{true}} \approx \text{se}(\hat{\gamma}_j)^{\text{imp}} \cdot \sqrt{\text{INFO}_j}$ (Figures S9 and S10).
 This is the first correction of summary statistics we consider in the simulations below.
@@
 Here MI consists in forming multiple (e.g.\ 20) datasets of hard-called genotypes by sampling them according to the imputation probabilities stored in the BGEN files, performing a GWAS on each of these datasets, and pooling the GWAS estimates.
-In simulations, we show that $\hat{\gamma}_j^{\text{MI}} \approx \hat{\gamma}_j^{\text{imp}} \cdot \text{INFO}_j$ and $Z_j^{\text{MI}} \approx Z_j^{\text{imp}} \cdot \text{INFO}_j$, where $Z = \hat{\gamma} / \text{se}(\hat{\gamma})$ (Figure \ref{fig:compare-mi}).
+In simulations, we show that $\hat{\gamma}_j^{\text{MI}} \approx \hat{\gamma}_j^{\text{imp}} \cdot \text{INFO}_j$ and $Z_j^{\text{MI}} \approx Z_j^{\text{imp}} \cdot \text{INFO}_j$, where $Z = \hat{\gamma} / \text{se}(\hat{\gamma})$ (Figure S11).
 This is the second correction of summary statistics we implement in the simulations below, along with $n_j \cdot \text{INFO}_j$ as the new per-variant sample sizes. 
 Finally, we consider an in-between solution as a third correction, using $\hat{\gamma}_j = \hat{\gamma}_j^{\text{imp}} \cdot \text{INFO}_j$, $\text{se}(\hat{\gamma}_j) = \text{se}(\hat{\gamma}_j)^{\text{imp}} \cdot \sqrt{\text{INFO}_j}$, and $n_j \cdot \text{INFO}_j$ as the per-variant sample sizes.
-Note that we have recomputed INFO scores for the subset of 362,307 European individuals used in this paper since they can differ substantially from the ones reported by the UK Biobank for the whole data (Figures \ref{fig:compare-info} and \ref{fig:compare-maf-info}).
+Note that we have recomputed INFO scores for the subset of 362,307 European individuals used in this paper since they can differ substantially from the ones reported by the UK Biobank for the whole data (Figures S12 and S13).
 
@@
 
-When performing similar simulations with a heritability of 4\% (instead of 20\%), correction ``sqrt\_info'' still provides slightly higher predictive performance for all methods (compared to no correction), yet the other two corrections sometimes provide lower predictive performance (Figure \ref{fig:simu-info-smallh2}).
+When performing similar simulations with a heritability of 4\% (instead of 20\%), correction ``sqrt\_info'' still provides slightly higher predictive performance for all methods (compared to no correction), yet the other two corrections sometimes provide lower predictive performance (Figure S14).
 In the real data applications hereinafter, we therefore choose to use the first correction, ``sqrt\_info'', which seems to more consistently provide better results, and is simple because it is somewhat equivalent to post-processing PGS effects by multiplying them by $\sqrt{\text{INFO}}$.
@@
 In addition to the LD reference panel of Northwestern European ancestry, we design an alternative reference panel based on 10,000 individuals from South Europe by using the ``Italy'' center defined in \citet{prive2021high}.
-Allele frequencies and pairwise correlations are quite similar between these two LD reference panels (Figure \ref{fig:compare-altpop}).
+Allele frequencies and pairwise correlations are quite similar between these two LD reference panels (Figure S15).
 When using this alternative LD reference panel instead of a well-matched one as in the previous sections, squared correlations between the polygenic scores and the simulated phenotypes drop from 0.173 to 0.166 for lassosum2, from 0.168 to 0.159 for LDpred2(-grid), from 0.174 to 0.169 for LDpred2-low-h2, from 0.142 to 0.136 for LDpred2-auto, from 0.162 to 0.147 for LDpred2-auto-rob, from 0.143 to 0.138 for LDpred2-inf, and from  0.163 to 0.155 for PRS-CS-auto (Figure \ref{fig:simu-altpop}, averaging over 10 simulations).
@@
 We also compute and test using a shrunk LD matrix (from GCTB) for this alternative LD reference panel; again, using this type of LD matrix seems beneficial for LDpred2-auto.
-Finally, under the same simulation scenario, but with a smaller simulated heritability of 4\% (instead of 20\%), there remains a small drop in predictive performance when using the alternative LD reference (Figure \ref{fig:simu-altpop-smallh2}).
+Finally, under the same simulation scenario, but with a smaller simulated heritability of 4\% (instead of 20\%), there remains a small drop in predictive performance when using the alternative LD reference (Figure S16).
 
@@
 Breast cancer GWAS summary statistics are interesting because they include results from two mega analyses \cite[]{michailidou2013large,michailidou2015genome,michailidou2017association}, which means that parameters reported in these GWAS summary statistics, such as INFO scores and sample sizes, are estimated with high precision.
-Imputation INFO scores for the OncoArray summary statistics are generally very good (mean of 0.968 after having restricted to HapMap3 variants, Figure \ref{fig:hist-info-brca}) and better than the ones from iCOGS (mean of 0.841, Figure \ref{fig:hist-info-brca2}), probably because the iCOGS chip included around 200K variants only, compared to more than 500K variants for the OncoArray.
+Imputation INFO scores for the OncoArray summary statistics are generally very good (mean of 0.968 after having restricted to HapMap3 variants, Figure 
+S17) and better than the ones from iCOGS (mean of 0.841, Figure S18), probably because the iCOGS chip included around 200K variants only, compared to more than 500K variants for the OncoArray.
 For both summary statistics, we compare the standard deviations (of genotypes) inferred from the reported allele frequencies (i.e.\ $\sqrt{2 f (1 - f)}$ where $f$ is the allele frequency, and denoted as sd\_af) versus the ones inferred from the GWAS summary statistics (Equation \eqref{eq:approx-sd-log}, and denoted as sd\_ss).
-As shown in figures \ref{fig:brca_onco_qc} and \ref{fig:brca_icogs_qc}, there is a clear trend with sd\_ss being lower than sd\_af as INFO decreases; indeed, using $\text{sd\_ss} / \sqrt{\text{INFO}}$ provides a very good fit for sd\_af, except for some variants of chromosome 6 and 8 for the OncoArray summary statistics.
-Most of these outlier variants are either in region 25-33 Mbp of chromosome 6 or in 8-12 Mbp of chromosome 8 (Figure \ref{fig:hist-bad-pos}), which are two known long-range LD regions \cite[]{price2008long}.
+As shown in figures \ref{fig:brca_onco_qc} and S19, there is a clear trend with sd\_ss being lower than sd\_af as INFO decreases; indeed, using $\text{sd\_ss} / \sqrt{\text{INFO}}$ provides a very good fit for sd\_af, except for some variants of chromosome 6 and 8 for the OncoArray summary statistics.
+Most of these outlier variants are either in region 25-33 Mbp of chromosome 6 or in 8-12 Mbp of chromosome 8 (Figure S20), which are two known long-range LD regions \cite[]{price2008long}.
 We hypothesize that this is due to using principal components (PCs) that capture LD structure instead of population structure, as covariates in the GWAS \cite[]{prive2020efficient}.
@@
 PC19 from the UK Biobank was previously reported to capture LD structure in region 70-91 Mbp of chromosome 6 \cite[]{prive2020efficient}.
-In these simulations, the same bias as in figure \ref{fig:brca_onco_qc}B is observed for the variants in this region (Figure \ref{fig:gwas_bad_pc}), confirming our hypothesis.
+In these simulations, the same bias as in figure \ref{fig:brca_onco_qc}B is observed for the variants in this region (Figure S21), confirming our hypothesis.
 
@@
 As for the correction using INFO scores, we use the first correction, ``sqrt\_info'', which is simple because it is equivalent to post-processing PGS effects by multiplying them by $\sqrt{\text{INFO}}$.
-Although results for both QC used are very similar, correcting using INFO scores slightly improves predictive performance when deriving polygenic scores based on iCOGS summary statistics (Figure \ref{fig:res_brca_icogs}).
+Although results for both QC used are very similar, correcting using INFO scores slightly improves predictive performance when deriving polygenic scores based on iCOGS summary statistics (Figure S22).
 All other improvements introduced before have little to no effect here, probably because misspecifications are much smaller than in the simulations.
@@
 We use other external GWAS summary statistics for which INFO scores are reported (see Table \ref{tab:sumstats}); they all have a very high mean INFO score (larger than 0.94), except for T1D-affy (0.885).
-QC plots comparing standard deviations usually show little deviation from the identity line (after the INFO score correction), except for coronary artery disease (CAD) summary statistics (Figures \ref{fig:cad_qc}-\ref{fig:t1d_illu_qc}).
+QC plots comparing standard deviations usually show little deviation from the identity line (after the INFO score correction), except for coronary artery disease (CAD) summary statistics (Figures S23-S27).
 Note that the popular CAD summary statistics used here come from a multi-ancestry GWAS meta-analysis with more than 20\% non-European samples \cite[]{nikpay2015comprehensive,prive2021using}.
-In figures \ref{fig:res_cad} and \ref{fig:res_brca_onco}-\ref{fig:res_t1d_aff}, we then provide similar results as figure \ref{fig:res_brca_icogs} for other phenotypes.
+In figures \ref{fig:res_cad} and S28-S32, we then provide similar results as figure S22 for other phenotypes.
 For MDD and PRCA, most changes introduced before have little to no impact on predictive performance.
@@
 We also investigate results for vitamin D GWAS summary statistics, which do not report INFO scores nor allele frequencies, as opposed to previous ones.
-The QC procedure is then less precise and uses allele frequencies from the LD reference (Figure \ref{fig:vitaminD_qc}).
-However, these summary statistics do report per-variant sample sizes, which cover a wide range of different values (Figure \ref{fig:hist-N-vitaminD}).
+The QC procedure is then less precise and uses allele frequencies from the LD reference (Figure S33).
+However, these summary statistics do report per-variant sample sizes, which cover a wide range of different values (Figure S34).
 Here we compare using either the maximum sample size or the true per-variant sample sizes when deriving polygenic scores, and an additional QC step ``qc2'' which refers to removing all variants with a per-variant sample size less than 70\% of the maximum one.
@@
 When the maximum sample size is used, this ``qc2'' is very important, especially for LDpred2-auto.
-Note that, when using the maximum sample size to derive ``sd\_ss'' in ``qc1'' (not the case here; we used the per-variant sample sizes), ``qc1'' would remove variants with underestimated standard deviations due to overestimated sample sizes, which would effectively remove variants with low sample sizes (Figure \ref{fig:qcplot}), similar to the ``qc2'' used here.
+Note that, when using the maximum sample size to derive ``sd\_ss'' in ``qc1'' (not the case here; we used the per-variant sample sizes), ``qc1'' would remove variants with underestimated standard deviations due to overestimated sample sizes, which would effectively remove variants with low sample sizes (Figure S35), similar to the ``qc2'' used here.
 
@@
 We then run LDpred2-auto-rob for CAD with different shrinkage multiplicative coefficients for the off-diagonal elements of the LD matrix, from 1 (LD matrix unchanged) to 0 (using the identity matrix instead).
-When using a strong shrinkage of 0.3 or 0.4, results for LDpred2-auto-rob are similar whatever the QC used (Figure \ref{fig:res_cad_shrinkage}), as previously observed for PRS-CS-auto.
+When using a strong shrinkage of 0.3 or 0.4, results for LDpred2-auto-rob are similar whatever the QC used (Figure S36), as previously observed for PRS-CS-auto.
 Results with LDpred2-auto-rob are always best when using ``qc2'', the most stringent QC; then no shrinkage (1) seems needed, whereas a strong shrinkage make predictive performance drop.
@@
 We then compare three different LD reference panels to use with these GWAS summary statistics (Methods section \ref{sec:altLD}).
-Interestingly, when looking at height and HDL cholesterol, where we could get the best predictive performance, using the smaller Japanese LD reference we have defined here seems to provide more predictive PGS than using the one using a larger set of East Asian individuals from the UKBB or the widely used East Asian subset of the 1000 Genomes (1000G) data (Figure \ref{fig:bbj}), except when using PRS-CS-auto to predict height.
+Interestingly, when looking at height and HDL cholesterol, where we could get the best predictive performance, using the smaller Japanese LD reference we have defined here seems to provide more predictive PGS than using the one using a larger set of East Asian individuals from the UKBB or the widely used East Asian subset of the 1000 Genomes (1000G) data (Figure S37), except when using PRS-CS-auto to predict height.
 Note that the PGS here are validated in people of broad East Asian ancestry in the UK biobank and for continuous outcomes to obtain moderate sample sizes for comparison.
@@
 		\hline
-		Using imputed dosages for LD computation &  & None required; LD seems correctly estimated (Figure \ref{fig:compare-cor}) \\
+		Using imputed dosages for LD computation &  & None required; LD seems correctly estimated (Figure S38) \\
 		\hline 
@@
 		\hline
-		Rounding of summary statistics (e.g.\ effect sizes and SEs) & Add unnecessary noise to the data & None, but this has almost no impact on the predictive performance (Figure \ref{fig:simu-rounded}) \\
+		Rounding of summary statistics (e.g.\ effect sizes and SEs) & Add unnecessary noise to the data & None, but this has almost no impact on the predictive performance (Figure S39) \\
 		\hline 

### 2022-08-05 paper/paper-misspec.tex
typo
add title to fig

@@
 Breast cancer GWAS summary statistics are interesting because they include results from two mega analyses \cite[]{michailidou2013large,michailidou2015genome,michailidou2017association}, which means that parameters reported in these GWAS summary statistics, such as INFO scores and sample sizes, are estimated with high precision.
-Imputation INFO scores for the OncoArray summary statistics are generally very good (mean of 0.968 after having restricted to HapMap3 variants, Figure 
-S17) and better than the ones from iCOGS (mean of 0.841, Figure S18), probably because the iCOGS chip included around 200K variants only, compared to more than 500K variants for the OncoArray.
+Imputation INFO scores for the OncoArray summary statistics are generally very good (mean of 0.968 after having restricted to HapMap3 variants, Figure S17) and better than the ones from iCOGS (mean of 0.841, Figure S18), probably because the iCOGS chip included around 200K variants only, compared to more than 500K variants for the OncoArray.
 For both summary statistics, we compare the standard deviations (of genotypes) inferred from the reported allele frequencies (i.e.\ $\sqrt{2 f (1 - f)}$ where $f$ is the allele frequency, and denoted as sd\_af) versus the ones inferred from the GWAS summary statistics (Equation \eqref{eq:approx-sd-log}, and denoted as sd\_ss).
@@
 	\centerline{\includegraphics[width=0.95\textwidth]{brca_onco_qc}}
-	\caption{Standard deviations inferred from the OncoArray breast cancer GWAS summary statistics using equation \eqref{eq:approx-sd-log} (\textbf{A:} Raw or \textbf{B:} dividing them by $\sqrt{\text{INFO}}$) versus the ones inferred from the reported GWAS allele frequencies $f_j$ (using $\sqrt{2 f_j (1 - f_j)}$).
+	\caption{Comparison of standard deviations for quality control of GWAS summary statistics.
+Standard deviations inferred from the OncoArray breast cancer GWAS summary statistics using equation \eqref{eq:approx-sd-log} (\textbf{A:} Raw or \textbf{B:} dividing them by $\sqrt{\text{INFO}}$) versus the ones inferred from the reported GWAS allele frequencies $f_j$ (using $\sqrt{2 f_j (1 - f_j)}$).
 Only 100,000 HapMap3 variants are represented, at random.}
@@
 
-\clearpage
+%\clearpage
 %\vspace*{3em}
@@
 %\clearpage
-\vspace*{3em}
+%\vspace*{3em}
 
